### 2018-04-30 model/thesis_draft/Draft.tex

\documentclass[a4paper,12pt]{article}
\usepackage{enumerate}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}

%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=1.5in,right=1in,marginparwidth=1.75cm]{geometry}

%% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{dsfont}
\usepackage[document]{ragged2e}
\usepackage{titlesec}
\usepackage{mdframed}
\usepackage{amssymb}

\usepackage[justification=centering,format=plain,font=it]{caption}
\usepackage{minted}
\usepackage{setspace}

\usetikzlibrary{shapes}

%\newcommand{\sectionbreak}{\clearpage}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{invariant}[theorem]{Invariant}

\setlength{\parskip}{1em}
%\titleformat{\section}[block]{\color{blue}\Large\bfseries\filcenter}{}{0.5em}{}

\title{Functionally Verified B+ Trees}
\author{Brian McSwiggen\\
        Advised by Professor Andrew Appel}

\begin{document}
\maketitle

\clearpage

\doublespacing

\section{Abstract}

In this thesis, I make progress toward a fully verified functional B+\,tree program in Coq. A verified functional B+\,tree would serve as the basis for the future verification of an imperative B+\,tree program through proving equivalence, which would have applications in databases and other systems. The primary focus of the properties and abstractions in this thesis is to examine and describe the theory of the B+\,tree cursor, which is central to the operations of a B+\,tree both functionally and imperatively. The functional implementation presented here covers the essential cursor operations as well as B+\,tree lookups and insertions, but does not include a delete function. The progress made toward verification includes the development of a complete abstract specification, the structural correctness of cursor creation and positioning, and the relation of cursors and B+\,tree operations to an element-list abstract cursor representation.

\clearpage

\section{Introduction and Motivation}

B+\,trees are an extension of the more commonly known B-trees, which are N-ary trees often used to organize data in file systems. The property that makes B+\,trees unique is that the \textit{values}, that is the data actually being held in the tree, are kept exclusively at the leaves. The interior nodes hold only keys and pointers to their children.

One interesting consequence of this property is that B+\,trees in practice can be augmented with a linked list between the leaf nodes connecting all of the values. This makes finding a value stored in the tree efficient (due to the tree structure) while still keeping range queries and iteration over the values easy (due to the linked list). Additionally, because B+\,trees don’t have to store as much data in the interior nodes, a single node can fit more pointers, meaning that the fanout can be larger and the tree needs fewer levels, so data can be accessed more quickly \cite{elmasri_navathe_2011}. A well-designed B+\,tree will store enough values per leaf that a single leaf node just fits into a single storage block, allowing the maximum possible amount of data retrieval with the fewest possible I/O operations.

Because this is a data structure commonly used by databases and file systems \cite{elmasri_navathe_2011}, it is a critical component underlying a great portion of the systems and software we rely on every day. On the one hand, that means that B+\,tree implementations used in practice are likely to be very well tested; on the other hand, bugs or failures have the potential to be wide-reaching and catastrophic.

That risk of failure is why rigorous testing practices are widely recognized as crucial to making good software that works correctly. However, in the most critical cases, more certainty may be necessary. The development of “formal methods” has provided an option for a greater degree of confidence in a range of critical applications. Formal methods encompass a range of approaches to applying strict mathematical rigor to software and hardware design, and “formal verification” in particular is the practice of rigorously proving algorithmic correctness. Because of the high certainty attained through formal methods, they have been applied to three categories of software: safety-critical, security-critical, and systems-critical programs.

Safety-critical software (or hardware) is a category of systems with direct impact on human life and safety. For example, if an airplane processor fails, if a nuclear reactor control has a bug, or if a defibrillator misfires, the consequences could very likely include death or serious injury. For that reason, software development for these systems has sometimes turned to formal methods to ensure correctness. In one example from as far back as the 1970s, NASA commissioned an aircraft control computer called SIFT which was subject to very strict failure-rate limits. Formal verification was used to ensure that the SIFT software for using redundant systems to detect and recover from hardware failures would work correctly \cite{225554}.

Security-critical software, while it does not always carry the same risk to human life, does carry the risk of compromising crucial cryptographic tools, data privacy and integrity, and more. Much such software is incredibly important to modern technology, but is also very intricate, hard to implement correctly, and hard to keep secure. Formal verification provides a means of reasoning about these tricky tools. In 2015, the Verified Software Toolchain group completed a full verification of an OpenSSL HMAC implementation, meaning that as long as SHA-256 is in fact cryptographically secure (which remains unproven), the HMAC is as well \cite{190894}.

Systems-critical software carries a great amount of risk because many other systems and pieces of software rely on it. While it may not itself carry risk to human life, or to cryptographic security, if any safety-critical or security-critical software runs on a system then it carries that risk indirectly. This has led a push to develop fully verified system stacks, such as the CLInc stack of compiler, assembler, kernel, and microprocessor, which were all verified correct in the 1980s \cite{225554}. A more relevant example to this project is the Verified Software Toolchain developed by Andrew Appel at Princeton, which also includes the CompCert verified C-language compiler developed by Xavier Leroy at INRIA \cite{Leroy:2009:FVR:1538788.1538814}.

Because of the many systems that directly or indirectly depend on B+\,trees, the verification of such an implementation falls under the systems-critical category. Formally proving a B+\,tree implementation correct increases our trust in the storage systems we use underneath any other piece of software, including any safety- or security-critical applications that may be built on top.

\clearpage

\section{Previous Work}

Formal verification is a rich field of study, and a large body of research has focused both on verifying software and on tools to make the verification of software easier. Search trees, as an important and complex class of data structures, have been the subject of previous verification. Recent work at Princeton and other universities has focused on the science and tools of specification and verification.

Various interesting search tree variants, in both functional\footnote{\textit{Functional} programming refers to a programming paradigm by which execution is the evaluation of a mathematical function, and all data are immutable (i.e. there is no program state). Well-known functional programming languages include Haskell, Lisp, and OCaml.} and imperative\footnote{\textit{Imperative} programming refers to a programming paradigm by which state-based programs are built up from individual statements, or commands. Well-known imperative programming languages include C, Java, and Python.} languages, have been the subject of verification work. An educational specification and proof of basic binary search trees can be found in Verified Functional Algorithms \cite{appel} by Professor Andrew Appel of Princeton University, who also published an efficient verified functional implementation of Red-Black trees in 2011 \cite{appel_2011}. Appel’s paper in turn builds upon the work of Filliâtre and Letouzey in 2004, in which they verified functional implementations of finite sets in three ways: using ordered lists, AVL trees, and Red-Black trees \cite{filliatre:hal-00150913}. This paper was a powerful demonstration of the use of Coq modules to contain both definitions and proven properties about those definitions, and this approach has been used here as well in the abstract specification.

Functional program verification such as the work done by Appel, Filliâtre, and Letouzey serves a practical purpose both because it can help find bugs in current functional libraries (as Filliâtre and Letouzey did with the OCaml Set module based on AVL trees \cite{filliatre:hal-00150913}) and because it can be extracted from Coq into live OCaml code and used in practice. However, many practical applications rely on imperative software. As an example in the domain of search trees, Xiwen Chen of York University provided a verified concurrent imperative binary search tree which he had proven linearizable, i.e. that any operation performs correctly as if it executed immediately \cite{chen_ruppert_breugel_2013}. Although it is more difficult to reason about imperative programs than about purely functional ones, doing imperative verifications is crucial for practical application of formal verification.

There are many approaches and tools which have been developed and can be used to reason about imperative software. Although this thesis focuses on the verification of a purely functional program, such a functional verification can serve as a crucial intermediate step in proving properties of an imperative program. The functional B+\,trees discussed here will be used to ultimately prove correctness of a B+\,tree implementation written in C. The tools and techniques that will be used for the \textit{imperative} proof provide a framework for what a useful \textit{functional} B+\,tree verification is, and are described in the following sections.

\subsection{Separation Logic}

A group at Princeton University and the National University of Singapore very recently included a demonstration of a verified binary search tree as part of a paper on separation logic and the “magic wand” connective (separating implication) \cite{cao_wang_hobor_appel_2017}. \textit{Separation logic} is a set of logic rules extending Hoare logic that facilitate proofs about variably sized imperative data structures (such as arrays or trees) by allowing the program state to be separated into disjoint parts about which properties can be proven \cite{1029817}.

Separation logic is crucial because having to reason about the entirety of the program state together quickly gets very complicated. On top of the desired properties, there have to be assertions for exactly what parts of the program state any given command affects, and assertions that no other part of the program is affected, and assertions that any two portions of a program intended to be affected differently are in fact disjoint \cite{1029817}. It is easy to see how the whole structure does not scale well within Hoare logic, which is why separation logic is so important for the verification of many imperative programs.

\subsection{Verified Software Toolchain}

\begin{figure}[h]
    \begin{center}
		\includegraphics[]{b}
        \caption{Diagram of VST components \protect\cite{vst_2017}}
    \end{center}
\end{figure}

Separation logic is crucial not just to a proof about a particular program, such as our B+\,trees implementation, but also to the entire verification stack that it is based on. Since the mid-2000s, a group at Princeton University led by Professor Andrew Appel has been developing the \textit{Verified Software Toolchain} \cite{vst_2017}, a stack of verified components that allow proofs to be written about source programs and apply to the compiled program as well. Each component of the toolchain, from the Verifiable C program logic to Xavier Leroy’s CompCert C compiler to the machine language operational semantics (and all of the interfaces between them), is designed for verification and proven correct \cite{10.1007/978-3-642-19718-5_1}.

Although the scope of this thesis does not extend into connecting the B+\,tree specification to an imperative C program that it represents, the functional specification produced could be useful for the verification of a C program. The strategy of functional specification used in this thesis is intended to work with the Verified Software Toolchain, so that a C program could be proven via the VST separation logic to be equivalent to this functional specification.

\subsection{DeepSpec}

The Verified Software Toolchain itself is part of a larger research group called DeepSpec. Since 2016, the DeepSpec group at Princeton, University of Pennsylvania, Yale, and MIT have been collaborating under an NSF Expedition in Computing grant to explore the science of deep specifications. This thesis is in part an attempt to extend that work by applying deep specification to a B+\,tree database program. Deep specifications, as described by the DeepSpec group, are program specifications with four special properties \cite{deepspec}:

\begin{enumerate}
\item They are \textbf{rich}, that is, they can describe the complex behavior of real programs.

\item They are \textbf{two-sided}. Specifications function analogously to interfaces in modular programming: they describe the boundary between two independent parts. A two-sided specification is one that is exercised from both sides of the boundary, i.e. it matches both the program it describes and the programs that use it.

\item They are \textbf{formal}, meaning they are written in exact mathematical terms. Specifications that are provided only as vague or intuitive definitions may have holes or errors that are hard to spot. Formal specifications can be machine-checked and tested.

\item They are \textbf{live}, that is, they are connected directly to the implementation and the client code. If either changes, then the machine-checked proofs may no longer pass; this ensures that when the proofs are verified we can be sure they describe the actual program.
\end{enumerate}

In the course of their work, the DeepSpec group and their collaborators are developing an extensive network of systems-related programs connected at deep specification interfaces. The goal is to be able to construct a verified stack from the operating system and applications like the B+\,tree database all the way down to the machine language and transistors at the core of the computer.

The previously mentioned Verified Software Toolchain being developed by Appel is one part of this stack. It also includes a verified LLVM (VeLLVM) developed by Steve Zcancewic at UPenn, a verified operating system (CertiKOS) developed by Zhong Shao at Yale, and many other tools, systems, and applications.

\clearpage

\section{Project Approach and Scope}

The strategy that this thesis targets for verifying imperative software is through a \textit{functional model}. A functional model describes the actual behavior of the operations, and in this case the model is an equivalent piece of software written in a functional programming language, but other functional models are possible. When reasoning about a programming language, for example, the functional model might be the small-step operational semantics of the language. For the purposes of this thesis, since functional programs are much easier to formally reason about, the functional model provides a bridge: we can prove the equivalence of the functional program and the imperative program, and then any proofs of correctness for the functional program are also valid proofs of the imperative one.

One step beyond the functional model is the \textit{abstract specification}. This describes, in the most pure and abstract terms possible, the goal for the program. The abstract specification defines what it \textit{means} to be a “correct implementation” of the desired program. Proving these properties of the functional model proves that the model is correct, and therefore any imperative program equivalent to the model is also correct.

The scope of this thesis covers a B+\,tree abstract specification, a functional implementation, and a partially formalized proof of the implementation’s correctness relative to the specification. Once completely formally verified, the functional program could be used as the functional model for an imperative program, but that step is not covered in this work.

The particular style of B+\,tree and set of operations which are targeted in this work are derived from a subset of the SQLite commands. SQLite uses B+\,trees to store information in relational databases. One particularity of SQLite is that it represents a cursor into the B+\,tree as a list of pointers to the split point at each level. This makes it easier to change the position of the cursor without needing to have a linked list at the leaves as mentioned above or having to retraverse the tree.

\clearpage

\section{B+ Tree Details} \label{sec:b+tree}

A B+\,tree is a type of ordered tree with three important properties:

\begin{enumerate}
\item Any B+\,tree has a \textit{max fanout} value $b$ which represents the maximum number of children that any node can hold. Every node in the B+\,tree except the root must have between $\lceil \frac{b}{2} \rceil$ and $b$ children; the root may have as few as two (if it is an interior node) or one (if it is a leaf node) \cite{elmasri_navathe_2011}.

\item All \textit{values} are stored in leaf nodes. Interior nodes only store keys and pointers to the child nodes. This means that an interior node with $n$ children only needs to store $n-1$ keys, and also that keys may repeat in the tree \cite{elmasri_navathe_2011}.

\item As new key-value pairs are inserted, the tree only grows up at the root, not down at the leaves: when a node gets too big and splits, both halves become children of the original parent node, and no new levels are added to the tree. The only way for a new level to be added is for the root node to split and create a new root node above it. Similarly, as key-value pairs are deleted, the tree shrinks at the root rather than at the leaves. This ensures that the whole tree always stays balanced, that is, the distance from any leaf to the root is the same as the distance from any other leaf to the root \cite{elmasri_navathe_2011}.
\end{enumerate}

An interior node in a B+\,tree might look like the following:

%Maybe come back and clean this up later%
\begin{table}[h]
\centering
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
$P_1$ & $K_1$ & $P_2$ & $K_2$ & ... & $K_{n-1}$ & $P_n$\\
\hline
\end{tabular}
\caption{B+\,tree interior node with $n-1$ keys and $n$ subtrees}
\label{tab:interior-node}
\end{table}

Where each $P_i$ is a pointer to a child node, and each $K_i$ is a key. The fanout $n$ is constrained by $\lceil \frac{b}{2} \rceil \leq n \leq b$, the keys $K$ must be sorted ($ \forall i, j: \ K_i < K_j$), and each key $K$ constrains the values in the child nodes to either side: for every key $X$ in the node at $P_i, K_{i-1} \leq X < K_i$. Of course, for the end pointers, only one side of that inequality will hold \cite{elmasri_navathe_2011}.

A leaf node holds values rather than child pointers, and must have exactly one key per value. Therefore, a leaf node might look like the following:

%Maybe come back and clean this up later%
\begin{table}[h]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
$(K_1,V_1)$ & $(K_2,V_2)$ & ... & $(K_n,V_n)$\\
\hline
\end{tabular}
\caption{B+\,tree leaf node with $n$ keys and $n$ values}
\label{tab:leaf-node}
\end{table}

In this case, each $V_i$ is some value stored by the B+\,tree (often, but not necessarily, a pointer to a block of data in memory), and each $K_i$ is once again a key. The keys must be sorted just as in the interior nodes, and it still holds that the fanout n is constrained by $\lceil \frac{b}{2} \rceil \leq n \leq b$ \cite{elmasri_navathe_2011}.

With this structure, we now consider how to do database operations on this structure. In particular, we describe Lookup (finding a key/value in a b+tree), Insert (adding a key/value pair), and Delete (removing a key/value pair).

For all three operations, we will consider the following b+tree for examples:

%Find a way to position better?%
%Find a way to italicize captions?%
\begin{figure}[h]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=40mm]
    \tikzstyle{level 2}=[sibling distance=15mm]
    \node {7} [->]
      child {node {3}
        child {node {1 \nodepart{two} 2}}
        child {node {3 \nodepart{two} 4 \nodepart{three} 6}}    
      } 
      child {node {9 \nodepart{two} 11}
        child {node {7 \nodepart{two} 8}}
        child {node {9 \nodepart{two} 10}}   
        child {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{A simple B+\,tree of height 3. Each interior node has 2 or 3 children.}
    \label{fig:demotree1}
\end{figure}

Notice that all keys and subtrees are properly ordered, that all leaves are at depth 2 (where the root is depth 0), and that keys may appear multiple times within the tree. The fanout is $b=3$, meaning each node may have two or three children (this simple example is actually a variant of a 2-3 tree). For these examples, the values at the leaves have been left out, since they are irrelevant to the tree operations.

\subsection{Lookup}

Finding a key-value pair in the tree proceeds from the root. At each level, we search through the list of keys to find the split where the value we want falls. Since every node has a sorted list of keys, we can do binary search within a list if desired. However, the number of children is often small enough that a simpler linear search works as well, especially when the fanout is optimized for the size of CPU cache lines, rather than full disk pages.

If we were to search for 10 in the example tree, we would first note that $7 < 10$ and go to the second child of the root. We would then note that $9 \leq 10 < 11$, and go to the middle child, where we would find 10.

\begin{figure}[h]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=40mm]
    \tikzstyle{level 2}=[sibling distance=15mm]
    \node {7} [->]
      child {node {3}
        child {node {1 \nodepart{two} 2}}
        child {node {3 \nodepart{two} 4 \nodepart{three} 6}}    
      } 
      child[ultra thick,red] {node {9 \nodepart{two} 11}
        child[thin, black] {node {7 \nodepart{two} 8}}
        child[ultra thick, red] {node {9 \nodepart{two} 10}}   
        child[thin, black] {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{To find 10, we search for 10 at each level and recursively descend.}
    \label{fig:demotree2}
\end{figure}

\subsection{Insert}

Insertion is simple if the leaf node has less than $b$ children: we simply find the correct place according to the list of keys, and add the value there. However, we must maintain that any given node has at most b children. Therefore, if a key is inserted which would go into a leaf that has b children already, then we have to split that node and add the middle key to the node above \cite{elmasri_navathe_2011}.

If that node is also full, then we have to split it in turn, and so on up to the root where, if necessary, the root will split and the tree will grow by one level \cite{elmasri_navathe_2011}.

For a leaf node, since every key must appear mapped to its value, we keep the middle key in the leaf as well. However, for any interior node that splits, the middle key which goes up to the next level is removed from the level that split \cite{elmasri_navathe_2011}.

If we were to insert the key-value pair $(5,V_5)$ in the example tree, we would have to first find where in a leaf node 5 should go (between 4 and 6, on the left). Then, seeing that the leaf would have 4 key-value pairs, we split it in two and copy 5 up to the next level.

%Could use some work%
\begin{figure}[h]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=50mm]
    \tikzstyle{level 2}=[sibling distance=15mm]
    \node {7} [->]
      child[ultra thick,red] {node {3 \nodepart{two} 5}
        child[thin, black] {node {1 \nodepart{two} 2}}
        child[ultra thick, red] {node {3 \nodepart{two} 4}}
        child[ultra thick, red] {node {5 \nodepart{two} 6}}    
      } 
      child {node {9 \nodepart{two} 11}
        child {node {7 \nodepart{two} 8}}
        child {node {9 \nodepart{two} 10}}   
        child {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{To add a mapping for 5, we have to first find where 5 goes in the tree, then add the new key-value pair and split nodes as necessary.}
    \label{fig:demotree3}
\end{figure}

\subsection{Delete}

Deletion works intuitively as the inverse of insertion. It is simple if the leaf node has at least $\lceil \frac{b}{2} \rceil + 1$ entries; then we can just remove the entry and leave the rest of the tree as it is. However, if the leaf node would be left with less than $\lceil \frac{b}{2} \rceil$ entries after deletion, then we need to redistribute entries in order to maintain the fanout. That can happen by moving over entries from adjacent nodes (and changing the keys in the parent node), or by merging a node with adjacent nodes to reduce the total number of children. In the latter case, this reduction in the number of children may cause the parent to have to redistribute or merge as well, potentially propagating up to the root and decreasing the depth of the tree by one \cite{elmasri_navathe_2011}.

If we were to delete the key 9 from our example tree, we would see that its leaf now has only 1 entry. Since both adjacent nodes are too small to redistribute, we instead have to merge them.

\begin{figure}[h]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=40mm]
    \tikzstyle{level 2}=[sibling distance=20mm]
    \node {7} [->]
      child {node {3}
        child {node {1 \nodepart{two} 2}}
        child {node {3 \nodepart{two} 4 \nodepart{three} 6}}    
      } 
      child[ultra thick,red] {node {11}
        child[ultra thick, red] {node {7 \nodepart{two} 8 \nodepart{three} 10}}
        child[thin, black] {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{When we delete 9, we may need to merge nodes together.}
    \label{fig:demotree4}
\end{figure}

\clearpage

\section{Abstract Specification} \label{sec:spec}

Our abstract specification of a B+\,tree is that of a lookup table. That is, it stores data indexed by keys, and allows that data to be arbitrarily fetched, changed, added to, or removed. However, we add one more layer to this particular table abstraction. B+\,trees rely on a cursor, which could be any way of pointing to a particular place in the table. We build this notion into our abstract specification, allowing us to prove correctness relative to this abstract notion of a cursor in addition to a table. Here, the specification is shown as written in Coq.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
Module Type CURSOR_TABLE.
 Parameter V: Type.
 Definition key := Z.
 Parameter table: Type.
 Parameter cursor: Type.
 Parameter empty_t: table.
\end{minted}
\end{singlespace}
\end{figure}

We parametrize our abstract cursor-table with a number of basic types. Any cursor-table must store a set of values of type \texttt{V}. It stores these within a table of type \texttt{table} (\texttt{empty\_t} being the most basic such table, storing no values) and accesses them using a cursor of type \texttt{cursor}. Any implementation of a cursor-table must provide these basic parameters, but the exact types are up to the specific implementation---this makes the module type abstract.

Notice that any cursor-table must also have a \texttt{key} type which is mapped to values. For our purposes we have hard-coded this as Coq’s built-in \texttt{Z} type for integers. Perhaps it would be better to abstract even further by allowing the key to be any type that has a compare function and satisfies some particular set of axioms; this was in fact the approach taken by Filliâtre and Letouzey in their verification of finite set implementations \cite{filliatre:hal-00150913}. However, the simplifying assumption that the keys will be integers is not an unreasonable one, and in particular it fits well with the imperative implementation that this functional specification could be used to verify.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
 Parameter make_cursor: key -> table -> cursor.
 Parameter get_table: cursor -> table.
 Parameter get_key: cursor -> option key.
 Parameter get: cursor -> option V.
 Parameter insert: cursor -> key -> V -> cursor.
 Parameter next: cursor -> cursor.
 Parameter prev: cursor -> cursor.
 Parameter first_cursor: table -> cursor.
 Parameter last_cursor: table -> cursor.
\end{minted}
\end{singlespace}
\end{figure}

Any cursor-table must also define a set of operations on the table. Again, the specific implementations are left abstract: they are constrained only by their type signatures and the required axioms about them discussed below.

The most central of these functions is the \texttt{make\_cursor} operation, which takes any key and a table and creates a cursor at the appropriate location for that key in the table. Such a cursor is then used as input to most other operations. The reverse of \texttt{make\_cursor} is \texttt{get\_table}, which translate from a cursor back into the table it points into. We also require two shortcuts for creating cursors at the extremes of a table, these are \texttt{first\_cursor} and \texttt{last\_cursor} and are important in the definition of later axioms.

Once we have set a cursor, a cursor-table should be able to use it to do any of the desired operations: \texttt{get} to retrieve the value stored at the cursor, \texttt{get\_key} to retrieve the key mapped to that value, and \texttt{insert} to update the tree with a new key-value pair. Deletion is not included as part of this specification, but could be axiomatized similarly to insertion.

The cursor should also be movable from its position: \texttt{next} should move it to the next key greater than the cursor’s current position, while \texttt{prev} should move it to the last key lesser than its current position.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
 Parameter abs_rel: table -> cursor -> Prop.
 Parameter key_rel: key -> cursor -> Prop.
 Parameter eq_cursor : cursor -> cursor -> Prop.
 Parameter cursor_correct: cursor -> Prop.
 Parameter table_correct: table -> Prop.
\end{minted}
\end{singlespace}
\end{figure}

We need various propositions that can describe the correctness and relationship of tables, cursors, and keys as well. Specifically, we need four propositions. The first, \texttt{abs\_rel}, defines what it means for a particular cursor to be a reference into a particular table, i.e. for the table and cursor to represent the same abstract relation. The second, \texttt{key\_rel}, defines what it means for a cursor to be positioned for a particular key $k$, i.e. that the key to the right of the cursor is greater than $k$ and the key to the left is at most $k$. The third, 
\texttt{eq\_cursor}, is a predicate for two cursors being abstractly equivalent even if they are not structurally equal. The fourth and fifth describe what it means for a cursor or table to be correct: since we don’t require every instance of the \texttt{cursor} and \texttt{table} types to be correct, it is possible to have a cursor or table that does not actually correspond to a valid instance of a \texttt{CURSOR\_TABLE}. For example, with B+\,trees it is possible to construct an unbalanced or unsorted tree, in which case the operations are not guaranteed to be correct. It would have been possible to make a more complicated B+\,tree type than the one described later in section \ref{sec:typedef} and enforce the correctness properties on every tree instance. However, this would have made working with that type more difficult.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
 Axiom make_cursor_rel: forall t k,
   abs_rel t (make_cursor k t).
 Axiom get_table_rel: forall t c,
   abs_rel t c <-> get_table c = t.
 Axiom first_rel: forall t,
   abs_rel t (first_cursor t).
 Axiom last_rel: forall t,
   abs_rel t (last_cursor t).
 Axiom next_rel: forall t c,
   abs_rel t c -> abs_rel t (next c).
 Axiom prev_rel: forall t c,
   abs_rel t c -> abs_rel t (prev c).
 Axiom correct_rel: forall t c,
   abs_rel t c -> (cursor_correct c <-> table_correct t).
\end{minted}
\end{singlespace}
\end{figure}

Given the above relations and propositions, we now want to ensure they satisfy certain desirable properties. \texttt{make\_cursor} and \texttt{get\_table} should function correctly relative to our table-cursor relation, as should \texttt{first\_cursor} and \texttt{last\_cursor}. Cursor repositioning (\texttt{next} and \texttt{prev}) should stay within the same table. Finally, correctness should translate between cursors and tables---if a cursor is correct, then so should its corresponding table be correct; and for any correct table, all cursors into that table should be correct.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
 Axiom insert_correct: forall k v c,
   cursor_correct c -> key_rel k c -> cursor_correct (insert c k v).
\end{minted}
\end{singlespace}
\end{figure}

Inserting a key-value pair into a correct cursor should in general return a new correct cursor. As a consequence of this, by \texttt{correct\_rel} and \texttt{get\_table\_rel} above, it also implies a new table (that is, \texttt{get\_table} of the new cursor) that is itself correct. However, notice that we add an additional restriction: the new cursor only has to be correct if it satisfies \texttt{key\_rel k c}, i.e. it was positioned properly for the key $k$ that was inserted. We don’t actually specify what should happen when inserting a key-value pair into a cursor in the wrong position. Depending on the implementation, it could work correctly, return the original cursor back unchanged (as our implementation does), or modify the cursor completely incorrectly. It doesn’t ultimately matter---we only require the behavior to be correct when the cursor is positioned correctly.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
 Axiom glast: forall t,
   get (last_cursor t) = None.
 Axiom gis: forall k v c,
   cursor_correct c -> key_rel k c ->
   get (make_cursor k (get_table (insert c k v))) = Some v.
 Axiom gio: forall j k v c,
   cursor_correct c -> key_rel k c -> ~ key_rel j c ->
   get (make_cursor j (get_table (insert c k v))) =
   get (make_cursor j (get_table c)).
\end{minted}
\end{singlespace}
\end{figure}

The next three axioms cover how \texttt{get} should behave. Specifically, we build up the behavior of \texttt{get} through modifications to the table. This reflects the idea that the behavior of \texttt{get} for a certain key should only be affected by certain modifications, and should ignore others. If we look up a particular key $k$ for which we have inserted some value $v$, we should get back that value $v$. On the other hand, if we look up a key $j$ into a table where we last inserted $k$, and $j$ and $k$ are not in the same range for the cursor $c$, then the behavior of \texttt{get} should be as if $k$ had never been inserted.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
 Axiom next_prev: forall c t,
   abs_rel t c -> ~ (c = last_cursor t) ->
   eq_cursor c (prev (next c)).
 Axiom prev_next: forall c t,
   abs_rel t c -> ~ (c = first_cursor t) ->
   eq_cursor c (next (prev c)).
 Axiom cursor_order: forall c k1 k2,
   cursor_correct c -> get_key c = Some k1 ->
   get_key (next c) = Some k2 -> lt_key k1 k2 = true.
End CURSOR_TABLE.
\end{minted}
\end{singlespace}
\end{figure}

Finally, we want to ensure that for any valid cursor except the last one, moving to the next position and then back to the first gives us back a cursor equivalent to the original; this is \texttt{next\_prev}. In the reverse direction, for any valid cursor except the first one, moving to the previous position and back gives us an equivalent cursor to the original cursor; this is \texttt{prev\_next}. Finally, we expect \texttt{next} and \texttt{prev} to position the cursor relative to the ordering of the table. If we call \texttt{next}, we expect the “next” key that it moves to to be the key immediately greater than the one the cursor was previously on.

Note that while all of these axioms do specify \textit{what} the cursor-table should do, they don’t specify \textit{how} it should do it. We implement this specification with B+\,trees, but it could as well be implemented with a simple binary tree, a linked list, an array, or many other data structures. That level of abstraction is crucial: we simplify our correctness properties to only the general results that we want to prove.

\subsection{A Small Example} \label{subsec:ex}

To demonstrate the application of the abstract specification as described above, consider a small example. Suppose we have a small relation built up by inserting $(3,V_3)$, then $(1,V_1)$, and finally $(2,V_2)$. Functionally, this might look like the following:

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
insert (insert (insert (make_cursor 0 empty_t) 3 V_3) 1 V_1) 2 V_2
\end{minted}
\end{singlespace}
\end{figure}

Let the cursor returned by the final insert be $c$. If we now try to retrieve the value for key $1$ from $c$, we want to be able to prove that we get $V_1$ back. This might look like the following:

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
get (make_cursor 1 (get_table c))
\end{minted}
\end{singlespace}
\end{figure}

If we peel off the outer \texttt{insert} of $c$, we see that we can apply \texttt{gio}:

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
get (make_cursor 1 (get_table (insert c 2 V_2))) =
get (make_cursor 1 (get_table c))
\end{minted}
\end{singlespace}
\end{figure}

Peeling off the next \texttt{insert}, we can now apply \texttt{gis} and prove that we return $V_1$:

%For some reason the figure is making this mess up??
%\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
get (make_cursor 1 (get_table (insert c 1 V_1))) = Some V_1
\end{minted}
\end{singlespace}
%\end{figure}

In other words, without any knowledge of the specific implementation of any of the operations, and using only the theorems defined by the abstract specification, we have proved that the value returned by get is the intended value placed into the table.

\clearpage

\section{B+ Tree Type Specification} \label{sec:typedef}

The first step of writing an effective functional model is determining how to translate the imperative concepts of the B+\,tree data structure into functional terms. Pointers and mutable data have to be turned into immutable data that can be recursively destructed. In making decisions about how to design the functional type, we have to achieve two goals: first, creating a simple type that will be easy to prove correctness properties of, and second, creating a type that will be easy to prove equivalent to the imperative program.

\subsection{Tree Data Structure Type}

We represent a B+\,tree as a mutually inductive type of \texttt{tree} and \texttt{treelist}.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
Inductive tree : Type :=
 | node : key -> treelist -> tree
 | final : treelist -> tree
 | val : key -> V -> tree
with treelist : Type :=
 | tl_nil : treelist
 | tl_cons : tree -> treelist -> treelist
\end{minted}
\end{singlespace}
\end{figure}

Intuitively, \texttt{tree} holds the key-subtree mapping: each \textit{element} within a node is a tree. The \texttt{final} tree type holds the single subtree not paired with a key, since interior nodes have $n+1$ subtrees for $n$ keys. Then, an entire node is a \texttt{treelist}. An immediate similarity is apparent between the \texttt{treelist} type and Coq’s standard list type. This makes it convenient to destruct a treelist to iterate over it.

It also makes the single treelist type work for both interior nodes (which map keys to subtrees without values, and use the \texttt{node} and \texttt{final} tree types) and leaf nodes (which map keys to values, and use the \texttt{val} tree type). Since both of those nodes can be expressed by the single type \texttt{tree}, a treelist can hold either one. This in turn makes the structure of the B+\,tree simpler, but at the cost of added complexity and assumptions when proving properties of a tree. The type itself does not enforce the structural property that a node is \textit{either} a leaf node with \texttt{val}s or an interior node with \texttt{node}s and exactly one \texttt{final}, so we have to assume or prove this whenever needed.

The choice of putting a final tree at the end, rather than having an initial keyless subtree at the beginning, is somewhat out of sync with the imperative implementation. However, it makes both correctness proofs and the functional implementation much simpler.

Imagine that we had an initial keyless subtree at the beginning. We want to say that a treelist is structurally correct if it has exactly one keyless subtree. Once we destruct a correct treelist and remove that initial subtree, the remainder suddenly only has nodes, and is no longer structurally correct! By placing final at the end instead, a destructed treelist is still structurally correct.

Additionally, putting final at the end makes searching for a particular key (e.g. when making a cursor) easier. At each step, we need to compare the key we are looking for against the key for this subtree. With an initial keyless subtree, we would have to look at the key held in the \textit{next} tree at each step. With a final at the end, we only have to compare against the key held in the current node; once we get to final we know that any remaining keys have to be there.

When relating this representation to the imperative representation, it suffices to shift subtrees by one. In other words, the imperative implementation’s keyless “first pointer” is related to the first key’s subtree in the functional implementation; the imperative first key’s subtree is the functional second key’s subtree. The functional final subtree, which is keyless, is related to the imperative implementation’s last key’s subtree.

\subsection{B+ Tree Cursor Type}

Imperatively a cursor is a pointer to a particular entry in a B+\,tree. Once a cursor points to a part of a B+\,tree, it can be used to retrieve the value stored there, update it to store something else, or insert an entirely new key-value pair to the tree. In a functional B+\,tree there are no pointers per se, and we cannot update the tree in-place in the same way since all data is immutable. However, it is still useful to represent the idea of a cursor “pointing at” a particular position in a B+\,tree.

The representation used was to structure a pointer into a treelist as the combination of that treelist and the index of the cursor’s position within it. The \texttt{cursor} type is then a list of indices combined with a list of treelists. This cursor structure was chosen to make proofs more straightforward, because it contains the original tree structure unmodified, and the \texttt{make\_cursor} operation structurally matches the \texttt{treelist} type (this is explained in more detail in section \ref{subsec:cursorop}). However, other operations are slightly complicated because of the choice to use indices: every step requires destructing the treelist to find the right position before we can get or update the keys and values stored there.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
Definition cursor : Type := list nat * list treelist.
\end{minted}
\end{singlespace}
\end{figure}

Other possible representations of a cursor were also considered when designing the B+\,tree types for the functional model.\footnote{The primary other representation considered was to think of a cursor as a \textit{split} in a tree: it “points” to a particular part of the B+\,tree by actually separating it into the trees and treelists to the left of that part vs those to the right. The cursor type is then a list of treelist splits. Representing a cursor this way makes operations on that cursor simpler: getting the (key,value) pair at the cursor only requires looking at the entry immediately on the right in the first treelist split; updating it can be easily done by updating the value, gluing the split back together, and recursively inserting it back into the next level up. Unfortunately, this representation of a cursor complicates the task of cursor creation, and by consequence complicates the process of proofs about that cursor.} The choice of this particular representation was made to prioritize simplicity of proofs over simplicity of implementation. The purpose of this functional model is \textit{not} to be efficient---it exists as a tool to be proven about; with that in mind, it is almost always better to prioritize proofs over code when there is a trade-off to be made.

Note that this representation of a cursor is as a pair of \textit{lists} of individual treelist pointers. This might seem strange, since an imperative implementation only needs to store a single pointer to a single position. However, representing a cursor as a list of pointers reflects the structure of the SQLite cursor mentioned earlier, which is similarly a list of pointers to the relevant location at each level. It also greatly simplifies the functional implementation and proofs to be able to destruct and reconstruct the B+\,tree on a per-level basis, rather than needing to start from the root for any operation.

Also note that the \texttt{cursor} type as defined here has no way to distinguish between pointing \textit{at} an element versus pointing \textit{in between} two elements. In fact, in an abstract sense, a cursor is always considered to represent a split between two keys. In situations where we are considering the element that the cursor is on (e.g. when looking up a value), by default we assume it is the first element to the right of the cursor’s position. However, it is important to note that the cursor does not need to be on a particular element. For example, a cursor into an entirely empty table is completely valid, and is represented as any index into the treelist \texttt{tl\_nil}.

This adds a potential point for confusion: two \textit{structurally} different cursors could \textit{abstractly} represent the same split. Consider the following example:

%Do this%
---------Need a graphic for this---------

Both arrows represent the split between keys 2 and 6 in this B+\,tree---the first could be the result of \texttt{make\_cursor 3 f}, and the second the result of \texttt{make\_cursor 5 f}. Abstractly they are equivalent, and any operations on them should do the same thing. But in our implementation they are not actually equal structures. So we have to ensure (and prove) that all of our B+\,tree operations can detect this and treat it appropriately.

\clearpage

\section{Functional Implementation} \label{sec:func}

\subsection{Element-List Abstraction}

The question of proving that our B+\,tree operations treat cursors appropriately relies on an intuitive understanding of how cursors should behave that we now need to formalize. Within our implementation, a B+\,tree cursor is a list of indices and treelists. In the abstract specification, a B+\,tree cursor is only defined by how our operations use it. This leaves a gap in formalization that makes it difficult to prove our operations correct or prove that they use cursors appropriately.

To solve this problem, we introduce an intermediate abstraction of cursors as an element-list. An element, in this sense, is any key-value pair that appears in the B+\,tree. The tree itself then is a list of these elements in order by increasing key. A cursor splits that list into the parts before the cursor’s position and the parts after it, and to simplify our operations we reverse the order of the first list. In other words, the \textit{element-list abstraction} of a cursor is a pair of lists, one to the “left” and one to the “right”, where the first element in each list is the element immediately adjacent to the cursor on each side, and the last element in each list is the extreme first or last element of the tree.

It is clear that under this abstraction, functionally equivalent cursors are represented the same. If two structurally different cursors in the tree point to the same split, then they must have the same element-list abstraction. Furthermore, our operations on these element-lists are very simple. To lookup the value at a cursor, we simply pop an item off the right list and return its value. To insert a new key-value pair we just add it to the left list (placing the cursor just after it); to update the value for a key already in the list we just replace the value at the beginning of the right list. Moving the cursor position is just popping items off of one list and concatenating them to the other.

\subsection{Cursor Operations} \label{subsec:cursorop}

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
Function make_cursor_rec (x: key) (f : treelist) (ci : list nat)
(ct : list treelist) (n : nat) : cursor :=
  match f with
  | tl_nil => (n::ci,ct)
  | tl_cons (node k f') f =>
    if lt_key k x then make_cursor_rec x f ci ct (S n)
    else make_cursor_rec x f' (n::ci) (f'::ct) O
  | tl_cons (final f') tl_nil =>
    make_cursor_rec x f' (n::ci) (f'::ct) O
  | tl_cons (val k v) f =>
    if lt_key k x then make_cursor_rec x f ci ct (S n)
    else (n::ci,ct)
  | _ => ([],[])
  end.

Definition make_cursor (x : key) (f : treelist) : cursor :=
  make_cursor_rec x f [] [f] O.
\end{minted}
\end{singlespace}
\end{figure}

Creating a cursor to point at a particular key $x$ happens in two stages. Within a particular treelist, we scan across each node, comparing the node’s key k against the target key x. This is what is happening in the \texttt{match} section: the treelist is decomposed horizontally into its component trees. As we scan across, we update a counter $n$ that represents the index within the treelist where the function currently is. Once we find the first key greater than or equal to $x$ (as in the \texttt{else} portion), we add that index onto the cursor that we are building and restart from 0 in the treelist below that index. If we ever hit a \texttt{final} tree, then we can assume this is the last tree in the treelist and descend into it without scanning the treelist further. For leaves, we will either hit a \texttt{val} that matches or we will hit \texttt{tl\_nil} because we have passed the end of the treelist. Either way, we can just append the index and return the cursor that has been built up.

Decomposing the creation of a cursor in this way helps to make induction and termination of this function easy. At each step we are destructing the treelist given and recursing on one of the two results. Since functional objects have no pointers and must be constructed from the data they hold, it is impossible to have an infinitely large tree, so it is easy for Coq to automatically prove the termination of this function. Similarly, when we are proving properties of \texttt{make\_cursor}, having a function that destructs the treelists naturally fits precisely with our induction scheme:

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
treelist_tree_rec :
forall (P : tree -> Prop) (P0 : treelist -> Prop),
(forall (k : key) (t : treelist), P0 t -> P (node k t)) ->
(forall t : treelist, P0 t -> P (final t)) ->
(forall (k : key) (v : V), P (val k v)) ->
P0 tl_nil ->
(forall t : tree, P t ->
  forall t0 : treelist, P0 t0 -> P0 (tl_cons t t0)) ->
forall t : treelist, P0 t.
\end{minted}
\end{singlespace}
\caption{The induction scheme requires a property \texttt{P} that is inductively assumed about every tree, and a property \texttt{P0} that is assumed and proved about every treelist. Having functions destruct a treelist in both directions matches this scheme well.}
\label{fig:induction}
\end{figure}

This works well for making a cursor, but we still have the issue of multiple structurally distinct cursors which are abstractly the same. In order to do any operations with a cursor, we need a way to handle these two cursors equivalently. For this purpose, we wrote two normalization functions. The first, \texttt{next\_node}, normalizes a cursor so it points directly to a \texttt{val} tree; this would be the position pointed to by \texttt{make\_cursor 5} in the example tree from the previous section. The second, \texttt{prev\_node}, is the reverse: it normalizes a cursor so it has a val tree directly behind it, i.e. the \texttt{make\_cursor 3} position in the example tree. For both functions, the functions should change the structure of the cursor without changing the abstract element-list of the cursor.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
Fixpoint next_node (cn : list nat) (cf : list treelist)
: option (cursor * key) :=
  match (cn,cf) with
  | (n::cn',f::cf') =>
    (match point n f with
     | (_,Some (node k _),tl_cons (node _ f') _) =>
       Some (((S n)::cn',f::cf'),k)
     | (_,Some (node k _),tl_cons (final f') _) =>
       Some (((S n)::cn',f::cf'),k)
     | (_,Some (val k v), _) => Some ((n::cn',f::cf'),k)
     | _ =>
       (match next_node cn' cf' with
        | Some ((n'::cn,f'::cf),k) =>
          (match point n' f' with
           | (_,Some (node _ f1),_) =>
             Some ((O::n'::cn,f1::f'::cf),k)
           | (_,Some (final f1),_) =>
             Some ((O::n'::cn,f1::f'::cf),k)
           | _ => None
           end)
        | _ => None
        end)
     end)
  | (_,_) => None
  end.
\end{minted}
\end{singlespace}
\end{figure}

The central functionality of \texttt{next\_node} is to travel up the treelist as far as necessary to find a treelist that has a next tree to descend down into. It then positions itself before the very first element of that tree. If it is able to find such a modified cursor, it returns both that cursor and the pivot key at the highest node it reached (i.e., the key that separates the original cursor position from the new one). If it is not able to find such a cursor---in other words, if the cursor was already at the last possible position in the tree---then it returns \texttt{None}.

Here, \texttt{point n f} is a function that decomposes a treelist $f$ into three parts: the treelist before index $n$, the tree at index $n$ (if it exists, \texttt{None} otherwise), and the treelist after index $n$. We assume \texttt{next\_node} is called from a leaf node, so if there is a \texttt{val} that the cursor is already pointing to then we are done, and no modification is necessary. If there is something next to point to that is a \texttt{node} or a \texttt{final}, then we can assume we are in the middle of a recursive call. Since there is something next, this is the pivot tree, so we move the index up by one and attach the associated key $k$.

If we match \texttt{point n f} and find something else, for example \texttt{None} or a \texttt{final} tree, then we are at the end of a node and cannot point to the next thing. Therefore, we must recursively call \texttt{next\_node} to get the node adjacent to this one. Then we can move down into the tree pointed to by that partial cursor, and position a new (potentially partial) cursor at the first index of that tree.

The structure of \texttt{prev\_node} is very similar, but operating in reverse it destructs the cursor index to examine the tree immediately \textit{before} the cursor’s position. If it is at the beginning of a node and there is nothing directly preceding it to point to, then it recursively calls \texttt{prev\_node} at the next level up in the cursor. The partial cursor it gets back is then grown with the treelist $f’$ pointed to and a pointer to the last element of $f’$.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
Fixpoint move_to_next (c : cursor) : cursor :=
  match c with (cn,cf) =>
  (match next_node cn cf with
   | Some ((n::cn',cf'),_) => (S n::cn',cf')
   | _ => c
   end)
  end.

Fixpoint move_to_prev (c : cursor) : cursor :=
  match c with (cn,cf) =>
  (match prev_node cn cf with
   | Some ((S n::cn',cf'),_) => (n::cn',cf')
   | _ => c
   end)
  end.
\end{minted}
\end{singlespace}
\end{figure}

With these normalization functions, actually moving the cursor becomes trivial. We normalize to the next or prev position to ensure there is an adjacent element to move the cursor to, then we change the cursor to point to it.

Furthermore, the correctness of these functions follows directly from the correctness of the normalization functions. Given proofs that \texttt{next\_node} and \texttt{prev\_node} both produce correct cursors, then \texttt{move\_to\_next} and \texttt{move\_to\_prev} clearly produce correct cursors as well because the only modification they make is in the last index of the cursor. We also prove that both \texttt{next\_node} and \texttt{prev\_node} preserve the element-list, so that they position the cursor before or after an element without changing its abstract representation. Then it is intuitive to prove that the element-list of \texttt{move\_to\_next} of a cursor c is the original cursor’s element-list, but with the first element of the “right” list moved to the “left” list. Similarly, \texttt{move\_to\_prev} is the original element-list, but with the first element of the “left” list moved to the “right” list. If there is no right (or respectively left) element, this implies that the cursor is at the very end (or very beginning) of the tree, and we prove that \texttt{next\_node} and \texttt{prev\_node} return \texttt{None} in that case, so \texttt{move\_to\_next} and \texttt{move\_to\_prev} just return the original cursor.

These properties lead directly to proofs of the \texttt{next\_prev} and \texttt{prev\_next} properties in the abstract specification: if the cursor is not the last cursor, it must have an element to its right in its element-list; then \texttt{move\_to\_next} moves that element to the left list, and \texttt{move\_to\_prev} moves it back to the right. Similarly, if a cursor is not the first cursor, \texttt{move\_to\_prev} moves an element to the right list, and \texttt{move\_to\_next} moves it back to the left.

\subsection{Search}

To simplify proofs of search properties, we have combined the core of \texttt{get} and \texttt{get\_key} into a single function \texttt{get\_tree} from which the key and value can easily be retrieved. This allows us to prove properties of \texttt{get\_tree} once, and easily relate them to both \texttt{get} and \texttt{get\_key}.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
Fixpoint get_tree (c : cursor) : option tree :=
  match c with (cn,cf) =>
  (match next_node cn cf with
   | Some ((n::_,f::_),_) => lin_search n f
   | _ => None
   end)
  end.
\end{minted}
\end{singlespace}
\end{figure}

Similarly to the \texttt{move\_to\_next} and \texttt{move\_to\_prev} operations, the \texttt{get\_tree} operation is vastly simplified by having set up \texttt{next\_node} and the element-list abstraction. We normalize with \texttt{next\_node} to ensure that the cursor points directly at an element, and then we can return that element. If \texttt{next\_node} doesn’t return a valid cursor, then we must be at the end of the tree, so there is no element for get to return.

The correctness property of this function relative to the element-list abstraction is simple, and can be easily applied to the properties of the abstract specification. Specifically, if $(k,v)$ is the element at the front of the right element-list of a (correct) cursor, then \texttt{get\_tree} should return \texttt{val k v}. If the cursor is correct---including that it points into a leaf node---then \texttt{next\_node} must be correct as well, so the treelist that \texttt{next\_node} returns will necessarily be a leaf node, and the tree that \texttt{get\_tree} returns must necessarily be a \texttt{val} tree. Furthermore, given that \texttt{next\_node} positions the cursor to point a tree without changing the element-list for the cursor, we know both that \texttt{get\_tree} will return that tree and that it must be at the front of the right element-list of \texttt{next\_node} of the cursor (and therefore of the original cursor as well). Therefore, the tree returned by \texttt{get\_tree} and the element at the beginning of the right element-list must be the same.

\subsection{Insertion}

Insertion into a functional B+\,tree is significantly more complicated than cursor manipulation or search operations. There are two possibilities for insertion. If the key is not already in the tree we want to insert it and position the cursor immediately after it (this makes it easy to quickly create a new B+\,tree or insert a range of key-value pairs). If the key is already in the tree we want to simply update the value and leave the cursor positioned as is. Both of these assume that the cursor is properly positioned to insert the key; if it is not then the cursor should simply be returned unchanged.

By comparison, the element-list abstraction of insertion is quite simple. Let $l$ be the left element-list, and let $(k,v)::r$ be the right element-list. To insert a key-value pair $(x,v’)$, if $k$ and $x$ are not equal, then it should result in element-lists of $(x,v’)::l$ and $(k,v)::r$. If $k$ and $x$ are equal, then the result should be $l$ and $(k,v’)::r$.

The functional B+\,tree implementation of insertion has to achieve several more goals. Most important, it has to maintain the fanout value $b$, as explained in section \ref{sec:b+tree}, so that every treelist contains between $\lceil \frac{b}{2} \rceil$ and $b$ subtrees (except for potentially the root). This means that inserting a key-value pair might require splitting the leaf it was inserted into, and inserting the new leaf into its parent node might require splitting that node, and so on up to the root. In order to handle this, we introduce a new type \texttt{splitpair}:

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
Inductive splitpair : Type :=
| One : treelist -> splitpair
| Two : treelist -> key -> treelist -> splitpair.
\end{minted}
\end{singlespace}
\end{figure}

A treelist that is the correct size and is not split will result in a \texttt{One}, but if a treelist is too large and must be split then it results in a \texttt{Two} containing each half of the split treelist as well as the key differentiating them that will be inserted up into the parent node.

At a high level, inserting into a treelist is a recursive process that destructs the cursor going up and recreates it going back down. For each level of the treelist, we receive a splitpair from the level below. We use the function \texttt{insert\_across} to place that splitpair’s treelist(s) into the position indicated by the cursor, then the function \texttt{insert\_up} (potentially) splits the new treelist and passes it up to be inserted in the next level of the cursor. The result it gets back then has the appropriate index and treelist appended back onto it. Since the cursor carries with it the whole structure of the B+\,tree, the whole cursor must be destructed and reconstructed with the update B+\,tree, even if no node needs to be split.

\subsection{Correctness Theorem}

We need one final theorem that pulls all of the other results together and states that the functional model satisfies all of the properties in the \texttt{Module}. In Coq this is completed by filling out an instance of the \texttt{CURSOR\_TABLE} module:

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
Module BT_Table <: CURSOR_TABLE.
  ...
End BT_Table.
\end{minted}
\end{singlespace}
\end{figure}

In more intuitive terms, we can summarize this as the following final theorem:

\begin{theorem}
The functional B+\,tree program described in section \ref{sec:func} satisfies all of the axioms laid out in the abstract specification described in section \ref{sec:spec}.
\end{theorem}

\clearpage

\section{Reflections on Specification}

As described above from the DeepSpec group, the goal of a deep specification has four parts: it should be rich, two-sided, formal, and live. All of these properties together ensure a high-quality specification that is both correct and useable. Since that is the goal of this thesis, it is worth reflecting on how well this particular specification project achieved those properties.

This specification certainly attempts to be both rich and formal. By modelling the desired program as an equivalent functional program in Coq’s functional programming language Gallina, and by proving its correctness properties using Coq, we have both described the desired functionality in detail and in mathematically rigorous terms. Nonetheless, this specification has room to improve in both richness and formality. The difficulty of a functional programming language to represent naturally imperative concepts such as a cursor makes the implementation at times inelegant and complicated. In this way, the specification struggles to accurately and in detail describe the imperative implementation being targeted. Furthermore, the complicated implementation makes formal proofs more difficult, and as a result this verification relies on informal intuitive proofs in some areas to fill in gaps that could not be proved rigorously in Coq. However, both of these shortcomings would be overcome in a complete implementation and verification. In that sense, the verification \textit{approach} taken is both rich and formal. The specification would also certainly be live, were it attached to an imperative program through machine-checked proofs using the Verified Software Toolchain.

The specification was not initially two-sided, since there is no current client for it, and this was the cause of some difficulty in composing it. However, this thesis does demonstrate a certain client side, even without an explicit client program.

Two-sidedness is important for two reasons: first, it provides a clear goal for the specification; and second, it ensures that the specification is actually useful. Without it, there is the risk that the specification’s theorems will be irrelevant to how a client would want to use it, or worse, that a future client program not directly connected to the original specification could misunderstand it and rely on properties that may not actually be true.

The issue of two-sidedness was able to be addressed through the abstract specification. Once the abstract specification was fairly well-developed, then this actually motivated some of the later changes that were made in the functional specification. For example, the decision to build the cursor into the module type resulted in a decision to change the inputs to the \texttt{get} and \texttt{insert} functions and consider cursor operations separately from search and insertion operations. Additionally, the need to formalize abstract axioms required consideration of the exact properties each function needed to have, and how they could best be implemented. In this way, the completed abstract specification functioned as an approximation for the process of exercising the specification from the client side. It helped to simplify the properties that needed to be proved into a concise description that would hopefully be clear and useful to any future client of the specification. Furthermore, the very brief example in section \ref{subsec:ex} demonstrates some of the applicability of the abstract specification developed.

Although this specification is not a completely deep specification, and could not be without a full imperative verification outside the scope of this thesis, the approach presented here makes significant progress toward a true deep specification. For all four of richness, formality, liveness, and two-sidedness, attempting to achieve these properties in this specification led to a better, more rigorous, and more useful specification.

\clearpage

\section{Conclusion}

This thesis describes the uses and complexities of B+\,trees and their verification. B+\,trees as a search tree variant are both very useful and very complicated, and the formalization and specification of their functional operation is an important step toward an applicable complete verification of an imperative B+\,tree implementation.

The detailed examination of the mechanics of B+\,tree cursors is an important development for formally understanding and describing B+\,trees. The abstract specification in this thesis presents one possible complete axiomatization of the operation of cursor tables such as a B+\,tree. The element-list abstraction gives an intuitive and workable simplification of a B+\,tree cursor that serves as a crucial intermediary for proving properties of B+\,tree cursors. It acts as a firewall between the complicated functioning of the cursor operations and the desired properties to prove: once the complex cursor operations have been proved equivalent to the simple element-list operations, it is much easier to work with the element-list and prove high-level properties about it without worrying about complicated analysis of the underlying functions and the large number of cases they introduce.

The formal proofs of the cursor and B+\,tree operations demonstrate the correctness and usefulness of these varying levels of abstractions. They also serve as evidence for this approach to specification and how it can be successfully applied to complex programs for useable verification interfaces.

\clearpage

\section{Future Directions}

This thesis has both practical applications and broader implications about specification in general. The functional specification developed in this thesis can be further developed and used as the basis for the verification of an imperative B+\,tree program, and such a B+\,tree database is currently being developed in C for verification at Princeton University by Oluwatosin Adewale and Aurèle Barrière. The process and approach to specification in this thesis is based on the deep specification framework being studied by the DeepSpec group, and this thesis highlights the successes and complications encountered with that approach. Further work with the specification approach used in this thesis could extend and improve specification techniques for complex imperative software and one-sided specifications.

\clearpage
\bibliographystyle{IEEEtran}
\bibliography{biblio}

\end{document}

### 2018-04-30 model/thesis_draft/ProofSummary.tex

% Use this template to write your solutions to COS 423 problem sets

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amsfonts, amsthm, amssymb, algorithm, graphicx, mathtools,xfrac}
\usepackage[noend]{algpseudocode}
\usepackage{fancyhdr, lastpage}
\usepackage[vmargin=1.20in,hmargin=1.25in,centering,letterpaper]{geometry}
\setlength{\headsep}{.50in}
\setlength{\headheight}{15pt}

% Landau notation
\DeclareMathOperator{\BigOm}{\mathcal{O}}
\newcommand{\BigO}[1]{\BigOm\left({#1}\right)}
\DeclareMathOperator{\BigTm}{\Theta}
\newcommand{\BigTheta}[1]{\BigTm\left({#1}\right)}
\DeclareMathOperator{\BigWm}{\Omega}
\newcommand{\BigOmega}[1]{\BigWm\left({#1}\right)}
\DeclareMathOperator{\LittleOm}{\mathrm{o}}
\newcommand{\LittleOh}[1]{\LittleOm\left({#1}\right)}
\DeclareMathOperator{\LittleWm}{\omega}
\newcommand{\LittleOmega}[1]{\LittleWm\left({#1}\right)}

% argmin and argmax
\newcommand{\argmin}{\operatornamewithlimits{argmin}}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}

\newcommand{\calP}{\mathcal{P}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Exp}{\mathbb{E}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\sign}{\mathrm{sign\ }}
\newcommand{\abs}{\mathrm{abs\ }}
\newcommand{\eps}{\varepsilon}
\newcommand{\zo}{\{0, 1\}}
\newcommand{\SAT}{\mathit{SAT}}
\renewcommand{\P}{\mathbf{P}}
\newcommand{\NP}{\mathbf{NP}}
\newcommand{\coNP}{\co{NP}}
\newcommand{\co}[1]{\mathbf{co#1}}
\renewcommand{\Pr}{\mathop{\mathrm{Pr}}}

% theorems, lemmas, invariants, etc.
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{invariant}[theorem]{Invariant}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{property}[theorem]{Property}
\newtheorem{proposition}[theorem]{Proposition}

% piecewise functions
\newenvironment{piecewise}{\left\{\begin{array}{l@{,\ }l}}
{\end{array}\right.}

% paired delimiters
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}
\DeclarePairedDelimiter{\len}{|}{|}
\DeclarePairedDelimiter{\set}{\{}{\}}

\makeatletter
\@addtoreset{equation}{section}
\makeatother
\renewcommand{\theequation}{\arabic{section}.\arabic{equation}}

% algorithms
\algnewcommand\algorithmicinput{\textbf{INPUT:}}
\algnewcommand\INPUT{\item[\algorithmicinput]}
\algnewcommand\algorithmicoutput{\textbf{OUTPUT:}}
\algnewcommand\OUTPUT{\item[\algorithmicoutput]}

% Formating Macros

\pagestyle{fancy}
\lhead{\sc \hmwkClass $\;\;\cdot \;\;$ \hmwkSemester $\;\;\cdot\;\;$
Proof Summary}
%\chead{\sc Problem \hmwkAssignmentNum.\hmwkProblemNum}
%\chead{}
\rhead{\em \hmwkAuthorName\ $($\hmwkAuthorID$)$}
\cfoot{}
\lfoot{}
\rfoot{\sc Page\ \thepage\ of\ \protect\pageref{LastPage}}
\renewcommand\headrulewidth{0.4pt}
\renewcommand\footrulewidth{0.4pt}

\fancypagestyle{problemstatement}
{
    \rhead{}
    \lfoot{}
}

%%%%%% Begin document with header and title %%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%% Header Information %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Shouldn't need to change these
\newcommand{\hmwkClass}{Thesis}
\newcommand{\hmwkSemester}{Spring 2018}

%%% Your name, in standard First Last format
\newcommand{\hmwkAuthorName}{Brian McSwiggen}
%%% Your NetID
\newcommand{\hmwkAuthorID}{bm14}

%%%%%%% begin Summary %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This document aims to give a brief and informal summary of the relevant lemmas and theorems for my thesis. Each section describes at a high level the goals of the relevant portion of the implementation, describes the important lemmas and theorems, and lists the associated Coq theorem names. Formatting issues have made it difficult to include the formal statement of each theorem, but they can all be found in BTrees.v.

\section{Element List Abstraction}

We represent a cursor abstractly as a pair of key-value lists. The first is the "left" list, of all the key-value pairs with keys \textit{less} than the cursor's current position; the second is the "right" list with the pairs \textit{greater} than the current position.

Both lists are ordered by distance to the cursor's position, i.e. the adjacent key-value pair to each side is the first element of the corresponding list. The first element of the tree is the last element of the "left" list. The last element of the tree is the last element of the "right" list.

We can translate a cursor into this paired-list using \texttt{cursor\_elements}; we can also translate a cursor into just the left or right lists using \texttt{cursor\_left} or \texttt{cursor\_right}, respectively.

\bigskip
\texttt{cursor\_elements'\_sides\_equiv}

\begin{lemma}
The pair produced by \texttt{cursor\_elements} is equal to the pair of \texttt{cursor\_left} and \texttt{cursor\_right}.
\end{lemma}

\begin{proof}
This has been formally proved in Coq.
\end{proof}

Both the right and left functions are called recursively and gradually build up a "base" of elements already added. For both directions, we process elements from inside outward. Therefore, we want to prove the following:

\bigskip
\texttt{right\_el\_interior} and \texttt{left\_el\_interior}

\texttt{left\_rec\_interior} and \texttt{right\_rec\_interior}

\begin{lemma}
For both the left and right functions, the "base" they are passed ends up at the beginning of the list, i.e. if the base is \texttt{b} and the end result is \texttt{l}, then there is some \texttt{l'} such that \texttt{l = b++l'}.
\end{lemma}

\begin{proof}
This has been formally proved in Coq.
\end{proof}

We also want to show that cursors are processed from the bottom up, and that the things reached from the bottom (which are consequently closer to the cursor position) end up in the base first. In other words, we want to prove the following:

\bigskip
\texttt{cursor\_right\_elements1} and \texttt{cursor\_left\_elements1}

\begin{lemma}
For both the left and right functions, if we split a cursor into two parts, then the result of computing the elements of the whole cursor is equal to the result of first computing the elements of the first part, then using that as the base for computing the elements of the second part.
\end{lemma}

\begin{proof}
This has been formally proved in Coq.
\end{proof}

Finally, we need to prove that the element list actually represents the tree:

\bigskip
\texttt{cursor\_right\_el\_sorted} and \texttt{cursor\_left\_el\_sorted}

\begin{theorem}
The element list for a correct cursor (i.e. of an in-order tree) is in order.
\end{theorem}

\begin{proof}
This has \textbf{not} yet been proved.
\end{proof}

\section{Abstract Cursor Position}

This abstract element-list works well for what a cursor should represent, but it has a problem for the actual structure of a cursor: two \textit{structurally} different cursors could be \textit{abstractly} the same! Consider one cursor pointing at the very end of a leaf, and another cursor pointing at the very beginning of the next leaf.

So, we need a way to work with these cursors. Specifically, we create \texttt{next\_node} and \texttt{prev\_node}, which ensure a cursor either has an element immediately to its right or has an element immediately to its left. If a cursor is already in an acceptable position, then it should not be changed.

Importantly, these functions should not alter the abstract element-list representation of the cursor. In order to prove this, we have to first prove an invariant of the recursive right and left functions:

\bigskip
\texttt{cursor\_right\_elements4}, uses \texttt{point\_next}

\begin{invariant}
Let $c$ be a correct (partial) cursor, and let $c'$ be the (partial) cursor returned by \texttt{next\_cursor} of $c$. Then if $l$ is the right portion of the element list for $c$, and $l'$ is the right portion of the element list for $c'$, then the two lists are the same, except for the subtree pointed to by $c'$ if the cursor is pointing to an internal node.
\end{invariant}

\begin{proof}
This has been \textbf{mostly} proved in Coq. The exceptions are a couple of admits that follow directly from the cursor being correct.
\end{proof}

\bigskip
\texttt{cursor\_left\_elements4}, uses \texttt{left\_el\_app}

\begin{invariant}
Let $c$ be a correct (partial) cursor, and let $c'$ be the (partial) cursor returned by \texttt{next\_cursor} of $c$. Then if $l$ is the left portion of the element list for $c$, and $l'$ is the left portion of the element list for $c'$, then the two lists are the same, except for the subtree pointed to by $c$ if the cursor is pointing to an internal node.
\end{invariant}

\begin{proof}
This has been \textbf{partially} proved in Coq.
\end{proof}

We need to ensure that if \texttt{next\_cursor} is called on any correct cursor, then the result is also correct.

\bigskip
\texttt{next\_correct\_struct} and \texttt{next\_node\_balanced}

\texttt{next\_node\_sorted} and \texttt{next\_node\_fanout}

\texttt{next\_node\_correct}

\begin{lemma}
\texttt{next\_cursor} preserves cursor correctness.
\end{lemma}

\begin{proof}
This has been \textbf{partially} proved in Coq.
\end{proof}

We also need to prove all of these properties for \texttt{prev\_cursor}; this has not yet been formally done in Coq.

We can now apply these invariants to \texttt{cursor\_elements} as a whole:

\bigskip
\texttt{cursor\_elements\_next and cursor\_elements\_prev}

\begin{theorem}
If $c$ is a correct cursor, then \texttt{next\_cursor} of $c$, \texttt{prev\_cursor} of $c$, and $c$ all have the same element-list representation.
\end{theorem}

\begin{proof}
This follows directly from the invariants stated above, but has not yet been formally proved in Coq because those proofs have not been completed.
\end{proof}

\section{Cursor Creation}

In order to create a cursor, we recursively build up a list of indices into treelists, starting from the root and going down to the leaf that contains (or would contain) the key the cursor is pointing at. We need to prove primarily two things: first, that this produces a correct cursor; and second, that the positioning of the cursor (relative to the abstract element-list representation) is correct.

\bigskip
\texttt{make\_cursor\_rec\_correct} and \texttt{make\_cursor\_correct}

\begin{theorem}
\texttt{make\_cursor} produces a structurally correct cursor.
\end{theorem}

\begin{proof}
This has been \textbf{partially} proved in Coq.
\end{proof}

\bigskip
\texttt{make\_cursor\_right} and \texttt{make\_cursor\_left}

\begin{theorem}
The cursor $c$ returned by \texttt{make\_cursor} for a given key $k$ has the property that the immediate right element has key greater than or equal to $k$, and the immediate left element has key less than $k$.
\end{theorem}

\begin{proof}
This has \textbf{not} yet been proved.
\end{proof}

\section{Cursor Movement}

For \texttt{move\_to\_next} and \texttt{move\_to\_prev}, we similarly need to prove structural correctness and correctness relative to the element list abstraction.

\bigskip
\texttt{move\_to\_next\_correct} and \texttt{move\_to\_prev\_correct}

\begin{theorem}
Both functions, given a correct cursor, produce a correct cursor.
\end{theorem}

\begin{proof}
This has been formally proved for \texttt{move\_to\_next}, but not for \texttt{move\_to\_prev}. The proof would follow the same structure, and will be easy.
\end{proof}

\bigskip
\texttt{move\_to\_next\_el}

\begin{theorem}
If $c$ has the element lists $(l,(k,v)::r)$, then next of $c$ has element lists $((k,v)::l,r)$.
\end{theorem}

\begin{proof}
This has \textbf{not} yet been proved.
\end{proof}

\bigskip
\texttt{move\_to\_prev\_el}

\begin{theorem}
If $c$ has the element lists $((k,v)::l,r)$, then prev of $c$ has element lists $(l,(k,v)::r)$.
\end{theorem}

\begin{proof}
This has \textbf{not} yet been proved.
\end{proof}

\bigskip
\texttt{move\_to\_next\_none} and \texttt{move\_to\_prev\_none}

\begin{theorem}
If $c$ has no element to its right (or to its left), then next of $c$ (or prev of $c$) is just $c$.
\end{theorem}

\begin{proof}
This has \textbf{not} yet been proved.
\end{proof}

\section{Get}

Now that we've set up the element-list abstraction, there's actually only one thing that we need to prove about \texttt{get}. Furthermore, the proof follows immediately from what we've already proven!

\bigskip
\texttt{get\_correct}

\begin{theorem}
\texttt{get} returns the key-value pair at the front of the right list in the element-list abstraction.
\end{theorem}

\begin{proof}
This has \textbf{not} yet been formally proved.

However, the proof will mirror the following intuitive proof:

We know that \texttt{next\_node} positions the cursor structurally right next to a key-value pair without changing the element-list representation. Then \texttt{get} returns that element directly to the right, which must also be the front of the right element-list.
\end{proof}

\section{Insert}

Insertion has three distinct cases to consider when inserting a key-value pair $(k,v)$:
\begin{enumerate}
    \item If the cursor is not correctly positioned for $k$, then it should return the cursor unchanged.
    \item If the cursor is correctly positioned, but $k$ is not in the tree, then it should add the new pair $(k,v)$.
    \item If the cursor is correctly positioned, and $k$ is in the tree (meaning it is the next key to the right in the element list), then it should replace the value for $k$ with $v$.
\end{enumerate}

We can prove a helpful lemma with regards to the first point:

\bigskip
\texttt{bad\_cursor\_insert\_same}

\begin{lemma}
If the key $k$ is not in the range of keys represented by this cursor, then \texttt{insert} returns back the original cursor.
\end{lemma}

\begin{proof}
This has \textbf{not} yet been proved.
\end{proof}

Now consider insertion of a key $k$ into a cursor that represents $k$. We first want to prove that insert is structurally correct, i.e. a correct cursor, once inserted into, gives back a correct cursor.

\bigskip
\texttt{insert\_correct}

\begin{theorem}
If $c$ is a cursor into a correct b+tree (balanced, fanout, and ordered), then for any key $k$ and value $v$, inserting $(k,v)$ into $c$ produces a cursor into a correct b+tree.
\end{theorem}

\begin{proof}
This has \textbf{not} yet been proved. In Coq, it would be most likely split into three separate proofs (one for each correctness property), for simplicity and clarity. However, for the statement now we have combined it with \textbf{Theorem 18}.
\end{proof}

\bigskip
\texttt{insert\_correct}

\begin{theorem}
If $c$ is a structurally correct cursor (each level contains the next, and the last level is a leaf node) then for any $k$ and $v$, inserting $(k,v)$ into $c$ produces a structurally correct cursor.
\end{theorem}

\begin{proof}
This has \textbf{not} yet been proved.
\end{proof}

Then, we want to prove what it does to the element-list abstraction, i.e. that it either replaces the next key-value pair or inserts this key-value pair at the cursor's position.

\bigskip
\texttt{insert\_eq\_elements}

\begin{theorem}
If the next key is equal to the key $k$ being inserted, then let $l$ be the left element list and $(k,v')::r$ be the right element list. Then inserting $(k,v)$ into the cursor produces a cursor with element list equal to: $(l,(k,v)::r)$.
\end{theorem}

\begin{proof}
This has \textbf{not} yet been proved.
\end{proof}

\bigskip
\texttt{insert\_neq\_elements}

\begin{theorem}
If the next key is not equal, then let $l$ be the left element list and $r$ be the right element list. Then insertion produces a cursor with element list equal to $((k,v)::l,r)$.
\end{theorem}

\begin{proof}
This has \textbf{not} yet been proved.
\end{proof}

%%%%%%% end Summary %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2018-05-17 papers/mcswiggen/McSwiggen-Thesis.tex

--- 2018-04-30 model/thesis_draft/Draft.tex
@@
-\documentclass[a4paper,12pt]{article}
+\documentclass[12pt]{article}
 \usepackage{enumerate}
@@
 %% Sets page size and margins
-\usepackage[a4paper,top=3cm,bottom=2cm,left=1.5in,right=1in,marginparwidth=1.75cm]{geometry}
+\usepackage[top=3cm,bottom=2cm,left=1.5in,right=1in,marginparwidth=1.75cm]{geometry}
 
@@
 \usepackage{minted}
+\usepackage{xpatch,letltxmacro}
+\LetLtxMacro{\cminted}{\minted}
+\let\endcminted\endminted
+\xpretocmd{\cminted}{\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}}{}{}
 \usepackage{setspace}
+\usepackage[title]{appendix}
+\usepackage{amsthm}
 
@@
 
+\newcommand{\nocontentsline}[3]{}
+\newcommand{\tocless}[2]{\bgroup\let\addcontentsline=\nocontentsline#1{#2}\egroup}
+
 \setlength{\parskip}{1em}
@@
 
-\title{Functionally Verified B+ Trees}
-\author{Brian McSwiggen\\
-        Advised by Professor Andrew Appel}
+%\title{Functionally Verified B+ Trees}
+%\author{Brian McSwiggen\\
+%        Advised by Professor Andrew Appel}
 
 \begin{document}
-\maketitle
+\doublespacing
+
+\thispagestyle{empty}
+\begin{sc}
+\vspace*{0in}
+\begin{center}
+    \LARGE The Theory and Verification of\\B+\,Tree Cursor Relations
+\end{center}
+\vspace{.5in}
+\begin{center}
+    Brian McSwiggen\\
+    Advisor: Professor Andrew Appel
+\end{center}
+\vspace{.5in}
+\begin{center}
+   Submitted in partial fulfillment\\
+of the requirements for the degree of\\
+Bachelor of Science in Engineering
+
+\includegraphics[width=0.25\textwidth]{logo.jpg}
+
+Department of Computer Science\\
+Princeton University\\
+May 2018
+\end{center}
+\end{sc}
 
 \clearpage
-
-\doublespacing
-
-\section{Abstract}
-
-In this thesis, I make progress toward a fully verified functional B+\,tree program in Coq.
+\pagenumbering{roman}
+
+\begin{center}
+    \LARGE \textbf{Abstract}
+\end{center}
+
+\textit{In this thesis, I make progress toward a fully verified functional B+\,tree program in the Coq proof assistant.
 A verified functional B+\,tree would serve as the basis for the future verification of an imperative B+\,tree program through proving equivalence, which would have applications in databases and other systems.
-The primary focus of the properties and abstractions in this thesis is to examine and describe the theory of the B+\,tree cursor, which is central to the operations of a B+\,tree both functionally and imperatively.
+The primary focus of the properties and abstractions covered in this thesis is to examine and describe the theory of the B+\,tree cursor, which is central to the operations of a B+\,tree both functionally and imperatively.
 The functional implementation presented here covers the essential cursor operations as well as B+\,tree lookups and insertions, but does not include a delete function.
-The progress made toward verification includes the development of a complete abstract specification, the structural correctness of cursor creation and positioning, and the relation of cursors and B+\,tree operations to an element-list abstract cursor representation.
+The progress made toward verification includes the development of a complete abstract specification, the structural correctness of cursor creation and positioning, and the relation of cursors and B+\,tree operations to an element-list abstract cursor representation.}
 
@@
 
+\begin{center}
+    \LARGE \textbf{Acknowledgements}
+\end{center}
+
+\begin{singlespace}
+\small
+\setlength{\parskip}{0.7em}
+As my time at Princeton comes rushing to an end, I would like to give a heartfelt thanks to the many, many people who helped me make it, who carried me through this wild place, and who sometimes had to drag me kicking and screaming.
+I couldn’t possibly fit all of you on this page, but I’m going to make the font small and try my best.
+
+First, to Professor Appel, I wouldn’t have any of this thesis or even know what formal verification is without having you as my advisor and my teacher.
+Thank you, for believing in me, for meeting with me every week, for giving me help and feedback as often as I asked for it, and for showing me all the cool things these tools can do.
+Also to Professor Walker, my second reader, thanks for teaching me functional programming in the first place.
+
+To Aurèle and Tosin, thank you both for taking the time to explain things to me and answer my questions.
+Thank you also for letting an undergrad contribute as much as he can, even though you probably could have done my whole thesis, better, in half the time.
+I hope what I’ve done is useful to you anyways.
+
+Thank you to all the professors and teachers and mentors and tutors and TAs that I have had throughout all my time in school, even before Princeton: Dr.
+McCall, Mr.
+Plummer, Madame Hecker, Dr.
+Kairet, Mrs.
+Mentor, and Mrs.
+Lewis, to name just a few.
+
+To my two crazy families on campus---Theatre Intime and Quad---you may have driven me crazy from stress but you also made it all worth it.
+I can’t imagine a better place to be stuck working until 2am.
+I’m so proud of what we accomplished, and I have so much faith in the new officers to do even better.
+
+To the Q33, forever in my heart, forever in my mind---but \textit{never} in the TV room.
+
+Kika, you made my senior year wonderful, and you’ve been the best companion I could have asked for on this wild ride.
+I have full confidence that your senior year will be even more wonderful, more fruitful, and produce a better thesis than this one.
+
+Some other people deserve a special shout-out for making sure I didn’t literally fail.
+Katie, for dragging me out of bed with Nutella for that god-awful 9am freshman physics precept.
+Julie, for placing bets on how long I could go without missing a class.
+Zach, for getting breakfast with me whenever I was worried I might sleep through a morning meeting.
+Swebb and Taylor, for always picking up the phone, even at ungodly hours of the night (or morning).
+Lydia, who has heard me rant about more stupid problems than anyone.
+Jillian, for knowing me since middle school, and for somehow remaining my friend even after finding out that I’m basically still a child.
+
+I of course have to thank all the people who are already PTL, and all my friends in the Class of 2017, for proving that it can in fact be done. I’m not sure I would believe it otherwise.
+
+Finally, to my family.
+My parents, for supporting me and guiding me (and even proofreading my thesis!).
+My brothers, for being role models, mentors, and friends.
+My Oma and Opa, for making it all possible.
+
+Thank you all. I love you all so much.
+\setlength{\parskip}{1em}
+\end{singlespace}
+
+\clearpage
+
+\begin{singlespace}
+\vspace*{1.5in}
+\begin{center}
+    This thesis represents my own work in accordance with University regulations.
+\end{center}
+\begin{flushright}
+    \textit{Brian McSwiggen}
+
+    May 7, 2018
+\end{flushright}
+\end{singlespace}
+
+\clearpage
+
+\begin{singlespace}
+\setlength{\parskip}{-0.1em}
+\tableofcontents
+\setlength{\parskip}{1em}
+\end{singlespace}
+
+\clearpage
+\pagenumbering{arabic}
+
 \section{Introduction and Motivation}
@@
 B+\,trees are an extension of the more commonly known B-trees, which are N-ary trees often used to organize data in file systems.
-The property that makes B+\,trees unique is that the \textit{values}, that is the data actually being held in the tree, are kept exclusively at the leaves.
+The property that makes B+\,trees unique is that the \textit{values}, that is the data actually being stored in the tree, are kept exclusively at the leaves.
 The interior nodes hold only keys and pointers to their children.
 
-One interesting consequence of this property is that B+\,trees in practice can be augmented with a linked list between the leaf nodes connecting all of the values.
-This makes finding a value stored in the tree efficient (due to the tree structure) while still keeping range queries and iteration over the values easy (due to the linked list).
-Additionally, because B+\,trees don’t have to store as much data in the interior nodes, a single node can fit more pointers, meaning that the fanout can be larger and the tree needs fewer levels, so data can be accessed more quickly \cite{elmasri_navathe_2011}.
+Databases and file systems often use B+\,trees, especially for indexing, because they can efficiently lookup a value and do a range query.
+Because B+\,trees don’t have to store as much data in the interior nodes, a single node can fit more pointers, meaning that the fanout can be larger and the tree needs fewer levels, so data can be accessed more quickly \cite{elmasri_navathe_2011}.
 A well-designed B+\,tree will store enough values per leaf that a single leaf node just fits into a single storage block, allowing the maximum possible amount of data retrieval with the fewest possible I/O operations.
+Additionally, because all of the values are in the leaves and not buried in higher nodes of the tree, it is relatively easy to do a query for all keys within a certain range.
+Some implementations augment the tree with a linked list between leaf nodes to make range queries even easier.
 
@@
 Security-critical software, while it does not always carry the same risk to human life, does carry the risk of compromising crucial cryptographic tools, data privacy and integrity, and more.
-Much such software is incredibly important to modern technology, but is also very intricate, hard to implement correctly, and hard to keep secure.
+A lot of security-critical software is incredibly important to modern technology, but is also very intricate, hard to implement correctly, and hard to keep secure.
 Formal verification provides a means of reasoning about these tricky tools.
@@
 
-Functional program verification such as the work done by Appel, Filliâtre, and Letouzey serves a practical purpose both because it can help find bugs in current functional libraries (as Filliâtre and Letouzey did with the OCaml Set module based on AVL trees \cite{filliatre:hal-00150913}) and because it can be extracted from Coq into live OCaml code and used in practice.
+Functional program verification such as the work done by Appel, Filliâtre, and Letouzey serves a practical purpose both because it can help find bugs in current functional libraries (as Filliâtre and Letouzey did with the OCaml Set module based on AVL trees \cite{filliatre:hal-00150913}) and because it can be extracted from Coq into live OCaml code to be used in practice.
 However, many practical applications rely on imperative software.
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
     \begin{center}
@@
 
-Separation logic is crucial not just to a proof about a particular program, such as our B+\,trees implementation, but also to the entire verification stack that it is based on.
+Separation logic is crucial not just to a proof about a particular program, such as this B+\,tree implementation, but also to the entire verification stack that it is based on.
 Since the mid-2000s, a group at Princeton University led by Professor Andrew Appel has been developing the \textit{Verified Software Toolchain} \cite{vst_2017}, a stack of verified components that allow proofs to be written about source programs and apply to the compiled program as well.
@@
 Although the scope of this thesis does not extend into connecting the B+\,tree specification to an imperative C program that it represents, the functional specification produced could be useful for the verification of a C program.
-The strategy of functional specification used in this thesis is intended to work with the Verified Software Toolchain, so that a C program could be proven via the VST separation logic to be equivalent to this functional specification.
+The strategy of functional specification used in this thesis is intended to work with the Verified Software Toolchain, so that a C program could be proved via the VST separation logic to be equivalent to this functional specification.
 
 \subsection{DeepSpec}
+\label{subsec:deepspec}
 
 The Verified Software Toolchain itself is part of a larger research group called DeepSpec.
-Since 2016, the DeepSpec group at Princeton, University of Pennsylvania, Yale, and MIT have been collaborating under an NSF Expedition in Computing grant to explore the science of deep specifications.
+Since 2016, the DeepSpec group at Princeton, the University of Pennsylvania, Yale, and MIT have been collaborating under an NSF Expedition in Computing grant to explore the science of deep specifications.
 This thesis is in part an attempt to extend that work by applying deep specification to a B+\,tree database program.
@@
 The strategy that this thesis targets for verifying imperative software is through a \textit{functional model}.
-A functional model describes the actual behavior of the operations, and in this case the model is an equivalent piece of software written in a functional programming language, but other functional models are possible.
+A functional model describes the actual behavior of the operations.
+In this case the model is an equivalent piece of software written in a functional programming language, but other functional models are possible.
 When reasoning about a programming language, for example, the functional model might be the small-step operational semantics of the language.
@@
 One particularity of SQLite is that it represents a cursor into the B+\,tree as a list of pointers to the split point at each level.
-This makes it easier to change the position of the cursor without needing to have a linked list at the leaves as mentioned above or having to retraverse the tree.
+This makes it easier to change the position of the cursor without needing to retraverse the tree.
 
@@
 %Maybe come back and clean this up later%
-\begin{table}[h]
+\begin{table}[ht]
 \centering
@@
 
-%Find a way to position better?%
-%Find a way to italicize captions?%
-\begin{figure}[h]
+\begin{figure}[ht]
     \centering
@@
 \subsection{Lookup}
+\label{subsec:lookup}
 
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
     \centering
@@
 Insertion is simple if the leaf node has less than $b$ children: we simply find the correct place according to the list of keys, and add the value there.
-However, we must maintain that any given node has at most b children.
-Therefore, if a key is inserted which would go into a leaf that has b children already, then we have to split that node and add the middle key to the node above \cite{elmasri_navathe_2011}.
+However, we must maintain that any given node has at most $b$ children.
+Therefore, if a key is inserted which would go into a leaf that has $b$ children already, then we have to split that node and add the middle key to the node above \cite{elmasri_navathe_2011}.
 
@@
 %Could use some work%
-\begin{figure}[h]
+\begin{figure}[ht]
     \centering
@@
 We build this notion into our abstract specification, allowing us to prove correctness relative to this abstract notion of a cursor in addition to a table.
-Here, the specification is shown as written in Coq.
-
-\begin{figure}[h]
+
+\textbf{Figure~\ref{fig:params}} and \textbf{Figure~\ref{fig:axioms}} show the specification as written in Coq.
+The \texttt{Module Type} describes the specification.
+An instance of \texttt{CURSOR\_TABLE} must provide a definition for every \texttt{Parameter} and a proof for every \texttt{Axiom}.
+The implementation developed in this thesis is described in \textbf{Section~\ref{sec:typedef}} and \textbf{Section~\ref{sec:func}}, and its relation to \texttt{CURSOR\_TABLE} is given in \textbf{Section~\ref{sec:proofs}}.
+
+\begin{figure}[hbp]
 \begin{singlespace}
@@
  Parameter empty_t: table.
-\end{minted}
-\end{singlespace}
-\end{figure}
-
-We parametrize our abstract cursor-table with a number of basic types.
-Any cursor-table must store a set of values of type \texttt{V}.
-It stores these within a table of type \texttt{table} (\texttt{empty\_t} being the most basic such table, storing no values) and accesses them using a cursor of type \texttt{cursor}.
-Any implementation of a cursor-table must provide these basic parameters, but the exact types are up to the specific implementation---this makes the module type abstract.
-
-Notice that any cursor-table must also have a \texttt{key} type which is mapped to values.
-For our purposes we have hard-coded this as Coq’s built-in \texttt{Z} type for integers.
-Perhaps it would be better to abstract even further by allowing the key to be any type that has a compare function and satisfies some particular set of axioms; this was in fact the approach taken by Filliâtre and Letouzey in their verification of finite set implementations \cite{filliatre:hal-00150913}.
-However, the simplifying assumption that the keys will be integers is not an unreasonable one, and in particular it fits well with the imperative implementation that this functional specification could be used to verify.
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+
+ (* Functions of the implementation *)
  Parameter make_cursor: key -> table -> cursor.
@@
  Parameter last_cursor: table -> cursor.
-\end{minted}
-\end{singlespace}
-\end{figure}
-
-Any cursor-table must also define a set of operations on the table.
-Again, the specific implementations are left abstract: they are constrained only by their type signatures and the required axioms about them discussed below.
-
-The most central of these functions is the \texttt{make\_cursor} operation, which takes any key and a table and creates a cursor at the appropriate location for that key in the table.
-Such a cursor is then used as input to most other operations.
-The reverse of \texttt{make\_cursor} is \texttt{get\_table}, which translate from a cursor back into the table it points into.
-We also require two shortcuts for creating cursors at the extremes of a table, these are \texttt{first\_cursor} and \texttt{last\_cursor} and are important in the definition of later axioms.
-
-Once we have set a cursor, a cursor-table should be able to use it to do any of the desired operations: \texttt{get} to retrieve the value stored at the cursor, \texttt{get\_key} to retrieve the key mapped to that value, and \texttt{insert} to update the tree with a new key-value pair.
-Deletion is not included as part of this specification, but could be axiomatized similarly to insertion.
-
-The cursor should also be movable from its position: \texttt{next} should move it to the next key greater than the cursor’s current position, while \texttt{prev} should move it to the last key lesser than its current position.
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+
+ (* Predicates of the implementation *)
  Parameter abs_rel: table -> cursor -> Prop.
@@
 \end{minted}
-\end{singlespace}
-\end{figure}
-
-We need various propositions that can describe the correctness and relationship of tables, cursors, and keys as well.
-Specifically, we need four propositions.
-The first, \texttt{abs\_rel}, defines what it means for a particular cursor to be a reference into a particular table, i.e.
-for the table and cursor to represent the same abstract relation.
-The second, \texttt{key\_rel}, defines what it means for a cursor to be positioned for a particular key $k$, i.e.
-that the key to the right of the cursor is greater than $k$ and the key to the left is at most $k$.
-The third,
-\texttt{eq\_cursor}, is a predicate for two cursors being abstractly equivalent even if they are not structurally equal.
-The fourth and fifth describe what it means for a cursor or table to be correct: since we don’t require every instance of the \texttt{cursor} and \texttt{table} types to be correct, it is possible to have a cursor or table that does not actually correspond to a valid instance of a \texttt{CURSOR\_TABLE}.
-For example, with B+\,trees it is possible to construct an unbalanced or unsorted tree, in which case the operations are not guaranteed to be correct.
-It would have been possible to make a more complicated B+\,tree type than the one described later in section \ref{sec:typedef} and enforce the correctness properties on every tree instance.
-However, this would have made working with that type more difficult.
-
-\begin{figure}[h]
+\caption{The parameters in the first half of the Coq module type.}
+\label{fig:params}
+\end{singlespace}
+\end{figure}
+
+A cursor table is a relation that uses positionable cursors to access and update stored values of type \texttt{V}.
+As such, any cursor table implementation must have both a \texttt{table} type to represent an entire relation and a \texttt{cursor} type to act as a pointer into a relation.
+
+\begin{figure}[p]
 \begin{singlespace}
 \begin{minted}{coq}
+ (* Theorems about the predicates *)
  Axiom make_cursor_rel: forall t k,
@@
    abs_rel t c -> (cursor_correct c <-> table_correct t).
-\end{minted}
-\end{singlespace}
-\end{figure}
-
-Given the above relations and propositions, we now want to ensure they satisfy certain desirable properties.
-\texttt{make\_cursor} and \texttt{get\_table} should function correctly relative to our table-cursor relation, as should \texttt{first\_cursor} and \texttt{last\_cursor}.
-Cursor repositioning (\texttt{next} and \texttt{prev}) should stay within the same table.
-Finally, correctness should translate between cursors and tables---if a cursor is correct, then so should its corresponding table be correct; and for any correct table, all cursors into that table should be correct.
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+
+ (* Correctness preservation *)
  Axiom insert_correct: forall k v c,
    cursor_correct c -> key_rel k c -> cursor_correct (insert c k v).
-\end{minted}
-\end{singlespace}
-\end{figure}
-
-Inserting a key-value pair into a correct cursor should in general return a new correct cursor.
-As a consequence of this, by \texttt{correct\_rel} and \texttt{get\_table\_rel} above, it also implies a new table (that is, \texttt{get\_table} of the new cursor) that is itself correct.
-However, notice that we add an additional restriction: the new cursor only has to be correct if it satisfies \texttt{key\_rel k c}, i.e.
-it was positioned properly for the key $k$ that was inserted.
-We don’t actually specify what should happen when inserting a key-value pair into a cursor in the wrong position.
-Depending on the implementation, it could work correctly, return the original cursor back unchanged (as our implementation does), or modify the cursor completely incorrectly.
-It doesn’t ultimately matter---we only require the behavior to be correct when the cursor is positioned correctly.
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+
+ (* Get/insert correctness *)
  Axiom glast: forall t,
@@
    get (make_cursor j (get_table c)).
-\end{minted}
-\end{singlespace}
-\end{figure}
-
-The next three axioms cover how \texttt{get} should behave.
-Specifically, we build up the behavior of \texttt{get} through modifications to the table.
-This reflects the idea that the behavior of \texttt{get} for a certain key should only be affected by certain modifications, and should ignore others.
-If we look up a particular key $k$ for which we have inserted some value $v$, we should get back that value $v$.
-On the other hand, if we look up a key $j$ into a table where we last inserted $k$, and $j$ and $k$ are not in the same range for the cursor $c$, then the behavior of \texttt{get} should be as if $k$ had never been inserted.
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+
+ (* Cursor movement correctness *)
  Axiom next_prev: forall c t,
-   abs_rel t c -> ~ (c = last_cursor t) ->
+   cursor_correct c -> abs_rel t c -> ~ (c = last_cursor t) ->
    eq_cursor c (prev (next c)).
  Axiom prev_next: forall c t,
-   abs_rel t c -> ~ (c = first_cursor t) ->
+   cursor_correct c -> abs_rel t c -> ~ (c = first_cursor t) ->
    eq_cursor c (next (prev c)).
@@
 \end{minted}
-\end{singlespace}
-\end{figure}
+\caption{The axioms in the second half of the Coq module type.}
+\label{fig:axioms}
+\end{singlespace}
+\end{figure}
+
+Notice that any cursor-table must also have a \texttt{key} type which is mapped to values.
+For our purposes we have hard-coded this as Coq’s built-in \texttt{Z} type for integers.
+Perhaps it would be better to abstract even further by allowing the key to be any type that has a compare function and satisfies some particular set of axioms; this was in fact the approach taken by Filliâtre and Letouzey in their verification of finite set implementations \cite{filliatre:hal-00150913}.
+However, the simplifying assumption that the keys will be integers is not an unreasonable one, and in particular it fits well with the imperative implementation that this functional specification could be used to verify.
+
+As suggested by the name of the module type, cursors are central to the operation of any cursor table.
+Appropriately, \texttt{make\_cursor} is the most central operation to the cursor table.
+All the other operations listed in \textbf{Figure~\ref{fig:params}} rely on a cursor produced by either \texttt{make\_cursor} or the related \texttt{first\_cursor} and \texttt{last\_cursor}.
+The functions \texttt{first\_cursor} and \texttt{last\_cursor} are needed to creating a cursor pointing at the very beginning or very end of a list, and these are primarily needed to define later \texttt{Axiom}s about the operation of a cursor table.
+The inverse of \texttt{make\_cursor} is \texttt{get\_table}: it extracts from a cursor the table that it points into.
+
+Once we have set a cursor, a cursor-table should be able to use it to do any of the desired operations: \texttt{get} to retrieve the value stored at the cursor, \texttt{get\_key} to retrieve the key mapped to that value, and \texttt{insert} to update the tree with a new key-value pair.
+Deletion is not included as part of this specification, but could be axiomatized similarly to insertion.
+The cursor should also be movable from its position: \texttt{next} should move it to the next key greater than the cursor’s current position, while \texttt{prev} should move it to the last key lesser than its current position.
+
+Beyond the definition of the implementation itself, a series of predicates about cursors and tables must also be defined.
+The first, \texttt{abs\_rel}, defines what it means for a particular cursor to be a reference into a particular table, i.e.
+for the table and cursor to represent the same abstract relation.
+Of course, this could be as simple as the predicate that \texttt{get\_table} on the cursor returns the table, but we leave room for the particular implementation to provide a more complex cursor-table correspondence that makes later proofs easier (as is the case for this implementation, as described in \textbf{Section~\ref{sec:func}}).
+The second, \texttt{key\_rel}, defines what it means for a cursor to be positioned for a particular key $k$, i.e.
+that the key to the right of the cursor is greater than $k$ and the key to the left is at most $k$.
+The third, \texttt{eq\_cursor}, is a predicate for two cursors being abstractly equivalent even if they are not structurally equal.
+
+The fourth and fifth predicates describe what it means for a cursor or table to be correct: since we don’t require every instance of the \texttt{cursor} and \texttt{table} types to be correct, it is possible to have a cursor or table that does not actually correspond to a valid instance of a \texttt{CURSOR\_TABLE}.
+For example, with B+\,trees it is possible to construct an unbalanced or unsorted tree, in which case the operations are not guaranteed to be correct.
+In this implementation, it would have been possible to make a more complicated B+\,tree type than the one described in \textbf{Section~\ref{sec:typedef}} and enforce the correctness properties on every tree instance.
+However, this would have made working with that type more difficult.
+
+Given the operation and predicate \texttt{Parameters}, we now want to ensure they satisfy certain desirable properties.
+These are the \texttt{Axiom}s of the module type shown in \textbf{Figure~\ref{fig:axioms}}.
+The first set of axioms correspond to the predicates themselves, and how they should behave.
+For example, \texttt{make\_cursor} and \texttt{get\_table} should function correctly relative to our table-cursor correspondence \texttt{abs\_rel}, as should the functions for the ends, \texttt{first\_cursor} and \texttt{last\_cursor}.
+Cursor repositioning (\texttt{next} and \texttt{prev}) should stay within the same table.
+Finally, correctness should translate between cursors and tables---if a cursor is correct, then so should its corresponding table be correct; and for any correct table, all cursors into that table should be correct.
+
+The axiom \texttt{insert\_correct} dictates that inserting a key-value pair into a correct cursor should in general return a new correct cursor.
+As a consequence of this, by \texttt{correct\_rel} and \texttt{get\_table\_rel} above, there must exist a new table (that is, \texttt{get\_table} of the new cursor) that is itself correct.
+However, notice that we add an additional restriction: the new cursor only has to be correct if it satisfies \texttt{key\_rel k c}, i.e.
+it was positioned properly for the key $k$ that was inserted.
+We don’t actually specify what should happen when inserting a key-value pair into a cursor in the wrong position.
+Depending on the implementation, it could work correctly, return the original cursor back unchanged (as our implementation does), or modify the cursor completely incorrectly.
+It doesn’t ultimately matter---we only \textit{require} the behavior to be correct when the cursor is positioned correctly.
+
+The next three axioms in \textbf{Figure~\ref{fig:axioms}} cover how \texttt{get} should behave in conjunction with \texttt{insert}.
+Specifically, we build up the behavior of \texttt{get} through modifications to the table.
+This reflects the idea that the behavior of \texttt{get} for a certain key should only be affected by certain modifications, and should ignore others.
+If we look up a particular key $k$ for which we have inserted some value $v$, we should get back that value $v$.
+On the other hand, if we look up a key
+$j$ into a table where we last inserted $k$, and $j$ and $k$ are not in the same range for the cursor $c$, then the behavior of \texttt{get} should be as if $k$ had never been inserted.
 
@@
 In the reverse direction, for any valid cursor except the first one, moving to the previous position and back gives us an equivalent cursor to the original cursor; this is \texttt{prev\_next}.
-Finally, we expect \texttt{next} and \texttt{prev} to position the cursor relative to the ordering of the table.
+We also expect \texttt{next} and \texttt{prev} to position the cursor relative to the ordering of the table.
 If we call \texttt{next}, we expect the “next” key that it moves to to be the key immediately greater than the one the cursor was previously on.
@@
 \subsection{A Small Example} \label{subsec:ex}
+\setlength{\parskip}{0em}
 
@@
 Suppose we have a small relation built up by inserting $(3,V_3)$, then $(1,V_1)$, and finally $(2,V_2)$.
-Functionally, this might look like the following:
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
-insert (insert (insert (make_cursor 0 empty_t) 3 V_3) 1 V_1) 2 V_2
-\end{minted}
-\end{singlespace}
-\end{figure}
+Functionally, this would look like the following:
+
+\begin{center}\texttt{
+    insert (insert (insert (make\_cursor 0 empty\_t) 3 V\_3) 1 V\_1) 2 V\_2
+}\end{center}
 
@@
 
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
-get (make_cursor 1 (get_table c))
-\end{minted}
-\end{singlespace}
-\end{figure}
+\begin{center}\texttt{
+    get (make\_cursor 1 (get\_table c))
+}\end{center}
 
@@
 
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
-get (make_cursor 1 (get_table (insert c 2 V_2))) =
-get (make_cursor 1 (get_table c))
-\end{minted}
-\end{singlespace}
-\end{figure}
+\begin{center}\texttt{
+    get (make\_cursor 1 (get\_table (insert c 2 V\_2))) \\
+    = \\
+    get (make\_cursor 1 (get\_table c))
+}\end{center}
 
@@
 
-%For some reason the figure is making this mess up??
-%\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
-get (make_cursor 1 (get_table (insert c 1 V_1))) = Some V_1
-\end{minted}
-\end{singlespace}
-%\end{figure}
+\begin{center}\texttt{
+    get (make\_cursor 1 (get\_table (insert c 1 V\_1))) = Some V\_1
+}\end{center}
 
@@
 
+\setlength{\parskip}{1em}
 \clearpage
@@
 
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
 Inductive tree : Type :=
@@
  | tl_cons : tree -> treelist -> treelist
-\end{minted}
-\end{singlespace}
-\end{figure}
+\end{cminted}
+\caption{The tree and treelist type definitions.}
+\end{figure}
+\end{singlespace}
 
@@
 \subsection{B+ Tree Cursor Type}
+\label{subsec:cursortype}
 
@@
 
-The representation used was to structure a pointer into a treelist as the combination of that treelist and the index of the cursor’s position within it.
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
+Definition cursor : Type := list nat * list treelist.
+\end{cminted}
+\caption{The cursor type definition.}
+\end{figure}
+\end{singlespace}
+
+In this representation, a pointer into a treelist is structured as the combination of that treelist and the index of the cursor’s position within it.
 The \texttt{cursor} type is then a list of indices combined with a list of treelists.
-This cursor structure was chosen to make proofs more straightforward, because it contains the original tree structure unmodified, and the \texttt{make\_cursor} operation structurally matches the \texttt{treelist} type (this is explained in more detail in section \ref{subsec:cursorop}).
-However, other operations are slightly complicated because of the choice to use indices: every step requires destructing the treelist to find the right position before we can get or update the keys and values stored there.
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
-Definition cursor : Type := list nat * list treelist.
-\end{minted}
-\end{singlespace}
-\end{figure}
+This cursor structure was chosen to make proofs more straightforward, because it contains the original tree structure unmodified, and the \texttt{make\_cursor} operation structurally matches the \texttt{treelist} type (this is explained in more detail in \textbf{Section~\ref{subsec:cursorop}}).
+However, other operations are slightly complicated because of the choice to use indices rather than pulling out the relevant tree or encoding the pointer in some other structural way.
+Every step in any operation requires destructing the treelist to find the right position before we can get or update the keys and values stored there.
 
@@
 when looking up a value), by default we assume it is the first element to the right of the cursor’s position.
-However, it is important to note that the cursor does not need to be on a particular element.
-For example, a cursor into an entirely empty table is completely valid, and is represented as any index into the treelist \texttt{tl\_nil}.
-
-This adds a potential point for confusion: two \textit{structurally} different cursors could \textit{abstractly} represent the same split. Consider the following example:
-
-%Do this%
----------Need a graphic for this---------
-
-Both arrows represent the split between keys 2 and 6 in this B+\,tree---the first could be the result of \texttt{make\_cursor 3 f}, and the second the result of \texttt{make\_cursor 5 f}.
+However, it is important to note that the cursor does not need to have any element to its right.
+For example, a cursor into an entirely empty table is completely valid, and is represented as index 0 into the treelist \texttt{tl\_nil}.
+
+This adds a potential point for confusion: two \textit{structurally} different cursors could \textit{abstractly} represent the same split.
+Consider the example in \textbf{Figure~\ref{fig:equivcursor}}.
+Both arrows represent the split between keys 2 and 3 in this B+\,tree.
 Abstractly they are equivalent, and any operations on them should do the same thing.
-But in our implementation they are not actually equal structures.
-So we have to ensure (and prove) that all of our B+\,tree operations can detect this and treat it appropriately.
+But in this implementation they are not actually equal structures, so we have to ensure (and prove) that all of our B+\,tree operations can detect this and treat it appropriately.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.7\textwidth]{equivcursor.jpg}
+    \caption{The white and grey cursors should be equivalent.}
+    \label{fig:equivcursor}
+\end{figure}
 
@@
 
-\subsection{Element-List Abstraction}
-
-The question of proving that our B+\,tree operations treat cursors appropriately relies on an intuitive understanding of how cursors should behave that we now need to formalize.
-Within our implementation, a B+\,tree cursor is a list of indices and treelists.
-In the abstract specification, a B+\,tree cursor is only defined by how our operations use it.
-This leaves a gap in formalization that makes it difficult to prove our operations correct or prove that they use cursors appropriately.
-
-To solve this problem, we introduce an intermediate abstraction of cursors as an element-list.
-An element, in this sense, is any key-value pair that appears in the B+\,tree.
-The tree itself then is a list of these elements in order by increasing key.
-A cursor splits that list into the parts before the cursor’s position and the parts after it, and to simplify our operations we reverse the order of the first list.
-In other words, the \textit{element-list abstraction} of a cursor is a pair of lists, one to the “left” and one to the “right”, where the first element in each list is the element immediately adjacent to the cursor on each side, and the last element in each list is the extreme first or last element of the tree.
-
-It is clear that under this abstraction, functionally equivalent cursors are represented the same.
-If two structurally different cursors in the tree point to the same split, then they must have the same element-list abstraction.
-Furthermore, our operations on these element-lists are very simple.
-To lookup the value at a cursor, we simply pop an item off the right list and return its value.
-To insert a new key-value pair we just add it to the left list (placing the cursor just after it); to update the value for a key already in the list we just replace the value at the beginning of the right list.
-Moving the cursor position is just popping items off of one list and concatenating them to the other.
-
 \subsection{Cursor Operations} \label{subsec:cursorop}
 
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+A typical implementation of a relational table would have a lookup function that would find a key in the table, such as the operation discussed in \textbf{Section~\ref{subsec:lookup}}.
+In an array that function might be a linear or binary search.
+In a binary search tree it would require descending into the left or right subtree of a node depending on whether the key being looked up is less than or greater than that node’s key.
+In this B+\,tree implementation, \texttt{make\_cursor} fills that role.
+However, rather than simply returning the value held at that key it instead returns a cursor that points to the part of the B+\,tree where the key is or would be located.
+
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
 Function make_cursor_rec (x: key) (f : treelist) (ci : list nat)
@@
   make_cursor_rec x f [] [f] O.
-\end{minted}
-\end{singlespace}
-\end{figure}
-
+\end{cminted}
+\caption{The cursor creation function.}
+\end{figure}
+\end{singlespace}
+
+The structure of the function is similar to what lookup would be in a binary search tree, but as it destructs the tree to find the subtree to descend into it constructs a record of where it went.
 Creating a cursor to point at a particular key $x$ happens in two stages.
-Within a particular treelist, we scan across each node, comparing the node’s key k against the target key x.
+Within a particular treelist, we scan across each node, comparing the node’s key $k$ against the target key $x$.
 This is what is happening in the \texttt{match} section: the treelist is decomposed horizontally into its component trees.
@@
 Since functional objects have no pointers and must be constructed from the data they hold, it is impossible to have an infinitely large tree, so it is easy for Coq to automatically prove the termination of this function.
-Similarly, when we are proving properties of \texttt{make\_cursor}, having a function that destructs the treelists naturally fits precisely with our induction scheme:
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+Similarly, when we are proving properties of \texttt{make\_cursor}, having a function that destructs the treelists naturally fits precisely with our induction scheme, shown in \textbf{Figure~\ref{fig:induction}}.
+
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
 treelist_tree_rec :
@@
 forall t : treelist, P0 t.
-\end{minted}
-\end{singlespace}
-\caption{The induction scheme requires a property \texttt{P} that is inductively assumed about every tree, and a property \texttt{P0} that is assumed and proved about every treelist.
+\end{cminted}
+\caption{The induction scheme requires a predicate \texttt{P} that is inductively assumed about every tree, and a predicate \texttt{P0} that is assumed and proved about every treelist.
 Having functions destruct a treelist in both directions matches this scheme well.}
@@
 \end{figure}
+\end{singlespace}
 
@@
 For this purpose, we wrote two normalization functions.
-The first, \texttt{next\_node}, normalizes a cursor so it points directly to a \texttt{val} tree; this would be the position pointed to by \texttt{make\_cursor 5} in the example tree from the previous section.
+The first, \texttt{next\_node}, normalizes a cursor so it points directly to a \texttt{val} tree; this would be the position pointed to by the grey arrow in \textbf{Figure~\ref{fig:equivcursor}} from \textbf{Section~\ref{subsec:cursortype}}.
 The second, \texttt{prev\_node}, is the reverse: it normalizes a cursor so it has a val tree directly behind it, i.e.
-the \texttt{make\_cursor 3} position in the example tree.
-For both functions, the functions should change the structure of the cursor without changing the abstract element-list of the cursor.
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+the white arrow in \textbf{Figure~\ref{fig:equivcursor}}.
+
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
 Fixpoint next_node (cn : list nat) (cf : list treelist)
@@
   end.
-\end{minted}
-\end{singlespace}
-\end{figure}
+\end{cminted}
+\caption{The normalization to point directly to a \texttt{val} tree. The structure of \texttt{prev\_node} is similar.}
+\end{figure}
+\end{singlespace}
 
@@
 If it is at the beginning of a node and there is nothing directly preceding it to point to, then it recursively calls \texttt{prev\_node} at the next level up in the cursor.
-The partial cursor it gets back is then grown with the treelist $f’$ pointed to and a pointer to the last element of $f’$.
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+The partial cursor it gets back is then grown with the treelist $f'$ pointed to and a pointer to the last element of $f'$.
+
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
 Fixpoint move_to_next (c : cursor) : cursor :=
@@
   end.
-\end{minted}
-\end{singlespace}
-\end{figure}
+\end{cminted}
+\caption{The Coq code for moving the cursor. With the normalization functions, actual cursor movement is simple.}
+\end{figure}
+\end{singlespace}
 
@@
 
-Furthermore, the correctness of these functions follows directly from the correctness of the normalization functions.
-Given proofs that \texttt{next\_node} and \texttt{prev\_node} both produce correct cursors, then \texttt{move\_to\_next} and \texttt{move\_to\_prev} clearly produce correct cursors as well because the only modification they make is in the last index of the cursor.
-We also prove that both \texttt{next\_node} and \texttt{prev\_node} preserve the element-list, so that they position the cursor before or after an element without changing its abstract representation.
-Then it is intuitive to prove that the element-list of \texttt{move\_to\_next} of a cursor c is the original cursor’s element-list, but with the first element of the “right” list moved to the “left” list.
-Similarly, \texttt{move\_to\_prev} is the original element-list, but with the first element of the “left” list moved to the “right” list.
-If there is no right (or respectively left) element, this implies that the cursor is at the very end (or very beginning) of the tree, and we prove that \texttt{next\_node} and \texttt{prev\_node} return \texttt{None} in that case, so \texttt{move\_to\_next} and \texttt{move\_to\_prev} just return the original cursor.
-
-These properties lead directly to proofs of the \texttt{next\_prev} and \texttt{prev\_next} properties in the abstract specification: if the cursor is not the last cursor, it must have an element to its right in its element-list; then \texttt{move\_to\_next} moves that element to the left list, and \texttt{move\_to\_prev} moves it back to the right.
-Similarly, if a cursor is not the first cursor, \texttt{move\_to\_prev} moves an element to the right list, and \texttt{move\_to\_next} moves it back to the left.
+\begin{figure}
+    \centering
+    \includegraphics[width=0.7\textwidth]{movecursor.jpg}
+    \caption{An example of moving the cursor.}
+    \label{fig:movecursor}
+\end{figure}
+
+For example, consider the first (white) cursor in \textbf{Figure~\ref{fig:movecursor}}.
+\texttt{move\_to\_next} would first normalize it to the position of the grey cursor, then increment the counter to position it at the black cursor.
+If we then called \texttt{move\_to\_prev} from the black cursor, either the white or the grey cursor could be a correct result; in this implementation it would return the grey cursor.
 
@@
 
-To simplify proofs of search properties, we have combined the core of \texttt{get} and \texttt{get\_key} into a single function \texttt{get\_tree} from which the key and value can easily be retrieved.
+To simplify proofs of search properties, we have combined the core of both \texttt{get} and \texttt{get\_key} into a single function \texttt{get\_tree} from which the key and value can easily be retrieved.
 This allows us to prove properties of \texttt{get\_tree} once, and easily relate them to both \texttt{get} and \texttt{get\_key}.
 
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
 Fixpoint get_tree (c : cursor) : option tree :=
@@
   end.
-\end{minted}
-\end{singlespace}
-\end{figure}
-
-Similarly to the \texttt{move\_to\_next} and \texttt{move\_to\_prev} operations, the \texttt{get\_tree} operation is vastly simplified by having set up \texttt{next\_node} and the element-list abstraction.
+\end{cminted}
+\caption{The generalized search function.}
+\end{figure}
+\end{singlespace}
+
+Similarly to the \texttt{move\_to\_next} and \texttt{move\_to\_prev} operations, the \texttt{get\_tree} operation is vastly simplified by having set up \texttt{next\_node}.
 We normalize with \texttt{next\_node} to ensure that the cursor points directly at an element, and then we can return that element.
-If \texttt{next\_node} doesn’t return a valid cursor, then we must be at the end of the tree, so there is no element for get to return.
-
-The correctness property of this function relative to the element-list abstraction is simple, and can be easily applied to the properties of the abstract specification.
-Specifically, if $(k,v)$ is the element at the front of the right element-list of a (correct) cursor, then \texttt{get\_tree} should return \texttt{val k v}.
-If the cursor is correct---including that it points into a leaf node---then \texttt{next\_node} must be correct as well, so the treelist that \texttt{next\_node} returns will necessarily be a leaf node, and the tree that \texttt{get\_tree} returns must necessarily be a \texttt{val} tree.
-Furthermore, given that \texttt{next\_node} positions the cursor to point a tree without changing the element-list for the cursor, we know both that \texttt{get\_tree} will return that tree and that it must be at the front of the right element-list of \texttt{next\_node} of the cursor (and therefore of the original cursor as well).
-Therefore, the tree returned by \texttt{get\_tree} and the element at the beginning of the right element-list must be the same.
+If \texttt{next\_node} doesn’t return a valid cursor, then we must be at the end of the tree, so there is no element for \texttt{get} to return.
 
@@
 There are two possibilities for insertion.
-If the key is not already in the tree we want to insert it and position the cursor immediately after it (this makes it easy to quickly create a new B+\,tree or insert a range of key-value pairs).
+If the key is not already in the tree then we want to insert the key-value pair and position the cursor immediately after it.
+(Positioning the new cursor after the pair inserted makes it easy to quickly create a new B+\,tree or insert a range of key-value pairs).
 If the key is already in the tree we want to simply update the value and leave the cursor positioned as is.
@@
 
-By comparison, the element-list abstraction of insertion is quite simple.
-Let $l$ be the left element-list, and let $(k,v)::r$ be the right element-list.
-To insert a key-value pair $(x,v’)$, if $k$ and $x$ are not equal, then it should result in element-lists of $(x,v’)::l$ and $(k,v)::r$.
-If $k$ and $x$ are equal, then the result should be $l$ and $(k,v’)::r$.
-
-The functional B+\,tree implementation of insertion has to achieve several more goals.
-Most important, it has to maintain the fanout value $b$, as explained in section \ref{sec:b+tree}, so that every treelist contains between $\lceil \frac{b}{2} \rceil$ and $b$ subtrees (except for potentially the root).
+Additionally, the insertion implementation has to maintain several B+\,tree invariants.
+The most complicated of these is maintaining the fanout value $b$, as explained in \textbf{Section~\ref{sec:b+tree}}, so that every treelist contains between $\lceil \frac{b}{2} \rceil$ and $b$ subtrees (except for potentially the root).
 This means that inserting a key-value pair might require splitting the leaf it was inserted into, and inserting the new leaf into its parent node might require splitting that node, and so on up to the root.
@@
 
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
 Inductive splitpair : Type :=
@@
 | Two : treelist -> key -> treelist -> splitpair.
+\end{cminted}
+\caption{The type used for splitting too-large treelists during insertion.}
+\end{figure}
+\end{singlespace}
+
+A treelist that is the correct size and is not split will result in a \texttt{One}.
+If a treelist is too large and must be split then it results in a \texttt{Two} containing each half of the split treelist as well as the key differentiating them that will be inserted up into the parent node.
+
+\begin{singlespace}
+\begin{figure}[hpt]
+\begin{minted}{coq}
+Fixpoint insert_across (s : splitpair) (f' : treelist) (n : nat)
+  : treelist :=
+  match f' with
+  | tl_cons (node k f) f' =>
+    (match n with
+     | O =>
+       (match s with
+        | One f => tl_cons (node k f) f'
+        | Two f1 k' f2 => tl_cons (node k' f1) (tl_cons (node k f2) f')
+        end)
+     | S n' => tl_cons (node k f) (insert_across s f' n')
+     end)
+  | tl_cons (final f) tl_nil =>
+    (match n with
+     | O =>
+       (match s with
+        | One f => tl_cons (final f) tl_nil
+        | Two f1 k' f2 => tl_cons (node k' f1) (tl_cons (final f2) f')
+        end)
+     | S n' => tl_cons (final f) tl_nil
+     end)
+  | _ => tl_nil (* Shouldn't ever be hit. *)
+  end.
+
+Fixpoint insert_up (s : splitpair) (cn : list nat) (cf : list treelist)
+  : cursor :=
+  match (cn,cf) with
+  | (n::cn,f::cf) =>
+    (match decide_split (insert_across s f n) with
+     | One f => (match insert_up (One f) cn cf with (cn,cf) =>
+                (n::cn,f::cf) end)
+     | Two f1 k f2 =>
+       (match insert_up (Two f1 k f2) cn cf with (cn,cf) =>
+        if (Nat.leb (treelist_length f1) n)
+        then ((n-(treelist_length f1))::cn, f2::cf)
+        else (n::cn,f1::cf) end)
+     end)
+  | (_,_) => ([],[])
+  end.
 \end{minted}
-\end{singlespace}
-\end{figure}
-
-A treelist that is the correct size and is not split will result in a \texttt{One}, but if a treelist is too large and must be split then it results in a \texttt{Two} containing each half of the split treelist as well as the key differentiating them that will be inserted up into the parent node.
+\caption{Some of the more important sections of the insertion code.}
+\label{fig:insert}
+\end{figure}
+\end{singlespace}
 
@@
 For each level of the treelist, we receive a splitpair from the level below.
-We use the function \texttt{insert\_across} to place that splitpair’s treelist(s) into the position indicated by the cursor, then the function \texttt{insert\_up} (potentially) splits the new treelist and passes it up to be inserted in the next level of the cursor.
+As shown in \textbf{Figure~\ref{fig:insert}}, we use the function \texttt{insert\_across} to place that splitpair’s treelist(s) into the position indicated by the cursor, then the function \texttt{insert\_up} (potentially) splits the new treelist and passes it up to be inserted in the next level of the cursor.
 The result it gets back then has the appropriate index and treelist appended back onto it.
-Since the cursor carries with it the whole structure of the B+\,tree, the whole cursor must be destructed and reconstructed with the update B+\,tree, even if no node needs to be split.
-
-\subsection{Correctness Theorem}
-
-We need one final theorem that pulls all of the other results together and states that the functional model satisfies all of the properties in the \texttt{Module}.
-In Coq this is completed by filling out an instance of the \texttt{CURSOR\_TABLE} module:
-
-\begin{figure}[h]
-\begin{singlespace}
-\begin{minted}{coq}
+\texttt{decide\_split} is a function that compares the length of the treelist to the fanout, and splits it in half if needed.
+Note that since the cursor carries with it the whole structure of the B+\,tree, the whole cursor must be destructed and reconstructed with the updated B+\,tree, even if no node needs to be split.
+
+\clearpage
+
+\section{Proofs}
+\label{sec:proofs}
+
+This section runs through the application of the functional implementation from \textbf{Section~\ref{sec:func}} to the abstract specification from \textbf{Section~\ref{sec:spec}}, as well as the intermediate lemmas used to get there.
+It includes all of the most important lemmas and theorems stated in the Coq files.
+Not every lemma or theorem stated here has been fully proved; the breakdown of what has and hasn't been proved can be found in \textbf{Appendix~\ref{app:proofs}} along with the associated Coq theorem statement.
+Additionally, not every lemma or theorem is included here; the more minor ones used can be found in the complete Coq files in \textbf{Appendix~\ref{app:code}}.
+
+\subsection{Correctness Predicates}
+
+Various predicates about the correctness of cursors and trees are useful for proving the correctness of operations on those trees.
+The statement of these predicates in Coq is presented here, along with a description of what the predicate entails.
+
+\begin{singlespace}
+\begin{figure}[hpt]
+\centering
+\begin{cminted}{coq}
+Inductive treelist_sorted : key -> key -> treelist -> Prop :=
+| ts_nil : forall ki kf, treelist_sorted ki kf tl_nil
+| ts_node : forall (ki ki' kf k : key) (f f' : treelist),
+    treelist_sorted ki' kf f -> (* forall x in f, ki' < x <= kf *)
+    treelist_sorted ki k f' -> (* forall x in f', ki < x <= k *)
+    lt_key ki' k = false -> (* k <= ki' *)
+    lt_key ki k = true -> (* ki < k *)
+    treelist_sorted ki kf (tl_cons (node k f') f)
+| ts_final : forall (ki ki' kf : key) (f f' : treelist),
+    treelist_sorted ki' kf f -> (* forall x in f, ki' < x <= kf *)
+    treelist_sorted ki ki' f' -> (* forall x in f', ki < x <= ki' *)
+    lt_key ki ki' = true -> (* ki < ki' *)
+    treelist_sorted ki kf (tl_cons (final f') f)
+| ts_val : forall (ki ki' kf k : key) (v : V) (f : treelist),
+    treelist_sorted ki' kf f -> (* forall x in f, ki' < x <= kf *)
+    lt_key ki' k = false -> (* k <= ki' *)
+    lt_key ki k = true -> (* ki < k *)
+    treelist_sorted ki kf (tl_cons (val k v) f).
+
+Definition sorted (f : treelist) : Prop :=
+  exists ki kf, treelist_sorted ki kf f.
+\end{cminted}
+\caption{The inductive property that all the keys of a treelist are ordered.}
+\label{fig:sorted}
+\end{figure}
+\end{singlespace}
+
+\textbf{Figure~\ref{fig:sorted}} shows the inductive \texttt{Prop} for the keys of a treelist being sorted in a certain range, which is \texttt{treelist\_sorted}.
+It also shows the more high-level \texttt{sorted}, which represents the idea that there is \textit{some} range in which a treelist is sorted.
+The four cases of the inductive proposition follow the four possible treelist-tree combinations.
+Any treelist that is just \texttt{tl\_nil} is sorted within any range.
+For a \texttt{node} or a \texttt{final} tree within a \texttt{tl\_cons} treelist, we build up sortedness based on the sortedness of the subtree, the sortedness of the remainder of the treelist, and (for \texttt{node}) the key of the tree.
+For a \texttt{val} tree, the tree's key must be within the new range and outside the range of the remainder of the treelist.
+
+\begin{singlespace}
+\begin{figure}[hpt]
+\centering
+\begin{cminted}{coq}
+Inductive balanced_treelist : nat -> treelist -> Prop :=
+| bf_nil : balanced_treelist 1 tl_nil
+| bf_val : forall k v f,
+    balanced_treelist 1 f -> (* f is a val treelist *)
+    balanced_treelist 1 (tl_cons (val k v) f)
+| bf_node : forall n k f f',
+    balanced_treelist n f -> (* f is balanced with n levels *)
+    balanced_treelist (S n) f' -> (* f' is balanced with n+1 levels *)
+    balanced_treelist (S n) (tl_cons (node k f) f')
+| bf_final : forall n f,
+    balanced_treelist n f -> (* f is balanced with n levels *)
+    balanced_treelist (S n) (tl_cons (final f) tl_nil).
+
+Definition balanced (f : treelist) : Prop :=
+  exists n, balanced_treelist n f.
+\end{cminted}
+\caption{The inductive property that all leaves in a treelist are the same distance from the root.}
+\label{fig:balanced}
+\end{figure}
+\end{singlespace}
+
+Intuitively, the property of a B+\,tree being balanced shown in \textbf{Figure~\ref{fig:balanced}} means that every leaf node is the same distance away from the root.
+However, since functional treelists are built from the bottom up, we flip that property: a balanced B+\,tree is when the treelist is the same distance $n$ from all leaves in its subtrees.
+We encode this property by giving any \texttt{val} tree a height of 1.
+Then any treelist \texttt{tl\_cons t f} has one more than the height of $t$, provided that $f$ also has that same height.
+
+\begin{singlespace}
+\begin{figure}[hpt]
+\centering
+\begin{cminted}{coq}
+Inductive fanout_restr : nat -> treelist -> Prop :=
+| fr_nil : fanout_restr O tl_nil
+| fr_val : forall n f k v,
+    n < b ->
+    fanout_restr n f ->
+    fanout_restr (S n) (tl_cons (val k v) f)
+| fr_node : forall n n' k f f',
+    n' > div_two b false ->
+    fanout_restr n' f' ->
+    n < b ->
+    fanout_restr n f ->
+    fanout_restr (S n) (tl_cons (node k f') f)
+| fr_final : forall n f,
+    n > div_two b false ->
+    fanout_restr n f ->
+    fanout_restr 1 (tl_cons (final f) tl_nil).
+
+Definition fanout (f : treelist) : Prop :=
+  exists n, fanout_restr n f.
+\end{cminted}
+\caption{The property that any node in a treelist has between $\lceil \frac{b}{2} \rceil$ and $b$ subtrees.}
+\label{fig:fanout}
+\end{figure}
+\end{singlespace}
+
+The last treelist property is that of the fanout restriction shown in \textbf{Figure~\ref{fig:fanout}}: every node in a B+\,tree with fanout $b$ must have between $\lceil \frac{b}{2} \rceil$ and $b$ subtrees.
+The \texttt{nat} parameter inductively builds up the length of the treelist, and can grow to any size between $0$ and $b$.
+However, we restrict it to being at least $\frac{b}{2}$ long when it is used as a subtree in a higher treelist.
+This makes the inductive \texttt{Prop} easy to construct, and also easily encodes the property that the root is allowed to have very few children.
+
+\begin{singlespace}
+\begin{figure}[hpt]
+\centering
+\begin{cminted}{coq}
+Inductive cursor_correct_struct : cursor -> Prop :=
+| cc_nil : cursor_correct_struct ([],[])
+| cc_first : forall n f, cursor_correct_struct ([n],[f])
+| cc_node : forall n n' k f f' ci ct,
+    cursor_correct_struct (n::ci,f::ct) ->
+    lin_search n f = Some (node k f') ->
+    cursor_correct_struct (n'::n::ci,f'::f::ct)
+| cc_final : forall n n' f f' ci ct,
+    cursor_correct_struct (n::ci,f::ct) ->
+    lin_search n f = Some (final f') ->
+    cursor_correct_struct (n'::n::ci, f'::f::ct).
+(* add that it ends at a leaf *)
+(* first f should be correct *)
+
+Inductive rec_prop (P : treelist -> Prop) : cursor -> Prop :=
+| rp_nil : rec_prop P ([],[])
+| rp_next : forall n f cn cf,
+    P f -> rec_prop P (cn,cf) -> rec_prop P (n::cn,f::cf).
+
+Definition cursor_correct (c : cursor) : Prop :=
+  cursor_correct_struct c /\
+  rec_prop balanced c /\
+  rec_prop sorted c /\
+  rec_prop fanout c.
+\end{cminted}
+\caption{Structural and treelist correctness properties of cursors.}
+\label{fig:cursorcorrect}
+\end{figure}
+\end{singlespace}
+
+\textbf{Figure~\ref{fig:cursorcorrect}} shows the various cursor correctness predicates.
+For a cursor to be \textit{structurally} correct means that each level of the lists is properly built from the previous one.
+If one level gives a treelist $f$ and an index $n$, then the treelist $f'$ at the next level should actually be contained at that index $n$ in the previous treelist $f$.
+
+A correct cursor also encapsulates the correctness of the B+\,tree it references.
+The inductive predicate \texttt{rec\_prop} allows any treelist predicate to be applied over all treelists in the cursor.
+This allows us to easily say that a correct cursor should not only be structurally correct, but should also reference a treelist that is balanced, sorted, and fanout-restricted.
+
+\subsection{Element-List Abstraction}
+
+The task of proving that our B+\,tree operations treat cursors appropriately relies on an intuitive understanding of how cursors should behave that we now need to formalize.
+Within our implementation, a B+\,tree cursor is a list of indices and treelists.
+In the abstract specification, a B+\,tree cursor is only defined by how our operations use it.
+This leaves a gap in formalization that makes it difficult to prove our operations correct or prove that they use cursors appropriately.
+
+To solve this problem, we introduce an intermediate abstraction of cursors as an element-list.
+An element, in this sense, is any key-value pair that appears in the B+\,tree.
+The entire tree can be thought about as a list of these elements in order by increasing key.
+A cursor splits that list into the parts before the cursor’s position and the parts after it.
+Finally, to simplify our operations we reverse the order of the first list.
+In other words, the \textit{element-list abstraction} of a cursor is a pair of lists, one to the “left” and one to the “right”, where the first element in each list is the element immediately adjacent to the cursor on each side, and the last element in each list is the extreme first or last element of the tree.
+
+It is clear that under this abstraction, functionally equivalent cursors are represented the same.
+If two structurally different cursors in the tree point to the same split, then they must have the same element-list abstraction.
+Furthermore, our operations on these element-lists are very simple.
+To lookup the value at a cursor, we simply pop an item off the right list and return its value.
+To insert a new key-value pair we just add it to the left list (placing the cursor just after it); to update the value for a key already in the list we just replace the value at the beginning of the right list.
+Moving the cursor position is just popping items off of one list and concatenating them to the other.
+
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
+Fixpoint cursor_right (cn : list nat) (cf : list treelist)
+  (base : list (key * V)) : list (key * V) :=
+  match (cn,cf) with
+  | (n::cn,f::cf) =>
+    (match point n f with (_,_,f') =>
+    cursor_right cn cf (right_el f' base) end)
+  | (_,_) => base
+  end.
+
+Fixpoint cursor_left (cn : list nat) (cf : list treelist)
+  (base : list (key * V)) : list (key * V) :=
+  match (cn,cf) with
+  | (n::cn,f::cf) =>
+    (match point n f with (f',_,_) =>
+    cursor_left cn cf (left_el f' base) end)
+  | (_,_) => base
+  end.
+
+Fixpoint cursor_elements' (cn : list nat) (cf : list treelist)
+  (l : list (key * V)) (r : list (key * V))
+  : (list (key * V)) * (list (key * V)) :=
+  match (cn,cf) with
+  | (n::cn,f::cf) =>
+    (match point n f with (f1,_,f2) =>
+    cursor_elements' cn cf (left_el f1 l) (right_el f2 r) end)
+  | (_,_) => (l,r)
+  end.
+\end{cminted}
+\caption{The functions used to generate the element-list abstraction of a cursor.}
+\label{fig:el}
+\end{figure}
+\end{singlespace}
+
+In order to actually work with this abstraction, we have to create a functional way to translate between a cursor and its abstraction.
+The element-lists can be generated by the function \texttt{cursor\_elements} shown in \textbf{Figure~\ref{fig:el}}, but since it is easier to reason in proofs about one side of the element-list at a time, \texttt{cursor\_left} and \texttt{cursor\_right} do the two sides separately.
+The following lemma describes this fact:
+
+\begin{lemma}[\texttt{cursor\_elements'\_sides\_equiv}]
+The pair that \texttt{cursor\_elements} returns is equal to the pair of \texttt{cursor\_left} and \texttt{cursor\_right}.
+\end{lemma}
+
+Both the right and left functions are called recursively and gradually build up a ``base'' of elements already added.
+For both directions, we process elements from the inside outward.
+For this, we have the following lemmas:
+
+\begin{lemma}[\texttt{left\_rec\_interior}; \texttt{right\_rec\_interior}]
+For both the left and right functions, the ``base'' they are passed ends up at the beginning of the list, i.e.
+if the base is \texttt{b} and the end result is \texttt{l}, then there is some \texttt{l'} such that \texttt{l = b++l'}.
+\end{lemma}
+
+We also want to show that cursors are processed from the bottom up, and that the things reached from the bottom (which are consequently closer to the cursor position) end up in the base first.
+
+\begin{lemma}[\texttt{cursor\_right\_elements1}; \texttt{cursor\_left\_elements1}]
+For both the left and right functions, if we split a cursor into two parts, then the result of computing the elements of the whole cursor is equal to the result of first computing the elements of the first part, then using that as the base for computing the elements of the second part.
+\end{lemma}
+
+Finally, we need to prove that the element-lists preserve the sortedness of the tree. (B+\,tree balance and fanout are irrelevant to the element-list, since this abstraction flattens the tree.)
+
+\begin{lemma}[\texttt{cursor\_right\_el\_sorted}; \texttt{cursor\_left\_el\_sorted}]
+The element list for a correct cursor (i.e. of an in-order tree) is in order.
+\end{lemma}
+
+\subsection{Cursor Normalization}
+\label{subsec:norm}
+
+The normalization functions \texttt{next\_node} and \texttt{prev\_node} allow the implementation to treat abstractly equivalent cursors the same.
+That relies on the properties that they don’t change the cursor’s abstraction and that they do generate correct cursors, which are formalized in the following lemmas.
+
+\begin{lemma}[\texttt{next\_node\_correct}]
+\texttt{next\_cursor} preserves cursor correctness.
+\end{lemma}
+
+\begin{lemma}[\texttt{prev\_node\_correct}]
+\texttt{prev\_cursor} preserves cursor correctness.
+\end{lemma}
+
+\begin{lemma}[\texttt{cursor\_elements\_next}; \texttt{cursor\_elements\_prev}]
+If $c$ is a correct cursor, then \texttt{next\_cursor} of $c$, \texttt{prev\_cursor} of $c$, and $c$ all have the same element-list representation.
+\end{lemma}
+
+\subsection{Cursor Creation}
+
+In order to create a cursor, we recursively build up a list of indices into treelists, starting from the root and going down to the leaf that contains (or would contain) the key the cursor is pointing at.
+We need to prove primarily two things: first, that this produces a correct cursor; and second, that the positioning of the cursor (relative to the abstract element-list representation) is correct.
+
+\begin{lemma}[\texttt{make\_cursor\_correct}]
+\texttt{make\_cursor} produces a correct cursor.
+\end{lemma}
+
+\begin{lemma}[\texttt{make\_cursor\_right}; \texttt{make\_cursor\_left}]
+The cursor $c$ returned by \texttt{make\_cursor} for a given key $k$ has the property that the immediate right element has key greater than or equal to $k$, and the immediate left element has key less than $k$.
+\end{lemma}
+
+\subsection{Cursor Movement}
+
+The correctness of the \texttt{move\_to\_next} and \texttt{move\_to\_prev} functions follows directly from the correctness of the normalization functions.
+Given proofs that \texttt{next\_node} and \texttt{prev\_node} both produce correct cursors, then \texttt{move\_to\_next} and \texttt{move\_to\_prev} clearly produce correct cursors as well because the only modification they make is in the last index of the cursor.
+
+\begin{lemma}[\texttt{move\_to\_next\_correct}; \texttt{move\_to\_prev\_correct}]
+Both functions, given a correct cursor, produce a correct cursor.
+\end{lemma}
+
+We stated in \textbf{Section~\ref{subsec:norm}} that both \texttt{next\_node} and \texttt{prev\_node} preserve the element-list, so that they position the cursor before or after an element without changing its abstract representation.
+Then it is intuitive to prove that the element-list of \texttt{move\_to\_next} of a cursor $c$ is the original cursor’s element-list, but with the first element of the “right” list moved to the “left” list.
+Similarly, \texttt{move\_to\_prev} is the original element-list, but with the first element of the “left” list moved to the “right” list.
+If there is no right (or respectively left) element, this implies that the cursor is at the very end (or very beginning) of the tree, and we prove that \texttt{next\_node} and \texttt{prev\_node} return \texttt{None} in that case, so that \texttt{move\_to\_next} and \texttt{move\_to\_prev} just return the original cursor.
+
+\begin{lemma}[\texttt{move\_to\_next\_el}]
+\label{lemma:move_to_next_el}
+If $c$ has the element lists $(l,(k,v)::r)$, then next of $c$ has element lists $((k,v)::l,r)$.
+\end{lemma}
+
+\begin{lemma}[\texttt{move\_to\_prev\_el}]
+\label{lemma:move_to_prev_el}
+If $c$ has the element lists $((k,v)::l,r)$, then prev of $c$ has element lists $(l,(k,v)::r)$.
+\end{lemma}
+
+\begin{lemma}[\texttt{move\_to\_next\_none}; \texttt{move\_to\_prev\_none}]
+\label{lemma:move_to_none}
+If $c$ has no element to its right (or to its left), then next of $c$ (or prev of $c$) is just $c$.
+\end{lemma}
+
+\subsection{Correctness of Get}
+\label{subsec:getcorrect}
+
+The correctness property of \texttt{get} relative to the element-list abstraction is simple, and can be easily applied to the properties of the abstract specification.
+Specifically, if $(k,v)$ is the element at the front of the right element-list of a (correct) cursor, then \texttt{get\_tree} should return \texttt{val k v}.
+
+\begin{lemma}[\texttt{get\_correct}]
+\texttt{get} returns the key-value pair at the front of the right list in the element-list abstraction.
+\end{lemma}
+
+If the cursor is correct---including that it points into a leaf node---then \texttt{next\_node} must be correct as well (from \textbf{Section~\ref{subsec:norm}}), so the treelist that \texttt{next\_node} returns will necessarily be a leaf node (this is part of the definition of correctness), and the tree that \texttt{get\_tree} returns must necessarily be a \texttt{val} tree.
+Furthermore, given that \texttt{next\_node} positions the cursor to point a tree without changing the element-list for the cursor, we know both that \texttt{get\_tree} will return that tree and that it must be at the front of the right element-list of \texttt{next\_node} of the cursor (and therefore of the original cursor as well).
+Therefore, the tree returned by \texttt{get\_tree} and the element at the beginning of the right element-list must be the same.
+
+\subsection{Correctness of Insert}
+
+By comparison to the functional implementation, the element-list abstraction of insertion is quite simple.
+Let $l$ be the left element-list, and let $(k,v)::r$ be the right element-list.
+Insertion has three distinct cases to consider when inserting a key-value pair $(x,v')$:
+\begin{enumerate}
+    \item If the cursor is not correctly positioned for $x$, then it should return the cursor unchanged.
+    \item If the cursor is correctly positioned, but $x$ is not in the tree, then it should result in element-lists of $(x,v')::l$ and $(k,v)::r$.
+    \item If the cursor is correctly positioned, and $x$ is in the tree (i.e. $x = k$), then the resulting element-lists should be $l$ and $(k,v')::r$.
+\end{enumerate}
+
+We can prove a helpful lemma with regards to the first point:
+
+\begin{lemma}[\texttt{bad\_cursor\_insert\_same}]
+If the key $x$ is not in the range of keys represented by this cursor, then \texttt{insert} returns back the original cursor.
+\end{lemma}
+
+Now consider insertion of a key $x$ into a cursor that does represent $x$.
+We first want to prove that insert is structurally correct, i.e.
+a correct cursor, once inserted into, gives back a correct cursor.
+
+\begin{lemma}[\texttt{insert\_correct}]
+\label{lemma:insert_correct}
+If $c$ is a correct cursor into a correct B+\,tree, then for any key $x$ and value $v$, inserting $(x,v)$ into $c$ produces a correct cursor into a correct B+\,tree.
+\end{lemma}
+
+Then, we want to prove what it does to the element-list abstraction, i.e. that it either replaces the next key-value pair or inserts this key-value pair at the cursor's position.
+
+\begin{lemma}[\texttt{insert\_eq\_elements}]
+If the next key is equal to the key $x$ being inserted, then let $l$ be the left element list and $(x,v')::r$ be the right element list.
+Then inserting $(x,v)$ into the cursor produces a cursor with element list equal to: $(l,(x,v)::r)$.
+\end{lemma}
+
+\begin{lemma}[\texttt{insert\_neq\_elements}]
+If the next key is not equal, then let $l$ be the left element list and $r$ be the right element list. Then insertion produces a cursor with element list equal to $((x,v)::l,r)$.
+\end{lemma}
+
+\subsection{Complete Correctness Theorem}
+
+We need one final theorem that pulls all of the other results together and states that the functional model satisfies all of the properties in the abstract specification.
+In Coq this is completed by filling out an instance of the \texttt{CURSOR\_TABLE} module type shown in \textbf{Figure~\ref{fig:params}} and \textbf{Figure~\ref{fig:axioms}}.
+We create a \texttt{Module} of type \texttt{CURSOR\_TABLE}:
+
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
 Module BT_Table <: CURSOR_TABLE.
@@
 End BT_Table.
-\end{minted}
-\end{singlespace}
-\end{figure}
-
-In more intuitive terms, we can summarize this as the following final theorem:
-
-\begin{theorem}
-The functional B+\,tree program described in section \ref{sec:func} satisfies all of the axioms laid out in the abstract specification described in section \ref{sec:spec}.
+\end{cminted}
+\caption{The Coq structure for creating an instance of a \texttt{Module Type}.}
+\end{figure}
+\end{singlespace}
+
+This \texttt{Module} contains all of the implementations of the \texttt{Parameters} and proofs of the \texttt{Axioms} in \texttt{CURSOR\_TABLE}.
+
+Many of the theorems in the module follow directly from properties proved above about the B+\,tree operations and the element-list abstraction.
+For example, consider \texttt{next\_prev}, which was stated in Coq terms in \textbf{Figure~\ref{fig:axioms}}.
+In English, the theorem is the following:
+
+\begin{theorem}[\texttt{next\_prev}]
+For any cursor $c$ that points into some table $t$, if $c$ is not in the last position of $t$ then \texttt{prev} of \texttt{next} of $c$ is equivalent to $c$.
 \end{theorem}
 
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
+Theorem next_prev: forall c t,
+   abs_rel t c -> ~ (c = last_cursor t) -> eq_cursor c (prev (next c)).
+\end{cminted}
+\caption{The formal Coq statement of \texttt{next\_prev}.}
+\end{figure}
+\end{singlespace}
+
+Note that ``equivalent'' uses the predicate \texttt{eq\_cursor}. In the definitions section of the \texttt{Module}, we define \texttt{eq\_cursor} as having equal element-list representations.
+
+\begin{singlespace}
+\begin{figure}[ht]
+\centering
+\begin{cminted}{coq}
+Definition eq_cursor c1 c2 : Prop :=
+  cursor_elements V c1 = cursor_elements V c2.
+\end{cminted}
+\caption{The formal Coq statement of \texttt{next\_prev}.}
+\end{figure}
+\end{singlespace}
+
+Therefore, the proof of \texttt{next\_prev} follows immediately from lemmas about \texttt{next} and \texttt{prev}: \textbf{Lemma~\ref{lemma:move_to_next_el} (\texttt{move\_to\_next\_el})}, \textbf{Lemma~\ref{lemma:move_to_prev_el} (\texttt{move\_to\_next\_el})}, and \textbf{Lemma~\ref{lemma:move_to_none} (\texttt{move\_to\_next\_none}; \texttt{move\_to\_prev\_none})}.
+If the cursor is not the last cursor, it must have an element to its right in its element-list; then \texttt{move\_to\_next} moves that element to the left list, and \texttt{move\_to\_prev} moves it back to the right.
+
+Many other proofs in the module would follow from the lemmas listed above.
+For example, \texttt{insert\_correct} (\textbf{Lemma~\ref{lemma:insert_correct}}) in the implementation is a more general version of the \texttt{insert\_correct} required in the abstract specification.
+Others that don't follow as exactly are still intuitive to understand how they would fit together.
+However, the module cannot be completed without first completing the various auxiliary lemmas listed above and others in the Coq files.
+Although this thesis makes significant progress toward connecting the implementation and the specification in a complete correctness proof, the verification is still only partial.
+
 \clearpage
@@
 
-As described above from the DeepSpec group, the goal of a deep specification has four parts: it should be rich, two-sided, formal, and live.
+As described in \textbf{Section~\ref{subsec:deepspec}} from the DeepSpec group, the goal of a deep specification has four parts: it should be rich, two-sided, formal, and live.
 All of these properties together ensure a high-quality specification that is both correct and useable.
-Since that is the goal of this thesis, it is worth reflecting on how well this particular specification project achieved those properties.
+Since that is a goal of this thesis, it is worth reflecting on how well this particular specification project achieved those properties.
 
@@
 It helped to simplify the properties that needed to be proved into a concise description that would hopefully be clear and useful to any future client of the specification.
-Furthermore, the very brief example in section \ref{subsec:ex} demonstrates some of the applicability of the abstract specification developed.
+Furthermore, the very brief example in \textbf{Section~\ref{subsec:ex}} demonstrates some of the applicability of the abstract specification developed.
 
@@
 The process and approach to specification in this thesis is based on the deep specification framework being studied by the DeepSpec group, and this thesis highlights the successes and complications encountered with that approach.
-Further work with the specification approach used in this thesis could extend and improve specification techniques for complex imperative software and one-sided specifications.
+Further work with the specification approach used in this thesis could extend and improve specification techniques for complex imperative software and specifications without an explicit client side.
 
 \clearpage
+
+\begin{appendices}
+
+\section{Proof Status Summary}
+\label{app:proofs}
+
+\setcounter{theorem}{0}
+\begin{singlespace}
+
+\tocless\subsection{Element-List Abstraction}
+
+\begin{lemma}[\texttt{cursor\_elements'\_sides\_equiv}]
+The pair that \texttt{cursor\_elements} returns is equal to the pair of \texttt{cursor\_left} and \texttt{cursor\_right}.
+\end{lemma}
+
+\begin{cminted}{coq}
+Theorem cursor_elements'_sides_equiv : forall cn cf l r,
+  cursor_elements' cn cf l r =
+  (cursor_left cn cf l, cursor_right cn cf r).
+\end{cminted}
+
+\begin{proof}
+This has been formally proved in Coq.
+\end{proof}
+
+\begin{lemma}[\texttt{left\_rec\_interior}; \texttt{right\_rec\_interior}]
+For both the left and right functions, the ``base'' they are passed ends up at the beginning of the list, i.e.
+if the base is \texttt{b} and the end result is \texttt{l}, then there is some \texttt{l'} such that \texttt{l = b++l'}.
+\end{lemma}
+
+\begin{cminted}{coq}
+Theorem left_rec_interior : forall cn cf l,
+  exists l', cursor_left cn cf l = (l++l').
+
+Theorem right_rec_interior : forall cn cf r,
+  exists r', cursor_right cn cf r = (r++r').
+\end{cminted}
+
+\begin{proof}
+This has been formally proved in Coq.
+\end{proof}
+
+\begin{lemma}[\texttt{cursor\_right\_elements1}; \texttt{cursor\_left\_elements1}]
+For both the left and right functions, if we split a cursor into two parts, then the result of computing the elements of the whole cursor is equal to the result of first computing the elements of the first part, then using that as the base for computing the elements of the second part.
+\end{lemma}
+
+\begin{cminted}{coq}
+Theorem cursor_right_elements1 : forall cn1 cf1 cn2 cf2 b,
+  length cn1 = length cf1 -> length cn2 = length cf2 ->
+  cursor_right (cn1++cn2) (cf1++cf2) b =
+  cursor_right cn2 cf2 (cursor_right cn1 cf1 b).
+
+Theorem cursor_left_elements1 : forall cn1 cf1 cn2 cf2 b,
+  length cn1 = length cf1 -> length cn2 = length cf2 ->
+  cursor_left (cn1++cn2) (cf1++cf2) b =
+  cursor_left cn2 cf2 (cursor_left cn1 cf1 b).
+\end{cminted}
+
+\begin{proof}
+This has been formally proved in Coq.
+\end{proof}
+
+\begin{lemma}[\texttt{cursor\_right\_el\_sorted}; \texttt{cursor\_left\_el\_sorted}]
+The element list for a correct cursor (i.e. of an in-order tree) is in order.
+\end{lemma}
+
+\begin{cminted}{coq}
+Theorem cursor_right_el_sorted : forall cn cf,
+  cursor_correct (cn,cf) -> sorted_less (cursor_right cn cf []).
+
+Theorem cursor_left_el_sorted : forall cn cf,
+  cursor_correct (cn,cf) -> sorted_more (cursor_left cn cf []).
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq.
+
+Informally, this follows from the order implied by \texttt{cursor\_correct}.
+We can introduce an invariant that as it processes the cursor, the list created so far is sorted and within certain bounds, and then the remainder of the cursor (whose element list, by the previous lemmas, will be placed after this first list) will be sorted and outside of those bounds.
+\end{proof}
+
+\tocless\subsection{Cursor Normalization}
+
+\begin{lemma}[\texttt{next\_node\_correct}]
+\texttt{next\_cursor} preserves cursor correctness.
+\end{lemma}
+
+\begin{cminted}{coq}
+Theorem next_node_correct : forall cn cf c k,
+  cursor_correct (cn,cf) -> next_node cn cf = Some (c,k) ->
+  cursor_correct c.
+\end{cminted}
+
+\begin{proof}
+This has been formally proved in Coq.
+\end{proof}
+
+\begin{lemma}[\texttt{prev\_node\_correct}]
+\texttt{prev\_cursor} preserves cursor correctness.
+\end{lemma}
+
+\begin{cminted}{coq}
+Theorem prev_node_correct : forall cn cf c k,
+  cursor_correct (cn,cf) -> prev_node cn cf = Some (c,k) ->
+  cursor_correct c.
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq. However, the structure of the proof would closely follow the proof of \texttt{next\_node\_correct}.
+\end{proof}
+
+\begin{lemma}[\texttt{cursor\_elements\_next}; \texttt{cursor\_elements\_prev}]
+If $c$ is a correct cursor, then \texttt{next\_cursor} of $c$, \texttt{prev\_cursor} of $c$, and $c$ all have the same element-list representation.
+\end{lemma}
+
+\begin{cminted}{coq}
+Theorem cursor_elements_next : forall cn cf c k,
+  cursor_correct_struct (cn,cf) ->
+  next_node cn cf = Some (c,k) ->
+  cursor_elements (cn,cf) = cursor_elements c.
+
+Lemma cursor_elements_prev : forall cn cf c k,
+  cursor_correct_struct (cn,cf) ->
+  prev_node cn cf = Some (c,k) ->
+  cursor_elements (cn,cf) = cursor_elements c.
+\end{cminted}
+
+\begin{proof}
+\texttt{cursor\_elements\_next} has been partially proved; \texttt{cursor\_elements\_prev} has not been proved.
+\end{proof}
+
+\tocless\subsection{Cursor Creation}
+
+\begin{lemma}[\texttt{make\_cursor\_correct}]
+\texttt{make\_cursor} produces a correct cursor.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma make_cursor_correct : forall x f,
+  cursor_correct_struct (make_cursor x f).
+\end{cminted}
+
+\begin{proof}
+This has been formally proved in Coq.
+\end{proof}
+
+\begin{lemma}[\texttt{make\_cursor\_right}; \texttt{make\_cursor\_left}]
+The cursor $c$ returned by \texttt{make\_cursor} for a given key $k$ has the property that the immediate right element has key greater than or equal to $k$, and the immediate left element has key less than $k$.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma make_cursor_right : forall cn cf x f k v l,
+  make_cursor x f = (cn,cf) ->
+  cursor_right cn cf [] = (k,v)::l ->
+  lt_key k x <> true.
+
+Lemma make_cursor_left : forall cn cf x f k v l,
+  make_cursor x f = (cn,cf) ->
+  cursor_left cn cf [] = (k,v)::l ->
+  lt_key k x = true.
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq.
+\end{proof}
+
+\tocless\subsection{Cursor Movement}
+
+\begin{lemma}[\texttt{move\_to\_next\_correct}; \texttt{move\_to\_prev\_correct}]
+Both functions, given a correct cursor, produce a correct cursor.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma move_to_next_correct : forall c,
+  cursor_correct c ->
+  cursor_correct (move_to_next c).
+
+Lemma move_to_prev_correct : forall c,
+  cursor_correct c ->
+  cursor_correct (move_to_prev c).
+\end{cminted}
+
+\begin{proof}
+\texttt{move\_to\_next\_correct} has been formally proved in Coq.
+
+\texttt{move\_to\_prev\_correct} has not been formally proved.
+\end{proof}
+
+\begin{lemma}[\texttt{move\_to\_next\_el}]
+If $c$ has the element lists $(l,(k,v)::r)$, then next of $c$ has element lists $((k,v)::l,r)$.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma move_to_next_el : forall c l r k v,
+  cursor_correct c ->
+  cursor_elements c = (l,(k,v)::r) ->
+  cursor_elements (move_to_next c) = ((k,v)::l,r).
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq.
+\end{proof}
+
+\begin{lemma}[\texttt{move\_to\_prev\_el}]
+If $c$ has the element lists $((k,v)::l,r)$, then prev of $c$ has element lists $(l,(k,v)::r)$.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma move_to_prev_el : forall c l r k v,
+  cursor_correct c ->
+  cursor_elements c = ((k,v)::l,r) ->
+  cursor_elements (move_to_next c) = (l,(k,v)::r).
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq.
+\end{proof}
+
+\begin{lemma}[\texttt{move\_to\_next\_none}; \texttt{move\_to\_prev\_none}]
+If $c$ has no element to its right (or to its left), then next of $c$ (or prev of $c$) is just $c$.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma move_to_next_none : forall cn cf,
+  cursor_correct (cn,cf) ->
+  (cursor_right cn cf [] = [] <-> move_to_next (cn,cf) = (cn,cf)).
+
+Lemma move_to_prev_none : forall cn cf,
+  cursor_correct (cn,cf) ->
+  (cursor_left cn cf [] = [] <-> move_to_prev (cn,cf) = (cn,cf)).
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq.
+\end{proof}
+
+\tocless\subsection{Correctness of Get}
+
+\begin{lemma}[\texttt{get\_correct}]
+\texttt{get} returns the key-value pair at the front of the right list in the element-list abstraction.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma get_correct : forall cn cf k v l,
+  cursor_correct (cn,cf) ->
+  cursor_right cn cf [] = (k,v)::l ->
+  get_tree (cn,cf) = Some (val k v).
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq. The informal proof laid out in \textbf{Section~\ref{subsec:getcorrect}} describes how it would be proved from various auxiliary lemmas.
+\end{proof}
+
+\tocless\subsection{Correctness of Insert}
+
+\begin{lemma}[\texttt{bad\_cursor\_insert\_same}]
+If the key $x$ is not in the range of keys represented by this cursor, then \texttt{insert} returns back the original cursor.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma bad_cursor_insert_same : forall c k v,
+  key_rel k c <> true -> insert k v c = c.
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq.
+\end{proof}
+
+\begin{lemma}[\texttt{insert\_correct}]
+If $c$ is a correct cursor into a correct B+\,tree, then for any key $x$ and value $v$, inserting $(x,v)$ into $c$ produces a correct cursor into a correct B+\,tree.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma insert_correct : forall c k v,
+  cursor_correct c -> cursor_correct (insert k v c).
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq.
+\end{proof}
+
+\begin{lemma}[\texttt{insert\_eq\_elements}]
+If the next key is equal to the key $x$ being inserted, then let $l$ be the left element list and $(x,v')::r$ be the right element list.
+Then inserting $(x,v)$ into the cursor produces a cursor with element list equal to: $(l,(x,v)::r)$.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma insert_eq_elements : forall c k v' v l r,
+  cursor_elements c = (l,(k,v')::r) ->
+  cursor_elements (insert k v c) = (l,(k,v)::r).
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq.
+\end{proof}
+
+\begin{lemma}[\texttt{insert\_neq\_elements}]
+If the next key is not equal, then let $l$ be the left element list and $r$ be the right element list. Then insertion produces a cursor with element list equal to $((x,v)::l,r)$.
+\end{lemma}
+
+\begin{cminted}{coq}
+Lemma insert_neq_elements : forall c k v l r,
+  get_key c <> Some k -> cursor_elements c = (l,r) ->
+  cursor_elements (insert k v c) = ((k,v)::l,r).
+\end{cminted}
+
+\begin{proof}
+This has not been formally proved in Coq.
+\end{proof}
+
+\end{singlespace}
+
+\clearpage
+
+\section{Coq Code}
+\label{app:code}
+
+\begin{spacing}{0.8}
+
+\tocless\subsection{Implementation (BTrees.v)}
+
+\small\inputminted[breaklines]{coq}{./BTrees.v}
+
+\clearpage
+
+\tocless\subsection{Specification (BTreesModule.v)}
+
+\small\inputminted[breaklines]{coq}{./BTreesModule.v}
+
+\end{spacing}
+
+\end{appendices}
+
+\clearpage
+\pagenumbering{gobble}
+
+\begin{singlespace}
 \bibliographystyle{IEEEtran}
 \bibliography{biblio}
+\end{singlespace}
 

### 2018-05-21 src/btreecomplexity.c
Added KVStore tests to btreecomplexity.c. Updated Makefile. Uncommented out helper function to handle deletion of a btree.

@@ -14,8 +14,11 @@
 #include <stdio.h>
 #include <stdlib.h>
 #include <time.h>
-#include "relation.h"
 #include <assert.h>
+#include <string.h>
+
+#include "relation.h"
+#include "kvstore.h"
 
 /*
  * Simple C Test Suite
@@ -34,8 +37,8 @@ static int compareInt(const void* first, const void* second){
     }
 }
 
-/* Return the time it takes to traverse a relation of numKeys keys*/
-static double timeNext(int numKeys) {
+/* Return the time it takes to traverse a relation of numKeys keys in order. */
+static double timeSequentialGets(int numKeys) {
     Relation_T bTree;
     Cursor_T cursor;
     int i;
@@ -63,8 +66,10 @@ static double timeNext(int numKeys) {
     return (double) (end - begin) / CLOCKS_PER_SEC;  
 }
 
-/* Return the time it takes to traverse a relation of numKeys keys*/
-static double timeRandomGets(int numKeys) {
+/* Return the time it visit the first and last key numkeys number of times. 
+ * The relation is of size numkeys. Alternates between first and last key based
+ * on parity of current iteration. */
+static double timeAlternateGets(int numKeys) {
     Relation_T bTree;
     Cursor_T cursor;
     int i;
@@ -83,14 +88,14 @@ static double timeRandomGets(int numKeys) {
         RL_PutRecord(cursor, i, NULL);
         testKeys[i] = i;
     }
-    
-    
-    UTIL_Shuffle(testKeys, numKeys);
-    
-    
+        
     begin = clock();
     for(i = 0; i < numKeys; i++) {
-        RL_MoveToKey(cursor, testKeys[i]);
+        if(i%2 == 0) {
+            RL_MoveToKey(cursor, testKeys[0]);
+        } else {
+            RL_MoveToKey(cursor, testKeys[numKeys - 1]);
+        }
     }
     end = clock();
     
@@ -103,17 +108,20 @@ static double timeRandomGets(int numKeys) {
 }
 
 
-/* Return the time it takes to traverse a relation of p keys*/
+/* Return the time it takes to traverse a relation of size numkeys in a 
+ * partially random order. A random sequence of the keys are generated. This sequence
+ * is partially sorted in batches of size interval.  */
 static double timePartiallySortedGets(int numKeys, int interval) {
[…]

### 2018-05-21 src/relation_mem.c

@@ -1167,35 +1167,32 @@ static void moveToLast(BtNode* node, Cursor* cursor, int level) {
 }  
 
 static void handleDeleteBtree(BtNode* node, void (* freeRecord)(void *)) {
-    /* int i; */
+    int i;
     
-    /* /\* Part of base case. In leaf node, free all records with  */
-    /*  * freeRecord function, if any. *\/ */
-    /* if (node->isLeaf) { */
-    /*     if (freeRecord == NULL){ */
-    /*         return; */
-    /*     } */
+    /* Part of base case. In leaf node, free all records with 
+     * freeRecord function, if any.  */
+    if (node->isLeaf) {
+        if (freeRecord == NULL){
+             return; 
+         } 
         
-    /*     for (i = 0; i < node->numKeys; i++) { */
-    /*         freeRecord((void *)node->entries[i].ptr.record); */
-    /*     } */
+         for (i = 0; i < node->numKeys; i++) { 
+             freeRecord((void *)node->entries[i].ptr.record); 
+         } 
         
-    /*     return; */
-    /* } */
+         return;
+    }
     
-    /* /\* Recursively delete every child subtree. If the child is a leaf, after  */
-    /*  * recursively deleting every child subtree, free the node. *\/ */
-    /* handleDeleteBtree(node->ptr0, freeRecord); */
-    /* if (node->ptr0->isLeaf) { */
-    /*     free(node->ptr0); */
-    /* } */
-    /* for(i = 0; i < node->numKeys; i++) { */
-    /*     handleDeleteBtree(node->entries[i].ptr.child, freeRecord); */
-    /*     if (node->entries[i].ptr.child->isLeaf) { */
-    /*         free(node->entries[i].ptr.child); */
-    /*     } */
-    /* } */
-  return;
+    /* Recursively delete every child subtree. If the child is a leaf, after
+     * recursively deleting every child subtree, free the node.  */
+    handleDeleteBtree(node->ptr0, freeRecord); 
+    free(node->ptr0);
+    for(i = 0; i < node->numKeys; i++) { 
+        handleDeleteBtree(node->entries[i].ptr.child, freeRecord); 
+        free(node->entries[i].ptr.child); 
+    } 
+  
+    return;
 }
 
 /* Check node size invariants */

### 2018-05-21 papers/adewale/ch-background/chapter-background.tex
Master's Thesis .pdf and .tex files.

\section{Background} \label{sec:Background}

In this section, we discuss concepts related to this project. We first give an overview of main-memory databases to illustrate why a main-memory key-value store is not only feasible but valuable. We discuss B-Trees and B+-Trees, and why they are valuable for databases. Finally we give an overview of the Masstree data structure that our key-value store is based on.

\subsection{Main Memory Databases} 
Main-memory databases are a class of storage systems where records are located entirely in memory as opposed to traditional database systems where records are located on external storage devices and loaded into memory as needed. As one might expect, these traditional systems are generally slower than their Main-memory counterparts, due to file I/O. In the past, these traditional database systems were necessary as commodity hardware did not have memory large enough to hold all the data in a database. However, much cheaper memory prices mean main-memory databases are a feasible storage solution. This is evidenced by the emergence of various main-memory database solutions and implementations, such as: SQL Server Hekaton, H-Store/VoltDB, Hyper, Silo (Masstree), SAP HANA and MemSQL. As file I/O is no longer a bottle neck, main-memory databases have more opportunities for improved performance. Furthermore, they have to make different design choices with regard to indexing, concurrency, caching, and partitioning from traditional databases as file I/O is no longer the main bottle neck. Moreover, because these databases are not file-based, they also have to address the issues of durability and logging. Faerber et al. \cite{main-mem-dbs} give a more in-depth survey of main-memory databases and how different implementations address the challenges and opportunities in this field of database design.

\begin{comment}
    should I mention prices of databases. 
\end{comment}

\subsection{Database Indexing and B+-Trees} \label{sec:indexB+-trees}

A key feature of databases is an indexing layer or component. The indexing layer enables fast retrieval/access of any record in a database. It reduces the running time of accessing any record in a file-based database with $N$ pages from $O(N)$ to $O(log _d N)$ (where d is a constant) in the case of B+-Trees, a tree-based index. Hash-based indices are also possible. Tree-based indices maintain an order between their records. As a result they can support range queries like: \say{Which students have a GPA between 2.5 and 3.5?}. Databases have at least one primary index and can have multiple secondary indices for each set of records. This enables fast operations on different fields of the records. One can view our key-value store as a primary index of a set of records.

\begin{figure}[hbtp]
    \centering
    \includegraphics[scale=0.3]{figures/ExampleBTree.png}
    \caption{Example \textbf{B-Tree} with 19 keys. $d = 2$.}
    \label{fig:exB-Tree}
\end{figure}

B+-Trees are a variant of B-Trees: a balanced search tree data structure. B-Trees were first introduced by Bayer and McCreight \cite{bayer1970organization} as a method for indexing records stored in files on secondary storage. Figure \ref{fig:exB-Tree} shows an example B-Tree. Each node in a B-Tree has an order $d$ that determines its capacity, i.e. how many keys it can hold. Each node, except the root, must have between $d$ and $2d$ keys. The root can have between $0$ and $2d$ keys. The pointer between two keys $a$ and $b$ contain keys whose values are greater than $a$ and smaller than $b$.

\begin{figure}[htbp]
    \centering
    \includegraphics[scale=0.3]{figures/ExampleBplusTree.png}
    \caption{Example \textbf{B+-Tree} with 15 keys. $d = 2$.}
    \label{fig:exB+-Tree}
\end{figure}

B+-Trees vary from their B-Tree counterparts in that their nodes are divided into a \textit{sequence} and \textit{index} set. Figure \ref{fig:exB+-Tree} shows a simple B+-Tree with one internal node (the root) and five leaf nodes holding the keys stored in the tree. The sequence set consists of the leaves of the B-Tree. The index set consists of the internal (non-leaf) nodes. So record keys are found in the leaves, while the internal node contain copies of keys that guide searches to the leaf node containing the desired record. These copies are the result of node splitting. A node split occurs when we try to insert a new key into a full node in order to make room for the new key. After a node split, a copy of the first key in the newly created second node and a pointer to the second node is pushed into the parent node of the first node. This process continues if the parent node is full. Furthermore, traditional B+-Trees have cross-links between leaves. This means the sequence set forms a (doubly) linked list. Restricting keys / records to the leaves leads to simpler deletion and enables fast sequential access of keys. Getting the next key in the sequence of current keys always takes constant time. A survey of B-Trees and their variants, including B+-Trees, was presented by Comer in \cite{comer1979ubiquitous}.

\subsection{Handling Variable-length Keys} \label{sec:VarLCPKeysMasstree}

\subsubsection{Variable-length Keys and Keys with Long Common Prefixes}

B+-Trees are an excellent data structure for record insertions, updates and retrieval. Any record in a B+-Tree containing a large number of records can be accessed with a few node accesses. Indeed the number of node accesses involved in an operation is a good cost model when analyzing the performance of file-based databases and even main-memory databases. However, it pays to consider the cost of string compares in B+-Trees in the presence of variable length keys that could possibly share long common prefixes. This is because aside from node accesses a fair amount of time is spent looking for the appropriate key in an internal node (to guide search) or in a leaf node (to retrieve the desired record), when searching for a record. The time spent comparing strings is more significant in main-memory databases where the cost of node accesses is relatively low compared to file-based databases.

\begin{figure}[htbp]
    \centering
    \includegraphics[scale=0.4]{figures/pagevariablelengthkeys.png}
    \caption{Example B+-Tree with $d = 2$.}
    \label{fig:exNodePageB+-Tree}
\end{figure}

Most B+-Tree implementations handle variable-length keys by adding an extra level of indirection. A page/node containing variable length keys in a file-based B+-Tree will look something like Figure \ref{fig:exNodePageB+-Tree} (from page 220 of \cite{ramakrishnan2000database}). VoltDB (a main memory database) also adds a level of indirection when storing long record fields. Figure \ref{fig:voltDBStructure} (from page 48 of \cite{main-mem-dbs}) shows how records are stored in VoltDB's storage layer. Here, record tuples are stored in fixed-sized blocks. If a tuple's field is greater than 8 bytes, it is stored in a variable-size block; the memory address of the item is then stored in the field's location in the tuple instead. In both of the presented cases,  there is a level of indirection involved in finding desired keys within a node. Aside from this being less cache friendly, there is the added complication of maintaining the free area within a node page. Moreover, it is possible that some space could be wasted due to fragmentation within a data page.

\begin{figure}[hbtp]
    \centering
    \includegraphics[scale=0.4]{figures/voltdbvariablelengthkey.png}
    \caption{VoltDB storage layout.}
    \label{fig:voltDBStructure}
\end{figure}

\begin{comment}
 VoltDB/H-Store (a main memory database)\cite{main-mem-dbs}, for example, has a fixed-size block pool containing table tuples. If a field is larger than 8-bytes, it is stored in a variable-length block instead. The field in the tuple contains the address of the block instead of the 8-byte data. Traditional file based databases use an "indirection vector"\cite{graefe2011modern} or "directory of slots"\cite{ramakrishnan2000database} with elements of fixed size that point to records in a data area in the page. Each page is a node in the B-Tree. Our implementation follows MassTree's approach instead of the standard approach. An overview of MassTree's datastructure is given below.
\end{comment}

\begin{table}[h]
\centering
\label{tab:exampleResourceURLTable}
\begin{tabular}{|l|c|}
\hline
Resource URL                & \multicolumn{1}{l|}{Unique Clients} \\ \hline \hline
/apis/v1/photos/albums      & 753                        \\ \hline
/apis/v1/photos/profile     & 486                        \\ \hline
/apis/v1/info/tutorial      & 1000                       \\ \hline
/apis/v1/info/preferences   & 1000                       \\ \hline
\end{tabular}
\caption{Example of HTTP resource URLs and unique client counts.}
\end{table}

B+-Trees can also run into issues with keys with long prefixes. Consider a web server that periodically processes it server logs and stores in a databases resources requested and the count of how many unique clients requested that resource. An example database table is shown in Table \ref{tab:exampleResourceURLTable}. The corresponding primary B+-Tree index on resource URLs would look like the B+-Tree shown in Figure \ref{fig:urlExampleB+-Tree} (only keys are shown). This simple example is illustrative of how B+-Trees handle long keys that possibly have long common prefixes. Space is wasted saving copies of long keys and performance falls. This is because when searching for any key in the tree, more character comparisons have to be made to determine the lexicographical relationship between two strings that share a long common prefix.

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.3]{figures/urlexamplebtree.png}
    \caption{B+-Tree index of Table \ref{tab:exampleResourceURLTable} (only keys are shown).}
    \label{fig:urlExampleB+-Tree}
\end{figure}

For the simple example in Table \ref{tab:exampleResourceURLTable}, these issues are not a cause for concern. However, one could imagine a much larger table where there are millions of rows where most keys share long common prefixes. These issues would be intolerable. Even if we take the prefix B+-Tree approach (see \cite{comer1979ubiquitous,ramakrishnan2000database}) - where instead of storing copies of keys in internal nodes, the smallest possible delimiter is used - space could still be wasted if there the common prefixes are long. In Figure \ref{fig:urlExamplePrefixB+-Tree} we see that the root of the prefix B+-Tree index of Table \ref{tab:exampleResourceURLTable} contains the delimiter key \say{/apis/v1/p}. If the largest key to the left of the root and the smallest key to the right of the root shared a longer common prefix, then delimiter key must be longer.

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.3]{figures/urlexampleprefixbtree.png}
    \caption{Prefix B+-Tree index of Table \ref{tab:exampleResourceURLTable} (only keys are shown).}
    \label{fig:urlExamplePrefixB+-Tree}
\end{figure}

\subsubsection{The solution: Masstree}

Given the aforementioned challenges, our key-value store is based on the Masstree system / data structure. This is to ensure our key-value store can properly handle variable length keys and keys sharing long common prefixes,.

Masstree \cite{masstree} is an in-memory network key-value storage server that supports range queries. It is also the name of the data structure that the server uses to store key-value pairs. The Masstree data structure is designed to efficiently support arbitrary-length keys, including keys with long common prefixes.  The Masstree system supports \texttt{$get_c(k)$}, \texttt{$put_c(k,v)$}, \texttt{$remove(k)$} and \texttt{$getrange_c(k,n)$} operations. As one might expect, \texttt{get} returns the value associated with a key \texttt{k}. \texttt{put} adds a key-value pair \texttt{(k,v)} into the key-value store or updates the value (\texttt{v}) associated with a key \texttt{k}. \texttt{getrange} returns up to the first \texttt{n} key-value pairs greater than or equal to \texttt{k} in the database. Finally the parameter c is a list of column numbers that allow a client to get or set portions of a key's value.

The authors of \cite{masstree} describe the Masstree data structure as a \say{trie-like concatenation of B+-Trees}. The structure of a Masstree tree is shown in Figure \ref{fig:MasstreeTreeStructure}. The tree has multiple layers containing one or more B+-Tree. The root layer contains at most one B+-Tree, the second layer contains at most $2^{64}$ trees - one for every possible key the first layer can have - and so on. Each B+-Tree and layer is responsible for some 8-byte slice of a key. Masstree's B+-Trees consist of interior nodes and border nodes(see Figure \ref{fig:MasstreenodeDS}).

\begin{figure}[htbp]
    \centering
    \includegraphics[scale=0.50]{figures/masstreenodestructures.png}
    \caption{Masstree Node Data Structures \cite{masstree}}
    \label{fig:MasstreenodeDS}
\end{figure}

Border nodes are similar to traditional B+-Tree leaf nodes. They have cross-links (to aid the \texttt{getrange} operation), keys and key values. However, Masstree's border nodes are unique in that their keys are identified by an 8-byte key slice and a key length. This means that a single Masstree B+-Tree can represent multiple (at most 10) keys with each key slice. At most 9 different keys for every possible prefix of the keyslice (keys with length 0 to length 8) and a 10th key in the event that there is a key suffix or a link to the next layer. Not all key slices can have 10 different keys. A key slice with no null bytes can have only two different keys. A key with a null byte at the end can have three different keys. Moreover, for each key in a border node, there is a value or a link to a B+-Tree in the next layer. Whether a key points to a value or another B+-Tree depends on the value of its key length field. If a key's key length is greater than 8, then the key has a suffix and a value or has a link to another layer instead of a value.

\begin{figure}[hbtp]
    \centering
    \includegraphics[scale=0.50]{figures/masstreedatastructure.png}
    \caption{Masstree Tree Structure \cite{masstree}}
    \label{fig:MasstreeTreeStructure}
\end{figure}

MassTree border nodes also contain keysuffix data structures in the event that a key's suffixes aren't shared by any other keys with the same prefix. That is, if we have the keys \texttt{"AB"} and \texttt{"ABCDEFGH!"} in our key-value store, there is no need to create a second layer to index the last byte of the second key. On the contrary, we store \texttt{"AB"} (2 bytes) and \texttt{"ABCDEFGH!"} (9 bytes) with their appropriate key slices and key lengths in a B+-Tree with a single node (layer 0). Then we set the keysuffix for the latter of the key slices to be \texttt{"!"}. However, if we have another key \texttt{"ABCDEFGH*"}, key slice \texttt{"ABCDEFGH"} must now point to a new layer. In addition, the string whose suffix is $"!"$ is inserted in the new layer with its original value and the new string whose suffix is \texttt{"*"} is inserted in the new layer with its value. Generally, if a key is $8h$ bytes long, its value will be stored at layer $x$ where $x < h$. So, if a key is 8 bytes or less, its value will be referenced by a keyslice in the root layer's B+-Tree. If a key is 40 bytes long, its value will be stored in one of layers 0 to 4. This is because key suffixes can be stored in a bordernode, if no other prefixes of the full key share in the keyslices up to a layer and part of the suffix. The keysuffix data structure means that unnecessary layers are not created when there are many relatively long keys that do not share common prefixes. On the other hand, if there are long keys that share common prefixes, then multiple layers (relatively small layers) would be responsible for a portion of the common prefix. Which keeps retrievals fast as shown in the Evaluation Section \ref{sec:Evaluation}.

\begin{comment}
Operations in MassTree can benefit from cache-locality compared to typical B+-Tree schemes where variable length keys are handled by an extra level of indirection (areas outside the tree to store long keys) (reference???). Operations also benefit by reducing space costs associated with storing multiple keys with common prefixes separately.
\end{comment}

### 2018-05-21 papers/adewale/ch-evaluation/chapter-evaluation.tex
Master's Thesis .pdf and .tex files.

\section{Evaluation} \label{sec:Evaluation}

In this section we evaluate our key-value store and the B+-Tree library. The following experiments were ran on a 13-inch early 2015 Macbook Pro\footnote{Processor: 2.7 GHz Intel Core i5. RAM: 8GB}.

\subsection{B+-Tree Experiments}

In this section we evaluate the following claims that our B+-Tree library: 
\begin{itemize}
    \item provides amortized constant time sequential (put or get) operation.
    \item benefits from performance improvements when a sequence of (put or get) operations is partially sorted by key.
\end{itemize}

To evaluate the performance of B+-Tree get operations, we timed how long it took to move to / get all the keys in a B+-Tree for different workloads. That is, we visited or retrieved all the keys in the B+-Tree sequentially, in partially-sorted order and randomly for different B+-Tree sizes. There are three types of partially sorted workloads: those sorted in batches of 15 key-value pairs (\textit{PS Batches 15}), workloads with 1000 sorted batches each with size N / 1000 (\textit{PS Batches N / 1000}) and, finally, workloads with 100 sorted batches each with size N / 100 (\textit{PS Batches N / 100}).  For each B+-Tree with size $N$, the B+-Tree was constructed by inserting all the keys from $0$ to $N - 1$. The results of this experiment are shown in Figure \ref{fig:B+GetsGraph}.

\begin{figure}[hbtp]
    \centering
    \includegraphics[scale=0.50]{figures/Btreegetsgraph.png}
    \caption{Time to visit or get all the keys in a B+-Tree for different B+-Tree sizes and different workloads.}
    \label{fig:B+GetsGraph}
\end{figure}

Similarly to evaluate the performance of B+-Tree put operations, we timed how long it took to insert a sequence of keys into the B+-Tree with the aforementioned workloads (sequential, partially-sorted and random). Results are shown in Figure \ref{fig:B+PutsGraph} Finally, we use the doubling hypothesis \cite{sedgewick2011algorithms} to empirically measure the order of growth of sequential and random operations . The doubling hypothesis states that if $T(N) \sim aN^blgN$ then $\frac{T(2N)}{T(N)} \sim 2^b$. So by calculating the log ratios of the times taken, we can estimate the order of growth of our B+-Tree under different workloads.

\begin{figure}[htbp]
    \centering
    \includegraphics[scale=0.50]{figures/Btreeinsertsgraph.png}
    \caption{Time to insert all keys from 0 to size - 1 into an empty B+-Tree for different sizes and different workloads.}
    \label{fig:B+PutsGraph}
\end{figure}

As we can see from Figures \ref{fig:B+GetsGraph} and \ref{fig:B+PutsGraph} sequential operations are very fast inx our B+-Tree, while random operations are the slowest. We also see that our B+-Trees have performance gains if a series of put or get operations are partially sorted into sorted batches. The larger the sorted batches are in proportion to the number of keys to be inserted or retrieved, the larger the performance gain. On a different note, we expected that alternately retrieving the first and last keys in the B+-Tree its size number of times would model the worst case performance, as each retrieval costs $2 \times height$ node accesses. Even though this should, in theory, model the B+-Tree's worst case behavior it does not in practice, because of caching. If the same two keys are constantly retrieved, the same same ancestor nodes are traversed each time to get from one key to the other. These nodes will be in higher levels of cache memory thereby keeping the cost of node accesses to retrieve each key very low. This experiment highlights the value of cache-friendly data structures when implementing main-memory databases and also highlights how costly main-memory look-ups can be in comparison to cache look-ups.

\begin{figure}[htbp]
    \centering
    \includegraphics[scale=0.50]{figures/BtreeSequentialNextGraph.png}
    \caption{Time to sequentially visit all the keys in a B+-Tree for different B+-Tree sizes}
    \label{fig:B+GetsSequentialGraph}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[scale=0.50]{figures/BtreeSequentialnsertsGraph.png}
    \caption{Time to sequentially insert all the keys from 0 to size - 1 into an empty B+-Tree}
    \label{fig:B+PutsSequentialGraph}
\end{figure}

Figures \ref{fig:B+GetsSequentialGraph} and \ref{fig:B+PutsSequentialGraph} allow us to better appreciate the speed of sequential operations. Our B+-Tree implementation can sequentially retrieve all keys from a B+-Tree of size 5 million in about 49 milliseconds; this is about 9.8 nanoseconds per next operation. We can sequentially insert 5 million keys into a B+-Tree in about 265 milliseconds; this is about 53 nanoseconds per next operation.

\begin{table}[hbtp]
\centering
\label{fig:logRatioGetAnalysis}
\begin{tabular}{|c|c|c|c|c|}
\hline
\multirow{2}{*}{Number of keys (millions)} & \multicolumn{2}{c|}{Move to Next} & \multicolumn{2}{c|}{Random Get} \\ \cline{2-5} 
                                           & Time (ms)       & log ratio       & Time (ms)      & log ratio      \\ \hline
0.5                                        & 5.03            &                 & 182.48         &                \\ \hline
1                                          & 9.89            & 0.975           & 420.07         & 1.203          \\ \hline
2                                          & 20.16           & 1.027           & 970.84         & 1.209          \\ \hline
4                                          & 42.45           & 1.074           & 2421.73        & 1.319          \\ \hline
\end{tabular}
\caption{Log ratio of running times of sequential and random get operations for different key sizes.}
\end{table}

\begin{table}[htbp]
\centering
\label{fig:logRatioPutAnalysis}
\begin{tabular}{|c|c|c|c|c|}
\hline
\multirow{2}{*}{Number of keys (millions)} & \multicolumn{2}{c|}{Sequential Put} & \multicolumn{2}{c|}{Random Put} \\ \cline{2-5} 
                                           & Time (ms)       & log ratio       & Time (ms)      & log ratio      \\ \hline
0.5                                        & 25.01           &                 & 148.83         &                \\ \hline
1                                          & 59.89           & 1.260           & 350.62         & 1.236          \\ \hline
2                                          & 107.29          & 0.841           & 805.68         & 1.200          \\ \hline
4                                          & 225.94          & 1.074           & 1973.40        & 1.292          \\ \hline
\end{tabular}
\caption{Log ratio of running times of sequential and random put operations for different key sizes.}
\end{table}

Using the doubling hypothesis we see that the time complexity of $N$ sequential put or get operations is effectively linear (see see \ref{fig:logRatioGetAnalysis} and \ref{fig:logRatioPutAnalysis}). The log ratio converges to approximately one. However for random operations, even though the log ratio rounds down to 1, the fractional portion hints at an extra logarithmic term. This is in line with our expectation that $N$ random put or get operations is linearithmic.

\subsection{KVStore Experiments}

In this section we evaluate the following claims that our key-value store library: 
\begin{itemize}
    \item performs well when its keys have long common prefixes
    \item benefits from performance improvements when a sequence of operations is partially sorted by key.
\end{itemize}

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.50]{figures/KVStoregetsgraph.png}
    \caption{Time to get all the keys in a KVStore for different sizes and different workloads. Keys are 30 bytes long and are not forced to have common prefixes. Keys are randomly generated.}
    \label{fig:KVStoreGetsGraph}
\end{figure}

Similarly to the B+-Tree experiments we evaluated how long it takes to retrieve all the keys in a KVStore object for different sizes and workloads, where each KVStore object is built with random keys. The results are in \ref{fig:KVStoreGetsGraph}. We see that, as expected sequential gets are very fast. We also see that partially sorting query keys gives performance gains over random query keys. However, the performance gains form partially sorting are not as large as one might expect. There could be different reasons for this. A possibility is that even if the next key being inserted or retrieved is different from the previous key on only the last few bytes, our current KVStore implementation still traverses the trie of B+-Trees from the first layer to the appropriate layer where the keys diverge. Even though we can skip all the layers where the key slices of the next key are the same as the previous key's key slices.

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.50]{figures/kvstoreprefixgraph.png}
    \caption{Time to get all the keys in a KVStore for different sizes for different common prefix lengths. Keys are 100 bytes long. We compare the behavior of get operations when the first 80 bytes are the same and the remaining 20 bytes are random versus when all 100 bytes are random. Keys are randomly generated.}
    \label{fig:KVStoreCommonPrefixes}
\end{figure}

From our discussion in section \ref{sec:VarLCPKeysMasstree}, the expectation is that a typical B+-Tree implementation and, by extension, a key-value store built from this B+-Tree would suffer performance degradation when most of its keys have long common prefixes. However, our KVStore implementation properly handles this scenario as shown in Figure \ref{fig:KVStoreCommonPrefixes}.In fact our KVStore does slightly better for keys with long common prefixes. This is probably because the first few layers of the KVStore consists of only two nodes each (a B+-Tree leaf node with one entry and a bordernode). Consequently, to get to any key all the nodes containing the common prefix must be visited and so these nodes will be in higher levels of cache memory.

### 2018-05-21 papers/adewale/ch-system/chapter-system.tex
Master's Thesis .pdf and .tex files.

\section{System} \label{sec:Our System}

\subsection{Operations} \label{sec:kvstore operations}
Our overall system is implemented as a C programming language library and currently exposes the following operations:

    \begin{itemize}
        \item \texttt{KVStore\_T KV\_NewKVStore(void);}
        \item \texttt{Bool KV\_Put(KVStore\_T kvStore, KVKey\_T key, const void* value);} %%% update to void in code.
        \item \texttt{const void* KV\_Get(KVStore\_T kvStore, KVKey\_T key);}
    \end{itemize}

\texttt{KV\_NewKVStore} returns a newly created key-value store instance. \texttt{get} retrieves the value associated with the given key in the store. \texttt{put} adds a key-value pair to the key-value store or updates the value associated with an existing key. Our key type \texttt{KVKey\_T} is created from an array of chars with its specified length. The return type of put will be updated to void because put should always succeed, unless perhaps memory allocation fails (which is extremely unlikely).

\subsection{Architecture Overview}

Our system is a client of two internal libraries: a \texttt{B+-tree library} and a \texttt{Bordernode library}. The B+-Tree library enables the creation of B+-Trees and operations on those B+-Trees. The Bordernode library implements a variant of Masstree's \cite{masstree} bordernodes (nodes at the border of two btree layers). See section \ref{sec:B+-tree implementation} for more details. We also implement a \texttt{utility library} for some common definitions and operations used across the different libraries.

\begin{comment}
    how much detail to go here?

    architecture diagram?
\end{comment}

\subsection{B+-Tree Implementation} \label{sec:B+-tree implementation}

Our system's B+-Tree data structure is similar to the traditional B+-Tree. However, we do not use cross links between leaf nodes in our implementation, but rather add a new independent auxiliary data structure called a B+-Tree cursor. The B+-Tree cursor can be thought of as an array of node pointers and corresponding pointers to entries within each node (Figure \ref{fig:exB+-Tree}). Each entry in an internal node consists of a key and the pointer to its right. The first entry has no key associated with it. It is simply a pointer. In leaf nodes entries consist of keys and pointers to the record/value they are associated with. The first element of the cursor points to an entry in the root node. The second element in the cursor points to an entry in the child of the entry pointed to in the root, and so on until an element of the cursor points to an entry in a leaf node. The cursor therefore points to all the entries (and nodes) in the unique path from an entry in the root to an entry in a leaf node.

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.30]{figures/BtreeCursor.png}
    \caption{B+-Tree Cursor Pointing at Key 25.}
    \label{fig:exB+Cursor}
\end{figure}

It is important to know that we were inspired to use this B+Tree with cursor concept from SQLite \cite{SQLite}. SQLite's B-Tree implementation supports both B+-Trees (called B*-Trees in their documentation) and B-Trees. B+-Trees are used as what SQLite calls "Table B-Trees", while B-Trees are used as "Index B-Trees". See \cite{SQLiteFileFormat} for SQLite's database file format documentation. The beauty of this data structure is that it enables \textbf{both} fast sequential gets and puts. More specifically, a sequential get or put operation takes amortized constant time. If these assertions are true, insertion of a set of sorted key-value pairs takes linear time and partially sorting a set of keys involved in a series of get and/or put operations can lead to large performance gains. It would also mean that databases using this B+-Tree scheme can reduce the time it takes for operations that build trees / tables from sorted data e.g. failure recovery, data replication. Moreover, if queries come in batches, they can be partially sorted to reduce processing time.

\begin{comment}
internal nodes lead search can have high fanout and take relatively less space. makes deletion easier.
cross links guarantee constant time.

With very large d, All this means we fetch few files when searching for a key or when

ADD figure of B+-Tree and B-trees and reference it.

Reference insertion and deletion algorithms.

THANK AURELE FOR WORK on this implementation

Draw picture of array and trees.
draw graph of cursor.
Proof of amortized constant time.
refer to both.

2 log n worst case time is what happens in a recursive implementation

examples of building b-tree from sorted or partially-sorted data.

algo for B+-Tree cursor insert.

ubiquitous b-tree - general info about b-trees and variant
modern b-tree techniques = more detailed exploration of b+-trees for database exploration
dbms book - algo for trad b+-trees

figures don't show all details, like other information in nodes / cursor
\end{comment}

\subsubsection {B+-Tree Cursor and Operations}

The B+-Tree Module interface has various methods for operating on B+-Trees, including: \texttt{MoveToFirst}, \texttt{MoveToNext} and \texttt{MoveToPrevious}, However, in this section we will focus on the \texttt{MoveToKey}, \texttt{MoveToNext} and \texttt{PutRecord} operations on B+-Trees with a FANOUT (maximum number of keys per node) of two. The FANOUT can be varied and in our implementation it is set to a default value of 15, similar to the node FANOUT of Masstree \cite{masstree}.

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.30]{figures/CursorAt5.png}
    \caption{B+-Tree Cursor Pointing at Key 5.}
    \label{fig:exB+Cursor5}
\end{figure}

\texttt{MoveToKey} and \texttt{MoveToKey} operations are simple if the desired key is in the same node. We just search for the correct key within the current node. For example, if the cursor is at key 5 in Figure \ref{fig:exB+Cursor5}, a \texttt{MoveToNext} or \texttt{MoveToKey(10)} operation should simply move the cursor to the next entry in the node, entry 10. However, if the desired key resides in a leaf node one or more nodes away from the current leaf node, there is more work involved. In the traditional implementation of B+-Trees, a move-to-next operation simply uses the cross-links between leaf nodes to get to the next node, and consequently the next key. Nevertheless, a random search always starts from the root. In our B+-Tree implementation a random search (or next operation to an entry in a sibling node) starts from the current leaf node and ascends and descends through the tree \textbf{as needed} to get to the desired key.

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.30]{figures/CursorAt10.png}
    \caption{B+-Tree Cursor Pointing at Key 10.}
    \label{fig:exB+Cursor10}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.30]{figures/CursorAt20.png}
    \caption{B+-Tree Cursor Pointing at Key 20.}
    \label{fig:exB+Cursor20}
\end{figure}

More specifically, in the case of a \texttt{MoveToNext} operation, the cursor first moves up to the closest ancestor node that it is not pointing to the last entry within. i.e. an ancestor node for which the cursor can move to the next entry within. Afterwards, the cursor then sets the pointer at that level to the next entry in the node and then descends to the entry's smallest leaf child. For example, consider a \texttt{MoveToNext} operation from key 10 (Figure \ref{fig:exB+Cursor10}), this moving the cursor's last pointer from entry 10 on node A to entry 20 on node C (Figure \ref{fig:exB+Cursor20}). To do this, the cursor first ascends to node B. Then it moves its entry pointer for that level to the next entry in node B. Finally it descends to the current entry in node B's smallest leaf child. If the cursor was originally at the last entry in node B, it would have to keep ascending till it either reaches the root or a node where its pointer for that node/level was not previously pointing to the last entry in the node. For example, if the cursor was at key 35 (node D), a move to the next entry (key 40) entails moving up to node E, moving the entry pointer in that node to the next entry and then descending to the left most leaf child.

For a random search, the cursor has to move up to the closest node that is definitely an ancestor of the desired search key. This is a node where the search key is greater than or equal to its first entry and strictly smaller than its last entry. The only exception to this rule is the root. The root is an ancestor of every node in the tree. After a definite ancestor of a key has been found the cursor can then descend down the tree from this node to the desired leaf entry. For example, let the cursor be at key 20 (Figure \ref{fig:exB+Cursor20}). What happens when \texttt{MoveToKey(50)} is called? To move to key 50, the cursor first ascends to node E. It does not go beyond node E as 50 is in between node E's smallest and greatest key. We then descend to node H then node I which contains key 50 (Figure \ref{fig:exB+Cursor20}). What if a call is then made to \texttt{MoveToKey(73)}? In this scenario, the cursor ascends all the way to the root R, and then descends to node E then node F and then node G. The cursor then points at key 73 (Figure \ref{fig:exB+Cursor73}). The cursor has to ascend all the way to node R (the root) because it is impossible from simply looking at node E's entries to know for sure that node E is definitely an ancestor of key 73. All node E's entries can inform us is that children of its greatest entry are greater than or equal to 60. It is possible that another child of E's parent could be the ancestor of key 73. This would be the case if the first key in node R is less than 73. This is an example of the worst case of random search with the B+-Tree scheme. In the worst case, a \texttt{MoveToKey} call takes time proportional to $2 log N$. $log N$ time to ascend to the root and $log N$ time to descend from the root to the apropriate leaf node. This is in contrast to random search in the traditional B+-Tree scheme which always takes time proportional to $log N$. However it is important to note that the cost to ascend will not be significant as most (if not all) of the nodes in the ascend path are likely to be in cache (more specifically higher levels of cache memory) because they have been recently visited. On the other hand, whenever we carry out a random search for a nearby node, we benefit from not needing to ascend to the root and from most of the nodes we visit being in higher levels of cache memory.

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.30]{figures/CursorAt50.png}
    \caption{B+-Tree Cursor Pointing at Key 50.}
    \label{fig:exB+Cursor50}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.30]{figures/CursorAt73.png}
    \caption{B+-Tree Cursor Pointing at Key 73.}
    \label{fig:exB+Cursor73}
\end{figure}

Sequential (next) operations are fast as well. Most of the time they entail simply moving to the next element in a node, a few times they entail a jump from a node to its right sibling node which often involves fewer node accesses and occasionally, in the worst case, involves $2h$ node accesses, where $h$ is the height of the tree. Inserts can benefit from using this scheme (i.e using a cursor that remembers its current location and the path to it). With sequential inserts, most of the time, the current node has enough space for the next key. Sometimes, we need to split the node (node splits are described briefly in \ref{sec:indexB+-trees}). Most of the time, a single split is enough. Sometimes the split propagates upward to one or more nodes. Occasionally, a node split propagates up multiple levels. As a result, we expect that a sequential put or random put \textbf{near} the cursor's current location will be quick and simply involves the cost of moving to a nearby location and inserting a key; both of these steps (move and put) are often fast if the key to be put into the tree is near the key the cursor is pointing to.

\begin{comment}
   Nevertheless, it is unlikely that, in a series of random searches / move-to-key calls, the desired key is always very far from the cursor's current position. Sure, except due to caching doing this constantly is very fast.
\end{comment}

In conclusion, aside from reducing the number of node accesses needed in move and put operations, the B+-Tree with cursor scheme benefits highly from cache-locality when the cursor does not move too far. It is less likely to access unnecessary nodes in the path from parent to the desired key, rather it is more likely to revisit previously visited nodes. For example, when traversing the tree from keys 5 to 35 (imagine a range query or sequential updates), our B+-Tree with cursor scheme visits only nodes A, B, C and D. Note we visit node B FANOUT (two) times. This implies that if a sequence of operations is partially sorted by key we can experience massive performance gains, subject to how much we sort the sequence of operations. It is important to note that SQLite \cite{SQLite} (which uses this B+-Tree with cursor scheme) does not always benefit if the cursor is at a key close to a search key. Indeed, SQLite's \texttt{sqlite3BtreeNext} function behaves similarly to our \texttt{MoveToNext} function: most of the time the next key would be in the current node and SQLite's cursors only ascends the tree as needed. Nevertheless, SQLite's \texttt{MoveTo} function always starts searching for the desired key from the root, unless the cursor is already at the correct position. This is in contrast to our implementation of \texttt{MoveToKey} which never begins search from the root. Therefore, a partially random sequence of operations would be expected to execute as quickly as a random sequence of operations and experience no performance gains in SQLite.

\subsubsection {Why Sequential Operations are Amortized Constant Time}

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.30]{figures/amortizedtimenextexample.png}
    \caption{Example B+-Tree with 18 keys and height 3 (cursor not shown). When sequentially visiting all keys in the B+-Tree, there are 17 total next operations: 9 blue next operations, 6 yellow next operations and 2 red next operations.}
    \label{fig:amortizedanalysisheight3}
\end{figure}

In this section, we will investigate the claim made in \ref{sec:indexB+-trees} that a sequential get or put operation takes amortized constant time. Figure \ref{fig:amortizedanalysisheight3} gives an intuition for why this might be the case. Traversing through the tree in Figure \ref{fig:amortizedanalysisheight3} from key 5 to key 83 takes 17 next operations. Most (about half) of these  operations are the low cost blue next operations.These blue next operations cost little as they simply involve moving the cursor to the next element in the node and involve zero extra node accesses. The other (roughly) half of next operations require a jump from one node to another. Most of these next operations require the cursor moving up only one level and then down to reach the sibling node containing the next entry. These are the yellow operations which require only two node accesses. Few of these next operations require the cursor moving up to the root and then down to the next entry in the sibling node. These are the red next operations, the most costly next operations.

\begin{figure}[p]
    \centering
    \includegraphics[scale=0.18]{figures/amortizedtimenextexamplehugerotated.png}
    \caption{Example B+-Tree with 54 keys and height 4 (cursor not shown). When sequentially visiting all keys in the B+-Tree, there are 53 total next operations: 27 blue next operations, 18 yellow next operations, 6 purple operations and 2 red next operations.}
    \label{fig:amortizedanalysisheight4}
\end{figure}

For our tree with 18 keys, in order of costliness there are:

    \begin{itemize}
        \item 2 red operations
        \item 6 yellow operations
        \item 9 blue operations
    \end{itemize}

If our tree was a level deeper, as in Figure \ref{fig:amortizedanalysisheight4}, in order of costliness there are:

    \begin{itemize}
        \item 2 red operations
        \item 6 purple operations
        \item 18 yellow operations
        \item 27 blue operations
    \end{itemize}

We see that the most costly jumps occur $2 * 3^0$ times, the next most costly operations occur $2 * 3^1$, the next most costly operations occur $2 * 3^2$ times and so on. The least costly operations, however, occur $3^h$ times where $h$ = the height (number of levels) of the tree. This implies that the most costly operations which are proportional to $log N$ where $N$ is the number of entries in the tree occur very infrequently. If each next operation was logarithmic on average, we will expect that most of the next operations are proportional to $log N$. Not only is this not the case, but we see that the more costly an operation, the (exponentially) less likely it is to occur. Put another way, a fraction of the time a next operation requires that we jump to the next node.  When jumping from a node to the next, a majority of the time moving up one level and then back down is sufficient. A fraction of the time when we jump between two nodes, we need to move up two levels and then back down. An even smaller fraction of the time we need to move up three levels and then back down, and so on and so forth.

With this intuition we can find an upper bound of the average cost of a next operation for a tree of any height:

\begin{equation} \label{eq:TnextaveExp1}
    T_{ave} < c + \frac{2}{F} + \frac{4}{F^2} + \frac{6}{F^3} + ... = c + \frac{2 \times 1}{F} + \frac{4 \times 2}{F^2} + \frac{6 \times 3}{F^3} + ...
\end{equation}

\begin{equation} \label{eq:TnextaveSeries}
    T_{ave} < c + \sum_{i=0}^{\infty} \frac{2i}{F^i} 
\end{equation}

where $F$ is the FANOUT - maximum number of keys - in each node in the B+-Tree.

In Equation \ref{eq:TnextaveExp1}, we see that a next operation always takes some constant plus 2 node accesses $\frac{1}{F}$ of the time, plus 4 node accesses $\frac{1}{F^2}$ of the time and so on. The frequency of costly operations decreases exponentially. This is an upper bound because in reality the probability of any jump is $\frac{1}{F}$, low cost and high cost jumps included. So this means that in reality, the most frequent lowest cost jumps occur less than $\frac{1}{F}$ of the time, the next lowest cost jumps occur less than $\frac{1}{F^2}$ of the time and so on.

The series in equation \ref{eq:TnextaveSeries} is the product of an \textbf{arithmetic} and a \textbf{geometric} series. It is an \textbf{arithmetico-geometric} series. We know that the sum from an arithmetico-geometric series is given by \cite{riley2011foundationmaths}:

\begin{equation} \label{eq:arith-geometricSeries}
    S = \sum_{n=0}^{\infty} (a + n d)r^n = \frac{a}{1-r} + \frac{rd}{(1-r)^2} 
\end{equation}

where $r < 1$.

Equation \ref{eq:TnextaveSeries} can be simplified with \ref{eq:arith-geometricSeries} to yield:

\begin{equation} \label{eq:TnextaveResult}
    T_{ave} < c + \frac{2F}{(F-1)^2} 
\end{equation}

From Equation \ref{eq:TnextaveResult}, we see that the average cost of a next operation for a tree of any height with fanout F is upper-bounded by some small cost plus $\frac{2F}{(F-1)^2}$ node accesses. This shows two things. Firstly, a next operation does not take logarithmic time (is not proportional to the height of the tree), but rather takes amortized constant time. Secondly, the cost of a next operation is inversely proportional to the FANOUT $F$ of the nodes in the tree.

Given the previous analysis, we can also find an upper bound for the cost of a sequential put operation. i.e. a put operation where the key=value pair to be updated or inserted is to be placed beside the current key. We know that the cost of a sequential put operation is the cost to move to the appropriate node (same node or sibling node) plus the cost of inserting the entry. We already know that the cost to move to the next key (next appropriate node) is less than the upper bound in Equation \ref{eq:TnextaveResult}. We know that inserting a key-value pair in a node takes some constant and might involve node splits. How often are node splits? Imagine a sequence of sequential inserts, where each key is not present in the tree. Most of the time a node will have space for an insert and not require any node splits (see \ref{sec:indexB+-trees}). A fraction of the time (proportional to $\frac{1}{F}$) there is a node split. A fraction of the time when there is a node split it propagates up to the next level, a fraction of the time when a split propagates by one extra level it propagates another level, and so on.

We can find the average cost to insert a new key into a node by simplifying the following inequality:

\begin{equation} \label{eq:TinsertAtLocation}
    T_{ins} < c + c_{split}(\frac{1}{F} + \frac{2}{F^2} + \frac{3}{F^3} + ...)
\end{equation}

That is the cost of inserting a new key-value pair is less than some constant plus the cost of a split that occurs at most some time proportional to $\frac{1}{F}$ plus the cost of two splits that occurs at most some time proportional to $\frac{1}{F^2}$ and so on. The constant$c_{split}$ is to account for the fact that splits occur in proportion to the fractions involved in the geometric series.

\begin{equation} \label{eq:TinsertAtLocationResult}
    T_{ins} < c + c_{split}\sum_{i=0}^{\infty} \frac{i}{F^i} = c + \frac{c_{split}F}{(F-1)^2} 
\end{equation}

Using Equation \ref{eq:arith-geometricSeries} we solve the arithmetico-geometric series in Equation \ref{eq:TinsertAtLocation} to find an upper bound on the cost of insertions.

Putting equation \ref{eq:TnextaveResult} and equation \ref{eq:TinsertAtLocationResult} together an upper bound on the cost of a sequential insert and, consequently, a sequential put is:

\begin{equation} \label{eq:Tseqput}
    T_{seqput} < T_{ins} + T_{ave} = c_1 + \frac{2F}{(F-1)^2} + c_2 + \frac{c_{split}F}{(F-1)^2} 
\end{equation}

As we can see from equation \ref{eq:Tseqput}, the time it takes for a sequential put is lest than some constant inversely proportional to the fanout $F$. Therefore sequential puts also take amortized constant time.

\begin{comment}
If we have d to the (h) keys, in the best case next takes constant time, in the worst case it take time proportional to h, as we have to go to the root and back down. however, if we have n random gets or get next operations this happens once every n times.
\end{comment}

\subsection{Key-value Store Implementation} \label{kvstore implementation}

In this section we discuss our key-value store implementation. Our key-value store is implemented in the C programming language as a library (called KVStore). It currently supports the operations outlined in section \ref{sec:kvstore operations}. It is a client of the B+-Tree module \ref{sec:B+-tree implementation} and the BorderNode module. Our B+-Tree module as is can only handle fixed length 8 byte keys. Therefore it must be extended to handle variable length keys. As mentioned in section \ref{sec:VarLCPKeysMasstree}, databases and storage systems that use B+-Trees tend to handle variable length keys by adding a level of indirection between a node's slot array and the record / record key in the node's data area; we also saw how B+-Trees are not optimized for handling keys with long common prefixes. We discussed how Masstree effectively handles variable-length keys and keys with long common prefixes. As a result, our KVStore Module is based on Masstree. It is also a trie-like concatenation of B+-Trees.

\begin{comment}
    For example, in SQLite (where records are stored in files by default), B+-Trees are used as a Table index, while ... index, and each node is a page on disk, variable length keys are .... Other external storage based databases use a similar / different approach (check databasemanagement book and other modern b-tree techniques). With main memory databases there is more flexibility to try different approaches. For example, this system handles variable length keys by this. That system handles variable length keys by that.

    Mainmemdbs page 49/52 h-store uses variable sized blocks. modern b-tree techniques indirection vector page 236. 220 database management textbook 220
\end{comment}

\subsubsection{Key-value Store Module and BorderNode Module}

Our key-value store module is based on Masstree. It can be viewed as a serial re-implementation of Masstree. However, there are a few differences. As mentioned previously, it is a client of the B+-Tree module \ref{sec:B+-tree implementation}. We also made the decision to implement border nodes separately from the B+-Tree; our B+-Tree's leaf nodes are not border nodes. Our B+-Tree module's leaf nodes (listing \ref{lst:btreenode}) have fixed length key slices and associated value, but do not have key length fields or key suffixes as in Masstree (see Figure \ref{fig:MasstreenodeDS}). Furthermore, our border nodes contain a key slice, an array of values and a key suffix (listing \ref{lst:bordernode}). Each border node is associated with a key slice and does not have an array of key lengths; key length is implicit in the position of the value in the value array. If there is a key suffix in a bordernode, that means that the 10th key does not have a link to the next layer, and vice versa.

\begin{lstlisting}[language=C, caption={B+-Tree node type definition}, captionpos=b, label = {lst:btreenode}]

typedef unsigned long Key;

struct Entry {
    Key key;
    Child_or_Record ptr;
};

struct BtNode {
    Bool isLeaf;
    Bool FirstLeaf;
    Bool LastLeaf;
    int numKeys;
    BtNode* ptr0;
    Entry entries[FANOUT];
};
\end{lstlisting}

This design of separating the concept of a bordernode from that of a B+-Tree into two separate modules for use by a key-value store allows for greater modularity. The B+-Tree and BorderNode modules can each easily be replaced with different implementations. It is important to also note that this implementation of Bordernodes use less memory when there are keys with multiple common prefixes. For example, if there are 10 keys with the keyslice "00000000", where each character is a NULL character. In Masstree there would be 10 key slices, 10 values (references) and 10 keylengths. In our scheme, there would be one key slice in the B+-Tree, one reference to a Bordernode pointed to by the entry in the leaf node, 10 values and 4 overhead fields in the Bordernode. In this scenario we use about half the space Masstree uses. However, when every key slice is unique, MassTree has a key slice, a value and a key length associated with each key, while we have a key slice, a border node reference and space for 10 values associated with each key. Note that we can optimize our Bordernode implementation's space usage by removing the keyslice field in the Bordernode and by using a more efficient bordernode implementation when there are only a few keys associated with the bordernode's keyslice. This field is redundant as it will be already present in a B+-Tree entry pointing to the bordernode.

\begin{lstlisting}[language=C, caption={Border node type definition}, captionpos=b, label={lst:bordernode}]
struct BorderNode {
    unsigned long keySlice;
    void* val[MAX_BN_SIZE];
    size_t keySuffixLength;
    char* keySuffix;
};
\end{lstlisting}

Another difference between our key-value store implementation and Masstree is that in order to leverage the benefits of our B+-Tree with cursor implementations we permanently associate each B+-Tree with a cursor. This is to ensure higher performance when the key involved in the next operation is lexicographically close to the key involved in the previous operation.

### 2018-05-21 verif/btrees/relation_mem.c
more on splitnode proof

@@ -541,13 +541,14 @@ static void splitnode(BtNode* node, Entry* entry, Bool isLeaf) {
 
       /* if the new entry came before an entry in the first node, 
        * then we need to update those entries in the first node.*/
+      node->numKeys = MIDDLE;
+      
       if(tgtIdx < MIDDLE) {
         for(i = tgtIdx; i < MIDDLE; i++) {
 	  node->entries[i].key = allEntries[i].key;
 	  node->entries[i].ptr.record = allEntries[i].ptr.record;
         }
       }
-      node->numKeys = MIDDLE;
 
       /*Copy entries to second node.*/
       for (i = MIDDLE; i < FANOUT + 1; i++) {

### 2018-05-21 papers/adewale/abstract.tex

In this paper, we discuss the implementation of a serial main-memory key-value store based on Masstree\cite{masstree}. Similar to Masstree, the key-value store is implemented as a trie-like tree of B+-Trees, where each B+-Tree is responsible for a fixed-length slice of a variable-length key. However, one of the major differences between our key-value store and Masstree is that our B+-tree implementation (a component of the key-value store) takes linear time to insert a set of sorted records. This is compared to a traditional B+-tree implementation that would take linearithmic time. Moreover, partially sorting a sequence of operation leads to substantial performance gains. This is made possible using a data structure for navigating B+-trees called a B+-tree cursor. As our next operation is amortized constant time, our B+-tree does not need to maintain cross links between leaf nodes. We also briefly show that this same data structure can be extended to the trie of B+-Trees to ensure amortized linear time for bulk insertion of key-value pairs in the key-value store. We were inspired with this idea of B+-Tree cursors from the SQLite \cite{SQLite} B-tree source code.

\begin{comment}
    Confirm sqlite idea of cursors, is it constant time indeed?
    Show how cursor in b-tree and masstree impl. is indeed constant amortized time. is it really a cursor of cursors. 
    Update abstract in the end.
\end{comment}

### 2018-05-21 papers/adewale/acknowledgements.tex

I would like to thank Thomas Schaffner, Chris Galea, and Donna Gabai for helping and/ or supporting me in my Thesis Presentation. I would like to thank Aur\`ele Barri\`ere for co-implementing the B+-Tree library. He was very instrumental in the B+-Tree library's final implementation, which leverages the B+-Tree with cursor concept. I would also like to thank Professor Andrew Appel for advising me throughout this thesis project and Professor Margaret Martonosi for being my thesis reader. I would like to also thank my parents, Segun and Victoria Adewale, for their support through out my life and education. I doubt I would have made it this far without them. Finally, I am ultimately grateful to God for bringing me this far and for the opportunities that lie ahead. Soli Deo Gloria!

### 2018-05-21 papers/adewale/ch-conclusion/chapter-conclusion.tex

\section{Conclusion}
In this Thesis Paper, we have described at a high-level the implementation of our main-memory serial key-value store. We gave a quick overview of main-memory databases, B+-Trees and Masstree in section \ref{sec:Background}. We discuss out key-value store implementation, B+-Tree with cursor implementation and why B+-Tree sequential operations take amortized constant time in section \ref{sec:Our System}. We also learn more about SQLite's \cite{SQLite} B+-Tree with cursor scheme and how it is used in our system. We evaluated the behaviour of our B+-Tree library and KVStore library in section \ref{sec:Evaluation}. Here we saw that the B+-Tree with cursor idea does indeed yield amortized constant time sequential\textbf{get} and \textbf{put} operations (traditional B+-Tree's can only guarantee constant time \textbf{get} operations). Moreover, we have shown that partially sorting a sequence of inputs leads to performance gains for our B+-Tree and KVStore. We have also shown that Masstree's trie of B+-Tree scheme can properly handle long common prefixes. Finally, we briefly presented a way to extend the B+-Tree cursor to a KVStore Cursor to further speed up sequential and partially sorted queries in our system.

### 2018-05-21 papers/adewale/ch-future-work/chapter-future-work.tex

\section{Future Work}

Our key-value store implementation is far from finished. There are a myriad of ways in which this key-value store can and should be improved upon. We will at least need to implement deletion for both the KVStore module and the B+-Tree module, among some other operations. It would also be good to add durability and support for concurrency. We would also like to extend the B+-Tree cursor concept to create and implement a KVStore Cursor (see Figure \ref{fig:KVStoreCursor1}).

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.50]{figures/kvcursor1.png}
    \caption{KVStore Cursor at node 4.}
    \label{fig:KVStoreCursor1}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.50]{figures/kvcursor2.png}
    \caption{KVStore Cursor at node 6.}
    \label{fig:KVStoreCursor2}
\end{figure}

The KVStore cursor is analogous to the B+-Tree cursor; ancestor nodes in the B+-Tree correspond to ancestor B+-Trees / layers in the KVStore. However, with a KVStore cursor, we would need to also store the key slice that each cursor is located at. This could drastically improve the cost of sequential inserts. This means that when the key of the next operation and the KVStore cursor's current key share a long common prefix, many layers in the KVStore data structure (the trie) will be skipped at the cost of only comparing the search key with the characters associated with each level of the KVStore cursor. We expect this cost to be significantly lower than traversing multiple layers of the overall tree data structure (which also involves navigating through each layer) to get to the desired key. However, it is important to note that if the search keys are totally random, there might be little or not benefit to this scheme. In fact, there would be the added overhead of maintaining KVStore cursor information.

\begin{comment}
Interesting how in the past sequential i/o was crucial and now this cursor of cursors idea grants great gains when one operates on tree data structures sequentially. 
\end{comment}

### 2018-05-21 papers/adewale/ch-intro/chapter-intro.tex

\section{Introduction} \label{sec:introduction}

In a world where software runs on computers with faster CPUs and large memory, there is a clear need for faster databases. Main-memory databases (where records reside primarily in memory) are a feasible solution. These databases are much faster than their disk-based counterparts, as they are not slowed down by a need for frequent I/O operations. Moreover, given advancements in semiconductor technology, main-memory databases have become a feasible storage solution; memory is much cheaper and much larger. In this paper we discuss an implementation of a serial main Memory key-value store based on Masstree \cite{masstree}. Our implementation uses SQLite's \cite{SQLite} concept of a B+-Tree cursor which enables put and get operations to run in \textbf{amortized constant time} when the cursor is near the desired location. We show that a set of sequential put or get operations to this modified B+-Tree run in linear time and that partially sorting a sequence of operations by key increases performance. We also show that this performance improvement under partially sorted workloads extends to out key-value store (trie of B+-Trees). We also briefly present the concept of a key-value store cursor to further improve the performance of out key-value store under sorted and partially sorted workloads.

\begin{comment}
    Experiments and proof that B+-Tree has amortized linear behavior
    Experiments / analysis showing that overall kv store can have amortized linear time.

\end{comment}

### 2018-08-26 memmgr/malloc.c
strengthen malloc_compatible in malloc_token

@@ -3,10 +3,8 @@
 #include <assert.h>
 #include <sys/mman.h> 
 
-/* DISCLAIMER: don't use this with unverified clients: malloc doesn't zero
-the allocated blocks so clients can trash the free lists. 
-A verified client can also trash the size field, owing to transparency
-of the definition of malloc_token... 
+/* ALERT: malloc doesn't zero the free-list pointer in allocated blocks,
+   so it is easy for an un-verified client to trash the free lists. 
 */
 
 /* restricted spec for our purposes

### 2018-08-27 papers/luke/report.tex
include the report

\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{bussproofs}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{tikz}
\usepackage[title]{appendix}

\makeatletter
\newcommand{\@chapapp}{\relax}%
\makeatother

\begin{document}
%
\title{Project Report on DeepSpecDB}

\author{Yixuan Chen\inst{1}}
\institute{University of Michigan}
\maketitle

\begin{abstract}
  Recent years have witnessed a rapid development of main-memory database
  systems thanks to the growingly affordable memory. DeepSpecDB is another
  main-memory database management system implemented in C with deep
  specification and end-to-end verification guaranteeing the correctness of the
  system.
\end{abstract}

\section{Introduction}

As the unit price of memory decreases over time, efforts have been put into
migrating the traditional disk-based applications into main memory. A classical
category of such applications are databases. For example, H-store~\cite{hstore},
MongoDB, Redis and Memcached are well-known database systems featuring
main-memory index and storage.

Meanwhile, database systems are usually barely specified and vulnerable to
subtle bugs. We believe that verified programs against deep specification is the
remedy for the database. Deep specification will ensure that the behavior will
be captured by the specification and proven correctly by machine-checked proof.
With the help of the Verified Software Toolchain~\cite{VST}, we are now able to
verify C programs against the C11 operational semantics. The DeepSpecDB project
aims to develop a main-memory database system with deep specification
guaranteeing the correctness of the system.

One of the previous development of main-memory database is
MassTree~\cite{masstree}. MassTree implements the \textit{trie over B+ tree} 
style key-value store for keys of variable length strings. The data-structure is
specially tailored for the cache of modern day processors. The original MassTree
is designed for single-key transactions, but later development of
Silo~\cite{Silo} and SiloR~\cite{SiloR} suggest its capability of indexing for
relational databases.

The index engine for databases is basically a mapping from index to rows.
Additionally, to support iterations and optimize bulk operations we need a way
to traverse through the index. For this purpose, a data structure called
\textit{Cursor} is introduced. A cursor is a thread-local object that points to
position inside some indexing data structure and supports local movements and
operations. MassTree supports cursors with minor modification to the data
structure, and is chosen to be the indexing data structure of variable length
strings in DeepSpecDB project.

Before my work, Aur\`ele Barri\`ere~\cite{aurele} has already verified the
correspondence between functional and C implementation of B+ tree with cursors.
Brian McSwiggen~\cite{brian} has developed a specification of B+ tree with
cursors. Oluwatosin Victor Adewale~\cite{oluwatosin} has developed the C
implementation for MassTree, however, the cursor for trie is not implemented.

In this report, I present my work on verification of trie with cursors. In
Section~\ref{modver}, I will present the methods I use in verification and the
reason behind them. In Section~\ref{spec}, I will discuss about the
specification we derived so far for any data structures with cursors. In
Section~\ref{trie}, I will introduce the instantiation of the specification with
trie as internal data structures. In Section~\ref{walkthrough}, I will walk
through a client routine for the indexing data structure, and present what the
end-to-end verification guarantee.

\section{Terminology}

In this report, I will use the term \emph{kv-map} (stands for key-value map) for
data structures that supports \texttt{put} and \texttt{get} operations similar
to a finite map.

% In this report, I will use the term \emph{table} for data structures that
% supports \texttt{put} and \texttt{get} operations the same way as a finite map
% (this is not a proper term in larger context of the database, and it's a
% historical issue should be resolved soon).

\section{Modular Verification}\label{modver}

Modularity is the way to achieve scalability. In this section, I will discuss
about the efforts in modularizing the verification, and present an overview
of the verification structure of the project.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{diagram.png}
  \caption{Verification Structure}\label{fig:struct}
\end{figure}

\subsection{Functional Specification of C Programs}

When verifying a C program against specification, there are generally two
approaches. One is to prove the specification directly within separation
logic. The other is to prove that the C program refines a functional program,
and the functional program is proved later to satisfy the specification. We
favor the latter method.

The functional specification modularizes the verification, and therefore
is beneficial in several aspects. When verifying the C program, we would never
want to deal with application-domain theories accompanied with low level
details in separation logic. By elevating the C program into functional
programs, we are decoupling the refinement proof from the property proofs, and
both proofs are easier and more maintainable. Meanwhile, it turns out that
deriving the perfect specification the first time is almost impossible: The
specification evolves as we are verifying the program. Without a functional
program in between, we will have to rewrite the entire proof. However, the
functional program usually captures all observable behaviors, and therefore the
refinement proof usually remains unchanged. Similarly, while the C program might
be modified for bug fixes, only the lower part of proof needs review, while the
application domain reasoning stays valid.

Concretely in this project, both the B+ trees and tries are implemented in C as
well as in Gallina, where the refinement relation between the two programs is
established using Verified Software Toolchain. After establishing the refinement
relation, we focus mainly on the functional programs for later verification.

\subsection{Modularizing the Components}

Using the functional programs is an effort to modularize the verification
vertically, we would also like to modularize horizontally. In this project,
indexing of database systems are separated from the other parts. Therefore it
naturally forms a verification boundary: Any indexing data structure will
implement a uniform specification, the only part exposed to the rest of the
project. As shown in figure~\ref{fig:struct}, everything surrounded in the bold
square is opaque to the rest of project, while the only interface is the
specification.

Within the index engine, MassTree is naturally divided into two modules: A
module that implements B+ tree with cursors and a module that relies on previous
module to implement trie with cursors. In the verification of trie, we went a
further step where we parameterized over the dependency: for any module that
implements the specification of table of integer-typed key with
cursors, we can derive a module that implements the same specification of
string-typed key with cursors using the trie data structure. Such
parameterization modularized the verification in the way that verification of
trie is now independent from any detail in implementation or verification of B+
trees. Also as a side effect, this approach \textbf{tested} the
\textit{two-sidedness} of the specification: As a consumer of the specification,
a successful verification implied that the specification was sufficient in
capturing the behaviors. Meanwhile as a provider, it made sure that only
reasonable properties was specified.

In Coq, this modularization is achieved through \texttt{Module} and
\texttt{Module Type}. The specification is expressed as a \texttt{Module Type}
where interface functions are expressed as parameters and properties are
expressed as axioms. A \texttt{Module} can implement a \texttt{Module Type} by
providing all the parameters and justifying all the axioms. The dependency of
modules are expressed as functors.

\section{Specification for Table with Cursors}\label{spec}

In this section I will introduce the specification for indexing data
structures. The specification can be generally divided into two parts: the part
where it guaranteed the correctness of the table itself and the part where it
provided essential utilities for clients' reasoning.

\subsection{Interface}

Spiritually, a table maintains the mapping from keys to values and a cursor
points into the table at a specific position and supports movement operations.
Concretely, such a data structure should provide the operations in
figure~\ref{fig:interface}.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  Definition key := KeyType.t.
  Parameter table: Type -> Type.
  Parameter cursor: Type -> Type.

  Section Types.
    Context {value: Type}.
    Parameter empty: table value -> Prop.

    Parameter make_cursor: key -> table value -> cursor value.
    Parameter first_cursor: table value -> cursor value.
    Parameter last_cursor: table value -> cursor value.
    Parameter next_cursor: cursor value -> table value -> cursor value.
    Parameter prev_cursor: cursor value -> table value -> cursor value.

    Parameter get: cursor value -> table value -> option (key * value).

    Parameter put: key -> value -> cursor value -> table value -> (* input *)
                   cursor value -> table value -> (* output *)
                   Prop.
    [...]
  End Types.
End ABSTRACT_TABLE.
\end{minted}
  \caption{Interface for Table with Cursors}\label{fig:interface}
\end{figure}

This is a typical interface of tables which supports operations for creation of
table (\texttt{empty}), creation of cursor (\texttt{make\_cursor},
\texttt{first\_cursor}, \texttt{last\_cursor}), movement of cursor
(\texttt{next\_cursor}, \texttt{prev\_cursor}), fetching data at
cursor (\texttt{get}) and updating the table (\texttt{put}). The types of
functions indicates clearly that all other functions except for \texttt{empty}
and \texttt{put} are immutable with respect to the table. The two table mutating
functions are defined as relations rather than functions, this is a compromise
to support cursors. I will discuss the reason and consequence of relational
specification in detail later in this section.

Apart from these functions, we also need some auxiliary predicates, as shown in
figure~\ref{fig:aux}. \texttt{abs\_rel} asserts that a cursor and a table are
correctly associated with each other (the abbreviation \texttt{abs} should
stands for \emph{abstraction}, but it's not a right word here, rather a
historical issue). \texttt{key\_rel} asserts that a cursor is associated with
a key in a table. One can imagine this predicate as \textit{the cursor is placed
at the position such that all the existing keys to the decreasing side are less
than this key, and all the existing keys to the opposite are no less than this
key}. This makes sense when one visualizes a table as a list of pairs of key and
value. This definition is closely related to the semantics of \texttt{get}, We
will see later how this is used in generalized tables. \texttt{eq\_cursor}
asserts that two cursors are equivalent. Coq equality (\texttt{eq}) does not
work well here, because of \texttt{eq} implies structural equality, and
structurally different cursors can be equivalent (especially for those table
with complex structures, such as B+ trees or tries). The final two predicates
\texttt{cursor\_correct} and \texttt{table\_correct} are the normal invariants.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Parameter abs_rel: cursor value -> table value -> Prop.
  Parameter key_rel: key -> cursor value -> table value -> Prop.
  Parameter eq_cursor : cursor value -> cursor value -> table value -> Prop.
  Parameter cursor_correct: cursor value -> Prop.
  Parameter table_correct: table value -> Prop.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Auxiliary Predicates of Table with Cursors}\label{fig:aux}
\end{figure}

\subsection{Semantics of Operations}

The functions and predicates themselves are not particularly interesting without
further specification, which defines the semantics of operations.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom empty_correct: forall t, empty t -> table_correct t.
  Axiom get_empty: forall t c, empty t -> abs_rel c t -> get c t = None.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Specification for \texttt{empty}}\label{fig:empty}
\end{figure}

For an empty table, it should satisfy the correctness invariant and also that we
are never able to get anything from cursor associated with the table.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom make_cursor_key: forall t k,
    table_correct t -> key_rel k (make_cursor k t) t.
  Axiom make_cursor_abs: forall t k,
    table_correct t -> abs_rel (make_cursor k t) t.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Specification for \texttt{make\_cursor}}\label{fig:make}
\end{figure}

For the \texttt{make\_cursor} operation, the cursor it made should always be
associated with the key and the table.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom first_cursor_abs: forall t,
    table_correct t -> abs_rel (first_cursor t) t.
  Axiom last_cursor_abs: forall t,
    table_correct t -> abs_rel (last_cursor t) t.
  Axiom get_last: forall t, get (last_cursor t) t = None.
  Axiom first_cursor_get_empty: forall t,
    table_correct t -> get (first_cursor t) t = None -> empty t.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Specification for \texttt{first\_cursor} and \texttt{last\_cursor}}\label{fig:first_last}
\end{figure}

\texttt{first\_cursor} and \texttt{last\_cursor} are dual operations. Both of
them should yield cursors associated with the table. Additionally, the
\texttt{last\_cursor} should point past the end of the table, meaning that we
can get nothing from it. The \texttt{first\_cursor} should always return a
binding, unless the table is empty.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom next_cursor_abs: forall t c,
    abs_rel c t -> abs_rel (next_cursor c t) t.
  Axiom prev_cursor_abs: forall t c,
    abs_rel c t -> abs_rel (prev_cursor c t) t.
  Axiom next_order: forall t c k1 k2,
    ~ eq_cursor c (last_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (next_cursor c t) t ->
    KeyType.lt k1 k2.
  Axiom prev_order: forall t c k1 k2,
    ~ eq_cursor c (first_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (prev_cursor c t) t ->
    KeyType.lt k2 k1.
  Axiom next_compat: forall t c k1 k2 k3,
    ~ eq_cursor c (last_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (next_cursor c t) t ->
    ~ key_rel k3 c t -> ~ key_rel k3 (next_cursor c t) t ->
    KeyType.lt k1 k3 /\ KeyType.lt k3 k2 ->
    False.
  Axiom prev_compat: forall t c k1 k2 k3,
    ~ eq_cursor c (first_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (prev_cursor c t) t ->
    ~ key_rel k3 c t -> ~ key_rel k3 (prev_cursor c t) t ->
    KeyType.lt k3 k1 /\ KeyType.lt k2 k3 ->
    False.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Specification for \texttt{next\_cursor} and \texttt{prev\_cursor}}\label{fig:next_prev}
\end{figure}

\texttt{next\_cursor} and \texttt{prev\_cursor} is another pair of dual
operations. Both of them should yield cursors associated with the table. Unless
at the end of table, movement should reach a cursor associated with keys of the
right order, they should be inverse to each other and they should move only to
adjacent positions.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom get_key_rel: forall t c k,
    abs_rel c t -> get_key c t = Some k -> key_rel k c t.

  Axiom put_correct: forall t1 t2 c1 c2 k v,
    abs_rel c1 t1 ->
    put k v c1 t1 c2 t2 ->
    table_correct t2.

  Axiom get_put_same: forall t1 t2 c1 c2 c3 k v,
    put k v c1 t1 c2 t2 ->
    abs_rel c1 t1 ->
    abs_rel c3 t2 ->
    key_rel k c3 t2 ->
    get c3 t2 = Some (k, v).
  Axiom get_put_diff: forall t1 t2 c1 c2 c3 c4 k1 k2 v,
    k1 <> k2 ->
    put k1 v c1 t1 c2 t2 ->
    abs_rel c1 t1 ->
    abs_rel c4 t1 ->
    key_rel k2 c4 t1 ->
    abs_rel c3 t2 ->
    key_rel k2 c3 t2 ->
    get c3 t2 = get c4 t1.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Specification for \texttt{get} and \texttt{put}}\label{fig:get_put}
\end{figure}

\texttt{get} operation are different from the \texttt{get} operation of finite
map library in the Coq standard library: \texttt{get} operation here should
fetch the first key to the increasing side of the cursor (In other word, the
largest key that is still associated with this cursor by \texttt{key\_rel}).
However, \texttt{get} and \texttt{put} should satisfy the normal permute and
shadow property as well. Additionally, \texttt{put} operation should keep a
correct table correct after mutation. Notice that, the \texttt{put} takes an
extra cursor as input but is not specified in the axioms. This cursor is there
so that some implementation might optimize the operation based on a nearby
cursor (for example, the B+ tree implemented by Aur\`ele and Oluwatosin), but it
will never affect the semantics of the operation.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom abs_correct: forall t c,
    abs_rel c t -> cursor_correct c /\ table_correct t.

  Axiom key_rel_eq_cursor: forall t c1 c2 k,
    key_rel k c1 t ->
    key_rel k c2 t ->
    abs_rel c1 t ->
    abs_rel c2 t ->
    eq_cursor c1 c2 t.

  Axiom eq_cursor_get: forall t c1 c2,
    abs_rel c1 t ->
    abs_rel c2 t ->
    eq_cursor c1 c2 t ->
    get c1 t = get c2 t.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Additional Specification}\label{fig:aux_spec}
\end{figure}

In addition to the semantics of operations, more specifications are there to
regulate the predicates. \texttt{abs\_rel} should imply that both the cursor and
table correct. Cursors associated with the same key should be equivalent and
equivalent cursors should return the same result.

It should be acknowledged that the current set of specification is not guaranteed
to be complete (and I'm pretty sure not). The completeness is improved along the
way of instantiating it with some concrete implementation and trying to justify
the properties (This is why I used the word \textbf{test} in previous section).
For later development, here is a list of possible issues in my mind.

\begin{enumerate}
\item The semantics of \texttt{key\_rel}, which is the most crucial part of the
  specification, is however not formalized. It might be formalized through the
  \texttt{flatten} function (see later section) and additional auxiliary
  definitions.
\item The two axioms \texttt{next\_compat}, \texttt{prev\_compat} have not yet
  been proved for either B+ tree or trie, and there are quite a few ways to
  express that ``movements will reach only nearby positions''. These two axioms
  might not be the final decision.
\item The additional input of cursor for \texttt{put} operation might be erased
  from the specification, since it does not affect the semantics at all.
\end{enumerate}

\subsection{Augmented Types, Nondeterminism and Relational Specification}

While we want to abstract low level details away from the concrete implementation
in C, our model cannot be fully abstracted. One of the issues is about
abstracting internal pointers.

Consider the following minimal example, where we have a \texttt{struct} has two
internal fields, \texttt{choice\_a} and \texttt{choice\_b}, bothing pointing to
an array of integers of known length. Without cursors, the data structure can be
represented as abstract data of type \texttt{list Z * list Z}.

\begin{figure}[htbp]
  \centering
\begin{minted}{c}
struct choice {
  int *choice_a;
  int *choice_b;
}
\end{minted}
\begin{minted}{coq}
Definition choice_rep (data: list Z * list Z) (p: val) :=
  EX pa: val, pb: val,
  data_at Tsh (Tstruct _choice noattr) (pa, pb) p *
  data_at Tsh (Tarray tint (Zlength (fst data))) (fst data) pa *
  data_at Tsh (Tarray tint (Zlength (snd data))) (snd data) pb. 
\end{minted}
  \caption{Toy Data Structure}\label{fig:choice}
\end{figure}

Now imagine we want to have a ``cursor'' that points at either of the choices.
It's impossible to express in the separation logic because the values have
been abstracted away. The remedy is to include the pointers in the abstract
data, which we call augmented types. For more details, Aurele's paper (cite?)
have a good explanation on this.

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Definition choice_rep (data: (val * list Z) * (val * list Z)) (p: val) :=
  data_at Tsh (Tstruct _choice noattr)
              (fst (fst data), fst (snd data)) p *
  data_at Tsh (Tarray tint (Zlength (snd (fst data))))
              (snd (fst data)) (fst (fst data)) *
  data_at Tsh (Tarray tint (Zlength (snd (snd data))))
              (snd (snd data)) (fst (snd data)).

Definition cursor_rep
  (data: (val * list Z) * (val * list Z)) (cursor: bool) (p: val) :=
  data_at Tsh (tptr tint)
              (if cursor then (fst (fst data)) else (fst (snd data))) p.
\end{minted}
  \caption{Toy Data Structure with Cursor}\label{fig:choice_cursor}
\end{figure}

However, such types will introduce another issue: The pointers are only given at
runtime from the \texttt{malloc} calls. They do not appear in the inputs and can
vary in different runs. The nondeterminism cannot be captured well if the
interface is defined with functions. Therefore for those functions that mutate
the table, we are compelled to define them relationally.

However, some properties implied by a function are no longer guaranteed.
Specifically, determinism and totality. Although we allow the nondeterminism
using the relational specification, we actually want to limit the nondeterminism
within the internal addresses, and no nondeterminism should be observed related
to data. In Aur\`ele's work, he parameterized the augmented type and proved the
theorem that such operations will always yields the same result if the
parameter type is instantiated as \texttt{unit}. In my work, I would claim 
that this is guaranteed as a side effect if a correct \texttt{flatten} function
is provided, which is presented in next subsection. Meanwhile, totality
guarantees that the function represented by the relation will always return.
This is not put into the specification currently, and it might be worthwhile in
later development.

\subsection{Demanding Client, Flattening}

Previous works on verification of B+ tree had assumed that the \texttt{value}
type are integers (\texttt{Int.int}). However, clients usually want a map over
arbitrary \texttt{value} types. Some clients even want to define inductive types
with the table. A minimal example is the finite-degree tree where at each node
it maintains a map from keys to children. However, such definition will never
pass the positiveness check, as a consequence of abstraction.

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Tree (K: UsualOrderedType) (NodeTable: Table K).
  Inductive tree: Type :=
  | node_of: NodeTable.table tree -> tree.
End Tree.
\end{minted}
  \caption{Finite-degree Tree}\label{fig:inductive}
\end{figure}

Even if the client does not require strict positiveness, a separation logic
client might need to quantify over all the bindings and assert that they are
well represented in the memory, as shown in Figure~\ref{fig:rep}. There have
been quite a few new functions declared without specification, and even more are
required to work with table mutations. As the amount of new functions suggest,
additional specification should be added so that separation logic clients live a
better life.

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Client (K: UsualOrderedType) (T: Table K).
  Parameter value: Type.
  Parameter value_rep: value -> val -> mpred.
  Definition t: Type := T.table value.

  Definition rep: (data: t) (p: val): mpred :=
    EX concrete_table: T.table val,
    !! (Table.keys concrete_table = Table.keys data) &&
    Table.table_rep concrete_table p *
    Table.iter_sepcon data concrete_table value_rep.
End Tree.
\end{minted}
  \caption{A Painful Separation Logic Client}\label{fig:rep}
\end{figure}

The remedy for both issues is the \texttt{flatten} function. \texttt{flatten}
turns a table into list of pairs of key and value, where the list is sorted in
the order of keys. Furthermore, The list have been proven to be an
implementation of the same interface as table with cursors. Formally, these
properties are specified in Figure~\ref{fig:flatten}.

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Type FLATTENABLE_TABLE (KeyType: UsualOrderedType)
                              <: ABSTRACT_TABLE KeyType.
  Include ABSTRACT_TABLE KeyType.
  Module Flattened := SortedListTable KeyType.
  Section Spec.
    Context {value: Type}.
    Parameter flatten: table value -> Flattened.table value.
    Axiom flatten_invariant: forall t,
        table_correct t ->
        Flattened.table_correct (flatten t) /\
        forall (k: key) (c1: cursor value) (c2: Flattened.cursor value),
          key_rel k c1 t -> Flattened.key_rel k c2 (flatten t) ->
          abs_rel c1 t -> Flattened.abs_rel c2 (flatten t) ->
          get c1 t = Flattened.get c2 (flatten t).
  End Spec.
End FLATTENABLE_TABLE.
\end{minted}
  \caption{Flattenable Table}\label{fig:flatten}
\end{figure}

Particularly, the sortedness guarantees that the list is always the same
regardless of the internal structure of original structure, given as fact in
Figure~\ref{fig:flatten_fact}. \texttt{Flattened.fput} is the functional version
of \texttt{put} for the list implementation, where we have no nondeterminism. As
an side effect, the nondeterminism of relational specification mentioned before
is dealt with as a side effect of the \texttt{flatten} function. Two more useful
facts about the list implementation have been given for separation logic client.

%TODO improvement

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Type FlattenableTableFacts (KeyType: UsualOrderedType)
                                  (Table: FLATTENABLE_TABLE KeyType).
  Include Table.
  Section Implication.
    Context {value: Type}.
    Theorem put_permute: forall t1 t2 c1 c2 k v,
      table_correct t1 ->
      abs_rel c1 t1 ->
      put k v c1 t1 c2 t2 ->
      flatten t2 = Flattened.fput k v (flatten t1).

    Theorem Forall_put (P: value -> Prop): forall t k v,
      Flattened.table_correct t ->
      P v ->
      Forall P (t) ->
      Forall P (Flattened.fput k v t).
    Theorem iter_sepcon_put (P: value -> mpred): forall t k v,
      Flattened.table_correct t ->
      P v * iter_sepcon P (flatten t) |--
      (option_wrap P (Flattened.get_exact k t)) *
      iter_sepcon P (Flattened.fput k v t).
End FLATTENABLE_TABLE.
\end{minted}
  \caption{Facts about Flattenable Table}\label{fig:flatten_fact}
\end{figure}

\subsection{\texttt{funspec} for C Interface Functions}

The \texttt{funspec} for C programs are listed in appendix. The
\texttt{funspec}s are generally of less interest because they closely follow
the \texttt{Module Type}. However, \texttt{funspec} are not abstracted into
\texttt{Module Type}, because of current limitation of Verifiable C. This might
be done in the future.

\section{Trie with Cursors}\label{trie}

In this section, I will demonstrate the instantiation of trie, informally and
formally.

\subsection{Informal Definition}

Before formal definition, I will introduce the data structure informally first.
MassTree implements a mapping from variable length strings to some value type.
Basically, MassTree organizes all the keys in a trie structure. However,
unlike traditional trie where each node indexes a single character, in MassTree
strings are divided into sequences of $k$-byte slices ($k$ is either 8 or
4 depending on word size). These segments are calculated into integers (if the
suffix are shorter, 0 is padded after), called \texttt{keyslice}. Each node
maps from \texttt{keyslice} to an auxiliary data structure called
\texttt{BorderNode}. \texttt{BorderNode} further distinguishes prefixes of
different length (prefixes of length $1$ to $k$ of this slice will share the
same \texttt{keyslice} value due to the padding). A \texttt{BorderNode} might
also point to either the next layer of trie or a client value whose key has a
suffix after current \texttt{keyslice} and no other key shares a prefix reaching
current \texttt{keyslice}. The mapping at each node is implemented using B+
Tree. For more detail, the MassTree paper~\cite{masstree} is a good reference.

% TODO: insert a graph here for MassTree

\subsection{Formal Model}

Figure~\ref{fig:typedef} defines the type for functional trie. A trie is defined
inductively, containing an address, a node table maps from \texttt{keyslice}s to
addresses and a flattened node table maps from \texttt{keyslice}s to a pair of
address and \texttt{BorderNode}. While \texttt{BorderNode} can optionally points
to next level of trie or final value, as represented by the \texttt{link} type.
One should have noticed that there is a duplication of information: both the
table form of node and list form of node is saved. We need the table form
because of we need the unmodified augmented types for representation in
separation logic. We need the list form to pass the positiveness check. This is
the original motivation to introduce \texttt{flatten}. A \texttt{BorderNode}
contains a list of prefix links, an optional suffix key field and a suffix link
field. When there is a suffix key, the suffix link field should point to a
final value. Otherwise, the suffix link field should optionally point to
another trie. Although the \texttt{BorderNode}'s ``value'' type is always the
same. Different endpoints (prefix links and suffix link) have different
requirement and not all three constructors are valid, future work might refine
this part better.

%TODO: bordernode definition

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module BorderNode.
  [...]
  Definition table: Type := list value * option string * value.
  Inductive cursor: Type :=
    | before_prefix: Z -> cursor
    | before_suffix: cursor
    | after_suffix: cursor.
  [...]
End BorderNode.

Module TrieKey <: UsualOrderedType := [...]. (* a module about strings *)
Module Keyslice <: UsualOrderedType := [...]. (* a module about integers *)

Module Trie (Node: FLATTENABLE_TABLE Keyslice) <: FLATTENABLE_TABLE KeyType.
  [...]
  Inductive trie: Type :=
  | trienode_of: val ->
                 Node.table val ->
                 Node.Flattened.table (val * BorderNode.table link) ->
                 trie
  with link: Type :=
  | value_of: value -> link
  | trie_of: trie -> link
  | nil: link.

  Definition cursor: Type := list
     (trie * Node.cursor val * (BorderNode.table link) * BorderNode.cursor).
  [...]
End Trie.
\end{minted}
  \caption{Trie Definition}\label{fig:typedef}
\end{figure}

Cursors is also defined in Figure~\ref{fig:typedef}, which is a list of
4-tuples. Each element of the list contains a cursor for this node and a cursor
for the \texttt{BorderNode}, and both the trie and the \texttt{BorderNode} is
included as well for convenience. A \texttt{BorderNode} cursor points either
before some prefix link, or at either side of the suffix link. The cursor is
defined in the same order as walking down the trie: the first one corresponds to
the root node and the last one corresponds to the leaf node. Notice that
\emph{the lexicographic order of cursors is equivalent to the lexicographic
  order of the strings}. This is crucial for the correctness of cursor
operations.

\subsection{Interface Functions}

\emph{Because of the length of functions and limitation of space, all functions
  in discussion are listed in appendix.}

\subsubsection{\texttt{empty}}

The \texttt{empty} predicate is defined straightforward. The root node of the trie
should contain only empty node table. There exists another possible definition
where we introduce a \texttt{empty} constructor for \texttt{trie} type, however.
Which one is better is still under question.

\subsubsection{\texttt{make\_cursor}}

is defined by walking down the trie recursively and collecting the routes it
chose. Notice that if it reached the place where a suffix existed in the trie, a
lexicographic comparison is required to determine whether the cursor is placed
before it or after it. Notice that, \texttt{make\_cursor} does not always
guarantee that it will actually point to a pair of key and value. Indeed,
locating the right pair to fetch is a complicated job.

% TODO: BorderNode.next_cursor

\subsubsection{\texttt{strict\_first\_cursor} and \texttt{normalize\_cursor}}

Before presenting the rest operations, these two auxiliary functions will be
introduced first. \texttt{strict\_first\_cursor} takes a trie and yields the
first cursor if there exists one (unless it's an empty table, there always
exists one). \texttt{normalize\_cursor} takes a cursor, which could potentially
pointing at any place in trie (\textit{e.g.}, an internal node of trie, an
endpoint leads to no pair, \texttt{after\_suffix} of a bordernode, etc.), and
try to locate the pair of key and value where \emph{the key is the least one in
  the increasing direction to the cursor}, if such a pair exists.

It's worth mentioning that the \texttt{normalize\_cursor} is a recursive
function, which calls itself recursively before any other branching is made.
This is important because in imperative C program, we can actually start from
the leaf cursor and climb up the spine. This direction will be faster when
the movement is limited to nearby area (which should be the case for
\texttt{normalize\_cursor}), and benefit from the cache locality. The style
appears in many other functions.

The \texttt{normalize\_cursor} and \texttt{strict\_first\_cursor}, however, is
incorrect without a stronger assumption. Specifically, these functions rely on
the fact that there is no dead end in the trie (a subtrie which contains no pair
of keys and values, will be formalized in \texttt{table\_correct} later). If
not, these functions can get trapped into dead end and therefore cause a false
negative result.

\subsubsection{\texttt{get}}

With the help of \texttt{normalize\_cursor}, \texttt{get} is straightforward.
First we traverse down the cursor to collect all the \texttt{keyslice}s and
reconstruct the key, as well as fetch the value pointing to at the end of
normalized cursor. Notice that although the \texttt{get} functions requires a
traverse to obtain the original key, to obtain the value requires only access
the leaf node. This is implemented in imperative program.

\subsubsection{\texttt{next\_cursor}}

\texttt{next\_cursor} is quite similar to \texttt{normalize\_cursor}, except
that it moves one step further. Therefore \texttt{next\_cursor} is defined such
that normalize the cursor first, and forcefully move forward the cursor one more
step.
\subsubsection{\texttt{first\_cursor}}

\texttt{first\_cursor} is a wrapper of \texttt{strict\_first\_cursor}. If the
auxiliary function failed to find a cursor, then \texttt{first\_cursor} will
return an empty cursor, which is regarded as \texttt{last\_cursor} for any
tables.

\subsubsection{\texttt{put} and \texttt{create\_pair}}

To implement the \texttt{put} function, another auxiliary function
\texttt{create\_pair} is required to handle a special case. Generally,
\texttt{put} function is similar to the \texttt{make\_cursor} function, both of
which walk down the trie with a key, deciding the route based on current
\texttt{keyslice}. However, the special case triggers when there is a suffix in
the \texttt{BorderNode}, and is different from the suffix of input key.
\texttt{create\_pair} takes two key and value pairs and return a \texttt{trie}
that contains them, which is linked to the original \texttt{BorderNode}.

\subsubsection{\texttt{table\_correct}}

For a table to be correct, the following properties should hold:

\begin{enumerate}
\item both the table form and list form should be correct
\item they should have the same content
\item the children \texttt{BorderNode}s should be correct recursively:
  \begin{enumerate}
  \item the prefix links should have the same length of \texttt{keyslice}'s
    length,
  \item prefix links should points only to values or nothing,
  \item suffix link should only points to nothing, a trie if there is no suffix
    key or a value if there is a suffix key,
  \item any sub trie should be correct recursively and non-empty,
  \item there should be at least a value or a subtrie pointed to.
  \end{enumerate}
\end{enumerate}

The final rule and ``non-empty'' sub trie property will guarantee that no dead
end exists in the trie. This property is naturally guaranteed with the current
implementation of \texttt{put}, \texttt{create\_pair} and \texttt{empty}.
However, I expected it to be problematic when we finally want to implement
\texttt{delete}, which cannot deal with it trivially.

Meanwhile, most of the correctness of \texttt{BorderNode} might be a result of
improperly designed types. It might be possible to improve this tedious
predicate by modify the type of \texttt{BorderNode}.

\subsubsection{\texttt{cursor\_correct}}

For a cursor to be correct, the following properties should hold:

\begin{enumerate}
\item if this is not the root cursor, then the ancestors should points to the
  current node,
\item the current table is correct,
\item both the cursors are correct,
\item by following the node's cursor, we shall obtain the \texttt{BorderNode}
  included in current level of cursor.
\end{enumerate}

While the correctness of cursor included the correctness of table for now, it's
actually not necessary. This just simplifies the proof and can be removed in
future to make the definition more clearly.

\subsubsection{Other Auxiliary Predicates}

The other auxiliary predicates are defined straightforward. \texttt{abs\_rel} is
defined as the conjunction of table and cursor correctness, as well as the top
most level of cursor points to the root of the trie. The rest levels are
then guaranteed by the cursor's correctness. Both \texttt{key\_rel} are defined
based on \texttt{normalized\_cursor}. A cursor is related to a key if the cursor
can be normalized to the same cursor as \texttt{make\_cursor k t} normalize to.
Two cursors are equivalent if they can be normalized to the same cursor.

\subsection{Current Status and Remaining Problems}

While most of the functions have been instantiated, some are still left as holes.
The \texttt{last\_cursor} and \texttt{prev\_cursor}, dual operation to
\texttt{first\_cursor} and \texttt{prev\_cursor}, should be easy to implement
with reference to the existing ones.

The specification is only partially finished, however. The properties about
\texttt{get} and \texttt{put} are mostly finished (while some holes remained
about theories in strings, and trivially held correctness proof), but little has
been done on the cursor movement operations. These operations shall expect more
difficulty than \texttt{get} and \texttt{put}.

The current MassTree implementation does not support zero-length key. To support
it, a \emph{zero-length prefix} entry should be added in \texttt{BorderNode}.
However, such entry will not be used in any other case than the zero-length key
and it means a significant waste of space.

To get the value from a cursor is cheaper than getting a key from it in trie,
because the former one needs only to access the leaf node while the latter one
needs to traverse through the cursor. This is not represented by the functional
model but only in the imperative program. In practice, the keys are usually
obtained to implement exact \texttt{get} (like the one in traditional finite
map, rather than \texttt{get} in current context). It's widely used in kv-map
but we imagine it will be less used by the client of index engine. Even though,
it might be worhwhile to provide a \texttt{get\_exact} primitive to avoid
overhead.

\section{Example Walk-through}\label{walkthrough}

In this section, I will demonstrate a typical client usage of the library, and
discuss what is guaranteed by the verification.

\subsection{Naming for C Programs}

Before go into the C program, I will briefly introduce the naming for the C
programs. The data structure that implements the mapping is named
\texttt{Index}, and the naming follows the intuition. The B+ tree naming are
prefixed with \texttt{I} (stands for integer), while the trie naming are
prefixed with \texttt{S} (stands for string).

\subsection{A Toy Client}

Imagine a client who puts a key and value and then immediately gets the value
associated with the same key, like the toy program in Figure~\ref{fig:toy_client}

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
SValue foobar(SIndex index, SKey k, SValue v) {
  SCursor c = Sfirst_cursor(index);
  Sput(k, v, c, index);
  Sfree_cursor(c);
  c = Smake_cursor(index, k);
  SValue ret;
  Bool success = Sget(c, index, &ret);
  Sfree_cursor(c);
  return ret;
}
\end{minted}
  \caption{A Toy Client}\label{fig:toy_client}
\end{figure}

Now suppose we run VST analysis over this function, as shown in the
Listing~\ref{lst:annotation}. We have already know that the the table is
correctly represented in the memory, as well as the key. The first call to
\texttt{Sfirst\_cursor} creates the cursor for the \texttt{put} operation. The
call to \texttt{put} will provides a new cursor and new table, and a predicate
guarantees that they are the result of the functional \texttt{put} operation.
After that, several calls to make more cursors and tries get the value we
have just put into the table. At the end of VST analysis, we know that the now
the new table is correctly represented in the memory, the return value should be
the value that the functional \texttt{get} returns, and there is no memory leak
or any undefined behavior during the execution thanks to the soundness of
Verifiable C provides.

Furthermore, according to the specification in the \texttt{Module Type}, we know
that the cursor made from the key should be associated to the cursor, which
means \texttt{key\_rel k (make\_cursor k t') t'}. Additionally, if \texttt{t'} is
the result of some \texttt{put} operation, the input key to it is \texttt{k},
and the input cursor to \texttt{get} is associated with the same key, then it
will definitely get the value \texttt{v}. Therefore, we can prove that the this
function \texttt{foobar} will always return the same value as the input.

\begin{listing}[htbp]
\centering
\{\texttt{table\_rep t index * key\_rep key k}\} \\
\texttt{SCursor c = Sfirst\_cursor(index);} \\
\{\texttt{table\_rep t index * key\_rep key k * cursor\_rep (first\_cursor t)
c}\} \\
\texttt{Sput(k, v, c, index);} \\
\{$\exists t'\exists c'$, \texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep
t' index * key\_rep key k * cursor\_rep c' c}\} \\
\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep t' index * key\_rep key
k * cursor\_rep c' c}\} \\
\texttt{Sfree\_cursor(c);} \\
\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep t' index * key\_rep key
k}\} \\
\texttt{c = Smake\_cursor(index, k);} \\
\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep t' index * key\_rep key
k * cursor\_rep (make\_cursor k t') c}\} \\
\texttt{SValue ret;} \\
\texttt{Bool success = Sget(c, index, \&ret);} \\
\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep t' index * key\_rep key
k * cursor\_rep (make\_cursor k t') c *} \\
\texttt{[ret] = match get\_value (make\_cursor k t') t' with Some v' => v' | None =>
nullval end *} \\
\texttt{success = if (get\_value (make\_cursor k t') t') then 1 else 0}\} \\
\texttt{Sfree\_cursor(c);} \\
\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep t' index * key\_rep key
k *} \\
\texttt{[ret] = match get\_value (make\_cursor k t') t' with Some v' => v' | None =>
nullval end *} \\
\texttt{success = if (get\_value (make\_cursor k t') t') then 1 else 0}\} \\
\texttt{return ret;} \\
\caption{Annotation for Toy Client}\label{lst:annotation}
\end{listing}

The complete stack of proof has almost been completed, with some holes about
string theories. Therefore, we have established an end-to-end verification,
provided that the B+ tree is verified as well.

\section{Future Work}

\subsection{Specification}

The specification has been at least implemented by a list of pairs. Not too much
changes will be expected, but I suppose it still needs some improvement. The
client of the index engine in the database system would have a better idea what
guarantee it want.

Meanwhile, connection from B+ tree's functional implementation to the
specification is still needed. Especially, the relational style \texttt{put} and
\texttt{empty} might require some work.

\subsection{MassTree}

As for now, most of the design and implementation works have been done, while
the verification is lagging behind. I would expect that the proof and from the C
programs to the functional programs wouldn't encounter too much difficulty.
Because for most of the proof it's collecting the branching decision as one
symbolic execution the code with help of VST, and then using the decisions to
guide the functional program's execution. However, the proof from functional
programs to the specification properties might be difficult. Especially, the
proof to reason the order of keys might require some auxiliary predicates over
the trie globally to assist the proof.

\subsection{New Operations}

So far, work has only been done on the \texttt{put} operation. The
\texttt{delete} operation is left untouched. I expect that this operation might
conflict with the current implementation of \texttt{first\_cursor} and
\texttt{next\_cursor}, as I have briefly mentioned in previous section.

\bibliographystyle{splncs04}
\bibliography{reference.bib}

\section*{Appendix A:\@\texttt{funspec} for Interface Functions of B+ Tree}

\begin{minted}[breaklines]{coq}
Definition Iempty_spec: ident * funspec :=
  DECLARE _Iempty
  WITH tt: unit
  PRE [ ]
  PROP ()
  LOCAL ()
  SEP ()
  POST [ tptr BTree.tindex ] EX t: @BTree.table val, EX pt: val,
  PROP (BTree.empty t)
  LOCAL (temp ret_temp pt)
  SEP (BTree.table_rep t pt).

Definition Iput_spec: ident * funspec :=
  DECLARE _Iput
  WITH k: Z, v: val,
       t: @BTree.table val, pt: val,
       c: @BTree.cursor val, pc: val
  PRE [ _key OF tuint, _value OF tptr tvoid, _cursor OF tptr BTree.tcursor, _index OF tptr BTree.tindex ]
  PROP (BTree.abs_rel c t)
  LOCAL (temp _key (Vint (Int.repr k)); temp _value v;
         temp _cursor pc; temp _index pt)
  SEP (BTree.table_rep t pt; BTree.cursor_rep c pc)
  POST [ tvoid ]
  EX new_t: @BTree.table val, EX new_c: @BTree.cursor val,
  PROP (BTree.put k v c t new_c new_t)
  LOCAL ()
  SEP (BTree.table_rep new_t pt; BTree.cursor_rep new_c pc).

Definition Imake_cursor_spec: ident * funspec :=
  DECLARE _Imake_cursor
  WITH k: Z,
       t: @BTree.table val, pt: val
  PRE [ _key OF tuint, _index OF tptr BTree.tindex ]
  PROP (0 <= k <= Int.max_unsigned; BTree.table_correct t)
  LOCAL (temp _key (Vint (Int.repr k)); temp _index pt)
  SEP (BTree.table_rep t pt)
  POST [ tptr BTree.tcursor ] EX pc: val,
  PROP ()
  LOCAL (temp ret_temp pc)
  SEP (BTree.table_rep t pt; BTree.cursor_rep (BTree.make_cursor k t) pc).

Definition Ifirst_cursor_spec: ident * funspec :=
  DECLARE _Ifirst_cursor
  WITH t: @BTree.table val, pt: val
  PRE [ _index OF tptr BTree.tindex ]
  PROP (BTree.table_correct t)
  LOCAL (temp _index pt)
  SEP (BTree.table_rep t pt)
  POST [ tptr BTree.tcursor ] EX pc: val,
  PROP ()
  LOCAL (temp ret_temp pc)
  SEP (BTree.table_rep t pt; BTree.cursor_rep (BTree.first_cursor t) pc).

Definition Iget_key_spec: ident * funspec :=
  DECLARE _Iget_key
  WITH c: @BTree.cursor val, pc: val,
       t: @BTree.table val, pt: val,
       pk: val
  PRE [ _cursor OF tptr BTree.tcursor, _index OF tptr BTree.tindex, _key OF tptr tuint ]
  PROP (BTree.abs_rel c t)
  LOCAL (temp _cursor pc; temp _index pt; temp _key pk)
  SEP (BTree.table_rep t pt; BTree.cursor_rep c pc; data_at_ Tsh tuint pk)
  POST [ tint ]
  PROP ()
  LOCAL (temp ret_temp (if BTree.get_key c t then (Vint Int.one) else (Vint Int.zero)))
  SEP (BTree.table_rep t pt; BTree.cursor_rep c pc;
       data_at Tsh tuint match BTree.get_key c t with
                         | Some k => (Vint (Int.repr k))
                         | None => Vundef
                         end pk).

Definition Iget_value_spec: ident * funspec :=
  DECLARE _Iget_value
  WITH c: @BTree.cursor val, pc: val,
       t: @BTree.table val, pt: val,
       pv: val
  PRE [ _cursor OF tptr BTree.tcursor, _index OF tptr BTree.tindex, _value OF tptr (tptr tvoid) ]
  PROP (BTree.abs_rel c t)
  LOCAL (temp _cursor pc; temp _index pt; temp _value pv)
  SEP (BTree.table_rep t pt; BTree.cursor_rep c pc; data_at_ Tsh (tptr tvoid) pv)
  POST [ tint ]
  PROP ()
  LOCAL (temp ret_temp (if BTree.get_value c t then (Vint Int.one) else (Vint Int.zero)))
  SEP (BTree.table_rep t pt; BTree.cursor_rep c pc;
       data_at Tsh (tptr tvoid) match BTree.get_value c t with
                                | Some v => v
                                | None => Vundef
                                end pv).
\end{minted}

\section*{Appendix B:\@Functional Program for Trie}

\begin{minted}[breaklines]{coq}
Module Trie (Node: FLATTENABLE_TABLE Keyslice)
            <: FLATTENABLE_TABLE KeyType.
  [...]
  Definition empty (t: trie) :=
    match t with
    | trienode_of _ trieform listform =>
      Node.empty trieform /\ Node.Flattened.empty listform
    end.

  Function make_cursor (k: key) (t: trie) {measure length k}: cursor :=
    let keyslice := get_keyslice k in
    match t with
    | trienode_of _ tableform listform =>
      match Node.Flattened.get_exact keyslice listform with
      | Some (_, bnode) =>
        if (Z_le_dec (Zlength k) keyslice_length) then
          [(t, Node.make_cursor keyslice tableform,
            bnode, BorderNode.before_prefix (Zlength k))]
        else
          match fst (BorderNode.get_suffix_pair bnode) with
          | None =>
            match BorderNode.get_suffix None bnode with
            | value_of _ => []
            | nil => [(t, Node.make_cursor keyslice tableform,
                       bnode, BorderNode.after_suffix)]
            | trie_of t' =>
              (t, Node.make_cursor keyslice tableform,
               bnode, BorderNode.before_suffix) :: make_cursor (get_suffix k) t'
            end
          | Some k' =>
            if (TrieKeyFacts.lt_dec k' (get_suffix k)) then
              [(t, Node.make_cursor keyslice tableform,
                bnode, BorderNode.after_suffix)]
            else
              [(t, Node.make_cursor keyslice tableform,
                bnode, BorderNode.before_suffix)]
          end
      | None => []
      end
    end.

  Function strict_first_cursor (t: trie) {measure trie_height t}: option cursor :=
    match t with
    | trienode_of _ tableform listform =>
      match Node.Flattened.get_value
              (Node.Flattened.first_cursor listform) listform with
      | Some (_, bnode) =>
        match BorderNode.first_cursor bnode with
        | BorderNode.before_prefix len =>
          match (BorderNode.get_prefix len bnode) with
          | value_of _ =>
            Some [(t, Node.first_cursor tableform,
                   bnode, BorderNode.before_prefix len)]
          | _ => None (* impossible *)
          end
        | BorderNode.before_suffix =>
          match snd (BorderNode.get_suffix_pair bnode) with
          | trie_of t' =>
            match strict_first_cursor t' with
            | Some c' => Some ((t, Node.first_cursor tableform,
                                bnode, BorderNode.before_suffix) :: c')
            | None => None (* impossible *)
            end
          | value_of _ =>
            Some [(t, Node.first_cursor tableform,
                   bnode, BorderNode.before_suffix)]
          | nil => None (* impossible *)
          end
        | BorderNode.after_suffix => None
        end
      | None => None
      end
    end.

  Fixpoint normalize_cursor (c: cursor): option cursor :=
    match c with
    | [] => None
    | (trienode_of addr tableform listform, table_cursor, bnode, bnode_cursor) :: c' =>
      let t := trienode_of addr tableform listform in
      match normalize_cursor c' with
      | Some c'' => Some ((t, table_cursor, bnode, bnode_cursor) :: c'')
      | None =>
        match BorderNode.next_cursor bnode_cursor bnode with
        | BorderNode.before_prefix len =>
          Some [(t, table_cursor, bnode, (BorderNode.before_prefix len))]
        | BorderNode.before_suffix =>
          match (snd (BorderNode.get_suffix_pair bnode)) with
          | nil => None
          | value_of _ => Some [(t, table_cursor, bnode, BorderNode.before_suffix)]
          | trie_of t' =>
            match strict_first_cursor t' with
            | Some c' =>
              Some ((t, table_cursor, bnode, BorderNode.before_suffix) :: c')
            | None =>
              None
            end
          end
        | BorderNode.after_suffix =>
          let table_cursor' := Node.next_cursor table_cursor tableform in
          match Node.get_key table_cursor' tableform with
          | Some key =>
            match Node.Flattened.get_value (Node.Flattened.make_cursor key listform) listform with
            | Some (_, bnode') =>
              match BorderNode.next_cursor (BorderNode.before_prefix 1) bnode' with
              | BorderNode.before_prefix len =>
                match (BorderNode.get_prefix len bnode') with
                | nil => None
                | value_of _ => Some [(t, table_cursor', bnode', (BorderNode.before_prefix len))]
                | trie_of t' => None
                end
              | BorderNode.before_suffix =>
                match (snd (BorderNode.get_suffix_pair bnode')) with
                | nil => None
                | value_of _ => Some [(t, table_cursor', bnode', BorderNode.before_suffix)]
                | trie_of t' =>
                  match strict_first_cursor t' with
                  | Some c' =>
                    Some ((t, table_cursor', bnode', BorderNode.before_suffix) :: c')
                  | None =>
                    None
                  end
                end
              | BorderNode.after_suffix => None
              end
            | None => None
            end
          | None => None
          end
        end
      end
    end.

  Fixpoint strict_next_cursor (c: cursor): cursor :=
    match c with
    | [] => []
    | [(trienode_of addr tableform listform, table_cursor, bnode, bnode_cursor)] =>
      match bnode_cursor with
      | BorderNode.before_prefix len =>
        if Z_lt_dec len keyslice_length then
          [(trienode_of addr tableform listform, table_cursor, bnode, BorderNode.before_prefix (len + 1))]
        else
          [(trienode_of addr tableform listform, table_cursor, bnode, BorderNode.after_suffix)]
      | BorderNode.before_suffix => [(trienode_of addr tableform listform, table_cursor, bnode, BorderNode.after_suffix)]
      | BorderNode.after_suffix =>
        let table_cursor' := Node.next_cursor table_cursor tableform in
        match Node.get_key table_cursor' tableform with
        | Some key =>
          match Node.Flattened.get_value (Node.Flattened.make_cursor key listform) listform with
          | Some bnode' =>
            [(trienode_of addr tableform listform, table_cursor, bnode, BorderNode.before_prefix 1)]
          | None => []
          end
        | None => [] (* should not be the case in the only scenario of usage *)
        end
      end
    | current_slice :: c' =>
      current_slice :: (strict_next_cursor c')
    end.

  Definition next_cursor (c: cursor) (t: table): cursor :=
    match normalize_cursor c with
    | Some c' =>
      strict_next_cursor c'
    | None =>
      []
    end.

  Definition first_cursor (c: cursor) (t: table): cursor :=
    match strict_first_cursor c with
    | Some c' => c'
    | None => []
    end.

  Fixpoint get_raw (c: cursor): option (key * value) :=
    match c with
    | (trienode_of _ tableform _, table_cursor, bnode, bnode_cursor) :: c' =>
      match Node.get_key table_cursor tableform with
      | Some keyslice =>
        match bnode_cursor with
        | BorderNode.before_prefix len =>
          match BorderNode.get_prefix len bnode with
          | value_of v => Some (reconstruct_keyslice (keyslice, len), v)
          | trie_of t' => None
          | nil => None
          end
        | BorderNode.before_suffix =>
          match (snd (BorderNode.get_suffix_pair bnode)) with
          | value_of v => Some (reconstruct_keyslice (keyslice, keyslice_length), v)
          | trie_of t' =>
            match get_raw c' with
            | Some (k', v') => Some (reconstruct_keyslice (keyslice, keyslice_length) ++ k', v')
            | None => None
            end
          | nil => None
          end
        | BorderNode.after_suffix => None
        end
      | None => None
      end
    | [] => None
    end.

  Definition get (c: cursor) (t: table): option (key * value) :=
    match normalize_cursor c with
    | Some c' =>
      get_raw c'
    | None => None
    end.

  Inductive create_pair (k1 k2: key) (v1 v2: value): cursor -> table -> Prop :=
    | create_pair_case1: forall emptylist emptytable listform tableform listcursor tablecursor bnode_addr addr,
    let keyslice1 := get_keyslice k1 in
    let keyslice2 := get_keyslice k2 in
    keyslice1 = keyslice2 ->
    Zlength k1 <= keyslice_length \/ Zlength k2 <= keyslice_length ->
    let bnode := BorderNode.put_value k1 (value_of v1) BorderNode.empty in
    let bnode := BorderNode.put_value k2 (value_of v2) bnode in
    Node.Flattened.empty emptylist ->
    Node.Flattened.put keyslice1 (bnode_addr, bnode)
                       (Node.Flattened.first_cursor emptylist) emptylist
                       listcursor listform ->
    Node.empty emptytable ->
    Node.put keyslice1 bnode_addr (Node.first_cursor emptytable) emptytable tablecursor tableform ->
    create_pair k1 k2 v1 v2
                [(trienode_of addr tableform listform, tablecursor, bnode, BorderNode.length_to_cursor (Zlength k1))]
                (trienode_of addr tableform listform)
    | create_pair_case2: forall emptylist emptytable listform tableform listcursor tablecursor bnode_addr c' t' addr,
      let keyslice1 := get_keyslice k1 in
      let keyslice2 := get_keyslice k2 in
      keyslice1 = keyslice2 ->
      Zlength k1 > keyslice_length /\ Zlength k2 > keyslice_length ->
      create_pair (get_suffix k1) (get_suffix k2) v1 v2 c' t' ->
      let bnode := BorderNode.put_suffix None (trie_of t') BorderNode.empty in
      Node.Flattened.empty emptylist ->
      Node.Flattened.put keyslice1 (bnode_addr, bnode)
                         (Node.Flattened.first_cursor emptylist) emptylist
                         listcursor listform ->
      Node.empty emptytable ->
      Node.put keyslice1 bnode_addr (Node.first_cursor emptytable) emptytable tablecursor tableform ->
      create_pair k1 k2 v1 v2
                  ((trienode_of addr tableform listform, tablecursor, bnode, BorderNode.before_suffix) :: c')
                  (trienode_of addr tableform listform)
    | create_pair_case3: forall emptylist emptytable listform1 listform2 tableform1 tableform2 listcursor1 listcursor2 tablecursor1 tablecursor2 bnode_addr1 bnode_addr2 addr,
      let keyslice1 := get_keyslice k1 in
      let keyslice2 := get_keyslice k2 in
      keyslice1 <> keyslice2 ->
      let bnode1 := BorderNode.put_value k1 (value_of v1) BorderNode.empty in
      let bnode2 := BorderNode.put_value k2 (value_of v2) BorderNode.empty in
      Node.Flattened.empty emptylist ->
      Node.Flattened.put keyslice2 (bnode_addr2, bnode2)
                         (Node.Flattened.first_cursor emptylist) emptylist
                         listcursor2 listform2 ->
      Node.Flattened.put keyslice1 (bnode_addr1, bnode1)
                         (Node.Flattened.first_cursor listform2) listform2
                         listcursor1 listform1 ->
      Node.empty emptytable ->
      Node.put keyslice2 bnode_addr2 (Node.first_cursor emptytable) emptytable tablecursor2 tableform2 ->
      Node.put keyslice1 bnode_addr1 (Node.first_cursor tableform2) tableform2 tablecursor1 tableform1 ->
      create_pair k1 k2 v1 v2
                  [(trienode_of addr tableform1 listform1, tablecursor1, bnode1, BorderNode.length_to_cursor (Zlength k1))]
                  (trienode_of addr tableform1 listform1).

    Inductive put (k: key) (v: value): cursor -> trie -> cursor -> trie -> Prop :=
    | put_case1: forall tableform listform listform' listcursor listcursor' bnode_addr bnode c addr,
      let keyslice := get_keyslice k in
      list_get_exact keyslice listform = Some (bnode_addr, bnode) ->
      Zlength k <= keyslice_length ->
      let bnode := BorderNode.put_prefix (Zlength k) (value_of v) bnode in
      Node.Flattened.abs_rel listcursor listform ->
      Node.Flattened.put keyslice (bnode_addr, bnode)
                         listcursor listform
                         listcursor' listform' ->
      let tablecursor := Node.make_cursor keyslice tableform in
      put k v c (trienode_of addr tableform listform)
          [(trienode_of addr tableform listform', tablecursor, bnode, BorderNode.before_prefix (Zlength k))]
          (trienode_of addr tableform listform')
    | put_case2: forall tableform listform listform' listcursor listcursor' bnode_addr bnode c t' c' t'' c'' addr,
      let keyslice := get_keyslice k in
      list_get_exact keyslice listform = Some (bnode_addr, bnode) ->
      Zlength k > keyslice_length ->
      BorderNode.get_suffix None bnode = trie_of t' ->
      put (get_suffix k) v c' t' c'' t'' ->
      let bnode := BorderNode.put_suffix None (trie_of t'') bnode in
      Node.Flattened.abs_rel listcursor listform ->
      Node.Flattened.put keyslice (bnode_addr, bnode)
                         listcursor listform
                         listcursor' listform' ->
      let tablecursor := Node.make_cursor keyslice tableform in
      put k v c (trienode_of addr tableform listform)
          ((trienode_of addr tableform listform', tablecursor, bnode, BorderNode.before_suffix) :: c'')
          (trienode_of addr tableform listform')
    | put_case3: forall tableform listform listform' listcursor listcursor' bnode_addr bnode c addr,
      let keyslice := get_keyslice k in
      list_get_exact keyslice listform = Some (bnode_addr, bnode) ->
      Zlength k > keyslice_length ->
      BorderNode.get_suffix_pair bnode = (None, nil) ->
      let bnode := BorderNode.put_suffix (Some (get_suffix k)) (value_of v) bnode in
      Node.Flattened.abs_rel listcursor listform ->
      Node.Flattened.put keyslice (bnode_addr, bnode)
                         listcursor listform
                         listcursor' listform' ->
      let tablecursor := Node.make_cursor keyslice tableform in
      put k v c (trienode_of addr tableform listform)
          [(trienode_of addr tableform listform', tablecursor, bnode, BorderNode.before_suffix)]
          (trienode_of addr tableform listform')
    | put_case4: forall tableform listform listform' listcursor listcursor' bnode_addr bnode c addr,
      let keyslice := get_keyslice k in
      list_get_exact keyslice listform = Some (bnode_addr, bnode) ->
      Zlength k > keyslice_length ->
      fst (BorderNode.get_suffix_pair bnode) = Some k ->
      let bnode := BorderNode.put_suffix (Some (get_suffix k)) (value_of v) bnode in
      Node.Flattened.abs_rel listcursor listform ->
      Node.Flattened.put keyslice (bnode_addr, bnode)
                         listcursor listform
                         listcursor' listform' ->
      let tablecursor := Node.make_cursor keyslice tableform in
      put k v c (trienode_of addr tableform listform)
          [(trienode_of addr tableform listform', tablecursor, bnode, BorderNode.before_suffix)]
          (trienode_of addr tableform listform')
    | put_case5: forall tableform listform listform' listcursor listcursor' bnode_addr bnode c c' t' k' v' addr,
      let keyslice := get_keyslice k in
      list_get_exact keyslice listform = Some (bnode_addr, bnode) ->
      Zlength k > keyslice_length ->
      BorderNode.get_suffix_pair bnode = (Some k', value_of v') ->
      get_suffix k <> k' ->
      create_pair (get_suffix k) k' v v' c' t' ->
      let bnode := BorderNode.put_suffix None (trie_of t') bnode in
      Node.Flattened.abs_rel listcursor listform ->
      Node.Flattened.put keyslice (bnode_addr, bnode)
                         listcursor listform
                         listcursor' listform' ->
      let tablecursor := Node.make_cursor keyslice tableform in
      put k v c (trienode_of addr tableform listform)
          ((trienode_of addr tableform listform', tablecursor, bnode, BorderNode.before_suffix) :: c')
          (trienode_of addr tableform listform')
    | put_case6: forall tableform tableform' listform listform' tablecursor tablecursor' listcursor listcursor' bnode_addr c addr,
      let keyslice := get_keyslice k in
      list_get_exact keyslice listform = None ->
      let bnode := BorderNode.put_value k (value_of v) BorderNode.empty in
      Node.Flattened.abs_rel listcursor listform ->
      Node.Flattened.put keyslice (bnode_addr, bnode)
                         listcursor listform
                         listcursor' listform' ->
      Node.abs_rel tablecursor tableform ->
      Node.put keyslice bnode_addr
                         tablecursor tableform
                         tablecursor' tableform' ->
      put k v c (trienode_of addr tableform listform)
          [(trienode_of addr tableform' listform', tablecursor', bnode, BorderNode.before_suffix)]
          (trienode_of addr tableform' listform').

    Inductive table_correct: table -> Prop :=
    | table_correct_intro tableform listform: forall addr,
      Node.table_correct tableform ->
      Node.Flattened.table_correct listform ->
      map fst (Node.flatten tableform) = map fst listform ->
      map snd (Node.flatten tableform) = map (compose fst snd) listform ->
      Forall (compose bordernode_correct (compose snd snd)) listform ->
      table_correct (trienode_of addr tableform listform)
    with
    bordernode_correct: @BorderNode.table link -> Prop :=
    | bordernode_correct_intro prefixes (k: option string) l:
      Zlength prefixes = keyslice_length ->
      Forall (fun l => is_value l \/ l = nil) prefixes ->
      (k = None -> subtrie_correct l) ->
      (k <> None -> is_value l) ->
      match k with
      | Some k' => 0 < Zlength k'
      | None => True
      end ->
      (* optional: no dangling(dead end) in the tree, useful for [next_cursor] and [first_cursor] *)
      (l <> nil \/ Exists (fun l => l <> nil) prefixes) ->
      bordernode_correct (prefixes, k, l)
    with
    subtrie_correct: link -> Prop :=
    | nil_correct: subtrie_correct nil
    | subtrie_correct_intro t:
      table_correct t ->
      (* optional: no dangling(dead end) in the tree, useful for [next_cursor] and [first_cursor] *)
      ~ empty t ->
      subtrie_correct (trie_of t).

    Fixpoint cursor_correct_aux (c: cursor) (p: option trie): Prop :=
      match c with
      | (trienode_of addr tableform listform, table_cursor, bnode, bnode_cursor) :: c' =>
        match p with
        | Some t => t = trienode_of addr tableform listform
        | None => True
        end /\
        table_correct (trienode_of addr tableform listform) /\
        Node.cursor_correct table_cursor /\
        BorderNode.cursor_correct bnode_cursor /\
        match Node.get_key table_cursor tableform with
        | Some key =>
          match Node.Flattened.get_value (Node.Flattened.make_cursor key listform) listform with
          | Some (_, bnode') =>
            bnode' = bnode /\
            match bnode_cursor with
            | BorderNode.before_prefix len =>
              match BorderNode.get_prefix len bnode with
              | trie_of t' => False
              | _ => c' = []
              end
            | BorderNode.before_suffix =>
              match (snd (BorderNode.get_suffix_pair bnode)) with
              | trie_of t' => cursor_correct_aux c' (Some t')
              | _ => c' = []
              end
            | BorderNode.after_suffix => c' = []
            end
          | None => False
          end
        | None => False
        end
      | [] => True
      end.

    Definition cursor_correct (c: cursor): Prop := cursor_correct_aux c None.

    Definition abs_rel (c: cursor) (t: table): Prop :=
      cursor_correct c /\ table_correct t /\
      match c with
      | (t', _, _, _) :: _ =>
        t' = t
      | [] => True
      end.

    Definition key_rel (k: key) (c: cursor) (t: table): Prop :=
      normalize_cursor c = normalize_cursor (make_cursor k t).

    Definition eq_cursor (c1 c2: cursor) (t: table):
      normalize_cursor c1 = normalize_cursor c2.
End Trie.
\end{minted}

\end{document}

### 2018-08-28 papers/luke/report.tex
turn table into map

@@
 further step where we parameterized over the dependency: for any module that
-implements the specification of table of integer-typed key with
+implements the specification of kv-map of integer-typed key with
 cursors, we can derive a module that implements the same specification of
@@
 
-\section{Specification for Table with Cursors}\label{spec}
+\section{Specification for KV-Map with Cursors}\label{spec}
 
@@
 structures. The specification can be generally divided into two parts: the part
-where it guaranteed the correctness of the table itself and the part where it
+where it guaranteed the correctness of the kv-map itself and the part where it
 provided essential utilities for clients' reasoning.
@@
 
-Spiritually, a table maintains the mapping from keys to values and a cursor
-points into the table at a specific position and supports movement operations.
+Spiritually, a kv-map maintains the mapping from keys to values and a cursor
+points into the kv-map at a specific position and supports movement operations.
 Concretely, such a data structure should provide the operations in
@@
   \begin{minted}{coq}
-Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
+Module Type KV_MAP (KeyType: UsualOrderedType).
   Definition key := KeyType.t.
-  Parameter table: Type -> Type.
+  Parameter map: Type -> Type.
   Parameter cursor: Type -> Type.
@@
     Context {value: Type}.
-    Parameter empty: table value -> Prop.
-
-    Parameter make_cursor: key -> table value -> cursor value.
-    Parameter first_cursor: table value -> cursor value.
-    Parameter last_cursor: table value -> cursor value.
-    Parameter next_cursor: cursor value -> table value -> cursor value.
-    Parameter prev_cursor: cursor value -> table value -> cursor value.
-
-    Parameter get: cursor value -> table value -> option (key * value).
-
-    Parameter put: key -> value -> cursor value -> table value -> (* input *)
-                   cursor value -> table value -> (* output *)
+    Parameter empty: map value -> Prop.
+
+    Parameter make_cursor: key -> map value -> cursor value.
+    Parameter first_cursor: map value -> cursor value.
+    Parameter last_cursor: map value -> cursor value.
+    Parameter next_cursor: cursor value -> map value -> cursor value.
+    Parameter prev_cursor: cursor value -> map value -> cursor value.
+
+    Parameter get: cursor value -> map value -> option (key * value).
+
+    Parameter put: key -> value -> cursor value -> map value -> (* input *)
+                   cursor value -> map value -> (* output *)
                    Prop.
@@
   End Types.
-End ABSTRACT_TABLE.
-\end{minted}
-  \caption{Interface for Table with Cursors}\label{fig:interface}
-\end{figure}
-
-This is a typical interface of tables which supports operations for creation of
-table (\texttt{empty}), creation of cursor (\texttt{make\_cursor},
+End KV_MAP.
+\end{minted}
+  \caption{Interface for KV-Map with Cursors}\label{fig:interface}
+\end{figure}
+
+This is a typical interface of kv-map which supports operations for creation of
+map (\texttt{empty}), creation of cursor (\texttt{make\_cursor},
 \texttt{first\_cursor}, \texttt{last\_cursor}), movement of cursor
 (\texttt{next\_cursor}, \texttt{prev\_cursor}), fetching data at
-cursor (\texttt{get}) and updating the table (\texttt{put}). The types of
+cursor (\texttt{get}) and updating the map (\texttt{put}). The types of
 functions indicates clearly that all other functions except for \texttt{empty}
-and \texttt{put} are immutable with respect to the table. The two table mutating
+and \texttt{put} are immutable with respect to the map. The two map mutating
 functions are defined as relations rather than functions, this is a compromise
@@
 Apart from these functions, we also need some auxiliary predicates, as shown in
-figure~\ref{fig:aux}. \texttt{abs\_rel} asserts that a cursor and a table are
+figure~\ref{fig:aux}. \texttt{abs\_rel} asserts that a cursor and a kv-map are
 correctly associated with each other (the abbreviation \texttt{abs} should
@@
 historical issue). \texttt{key\_rel} asserts that a cursor is associated with
-a key in a table. One can imagine this predicate as \textit{the cursor is placed
+a key in a kv-map. One can imagine this predicate as \textit{the cursor is placed
 at the position such that all the existing keys to the decreasing side are less
 than this key, and all the existing keys to the opposite are no less than this
-key}. This makes sense when one visualizes a table as a list of pairs of key and
+key}. This makes sense when one visualizes a kv-map as a list of pairs of key and
 value. This definition is closely related to the semantics of \texttt{get}, We
-will see later how this is used in generalized tables. \texttt{eq\_cursor}
+will see later how this is used in generalized kv-map. \texttt{eq\_cursor}
 asserts that two cursors are equivalent. Coq equality (\texttt{eq}) does not
 work well here, because of \texttt{eq} implies structural equality, and
-structurally different cursors can be equivalent (especially for those table
+structurally different cursors can be equivalent (especially for those kv-map
 with complex structures, such as B+ trees or tries). The final two predicates
-\texttt{cursor\_correct} and \texttt{table\_correct} are the normal invariants.
+\texttt{cursor\_correct} and \texttt{map\_correct} are the normal invariants.
 
@@
   \begin{minted}{coq}
-Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
-  [...]
-  Parameter abs_rel: cursor value -> table value -> Prop.
-  Parameter key_rel: key -> cursor value -> table value -> Prop.
-  Parameter eq_cursor : cursor value -> cursor value -> table value -> Prop.
+Module Type KV_MAP (KeyType: UsualOrderedType).
+  [...]
+  Parameter abs_rel: cursor value -> map value -> Prop.
+  Parameter key_rel: key -> cursor value -> map value -> Prop.
+  Parameter eq_cursor : cursor value -> cursor value -> map value -> Prop.
   Parameter cursor_correct: cursor value -> Prop.
-  Parameter table_correct: table value -> Prop.
-  [...]
-End ABSTRACT_TABLE.
-\end{minted}
-  \caption{Auxiliary Predicates of Table with Cursors}\label{fig:aux}
+  Parameter map_correct: map value -> Prop.
+  [...]
+End KV_MAP.
+\end{minted}
+  \caption{Auxiliary Predicates of KV-Map with Cursors}\label{fig:aux}
 \end{figure}
@@
   \begin{minted}{coq}
-Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
-  [...]
-  Axiom empty_correct: forall t, empty t -> table_correct t.
+Module Type KV_MAP (KeyType: UsualOrderedType).
+  [...]
+  Axiom empty_correct: forall t, empty t -> map_correct t.
   Axiom get_empty: forall t c, empty t -> abs_rel c t -> get c t = None.
   [...]
-End ABSTRACT_TABLE.
+End KV_MAP.
 \end{minted}
@@
 
-For an empty table, it should satisfy the correctness invariant and also that we
-are never able to get anything from cursor associated with the table.
+For an empty map, it should satisfy the correctness invariant and also that we
+are never able to get anything from cursor associated with the kv-map.
 
@@
   \begin{minted}{coq}
-Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
+Module Type KV_MAP (KeyType: UsualOrderedType).
   [...]
   Axiom make_cursor_key: forall t k,
-    table_correct t -> key_rel k (make_cursor k t) t.
+    map_correct t -> key_rel k (make_cursor k t) t.
   Axiom make_cursor_abs: forall t k,
-    table_correct t -> abs_rel (make_cursor k t) t.
-  [...]
-End ABSTRACT_TABLE.
+    map_correct t -> abs_rel (make_cursor k t) t.
+  [...]
+End KV_MAP.
 \end{minted}
@@
 For the \texttt{make\_cursor} operation, the cursor it made should always be
-associated with the key and the table.
+associated with the key and the kv-map.
 
@@
   \begin{minted}{coq}
-Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
+Module Type KV_MAP (KeyType: UsualOrderedType).
   [...]
   Axiom first_cursor_abs: forall t,
-    table_correct t -> abs_rel (first_cursor t) t.
+    map_correct t -> abs_rel (first_cursor t) t.
   Axiom last_cursor_abs: forall t,
-    table_correct t -> abs_rel (last_cursor t) t.
+    map_correct t -> abs_rel (last_cursor t) t.
   Axiom get_last: forall t, get (last_cursor t) t = None.
   Axiom first_cursor_get_empty: forall t,
-    table_correct t -> get (first_cursor t) t = None -> empty t.
-  [...]
-End ABSTRACT_TABLE.
+    map_correct t -> get (first_cursor t) t = None -> empty t.
+  [...]
+End KV_MAP.
 \end{minted}
@@
 \texttt{first\_cursor} and \texttt{last\_cursor} are dual operations. Both of
-them should yield cursors associated with the table. Additionally, the
-\texttt{last\_cursor} should point past the end of the table, meaning that we
+them should yield cursors associated with the kv-map. Additionally, the
+\texttt{last\_cursor} should point past the end of the kv-map, meaning that we
 can get nothing from it. The \texttt{first\_cursor} should always return a
-binding, unless the table is empty.
+binding, unless the map is empty.
 
@@
   \begin{minted}{coq}
-Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
+Module Type KV_MAP (KeyType: UsualOrderedType).
   [...]
@@
   [...]
-End ABSTRACT_TABLE.
+End KV_MAP.
 \end{minted}
@@
 \texttt{next\_cursor} and \texttt{prev\_cursor} is another pair of dual
-operations. Both of them should yield cursors associated with the table. Unless
-at the end of table, movement should reach a cursor associated with keys of the
+operations. Both of them should yield cursors associated with the kv-map. Unless
+at the end of kv-map, movement should reach a cursor associated with keys of the
 right order, they should be inverse to each other and they should move only to
@@
     put k v c1 t1 c2 t2 ->
-    table_correct t2.
+    map_correct t2.
 
@@
 shadow property as well. Additionally, \texttt{put} operation should keep a
-correct table correct after mutation. Notice that, the \texttt{put} takes an
+correct kv-map correct after mutation. Notice that, the \texttt{put} takes an
 extra cursor as input but is not specified in the axioms. This cursor is there
@@
   \begin{minted}{coq}
-Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
+Module Type KV_MAP (KeyType: UsualOrderedType).
   [...]
   Axiom abs_correct: forall t c,
-    abs_rel c t -> cursor_correct c /\ table_correct t.
+    abs_rel c t -> cursor_correct c /\ map_correct t.
 
@@
 regulate the predicates. \texttt{abs\_rel} should imply that both the cursor and
-table correct. Cursors associated with the same key should be equivalent and
+map correct. Cursors associated with the same key should be equivalent and
 equivalent cursors should return the same result.
@@
 interface is defined with functions. Therefore for those functions that mutate
-the table, we are compelled to define them relationally.
+the kv-map, we are compelled to define them relationally.
 
@@
 arbitrary \texttt{value} types. Some clients even want to define inductive types
-with the table. A minimal example is the finite-degree tree where at each node
+with the kv-map. A minimal example is the finite-degree tree where at each node
 it maintains a map from keys to children. However, such definition will never
@@
 \begin{minted}{coq}
-Module Tree (K: UsualOrderedType) (NodeTable: Table K).
+Module Tree (K: UsualOrderedType) (NodeMap: KV_MAP K).
   Inductive tree: Type :=
-  | node_of: NodeTable.table tree -> tree.
+  | node_of: NodeMap.map tree -> tree.
 End Tree.
@@
 been quite a few new functions declared without specification, and even more are
-required to work with table mutations. As the amount of new functions suggest,
+required to work with map mutations. As the amount of new functions suggest,
 additional specification should be added so that separation logic clients live a
@@
 \begin{minted}{coq}
-Module Client (K: UsualOrderedType) (T: Table K).
+Module Client (K: UsualOrderedType) (T: KV_MAP K).
   Parameter value: Type.
   Parameter value_rep: value -> val -> mpred.
-  Definition t: Type := T.table value.
+  Definition t: Type := T.map value.
 
   Definition rep: (data: t) (p: val): mpred :=
-    EX concrete_table: T.table val,
-    !! (Table.keys concrete_table = Table.keys data) &&
-    Table.table_rep concrete_table p *
-    Table.iter_sepcon data concrete_table value_rep.
+    EX concrete_map: T.map val,
+    !! (T.keys concrete_map = T.keys data) &&
+    T.map_rep concrete_map p *
+    T.iter_sepcon data concrete_map value_rep.
 End Tree.
@@
 The remedy for both issues is the \texttt{flatten} function. \texttt{flatten}
-turns a table into list of pairs of key and value, where the list is sorted in
+turns a map into list of pairs of key and value, where the list is sorted in
 the order of keys. Furthermore, The list have been proven to be an
-implementation of the same interface as table with cursors. Formally, these
+implementation of the same interface as map with cursors. Formally, these
 properties are specified in Figure~\ref{fig:flatten}.
@@
 \begin{minted}{coq}
-Module Type FLATTENABLE_TABLE (KeyType: UsualOrderedType)
-                              <: ABSTRACT_TABLE KeyType.
-  Include ABSTRACT_TABLE KeyType.
-  Module Flattened := SortedListTable KeyType.
+Module Type FLATTENABLE_MAP (KeyType: UsualOrderedType)
+                              <: KV_MAP KeyType.
+  Include KV_MAP KeyType.
+  Module Flattened := SortedListMap KeyType.
   Section Spec.
     Context {value: Type}.
-    Parameter flatten: table value -> Flattened.table value.
+    Parameter flatten: map value -> Flattened.map value.
     Axiom flatten_invariant: forall t,
-        table_correct t ->
-        Flattened.table_correct (flatten t) /\
+        map_correct t ->
+        Flattened.map_correct (flatten t) /\
         forall (k: key) (c1: cursor value) (c2: Flattened.cursor value),
@@
   End Spec.
-End FLATTENABLE_TABLE.
-\end{minted}
-  \caption{Flattenable Table}\label{fig:flatten}
+End FLATTENABLE_MAP.
+\end{minted}
+  \caption{Flattenable Map}\label{fig:flatten}
 \end{figure}
@@
 \begin{minted}{coq}
-Module Type FlattenableTableFacts (KeyType: UsualOrderedType)
-                                  (Table: FLATTENABLE_TABLE KeyType).
-  Include Table.
+Module Type FlattenableMapFacts (KeyType: UsualOrderedType)
+                                (Map: FLATTENABLE_MAP KeyType).
+  Include Map.
   Section Implication.
@@
     Theorem put_permute: forall t1 t2 c1 c2 k v,
-      table_correct t1 ->
+      map_correct t1 ->
       abs_rel c1 t1 ->
@@
     Theorem Forall_put (P: value -> Prop): forall t k v,
-      Flattened.table_correct t ->
+      Flattened.map_correct t ->
       P v ->
@@
     Theorem iter_sepcon_put (P: value -> mpred): forall t k v,
-      Flattened.table_correct t ->
+      Flattened.map_correct t ->
       P v * iter_sepcon P (flatten t) |--
@@
       iter_sepcon P (Flattened.fput k v t).
-End FLATTENABLE_TABLE.
-\end{minted}
-  \caption{Facts about Flattenable Table}\label{fig:flatten_fact}
+End FLATTENABLE_MAP.
+\end{minted}
+  \caption{Facts about Flattenable Map}\label{fig:flatten_fact}
 \end{figure}
@@
 Figure~\ref{fig:typedef} defines the type for functional trie. A trie is defined
-inductively, containing an address, a node table maps from \texttt{keyslice}s to
-addresses and a flattened node table maps from \texttt{keyslice}s to a pair of
+inductively, containing an address, a node maps from \texttt{keyslice}s to
+addresses and a flattened node maps from \texttt{keyslice}s to a pair of
 address and \texttt{BorderNode}. While \texttt{BorderNode} can optionally points
@@
 One should have noticed that there is a duplication of information: both the
-table form of node and list form of node is saved. We need the table form
+map form of node and list form of node is saved. We need the map form
 because of we need the unmodified augmented types for representation in
@@
 contains a list of prefix links, an optional suffix key field and a suffix link
-field. When there is a suffix key, the suffix link field should point to a
-final value. Otherwise, the suffix link field should optionally point to
-another trie. Although the \texttt{BorderNode}'s ``value'' type is always the
-same. Different endpoints (prefix links and suffix link) have different
-requirement and not all three constructors are valid, future work might refine
-this part better.
+field. When there is a suffix key, the suffix link field should point to a final
+value. Otherwise, the suffix link field should optionally point to another trie.
+Although the \texttt{BorderNode}'s ``value'' type is always the same. Different
+endpoints (prefix links and suffix link) have different requirement and not all
+three constructors are valid, future work might refine this part better.
 
@@
   [...]
-  Definition table: Type := list value * option string * value.
+  Definition map: Type := list value * option string * value.
   Inductive cursor: Type :=
@@
 
-Module Trie (Node: FLATTENABLE_TABLE Keyslice) <: FLATTENABLE_TABLE KeyType.
+Module Trie (Node: FLATTENABLE_MAP Keyslice) <: FLATTENABLE_MAP KeyType.
   [...]
@@
   | trienode_of: val ->
-                 Node.table val ->
-                 Node.Flattened.table (val * BorderNode.table link) ->
+                 Node.map val ->
+                 Node.Flattened.map (val * BorderNode.map link) ->
                  trie
@@
   Definition cursor: Type := list
-     (trie * Node.cursor val * (BorderNode.table link) * BorderNode.cursor).
+     (trie * Node.cursor val * (BorderNode.map link) * BorderNode.cursor).
   [...]
@@
 The \texttt{empty} predicate is defined straightforward. The root node of the trie
-should contain only empty node table. There exists another possible definition
-where we introduce a \texttt{empty} constructor for \texttt{trie} type, however.
-Which one is better is still under question.
+should contain only empty node. There exists another possible definition where
+we introduce a \texttt{empty} constructor for \texttt{trie} type, however. Which
+one is better is still under question.
 
@@
 introduced first. \texttt{strict\_first\_cursor} takes a trie and yields the
-first cursor if there exists one (unless it's an empty table, there always
+first cursor if there exists one (unless it's an empty map, there always
 exists one). \texttt{normalize\_cursor} takes a cursor, which could potentially
@@
 the fact that there is no dead end in the trie (a subtrie which contains no pair
-of keys and values, will be formalized in \texttt{table\_correct} later). If
+of keys and values, will be formalized in \texttt{map\_correct} later). If
 not, these functions can get trapped into dead end and therefore cause a false
@@
 return an empty cursor, which is regarded as \texttt{last\_cursor} for any
-tables.
+maps.
 
@@
 
-\subsubsection{\texttt{table\_correct}}
-
-For a table to be correct, the following properties should hold:
+\subsubsection{\texttt{map\_correct}}
+
+For a trie to be correct, the following properties should hold:
 
 \begin{enumerate}
-\item both the table form and list form should be correct
+\item both the map form and list form should be correct
 \item they should have the same content
@@
   current node,
-\item the current table is correct,
+\item the current trie is correct,
 \item both the cursors are correct,
@@
 
-While the correctness of cursor included the correctness of table for now, it's
+While the correctness of cursor included the correctness of map for now, it's
 actually not necessary. This just simplifies the proof and can be removed in
@@
 The other auxiliary predicates are defined straightforward. \texttt{abs\_rel} is
-defined as the conjunction of table and cursor correctness, as well as the top
+defined as the conjunction of map and cursor correctness, as well as the top
 most level of cursor points to the root of the trie. The rest levels are
@@
 Now suppose we run VST analysis over this function, as shown in the
-Listing~\ref{lst:annotation}. We have already know that the the table is
+Listing~\ref{lst:annotation}. We have already know that the the map is
 correctly represented in the memory, as well as the key. The first call to
 \texttt{Sfirst\_cursor} creates the cursor for the \texttt{put} operation. The
-call to \texttt{put} will provides a new cursor and new table, and a predicate
+call to \texttt{put} will provides a new cursor and new map, and a predicate
 guarantees that they are the result of the functional \texttt{put} operation.
 After that, several calls to make more cursors and tries get the value we
-have just put into the table. At the end of VST analysis, we know that the now
-the new table is correctly represented in the memory, the return value should be
+have just put into the map. At the end of VST analysis, we know that the now
+the new map is correctly represented in the memory, the return value should be
 the value that the functional \texttt{get} returns, and there is no memory leak
@@
 \centering
-\{\texttt{table\_rep t index * key\_rep key k}\} \\
+\{\texttt{map\_rep t index * key\_rep key k}\} \\
 \texttt{SCursor c = Sfirst\_cursor(index);} \\
-\{\texttt{table\_rep t index * key\_rep key k * cursor\_rep (first\_cursor t)
+\{\texttt{map\_rep t index * key\_rep key k * cursor\_rep (first\_cursor t)
 c}\} \\
 \texttt{Sput(k, v, c, index);} \\
-\{$\exists t'\exists c'$, \texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep
+\{$\exists t'\exists c'$, \texttt{!!(put k v (first\_cursor t) t c' t') \&\& map\_rep
 t' index * key\_rep key k * cursor\_rep c' c}\} \\
-\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep t' index * key\_rep key
+\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& map\_rep t' index * key\_rep key
 k * cursor\_rep c' c}\} \\
 \texttt{Sfree\_cursor(c);} \\
-\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep t' index * key\_rep key
+\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& map\_rep t' index * key\_rep key
 k}\} \\
 \texttt{c = Smake\_cursor(index, k);} \\
-\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep t' index * key\_rep key
+\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& map\_rep t' index * key\_rep key
 k * cursor\_rep (make\_cursor k t') c}\} \\
@@
 \texttt{Bool success = Sget(c, index, \&ret);} \\
-\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep t' index * key\_rep key
+\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& map\_rep t' index * key\_rep key
 k * cursor\_rep (make\_cursor k t') c *} \\
@@
 \texttt{Sfree\_cursor(c);} \\
-\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& table\_rep t' index * key\_rep key
+\{\texttt{!!(put k v (first\_cursor t) t c' t') \&\& map\_rep t' index * key\_rep key
 k *} \\
@@
   SEP ()
-  POST [ tptr BTree.tindex ] EX t: @BTree.table val, EX pt: val,
+  POST [ tptr BTree.tindex ] EX t: @BTree.map val, EX pt: val,
   PROP (BTree.empty t)
   LOCAL (temp ret_temp pt)
-  SEP (BTree.table_rep t pt).
+  SEP (BTree.map_rep t pt).
 
@@
   WITH k: Z, v: val,
-       t: @BTree.table val, pt: val,
+       t: @BTree.map val, pt: val,
        c: @BTree.cursor val, pc: val
@@
          temp _cursor pc; temp _index pt)
-  SEP (BTree.table_rep t pt; BTree.cursor_rep c pc)
+  SEP (BTree.map_rep t pt; BTree.cursor_rep c pc)
   POST [ tvoid ]
-  EX new_t: @BTree.table val, EX new_c: @BTree.cursor val,
+  EX new_t: @BTree.map val, EX new_c: @BTree.cursor val,
   PROP (BTree.put k v c t new_c new_t)
   LOCAL ()
-  SEP (BTree.table_rep new_t pt; BTree.cursor_rep new_c pc).
+  SEP (BTree.map_rep new_t pt; BTree.cursor_rep new_c pc).
 
@@
   WITH k: Z,
-       t: @BTree.table val, pt: val
+       t: @BTree.map val, pt: val
   PRE [ _key OF tuint, _index OF tptr BTree.tindex ]
-  PROP (0 <= k <= Int.max_unsigned; BTree.table_correct t)
+  PROP (0 <= k <= Int.max_unsigned; BTree.map_correct t)
   LOCAL (temp _key (Vint (Int.repr k)); temp _index pt)
-  SEP (BTree.table_rep t pt)
+  SEP (BTree.map_rep t pt)
   POST [ tptr BTree.tcursor ] EX pc: val,
@@
   LOCAL (temp ret_temp pc)
-  SEP (BTree.table_rep t pt; BTree.cursor_rep (BTree.make_cursor k t) pc).
+  SEP (BTree.map_rep t pt; BTree.cursor_rep (BTree.make_cursor k t) pc).
 
@@
   DECLARE _Ifirst_cursor
-  WITH t: @BTree.table val, pt: val
+  WITH t: @BTree.map val, pt: val
   PRE [ _index OF tptr BTree.tindex ]
-  PROP (BTree.table_correct t)
+  PROP (BTree.map_correct t)
   LOCAL (temp _index pt)
-  SEP (BTree.table_rep t pt)
+  SEP (BTree.map_rep t pt)
   POST [ tptr BTree.tcursor ] EX pc: val,
@@
   LOCAL (temp ret_temp pc)
-  SEP (BTree.table_rep t pt; BTree.cursor_rep (BTree.first_cursor t) pc).
+  SEP (BTree.map_rep t pt; BTree.cursor_rep (BTree.first_cursor t) pc).
 
@@
   WITH c: @BTree.cursor val, pc: val,
-       t: @BTree.table val, pt: val,
+       t: @BTree.map val, pt: val,
        pk: val
@@
   LOCAL (temp _cursor pc; temp _index pt; temp _key pk)
-  SEP (BTree.table_rep t pt; BTree.cursor_rep c pc; data_at_ Tsh tuint pk)
+  SEP (BTree.map_rep t pt; BTree.cursor_rep c pc; data_at_ Tsh tuint pk)
   POST [ tint ]
@@
   LOCAL (temp ret_temp (if BTree.get_key c t then (Vint Int.one) else (Vint Int.zero)))
-  SEP (BTree.table_rep t pt; BTree.cursor_rep c pc;
+  SEP (BTree.map_rep t pt; BTree.cursor_rep c pc;
        data_at Tsh tuint match BTree.get_key c t with
@@
   WITH c: @BTree.cursor val, pc: val,
-       t: @BTree.table val, pt: val,
+       t: @BTree.map val, pt: val,
        pv: val
@@
   LOCAL (temp _cursor pc; temp _index pt; temp _value pv)
-  SEP (BTree.table_rep t pt; BTree.cursor_rep c pc; data_at_ Tsh (tptr tvoid) pv)
+  SEP (BTree.map_rep t pt; BTree.cursor_rep c pc; data_at_ Tsh (tptr tvoid) pv)
   POST [ tint ]
@@
   LOCAL (temp ret_temp (if BTree.get_value c t then (Vint Int.one) else (Vint Int.zero)))
-  SEP (BTree.table_rep t pt; BTree.cursor_rep c pc;
+  SEP (BTree.map_rep t pt; BTree.cursor_rep c pc;
        data_at Tsh (tptr tvoid) match BTree.get_value c t with
@@
 \begin{minted}[breaklines]{coq}
-Module Trie (Node: FLATTENABLE_TABLE Keyslice)
-            <: FLATTENABLE_TABLE KeyType.
+Module Trie (Node: FLATTENABLE_MAP Keyslice)
+            <: FLATTENABLE_MAP KeyType.
   [...]
@@
     match t with
-    | trienode_of _ tableform listform =>
+    | trienode_of _ mapform listform =>
       match Node.Flattened.get_exact keyslice listform with
@@
         if (Z_le_dec (Zlength k) keyslice_length) then
-          [(t, Node.make_cursor keyslice tableform,
+          [(t, Node.make_cursor keyslice mapform,
             bnode, BorderNode.before_prefix (Zlength k))]
@@
             | value_of _ => []
-            | nil => [(t, Node.make_cursor keyslice tableform,
+            | nil => [(t, Node.make_cursor keyslice mapform,
                        bnode, BorderNode.after_suffix)]
             | trie_of t' =>
-              (t, Node.make_cursor keyslice tableform,
+              (t, Node.make_cursor keyslice mapform,
                bnode, BorderNode.before_suffix) :: make_cursor (get_suffix k) t'
@@
             if (TrieKeyFacts.lt_dec k' (get_suffix k)) then
-              [(t, Node.make_cursor keyslice tableform,
+              [(t, Node.make_cursor keyslice mapform,
                 bnode, BorderNode.after_suffix)]
             else
-              [(t, Node.make_cursor keyslice tableform,
+              [(t, Node.make_cursor keyslice mapform,
                 bnode, BorderNode.before_suffix)]
@@
     match t with
-    | trienode_of _ tableform listform =>
+    | trienode_of _ mapform listform =>
       match Node.Flattened.get_value
@@
           | value_of _ =>
-            Some [(t, Node.first_cursor tableform,
+            Some [(t, Node.first_cursor mapform,
                    bnode, BorderNode.before_prefix len)]
@@
             match strict_first_cursor t' with
-            | Some c' => Some ((t, Node.first_cursor tableform,
+            | Some c' => Some ((t, Node.first_cursor mapform,
                                 bnode, BorderNode.before_suffix) :: c')
@@
           | value_of _ =>
-            Some [(t, Node.first_cursor tableform,
+            Some [(t, Node.first_cursor mapform,
                    bnode, BorderNode.before_suffix)]
@@
     | [] => None
-    | (trienode_of addr tableform listform, table_cursor, bnode, bnode_cursor) :: c' =>
-      let t := trienode_of addr tableform listform in
+    | (trienode_of addr mapform listform, map_cursor, bnode, bnode_cursor) :: c' =>
+      let t := trienode_of addr mapform listform in
       match normalize_cursor c' with
-      | Some c'' => Some ((t, table_cursor, bnode, bnode_cursor) :: c'')
+      | Some c'' => Some ((t, map_cursor, bnode, bnode_cursor) :: c'')
       | None =>
@@
         | BorderNode.before_prefix len =>
-          Some [(t, table_cursor, bnode, (BorderNode.before_prefix len))]
+          Some [(t, map_cursor, bnode, (BorderNode.before_prefix len))]
         | BorderNode.before_suffix =>
@@
           | nil => None
-          | value_of _ => Some [(t, table_cursor, bnode, BorderNode.before_suffix)]
+          | value_of _ => Some [(t, map_cursor, bnode, BorderNode.before_suffix)]
           | trie_of t' =>
@@
             | Some c' =>
-              Some ((t, table_cursor, bnode, BorderNode.before_suffix) :: c')
+              Some ((t, map_cursor, bnode, BorderNode.before_suffix) :: c')
             | None =>
@@
         | BorderNode.after_suffix =>
-          let table_cursor' := Node.next_cursor table_cursor tableform in
-          match Node.get_key table_cursor' tableform with
+          let map_cursor' := Node.next_cursor map_cursor mapform in
+          match Node.get_key map_cursor' mapform with
           | Some key =>
@@
                 | nil => None
-                | value_of _ => Some [(t, table_cursor', bnode', (BorderNode.before_prefix len))]
+                | value_of _ => Some [(t, map_cursor', bnode', (BorderNode.before_prefix len))]
                 | trie_of t' => None
@@
                 | nil => None
-                | value_of _ => Some [(t, table_cursor', bnode', BorderNode.before_suffix)]
+                | value_of _ => Some [(t, map_cursor', bnode', BorderNode.before_suffix)]
                 | trie_of t' =>
@@
                   | Some c' =>
-                    Some ((t, table_cursor', bnode', BorderNode.before_suffix) :: c')
+                    Some ((t, map_cursor', bnode', BorderNode.before_suffix) :: c')
                   | None =>
@@
     | [] => []
-    | [(trienode_of addr tableform listform, table_cursor, bnode, bnode_cursor)] =>
+    | [(trienode_of addr mapform listform, map_cursor, bnode, bnode_cursor)] =>
       match bnode_cursor with
@@
         if Z_lt_dec len keyslice_length then
-          [(trienode_of addr tableform listform, table_cursor, bnode, BorderNode.before_prefix (len + 1))]
+          [(trienode_of addr mapform listform, map_cursor, bnode, BorderNode.before_prefix (len + 1))]
         else
-          [(trienode_of addr tableform listform, table_cursor, bnode, BorderNode.after_suffix)]
-      | BorderNode.before_suffix => [(trienode_of addr tableform listform, table_cursor, bnode, BorderNode.after_suffix)]
+          [(trienode_of addr mapform listform, map_cursor, bnode, BorderNode.after_suffix)]
+      | BorderNode.before_suffix => [(trienode_of addr mapform listform, map_cursor, bnode, BorderNode.after_suffix)]
       | BorderNode.after_suffix =>
-        let table_cursor' := Node.next_cursor table_cursor tableform in
-        match Node.get_key table_cursor' tableform with
+        let map_cursor' := Node.next_cursor map_cursor mapform in
+        match Node.get_key map_cursor' mapform with
         | Some key =>
@@
           | Some bnode' =>
-            [(trienode_of addr tableform listform, table_cursor, bnode, BorderNode.before_prefix 1)]
+            [(trienode_of addr mapform listform, map_cursor, bnode, BorderNode.before_prefix 1)]
           | None => []
@@
 
-  Definition next_cursor (c: cursor) (t: table): cursor :=
+  Definition next_cursor (c: cursor) (t: map): cursor :=
     match normalize_cursor c with
@@
 
-  Definition first_cursor (c: cursor) (t: table): cursor :=
+  Definition first_cursor (c: cursor) (t: map): cursor :=
     match strict_first_cursor c with
@@
     match c with
-    | (trienode_of _ tableform _, table_cursor, bnode, bnode_cursor) :: c' =>
-      match Node.get_key table_cursor tableform with
+    | (trienode_of _ mapform _, map_cursor, bnode, bnode_cursor) :: c' =>
+      match Node.get_key map_cursor mapform with
       | Some keyslice =>
@@
 
-  Definition get (c: cursor) (t: table): option (key * value) :=
+  Definition get (c: cursor) (t: map): option (key * value) :=
     match normalize_cursor c with
@@
 
-  Inductive create_pair (k1 k2: key) (v1 v2: value): cursor -> table -> Prop :=
-    | create_pair_case1: forall emptylist emptytable listform tableform listcursor tablecursor bnode_addr addr,
+  Inductive create_pair (k1 k2: key) (v1 v2: value): cursor -> map -> Prop :=
+    | create_pair_case1: forall emptylist emptymap listform mapform listcursor mapcursor bnode_addr addr,
     let keyslice1 := get_keyslice k1 in
@@
                        listcursor listform ->
-    Node.empty emptytable ->
-    Node.put keyslice1 bnode_addr (Node.first_cursor emptytable) emptytable tablecursor tableform ->
+    Node.empty emptymap ->
+    Node.put keyslice1 bnode_addr (Node.first_cursor emptymap) emptymap mapcursor mapform ->
     create_pair k1 k2 v1 v2
-                [(trienode_of addr tableform listform, tablecursor, bnode, BorderNode.length_to_cursor (Zlength k1))]
-                (trienode_of addr tableform listform)
-    | create_pair_case2: forall emptylist emptytable listform tableform listcursor tablecursor bnode_addr c' t' addr,
+                [(trienode_of addr mapform listform, mapcursor, bnode, BorderNode.length_to_cursor (Zlength k1))]
+                (trienode_of addr mapform listform)
+    | create_pair_case2: forall emptylist emptymap listform mapform listcursor mapcursor bnode_addr c' t' addr,
       let keyslice1 := get_keyslice k1 in
@@
                          listcursor listform ->
-      Node.empty emptytable ->
-      Node.put keyslice1 bnode_addr (Node.first_cursor emptytable) emptytable tablecursor tableform ->
+      Node.empty emptymap ->
+      Node.put keyslice1 bnode_addr (Node.first_cursor emptymap) emptymap mapcursor mapform ->
       create_pair k1 k2 v1 v2
-                  ((trienode_of addr tableform listform, tablecursor, bnode, BorderNode.before_suffix) :: c')
-                  (trienode_of addr tableform listform)
-    | create_pair_case3: forall emptylist emptytable listform1 listform2 tableform1 tableform2 listcursor1 listcursor2 tablecursor1 tablecursor2 bnode_addr1 bnode_addr2 addr,
+                  ((trienode_of addr mapform listform, mapcursor, bnode, BorderNode.before_suffix) :: c')
+                  (trienode_of addr mapform listform)
+    | create_pair_case3: forall emptylist emptymap listform1 listform2 mapform1 mapform2 listcursor1 listcursor2 mapcursor1 mapcursor2 bnode_addr1 bnode_addr2 addr,
       let keyslice1 := get_keyslice k1 in
@@
                          listcursor1 listform1 ->
-      Node.empty emptytable ->
-      Node.put keyslice2 bnode_addr2 (Node.first_cursor emptytable) emptytable tablecursor2 tableform2 ->
-      Node.put keyslice1 bnode_addr1 (Node.first_cursor tableform2) tableform2 tablecursor1 tableform1 ->
+      Node.empty emptymap ->
+      Node.put keyslice2 bnode_addr2 (Node.first_cursor emptymap) emptymap mapcursor2 mapform2 ->
+      Node.put keyslice1 bnode_addr1 (Node.first_cursor mapform2) mapform2 mapcursor1 mapform1 ->
       create_pair k1 k2 v1 v2
-                  [(trienode_of addr tableform1 listform1, tablecursor1, bnode1, BorderNode.length_to_cursor (Zlength k1))]
-                  (trienode_of addr tableform1 listform1).
+                  [(trienode_of addr mapform1 listform1, mapcursor1, bnode1, BorderNode.length_to_cursor (Zlength k1))]
+                  (trienode_of addr mapform1 listform1).
 
     Inductive put (k: key) (v: value): cursor -> trie -> cursor -> trie -> Prop :=
-    | put_case1: forall tableform listform listform' listcursor listcursor' bnode_addr bnode c addr,
+    | put_case1: forall mapform listform listform' listcursor listcursor' bnode_addr bnode c addr,
       let keyslice := get_keyslice k in
@@
                          listcursor' listform' ->
-      let tablecursor := Node.make_cursor keyslice tableform in
-      put k v c (trienode_of addr tableform listform)
-          [(trienode_of addr tableform listform', tablecursor, bnode, BorderNode.before_prefix (Zlength k))]
-          (trienode_of addr tableform listform')
-    | put_case2: forall tableform listform listform' listcursor listcursor' bnode_addr bnode c t' c' t'' c'' addr,
+      let mapcursor := Node.make_cursor keyslice mapform in
+      put k v c (trienode_of addr mapform listform)
+          [(trienode_of addr mapform listform', mapcursor, bnode, BorderNode.before_prefix (Zlength k))]
+          (trienode_of addr mapform listform')
+    | put_case2: forall mapform listform listform' listcursor listcursor' bnode_addr bnode c t' c' t'' c'' addr,
       let keyslice := get_keyslice k in
@@
                          listcursor' listform' ->
-      let tablecursor := Node.make_cursor keyslice tableform in
-      put k v c (trienode_of addr tableform listform)
-          ((trienode_of addr tableform listform', tablecursor, bnode, BorderNode.before_suffix) :: c'')
-          (trienode_of addr tableform listform')
-    | put_case3: forall tableform listform listform' listcursor listcursor' bnode_addr bnode c addr,
+      let mapcursor := Node.make_cursor keyslice mapform in
+      put k v c (trienode_of addr mapform listform)
+          ((trienode_of addr mapform listform', mapcursor, bnode, BorderNode.before_suffix) :: c'')
+          (trienode_of addr mapform listform')
+    | put_case3: forall mapform listform listform' listcursor listcursor' bnode_addr bnode c addr,
       let keyslice := get_keyslice k in
@@
                          listcursor' listform' ->
-      let tablecursor := Node.make_cursor keyslice tableform in
-      put k v c (trienode_of addr tableform listform)
-          [(trienode_of addr tableform listform', tablecursor, bnode, BorderNode.before_suffix)]
-          (trienode_of addr tableform listform')
-    | put_case4: forall tableform listform listform' listcursor listcursor' bnode_addr bnode c addr,
+      let mapcursor := Node.make_cursor keyslice mapform in
+      put k v c (trienode_of addr mapform listform)
+          [(trienode_of addr mapform listform', mapcursor, bnode, BorderNode.before_suffix)]
+          (trienode_of addr mapform listform')
+    | put_case4: forall mapform listform listform' listcursor listcursor' bnode_addr bnode c addr,
       let keyslice := get_keyslice k in
@@
                          listcursor' listform' ->
-      let tablecursor := Node.make_cursor keyslice tableform in
-      put k v c (trienode_of addr tableform listform)
-          [(trienode_of addr tableform listform', tablecursor, bnode, BorderNode.before_suffix)]
-          (trienode_of addr tableform listform')
-    | put_case5: forall tableform listform listform' listcursor listcursor' bnode_addr bnode c c' t' k' v' addr,
+      let mapcursor := Node.make_cursor keyslice mapform in
+      put k v c (trienode_of addr mapform listform)
+          [(trienode_of addr mapform listform', mapcursor, bnode, BorderNode.before_suffix)]
+          (trienode_of addr mapform listform')
+    | put_case5: forall mapform listform listform' listcursor listcursor' bnode_addr bnode c c' t' k' v' addr,
       let keyslice := get_keyslice k in
@@
                          listcursor' listform' ->
-      let tablecursor := Node.make_cursor keyslice tableform in
-      put k v c (trienode_of addr tableform listform)
-          ((trienode_of addr tableform listform', tablecursor, bnode, BorderNode.before_suffix) :: c')
-          (trienode_of addr tableform listform')
-    | put_case6: forall tableform tableform' listform listform' tablecursor tablecursor' listcursor listcursor' bnode_addr c addr,
+      let mapcursor := Node.make_cursor keyslice mapform in
+      put k v c (trienode_of addr mapform listform)
+          ((trienode_of addr mapform listform', mapcursor, bnode, BorderNode.before_suffix) :: c')
+          (trienode_of addr mapform listform')
+    | put_case6: forall mapform mapform' listform listform' mapcursor mapcursor' listcursor listcursor' bnode_addr c addr,
       let keyslice := get_keyslice k in
@@
                          listcursor' listform' ->
-      Node.abs_rel tablecursor tableform ->
+      Node.abs_rel mapcursor mapform ->
       Node.put keyslice bnode_addr
-                         tablecursor tableform
-                         tablecursor' tableform' ->
-      put k v c (trienode_of addr tableform listform)
-          [(trienode_of addr tableform' listform', tablecursor', bnode, BorderNode.before_suffix)]
-          (trienode_of addr tableform' listform').
-
-    Inductive table_correct: table -> Prop :=
-    | table_correct_intro tableform listform: forall addr,
-      Node.table_correct tableform ->
-      Node.Flattened.table_correct listform ->
-      map fst (Node.flatten tableform) = map fst listform ->
-      map snd (Node.flatten tableform) = map (compose fst snd) listform ->
+                         mapcursor mapform
+                         mapcursor' mapform' ->
+      put k v c (trienode_of addr mapform listform)
+          [(trienode_of addr mapform' listform', mapcursor', bnode, BorderNode.before_suffix)]
+          (trienode_of addr mapform' listform').
+
+    Inductive map_correct: map -> Prop :=
+    | map_correct_intro mapform listform: forall addr,
+      Node.map_correct mapform ->
+      Node.Flattened.map_correct listform ->
+      map fst (Node.flatten mapform) = map fst listform ->
+      map snd (Node.flatten mapform) = map (compose fst snd) listform ->
       Forall (compose bordernode_correct (compose snd snd)) listform ->
-      table_correct (trienode_of addr tableform listform)
+      map_correct (trienode_of addr mapform listform)
     with
-    bordernode_correct: @BorderNode.table link -> Prop :=
+    bordernode_correct: @BorderNode.map link -> Prop :=
     | bordernode_correct_intro prefixes (k: option string) l:
@@
     | subtrie_correct_intro t:
-      table_correct t ->
+      map_correct t ->
       (* optional: no dangling(dead end) in the tree, useful for [next_cursor] and [first_cursor] *)
@@
       match c with
-      | (trienode_of addr tableform listform, table_cursor, bnode, bnode_cursor) :: c' =>
+      | (trienode_of addr mapform listform, map_cursor, bnode, bnode_cursor) :: c' =>
         match p with
-        | Some t => t = trienode_of addr tableform listform
+        | Some t => t = trienode_of addr mapform listform
         | None => True
         end /\
-        table_correct (trienode_of addr tableform listform) /\
-        Node.cursor_correct table_cursor /\
+        map_correct (trienode_of addr mapform listform) /\
+        Node.cursor_correct map_cursor /\
         BorderNode.cursor_correct bnode_cursor /\
-        match Node.get_key table_cursor tableform with
+        match Node.get_key map_cursor mapform with
         | Some key =>
@@
 
-    Definition abs_rel (c: cursor) (t: table): Prop :=
-      cursor_correct c /\ table_correct t /\
+    Definition abs_rel (c: cursor) (t: map): Prop :=
+      cursor_correct c /\ map_correct t /\
       match c with
@@
 
-    Definition key_rel (k: key) (c: cursor) (t: table): Prop :=
+    Definition key_rel (k: key) (c: cursor) (t: map): Prop :=
       normalize_cursor c = normalize_cursor (make_cursor k t).
 
-    Definition eq_cursor (c1 c2: cursor) (t: table):
+    Definition eq_cursor (c1 c2: cursor) (t: map):
       normalize_cursor c1 = normalize_cursor c2.

### 2019-08-20 papers/Pablo/report_Pablo.tex
Added my report

\documentclass[11pt]{article}

\usepackage{a4wide}
\usepackage{caption, subcaption, tabu, listings, latexsym, url}
\usepackage[dvipsnames]{xcolor}
\usepackage{stmaryrd}
\usepackage{geometry}
\usepackage{amsmath, amsfonts}
\usepackage[inline]{enumitem}
\usepackage[]{algorithm2e}
\geometry{margin=1.995cm}

% Coq style (inspired from a file of Assia Mahboubi)

\definecolor{ltblue}{rgb}{0,0.4,0.4}
\definecolor{dkblue}{rgb}{0,0.1,0.6}
\definecolor{dkgreen}{rgb}{0,0.35,0}
\definecolor{dkviolet}{rgb}{0.3,0,0.5}
\definecolor{dkred}{rgb}{0.5,0,0}

\DeclareMathOperator{\lookup}{\mathsf{lookup}\;}
\DeclareMathOperator{\Insert}{\mathsf{insert}\;}
\DeclareMathOperator{\some}{\mathsf{Some}\;}

\lstdefinelanguage{Coq}{ 
%
% Anything betweeen $ becomes LaTeX math mode
mathescape=true,
%
% Comments may or not include Latex commands
texcl=false, 
%
% Vernacular commands
morekeywords=[1]{Section, Module, End, Require, Import, Export,
  Variable, Variables, Parameter, Parameters, Axiom, Hypothesis,
  Hypotheses, Notation, Local, Tactic, Reserved, Scope, Open, Close,
  Bind, Delimit, Definition, Let, Ltac, Fixpoint, Function, CoFixpoint, Add,
  Morphism, Relation, Implicit, Arguments, Unset, Contextual,
  Strict, Prenex, Implicits, Inductive, CoInductive, Record,
  Structure, Canonical, Coercion, Context, Class, Global, Instance,
  Program, Infix, Theorem, Lemma, Corollary, Proposition, Fact,
  Remark, Example, Proof, Goal, Save, Qed, Defined, Hint, Resolve,
  Rewrite, View, Search, Show, Print, Printing, All, Eval, Check,
  Projections, inside, outside, Def},
%
% Gallina
morekeywords=[2]{forall, exists, exists2, fun, fix, cofix, struct,
  match, with, end, as, in, return, let, if, is, then, else, for, of,
  nosimpl, when},
%
% Sorts
morekeywords=[3]{Type, Prop, Set, true, false, option},
%
% Various tactics, some are std Coq subsumed by ssr, for the manual purpose
morekeywords=[4]{pose, set, move, case, elim, apply, clear, hnf,
  intro, intros, generalize, rename, pattern, after, destruct,
  induction, using, refine, inversion, injection, rewrite, congr,
  unlock, compute, ring, field, fourier, replace, fold, unfold,
  change, cutrewrite, simpl, have, suff, wlog, suffices, without,
  loss, nat_norm, assert, cut, trivial, revert, bool_congr, nat_congr,
  symmetry, transitivity, auto, split, left, right, autorewrite},
%
% Terminators
morekeywords=[5]{by, done, exact, reflexivity, tauto, romega, omega,
  assumption, solve, contradiction, discriminate},
%
% Control
morekeywords=[6]{do, last, first, try, idtac, repeat},
%
% Comments delimiters, we do turn this off for the manual
morecomment=[s]{(*}{*)},
%
% Spaces are not displayed as a special character
showstringspaces=false,
%
% String delimiters
morestring=[b]",
morestring=[d],
%
% Size of tabulations
tabsize=3,
%
% Enables ASCII chars 128 to 255
extendedchars=false,
%
% Case sensitivity
sensitive=true,
%
% Automatic breaking of long lines
breaklines=false,
%
% Default style fors listings
basicstyle=\small,
%
% Position of captions is bottom
captionpos=b,
%
% flexible columns
basewidth={2em, 0.5em},
columns=[l]flexible,
%
% Style for (listings') identifiers
identifierstyle={\ttfamily\color{black}},
% Style for declaration keywords
keywordstyle=[1]{\ttfamily\bfseries\color{dkviolet}},
% Style for gallina keywords
keywordstyle=[2]{\ttfamily\bfseries\color{dkgreen}},
% Style for sorts keywords
keywordstyle=[3]{\ttfamily\bfseries\color{ltblue}},
% Style for tactics keywords
keywordstyle=[4]{\ttfamily\color{dkblue}},
% Style for terminators keywords
keywordstyle=[5]{\ttfamily\color{dkred}},
%Style for iterators
%keywordstyle=[6]{\ttfamily\color{dkpink}},
% Style for strings
stringstyle=\ttfamily,
% Style for comments
commentstyle={\ttfamily\itshape\color{dkgreen}},
%
%moredelim=**[is][\ttfamily\color{red}]{/&}{&/},
literate=
    {forall}{{\color{dkgreen}{$\forall\;$}}}1
    {exists}{{$\exists\;$}}1
    {<-}{{$\leftarrow\;\;$}}1
    {=>}{{$\Rightarrow\;\;$}}1
    {==}{{\code{==}\;}}1
    {==>}{{\code{==>}\;}}1
%    {:>}{{\code{:>}\;}}1
    {->}{{$\rightarrow\;\;$}}1
    {<-->}{{$\longleftrightarrow\;\;$}}1
    {<->}{{$\leftrightarrow\;\;$}}1
    {<==}{{$\leq\;\;$}}1
    {\#}{{$^\star$}}1 
    {\\o}{{$\circ\;$}}1 
%    {\@}{{$\cdot$}}1 
    {\/\\}{{$\wedge\;$}}1
    {\\\/}{{$\vee\;$}}1
%    {++}{{\code{++}}}1
    {~}{{\ }}1
    {\@\@}{{$@$}}1
    {\\mapsto}{{$\mapsto\;$}}1
    {\\hline}{{\rule{\linewidth}{0.5pt}}}1
%
}[keywords,comments,strings]

\lstnewenvironment{coq}{\lstset{language=Coq}}{}

% pour inliner dans le texte
\def\coqe{\lstinline[language=Coq, basicstyle=\small]}
% pour inliner dans les tableaux / displaymath...
\def\coqes{\lstinline[language=Coq, basicstyle=\scriptsize]}

% END Coq style

\def\ce{\lstinline[language=C, breaklines=true, columns=fixed]}

\newcommand{\wand}{-\kern-.6em\raisebox{-.659ex}{*}\ } % magic wand

%\definecolor{backgroundcolor}{RGB}{240, 245, 255}

%\lstset{backgroundcolor=\color{backgroundcolor}, mathescape, breaklines = true, basicstyle = \ttfamily\small}

\usepackage{tikz}
\usetikzlibrary{shapes, fit, positioning, arrows.meta}

\setlength{\parskip}{0.3\baselineskip}

\begin{document}

\title{\vspace{-1cm}Contributions to a Verified Database Management System\\[.3cm]
\large Internship report - MPRI research master program}

\author{Pablo \textsc{Le Hénaff}\\[.5cm]
Supervisors: Andrew \textsc{Appel}, Lennart \textsc{Beringer}\\[.5cm]
Princeton University, Princeton, NJ 08540, United States of America}

\date{April 1st - August 31st, 2019}

\maketitle

\subsection*{General context}

\emph{Software verification} is the area of computer science research devoted to the study and development of safe programming technology.
Safe programs are free of bugs: they have no unexpected behavior and comply with their specification.
Formal proof systems like \emph{the Coq proof assistant} are computer applications that have been developed to express statements in a precise, logical, and mathematical language,
and check their correctness.

The \emph{Verified Software Toolchain} (VST) is a research project that explores the end-to-end formal verification of programs written in C.
VST includes \emph{Verifiable C}~\cite{PLCC}, a tool to prove the correctness of C code using \emph{separation logic}, embedded inside Coq and developed at Princeton University.

\emph{Relational database management systems} (RDBMS) are programs that are used to store large amounts of data organised in tables.
They are widespread in data-intensive computer applications, such as web servers.
A RDBMS can handle a wide variety of requests from the client, for instance to retrieve filtered or sorted data, or combine multiple sources of information.
Such requests are expressed in a query language, the most well-known of them being SQL (for \emph{Structured Query Language}).
Queries are compiled down in several stages to a combination of efficient low-level algorithms by the RDBMS.

The topic of this report lies at the intersection of RDBMS engineering and state-of-the-art software verification using Coq and Verifiable C.

\subsection*{Research problem}

Despite their omnipresence, there currently exists no usable, efficient RDBMS that has been formally guaranteed to execute safely.

RDBMSs used nowadays in real-world applications are highly complex and tightly engineered systems,
thus proving their functional correctness using current techniques is impossible~\cite{DCWS}.% datacert website

\emph{DeepSpecDB} is an ongoing effort to program some components of a SQL RDBMS in efficient C and explore their VST verification.

\subsection*{My contributions}

I have made contributions to DeepSpecDB on the following aspects:
\begin{itemize}[itemsep=1pt, topsep=0pt]
\renewcommand\labelitemi{--}
\item I reviewed current research work on SQL and RDBMS Coq formalization for use as functional model (section~\ref{dbms});
\item I implemented in C the \emph{index join} algorithm and some other algorithms along the way (section~\ref{implementation});
\item I achieved the VST verification of a hashtable program, mapping integer keys to pointer values, that was used in my C code;
  I instantiated for that purpose a variant of a recently published separation logic technique (section~\ref{hashtable});
\item I formalized in Gallina the invariants and operations of the \emph{B+tree} data structure at the heart of our system, and proved their properties (section~\ref{btree_section}).
\end{itemize}

\subsection*{Arguments supporting their validity}

All the algorithms and data structures at stake are used in commercial, industry-grade systems and have shown many proofs of their efficiency for real-world applications.
The only known equivalent project, DataCert~\cite{DCWS}, has been developing in Coq the upper part of a verified query compilation pipeline.
They plan on compiling that dependently-typed code down to assembly using CertiCoq~\cite{CVCC}, a technology that is not ready yet and whose main focus is not performance.

DeepSpecDB focuses on the last stage of the pipeline: low-level data management.
Writing C code directly and using Verifiable C to give correctness guarantees is a better approach here,
because performance is crucial.
Using the low-level memory management features of C, it is possible to write byte-precise optimized data manipulating algorithms.

\subsection*{Summary and future work}

My contributions have advanced the current state of DeepSpecDB, both of the implementation and verification sides.

On the short term, the next task will be to complete the modular VST verification of the various data structures and algorithms that are already implemented.
More RDBMS algorithms should be added: my code only implements the basic ones, while commercial RDBMSs have a wide panel of algorithms to choose from to optimize a request.

Finally, the low-level algorithms could be interfaced with a program that parses SQL queries and generates query execution plans.
That program could be a commercial RDBMS optimizer paired with a verified plan checker using work from DataCert on SQL semantics, to reach a completely verified system usable for real-world tasks.

\subsection*{Notes and acknowledgments}

This work was funded by \emph{DeepSpec}\footnote{\url{www.deepspec.org}}, an \emph{Expedition in Computing} funded by the \emph{National Science Foundation}\footnote{\url{www.nsf.gov}}.
I also received a stipend from \emph{École polytechnique} as a student of the \emph{Ingénieur polytechnicien} program.

I am thankful to Professor Andrew Appel for inviting me to Princeton University, for his enlightening supervision and for the appreciated financial support.
Thank you Lennart Beringer for your kind help. Thank you Anastasiya Kravchuk-Kirilyuk for being such a cheerful collaborator.
I really enjoyed getting to know the programming languages group and learnt a lot during my time in Princeton.

\newpage

\section{Background and state of the art}

\subsection{Verifiable C}

Verifiable C is a Coq framework and package for proving the functional correctness of C code in an interactive way.
It implements a program logic for C. That logic is a kind of separation logic which allows to reason about imperative programs with pointers one statement at a time.
Full correctness of an executable binary compiled from C can be achieved using VST the following way:
Verifiable C is used to check the correctness of the C code using Coq and all the Verifiable C automation library,
while the \emph{CompCert certified compiler} generates assembly code that is guaranteed to satisfy the specifications of the C language.

That pipeline is worthy of trust since Verifiable C's logic has been mecanically proved sound with respect to the C semantics standard implemented by CompCert.
This proof of soundness can be interpreted as in the following statement:
\begin{center}\emph{Any behavior proven using Verifiable C on a C program\\
is guaranteed to be a behavior of the resulting assembly code generated by CompCert}.\end{center}

Correctness proofs using Verifiable C consist in matching a low-level, imperative C function with a high-level, functional \emph{specification}
taking the form of a Hoare triple $\{ P \} s \{ Q \}$, where $P$ and $Q$ are Verifiable C predicates.
That triple means that when the heap, global and local variables of the program are in a state satisfying predicate $P$,
if statement $s$ is executed by the machine and terminates, the heap and variables will be in a state satisfying $Q$ afterwards.
That specification involves the \emph{functional model}, an implementation in Gallina of C data structures and functions.

Functional languages like Coq are intrinsically easier to reason about than C, because of the immutable nature of their variables: the notion of heap does not exist.
Moreover, the \emph{calculus of inductive constructions} and \emph{dependent types} implemented by Coq are used to
express and prove properties of Coq functions.

Therefore, proving correctness of C code using Verifiable C amounts not only to proving that the C program is a refinement of a functional model expressed in Gallina:
once the high-level specification is proved correct, it is possible to prove further correctness properties on that functional model using Coq,
which by transitivity will also hold on the generated binary.

Any Coq definition or library can be used as a functional model.
However, Verifiable C promotes some standard choices:
for instance, it is strongly advised to use the \coqe{Z} type of integers, even for non-negative numbers (as opposed to Peano natural numbers encoded with the successor function),
as it makes arithmetic reasoning easier and more efficient.

\subsection{Relational DBMS}
\label{dbms}
A database management system executes requests issued by the client to query, add, alter, remove, or reorganise data stored in a database.

DeepSpecDB falls in the category of \emph{relational} DBMS: we will only consider the relational model of data, which is the most commonly used and studied.
Newer DBMS can accomodate data in other forms, such as graph data or semi-structured data like XML documents. These approaches are nowadays referred to by the concept of \emph{NoSQL}.

RDBMS design involves both a
\begin{enumerate*}[label=(\roman*)]\item theoretical, abstract description of the data - the relational data model - that serves to compile queries into an
\item algorithmic, concrete implementation of data management operations.\end{enumerate*}
This section's explanations will be interleaved with short descriptions of recent research work in the formal methods community that I have studied to understand
what formal description of some database system components we could use to specify our DeepSpecDB system.

In particular, \emph{DataCert} is an ongoing research project at the intersection of formal methods, the theory of databases and their functional implementation.
It is being developed in Lille, Lyon, and Orsay in France and is funded by the Agence Nationale de la Recherche.
DataCert plans to interact with the DeepSpec Expedition in Computing by using the CertiCoq verified compiler to generate binary executables~\cite{DCWSRM}.

\subsubsection{Relational model and algebra}
\label{relmodel}
In the relational model, data is organized into \emph{relations}, that are \emph{multisets} of \emph{tuples} sharing the same \emph{attributes}.
A multiset, or \emph{bag}, is a set that can contain multiple occurences of a same element. The number of occurences of a given element is called its \emph{multiplicity}.
In a relation, the ordered list of attributes common to all tuples is called the \emph{schema} in the database jargon.
Each attribute has an associated \emph{domain} with corresponds to the Coq notion of \coqe{Type}.
Intuitively, tuples are rows in a table, the attributes (or fields) are columns of the table and the schema is the header, or specification of the columns.
Each element in a tuple must belong to the domain of the corresponding attribute in the schema.

Malecha et al.~\cite{TVRDMS} use a Coq inductive definition for tuples, reproduced in figure~\ref{TVRDMS_tuple}, based on a list of Coq types called \coqe{Schema}.
Also included is a query in their SQL-like language, illustrating how fields are selected using their position in the schema.

\begin{figure}
\begin{minipage}{.45\textwidth}
  \begin{coq}
    Fixpoint Tuple (A: Schema): Set :=
    match A with
    | nil => unit
    | n::t => $\llbracket$n$\rrbracket$ * Tuple t
    end.
  \end{coq}
\end{minipage}\hfill
\begin{minipage}{.50\textwidth}
\begin{lstlisting}[basicstyle=\ttfamily\small]
SELECT 0, 1, 2 FROM
  (JOIN tbl1 , tbl2 ON col 0 = col 0)
WHERE col 0 = "hello world"
      AND col 1 < col 3
\end{lstlisting}
\end{minipage}
  \caption{A first tuple formal model and a sample SQL-like query from~\cite{TVRDMS}}
  \label{TVRDMS_tuple}
\end{figure}

Benzaken et al.~\cite{CFRDM} define another version of a tuple type. It is reproduced in figure~\ref{datacert_tuples}.
A tuple record \coqe{Tuple.Rcd} holds Coq types for relational data types, attributes, values and abstract tuples.
Thus each attribute is identified by a distinct name in the \coqe{attribute} type.
The \coqe{support} of a tuple is the finite set of attributes it covers. The \coqe{dot} function is projection on a single attribute.
The \coqe{mk_tuple} parameter abstractly builds a tuple from its support and dot/projection function.

While it looks more complex than the previous tuples, this formalization of relational tuples is far more realistic.
Indeed, the main flaw of the previous formalization is that attributes are denoted by their position in the tuple's schema, not by a given identifier:
this is called the \emph{unnamed} relational model.
DataCert implements a \emph{named} version of the relational model. This version is also implemented in all commercial RDBMSs.
While the named and unnamed relational models are theoretically equivalent~\cite{CFRDM},
the unnamed version does not allow attributes to carry semantic information nor does it allow RDBMSs to choose optimized algorithms and indexing data structures based on a given attribute.

\begin{figure}
  \begin{coq}
    Module Tuple.
    Record Rcd : Type := mk_R {
      [...]
      A : Fset.Rcd attribute;
      tuple : Type;
      support : tuple -> set A;
      dot : tuple -> attribute -> value;
      mk_tuple : set A -> (attribute -> value) -> tuple;
      [...] }
  \end{coq}
  \caption{The more realistic DataCert tuples, from~\cite{CFRDM}}
  \label{datacert_tuples}
\end{figure}
Relations are defined in both cases as bags of tuples respectful of the relation's schema.

The \emph{relational algebra} is an abstract algebraic structure that describes selection of relational data. The most commonly used relational algebra operators and their properties are thoroughly described in Garcia-Molina et al.~\cite{DSCB} and summed up in figure~\ref{relational_operators}, where $R$ and $R'$ are relations. The grouping and renaming operators ($\gamma$ and $\rho$ respectively), as well as exotic variants of join, are omitted for concision.
The join operator we consider is a restricted version of the \emph{theta-join}, that is, it accepts an equality predicate to identify attributes with potentially different name.
This type of join is called the \emph{equijoin}, and the attributes that are on both sides of the equality in the predicate are called \emph{foreign keys}.

\begin{figure}
\centering
\begin{tabu} to 0.9\textwidth {|c|X[cm]|}
  \hline
  \textbf{Operator} & \textbf{Meaning}\\
  \hline
  $\sigma_p(R)$ & Selects tuples in $R$ that satisfy predicate $p$\\
  \hline
  $\pi_{A_1,\dots,A_n}$ & Projects tuples in $R$ on attributes $A_1, \dots, A_n$\\
  \hline
  $R \bowtie_{\bigwedge_{i = 1}^n R.A_i = R'.B_i} R'$ & \strut Joins $R$ and $R'$ by identifying for $i \in \{1, \dots, n\}$
  field $A_i$ from $R$ with field $B_i$ from $R'$ \strut\\
  \hline
  $\cup$, $\cap$, $\backslash$ & Usual set operations with bag semantics\\
  \hline 
\end{tabu}
\caption{Relational algebra operators}
\label{relational_operators}
\end{figure}

After being parsed into an abstract syntax tree, a SQL request is interpreted by the RDBMS as a term of the relational algebra.
DataCert is the most advanced project providing detailed and complete mechanized semantics for the relational algebra~\cite{CFRDM}
as well as for SQL~\cite{CMFSRSQ}~\cite{CFSEE}.

To illustrate relational algebra operators and their meaning, figure~\ref{ex_relalg} shows a typical example
of two tables sharing some information (fields \textbf{N} in A and \textbf{n} in B) that are being joined.
The result is filtered and projected to execute a SQL query.
One can notice that the equijoin is a somewhat silly operator, as it produces a new relation where multiple columns (\textbf{n} and \textbf{k} in the example) contain the exact same information under different attribute names.
This behavior is the one implemented in commercial RDBMSs.

The \emph{natural join} will join those attributes from both relations that have exactly the same name, and merge the resulting identical columns.
My implementation of the index join, described in section~\ref{implementation}, is made smarter than the traditional equijoin by merging identical columns like the natural join.
The attribute name of the new merged column is the one of the left relation.
It is unclear whether this behavior is a good thing or not; anyway, switching to the silly version is only a matter of deleting a few lines of code.

\begin{figure}
  \def\arraystretch{1.2}
  \begin{subfigure}{0.2\textwidth}
    \centering
      \begin{tabular}{|c|c|}
        \hline
        \multicolumn{2}{|c|}{a}\\
        \hline
        \textbf{n} & \textbf{lit} \\
        \hline
        0 & zero\\
        \hline
        1 & one\\
        \hline
        2 & two\\
        \hline
        3 & three\\
        \hline
      \end{tabular}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.2\textwidth}
    \centering
      \begin{tabular}{|c|c|}
        \hline
        \multicolumn{2}{|c|}{b}\\
        \hline
        \textbf{k} & \textbf{even} \\
        \hline
        0 & \textsf{T}\\
        \hline
        1 & \textsf{F}\\
        \hline
        2 & \textsf{T}\\
        \hline
        3 & \textsf{F}\\
        \hline
        4 & \textsf{T}\\
        \hline
      \end{tabular}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.4\textwidth}
    \centering
      \begin{tabular}{|c|c|c|c|}
        \hline
        \multicolumn{4}{|c|}{$\mathrm{a} \bowtie_{\mathrm{a.n} = \mathrm{b.k}} \mathrm{b}$}\\
        \hline
        \textbf{n} & \textbf{lit} & \textbf{k} & \textbf{even}\\
        \hline
        0 & zero & 0 & \textsf{T}\\
        \hline
        1 & one & 1 & \textsf{F}\\
        \hline
        2 & two & 2 & \textsf{T}\\
        \hline
        3 & three & 3 & \textsf{F}\\
        \hline
      \end{tabular}
  \end{subfigure}\\[.5cm]
    \hfill
  \begin{subfigure}{0.45\textwidth}
    \centering
      \begin{tabular}{|c|c|c|c|}
        \hline
        \multicolumn{4}{|c|}{$\sigma_{\mathrm{even} = \mathsf{T}}(\mathrm{a} \bowtie_{\mathrm{a.n} = \mathrm{b.k}} \mathrm{b})$}\\
        \hline
        \textbf{n} & \textbf{lit} & \textbf{k} & \textbf{even} \\
        \hline
        0 & zero & 0 & \textsf{T}\\
        \hline
        2 & two & 2 & \textsf{T}\\
        \hline
      \end{tabular}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.45\textwidth}
    \centering
      \begin{tabular}{|c|c|}
        \hline
        \multicolumn{2}{|c|}{$\pi_{\mathrm{n}, \mathrm{lit}}(\sigma_{\mathrm{even} = \mathsf{T}}(\mathrm{a} \bowtie_{\mathrm{a.n} = \mathrm{b.k}} \mathrm{b}))$}\\
        \hline
        \textbf{n} & \textbf{lit} \\[.1cm]
        \hline
        0 & zero\\
        \hline
        2 & two\\
        \hline
      \end{tabular}
  \end{subfigure}\\[.5cm]
  \begin{subfigure}{\textwidth}\centering
    \lstinline[language=SQL, mathescape]{SELECT n, lit FROM a JOIN b ON a.n = b.k WHERE even = $\mathsf{T}$}
  \end{subfigure}
  \caption{Illustration of the relational algebra}
  \label{ex_relalg}
\end{figure}

\begin{figure}
  \begin{minipage}{.4\textwidth}
  \begin{lstlisting}[basicstyle=\footnotesize, caption={Output of EXPLAIN}]
Hash Join
  Hash Cond: (b.k = a.n)
  ->  Seq Scan on b
        Filter: (even IS TRUE)
  ->  Hash
        ->  Seq Scan on a
  \end{lstlisting}
  \end{minipage}
  \begin{minipage}{.5\textwidth}
    \begin{tikzpicture}[sibling distance=15em,
  every node/.style = {shape=rectangle,
    draw, align=center}]]
  \node {Hash join\\ b.k = a.n}
    child { node {Sequential scan on b\\ filter output on: even = \textsf{T}} }
    child { node (a) {Hash\\ on attribute n}
      child { node (b) {Sequential scan on a} } };

    \node[fit=(a) (b), rectangle, draw=red, label=below:{\color{red} Index scan on a}] {};
\end{tikzpicture}

  \end{minipage}
  \caption{The query execution plan associated to figure~\ref{ex_relalg}}
  \label{postgres_QEP}
\end{figure}

\subsubsection{Query execution}
\label{query_execution}
Given a relational algebra term, the DBMS elaborates a concrete \emph{query execution plan} (QEP).
A QEP takes the form of a binary tree whose nodes are \emph{physical, low-level algorithms} that effectively treat and output the data in memory.
A same relational algebra term is implemented by several QEPs, with different expected running times.
These running times vary in particular with the size of tables: a small relation will be more efficiently treated as is,
while a large relation may benefit from building dedicated indexing structures.
The RDBMS tries to choose the fastest QEP to execute a query.
A detailed exposition of PostgreSQL QEPs can be found in~\cite{postgresql_manual}, section 14.1: \emph{Using EXPLAIN}.
The QEP generated by PostgreSQL for the query presented in figure~\ref{ex_relalg} is detailed in figure~\ref{postgres_QEP}.

Notice how the join comes last in the QEP while it came first in the relational algebra term.
Indeed, multiple abstract terms (respectively QEPs) are available to describe (resp. execute) a same query.
The rewrite rules that transform relational algebra terms while preserving their interpretation are called \emph{algebraic equivalences}.
They have been proven correct by DataCert and are briefly described in Benzaken et al.~\cite{CFRDM}.

The structure on which a QEP is built is known as the \emph{physical algebra}.
Low-level algorithms are operators of that algebra, while terms are either \emph{iterators} or \emph{indices}.

\paragraph{Iterators.}

A tuple iterator is an abstract list-like data container on which it is possible to perform the following operations:
\begin{itemize}[topsep=0pt, noitemsep]
\item the \emph{init} operation allocates whatever data structure in memory is necessary for iterating data: it is a required preliminary operation;
\item the \emph{next} operation produces a tuple (or tuple address) and alters the internal data structures of the iterator to reflect it;
\item the \emph{close} operation terminates the iterator's operation, by e.g. deallocating memory.
\end{itemize}

The QEP ultimately generates an iterator over all the tuples that were selected by the original request.
The abstract representation of an iterator (called \coqe{Cursor}...) from~\cite{CFSEE} is partially reproduced in figure~\ref{iterator_model}.

\subparagraph{Cursors.} The concept of iterator is inseparable from the notion of \emph{cursor}. A cursor serves to remember previous operation of the iterator, and points to whatever corresponds to the current state of the iteration. Most iterators will use an adequate cursor internally. We have defined cursors for various data structures, namely association lists, hashtables and B+trees (see section~\ref{btree_section}).

\begin{figure}
  \begin{minipage}{0.45\textwidth}
  \begin{coq}
Record Cursor (elt : Type) : Type := {
  cursor: Type;
  next: cursor -> result elt * cursor;
  has_next: cursor -> Prop;
  \end{coq}\end{minipage}\hfill
  \begin{minipage}{0.45\textwidth}
  \begin{coq}
  reset: cursor -> cursor;
  collection: cursor -> list elt;
  visited: cursor -> list elt;
  coherent: cursor -> Prop; [...] }
  \end{coq}\end{minipage}
  \caption{DataCert iterators, without axioms}
  \label{iterator_model}
\end{figure}

\paragraph{Indices.}

Database indices are efficient implementations of the \emph{finite map} interface, that can quickly tell the DBMS where a tuple is situated in memory.

Finite maps represent finite functional relations between keys and values, that is, relations that can be represented by a (partial) function,
where keys map to a unique value.
Let $\mathcal{M}(K, V)$ be the abstract type of finite maps storing keys of type $K$ and values of type $V$.
The minimal interface has two operations $\lookup\!: K \rightarrow \mathcal{M}(K,V) \rightarrow \mathsf{option}\ V$
and $\Insert\!: K \rightarrow V \rightarrow \mathcal{M}(K,V) \rightarrow \mathcal{M}(K,V)$, where $\mathsf{option}$ is the usual option functor.
It also has an \emph{empty} map $m_\emptyset: \mathcal{M}(K,V)$.
These operations must satisfy the following axioms:
\begin{equation} \begin{array}{rcl}
  \forall k, \quad \lookup k\ m_\emptyset & = & \mathsf{None} \\[.2cm]
\forall k\ \forall k'\ \forall m, \quad
\lookup k'\ (\Insert k\ v\ m) & = & \left\{
\begin{array}{cl}
  \some v & \mathrm{if}\quad k = k'\\
  \lookup k'\ m  & \mathrm{otherwise.}
\end{array} \right. 
\end{array} \label{map_axioms}\end{equation}

\textsf{lookup} can be used to define an implementation-agnostic equivalence relation $\equiv$ on maps, stating that two maps hold exactly the same mappings:
\[ \forall m\ \forall m',\quad m \equiv m' \qquad \overset{\mathrm{def}}{\Leftrightarrow} \qquad \forall k,\quad \lookup k\ m = \lookup k\ m' \]
which then yields the properties:
\[\begin{array}{rrcl}
& \Insert k\ v'\ (\Insert k\ v\ m) & \equiv & \Insert k\ v'\ m \\
\text{and if } k \neq k', & \Insert k'\ v'\ (\Insert k\ v\ m) & \equiv & \Insert k\ v\ (\Insert k'\ v'\ m)\ .
\end{array}\]

\subparagraph{Concatenated indices.}
Using C pointers, it is possible to accomodate any type for values by refering to their addresses in memory.
However, known concrete implementations of indices usually handle atomic key types only, like integers (B+trees) or strings (tries).
But what if we want to index data on a product type $K_1 \times K_2$?

The solution consists in implementing $\mathcal{M}(K_1 \times K_2, V)$ as $\mathcal{M}(K_1, \mathcal{M}(K_2, V))$.
Given $(k_1, k_2): K_1 \times K_2$ and $m: \mathcal{M}(K_1, \mathcal{M}(K_2, V)$, we implement the interface for $\mathcal{M}(K_1 \times K_2, V)$ by posing
\begin{align*}
\tilde{m}_\emptyset & := m_\emptyset \\
\mathsf{lookup\_or\_empty}\ k\ m & := \mathrm{if}\ \lookup k\ m = \mathsf{Some}\ v \ \mathrm{then}\ v \ \mathrm{else}\ m_\emptyset\\
\widetilde{\lookup} (k_1, k_2)\ m & := \lookup k_2\ (\mathsf{lookup\_or\_empty}\ k_1\ m)\\
\widetilde{\Insert} (k_1, k_2)\ v\ m & := \Insert k_1\ (\Insert k_2\ v\ (\mathsf{lookup\_or\_empty}\ k_1\ m)\ m\ .
\end{align*}

The proof that these definitions verify the axioms~\eqref{map_axioms} is easy and, due to space constraints, left to the reader.
\emph{Concatenated indices} refer to such maps where the key type is composite;
they are used to index tuples based on their projection on a list of attributes called a \emph{subschema}.

\subparagraph{Primary and secondary indices.}
\emph{Entity integrity} is a database concept meaning that all the tuples in a database must be
uniquely identifiable and capable of being located~\cite{DGS}.
To maintain it, RDBMSs require relations to have a \emph{primary key},
i.e. a subschema of the relation on which each tuple's projection is unique.
While it is possible to define as primary key the whole schema if no two tuples are identical,
that would be less performant than finding an adequate and minimal primary key.

RDBMSs systematically build an index on the primary key of a relation: it is called the \emph{primary index} and defines the relation's content.
That primary index is usually a B+tree when the primary key is composed of a single integer field, because they allow ordered iteration (see section~\ref{btree_section}).

RDBMSs can also build \emph{secondary indices} indexing tuples on other attributes.
Since tuples do not necessarily have unique projections on these attributes,
a secondary index implements a finite map whose values are lists or iterators,
or equivalently a kind of finite map interface with duplicates.

\paragraph{Iterator- and index-generating algorithms}
Here is a description of the database algorithms that I implemented.
\begin{itemize}[itemsep=1pt, topsep=0pt]
\item The \emph{sequential scan} outputs an iterator over \emph{all} the tuples stored in a relation $R$.
It relies on the primary index of $R$ and doesn't take any other input, thus lies at a leaf of the QEP.
\item The \emph{index scan} indexes tuples in a relation based on their projection on a given sub-schema.
Its pseudo-code is in algorithm~\ref{index_scan_algo}.
The $\Insert^\text{list}$ function implements the insertion of an element $v: V$ at key $k: K$ in a ``finite map with duplicates'' of type $\mathcal{M}(K,\ \mathsf{list}\ V)$:
if no mapping is found at key $k$, a singleton list $[v]$ is inserted. Otherwise, the current list $l_k$ of values corresponding to key $k$ is retrieved, $v$ added to it, and the new list $v :: l_k$ is inserted back.

\begin{algorithm}
 \KwData{A relation $R$ and a subschema $(A_i)_{i=1}^n$ of $R$.}
 \KwResult{A secondary index based on that subschema.}
 $m: \mathcal{M}(D_1 \times \dots \times D_n,\ \mathsf{list}\ P) := m_\emptyset$\\
 \qquad where $\forall i$, $D_i$ is the domain of $A_i$ and $P$ is the type of heap pointers\;
 $\mathsf{it}$ := the sequential scan iterator on $R$\;
 $\mathsf{it.init()}$\;
 \While{$\mathsf{it}$ is not empty}{
  (tuple) $t$ := $\mathsf{it.next()}$\;
  $t_\text{proj} := \pi_{(A_i)_i}(t)$\;
  $m := \Insert^\text{list} t_\text{proj}\ \&t\ m$;
 } $\mathsf{it.close()}$\;
 \KwRet{$m$}
 \caption{The index scan algorithm}\label{index_scan_algo}
\end{algorithm}

\item The \emph{index join} is a fast implementation of the $\bowtie$ abstract operator.
  When computing $R \bowtie R'$ and by analogy with the nested-loop algorithm, $R$ will be called the \emph{outer relation}
  and $R'$ the \emph{inner relation}.
  The index join uses a secondary index to lookup all tuples from the inner relation that can be joined with a given tuple from the outer relation.
  Its interface is in algorithm~\ref{index_join_algo} and its implementation in C is partially detailed in section~\ref{implementation}.
  The interested reader can find detailed complexity analysis as well as descriptions of other efficient algorithms for joining relations in Mishra and Eich~\cite{JPRD}.
  \begin{algorithm}
    \KwData{An iterator $\mathsf{it}$ on relation $R^\text{out}$ with schema $(A_i)_i$,\\
      the schema $(B_j)_j$ of $R^\text{in}$,\\
      a list of pairs of foreign keys $(A_{i_k}, B_{j_k})_{k = 1}^m$,\\
      a secondary index $m: \mathcal{M}(D_1 \times \cdots \times D_m,\ \mathsf{list}\ P)$ on the subschema $(B_{j_k})_k$ of $R^\text{in}$}
    \KwResult{An iterator on $R^\text{out} \bowtie_{\bigwedge_k (A_{i_k} = B_{j_k})} R^\text{in}$}
    \caption{The classic index (equi)join algorithm: interface}
    \label{index_join_algo}
  \end{algorithm}
\end{itemize}
\section{Implementation of some low-level algorithms in C}
\label{implementation}
When I started my internship, DeepSpecDB already had some libraries implementing the main data structures used to index tuples' addresses in memory:
B+trees~\cite{Tosin}~\cite{Aurele}, hashtables, association lists.
I developed database code that would use these libraries and implement the sequential scan, index scan and index join algorithms.

\subsection{Object-oriented programming in C}

Assume that in a query execution plan, an algorithm $A$ takes input from iterators $\textit{it}_1$ and $\textit{it}_2$.
The implementation of $A$ only needs to have access to the \emph{init}, \emph{next} and \emph{close} operations of $\textit{it}_1$ and $\textit{it}_2$.
Moreover, we don't want the implementation of $A$ to depend on what algorithm yields $\textit{it}_1$ or $\textit{it}_2$.
This modularity is achieved through an \emph{object-oriented} approach.

Object-oriented (OO) programming in C needs explicit method closures, as the language doesn't have built-in OO patterns.
A C object can be implemented as a structure with a pointer to a \emph{method table}, that stores function pointers to methods,
and a pointer to an \emph{environment} that stores private data.
A \emph{class interface} or \emph{abstract class} or \emph{signature} specifies the names of the methods and the types they must have.
An object \emph{class} describes a set of objects that have common methods with the same implementation.
A class can implement an interface, in which case the interface's methods must be a subset of the class's methods.
In our case, the abstract class or interface is ``the iterators'' and the classes are the physical iterator-generating algorithms.

The C type of environments can vary between different classes implementing the same interface,
in order to account for different method implementation.
That feature is achieved through pointer casts. The generic environment pointer has type \ce{void*}.

For example, instances of the sequential scan iterator class have an environment composed of a B+tree primary index pointing to a relation, and a cursor on that B+tree.
The cursor is initialized to the first position by \emph{init}, and moved to the next record by \emph{next}.
However, the methods and environment of the index join are completely different (see next section).

\noindent\begin{minipage}{.45\textwidth}
  \begin{lstlisting}[language=C, basicstyle=\small]
    struct methods {
      void (*init) (void* env);
      const void* (*next) (void* env);
      void (*close) (void* env);
    };
\end{lstlisting}\end{minipage}\hfill
\begin{minipage}{.45\textwidth}
  \begin{lstlisting}[language=C, basicstyle=\small]
    typedef struct iterator_t {
      struct methods *mtable;
      void* env;
    } *iterator;
  \end{lstlisting}
\end{minipage}\\

It is particularily unsafe to call methods directly, as there is no restriction on what the environment pointer actually points to.
I therefore also defined convenient wrappers around an iterator's methods, like this \ce{get_next} function, which guarantees that
an iterator's \ce{next} method is always called with the iterator's environment:
\begin{lstlisting}[language=C, basicstyle=\small]
  const void* get_next(iterator it) { return it->mtable->next(it->env); };
\end{lstlisting}

\subsection{The index join: implementation}\label{actual_implementation}
\def\inlinec{\lstinline[language=C, basicstyle=\normalsize, columns=fixed]}

This section details the implementation of the \ce{next} method of the index join iterator.
That code implements the relational algebraic equijoin, modified to delete duplicate columns, as explained in section~\ref{relmodel}.

{\lstset{language=C, basicstyle=\footnotesize, numbers=left, mathescape}
\begin{lstlisting}
const void* index_join_next(void* env) {
  struct index_join_env* e = (struct index_join_env*) env;$\label{env_cast}$
\end{lstlisting}
\lstset{firstnumber=last}

The defined function has the expected prototype for a \inlinec{next} method.
Line~\ref{env_cast} casts the generic environment pointer into the index join's own environment pointer type.
The index join environment holds essential data like the schema of both relations, the outer iterator,
and the index on the inner relation.

\begin{lstlisting}
  while((e->current_inner == NULL || fifo_empty(e->current_inner))
        && (e->current_outer = get_next(e->outer)) ) {
    Key proj = get_projection(e->outer_attrs, e->current_outer, e->outer_join_attrs);
    e->current_inner = (fifo*) index_lookup(e->ind_on_inner_sch, e->ind_on_inner, proj);
  };
\end{lstlisting}

If the inner relation has no more tuple that matches the projection on join attributes of the current outer tuple,
we query a new outer tuple from the outer iterator using \inlinec{get_next}.
In that case, we also retrieve the list (\inlinec{fifo}) of inner tuples from the inner index
that can be joined with the current outer tuple.

\begin{lstlisting}
  if(!e->current_outer) return NULL;$\label{after_if}$
\end{lstlisting}

That \lstinline[columns = fixed]{if} statement is triggered when the outer tuple iterator is now empty: the index join iterator is now also empty,
thus the \emph{next} method returns the null pointer.

Past line~\ref{after_if}, we now have an outer tuple \inlinec{e->current_outer} that has to be joined with \emph{all} the inner tuples
that are stored in the inner tuple list \inlinec{e->current_inner}.

\begin{lstlisting}[language=C]
  size_t outer_t_length = attribute_list_length(e->outer_attrs);$\label{all1}$
  size_t inner_t_length = attribute_list_length(e->inner_attrs);$\label{all2}$
  size_t join_length = attribute_list_length(e->inner_join_attrs);$\label{all3}$
\end{lstlisting}

Lines~\ref{all1}, \ref{all2} and \ref{all3} calculate the respective number of attributes in
the outer tuples,
the inner tuples, and
the list of inner attributes that are going to be identified with the attributes from \inlinec{outer_join_attrs} by the join algorithm.
In the actual implementation, these numbers are calculated only once by \emph{init}.

Crucial to understanding the rest of the code are the following statements:
\begin{enumerate}[noitemsep, topsep=0pt, label=(\alph*)]
\item a single tuple is stored contiguously in memory, and
\item independently of its domain, a field in a tuple occupies exactly \inlinec{sizeof(void*)} bytes of memory, which is 8 on a 64-bits machine.
This does \emph{not} depend on whether the domain of that field is integer or string.
Integer entries in our system are \inlinec{unsigned long} and thus occupy 8 bytes and a string field is a pointer to the string characters, which is also 8 bytes.
Note that unfortunately, these facts are not specified in the C standard, thus the code may not be portable.
\end{enumerate}

\begin{lstlisting}
  void* new_t = malloc(sizeof(void*)*(outer_t_length + inner_t_length - join_length + 1));
  memcpy(new_t, e->current_outer, sizeof(void*) * outer_t_length);
\end{lstlisting}

We allocate the memory space for the new joined tuple, calculating its size (in multiples of 8 bytes) using the length of the various attribute lists.
Data from the outer tuple is entirely copied at the beginning of that new location.

\begin{lstlisting}
  const void* inner_t = fifo_get(e->current_inner)->data;
  attribute_list l = e->inner_attrs;
\end{lstlisting}

We retrieve the next inner tuple in the list and create an alias for the list of the inner schema.

\begin{lstlisting}
  size_t n = 0;
  for(; l != NULL; l=l->next) {
\end{lstlisting}

This loop will copy in the new joined tuple all fields in \inlinec{inner_t} that are not already present.

\begin{lstlisting}
    size_t inner_t_ofs = get_offset(e->inner_attrs, l->id, l->domain);
\end{lstlisting}

\inlinec{get_offset(attrs, attr_name, attr_domain)} returns the integer index of \inlinec{attr_name} in the attribute list \inlinec{attrs}, or a negative number if \inlinec{attr_name} is not found.
The \inlinec{attr_domain} parameter allows to fail if \inlinec{attr_name} exists in attrs, but with a different domain.
An optimized version of the code would calculate the offsets only once at initialisation.

\begin{lstlisting}
    if(get_offset(e->inner_join_attrs, l->id, l->domain) < 0) {
      memcpy((void*) (((size_t*) new_t) + outer_t_length + n),
          (void*) ((size_t*) inner_t + inner_t_ofs), sizeof(size_t));
      n++;
    };
  }
  return new_t;
};
\end{lstlisting}
}

Finally, the index join is packaged into the iterator structure with its other methods whose details have been omitted here:

\begin{lstlisting}[language=C, basicstyle=\footnotesize, mathescape]
struct methods index_join_iterator_mtable =
     { &index_join_init, &index_join_next, &index_join_close };

iterator index_join($[...]$) {
  $[...]$
  struct index_join_env *env = malloc(sizeof(struct index_join_env)); if(!env) exit(1);
  $[... \text{initialize the environment structure} ...]$
  iterator it = malloc(sizeof(struct iterator_t)); if(!it) exit(1);
  it->mtable = &index_join_iterator_mtable;
  it->env = (void*) env;
  return it;
}
\end{lstlisting}

\subsection{Verification: perspective}

Verification of object-oriented C code using Verifiable C is described by Appel et al.~\cite{PLCC}, chapter 29: \emph{Dependently typed C programs}.
The method that is presented will involve defining a generic Coq record type for the \inlinec{iterator} abstract class that contains various definitions about the separation logic representation of iterators
and proof of statements about these definitions.
That Coq record defines what should be common to all implementations of iterators with respect to Verifiable C logic.

I have attempted at verifying that code, however I realized that it was too early to do so since the building blocks
like indices were not done yet.
Also, given that specifications are usually subtly modified as the proof is carried out,
I will not reproduce here my attempts as they have not yet be proven to be fully correct.
No Verifiable C verification has ever been attempted on such a large object-oriented code base.

\section{A verified hashtable program}
\label{hashtable}
I achieved the verification of an implementation of \emph{hashtables with external chaining}, that was used in our implementation of concatenated indices. The Verifiable C tutorial~\cite{VCT} guides students through the verification of a hashtable implementing a string multiset algebraic datatype by mapping string keys to their multiplicity in the multiset. The separation logic proof technique that is used is called \emph{Magic wand as frame} and is the object of a recent paper~\cite{MWAF} co-authored by my supervisor.

I will explain in this section how the Verifiable C verification of DeepSpecDB's integer hashtables led me to instantiate the magic-wand-as-frame technique in an interesting way.

\subsection{Purpose and C code}

As explained in section~\ref{query_execution}, we need to be able to build various kinds of indices for the join algorithm. Each of our attribute domains - strings or integers for now - has its own implementation of a key-pointer store: string association lists and integer hashtables, respectively. A concatenated index is implemented as an index pointing to other indices.

The hashtable data structure consist of an array of pointers to linked lists, called buckets. The bucket at array position $i$ contains any entry whose integer key $k$ verifies $k\ \mathrm{mod}\ N = i$, where $N$ stands for the total number of buckets in the array. \emph{External chaining} means that the hashtable handles collisions of hashes of different keys by growing the linked list buckets.
We use a large and prime $N$, namely 65599, to distribute entries evenly on average and guarantee fast lookups.

The insertion operation has a main procedure \ce{inthash_insert} and an auxiliary function \break\ce{inthash_insert_list} that are both reproduced in figure~\ref{inthash_insert}.

\begin{figure}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting}[language=C,basicstyle=\small]
struct icell {
  size_t key;
  void *value;
  struct icell *next;
};
\end{lstlisting}
\end{minipage}\hfill
\begin{minipage}{0.5\textwidth}
\begin{lstlisting}[language=C, basicstyle=\small]
struct inthash {
  struct icell *buckets[N];
};
\end{lstlisting}
\end{minipage}
\begin{lstlisting}[language=C,basicstyle=\small, numbers=left, escapechar=|]
struct icell *inthash_insert_list (struct icell **r0, size_t key) {
  struct icell *p, **r;
  for(r=r0; ; r=&p->next) { |\label{begin_loop}|
    p = *r;
    if (!p) {
      p = new_icell(key,NULL,NULL);
      *r = p;
      return p;
    }
    if (p->key==key) {
      return p;
    }
  } |\label{end_loop}|
}

void* inthash_insert (inthash_t p, size_t key, void *value) {
  struct icell *q;
  void *old;
  q=inthash_insert_list (& p->buckets[key%N], key);
  old=q->value;
  q->value=value;
  return old;
}
\end{lstlisting}
\caption{Hashtable definition and insertion}
\label{inthash_insert}
\end{figure}

\begin{figure}
\begin{tikzpicture}

\node[rectangle split, draw, rectangle split parts = 4, label={120:p}] (bs) {0 \nodepart{two} 1 \nodepart{three} \strut $\vdots$ \nodepart{four} $N - 1$};

\node[rectangle split, rectangle split part align=base, rectangle split horizontal, draw, rectangle split parts = 3, above right = 1mm and 3cm of bs.one] (b11) {$k^0_1$ \nodepart{two} $v^0_1$ \nodepart{three} $p^0_1$};

\draw[{Circle}->] (bs.one east) to [out=30,in=200] node[above]{\scriptsize{Bucket 0}} (b11.one west);

\node[rectangle split, rectangle split part align=base, rectangle split horizontal, draw, rectangle split parts = 3, right = 2cm of b11.east] (b12) {$k^0_2$ \nodepart{two} $v^0_2$ \nodepart{three} $p^0_2$};

\draw[{Circle}->] (b11.east) to [out=-20,in=170] (b12.one west);

\node[rectangle split, rectangle split part align=base, rectangle split horizontal, draw, rectangle split parts = 3, right = 4cm of b12.east] (b1n) {$k^0_{n_0}$ \nodepart{two} $v^0_{n_0}$ \nodepart{three} $p^0_{n_0}$};

\draw[{Circle}->, dashed] (b12.east) to [out=30,in=200] (b1n.one west);

\node[rectangle split, rectangle split part align=base, rectangle split horizontal, draw, rectangle split parts = 3, below right = 1mm and 3cm of bs.two, label={-90:$\vdots$}] (b21) {$k^1_1$ \nodepart{two} $v^1_1$ \nodepart{three} $p^1_1$};

\draw[{Circle}->] (bs.two east) to [out=-20,in=160] node[below]{\scriptsize{Bucket 1}} (b21.one west);

\draw[{Circle}-, dashed] (b21.east) to [out=30,in=200] +(3, -.2);

\draw[{Circle}-, dashed] (bs.four east) to [out=-10,in=150] +(2, -.3);

\node[rectangle, fit=(b12.south west) (b1n.north east), draw = red!70, label={\color{BrickRed} $p^0_1 \leadsto l_2$}] {};

\node[rectangle, fit=(bs.one split west) (b11.north east), draw=green!70,
  label={\color{ForestGreen} \scriptsize{$p^0_1 \leadsto l_2 \wand \mathbf{listboxrep}\ p\ ((k^0_1, v^0_1) :: l_2)$}}] {};

\end{tikzpicture}
\caption{Hashtable illustration and separation logic predicates, where $l_2 := [(k^0_2, v^0_2), \dots, (k^0_{n_0}, v^0_{n_0})]$ and $p$ is the address of the first bucket pointer.}
\label{hashtable_drawing}
\end{figure}

The \ce{inthash_insert_list} function returns a pointer to the adequate cell in a bucket, by creating a new cell if needed.
The pointed cell has to be updated with the new value by the \ce{inthash_insert} function, which itself returns the old value.

\subsection{Separation logic representation}

Notice that the \ce{inthash_insert_list} function can modify the pointer in the bucket array, in the case the bucket is empty, hence the use of a pointer to a pointer in the arguments of the function. The paper and the Verifiable C tutorial describe such a situation using a \textbf{listboxrep} predicate, along with the standard separation logic representation predicate for linked lists:
\[ \begin{array}{ll}
  p \leadsto \mathrm{nil} & :=\ p = \mathrm{null}\\
  p \leadsto (k, v)::\mathit{tl} & :=\ p.\mathsf{key} \mapsto k * p.\mathsf{value} \mapsto v * \exists q, p.\mathsf{next} \mapsto q * q \leadsto \mathit{tl}\\[.3cm]
  \mathbf{listboxrep}\ l\ p & := \exists q, p \mapsto q * q \leadsto l \end{array} \]

The notations are as follows:
\begin{itemize}[noitemsep, topsep=0pt]
  \item $*$ is the separating conjunction: a heap $m$ satisfying $A * B$ (written $m \models A * B$) can be split into two disjoint parts $m_1$ and $m_2$ such that $m_1 \models A$ and $m_2 \models B$;
  \item $p \mapsto v$ means that value $v$ is stored at memory address $p$ on the heap; 
  \item $\mathrm{nil}$ is the empty list and $::$ is the constructor of functional lists - in this case, association lists or lists of pairs;
  \item finally, $\mathrm{null}$ is the null pointer.
\end{itemize}

The separating implication will be denoted $A \wand B$. By definition, $m \models A \wand B$ if and only iff for all $m'$ disjoint from $m$ such that $m' \models A$, we have $m \oplus m' \models B$, where $\oplus$ is heap disjoint union. Intuitively, $A \wand B$ describes a heap $A$ where a part $B$ is missing.

The magic-wand-as-frame paper advocates the use of that separating implication, or \emph{magic wand}, to describe partial data structures that appear on the heap when a recursive data structure (e.g. binary tree, linked list) is traversed.  The magic wand then allows for elegant loop invariants that \emph{do not} require additional inductive predicates for partial structures to be defined (e.g. list segments or partial trees). The paper shows that this technique can be used when the traversal doesn't modify the heap (e.g. for lookups) but also and more interestingly \emph{when it does}, like here. In the latter case, the wand conjunct has to use a universal quantifier over the remaining sub-structure, to be modified by the traversal. More details are given below.

In the body of \ce{inthash_insert_list} comes a loop (lines~\ref{begin_loop} to~\ref{end_loop}) in which the \emph{pointer value} stored at address \ce{r} goes through the bucket until reaching the end of the linked list or the current mapping for key $k$. With respect to the magic-wand-as-frame technique, I used the following loop invariant:

\[ \begin{array}{rl}
  \exists (l_1, l_2),\ l = l_1l_2\ \wedge\ \llbracket \mathtt{r} \rrbracket = r\ \wedge\ \llbracket \mathtt{r_0} \rrbracket = r_0\ \wedge\ \llbracket \mathtt{k} \rrbracket = k\ *\
   \mathbf{listboxrep}\ r\ l_2\ * & \\
   \forall l_2',\ \mathbf{listboxrep}\ r\ l_2'\ \wand\ \mathbf{listboxrep}\ r_0\ (l_1l_2') & \end{array}\]

The concatenation of two lists $l_1$ and $l_2$ is written $l_1l_2$ above. The universal quantification on the sublist whose starting address is stored at $r$ before the wand means that using that loop invariant, it is possible to alter that list when needed. In the case of \ce{inthash_insert_list}, that happens when a new cell has to be added at the end, because the key being searched wasn't found. In that case, the new cell will be given the \ce{NULL} mapping. That value will be modified by the \ce{inthash_insert} function.

When specifying \ce{inthash_insert_list}, it is necessary to include the fact that the returned value is pointing to the list cell that needs to be modified by \ce{inthash_insert}. I thus came up with the following specification:
\[ \begin{array}{cl}
   \mathrm{(precondition)} & \{ \llbracket \mathtt{p} \rrbracket = p \wedge \llbracket \mathtt{k} \rrbracket = k \wedge \mathbf{listboxrep}\ l\ p \}\\
  \mathrm{(statement)} & \qquad p_\mathrm{ret} = \mathtt{inthash\_insert\_list(p, k)}\\
  \mathrm{(postcondition)} & \{ \exists (r, v_0, \mathit{tl}),\ r \mapsto p_\mathrm{ret} * p_\mathrm{ret} \leadsto (k, v_0)::\mathit{tl}\ *\\
  & \qquad \qquad \forall v,\ r \mapsto p_\mathrm{ret} * p_\mathrm{ret} \leadsto (k, v)::tl \wand \mathbf{listboxrep}\ (\mathbf{add}\ k\ v\ l)\ p\}
\end{array} \]

Here, $\mathbf{add}$ is the insertion on functional association lists.
To have a sufficiently strong postcondition, it is \emph{not possible} to use the \textbf{listboxrep} predicate for the two first conjuncts and before the wand in the postcondition.
If we do, because of the existential quantification in \textbf{listboxrep} on the value stored at $r$, we lose the information that the returned value $p_\mathrm{ret}$ points to a sublist whose head key is the inserted key.
That's why, contrary to the paper's examples, I had to unfold the definition of \textbf{listboxrep}.

Notice how adequate and specific to the problem that specification is. Once \ce{inthash_insert_list} is proven to conform to that specification, the proof of \ce{inthash_insert} - with a natural specification matching the new hashtable in the heap with the same hashtable where key $k$ now maps to value $v$ - goes very smoothly, because most of the work is done in that auxiliary proof.

Finally, the C code could have been modified in the following way: pass the new value to\linebreak\ce{inthash_insert_list}, and make it perform the update and return the old value. In that case, we wouldn't have had to give a magic-wand postcondition for \ce{inthash_insert_list}, which may have been a little simpler. However, it is interesting to see that the magic-wand-as-frame technique can be adapted to various situations like this piece of C code.

\section{A verified B+tree functional implementation}
\label{btree_section}
B+trees are a specific kind of multiway search trees that store ordered mappings of a key to an associated value. B+trees are derived from B-trees, which were originally described by Bayer and McCreight~\cite{OMLOI}. B+trees have been designed to handle massive amounts of data while allowing for the shortest lookup time. Such trees are typically very wide and shallow, so that the number of nodes fetched from cold areas of cache or main memory on a path from the root to a leaf is small. The number of children of a node is also chosen so as to minimize cache misses, that is, so that one node of a B+tree is the same size as the block size of one layer of the memory hierarchy (disk or cache).

This section describes the B+tree data structures and its invariants, then exposes some challenges I faced during formalization in Coq.

\subsection{Presentation of B+trees}
\label{btree_maths}
Informally, B+tree nodes have the following structure. A \emph{leaf node} will contain a number of \emph{entries} which are key-value pairs and solely constitute the information contained in the abstract key-value store.
An \emph{internal node} contains a \emph{zero} (or \emph{leftmost}) \emph{pointer} and a list of key-child pointer pairs - also called entries. The entry keys guide the lookup operation towards the right mapping in the appropriate leaf. The lookup top-to-bottom traversal is similar to that of any binary search tree - except that the number of children is greater. Contrary to B-trees, B+tree internal nodes do not carry any mapping information: their only purpose is to make lookups fast using intermediate keys.

Formally, the inductive definition of the base type of the B+trees is the following:
\[ t \quad = \quad T(t_0, (k_i, t_i)_{i = 1}^{n_t})\quad | \quad L((k_i, v_i)_{i = 1}^{n_t}) \]
where $t, t_0, t_1, \dots, t_{n_t}$ are trees, the $k_i$'s are keys, the $v_i$'s are values, a pair $(k, x)$ where $k$ is a key and $x$ is either a tree or a value is called an \emph{entry}, and the number $n_t$ of entries is tree-dependent.
The words \emph{tree} and \emph{node} will be used interchangeably in the rest of the explanations. The constructor $T$ refers to internal nodes, while $L$ refers to leaves.

Along with that definition, we will use the usual notions of \emph{child}, \emph{tree path} and \emph{path length}.
An internal node $T(t_0, (k_i, t_i)_{i = 1}^{n_t})$ has $n_t + 1$ children, which are trees $t_0, t_1, \dots, t_{n_t}$. That is, the leftmost pointer is a child, and so are the trees in the entries.
Leaf nodes have no child. If $t_\text{child}$ is a child of $t_\text{parent}$, we write $\mathsf{child}\ t_\text{child}\ t_\text{parent}$.
A path will be an ordered collection of distinct nodes $(t_0, t_1, \dots, t_d)$ such that $\forall i \in \{0, \dots, d - 1\},$ $t_{i + 1}$ is a child of $t_i$.
Such a path is said to have length $d$, which is the number of edges it is composed of.

A node $t_\text{child}$ is said to be a \emph{subnode} of another node $t_\text{parent}$ if there is a path from $t_\text{parent}$ to $t_\text{child}$.
That defines a predicate and we write $\mathsf{subnode}\ t_\text{child}\ t_\text{parent}$.
The $\mathsf{subnode}$ relation is the reflexive transitive closure of the $\mathsf{child}$ relation.

Given such a multiway tree $t$ and a positive integer $m \in \mathbb{N}_{>0}$, we say that $t$ is a B+tree of order~$m$ if it satisfies the following invariants.
\begin{enumerate}
  \itemsep0em
  \item $t$ is \emph{balanced} in the sense that \emph{all paths from t to one of its leaves have the same length}. That length is called the \emph{depth} of the B+tree.

    That property guarantees that looking up any key in the tree will have the same time complexity.
    It is an even stricter balancing invariant than the one of \emph{AVL trees}, which are binary search trees allowing pairs of subtrees to have a height difference of 1.
    This invariant is concomitant to the way an insertion is performed:
    as we will see, during B$\pm$tree insertion, a new node is always a sibling of another preexisting node, or a new root; in other words, B$\pm$trees grow in height at the root, not the leaves.
    That behavior is inherited from 2-3 trees, of which B-trees are a generalization.
  \item $t$ is \emph{well-sized}, that is $t$ has at most $2 * m$ entries and any strict subnode of $t$ has between $m$ and $2 * m$ entries.
    If $t$ is not a leaf, then $t$ must also have at least 1 entry: no internal node has an empty list of entries.

    That invariant guarantees that the complete B$\pm$tree will have a number of nodes that is not too high compared to the number of mappings it contains. The interest of that invariant is to rule out nearly-empty nodes.
In the early days of databases, all the data was stored on slow discs, and a B+tree node's entries would be stored in a dedicated sector or block.
Having nodes with few entries would increase disk usage as well as the total depth of the tree, which decreases cache locality of the program and makes lookups slower.
    The same reasoning applies today to RAM blocks and/or cache lines.
  \item \label{} $t$ is a \emph{search tree}. That property is crucial not for performance, but for proper operation.

    For a leaf node $L((k_i, v_i)_{i = 1}^{n_t})$, we demand that the list of keys $k_1, \dots, k_{n_t}$ be sorted, with respect to whatever strict total order the type of keys comes equipped with.
    A B+tree doesn't allow duplicate keys, however an implementation allowing them seems feasible.

    For an internal node $t = T(t_0, (k_i, t_i)_{i = 1}^{n_t})$, we require that
    \begin{enumerate}
      \item the list of keys $k_1, \dots, k_{n_t}$ is sorted;
      \item all the children of $t$ are search trees;
      \item all keys $k$ in $t_0$ are such that $k \le k_0$ (remember that, if the tree is well-sized, then $n_t > 0$, so $k_0$ exists);
      \item for any $i \in \{1, \dots, n_t - 1\}$, the keys $k$ in $t_i$ are such that $k_i < k \le k_{i + 1}\ $; and
      \item the keys $k$ in $t_{n_t}$ are such that $k > k_{n_t}$.
    \end{enumerate}

\end{enumerate}

\begin{figure}
\centering
\begin{tikzpicture}
  \node[draw, rectangle split, rectangle split horizontal, rectangle split empty part width=-10, rectangle split parts = 9, inner sep = 2mm] (n)
       {\nodepart{two} $\quad k_1 \quad$ \nodepart{three} \nodepart{four} $\quad k_2 \quad$ \nodepart{five} \nodepart{six} $\quad \dots \quad$ \nodepart{seven} \nodepart{eight} $\quad k_{n_t} \quad$ \nodepart{nine}};

  \node[above=2mm of n.one north] (lm) {\small{0}}
    edge[->] (n.one north);

  \node[above=2mm of n.three north] {\small{1}}
    edge[->] (n.three north);

  \node[above=2mm of n.five north] {\small{2}}
    edge[->] (n.five north);

  \node[above=2mm of n.seven north] {\small{$n_t - 1$}}
    edge[->] (n.seven north);

  \node[above=2mm of n.nine north] {\small{$n_t$}}
    edge[->] (n.nine north);

  \node[ellipse, dashed, draw, below left= 1.2cm of n.one south] (t0) {\begin{tabular}{c} \small{$t_0$ (leftmost child)} \\ $k \le k_1$ \end{tabular} };

  \node[ellipse, dashed, draw, below right= 1cm and -.8cm of n.three south] (t1) {\begin{tabular}{c} $t_1$ \\ $k_1 < k \le k_2$ \end{tabular}};

  \node[below right = 1.4cm and .2cm of n.seven south] {\dots};

  \node[ellipse, dashed, draw, below right= 1.2cm of n.nine south] (tnt) {\begin{tabular}{c} $t_{n_t}$ \\ $k > k_{n_t}$ \end{tabular}};

  \path[-{Latex[width=1.5mm, length=2mm]}, dashed] (n.one south) edge (t0.20);
  \path[-{Latex[width=1.5mm, length=2mm]}, dashed] (n.three south) edge (t1.north);
  \path[-{Latex[width=1.5mm, length=2mm]}, dashed] (n.nine south) edge (tnt.140);

\end{tikzpicture}
\caption{A B+tree internal node. The numbers above indicate child positions used in B+tree cursors.}
\end{figure}

\begin{figure}
\centering
\begin{tikzpicture}[edge from parent/.style={draw=none}]
\tikzstyle{btree_node}=[rectangle split, rectangle split horizontal, rectangle split parts = 5, rectangle split empty part width=-10, draw]

\node [fill=red!10, rectangle split, rectangle split ignore empty parts, draw] at (-3, 0.7) (cursor) {1 \nodepart{two} 2 \nodepart{three} 0};

\node[btree_node, rectangle split part fill = {blue!10, blue!10, blue!50}, rectangle split parts = 3] (root) { \nodepart{two} 15 }
  child[sibling distance = 50mm] {node[btree_node] (d10) {\nodepart{two} 3 \nodepart{four} 7}
    child[sibling distance=15mm] {node[btree_node] (d20) {\nodepart{two} 1 \nodepart{four} 3}}
    child {node[btree_node] (d21) {\nodepart{two} 5 \nodepart{four} 6}}
    child[sibling distance=18mm] {node[btree_node] (d22) {\nodepart{two} 11 \nodepart{four} 15}}    
  } 
  child[sibling distance=115mm] {node[btree_node, rectangle split parts = 9, rectangle split part fill = {blue!10, blue!10, blue!10, blue!10, blue!50, blue!10}] (d11) {\nodepart{two} 21 \nodepart{four} 28 \nodepart{six} 32 \nodepart{eight} 50}
    child[sibling distance=20mm] {node[btree_node] (d23) {\nodepart{two} 17 \nodepart{four} 20}}
    child[sibling distance=20mm] {node[btree_node] (d24) {\nodepart{two} 22 \nodepart{four} 25}}
    child {node[btree_node, rectangle split part fill = {blue!50, blue!10}] (d25) {\nodepart{two} 29 \nodepart{four} 32}}    
    child[sibling distance=24mm] {node[btree_node, rectangle split parts = 7] (d26) {\nodepart{two} 34 \nodepart{four} 38 \nodepart{six} 44}}    
    child[sibling distance=25mm] {node[btree_node, rectangle split parts = 7] (d27) {\nodepart{two} 53 \nodepart{four} 54 \nodepart{six} 60}}    
  };
\path [->] (root.one south) edge (d10.north west);
\path [->] (root.three south) edge (d11.north west);
\path [->] (d10.one south) edge (d20.north west);
\path [->] (d10.three south) edge (d21.north west);
\path [->] (d10.five south) edge (d22.north west);
\path [->] (d11.one south) edge (d23.north west);
\path [->] (d11.three south) edge (d24.north west);
\path [->] (d11.five south) edge (d25.north west);
\path [->] (d11.seven south) edge (d26.north west);
\path [->] (d11.nine south) edge (d27.north west);

%cursor arrows
\path [{Circle[length=2mm]}->, gray!80, bend left = 15, dashed] (cursor.one east)+(-1mm,0) edge (root.north west);
\path [{Circle[length=2mm]}->, gray!80, bend right = 15, dashed] (cursor.two east)+(-1mm,.3mm) edge (d11.north west);
\path [{Circle[length=2mm]}->, gray!80, bend right = 15, dashed] (cursor.three east)+(-1mm,.5mm) edge (d25.north west);

\node[below=0 of root.three south] {\small{1}};
\node[below=0 of d11.five south] {\small{2}};
\node[below=0 of d25.one south] {\small{0}};

\end{tikzpicture}
\caption{A sample B+tree of depth 2 showing the ``search tree'' property, with parameter $m = 2$.
The cursor in light red points to key 29. The nodes on the path designated by the cursor are in light blue.
The numbers in the cursor represent the position of the adequate child or record in the node that is pointed to. The corresponding pointers in the B+tree are highlighted in blue.}
\label{bptree}
\end{figure}

\subsection{Coq model and proofs}

To my knowledge, the only proof assistant formalization of B+trees in the literature is described by Malecha et al.~\cite{TVRDMS}.
Their datatype for B+trees includes a \coqe{nat} height parameter, which allows Coq to check the termination of recursion easily.
They define a tree traversal that is supposed to serve for the implementation of different B+tree operations.
However, they only really use it for insertion, which is the only operation they define along with lookup.
I didn't define or verify the removal operation either, which is expected to be quite lengthy.
However, my formalization is made to match the \emph{cursored} operations we use in our C code.

A B+tree cursor is a path from the root to a leaf record, that is kept in memory by the B+tree C program.
They can be found in the B+tree implementation of SQLite~\cite{DGS}.
They allow speedups on large trees when consecutive lookups and insertions are made \emph{locally} in the tree.
Moreover, they are an alternative to cross-leaf pointers used in most implementations of B+trees~\cite{TVRDMS}~\cite{DSCB} for iterating over all sorted mappings contained in the tree.
A cursor is presented in figure~\ref{bptree}.

Using a cursor, the lookup and insertion operations do not have to traverse the tree from the root.
The leaf where the key is found is reached from the current position of the cursor.
If a sequence of keys is inserted in a sorted or mostly-sorted order, the cursor can provide a noticeable speedup,
because the cursor would move very little, and would trigger fewer cache misses compared to cursorless insertions or lookups, that would need to traverse the whole tree from top to bottom.

My formalization uses a Coq list to describe the entries in a B+tree node, which feels like the most natural way to define multiway search trees.
It takes the form of a Coq functor, and is inspired from the \coqe{FMapAVL} file~\cite{CSTDLAVL} from Coq's standard library,
which implements the finite map interface in \coqe{FMapInterface} using fully verified AVL trees.
However, that part of the standard library is unfortunately deprecated and doesn't seem to be maintained so I chose not to rely on it.

\begin{coq}
Module Type BPlusTreeParams.
  Module InfoTyp.
    Parameter t: Type.
  End InfoTyp.
  Parameter min_fanout: Z.
  Axiom min_fanout_pos: min_fanout > 0.
End BPlusTreeParams.

Module Raw (X: Orders.UsualOrderedTypeFull') (Params: BPlusTreeParams).
  Notation key := X.t.
  Definition entries (elt: Type): Type := list (key * elt).

  Inductive node: Type :=
  | leaf (records: entries elt) (info: Info.t): node
  | internal (ptr0: node) (children: entries node) (info: Info.t): node.

  [...]
End Raw.
\end{coq}

Each node stores an \coqe{Info.t} informative value, with no constraint.
The \coqe{min_fanout} parameter corresponds to the $m$ B+tree parameter above.
Keys need to be equipped with a total order.
I have investigated the use of any equivalence relation for equality of keys,
however that was not needed for our purpose and required a lot more work to prove that the non-Leibniz equality was respectful of all other definitions, which is needed for setoid rewriting.
Thus the key type is an instance of \coqe{UsualOrderedType} and not \coqe{OrderedType}.

A \emph{nested inductive type} occurs in the \coqe{children} argument of the \coqe{internal} constructor, as it is a list whose element type contains the inductive type \coqe{node} being defined.
Coq can handle nested inductive types, however I had to convince the termination checker that my inductive definitions on \coqe{node} were terminating, either by structural recursion or using well-founded relations.
An alternative definition uses mutual inductive types to redefine a list of children for \coqe{node}, isomorphic to the Coq type \coqe{list}, but is very impractical given that all definitions and lemmas about lists cannot be reused.

For such nested inductive types, the induction principle automatically defined by Coq is too weak. A proper induction principle can however be proved.
Different choices are possible. I chose the \coqe{Forall} predicate for the induction hypothesis:
\begin{coq}
Lemma node_ind (P: node -> Prop)
  (hleaf: forall records info, P (leaf records info))
  (hinternal: forall ptr0 children info,
      P ptr0 -> Forall P (map snd children) -> P (internal ptr0 children info)):
$\forall$ n, P n.
\end{coq}

Two methods are documented for recursive definitions on such nested inductive types:
\paragraph{Well-founded relations.}
The first method is to use a well-founded relation, or to use a measure and a predefined relation (e.g. less-than on natural numbers).
My take on that was inspired by what is done in the Equations package~\cite{ER}.
Equations has tactics and a command \coqe{Derive Subterm} that try to define automatically a \emph{subterm} relation for recursive inductive types and prove its well-foundedness.
In my case, I defined the following:
\begin{coq}
Inductive child: relation node :=
| child_ptr0: forall ptr0 records info,
    child ptr0 (internal ptr0 records info)
| child_entry: forall n ptr0 children info,
    In n (map snd children) ->
    child n (internal ptr0 children info).

Lemma child_wf: well_founded child.
\end{coq}

I used that technique to define the \coqe{get_cursor} function which, given a key $k$ and a B+tree,
returns a list of nodes and node indices on the path from the root to where the B+tree stores (or would be storing) $k$.
That ``list of indices and nodes'' cursor type is the reason why I didn't encode the height of a tree (and its balancing property) inside the \coqe{node} datatype itself.
\begin{coq}
Function get_cursor (k: key) (root: node) {wf child root}: list (Z * node) :=
match root with
| leaf records info => [(find_index k records, root)]
| internal ptr0 children info =>
    let i := find_index k children in
    (i, root)::get_cursor k (if i $=_\mathrm{dec}$ 0 then ptr0
                                else Znth (i - 1) (map snd children))
end.
Proof. [...] Defined.
\end{coq}
The \coqe{find_index} function returns a number indicating which child of an internal node should contain a given key,
or in the case of a leaf, at which position in the entries list a mapping with a given key is or should be inserted.

\paragraph{Structural recursion and nested fixpoints.}
Another method is to use basic structural recursion.
But Coq won't accept all fixpoint definitions if the structurally smaller argument is a \coqe{node}, even though most of the time termination is obvious to the user.

The following \emph{flatten} function, which transforms a B+tree into an ordered association list, works straight out-of-the-box:
\begin{coq}
Fixpoint flatten (n: node) {struct n}: entries elt :=
match n with
| leaf records _ => records
| internal ptr0 children _ => flatten ptr0 ++ flat_map (flatten $\circ$ snd) children
end.
\end{coq}

However, to define the \coqe{search_tree} predicate formally described in section~\ref{btree_maths},
one needs to define the treatment of the list of children using a \emph{nested fixpoint} for Coq to understand that the recursive call happens on a structurally decreasing argument.
That technique is described in various sources like in Chlipala~\cite{CPDT}, Bertot and Castran~\cite{CoqArt} and the Coq-Club mailing list archives. 
\begin{coq}
Fixpoint search_tree (min max: key) (n: node) {struct n}: Prop :=
match n with
| leaf records info => Sorted ($<_\mathrm{key}$) (map fst records) /\
    min $\le_\mathrm{key}$ max /\
    Forall (fun k => min $\le_\mathrm{key}$ k /\ k $\le_\mathrm{key}$ max) (map fst records)
| internal ptr0 children info => exists max_ptr0,
    search_tree min max_ptr0 ptr0 /\
    (fix st_children l min: Prop :=
    match l with
    | nil => min $\le_\mathrm{key}$ max
    | (k, n)::tl =>
        min $\le_\mathrm{key}$ k /\
        exists min_n max_n,
          k $<_\mathrm{key}$ min_n /\
          search_tree min_n max_n n /\
          st_children tl max_n
        end) children max_ptr0
    end.
\end{coq}

The meaning of \coqe{search_tree} $m\;M\;n$ is the following: $n$ is a search tree, and all its keys $k$ are such that $m \le_\mathrm{key} k \le_\mathrm{key} M$.

Among the properties proved about \coqe{search_tree} is the sortedness (with respect to key order) of the association list obtained by flattening a node satisfying \coqe{search_tree}.
I also defined the \emph{insertion with cursor} B+tree operation. Its code is not reproduced here for conciseness.
The insertion uses a list of nodes and integer positions output by \coqe{get_cursor}, and takes a key and a value to insert in the B+tree.
Starting from the leaf pointed to by the cursor, it inserts the new entry at the position given by the integer index.
If the node becomes too filled to satisfy the well-sized invariant, it is split in 2 distinct nodes, that need to be updated or inserted in the parent.
The procedure goes up the tree along the cursor, updating and inserting a new child as needed:
if internal nodes become too large, they are also split and generate a new node to insert above them.
The tree grows with a new root if the current root has itself to be split.

Finally, the main results of my development\footnote{Available in the DeepSpecDB repository at \url{www.github.com/PrincetonUniversity/DeepSpecDB/blob/master/verif/btrees/btrees_functional.v}} is the conservation of the different B+tree invariants after insertion.
For instance, I proved the following statement (400 lines of code):
\begin{coq}
Theorem insert_search_tree: forall root m M k0 e,
    search_tree m M root -> search_tree (key_min m k0) (key_max M k0) (insert k0 e root).
\end{coq}
\coqe{key_min} and \coqe{key_max} being defined using the functor \coqe{GenericMinMax} from the standard library.

\section{Conclusion}

My work has spanned several facets of the design of verified low-level database algorithms.
While formal methods are nowadays commonly used for ensuring that critical systems comply with their specification,
they have surprinsingly seldom been applied to data-centric systems.
Therefore, an in-depth study of the topic was a required first step.
The implementation of the standard algorithms led me to use libraries
that had been incompletely verified, or not verified at all.
I contributed to their Verifiable C verification (hashtables) as well as their Coq formalization and verification (B+trees).

Overall, I became very experienced with Coq, I gained a deep understanding of the operation of RDBMSs, and
I could put my knowledge of separation logic in practice using Verifiable C.

\newpage
\bibliographystyle{plain}
\bibliography{biblio}
\end{document}

### 2019-08-20 verif/report_Pablo.tex
Corrected 2 small mistakes

--- 2019-08-20 papers/Pablo/report_Pablo.tex
@@
 \end{tikzpicture}
-\caption{A sample B+tree of depth 2 showing the ``search tree'' property, with parameter $m = 2$.
+\caption{A sample B+tree of order 2 and depth 2 showing the ``search tree'' property.
 The cursor in light red points to key 29. The nodes on the path designated by the cursor are in light blue.
@@
 one needs to define the treatment of the list of children using a \emph{nested fixpoint} for Coq to understand that the recursive call happens on a structurally decreasing argument.
-That technique is described in various sources like in Chlipala~\cite{CPDT}, Bertot and Castran~\cite{CoqArt} and the Coq-Club mailing list archives. 
+That technique is described in various sources like in Chlipala~\cite{CPDT}, Bertot and Castéran~\cite{CoqArt} and the Coq-Club mailing list archives. 
 \begin{coq}

### 2020-06-04 concurrency/Verification of Concurrent Binary Search Tree.tex
Latex template for documentation

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

\documentclass[11pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc}

\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round

\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{semantic}
\usepackage{url}

\bibliographystyle{plain}% the recommended bibstyle

\title{Verifying a Binary Search Tree with Fine-Grained Locking }
\author{William Mansky}
\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle

\section{Introduction}

\section{Background}
\subsection{VST and Iris}
\subsection{Atomic Specifications}

\section{Safety Proofs}
\subsection{Lock invariants for hand-over-hand locking}

\section{Correctness Proofs}
\subsection{Locking and atomicity} 
\subsection{Insert and Lookup}
\subsection{Delete}

\section{Related Work}
\section{Conclusion}

\end{document}

### 2020-06-20 papers/luke/dsss18-pre.tex
include local writings/slides

\documentclass{beamer}
\usetheme{Copenhagen}
\usepackage[utf8]{inputenc}
\usepackage{minted}
\usepackage{tikz}
\usetikzlibrary{shapes}
\setbeamertemplate{headline}{}
\AtBeginSection[]
{
  \begin{frame}
    \frametitle{Table of Contents}
    \tableofcontents[currentsection]
  \end{frame}
}

\title{Specifying Tables with Cursors}
\author{Yixuan Chen\inst{1} \and Aur\`ele Barri\`ere\inst{2} \and Brian McSwiggen\inst{3}}
\institute{
  \inst{1}University of Michigan \and \inst{2} \'Ecole Normale Sup\'erieure de
  Rennes \and
  \inst{3}Princeton University
}
\date{DeepSpec Summer School, 2018}

\begin{document}

\frame{\titlepage}

\begin{frame}[t,fragile]
\frametitle{What is a Table?}

\pause

Table is a collection of pair of key and value that supports \texttt{get} and
\texttt{put} operations. \pause

\begin{onlyenv}<3>
  \begin{block}{Specification for a Table}
    \begin{minted}[fontsize=\footnotesize]{coq}
Module Type Table (K: DecidableType).
  Parameter table: Type -> Type.
  Section Types.
    Variable elt: Type.
    Parameter empty: table elt.
    Parameter get: K.t -> table elt -> option elt.
    Parameter put: K.t -> elt -> table elt -> table elt.
    [...]
  End Types.
End Table.
    \end{minted}
  \end{block}
\end{onlyenv}

\begin{onlyenv}<4-5>
  \begin{block}{Specification for a Table(Cont.)}
    \begin{minted}[fontsize=\footnotesize]{coq}
Module Type Table (K: DecidableType).
  [...]
  Section Spec.
    [...]
    Axiom get_empty: get k empty = None.
    Axiom put_shadow: get k (put k e t) = Some e.
    Axiom put_permute: k1 <> k2 -> get k1 (put k2 e t) = get k1 t.
  End Spec.
End Table.
    \end{minted}
  \end{block}
\end{onlyenv}

\only<5>{A spec of table resembles to the \texttt{FMap} in Coq's standard library.}

\end{frame}

\begin{frame}
\frametitle{What's the problem?}

\begin{itemize}
\item<2-> Performance
  \begin{itemize}
  \item Operations are independent, serious issue for iterators
  \item Range queries are not modeled
  \end{itemize}
\item<3-> Engineering
  \begin{itemize}
  \item Difficulty when incorporating into recursive definitions as a client
  \end{itemize}
\end{itemize}

\end{frame}

\section{Table with Cursors}

\begin{frame}[t]
\frametitle{Table with Cursors}

\begin{itemize}
\item<1-> Cursors are positions into the table
\item<2-> Each cursor is associated with key(s)
  \begin{itemize}
  \item<3-> In principle, the table should be separated by the cursor and the
    associated key in the same way
    \begin{center}
      \begin{tikzpicture}
        \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle
        split ignore empty parts,draw] \tikzstyle{every node}=[bplus]
        \node {cursor(140)} [->]
        child {node[rectangle split parts=6] {124\nodepart{two} 130\nodepart{three} 135\nodepart{four} 142\nodepart{five} 197\nodepart{six} 201}};
      \end{tikzpicture}
    \end{center}
  \end{itemize}
\item<4-> However, it can be anything that implements the interface
  functions
  \begin{itemize}
  \item A stupid one could simply save the key in the cursor 
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}[t,fragile]
\frametitle{Interface of Cursored Table}

\begin{onlyenv}<1>
  \begin{block}{Specification for a Cursored Table}
    \begin{minted}[fontsize=\footnotesize]{coq}
Module Type CursoredTable (K: OrderedType).
  Parameter table: Type -> Type.
  Parameter cursor: Type -> Type.
  Section Types.
    Variable elt: Type.
    Parameter make_cursor: K.t -> table elt -> cursor elt.
    Parameter get: cursor elt -> table elt -> option (K.t * elt).
    Parameter put: cursor elt -> elt -> table elt ->
                   (cursor elt * table elt).
  End Types.
End Table.
    \end{minted}
  \end{block}
\end{onlyenv}

\begin{onlyenv}<2>
  \begin{block}{Specification for a Cursored Table(Count.)}
    \begin{minted}[fontsize=\footnotesize]{coq}
Module Type CursoredTable (K: OrderedType).
  Parameter table: Type -> Type.
  Parameter cursor: Type -> Type.
  Section Types.
    [...]
    Parameter next_cursor: cursor elt -> table elt -> cursor elt.
    Parameter prev_cursor: cursor elt -> table elt -> cursor elt.
    Parameter first_cursor: table elt -> cursor elt.
    Parameter last_cursor: table elt -> cursor elt.
  End Types.
End Table.
    \end{minted}
  \end{block}
\end{onlyenv}

\begin{onlyenv}<3>
  \begin{block}{Specification for a Cursored Table(Count.)}
    \begin{minted}[fontsize=\footnotesize]{coq}
Module Type CursoredTable (K: OrderedType).
  Parameter table: Type -> Type.
  Parameter cursor: Type -> Type.
  Section Types.
    [...]
    Parameter table_correct: table elt -> Prop.
    Parameter cursor_correct: table elt -> Prop.
    Parameter abs_rel: cursor elt -> table elt -> Prop.
    Parameter key_rel: K.t -> cursor elt -> table elt -> Prop.
    Parameter eq_cursor: cursor elt -> cursor elt ->
                         table elt -> Prop.
  End Types.
End Table.
    \end{minted}
  \end{block}
\end{onlyenv}

\begin{onlyenv}<4>
  \begin{block}{Specification for a Cursored Table(Count.)}
    \begin{minted}[fontsize=\footnotesize]{coq}
Module Type CursoredTable (K: OrderedType).
  Parameter table: Type -> Type.
  Parameter cursor: Type -> Type.
  Section Specs.
    [...]
    Axiom make_cursor_key:
      table_correct t -> key_rel k (make_cursor k t) t.
    Axiom put_shadow:
      key_rel k c1 (snd (put k e c2 t)) ->
      key_rel k c2 t ->
      get c1 (snd (put k e c2 t)) = Some (k, e).
  End Specs.
End Table.
    \end{minted}
  \end{block}
\end{onlyenv}

\begin{onlyenv}<5>
  \begin{block}{Specification for a Cursored Table(Count.)}
    \begin{minted}[fontsize=\footnotesize]{coq}
Module Type CursoredTable (K: OrderedType).
  Parameter table: Type -> Type.
  Parameter cursor: Type -> Type.
  Section Specs.
    [...]
    Axiom next_order:
      ~ eq_cursor c (last_cursor t) t ->
      key_rel k1 c t -> key_rel k2 (next_cursor c t) t ->
      KeyType.lt k1 k2.
    Axiom next_compact:
      ~ eq_cursor c (last_cursor t) t ->
      key_rel k1 c t -> key_rel k2 (next_cursor c t) t ->
      forall c' k3,
      ~ eq_cursor c c' t -> key_rel k3 c' t ->
      lt k3 k1 \/ gt k3 k2.
  End Specs.
End Table.
    \end{minted}
  \end{block}
\end{onlyenv}

% TODO:

\end{frame}

\section{Table with Flattening}

\begin{frame}[t,fragile]
\frametitle{A Functional Client's Perspective}

Suppose we are going to define a tree with finite degrees. \\

\begin{onlyenv}<2>
  \begin{block}{Tradition Definition}
\begin{minted}[fontsize=\footnotesize]{coq}
Inductive tree: Type :=
| node_of: list tree -> tree.
\end{minted}
  \end{block}
\end{onlyenv}

\only<3->{Now suppose we have a key associated with each subtree.}

\begin{onlyenv}<4->
  \begin{alertblock}{Definition as a Client of Table}
\begin{minted}[fontsize=\footnotesize]{coq}
Module Tree (K: DecidableType) (NodeTable: Table K).
  Inductive tree: Type :=
  | node_of: NodeTable.table tree -> tree.
End Tree.
\end{minted}
  \end{alertblock}
\end{onlyenv}

\only<5->{This doesn't work! Coq has no knowledge about positiveness about
  \texttt{NodeTable.table}.}

\end{frame}

\begin{frame}
\frametitle{Remedy for the Problem}

To make the module type useful for such a client, there are possible several
solutions.

\begin{itemize}
\item<2-> Turn off the positive check
  \begin{itemize}
  \item<3-> not the ideal solution
  \end{itemize}
\item<4-> Define the strict-positiveness as part of the spec., and
  let the implementation prove it
  \begin{itemize}
  \item<5-> seems to be too much difficulty for an undergrad
  \end{itemize}
\item<6-> Define \texttt{flatten} function that turns the table into some
  already known strict positive type
  \begin{itemize}
  \item<7-> our choice!
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}[t,fragile]
\frametitle{Choice of Flattened Form}

There are two candidates for the flatten target, either a partial function or a
list of pairs.

\begin{onlyenv}<2-4>
  \begin{block}{Definition for Some Toy Tree}
\begin{minted}[fontsize=\footnotesize]{coq}
Inductive tree: Type := node_of: Z -> (Z -> option tree) -> tree.
\end{minted}
  \end{block}
\end{onlyenv}

\begin{onlyenv}<3-4>
  \begin{alertblock}{Representation Predicate for Partial Function Form}
\begin{minted}[fontsize=\footnotesize]{coq}
Fixpoint tree_rep {cs: compspecs} (k: Z) (t: tree): mpred :=
  match t with
  | node_of n subtrees =>
    EX p: val, data_at Tsh tint (Vint (Int.repr n)) p *
    (ALL k': Z, ALL t': tree,
     !! (subtrees k' = Some t') -* (tree_rep k' t'))
  end.
\end{minted}
  \end{alertblock}
\end{onlyenv}

\only<4>{It turns out we need to \textit{flatten} the partial function again
  into a list to make it work.}

\begin{onlyenv}<5-6>
  \begin{block}{Definition for Some Toy Tree}
\begin{minted}[fontsize=\footnotesize]{coq}
Inductive tree: Type := node_of: Z -> list (Z * tree) -> tree.
\end{minted}
  \end{block}
\end{onlyenv}

\begin{onlyenv}<6>
  \begin{block}{Representation Predicate for List Form}
\begin{minted}[fontsize=\footnotesize]{coq}
Fixpoint tree_rep {cs: compspecs} (k: Z) (t: tree): mpred :=
  match t with
  | node_of n subtrees =>
    EX p: val, data_at Tsh tint (Vint (Int.repr n)) p *
    iter_sepcon (fun b => tree_rep (fst b) (snd b)) subtrees
  end.
\end{minted}
  \end{block}
\end{onlyenv}

\end{frame}

\begin{frame}[t, fragile]
\frametitle{Strengthen the Flattening}

Guaranteeing the shape of flattened table is not satisfiable for the client,
because all the beautiful properties of the table are lost. \\ \pause

It would be great if the flattened list still provides \texttt{get} and
\texttt{put} functions and certain properties. \pause

\begin{onlyenv}<3>
  \begin{block}{List Table}
\begin{minted}[fontsize=\footnotesize]{coq}
Module ListTable (K: DecidableType) <: Table K.
  Definition table (elt: Type): Type := list (K.t * elt).
  [...]
End Module.
\end{minted}
  \end{block}
\end{onlyenv}

\begin{onlyenv}<4>
  \begin{block}{Flattenable Table}
\begin{minted}[fontsize=\footnotesize]{coq}
Module Type FlattenableTable (K: DecidableType) <: Table K.
  Include Table K.
  Module Flattened := ListTable K.t.
  Section Types.
    Variable elt: Type.
    Parameter flatten: table elt -> Flattened.table elt.
  End Types. 
End Module.
\end{minted}
  \end{block}
\end{onlyenv}

\begin{onlyenv}<5>
  \begin{block}{Flattenable Table(Count.)}
\begin{minted}[fontsize=\footnotesize]{coq}
Module Type FlattenableTable (K: DecidableType) <: Table K.
  [...]
  Section Specs.
    Axiom flatten_correct: forall t,
      table_correct t ->
      Flattened.table_correct (flatten t) /\
      forall (k: K.t), get k t = Flattened.get k (flatten t).
  End Specs. 
End Module.
\end{minted}
  \end{block}
\end{onlyenv}

\begin{onlyenv}<6>
  \begin{block}{Some Useful Flattenable Table Facts}
\begin{minted}[fontsize=\footnotesize]{coq}
Module FlattenableTableFacts
   (K: DecidableType) (T: FlattenableTable K).
  Include T.
  [...]
  Theorem put_permute (k: key) (v: elt) (t: table elt):
    table_correct t ->
    flatten (put k v t) = put k v (flatten t). 
End Module.
\end{minted}
  \end{block}
\end{onlyenv}

\end{frame}

\begin{frame}
\frametitle{Extending Flattening to Cursored Table}

It turns out, the \texttt{flatten} can be extended to table with cursors as
well. \\ \pause

Of course, there is much more effort when proving with cursors.

\end{frame}

\section{Progress and the Future}

\begin{frame}
\frametitle{What Are Done So Far}

\begin{itemize}
\item<1-> We have a relatively complete spec. for the cursored tables (Brian
  and Luke)
\item<2-> We have verified the implementation of B+-Trees in C with respects to
  the functional implementation (Aurele)
\item<3-> An cursored table implemented by sorted list and related facts have
  been (mostly) proven for clients
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{What Will be in the Future}

\begin{itemize}
\item An ongoing effort to verify a trie as both an client and provider of
  this spec.
  \begin{itemize}
  \item As a verification of the correctness of the spec.
  \end{itemize}
  \pause
\item DeepSpecDB will be using the spec. as the index engine for the
  relational database
\end{itemize}

\end{frame}

\begin{frame}
  \begin{center}
    \Large{Thanks!}
  \end{center}

\end{frame}

\end{document}

### 2020-07-13 verif/btrees/assert.h
db_client helpers

@@ -1,0 +1,111 @@
+/*-
+ * Copyright (c) 1992, 1993
+ *	The Regents of the University of California.  All rights reserved.
+ * (c) UNIX System Laboratories, Inc.
+ * All or some portions of this file are derived from material licensed
+ * to the University of California by American Telephone and Telegraph
+ * Co. or Unix System Laboratories, Inc. and are reproduced herein with
+ * the permission of UNIX System Laboratories, Inc.
+ *
+ * Redistribution and use in source and binary forms, with or without
+ * modification, are permitted provided that the following conditions
+ * are met:
+ * 1. Redistributions of source code must retain the above copyright
+ *    notice, this list of conditions and the following disclaimer.
+ * 2. Redistributions in binary form must reproduce the above copyright
+ *    notice, this list of conditions and the following disclaimer in the
+ *    documentation and/or other materials provided with the distribution.
+ * 3. All advertising materials mentioning features or use of this software
+ *    must display the following acknowledgement:
+ *	This product includes software developed by the University of
+ *	California, Berkeley and its contributors.
+ * 4. Neither the name of the University nor the names of its contributors
+ *    may be used to endorse or promote products derived from this software
+ *    without specific prior written permission.
+ *
+ * THIS SOFTWARE IS PROVIDED BY THE REGENTS AND CONTRIBUTORS ``AS IS'' AND
+ * ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+ * IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE
+ * ARE DISCLAIMED.  IN NO EVENT SHALL THE REGENTS OR CONTRIBUTORS BE LIABLE
+ * FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
[…]

### 2020-07-13 verif/indices/client/db_client.c

@@ -1,58 +1,62 @@
 #include "db_client.h"
 #include "relation.h"
+#include "relation_mem.c"
 #include <stdlib.h>
 #include <stddef.h>
 #include <string.h>
 #include <stdio.h>
+#include "assert.h"
 
 // init all btree methods
-void* btree_create_index () {
-    return RL_NewRelation();
+DB_Cursor_T btree_create_index () {
+    return (DB_Cursor_T) RL_NewRelation();
 }
 
-void* btree_create_cursor (void* env) {
+DB_Cursor_T btree_create_cursor (Index_T env) {
     Relation_T tree = (Relation_T) env;
     BtreeCursor_T cursor = malloc(sizeof(struct BtreeCursor));
+    assert(cursor != NULL);
     cursor->btree = tree;
     cursor->cur = RL_NewCursor(tree);
-    return cursor;
+    return (DB_Cursor_T) cursor;
 }
 
-size_t btree_cardinality (void* env) {
+size_t btree_cardinality (DB_Cursor_T env) {
     BtreeCursor_T cursor = (BtreeCursor_T) env;
     return RL_NumRecords(cursor->cur);
 }
 
-Bool btree_move_to_next (void* env) {
+Bool btree_move_to_next (DB_Cursor_T env) {
     BtreeCursor_T cursor = (BtreeCursor_T) env;
     return RL_MoveToNextValid(cursor->cur);
 }
 
-Bool btree_move_to_first (void* env) {
+Bool btree_move_to_first (DB_Cursor_T env) {
     BtreeCursor_T cursor = (BtreeCursor_T) env;
     return RL_MoveToFirst(cursor->cur);
 }
 
-void* btree_go_to_key (void* env, entry key) {
+Bool btree_go_to_key (DB_Cursor_T env, Key_T key) {
     BtreeCursor_T cursor = (BtreeCursor_T) env;
-    RL_MoveToKey(cursor->cur, (key.cont).int_cont);
-    return cursor;
+    entry* k = (entry*) key;
+    goToKey(cursor->cur, k->int_cont);
+    return RL_CursorIsValid(cursor->cur);
 }
 
-Bool btree_cursor_has_next (void* env) {
+Bool btree_cursor_has_next (DB_Cursor_T env) {
     BtreeCursor_T cursor = (BtreeCursor_T) env;
     return RL_CursorIsValid(cursor->cur);
 }
 
-void* btree_put_record (void* env, entry key, const void* value) {
[…]

### 2020-07-13 verif/indices/client/db_client.h

@@ -8,7 +8,7 @@
 /*=========== ENTRIES AND ATTRIBUTES =========== */
 
 // types of entries that can appear in the table
-typedef enum domain { Int, Str } domain;
+typedef enum domain { Int = 0, Str = 1 } domain;
 
 // list of "columns" in the table
 typedef struct attribute_list {
@@ -24,49 +24,37 @@ typedef struct index_attributes {
 
 // a single value in the table
 // does this need to have size_t?
-typedef union contents {
+typedef union entry {
     size_t int_cont;
     char* str_cont;
-} contents;
-
-// an entry is an int val/string val + domain specifying which one it is
-typedef struct entry {
-    domain dom;
-    contents cont;
 } entry;
-
-/*=========== ORDERED INDEX =========== */
-
-// env should be cursor? make cursor struct
-
-typedef struct OrdIndexMtable {
-    //basic
-    void* (*create_index) ();
-    void* (*create_cursor) (void* env);
-    size_t (*cardinality) (void* env);
-    
-    //cursor
-    Bool (*move_to_next) (void* env);
-    Bool (*move_to_first) (void* env);
-    void* (*go_to_key) (void* env, entry key);
-    Bool (*cursor_has_next) (void* env);
-    
-    //insert & lookup
-    void* (*put_record) (void* env, entry key, const void* value);
-    const void* (*get_record) (void* env);
-} OrdIndexMtable;
-
-/*=========== BTREE INDEX =========== */
 
 typedef struct BtreeCursor {
     Relation_T btree;
     Cursor_T cur;
 }* BtreeCursor_T;
 
-//typedef struct BtreeIndex {
-//    OrdIndexMtable* methods;
-//    BtreeCursor* index;
-//    index_attributes attrs;
-//} BtreeIndex;
+/*=========== ORDERED INDEX =========== */
+typedef struct value_t *Value_T;
+typedef struct key_t *Key_T;
+typedef struct db_cursor_t *DB_Cursor_T;
+typedef struct index_t *Index_T;
+
+typedef struct OrdIndexMtable {
+    //basic
+    DB_Cursor_T (*create_index) ();
+    DB_Cursor_T (*create_cursor) (Index_T env);
+    size_t (*cardinality) (DB_Cursor_T env);
+    
+    //cursor
+    Bool (*move_to_next) (DB_Cursor_T env);
+    Bool (*move_to_first) (DB_Cursor_T env);
[…]

### 2020-07-13 verif/indices/client/relation_mem.c

@@ -12,7 +12,7 @@
  */
 
 #include "relation.h"
-#include <assert.h>
+#include "assert.h"
 /* #include <stdlib.h> */
 #include <string.h>
 #include <stdio.h>

### 2020-07-13 papers/luke/poplsrc19/extended_abstract.tex

\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\acmConference[POPL SRC'19]{ACM SIGPLAN Symposium on Principles of Programming
  Languages -- Student Research Competition}{January 13--19,
  2019}{Cascais/Lisbon, Portugal} \acmYear{2019}
\acmISBN{}
\acmDOI{}
\startPage{1}

\bibliographystyle{ACM-Reference-Format}
\usepackage{booktabs}
\usepackage{subcaption}
\begin{document}

\title[]{Verification of a Cache-optimized Data Structure}
\author{Yixuan Chen} \affiliation{ \institution{University of Michigan} }
\email{xlk@umich.edu}

\author{Aur\`ele Barri\`ere} \affiliation{ \institution{\'Ecole Normale
    Sup\'erieure de Rennes} } \email{aurele.barriere@ens-rennes.fr}

\author{Lennart Beringer} \affiliation{ \institution{Princeton University} }
\email{eberinge@cs.princeton.edu}

\author{Andrew W. Appel} \affiliation{ \institution{Princeton University} }
\email{appel@cs.princeton.edu}

\begin{abstract}
  Recent years have witnessed a rapid development of main-memory database
  systems thanks to the growingly affordable memory. DeepSpecDB is a main-memory
  database management system implemented in C with deep specifications and
  end-to-end verification guaranteeing the correctness of the system. In this
  paper, we will present current status of DeepSpecDB on verification of index
  data structures.
\end{abstract}

\keywords{formal verification, deep specifications, database}

\maketitle

\section{Introduction}

As the unit price of memory decreases over time, efforts have been put into
migrating the traditional disk-based applications into main memory. A classical
category of such applications are databases. For example, H-store~\cite{hstore},
MongoDB, Redis and Memcached are well-known database systems featuring
main-memory index and storage.

Meanwhile, database systems are the kind of critical applications where security
vulnerabilities can lead to catastrophic consequences. We believe that formally
verifying functional correctness of programs with respect to deep specifications
is the remedy. Deep specifications are rich, live, formal, and two-sided,
ensuring that the behavior will be captured by the specifications and proved
correctly by machine-checked proofs. With the help of the Verified Software
Toolchain~\cite{VST}, we are now able to verify C programs against the
CompCert~\cite{compcert} operational semantics. The DeepSpecDB project aims to
design and verify a main-memory database system with deep specification
guaranteeing the correctness of the system. In this paper, we present our
efforts on the indexing data structures, which databases use to organize the
data and speed up data access.

\section{Overview of the data structure}

We design our database index based on MassTree~\cite{masstree}. In general, the
data structures maps from variable-length strings to values. MassTree maintains
the mapping by organizing strings in a \textit{trie over B+ trees}: each node of
traditional trie is now replaced with a B+ tree and a trie node now corresponds
to a fixed-length slice of key rather than a single character. To account for
the possible residual suffix of keys, a data structure named \textit{border
  node} is introduced. The \textit{border node} replaces the leaf node of
original B+ tree, and can potentially point to the next layer of trie nodes or
client values corresponding to different prefixes of current slice. The length
of the slice depends on the CPU word size, so that we can express the slice as
an integer and speed up the operations. The fanout of the B+ trees as well as
other parameters of MassTree are chosen so that internal structures can fit well
in cache line which improves the performance. Although the original MassTree was
designed for key-value databases, later development of Silo~\cite{Silo} and
SiloR~\cite{SiloR} suggest its capability of indexing for relational databases.

In the original MassTree design, the \textit{border node}s are linked with each
other to facilitate range queries and removals. We decide to remove the links in
favor of a new structure called \textit{cursor} inspired by SQLite. Abstractly,
if all keys are organized in an ordered list, a \textit{cursor} should point
between two adjacent keys and any key bounded by the two can be associated with
the \textit{cursor}. Concretely, the \textit{cursor} for B+ trees is implemented
with a list of pointers pointing at various internal nodes and the
\textit{border node}, and the indexes into those nodes. The list should also be
the trace when one tries to access the \textit{border node} given any slice of
key associated with the cursor from the root node. The \textit{cursor} for a
trie is a list of B+ tree cursors, which is also the trace when one tries to
access the client value given any key associated with the cursor.

\textit{Cursor}s provide similar sequential performance of range queries to
links because one can access the next or previous record in the structure by
traversing upward and then downward with the help of \textit{cursor}s, which
should cost only amortized constant time. Moreover, the traversal should hit
cache frequently considering the locality of range queries. We favor
\textit{cursor}s over links in the data structures because we anticipate better
concurrent performance compared to implementation with links for two reasons.
First, \textit{cursor}s can be allocated as thread local objects leading to less
contentions. Second, we expect \textit{cursor}s to perform better when
overwriting update is not available (updating the links in this case is
expensive), which happens in the optimistic concurrency control variant from
Silo.

\section{Modular Coq-based verification using VST}

The verification proceeds in a modular approach. In the database, the B+ trees
are also used as the index for integer keys, and we build tries on exactly the
same interface exposed to the rest of database. We organize the verification in
the same structure as we write the program. The verification of B+
trees~\cite{aurele} abstracts the implementation into specification. Tries are
then verified given only the specification of B+ trees. In another sense, the
verification holds for any data structure that is verified with the same
specification. Furthermore, the specification for the database index is
parameterized by key types, which means the specification of tries is the same
as that of B+ trees modulo the theory of keys. This also enables us to use the
verification of trie as a test for the specification: only if the specification
is neither over-restricted nor under-specified can the proofs go through.

In another direction, the application domain proof is modularized away from the
concrete semantics in verification of both data structures. That is, C programs
are proved to correctly implement corresponding functional programs. The
refinement is established in Verifiable C using Verified Software Toolchain. The
functional programs are proved to conform to the specification, which is usually
properties about keys, values and interaction with the data structures. The
benefit of modularization is significant: we need not care about the low level
details, such as pointers and memory management when reasoning about functional
correctness. We can also forget the application logic when we are in the
separation logic world.

\section{Lessons learned}

During the verification, we also discover several methodologies that is useful
for verification of complex data structures, especially with support of
\textit{cursor}s.

\subsection{Augmented types}

When using separation logic, one needs to derive the abstract data types
corresponding to the concrete ones, where all the addresses are typically
abstracted away (for example, in the proof for list reversal~\cite{reynolds}).
However, a \textit{cursor} might point to internal nodes of a data structure
which is often achieved by saving copies of pointers in C. We find it hard to
express the same relation using abstract cursor and the abstract data structure.
One promising solution is to use magic wand~\cite{wand}, but this approach turns
out to be unsound. Therefore we introduce the augmented types, where we include
the addresses in the abstract data type in order to express the relation between
\textit{cursor} and the data structure without violating the soundness.

Augmented types, however, introduce another problem: the addresses of data
structures are usually determined by allocation at runtime and data structures
located in different address space can express the same abstract data. To
account for the non-determinism, we type our specifications for mutating
operations as predicates rather than functions. It would be ideal if we can
still justify the determinism excluding the addresses, which we will see in the
next section.

\subsection{Flattening}

Coq has a very strict type system to ensure the strict positiveness of
inductive types and termination of recursive functions. It ensures the soundness
of the system, but also adds to the difficulty to define \textit{trie over B+
  trees}: with B+ trees hidden from trie's definition, Coq cannot check the strict
positiveness and rejects the definition. We introduce the concept of
\textit{flattening} to pass the type check. In short, an extra function
\textit{flatten} is required in the specification which turns an abstract data
structure into an ordered list of key-value pairs containing the same set of
data. We can then define the data structure for trie based on \textit{flattened}
B+ trees, which can pass Coq's type check without problem.

\textit{Flattening} excels among other candidate solutions to the above problem
because it synergizes with other components of verification well. We find
similar idea in Tobias Nipkow's verification of search trees~\cite{nipkow},
where \textit{flattening} is used to verify the functional correctness. We find
it also helpful in our verification: many theorems would have been proved
separately for different data structures can now be proved once for the
\textit{flattened} list and work for all. \textit{Flattening} also implies the
determinism of data structure mutations: if the mutating operations commute with
the \textit{flatten} function, we are guaranteed the determinism by the
orderedness of the \textit{flattened} list.

\section{Current status and the future}

DeepSpecDB is still an ongoing project
(\url{https://github.com/PrincetonUniversity/DeepSpecDB}). We have almost
finished the verification for B+ trees, and basic operations for tries. However,
the current verification targets only sequential version of the database. To
achieve high performance, we are going to rewrite the program and the
verification to support concurrency. We will also benchmark different
implementations to confirm our expectation of the performance. Furthermore,
there are efforts to integrate the verification of index engines with other
components of database, such as query compilers. We expect an end-to-end
verification will be established from the semantics of SQL to the semantics of
assembly code.

%% Bibliography
\bibliography{../reference.bib}

\end{document}

### 2020-09-24 concurrency/Verification of Concurrent Binary Search Tree.tex
first-pass explanation of fine-grained locking + atomicity

@@
 \section{Correctness Proofs}
-\subsection{Locking and atomicity} 
+\subsection{Fine-grained locking and atomicity}
+
+To prove that an operation on a data structure satisfies a logically atomic specification, we must show that there are no visible intermediate states of the operation, i.e., that other threads see the data structure as unchanged until the linearization point at which the operation takes effect.
+This is often implemented with either a lock-free series of atomic memory accesses, in which case all but the last access must make changes that are considered ``invisible'', or a coarse-grained lock, in which case any changes may be made and no other thread can access the data structure until the operation is complete.
+Fine-grained locking presents an interesting middle ground between these two approaches: other threads may continue to access the data structure as long as they do not require a section that is currently locked, and the visible changes may be implemented with multiple non-atomic operations in the locked section.
+
+In an Iris-style separation logic, the lock-free approach means that each atomic operation gets access to the state of the data structure stored in an atomic shift, and must either maintain its previous state or (at the linearization point) move to a new one; in the coarse-grained approach we can store the state of the data structure in a lock invariant, and the function has access to all of it throughout the critical section.
+With fine-grained locking, neither of these approaches quite applies as-is: there are no atomic operations by which we access the state of the data structure, but neither is there a lock whose invariant controls access to the entire structure.
+However, we can take a modified form of the lock-free approach, where we treat each critical section as an atomic operation that can access the atomic shift.
+Typically, we will perform a sequence of non-atomic operations in the body of the critical section without access to the global state of the data structure, and then (before exiting the critical section) access the atomic shift and show that we have either maintained the original state or satisfied the postcondition of the operation.
+In this section, we describe a general approach to associating fine-grained locks with pieces of the abstract state of a data structure, allowing them to interact with atomic shifts and be used to prove atomic specifications.
+(Note that this approach is not novel to us; it is adapted from something Ralf Jung worked on.
+We need to be more clear about the relationship.)
+
+technical material: definition of sync\_inv, sync\_commit and sync\_rollback
+
 \subsection{Insert and Lookup}

### 2020-09-25 concurrency/Verification of Concurrent Binary Search Tree.tex
paper progress

--- 2020-06-04 concurrency/Verification of Concurrent Binary Search Tree.tex
@@
 \usepackage{url}
+\usepackage{subcaption}
+\usepackage{listings}
 
@@
 \section{Safety Proofs}
+
+\begin{figure}[htb]
+\begin{subfigure}[t]{0.45\textwidth}
+\begin{verbatim}
+
+typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
+typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
+typedef struct tree_t **treebox;
+
+void insert (treebox t, int x, void *value) {
+  struct tree_t *tgt = *t;
+  struct tree *p;
+  void *l = tgt->lock;
+  acquire(l);
+  for(;;) {
+    p = tgt->t;
+    if (p==NULL) {
+      tree_t *p1 = (struct tree_t *) surely_malloc (sizeof *tgt);
+      tree_t *p2 = (struct tree_t *) surely_malloc (sizeof *tgt);
+      p1 ->t = NULL;
+      p2 ->t = NULL;
+      lock_t *l1 = (lock_t *) surely_malloc(sizeof(lock_t));
+      makelock(l1);
+      p1->lock = l1;
+      release2(l1);
+      lock_t *l2 = (lock_t *) surely_malloc(sizeof(lock_t));
+      makelock(l2);
+      p2->lock = l2;
+      release2(l2);
+      p = (struct tree *) surely_malloc (sizeof *p);
+      tgt->t = p;
+      p->key=x; p->value=value; p->left=p1; p->right=p2;
+      release2(l);
+      return;
+    } else {
+      int y = p->key;
+      if (x<y){
+      	tgt = p->left;
+        void *l_old = l;
+        l = tgt->lock;
+        acquire(l);
+        release2(l_old);
+      } else if (y<x){
+        tgt = p->right;
+        void *l_old = l;
+        l = tgt->lock;
+        acquire(l);
+        release2(l_old);
+      }else {
+      	p->value=value;
+        release2(l);
+      	return;
+      }
+    }
+  }
+} 
+\end{verbatim} 
+\end{subfigure}
+
+\begin{subfigure}[t]{0.45\textwidth}
+ \begin{verbatim}
+
+void *lookup (treebox t, int x) {
+  struct tree *p; void *v;
+  struct tree_t *tgt;
+  tgt = *t;
+  void *l = tgt->lock;
+  acquire(l);
+  p = tgt->t;
+  while (p!=NULL) {
+    int y = p->key;
+    if (x<y){
+      tgt=p->left;
+      void *l_old = l;
+      l = tgt->lock;
+      acquire(l);
+      p=tgt->t;
+      release2(l_old);
+    }else if (y<x){
+      tgt=p->right;
+      void *l_old = l;
+      l = tgt->lock;
+      acquire(l);
+      p=tgt->t;
+      release2(l_old);
+    }else {
+      v = p->value;
+      release2(l);
+      return v;
+    }
+  }
+  release2(l);
+  return NULL;
+}
+
+\end{verbatim}
+\end{subfigure}
+\caption{Concurrent Binary Search Tree}
+\end{figure}
+
+In this section, we will focus on safety proofs of concurrent binary search tree.
+Here safety means thread-safe.
+An operation of any concurrent data-structure is thread-safe if it functions correctly during simultaneous execution by multiple threads.
+In VST, we can use $\mathsf{lock\_inv}$ assertion to assert that there exists a lock in memory with a given invariant: $\mathsf{lock\_inv}\ \mathsf{sh}\ p\ R$ means that the current thread owns share $\mathsf{sh}$ of a lock at location $p$ with invariant $R$.
+We use the term \emph{lock invariant} for $R$, which is a predicate representing the resources protected by a lock.
+The $\mathsf{lock\_inv}$ predicate is sufficient to prove the safety properties of the concurrent data-structure.
+We have taken binary search tree with fine-grained concurrent control implemented using hand-over-hand locking mechanism.
+In hand-over-hand locking mechanism, we acquire the lock for a successor before releasing the lock for a predecessor.
+The code for $\texttt{insert}$ and $\texttt{lookup}$ methods of concurrent binary search tree that we are interested to verify is presented in Figure 1.
+
 \subsection{Lock invariants for hand-over-hand locking}
+A node of CBST have the pointer to its children, so lock invariant of a node should have the knowledge about its children.   
+Whenever we release the lock for the parent node, we have to recover all resources that the parent node accessed while acquiring the lock.
+By releasing the parent node's lock, we would lose the information about the lock\_inv of the node and we wouldn't be able to release the node's lock later.
+To solve this problem, we need to use $recursive$ lock, one whose invariant includes a share of lock itself.
+In VST, we can make such invariant using $\mathsf{selflock}$ function along with the lemma $\mathsf{selflock\_eq}$: $\forall Q\ \mathit{sh}\ p, \mathsf{selflock}\ Q\ \mathit{sh}\ p = Q * \triangleright \mathsf{lock\_inv}\ \mathit{sh}\ p\ (\mathsf{selflock}\ Q\ \mathit{sh}\ p)$.
+We can define $lock\ invariants$ and $\mathsf{lock\_inv}$ predicate for $recursive$ lock $lock$ as follows: $\mathsf{lock\_inv}\ lsh1\ p\ (\mathsf{selflock}\ P\ lsh2\ lock)$ where P represents the knowledge about the children of current node, $lsh1$ and $lsh2$ are the two halves of the writable share $Ews$.
 
+\subsection{Specification and Verification}
 \section{Correctness Proofs}
+In the previous section, we showed that how to prove CBST is safe, but not it is functionally correct.
+To prove the correctness, our threads need to be able to record the information about the actions they have performed on the shared state, instead of sealing all knowledge of the shared data structure inside the lock invariant.
+We can accomplish this with ghost variables, a simple form of auxiliary state.
+Aside from the $\texttt{permission}$ on shared resources held by each thread that we used in the safety proofs, the verification of correctness properties also involves the $\texttt{ghost state}$.
+In VST, any Coq type can be used as ghost state, as long as we can describe what happens when two elements of that type are joined together.
+We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \textsf{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.
+
+Linearizability is one of the most popular strong correctness condition for the most of concurrent data-structure, which implies that every operation appears to take place atomically, in some order, consistent with the real-time ordering of those operations.
+
 \subsection{Locking and atomicity} 
+In order to record the action each thread performed as a linear history, we will define global invariants using the ghost state, which are similar to lock invariants but are not associated with any particular memory location.
+Instead, a global invariant is true before and after every step of a program, acting as a publicly accessible resource.
+A program instruction can use the contents of global invariant if it can guarantee that no one will ever see an intermediate state in which the invariant does not hold.
 \subsection{Insert and Lookup}
+
 \subsection{Delete}

### 2020-09-25 concurrency/Verification of Concurrent Binary Search Tree.tex

@@
 Linearizability is one of the most popular strong correctness condition for the most of concurrent data-structure, which implies that every operation appears to take place atomically, in some order, consistent with the real-time ordering of those operations.
-
-\subsection{Locking and atomicity} 
 In order to record the action each thread performed as a linear history, we will define global invariants using the ghost state, which are similar to lock invariants but are not associated with any particular memory location.
@@
 A program instruction can use the contents of global invariant if it can guarantee that no one will ever see an intermediate state in which the invariant does not hold.
+
+\subsection{Fine-grained locking and atomicity}
+
+To prove that an operation on a data structure satisfies a logically atomic specification, we must show that there are no visible intermediate states of the operation, i.e., that other threads see the data structure as unchanged until the linearization point at which the operation takes effect.
+This is often implemented with either a lock-free series of atomic memory accesses, in which case all but the last access must make changes that are considered ``invisible'', or a coarse-grained lock, in which case any changes may be made and no other thread can access the data structure until the operation is complete.
+Fine-grained locking presents an interesting middle ground between these two approaches: other threads may continue to access the data structure as long as they do not require a section that is currently locked, and the visible changes may be implemented with multiple non-atomic operations in the locked section.
+
+In an Iris-style separation logic, the lock-free approach means that each atomic operation gets access to the state of the data structure stored in an atomic shift, and must either maintain its previous state or (at the linearization point) move to a new one; in the coarse-grained approach we can store the state of the data structure in a lock invariant, and the function has access to all of it throughout the critical section.
+With fine-grained locking, neither of these approaches quite applies as-is: there are no atomic operations by which we access the state of the data structure, but neither is there a lock whose invariant controls access to the entire structure.
+However, we can take a modified form of the lock-free approach, where we treat each critical section as an atomic operation that can access the atomic shift.
+Typically, we will perform a sequence of non-atomic operations in the body of the critical section without access to the global state of the data structure, and then (before exiting the critical section) access the atomic shift and show that we have either maintained the original state or satisfied the postcondition of the operation.
+In this section, we describe a general approach to associating fine-grained locks with pieces of the abstract state of a data structure, allowing them to interact with atomic shifts and be used to prove atomic specifications.
+(Note that this approach is not novel to us; it is adapted from something Ralf Jung worked on.
+We need to be more clear about the relationship.)
+
+technical material: definition of sync\_inv, sync\_commit and sync\_rollback
+
 \subsection{Insert and Lookup}

### 2020-10-06 concurrency/Verification of Concurrent Binary Search Tree.tex
paper progress in technical part

@@
 % !TEX encoding = UTF-8 Unicode
+%% For double-blind review submission
+\documentclass[acmsmall,screen]{acmart}\settopmatter{printfolios=true}
+%% For single-blind review submission
+%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
+%% For final camera-ready submission
+%\documentclass[acmlarge]{acmart}\settopmatter{}
 
-\documentclass[11pt]{article} % use larger type; default would be 10pt
+%% Note: Authors migrating a paper from PACMPL format to traditional
+%% SIGPLAN proceedings format should change 'acmlarge' to
+%% 'sigplan,10pt'.
+
+%% Some recommended packages.
+\usepackage{booktabs}   %% For formal tables:
+                        %% http://ctan.org/pkg/booktabs
+\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
+                        %% http://ctan.org/pkg/subcaption
+
+\makeatletter\if@ACM@journal\makeatother
+%% Journal information (used by PACMPL format)
+%% Supplied to authors by publisher for camera-ready submission
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{1}
+\acmArticle{1}
+\acmYear{2017}
+\acmMonth{10}
+%\acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+\else\makeatother
+%% Conference information (used by SIGPLAN proceedings format)
+%% Supplied to authors by publisher for camera-ready submission
+\acmConference[OOPSLA'17]{Conference on Object-Oriented Programming Systems, Languages, and Applications}{Oct 25-27, 2017}{Vancouver, Canada}
+\acmYear{2017}
+\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+%\acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+\fi
+
+%% Copyright information
+%% Supplied to authors (based on authors' rights management selection;
+%% see authors.acm.org) by publisher for camera-ready submission
+\setcopyright{acmcopyright}
+\acmPrice{}
+\acmDOI{10.1145/3133911}
+\acmYear{2017}
+\copyrightyear{2017}
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{OOPSLA}
+\acmArticle{1}
+\acmMonth{10}
+
+%% Bibliography style
+\bibliographystyle{ACM-Reference-Format}
+%% Citation style
+%% Note: author/year citations are required for papers published as an
+%% issue of PACMPL.
+\citestyle{acmauthoryear}   %% For author/year citations
+
+\usepackage{uri}
+\usepackage{mathtools}
+\usepackage{amsthm}
+%\usepackage{amssymb}
+\usepackage{semantic}
+\usepackage{graphicx}
+\usepackage{cases}
+\usepackage{hyperref}
+\usepackage{stmaryrd}
+\usepackage{listings}
+%\usepackage{lstlangcoq}
+\renewcommand{\lstlistingname}{Figure}
+
+\clubpenalty = 10000
+\widowpenalty = 10000
+\displaywidowpenalty = 10000
+
+\lstset{language=C,basicstyle=\ttfamily,mathescape=true,columns=fullflexible}
+
+\newcommand{\TODO}[1]{\textbf{\textcolor{red}{[ TODO: #1]}}}
+%\newcommand{\boxdotright}{\!\mathrel\boxdot\joinrel\rightarrow\!}
+\newcommand{\islock}{\boxdotright}
+\newcommand{\ghost}[2]{#1 @ #2}
+\newcommand{\isaex}{\!\mathrel\odot\joinrel\rightarrow\!}
+\newcommand{\xisaex}[1]{\!\mathrel\odot\joinrel\xrightarrow{#1}\!}
+\renewcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
+\newcommand{\emp}{\mathsf{emp}}
+
+\newcommand{\us}{$\mu$s}
+
+%\newcommand{\ignore}[1]{}
+
+\hyphenation{Comp-Cert}
 
 \usepackage[utf8]{inputenc}
-
-\usepackage{geometry} % to change the page dimensions
-\geometry{a4paper} % or letterpaper (US) or a5paper or....
-% \geometry{margin=2in} % for example, change the margins to 2 inches all round
+\usepackage{geometry}
+\geometry{a4paper}
 
@@
 \usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
-\usepackage{amssymb}
-\usepackage{mathtools}
+%\usepackage{amssymb}
 \usepackage{semantic}
 \usepackage{url}
-\usepackage{subcaption}
 \usepackage{listings}
-
-\bibliographystyle{plain}% the recommended bibstyle
 
@@
 
+\catcode`\@=11
 \section{Introduction}
@@
 \section{Safety Proofs}
-
-\begin{figure}[htb]
-\begin{subfigure}[t]{0.45\textwidth}
-\begin{verbatim}
-
+\begin{figure}[htp]
+\begin{subfigure}[t]{\textwidth}
+\begin{lstlisting}[language = C]
 typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
@@
   struct tree *p;
-  void *l = tgt->lock;
+  void *l = tgt>lock;
   acquire(l);
@@
 } 
-\end{verbatim} 
+\end{lstlisting} 
 \end{subfigure}
-
-\begin{subfigure}[t]{0.45\textwidth}
- \begin{verbatim}
-
+\caption{Insert Method}
+\label{insert}
+\end{figure}     
+\begin{figure}[htp]
+\begin{subfigure}[t]{\textwidth}
+ \begin{lstlisting}
 void *lookup (treebox t, int x) {
@@
 }
-
-\end{verbatim}
+\end{lstlisting}
 \end{subfigure}
-\caption{Concurrent Binary Search Tree}
+\caption{Lookup Method}
+\label{lookup}
 \end{figure}
-
 In this section, we will focus on safety proofs of concurrent binary search tree.
@@
 \section{Correctness Proofs}
-In the previous section, we showed that how to prove CBST is safe, but not it is functionally correct.
+In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct.
 To prove the correctness, our threads need to be able to record the information about the actions they have performed on the shared state, instead of sealing all knowledge of the shared data structure inside the lock invariant.
@@
 
-Linearizability is one of the most popular strong correctness condition for the most of concurrent data-structure, which implies that every operation appears to take place atomically, in some order, consistent with the real-time ordering of those operations.
-In order to record the action each thread performed as a linear history, we will define global invariants using the ghost state, which are similar to lock invariants but are not associated with any particular memory location.
-Instead, a global invariant is true before and after every step of a program, acting as a publicly accessible resource.
-A program instruction can use the contents of global invariant if it can guarantee that no one will ever see an intermediate state in which the invariant does not hold.
+%Linearizability is one of the most popular strong correctness condition for the most of %concurrent data-structure, which implies that every operation appears to take place atomically, %in some order, consistent with the real-time ordering of those operations.
+%In order to record the action each thread performed as a linear history, we will define global %invariants using the ghost state, which are similar to lock invariants but are not associated %with any particular memory location.
+Instead, a global invariant is true before and after every %step of a program, acting as a publicly accessible resource.
+A program instruction can use the %contents of global invariant if it can guarantee that no one will ever see an intermediate state %in which the invariant does not hold.
 
@@
 \subsection{Insert and Lookup}
+In this section, we will verify an atomic specification for insert and lookup method of Binary Search Tree with fine-grained concurrency.
+In order to write the atomic specification, we need to define a predicate that describes the state of the data structure, both concrete and abstract.
+We know that abstract state of data structure corresponds to the public pre- and postcondition of an atomic specification and concrete information of the node that a thread needs to access the data structure corresponds to the private pre- and postcondition of an atomic specification.
+For instance, an atomic specification for any function might look like as follows:
+\begin{align*}\forall t.\ \langle \mathsf{nodebox\_rep}(p,g\_root)\ |\ \mathsf{tree\_rep}(g\_root, t)\rangle\ \texttt{cbst\_op}(p)\ \langle \mathsf{nodebox\_rep}(p,g\_root)\ |\ \mathsf{tree\_rep}(g\_root, t')\rangle
+\end{align*}
+where $p$ and $g\_root$ are the pointer and $gname$ representing the root node of a tree, and $t$ and $t'$ are the abstract state of a tree before and after the function's execution.
+The $nodebox\_rep$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in section 3.
+In this case, the lock invariant of recursive lock have, along with the information about the children node,$my\_half$ predicate which holds the partial information describing the current thread's contribution to the shared state (i.e.
+$tree\_rep$).
+The $tree\_rep$ predicate describes the abstract state of the data structure in terms of collection of ghost states each describing the information about separate node in the tree.
+\subsubsection{Ghost States and Range}
+The main challenge in proving the properties of concurrent data structure is to abstract the data structure which helps to reason about it in the concurrent setting.
+According to the flow interface paper, we can use the product of lower-bound and upper-bound on each node of binary tree by propagating from each node to its child the appropriate bounds on the values presents in the sub-tree.
+We use the term $range$ to represent the product of lower- and upper-bound.
+The range for root node is always $(-\infty,+\infty)$ and we can calculate the range for all nodes in the tree by enforcing each node to propagates appropriate bounds.
+For the left child, range will be lower-bound for the node(lower-bound) and a key in the node (upper-bound), while for the right child range will be a key in the node (lower-bound) and upper-bound for the node (upper-bound).
+This properties in each node help us to prove that the thread have changed the state of data-structure to the valid state.
+For instance, we confirm that a key-value have been inserted in the right place if a key is inside the range of the node.
 
+We introduce the $range$ in our proof using the ghost variables.
+This $ranges$ along with other information about the node represent the abstract state of the data-structure.
+We need another type of ghost state to track the collection of all these per-node ghost states.
+So, we have two kind of ghost state in our proof: one is per-node ghost state which product of range and node information and another is over-all ghost states which represent the set of all per-node ghost variable.
+Both type of ghost states follow the $\texttt{reference}$ pattern in which each thread holds partial information describing its contribution to the shared state, and the shared resource holds a "reference" copy that records all of the contributions.
+
+For per-node ghost state, we first define the range ghost instance as follows:
+\begin{verbatim}
+ Program Instance range_ghost : Ghost :=
+  { G := (number*number); valid g := True; Join_G a b c := c =  merge_range a b }.
+\end{verbatim}
+where number $\subseteqq (-\infty,+\infty)$ and $merge\_range$ function merges two ranges into one.
+We extend this range information with the other information about the node: key, value and $gname$ for its child.
+We define our total ghost states as follows:
+\begin{verbatim}
+Definition ghost_info : Type := (key * val * gname * gname)%type.
+Instance node_ghost : Ghost := prod_PCM range_ghost (exclusive_PCM (option ghost_info)).
+\end{verbatim}
+
+Now we can create the contribution part and reference part discussed above using $\texttt{my\_half}$ and $\texttt{public\_half}$ function in VST.
+$\texttt{my\_half}$ predicate will be used in the private pre- and post-condition of atomic specification while  $\texttt{public\_half}$ will be used in the public pre- and post-condition of atomic specification.
+
+We define the contribution and reference part of the over-all ghost state as follows:
+\begin{verbatim}
+Definition ghost_ref g r1 := ghost_reference(P := set_PCM) r1 g.
+Definition in_tree g r1 := EX sh: share, ghost_part(P := set_PCM) sh (Ensembles.Singleton r1) g.
+\end{verbatim}
+We use $ghost\_ref$ and $in\_tree$ in the public and private part of the atomic specification respectively.
+The global invariant, which we introduced in our atomic specification as $tree\_rep$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
+\begin{align*}\mathsf{tree\_rep}(T,g) \triangleq [\ \forall n. \mathsf{public\_half\textsubscript{gn}} \ (range_n* ghost\_info_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
+
+Similarly,  $\texttt{nodebox\_rep}$ predicate ties together $\texttt{my\_half}$ and $\texttt{in\_tree}$ inside the $\texttt{lock\_inv}$ assertion of each node's lock as follows:
+\begin{align*}\mathsf{nodebox\_rep}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ &\mathsf{lock\_inv\textsubscript{lock}}(\mathsf{my\_half\textsubscript{g\_root}} *\ \mathsf{in\_tree}\ g\ g\_root\ *\ \exists\ pa,\ pb,\ ga,\ gb, \\&tp\mapsto\ (pa * pb)\ *\  \mathsf{nodebox\_rep}(pa,g,ga)\ *\ \mathsf{nodebox\_rep}(pb,g,gb))\end{align*}
+
+\subsubsection{insert}
+The code for the insert method of our concurrent binary search tree is shown in Figure~\ref{insert}.
+A node has a $lock$, key, value, and pointers to the left and right child.
+Each leaf node in tree are empty node with lock.
+Whenever a thread try to insert key-value pair in the tree, it first spans the tree to find the right position for new key using hand-over-hand locking mechanism; acquire the lock for child before releasing the lock for node.
+After locating right leaf node to insert new key-value, thread creates two new empty leaf nodes, insert key-value in the current node, and link the newly created leaf nodes to the current node.
+If the key already exist in the tree, then thread simply swaps the old value associated with node with the new value keeping the child pointers as it is.
+
+The atomic specification for insert method can be written as follows:
+\begin{align*} \forall t.\ \langle &\mathsf{nodebox\_rep}(p,g\_root)\ |\ \mathsf{tree\_rep}(g\_root, t)\rangle \end{align*} 
+$$\texttt{insert}(p,k,v)$$ 
+\begin{align*}\langle \mathsf{nodebox\_rep}(p,g\_root)\ |\ \mathsf{tree\_rep}(g\_root, t[k\mapsto v])\rangle \end{align*}
+We are not guaranteed that the state of  a tree at the end of the insertion will be $t[k\mapsto v]$ where $t$ is the tree state when insert is called; rather, we know that p always holds some tree during the function's execution, and at some point the function will take that tree $t$, add $[k\mapsto v]$,and then eventually return while in the meantime another thread may have modified the state of tree from $t[k\mapsto v]$ to any other arbitrary state that the current thread do not know.
+Since we have given the specification which is strong enough to specify both safety and correctness properties of insert method, we can now verify that body of the function satisfy that specification using various VST and Iris (encoded in VST) tactics.
+To prove the loop inside insert method, we need loop-invariant as follows:
+\begin{align*} \mathsf{insert\_inv}(b, x, g\_root, range,info) \triangleq\ &\exists\ lock,\ g\_current,\ np,\ !!(x\in range)\ \&\& \ \mathsf{my\_half}(g\_current,range,info)\ *\ R\ np\ * \\&\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
+
+Where $P_p$ and $Q_p$ are $\mathsf{tree\_rep}(g\_root, t)$ and $\mathsf{tree\_rep}(g\_root, t[k\mapsto v])$ respectively.
+This loop invariant has some existential variables which characterized the state of each iteration of loop in the C code.
+We use $forward\_while$ tactic just before the loop with above $insert\_inv$ which leaves four subgoals.
+The first subgoal is to prove that the current precondition in our proof satisfies the insert invariant, the second subgoal is about type-checking the loop-test expression which is often solve automatically by $entailer!$ tactic, the third subgoal is to prove that the loop body preserves the loop invariant for that we must forward-symbolic-execute through the loop body, and final subgoal is to prove that the post condition of the loop body entails the insert invariant.
+
+Figure \ref{insertproof} shows the $insert$ function annotated with separation logic specification. 
+\begin{figure}[htp]
+\begin{subfigure}[t]{1\textwidth}
+ $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+ \vspace*{-10pt}
+\begin{lstlisting}[language = C]
+void insert (treebox t, int x, void *value) {
+  struct tree_t *tgt = *t;
+  ....
+  acquire(l);
+ \end{lstlisting}  
+ $$\left\{\begin{array}{l} \mathsf{my\_half}(g\_root,(-\infty,+\infty),info)\ *\ R\ b\ *\ \mathsf{lock\_inv}(l,lsh2,R')\ *\ \\
+ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$ $$\Rrightarrow$$  $$\left\{\begin{array}{l} insert\_inv \end{array}\right\}$$ 
+  \vspace*{-10pt}
+  \begin{lstlisting}[language = C]
+  for(;;) {
+       \end{lstlisting}   
+   $$\left\{\begin{array}{l} insert\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}!!(x\in range)\&\&\mathsf{my\_half}(g\_current,range,info)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+ \vspace*{-10pt}      
+      \begin{lstlisting}[language = C]
+    p = tgt->t;
+    if (p==NULL) {
+      tree_t *p1 = (struct tree_t *) surely_malloc (sizeof *tgt);
+      tree_t *p2 = (struct tree_t *) surely_malloc (sizeof *tgt);
+      ....
+      p = (struct tree *) surely_malloc (sizeof *p);
+      tgt->t = p;
+      p->key=x; p->value=value; p->left=p1; p->right=p2;
+           \end{lstlisting} 
+  $$\left\{\begin{array}{l} \mathsf{own}\ g1\ a1\ \mathit{pp}\ *\ \mathsf{own}\ g2\ a2\ \mathit{pp}\ *\ \mathsf{my\_half}(g\_current,range,None)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+ $$\Downarrow{\textbf{sync\_commit}}$$
+$$\left\{\begin{array}{l} \mathsf{my\_half}(g\_current,range,None)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+ \vspace*{-10pt}
+        \begin{lstlisting}[language = C]
+      release2(l);
+         \end{lstlisting}
+       $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+        \vspace*{-10pt}
+         \begin{lstlisting}[language = C]
+      return;
+    } else {
+      int y = p->key;
+      if (x<y){
+      	....
+      } else if (y<x){
+        ....
+      }else {
+      	p->value=value;
+      	\end{lstlisting} 
+$$\left\{\begin{array}{l} \mathsf{my\_half}(g\_current,range,None)*\ R\ np\ *
+\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+ $$\Downarrow{\textbf{sync\_commit}}$$
+$$\left\{\begin{array}{l} \mathsf{my\_half}(g\_current,range,None)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+
+      	\begin{lstlisting}[language = C]
+        release2(l);
+        \end{lstlisting}
+        $$\left\{\begin{array}{l}  \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+        \begin{lstlisting}[language = C] 
+      	return;
+      }
+    }
+  }
+
+} 
+\end{lstlisting}
+\end{subfigure}
+\caption{The \lstinline{insert} function annotated with separation logic specification}
+\label{insertproof}
+\end{figure}
+
+\subsubsection{lookup}
 \subsection{Delete}

### 2020-10-07 concurrency/Verification of Concurrent Binary Search Tree.tex
small formatting fixes

@@
 \makeatletter\if@ACM@journal\makeatother
-%% Journal information (used by PACMPL format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{1}
-\acmArticle{1}
-\acmYear{2017}
-\acmMonth{10}
 %\acmDOI{10.1145/nnnnnnn.nnnnnnn}
@@
 %% Supplied to authors by publisher for camera-ready submission
-\acmConference[OOPSLA'17]{Conference on Object-Oriented Programming Systems, Languages, and Applications}{Oct 25-27, 2017}{Vancouver, Canada}
-\acmYear{2017}
-\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
 %\acmDOI{10.1145/nnnnnnn.nnnnnnn}
@@
 %% see authors.acm.org) by publisher for camera-ready submission
-\setcopyright{acmcopyright}
-\acmPrice{}
-\acmDOI{10.1145/3133911}
-\acmYear{2017}
-\copyrightyear{2017}
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{OOPSLA}
-\acmArticle{1}
-\acmMonth{10}
 
@@
 \newcommand{\xisaex}[1]{\!\mathrel\odot\joinrel\xrightarrow{#1}\!}
-\renewcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
+\newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
 \newcommand{\emp}{\mathsf{emp}}
@@
 
-\title{Verifying a Binary Search Tree with Fine-Grained Locking }
+\title{Verifying a Binary Search Tree with Fine-Grained Locking}
+\author{Roshan Sharma}
+\author{Shengyi Wang}
+\author{Lennart Beringer}
 \author{William Mansky}
+
 \date{} % Activate to display a given date or no date (if empty),
@@
 By releasing the parent node's lock, we would lose the information about the lock\_inv of the node and we wouldn't be able to release the node's lock later.
-To solve this problem, we need to use $recursive$ lock, one whose invariant includes a share of lock itself.
+To solve this problem, we need to use \emph{recursive} lock, one whose invariant includes a share of lock itself.
 In VST, we can make such invariant using $\mathsf{selflock}$ function along with the lemma $\mathsf{selflock\_eq}$: $\forall Q\ \mathit{sh}\ p, \mathsf{selflock}\ Q\ \mathit{sh}\ p = Q * \triangleright \mathsf{lock\_inv}\ \mathit{sh}\ p\ (\mathsf{selflock}\ Q\ \mathit{sh}\ p)$.
-We can define $lock\ invariants$ and $\mathsf{lock\_inv}$ predicate for $recursive$ lock $lock$ as follows: $\mathsf{lock\_inv}\ lsh1\ p\ (\mathsf{selflock}\ P\ lsh2\ lock)$ where P represents the knowledge about the children of current node, $lsh1$ and $lsh2$ are the two halves of the writable share $Ews$.
+We can define \emph{lock invariants} and $\mathsf{lock\_inv}$ predicate for \emph{recursive} lock $\mathit{lock}$ as follows: $\mathsf{lock\_inv}\ \mathit{lsh1}\ p\ (\mathsf{selflock}\ P\ \mathit{lsh2}\ \mathit{lock})$ where P represents the knowledge about the children of current node, $\mathit{lsh1}$ and $\mathit{lsh2}$ are the two halves of the writable share $\mathit{Ews}$.
 
@@
 We can accomplish this with ghost variables, a simple form of auxiliary state.
-Aside from the $\texttt{permission}$ on shared resources held by each thread that we used in the safety proofs, the verification of correctness properties also involves the $\texttt{ghost state}$.
+Aside from the \emph{permission} on shared resources held by each thread that we used in the safety proofs, the verification of correctness properties also involves the \emph{ghost state}.
 In VST, any Coq type can be used as ghost state, as long as we can describe what happens when two elements of that type are joined together.
-We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \textsf{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.
+We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \emph{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.
 
@@
 For instance, an atomic specification for any function might look like as follows:
-\begin{align*}\forall t.\ \langle \mathsf{nodebox\_rep}(p,g\_root)\ |\ \mathsf{tree\_rep}(g\_root, t)\rangle\ \texttt{cbst\_op}(p)\ \langle \mathsf{nodebox\_rep}(p,g\_root)\ |\ \mathsf{tree\_rep}(g\_root, t')\rangle
+\begin{align*}\forall t.\ \langle \mathsf{nodebox\_rep}(p,g_{\mathit{root}})\ |\ \mathsf{tree\_rep}(g_{\mathit{root}}, t)\rangle\ \texttt{cbst\_op}(p)\ \langle \mathsf{nodebox\_rep}(p,g_{\mathit{root}})\ |\ \mathsf{tree\_rep}(g_{\mathit{root}}, t')\rangle
 \end{align*}
-where $p$ and $g\_root$ are the pointer and $gname$ representing the root node of a tree, and $t$ and $t'$ are the abstract state of a tree before and after the function's execution.
-The $nodebox\_rep$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in section 3.
-In this case, the lock invariant of recursive lock have, along with the information about the children node,$my\_half$ predicate which holds the partial information describing the current thread's contribution to the shared state (i.e.
-$tree\_rep$).
-The $tree\_rep$ predicate describes the abstract state of the data structure in terms of collection of ghost states each describing the information about separate node in the tree.
+where $p$ and $g_{\mathit{root}}$ are the pointer and $gname$ representing the root node of a tree, and $t$ and $t'$ are the abstract state of a tree before and after the function's execution.
+The $\mathsf{nodebox\_rep}$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in section 3.
+In this case, the lock invariant of recursive lock have, along with the information about the children node,$\mathit{my\_half}$ predicate which holds the partial information describing the current thread's contribution to the shared state (i.e.
+$\mathit{tree\_rep}$).
+The $\mathit{tree\_rep}$ predicate describes the abstract state of the data structure in terms of collection of ghost states each describing the information about separate node in the tree.
 \subsubsection{Ghost States and Range}
@@
 So, we have two kind of ghost state in our proof: one is per-node ghost state which product of range and node information and another is over-all ghost states which represent the set of all per-node ghost variable.
-Both type of ghost states follow the $\texttt{reference}$ pattern in which each thread holds partial information describing its contribution to the shared state, and the shared resource holds a "reference" copy that records all of the contributions.
+Both type of ghost states follow the \emph{reference} pattern in which each thread holds partial information describing its contribution to the shared state, and the shared resource holds a "reference" copy that records all of the contributions.
 
@@
 \end{verbatim}
-where number $\subseteqq (-\infty,+\infty)$ and $merge\_range$ function merges two ranges into one.
-We extend this range information with the other information about the node: key, value and $gname$ for its child.
+where number $\subseteqq (-\infty,+\infty)$ and $\texttt{merge\_range}$ function merges two ranges into one.
+We extend this range information with the other information about the node: key, value and $\texttt{gname}$ for its child.
 We define our total ghost states as follows:
@@
 \end{verbatim}
-We use $ghost\_ref$ and $in\_tree$ in the public and private part of the atomic specification respectively.
-The global invariant, which we introduced in our atomic specification as $tree\_rep$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
+We use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part of the atomic specification respectively.
+The global invariant, which we introduced in our atomic specification as $\mathsf{tree\_rep}$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
 \begin{align*}\mathsf{tree\_rep}(T,g) \triangleq [\ \forall n. \mathsf{public\_half\textsubscript{gn}} \ (range_n* ghost\_info_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}

### 2020-10-08 concurrency/Verification of Concurrent Binary Search Tree.tex
first pass explanation for insert and lookup

@@
 \newcommand{\xisaex}[1]{\!\mathrel\odot\joinrel\xrightarrow{#1}\!}
-\newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
+\renewcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
 \newcommand{\emp}{\mathsf{emp}}
@@
 \subsection{Fine-grained locking and atomicity}
+\label{atomicity}
 
@@
 This loop invariant has some existential variables which characterized the state of each iteration of loop in the C code.
-We use $forward\_while$ tactic just before the loop with above $insert\_inv$ which leaves four subgoals.
-The first subgoal is to prove that the current precondition in our proof satisfies the insert invariant, the second subgoal is about type-checking the loop-test expression which is often solve automatically by $entailer!$ tactic, the third subgoal is to prove that the loop body preserves the loop invariant for that we must forward-symbolic-execute through the loop body, and final subgoal is to prove that the post condition of the loop body entails the insert invariant.
+We use $\texttt{forward\_call}$ tactic just before the loop with the above $\mathsf{insert\_inv}$ which leaves four subgoals.
+The first subgoal is to prove that the current precondition in our proof satisfies the insert invariant, the second subgoal is about type-checking the loop-test expression which is often solve automatically by $\texttt{entailer!}$ tactic, the third subgoal is to prove that the loop body preserves the loop invariant for that we must forward-symbolic-execute through the loop body, and final subgoal is to prove that the post condition of the loop body entails the insert invariant.
 
-Figure \ref{insertproof} shows the $insert$ function annotated with separation logic specification. 
+Figure \ref{insertproof} shows the $\texttt{insert}$ function annotated with separation logic specification.
+The proof starts with $\mathsf{nodebox\_rep}$ and $\mathsf{atomic\_shift}$ as the precondition.
+After acquiring the lock for the root node, the thread accesses the information inside the lock invariant of the root node.
+To prove the loop body, we use the $\texttt{forward\_call}$ tactic with $\mathsf{insert\_inv}$ defined above and do the forward symbolic execution through the function body.
+The first sub-goal given by  $\texttt{forward\_call}$ is simple and proved by instantiating the existential variables $\texttt{lock}$, $\texttt{g\_current}$, and $\texttt{np}$ with $\texttt{l}$, $\texttt{g\_root}$, and $\texttt{b}$ since we know that the current node is the root node.
+The main focus of the proof will be in the third sub-goal where we have to show that the loop body preserves the loop invariant.
+Once we locate the empty node for inserting new key-value pair (i.e inside the $\texttt{if}$ clause), we open the global invariant encoded in the $\mathsf{atomic\_shift}$, create the $\texttt{gnames}$ for two new empty child nodes of current node, and prove that insertion satisfy the public postcondition $\mathsf{Q\_p}$ with the help of $\texttt{sync\_commit}$ lemma described in section \ref{atomicity}.
+This is the linearization point of the $\texttt{insert}$ operation and must be done at any point in the proof, but before releasing the current node's lock.
+Modification to the global invariant, collection of $\mathsf{public\_half}$ for all nodes, is one of the most difficult task in the proof which we accomplished by defining the separate $\emph{extract}$ lemma as follows:
+\begin{verbatim}
+Lemma extract_lemmas_for_treerep2: forall t g g_root g_in x v, tree_rep2 g g_root t * in_tree g g_in |-- 
+EX n, EX n0 ,EX o:option ghost_info, public_half g_in (n, n0, Some o) *(|==>  (EX g1 g2:gname,
+( !!(o = None /\ (check_key_exist' x (n,n0) = true)) && public_half g_in (n, n0, Some (Some(x,v,g1,g2))))
+ -* tree_rep2 g g_root (insert x v t) * my_half g1 Tsh(n, Finite_Integer x, Some(@None ghost_info))*
+ my_half g2 Tsh (Finite_Integer x, n0, Some(@None ghost_info))* in_tree g g_in * in_tree g g1 * in_tree g g2)).
+  \end{verbatim}
+With the help of $\mathsf{in\_tree}$ assertion, we locate the node where we want to insert a key-value pair for that $o$ must be $\mathsf{None}$ which means the current node is a empty node and key $x$ must be inside the range associated with the current node (i.e.
+(n,n0)).
+We proved this lemma by applying induction in $t$.
+
+Another important step in the $\texttt{insert}$ proof is the case when we find the key already exists in the tree (i.e.
+the last $\texttt{else}$ clause in C code) where the old value associated with key is overridden by new value keeping child pointers as it is.
+We augment the above $\emph{extract}$ lemma with the following clause, since we don't create new $\texttt{gnames}$ for children here, and prove the lemma in the same induction over the tree $t$.
+\begin{verbatim}
+&&(|==> (ALL g1 g2:gname, ALL (v0:val),(!!(o = Some(x,v0,g1,g2) /\ (check_key_exist' x (n,n0) = true)) && 
+public_half g_in (n, n0,Some(Some(x,v, g1,g2))))-* tree_rep2 g g_root (insert x v t) * in_tree g g_in))
+ \end{verbatim}
+
 \begin{figure}[htp]
@@
 \end{subfigure}
-\caption{The \lstinline{insert} function annotated with separation logic specification}
+\caption{The $\texttt{insert}$ function annotated with separation logic specification}
 \label{insertproof}
@@
 \subsubsection{lookup}
+
+The code for the $\texttt{lookup}$ method of our concurrent binary search tree is shown in Figure~\ref{lookup}.
+It takes the location of the root pointer and a key as the arguments.
+A thread spans the tree to find a given key using hand-over-hand locking mechanism; acquire the lock for child before releasing the lock for node.
+Once a thread finds the key in the tree, it gets the value associated with that key, release the current node's lock, and return the value.
+
+The atomic specification for $\texttt{lookup}$ method can be written as follows:
+\begin{align*} \forall t.\ \langle &\mathsf{nodebox\_rep}(p,g\_root)\ |\ \mathsf{tree\_rep}(g\_root, t)\rangle \end{align*} 
+$$\texttt{lookup}(p,k)$$ 
+\begin{align*}\langle\lambda v.\ \mathsf{nodebox\_rep}(p,g\_root)\ |\ (v = \mathsf{lookup}(p,k))\ \&\&\ \mathsf{tree\_rep}(g\_root, t)\rangle \end{align*}
+
+This specification is almost similar to the specification given to the $\texttt{insert}$ method except the content of the global invariant.
+Since we have given the specification which is strong enough to specify both safety and correctness properties of $\texttt{lookup}$ method, we can now verify that body of the function satisfy this specification similar to the verification done in $\texttt{insert}$ proof.
+To prove the loop inside the lookup method, we need loop-invariant as follows:
+\begin{align*} \mathsf{lookup\_inv}(b, x, g\_root, range,info) \triangleq\ &\exists\ lock,\ g\_current,\ np,\ !!(x\in range)\ \&\& \ \mathsf{my\_half}(g\_current,range,info)\ *\ R\ np\ * \\&\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
+
+Where $P_p$ and $Q_p$ are $\mathsf{tree\_rep}(g\_root, t)$ and $(v = \mathsf{lookup}(p,k))\ \&\&\ \mathsf{tree\_rep}(g\_root, t)$ respectively.
+The proof steps for the $\texttt{lookup}$ verification are almost similar to the steps used in $\texttt{insert}$ verification with few differences which we discuss below.
+We use $\texttt{forward\_while}$ tactic just before the loop with the $\mathsf{lookup\_inv}$ defined above which leaves four subgoals as in $\texttt{insert}$ proof.
+
+Figure \ref{lookupproof} shows the $\texttt{lookup}$ function annotated with separation logic specification.
+The proof starts with $\mathsf{nodebox\_rep}$ and $\mathsf{atomic\_shift}$ assertions as the precondition and follow the same approach as in $\texttt{insert}$ proof for the verification of loop body.
+Once we find the given key (i.e.
+the last $\texttt{else}$ clause in C code) in the tree, we need to confirm that the current state of the tree in the global invariant agree with the fact that the current node, represented by $g\_\texttt{current}$ in the global invariant, actually contains the key-value pair $(k, v)$ in the ghost state represented $g\_\texttt{current}$.
+We accomplish this by opening the global invariant encoded in the $\mathsf{atomic\_shift}$, and proving that lookup satisfy the public postcondition $\mathsf{Q\_p}$ with the help of $\texttt{sync\_rollback}$ lemma described in section \ref{atomicity}.
+This is the linearization point of the $\texttt{lookup}$ operation and must be done before releasing the current node's lock.
+To locate the current node (i.e.$g\_\texttt{current}$) in the global invariants, we again augment the right hand-side of $\emph{extract}$ lemma used in the $\texttt{insert}$ proof with the following clause:
+\begin{verbatim}
+ && ( public_half g_in (n, n0, Some o) -* (tree_rep2 g g_root t * in_tree g g_in ))
+ \end{verbatim}
+
+\begin{figure}[htp]
+\begin{subfigure}[t]{1\textwidth}
+ $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+ \vspace*{-10pt}
+\begin{lstlisting}[language = C]
+void *lookup (treebox t, int x) {
+  struct tree *p; void *v;
+  struct tree_t *tgt;
+  tgt = *t;
+  void *l = tgt->lock;
+  acquire(l);
+  p = tgt->t;
+ \end{lstlisting}  
+ $$\left\{\begin{array}{l} \mathsf{my\_half}(g\_root,(-\infty,+\infty),info)\ *\ R\ b\ *\ \mathsf{lock\_inv}(l,lsh2,R')\ *\ \\
+ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$ $$\Rrightarrow$$  $$\left\{\begin{array}{l} lookup\_inv \end{array}\right\}$$ 
+  \vspace*{-10pt}
+  \begin{lstlisting}[language = C]
+    while (p!=NULL) {
+       \end{lstlisting}   
+   $$\left\{\begin{array}{l} lookup\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}!!(x\in range)\&\&\mathsf{my\_half}(g\_current,range,info)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+ \vspace*{-10pt}      
+      \begin{lstlisting}[language = C]
+    if (x<y){
+      tgt=p->left;
+      ....
+    }else if (y<x){
+      tgt=p->right;
+     ....
+    }else {
+    v = p->value;
+           \end{lstlisting} 
+  $$\left\{\begin{array}{l} \mathsf{my\_half}(g\_current,range,info)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+ $$\Downarrow{\textbf{sync\_rollback}}$$
+$$\left\{\begin{array}{l} \mathsf{my\_half}(g\_current,range,info)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+ \vspace*{-10pt}
+        \begin{lstlisting}[language = C]
+      release2(l);
+         \end{lstlisting}
+       $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+        \vspace*{-10pt}
+         \begin{lstlisting}[language = C]
+       return v;
+    }
+  }
+  release2(l);
+  return NULL;
+  }
+ \end{lstlisting} 
+\end{subfigure}
+\caption{The $\texttt{lookup}$ function annotated with separation logic specification}
+\label{lookupproof}
+\end{figure}
+
 \subsection{Delete}

### 2021-04-13 concurrency/Verification of Concurrent Binary Search Tree.tex
Add two paragraphs.

@@
 \newcommand{\xisaex}[1]{\!\mathrel\odot\joinrel\xrightarrow{#1}\!}
-\renewcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
+%% \newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
 \newcommand{\emp}{\mathsf{emp}}
@@
 \begin{document}
+\acmConference{Some conference}
 \maketitle
@@
 \subsection{Atomic Specifications}
+Atomic Hoare triples were introduced in TaDA~\cite{TaDA}.
 
@@
 \end{align*}
-where $p$ and $g_{\mathit{root}}$ are the pointer and $gname$ representing the root node of a tree, and $t$ and $t'$ are the abstract state of a tree before and after the function's execution.
+where $p$ and $g_{\mathit{root}}$ are the pointer and $\mathit{gname}$ representing the root node of a tree, and $t$ and $t'$ are the abstract state of a tree before and after the function's execution.
 The $\mathsf{nodebox\_rep}$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in section 3.
@@
 The global invariant, which we introduced in our atomic specification as $\mathsf{tree\_rep}$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
-\begin{align*}\mathsf{tree\_rep}(T,g) \triangleq [\ \forall n. \mathsf{public\_half\textsubscript{gn}} \ (range_n* ghost\_info_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
+\begin{align*}\mathsf{tree\_rep}(T,g) \triangleq [\ \forall n.
+\mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
 
@@
 \end{verbatim}
+
+In the rest two cases, the $\texttt{insert}$ method compares the key
+of the current node with the key $\texttt{x}$ and goes down into the
+corresponding left or right branch. We need to show that the range in
+the ghost state shrinks correctly. To achieve this goal we open the
+global invariant encoded in the $\mathsf{atomic\_shift}$ to retrieve
+the necessary information. But this time these operations maintain the
+original state. For the left branch case, we accomplish the proof by
+the following lemma:
+\begin{verbatim}
+Lemma in_tree_left_range:
+  forall (B: Type) (b: tree -> B -> mpred) (Q : B → mpred) (x x0: Z) (g g_root : gname)
+    (inv_names : invG) (v: val) (g_in ga gb: gname) (r a: node_info),
+    check_key_exist' x r.1 = true -> r.2 = Some (Some (x0, v, ga, gb)) -> x < x0 ->
+    atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
+                 b Q * my_half g_in gsh1 r * in_tree g g_in * my_half ga gsh1 a
+    |-- atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
+    b Q * my_half g_in gsh1 r *
+    (EX ba, !! (less_than_equal ba r.1.1 = true /\
+                range_inclusion a.1 (ba, Finite_Integer x0) = true) &&
+            (in_tree g g_in * my_half ga gsh1 (ba, Finite_Integer x0, a.2))).  
+\end{verbatim}
+The lemma for the right branch case is similar. Both use the
+$\texttt{sync\_rollback}$ lemma described in section \ref{atomicity}.
 
@@
 This is the linearization point of the $\texttt{lookup}$ operation and must be done before releasing the current node's lock.
+A critical difference from $\texttt{insert}$ is that the $\texttt{lookup}$ method does not change the shape the tree.
+With the help of the $\texttt{sync\_commit\_same}$, we need no longer the heavyweight $\emph{extract}$ lemma but instead a simpler $\emph{ramif}$ lemma to locate the current node in the global invariants:
+\begin{verbatim}
+Lemma ghost_tree_rep_public_half_ramif: forall tg g_root r_root g_in,
+Ensembles.In (find_ghost_set tg g_root) g_in -> ghost_tree_rep tg g_root r_root |-- 
+EX r: node_info, !! (range_info_in_tree r r_root tg) && (public_half g_in r * 
+(public_half g_in r -* ghost_tree_rep tg g_root r_root)).
+\end{verbatim}
+
 To locate the current node (i.e.$g\_\texttt{current}$) in the global invariants, we again augment the right hand-side of $\emph{extract}$ lemma used in the $\texttt{insert}$ proof with the following clause:
@@
 
+%% Bibliography
+\bibliography{sources}
+
 \end{document}

### 2021-04-13 concurrency/Verification of Concurrent Binary Search Tree.tex
added more content in writting

--- 2020-10-07 concurrency/Verification of Concurrent Binary Search Tree.tex
@@
 \usepackage{listings}
+\usepackage{iris}
 %\usepackage{lstlangcoq}
@@
 \newcommand{\islock}{\boxdotright}
-\newcommand{\ghost}[2]{#1 @ #2}
 \newcommand{\isaex}{\!\mathrel\odot\joinrel\rightarrow\!}
 \newcommand{\xisaex}[1]{\!\mathrel\odot\joinrel\xrightarrow{#1}\!}
-\newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
+%% \newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
 \newcommand{\emp}{\mathsf{emp}}
 
+\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
+\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
+\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
+\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
+\newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
 \newcommand{\us}{$\mu$s}
@@
 %\usepackage{amssymb}
-\usepackage{semantic}
 \usepackage{url}
@@
 \section{Introduction}
+Modern CPU design trends have shifted towards multicore processing over the last two decades.
+This is because the computing power of computers has reached its limit in single-threaded execution, while the quantity of transistors per single-core CPU are increasing as described by Moore's law \cite{591665}.
+CPUs have multiple cores now because that is the only way to get more transistors per chip, and achieve a desirable throughput.
+However, doubling the quantity of cores on one chip does not immediately imply an increase in the performance by a factor of two.
+Programs that we aim to run in multiple cores have to be carefully designed to efficiently divide the workload into threads, and facilitate non-conflicting access to the shared data by multiple threads in parallel.
+These challenges is resolved by designing the concurrent data structures, algorithms that provide thread-safe access to the shared data by storing and organizing data in a standard way.
+Concurrent data structures are now widely employed in real-world applications like database algorithms, file systems, networking, e-commerce, etc.
+Designing these data structures has been considered as one of the difficult tasks in the software development fields.
+Despite the careful design, implementation, and thorough unit testing of these algorithms, it is very likely to expect subtle bugs and unexpected behaviors of the program.
+This is because threads interact with each other in complex and subtle ways such that it is almost impossible to cover every corner case while testing.
+Therefore, it is important to use a precise way of proving the correctness of concurrent data structures which provides a guarantee that the given algorithm correctly implements the specified concurrent data structures.
 
+Formal verification is an active field of research that aims to prove or disprove the correctness of algorithms with respect to a certain formal specification, using mathematical techniques.
+Model checking and theorem proving  (also called deductive reasoning) are the two main approaches in this field.
+Model checking builds the mathematical model of a system as a collection of states and actions that move from one state to another.
+The model checker then outputs yes if the given model meets the given specifications and produces a counterexample otherwise, after a systematic and exhaustive exploration of a mathematical model.
+Model checking approach only works for finite models and certain infinite models that can be easily abstracted into finite sets of states.
+While it is a simple and automated operation, it does not scale well to large and complex systems where the search spaces are typically too large to be explored in a reasonable amount of time with a reasonable amount of space.
+Theorem proving, on the other hand, is a more powerful technique than model checking which can deal with an infinite state space.
+It works by generating a collection of mathematical proof obligations from the system and its specifications and proving these obligations using either proof assistants (such as Coq, Isabelle, PVS, HOL) or using automatic theorem provers (such as SMT solvers).
+Theorem proving allows us to work on more accurate representations of the system and express any properties, but most proofs have to be done manually which requires time and expertise.
+Since the binary search tree that we target for verification is a concurrent data structure of unbounded key-value pairs, and processes.
+The concurrency considerably increases the state space, and we also want to work with a model that's connected to the actual behavior of C programs, while model checking usually works on more abstract models.
+So, we use the theorem proving approach to formalize the proof of correctness.
+
+In this thesis, we describe the formal verification of the correctness of concurrent binary search tree with fine-grained locking.
+We implement the BST algorithms in C and use the Verified Software Toolchain(VST) \cite{10.5555/2670099} to verify with respect to given specification.
+The concurrent separation logic (CSL) is employed to reason about behavior of the program implemented in the C programming language.
+Separation logic (SL) \cite{1029817} is an extension of Hoare logic \cite{10.1145/363235.363259}, way of reasoning about programs using a set of logical rules.
+SL helps to locally reason about programs that play with pointer data structures, where specifications and proofs of a program component only reflect the chunk of memory that the component uses, not the entire global state of the system.
+Concurrent separation logic (CSL) is a variant of the separation logic for reasoning about the concurrent programs.
+The version of CSL formalized in VST is rooted on the work of Hobor et al.
+\cite{10.1007/978-3-540-78739-6_27}, but extended with higher-order ghost state in the style of Iris \cite{10.1145/2951913.2951943}.
 \section{Background}
 \subsection{VST and Iris}
+We use the Verified Software Toolchain \cite{10.5555/2670099}  to prove the correctness of the concurrent binary search tree.
+VST is a Coq-based proof system that helps to prove separation logic properties of C programs with the assurance that the properties will be preserved in compiled code.
+First, we write programs in C programming and using CompCert’s \emph{clightgen}, the ASTs for C programs can be generated in Coq.
+We then state and prove specifications for them using Verifiable C.
+Verifiable C is a higher-order separation logic for reasoning about the functional correctness of C programs and carries a soundness proof that links high-level specifications to assembly code generated by the CompCert Verified compiler.
+Verifiable C supports reasoning about two separate kinds of concurrency:  Pthreads-style concurrency with locks (used by implementing blocking data structures that can be fine-grained or coarse-grained ), and concurrency using C11 atomic operations (i.e.
+lock-free data structures).
+
+\subsection{Ghost State and Global Invariants}
+We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program.
+But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness.
+For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads.
+But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.
+\begin{figure}[htb]
+\centering
+$\texttt{x = 0;}$\\
+$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\hspace{30px}\pi = \pi_1\ .\ \pi_2$\\
+$\begin{array}{l || l}
+\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
+\texttt{acquire(l);} & \texttt{acquire(l);}\\
+\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
+\texttt{x++;} & \texttt{x++;}\\
+\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
+\texttt{release(l);} & \texttt{release(l);}\\
+\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
+\end{array}$\\
+$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
+\caption{The increment example annotated with separation logic assertion}
+\label{figure1}
+\end{figure}
+We must preserve the connection between the value of shared resources and the work performed by each thread.
+A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread.
+They do not appear in the original program.
+Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}.
+In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together.
+VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} (\textsf{gname} in Coq) associated with the value $a$, which may be of any type.
+We will see different kind of ghost states used to verify binary search tree in Chapter \ref{chapter4}.
+
+The main idea behind the logic from Iris \cite{10.1145/3022670.2951943}, a mechnized higher-order concurrent separation logic framework, is the construction of \emph{global invariant} as \emph{ghost state}.
+\emph{Global Invariant} is an invariant on the global (ghost and physical) state of program.
+We can open any invariant but must close it again before taking any steps of execution unless those execution are atomic.
+Thread can use the contents of \emph{global invariant} during atomic operation if it can guarantee that no one will ever see an intermediate state in which invariant does not hold.
+The rules from Jung et al.\cite{10.1145/3022670.2951943} for creating and opening invariants are:
+
+$$\inference[\textsf{inv\_alloc}]{}{\triangleright P \vdash \pvs[E] \mathsf{EX}\ i : \mathsf{iname}, \knowInv{i}{P}}$$
+$$\inference[\textsf{inv\_open}]{i \in E}{\knowInv{i}{P} \vdash \pvs[E][E \setminus i] \triangleright P * (\triangleright P \wand \pvs[E \setminus i][E] \mathsf{emp})}$$
+
+where invariants are provided in the form of assertion  $\knowInv{i}{P}$, and states that $P$ is maintained as an invariant on the global state with name $i$.
+The operator $\pvs[][]$ is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e.
+$\knowInv{i}{..\knowInv{i}{P}..}$).
 \subsection{Atomic Specifications}
+\emph{Linearizability} is one of the most popular traditional correctness condition for concurrent data structures.
+It is defined in terms of a set of high-level operations, such as insert and remove for a set; if a data structure is linearizable, any operation performed on it tends to happen atomically and in some order that corresponds to the real-time ordering of such operations.
+For instance, if operation $A$ is completed before operation $B$, then $B$ should logically follow $A$.
+We can prove the linearizability property of concurrent data structures using \emph{global invariants}.
+Suppose we have an invariant that said ``there exists a linear history of operations performed on the data structures, stored in some ghost state, and the current  state of the data structure is the one that would result from applying each of those operations in order".
+Then this directly implies that data structure is linearizable.
+We can allow a data structure operation to access the content of global invariants, treating that operation as an atomic operation if we are able prove that that operation \emph{appears to execute atomically} at some linearization point.
+This idea was introduced in the TaDA logic \cite{tada} in the form of \emph{atomic Hoare triples} as shown below, and called as  \emph{Logical atomicity}.
+$$\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$
+
+where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of shared data structure.
+$P_l$ and $Q_l$ are \emph{local} to each thread
+and do not interfere with global state of a data structure, while $P_p$ and $Q_p$ are public to all threads and must satisfy following rule: $P_p$ must be true at every point from the beginning of the function until the linearization point, for some abstract value $a$, and $Q_p$ must be true at the lineariztion point, but may not still true by the end of the function.
+So, at the linearization point the operation atomically transitions from $P_p$ to $Q_p$.
+
+VST has encoded an atomic Hoare triple, and provide the way that matches the notation VST uses for normal specifications.
+Such specification is called \emph{atomic specification}, and can be formalized in Coq as shown in  \ref{atomic_spec}.
+This is how we write pre- and postcondition in VST.
+\begin{figure}[htb]
+\centering
+\begin{verbatim}
+Program Definition insert_spec :=
+DECLARE _insert
+ATOMIC TYPE W OBJ a INVS Ei Eo
+WITH ...
+PRE [ ... ]
+  PROP (...)
+  LOCAL (...)
+  SEP (P_l) | (P_p)
+POST [ ... ]
+  PROP ()
+  LOCAL ()
+  SEP (Q_l) | (Q_p)
+\end{verbatim}
+\caption{Atomic Specification in VST}
+\label{atomic_spec}
+\end{figure}
+\texttt{W} is the \textsf{TypeTree} representing the type of arguments passed with \texttt{WITH} clause; \texttt{a} is the abstract state for the triple; \texttt{Ei} and \texttt{Eo} are the sets of invariants names inside and outside the triple.
+The \texttt{PROP} clause describes things that that are true independent of program state, the \texttt{LOCAL} clause describes the values contained in C local variables, and the \texttt{SEP} clause represents the \emph{separating conjunction} (*) of \emph{spatial predicates}, predicates on some part of the memory.
+In VST, while proving that a function implements an atomic specification, the precondition will contains an $\mathsf{atomic\_shift}$ assertion with the public pre- and postcondition, and the masks inside it.
+This atomic shift can be accessed through following two rules:
+$$\inference[\textsf{atomic\_commit}]{\forall a, R * P_p\ a \Rrightarrow \mathsf{EX}\ y,\ Q_p\ a\ y * R'\ y}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
+$$\inference[\textsf{atomic\_rollback}]{\forall a, R * P_p\ a \Rrightarrow P_p\ a * R'}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R'}$$
+During commit, we must provide resources $R$ which, in combination with the public precondition $P_p$, allow us to prove the public postcondition $Q_p$.
+Then, we gain access to an assertion $Q$ required by the postcondition of the function, and leftover resources $R'$.
+During rollback, we provide resources $R$ which, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as leftover resources $R'$.
+The rollback is specially used to learn some relationship between pieces of information (e.g.
+ghost state) stored in the public precondition, while the commit is used to prove the public postcondition from the precondition, and obtain an assertion $Q$.
+To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of roolbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.
 
 \section{Safety Proofs}
+\label{safety}
 \begin{figure}[htp]
@@
 An operation of any concurrent data-structure is thread-safe if it functions correctly during simultaneous execution by multiple threads.
+A thread can update the data structure if it holds a sufficiently large share of the data structure so that no other thread can possibly read it.
+If we want to have a data structure that can be modified by multiple threads, we must move shares between threads via locks.
 In VST, we can use $\mathsf{lock\_inv}$ assertion to assert that there exists a lock in memory with a given invariant: $\mathsf{lock\_inv}\ \mathsf{sh}\ p\ R$ means that the current thread owns share $\mathsf{sh}$ of a lock at location $p$ with invariant $R$.
@@
 The $\mathsf{lock\_inv}$ predicate is sufficient to prove the safety properties of the concurrent data-structure.
-We have taken binary search tree with fine-grained concurrent control implemented using hand-over-hand locking mechanism.
+
+%We have taken binary search tree with fine-grained locking.
 In hand-over-hand locking mechanism, we acquire the lock for a successor before releasing the lock for a predecessor.
@@
 Whenever we release the lock for the parent node, we have to recover all resources that the parent node accessed while acquiring the lock.
-By releasing the parent node's lock, we would lose the information about the lock\_inv of the node and we wouldn't be able to release the node's lock later.
-To solve this problem, we need to use \emph{recursive} lock, one whose invariant includes a share of lock itself.
-In VST, we can make such invariant using $\mathsf{selflock}$ function along with the lemma $\mathsf{selflock\_eq}$: $\forall Q\ \mathit{sh}\ p, \mathsf{selflock}\ Q\ \mathit{sh}\ p = Q * \triangleright \mathsf{lock\_inv}\ \mathit{sh}\ p\ (\mathsf{selflock}\ Q\ \mathit{sh}\ p)$.
-We can define \emph{lock invariants} and $\mathsf{lock\_inv}$ predicate for \emph{recursive} lock $\mathit{lock}$ as follows: $\mathsf{lock\_inv}\ \mathit{lsh1}\ p\ (\mathsf{selflock}\ P\ \mathit{lsh2}\ \mathit{lock})$ where P represents the knowledge about the children of current node, $\mathit{lsh1}$ and $\mathit{lsh2}$ are the two halves of the writable share $\mathit{Ews}$.
+By releasing the parent node's lock, we would lose the information about the lock\_inv of the node and we would not be able to release the node's lock later.
+To solve this problem, we use \emph{recursive} lock, one whose invariant includes a share of lock itself so that we can retain lock\_inv assertion for a node even after releasing its parent's lock.
+In VST, we can make such an invariant using $\mathsf{selflock}$ function along with the lemma $\mathsf{selflock\_eq}$: $\forall Q\ \mathit{sh}\ p, \mathsf{selflock}\ Q\ \mathit{sh}\ p = Q * \mathsf{lock\_inv}\ \mathit{sh}\ p\ (\mathsf{selflock}\ Q\ \mathit{sh}\ p)$.
+We can define $\mathsf{lock\_inv}$ predicate for \emph{recursive} lock $\mathit{lock}$ as follows: $\mathsf{lock\_inv}\ \mathit{sh1}\ p\ (\mathsf{selflock}\ P\ \mathit{sh2}\ \mathit{lock})$ where $\mathit{P}$ represents the knowledge about the children of current node, $\mathit{sh1}$ and $\mathit{sh2}$ are the two halves of the writable share.
 
 \subsection{Specification and Verification}
+We begin the verification by defining a predicate $\mathsf{nodebox\_rep}(p)$ that describes the concrete representation of the tree, where $p$ is the pointer to the root node.
+The $\texttt{nodebox\_rep}$ predicate ties together $\texttt{lock\_inv}$ assertion for each recursive lock as follows:
+\begin{align*}
+ &\mathsf{nodebox\_rep}_{\pi}(p) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \exists\ pa\ pb\ \ tp\mapsto\ (pa * pb)\ *\  \mathsf{nodebox\_rep}_{0.5}(pa)\ *\ \mathsf{nodebox\_rep}_{0.5}(pb)  \end{align*}
+
+ The ownership of each node's lock is divided into two halves: one half is owned by \emph{lock\ invariant} and another half is owned by node itself.
+But, in case of root node, we need to split one half into multiple pieces, and share those pieces to all threads such that all of them initially have the knowledge of root node.
+
 \section{Correctness Proofs}
@@
 We need to be more clear about the relationship.)
+ We use the following rules to associate fine-grained locks with abstract state of a data structure:
+\begin{align*}
+    \textsf{sync\_inv}(g,sh,R) \triangleq  \exists\ a,\ R\ g\ a\ *\ \textsf{my\_half}(g,sh,a)   \end{align*}
+$$\inference[\textsf{sync\_commit}]{\forall a, R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * \exists x_0'\ x_1'.\ (x_0, x_1) \leadsto (x_0', x_1') \land \\(\mathsf{my\_half}(g, \mathit{sh}, x_0') * \mathsf{public\_half}(g, x_1') \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) *  \textsf{sync\_inv}(g,sh,R) \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
+$$\inference[\textsf{sync\_rollback}]{\forall a, R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{public\_half}(g, x_1) \Rrightarrow P_p\ a * R')}{\begin{array}{c}\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{sync\_inv}(g,sh,R)  \Rrightarrow \\\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{sync\_inv}(g,sh,R')\end{array}}$$
+
+The $\mathsf{sync\_inv}$ predicate associates the contribution part of ghost state $a$ for a node with ghost name $g$, along with the other concrete information about a node parameterized on $a$.
+In the atomic specification, we protect the $\mathsf{sync\_inv}$ using lock inside lock invariant.
+So, once we acquire the lock, we get access to $\mathsf{sync\_inv}$.
+The $\mathsf{sync\_commit}$ and $\mathsf{sync\_rollback}$, specialized with ghost state with \emph{reference pattern} ($\mathsf{my\_half}$ and $\mathsf{public\_half}$), are the versions of $\mathsf{atomic\_commit}$ and $\mathsf{atomic\_rollback}$ described in Chapter \ref{chapter2}.
+In a commit, we must provide $\mathsf{sync\_inv}$ with resources $R$ that, in combination with the public precondition, allow us to prove the public postcondition; we then gain access to an assertion Q required by the postcondition of the function, and leftover resources $R'$.
+The state of the node $g$ in public precondition (described by an assertion $\mathsf{public\_half}$) must be joinable with the state described by an assertion $\mathsf{sync\_inv}$.
+In a rollback, we provide $\mathsf{sync\_inv}$ with resources $R$ that, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as $\mathsf{sync\_inv}$ with the leftover resources $R$.
+We must use these rules before releasing the node's lock (we lose the $\mathsf{sync\_inv}$ after releasing the lock).
 
 technical material: definition of sync\_inv, sync\_commit and sync\_rollback
-
-\subsection{Insert and Lookup}
-In this section, we will verify an atomic specification for insert and lookup method of Binary Search Tree with fine-grained concurrency.
-In order to write the atomic specification, we need to define a predicate that describes the state of the data structure, both concrete and abstract.
-We know that abstract state of data structure corresponds to the public pre- and postcondition of an atomic specification and concrete information of the node that a thread needs to access the data structure corresponds to the private pre- and postcondition of an atomic specification.
-For instance, an atomic specification for any function might look like as follows:
-\begin{align*}\forall t.\ \langle \mathsf{nodebox\_rep}(p,g_{\mathit{root}})\ |\ \mathsf{tree\_rep}(g_{\mathit{root}}, t)\rangle\ \texttt{cbst\_op}(p)\ \langle \mathsf{nodebox\_rep}(p,g_{\mathit{root}})\ |\ \mathsf{tree\_rep}(g_{\mathit{root}}, t')\rangle
-\end{align*}
-where $p$ and $g_{\mathit{root}}$ are the pointer and $gname$ representing the root node of a tree, and $t$ and $t'$ are the abstract state of a tree before and after the function's execution.
-The $\mathsf{nodebox\_rep}$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in section 3.
-In this case, the lock invariant of recursive lock have, along with the information about the children node,$\mathit{my\_half}$ predicate which holds the partial information describing the current thread's contribution to the shared state (i.e.
-$\mathit{tree\_rep}$).
-The $\mathit{tree\_rep}$ predicate describes the abstract state of the data structure in terms of collection of ghost states each describing the information about separate node in the tree.
+\subsection{Specification of the algorithm}
+In addition to program state, the verification of concurrent program involves two aspects which do not appear in actual C program: the \emph{shares} on shared resources and \emph{ghost state} described in Section \ref{safety}.
+We need to extend our specification in Section\ref{safety} with \emph{ghost state} to prove the functional correctness.
 \subsubsection{Ghost States and Range}
-The main challenge in proving the properties of concurrent data structure is to abstract the data structure which helps to reason about it in the concurrent setting.
-According to the flow interface paper, we can use the product of lower-bound and upper-bound on each node of binary tree by propagating from each node to its child the appropriate bounds on the values presents in the sub-tree.
-We use the term $range$ to represent the product of lower- and upper-bound.
-The range for root node is always $(-\infty,+\infty)$ and we can calculate the range for all nodes in the tree by enforcing each node to propagates appropriate bounds.
-For the left child, range will be lower-bound for the node(lower-bound) and a key in the node (upper-bound), while for the right child range will be a key in the node (lower-bound) and upper-bound for the node (upper-bound).
+ The main challenge in proving the properties of concurrent data structure is to abstract the data structure in a way that helps to reason about it in the concurrent setting.
+Krishna et al.
+\cite{krishna2017flow} mention in their paper that we can abstract binary search tree using appropriate bounds in each node: we can use the product of \emph{lower-bound} and \emph{upper-bound} on each node of binary tree by propagating from each node to its child the appropriate bounds on the values presents in the sub-tree.
+The \emph{lower-bound} and \emph{upper-bound} help us to prove that we perform the operation in the right node: for instance, if we $\mathsf{insert}$ a key in the tree, then key should be inside those two bounds.
+We use the term \emph{range} to represent the product of lower- and upper-bound.
+\ref{range_bst} shows an example of BST with each node labeled with appropriate bound propagated all the way from \emph{root} to empty \emph{leaf} nodes.
+\begin{figure}[htb]
+\centering
+\includegraphics[width=70mm,scale=0.5]{FIG/range_prop.png}
+\caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
+\label{range_bst}
+\end{figure}
+The range for root node is always $(-\infty,+\infty)$ and then, we can calculate the range for all nodes in the tree by enforcing each node to propagates appropriate bounds.
+For the left child, range will be lower-bound for the node (lower-bound) and the key in the node (upper-bound), while for the right child range will be a key in the node (lower-bound) and upper-bound for the node (upper-bound).
 This properties in each node help us to prove that the thread have changed the state of data-structure to the valid state.
-For instance, we confirm that a key-value have been inserted in the right place if a key is inside the range of the node.
-
+For instance, we can confirm that a key has been inserted in the right place if a key is inside the range of the node.
+Suppose we want to insert a key $38$ in \ref{range_bst}, then we will locate \emph{right} child of $35$ as the right place for $38$.
+During insertion, we can confirm that the node is indeed right place for $38$ by checking that $38$ is in the range (35,40).
 We introduce the $range$ in our proof using the ghost variables.
 This $ranges$ along with other information about the node represent the abstract state of the data-structure.
-We need another type of ghost state to track the collection of all these per-node ghost states.
-So, we have two kind of ghost state in our proof: one is per-node ghost state which product of range and node information and another is over-all ghost states which represent the set of all per-node ghost variable.
-Both type of ghost states follow the \emph{reference} pattern in which each thread holds partial information describing its contribution to the shared state, and the shared resource holds a "reference" copy that records all of the contributions.
-
-For per-node ghost state, we first define the range ghost instance as follows:
+We have used two kind of ghost state in our proof: \emph{per-node} ghost state and \emph{over-all} ghost state.\\
+\textbf{Per-node Ghost State}: The per-node ghost state is the product of the range and other information about node (key, value, and ghost name for its child nodes), created for each node.
+In VST, to use any Coq type as a ghost state, we need to create an instance of \texttt{Ghost} typeclass, \emph{separation algebra} with associated \texttt{join} relation, with an additional \texttt{valid} predicate marking those elements of the algrebra that can be used in ghost assertions.
+\texttt{Ghost} typeclass is related to the PCMs described in \ref{chapter2}.
+For instance, ghost instance for range can be created as follows:
 \begin{verbatim}
@@
 \end{verbatim}
-where number $\subseteqq (-\infty,+\infty)$ and $\texttt{merge\_range}$ function merges two ranges into one.
+Where \texttt{number} $\in (-\infty,+\infty)$ and $\texttt{merge\_range}$ is a function that merges two ranges into one.
 We extend this range information with the other information about the node: key, value and $\texttt{gname}$ for its child.
-We define our total ghost states as follows:
+We define our total ghost state as follows:
 \begin{verbatim}
@@
 \end{verbatim}
+The \texttt{exclusive\_PCM} is an exlusive algebra, which means that \texttt{Some} only joins with \texttt{None}, not with another \texttt{Some} (even the same one).
+Now, we use ghost assertion to represent this instance.
+This ghost state follow \emph{reference} pattern: each thread holds partial information describing its contribution to the shared state, and the shared resources hold a "reference" copy that records all of the contributions.
+VST provides \texttt{my\_half} and \texttt{public\_half} assertions for this: contribution element $\mathsf{my\_half}\ \mathit{g}\ \mathit{sh}\ a$ and reference element $\mathsf{public\_half}\ \mathit{g}\ r$.
+To join two elements, we combine the shares and values of the contributions, and require that the elements contain at most one reference between them to ensure the uniqueness of the reference value.
+When \texttt{my\_half} element has the full share, it is guaranteed to be equal to the reference element.
+In our case, \texttt{my\_half} assertion will be local to a thread (that means it will stay in private pre- and postcondition of atomic specification), and  \texttt{public\_half} will be shared with all threads (that means public pre-and postcondition).
 
-Now we can create the contribution part and reference part discussed above using $\texttt{my\_half}$ and $\texttt{public\_half}$ function in VST.
-$\texttt{my\_half}$ predicate will be used in the private pre- and post-condition of atomic specification while  $\texttt{public\_half}$ will be used in the public pre- and post-condition of atomic specification.
-
-We define the contribution and reference part of the over-all ghost state as follows:
+\textbf{Over-all Ghost State}: We have constructed the per-node ghost state for each node, using unique ghost variable for each node.
+But if we are given a ghost variable (\emph{gname}) to access the resources in the node labeled with that variable, we do not any clue about whether or not that variable exists in the tree.
+So, we need another kind of ghost state, which represents the a set of all per-node ghost variable.
+Again We can use ``reference" pattern to define this ghost state:
 \begin{verbatim}
 Definition ghost_ref g r1 := ghost_reference(P := set_PCM) r1 g.
-Definition in_tree g r1 := EX sh: share, ghost_part(P := set_PCM) sh (Ensembles.Singleton r1) g.
+Definition in_tree g g1 := EX sh: share, ghost_part(P :=set_PCM) sh (Ensembles.Singleton g1) g.
 \end{verbatim}
@@
 The global invariant, which we introduced in our atomic specification as $\mathsf{tree\_rep}$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
-\begin{align*}\mathsf{tree\_rep}(T,g) \triangleq [\ \forall n. \mathsf{public\_half\textsubscript{gn}} \ (range_n* ghost\_info_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
+\begin{align*}\mathsf{tree\_rep}(T,g) \triangleq [\ \forall n.
+\mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
+ Later while writing atomic specification, we will use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part respectively.
+Once we open \emph{global invariant}, we get access to $\mathsf{ghost\_ref}$ which will be used with $\mathsf{in\_tree}$ to establish the fact that node actually exists in the tree.
+We proved the following lemma for this:
+$$
+Lemma\ \mathbf{\ node\_exist\_in\_tree : }\ \forall\ g\ s\ g\_in,\  \mathbf{in\_tree}\ g\ g\_in  * \mathbf{ghost\_ref}\ g\ s\ \Rightarrow\ g\_in\ \in s .
+$$
+\subsubsection{Combining all pieces in atomic specification}
+We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification.
+The private pre- and postcondition of the atomic specification is given by defining the predicate $\mathsf{nodebox\_rep}(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and over-all ghost state is represented by ghost name $g$.
+The $\mathsf{nodebox\_rep}$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
+\begin{align*}
+    &\mathsf{nodebox\_rep}_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \mathsf{nodebox\_rep}_{0.5}(pa,g,ga) \\& *\ \mathsf{nodebox\_rep}_{0.5}(pb,g,gb))\end{align*}
 
-Similarly,  $\texttt{nodebox\_rep}$ predicate ties together $\texttt{my\_half}$ and $\texttt{in\_tree}$ inside the $\texttt{lock\_inv}$ assertion of each node's lock as follows:
-\begin{align*}\mathsf{nodebox\_rep}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ &\mathsf{lock\_inv\textsubscript{lock}}(\mathsf{my\_half\textsubscript{g\_root}} *\ \mathsf{in\_tree}\ g\ g\_root\ *\ \exists\ pa,\ pb,\ ga,\ gb, \\&tp\mapsto\ (pa * pb)\ *\  \mathsf{nodebox\_rep}(pa,g,ga)\ *\ \mathsf{nodebox\_rep}(pb,g,gb))\end{align*}
+The $\mathsf{nodebox\_rep}$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in Section \ref{safety}.
+In this case, the lock invariant of recursive lock have, along with the information about the children node,$\mathsf{my\_half}$ predicate which holds the partial information describing the current thread's contribution to the shared state.
+
+The public pre- and postcondition of the atomic specification is given by defining the predicate $\mathsf{tree\_rep}(T,g)$ that describes the abstract state of the binear search tree $T$  in terms of collection of ghost states each describing the information about separate node in the tree..
+The $\mathsf{tree\_rep}$ predicate ties together $\mathsf{public\_half}$ part of per-node ghost states and $\mathsf{ghost\_ref}$ of over-all ghost state as follows: \begin{align*}&\mathsf{ghost\_tree\_rep}(tg,range, g, g\_current) \triangleq\ \exists\ k, v, ga, gb\ \in tg,\ghost{\mathsf{public\_half}\ (range, Some(k,v,ga,gb))}{g\_current}\\& *\ \mathsf{ghost\_tree\_rep}(left(tg),(left(range),k),g,ga)\ *\ \mathsf{ghost\_tree\_rep}(right(tg),(k, right(range)),g,gb) \end{align*}
+\begin{align*}\mathsf{tree\_rep}(T,g,g\_root) \triangleq \ \exists\ (tg:ghost\_tree),\ &\mathsf{pure\_tree}(tg)\ =\ T\ \land\ \mathsf{ghost\_tree\_rep}(tg,(-\infty, +\infty),g,g\_root)&\\ *\ \ghost{\mathsf{ghost\_ref}\ (\mathsf{find\_ghost\_set}(tg))}{g}\end{align*}
+Where $tg$ is called \emph{ghost tree}, which can be existentially quantified, extend the node in pure tree $T$ with actual pointer and ghost names for child nodes.
+
+The atomic specification for any binary search tree's operation \texttt{op} can be written as follows:
+\begin{align*}\forall t.\ \langle \mathsf{nodebox\_rep}(p,g,g_{\mathit{root}})\ |\ \mathsf{tree\_rep}(t,g)\rangle\ \texttt{op}(p)\ \langle \mathsf{nodebox\_rep}(p,g,g_{\mathit{root}})\ |\ \mathsf{tree\_rep}(t', g)\rangle
+\end{align*}
+
+Where $t$ and $t'$ are the abstract state of a tree before and after the function's execution.
+
+\subsection{Insert and Lookup}
 
 \subsubsection{insert}
-The code for the insert method of our concurrent binary search tree is shown in Figure~\ref{insert}.
-A node has a $lock$, key, value, and pointers to the left and right child.
+The C outline-code for the insert method of binary search tree is shown in \ref{insert}.
+A node has a lock, key, value, and pointers to the left and right child.
 Each leaf node in tree are empty node with lock.
@@
 The atomic specification for insert method can be written as follows:
-\begin{align*} \forall t.\ \langle &\mathsf{nodebox\_rep}(p,g\_root)\ |\ \mathsf{tree\_rep}(g\_root, t)\rangle \end{align*} 
+$$\forall t.\ \langle \mathsf{nodebox\_rep}(p,g,g\_root)\ |\ \mathsf{tree\_rep}(t,g)\rangle $$ 
 $$\texttt{insert}(p,k,v)$$
-\begin{align*}\langle \mathsf{nodebox\_rep}(p,g\_root)\ |\ \mathsf{tree\_rep}(g\_root, t[k\mapsto v])\rangle \end{align*}
-We are not guaranteed that the state of  a tree at the end of the insertion will be $t[k\mapsto v]$ where $t$ is the tree state when insert is called; rather, we know that p always holds some tree during the function's execution, and at some point the function will take that tree $t$, add $[k\mapsto v]$,and then eventually return while in the meantime another thread may have modified the state of tree from $t[k\mapsto v]$ to any other arbitrary state that the current thread do not know.
-Since we have given the specification which is strong enough to specify both safety and correctness properties of insert method, we can now verify that body of the function satisfy that specification using various VST and Iris (encoded in VST) tactics.
-To prove the loop inside insert method, we need loop-invariant as follows:
-\begin{align*} \mathsf{insert\_inv}(b, x, g\_root, range,info) \triangleq\ &\exists\ lock,\ g\_current,\ np,\ !!(x\in range)\ \&\& \ \mathsf{my\_half}(g\_current,range,info)\ *\ R\ np\ * \\&\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
+$$\langle \mathsf{nodebox\_rep}(p,g,g\_root)\ |\ \mathsf{tree\_rep}(t[k\mapsto v],g)\rangle $$
+We are not guaranteed that the state of  a tree at the end of the insertion will be $t[k\mapsto v]$ where $t$ is the tree state when insert is called; rather, we know that p always holds some tree during the function's execution, and at some point the function will take that tree $t$, add $[k\mapsto v]$, and then eventually return, while in the meantime another thread may have modified the state of tree from $t[k\mapsto v]$ to any other arbitrary state that the current thread do not know.
 
-Where $P_p$ and $Q_p$ are $\mathsf{tree\_rep}(g\_root, t)$ and $\mathsf{tree\_rep}(g\_root, t[k\mapsto v])$ respectively.
+The \emph{lock}, used in the implementation of functions, provides \texttt{acquire} and \texttt{release} functions, which allow threads to transfer ownership of resource invariants.
+Their pre- and postconditions are as follows:
+$$\{!!\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{acquire}(\ell)\ \{R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$
+$$\{!!(\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{exclusive}\ R) * R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{release}(\ell)\ \{\mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$
+
+The resource invariant $R$ need to be \emph{exclusive}, i.e, it can only hold once in any given state. This allows us to know that if the current thread holds the invariants, it also hold the lock.
+
+Since we have given the specification that is strong enough to specify both the safety and the functional correctness properties of insert method, we can now verify that body of the function satisfies the atomic specification given above using various VST and Iris (encoded in VST) tactics.
+Once we have outermost specification for a function, the another creative part in verification is finding the right loop-invariant.
+We defines the loop invariant for the insert function as follows:
+\begin{align*} &\mathsf{insert\_inv}(b, x, g, g\_root) \triangleq\ \exists\ \mathit{lock},\mathit{g\_current},\mathit{np},\mathit{range},\mathit{info},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{g\_current}\ *\ R\ \mathit{np}\\& * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \mathsf{nodebox\_rep}(b,g,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
+
+Where $\mathsf{P_p}$ and $\mathsf{Q_p}$ are $\mathsf{tree\_rep}(t,g)$ and $\mathsf{tree\_rep}(t[k\mapsto v],g)$ respectively.
+The resources $\mathsf{R}$, parameterized by a node pointer $np$, describes the concrete information about a node.
+Also, $x$ is the key to be inserted that must be in the bound $\mathsf{range}$.
 This loop invariant has some existential variables which characterized the state of each iteration of loop in the C code.
-We use $forward\_while$ tactic just before the loop with above $insert\_inv$ which leaves four subgoals.
-The first subgoal is to prove that the current precondition in our proof satisfies the insert invariant, the second subgoal is about type-checking the loop-test expression which is often solve automatically by $entailer!$ tactic, the third subgoal is to prove that the loop body preserves the loop invariant for that we must forward-symbolic-execute through the loop body, and final subgoal is to prove that the post condition of the loop body entails the insert invariant.
 
-Figure \ref{insertproof} shows the $insert$ function annotated with separation logic specification. 
+Figure \ref{insertproof} shows the $\texttt{insert}$ function annotated with separation logic specification.
+The proof starts with $\mathsf{nodebox\_rep}$ and $\mathsf{atomic\_shift}$ as the precondition.
+After acquiring the lock for the root node, the thread accesses the information inside the lock invariant of the root node.
+In order to prove $\mathsf{while}$ loop, we show that the precondition satisfies the $\mathsf{insert\_inv}$, then prove that the loop body preserves the loop invariant ( cases inside $\mathsf{else if}$ clauses at line 15 and 17 in \ref{insert}).
+Once we locate the empty node for inserting new key-value pair (i.e inside the first $\mathsf{if}$ clause), we open the global invariant encoded in the $\mathsf{atomic\_shift}$, create the ghost names $\texttt{g1}$ and $\texttt{g2}$ for two new empty child nodes of current node, and prove that insertion satisfy the public postcondition $\mathsf{tree\_rep}(t[k\mapsto v],g)$ with the help of $\mathsf{sync\_commit}$ rule described in section \ref{atomicity}.
+This is the \emph{linearization point} of the $\mathsf{insert}$ operation and must be done at any point in the proof, but before releasing the current node's lock (i.e in critical section).
+In order to prove the public postcondition $\mathsf{tree\_rep}(t[k\mapsto v],g)$, we need to first locate the current node in $\mathsf{tree\_rep}(t,g)$ using ghost name $g\_current$, and then perform necessary modification such that the final state of the global invariant satisfies $\mathsf{tree\_rep}(t[k\mapsto v],g)$.
+We create and prove separate lemma for this, called $\emph{extract}$ lemma which is depicted in the \ref{extract_insert}.
+
+With the help of $\mathsf{in\_tree}$ assertion, we locate a leaf node denoted by ghost name $g\_current$ in the global state $\mathsf{tree\_rep}(t,g)$.
+Then we prove the \emph{extract} lemma, which says that the state of tree after adding new key-value pair (k,v) will be $\mathsf{tree\_rep}(t[k\mapsto v],g)$, by applying induction in $t$.
+
+Another important step in the $\mathsf{insert}$ proof is the case when we find that the key to be inserted already exists in the tree (i.e.
+the last $\texttt{else}$ clause in C code).
+The modification of the global state of the tree t is shown in \ref{extract_insert2}.
+In that case, the old value $v'$ associated with key is overridden by new value $v$ keeping a key $40$ and child pointers as as they are.
+This is also the \emph{linearization point} of the $\mathsf{insert}$ operation.
+We augment the same $\emph{extract}$ lemma with the clause that states the scenario in \ref{extract_insert2}, and prove the lemma in the same induction over the tree $t$.
+
+In the rest two cases, the $\texttt{insert}$ method compares the key
+of the current node with the key $\texttt{x}$ and goes down into the
+corresponding left or right branch. We need to show that the range in
+the ghost state shrinks correctly. To achieve this goal we open the
+global invariant encoded in the $\mathsf{atomic\_shift}$ to retrieve
+the necessary information. But this time these operations maintain the
+original state. For the left branch case, we accomplish the proof by
+the following lemma:
+\begin{verbatim}
+Lemma in_tree_left_range:
+  forall (B: Type) (b: tree -> B -> mpred) (Q : B → mpred) (x x0: Z) (g g_root : gname)
+    (inv_names : invG) (v: val) (g_in ga gb: gname) (r a: node_info),
+    check_key_exist' x r.1 = true -> r.2 = Some (Some (x0, v, ga, gb)) -> x < x0 ->
+    atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
+                 b Q * my_half g_in gsh1 r * in_tree g g_in * my_half ga gsh1 a
+    |-- atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
+    b Q * my_half g_in gsh1 r *
+    (EX ba, !! (less_than_equal ba r.1.1 = true /\
+                range_inclusion a.1 (ba, Finite_Integer x0) = true) &&
+            (in_tree g g_in * my_half ga gsh1 (ba, Finite_Integer x0, a.2))).  
+\end{verbatim}
+The lemma for the right branch case is similar. Both use the
+$\texttt{sync\_rollback}$ lemma described in section \ref{atomicity}.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=150mm,scale=0.5]{FIG/extract_insert.png}
+\caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation }
+\label{extract_insert}
+\end{figure}
+\begin{figure}[htb]
+\centering
+\includegraphics[width=150mm,scale=0.5]{FIG/extract_insert_2.png}
+\caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation, where key $k$ already exists in the tree $t$ }
+\label{extract_insert2}
+\end{figure}
 \begin{figure}[htp]
@@
  $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
- \vspace*{-10pt}
-\begin{lstlisting}[language = C]
-void insert (treebox t, int x, void *value) {
-  struct tree_t *tgt = *t;
-  ....
-  acquire(l);
+ \vspace*{-20pt}
+\begin{lstlisting}[language = C,numbers = none]
+insert(treebox t, int x, void *value)
+  tree_t *tgt = *t;
+  acquire(tgt->lock)
  \end{lstlisting}  
- $$\left\{\begin{array}{l} \mathsf{my\_half}(g\_root,(-\infty,+\infty),info)\ *\ R\ b\ *\ \mathsf{lock\_inv}(l,lsh2,R')\ *\ \\
- \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$ $$\Rrightarrow$$  $$\left\{\begin{array}{l} insert\_inv \end{array}\right\}$$ 
-  \vspace*{-10pt}
-  \begin{lstlisting}[language = C]
+ $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}((-\infty,+\infty),info)}{g\_root}\ *\ R\ b\ *\ \mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\
+ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} insert\_inv \end{array}\right\}$$ 
+  \begin{lstlisting}[language = C,numbers = none]
   for(;;) {
        \end{lstlisting}   
-   $$\left\{\begin{array}{l} insert\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}!!(x\in range)\&\&\mathsf{my\_half}(g\_current,range,info)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
- \vspace*{-10pt}      
-      \begin{lstlisting}[language = C]
+   $$\left\{\begin{array}{l} insert\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ \\*\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+      \begin{lstlisting}[language = C,numbers = none]
     p = tgt->t;
-    if (p==NULL) {
-      tree_t *p1 = (struct tree_t *) surely_malloc (sizeof *tgt);
-      tree_t *p2 = (struct tree_t *) surely_malloc (sizeof *tgt);
-      ....
-      p = (struct tree *) surely_malloc (sizeof *p);
-      tgt->t = p;
+    if (p==NULL)
+      tree_t *p1, *p2 = malloc();
+      p = malloc(); tgt->t = p;
       p->key=x; p->value=value; p->left=p1; p->right=p2;
            \end{lstlisting} 
-  $$\left\{\begin{array}{l} \mathsf{own}\ g1\ a1\ \mathit{pp}\ *\ \mathsf{own}\ g2\ a2\ \mathit{pp}\ *\ \mathsf{my\_half}(g\_current,range,None)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
- $$\Downarrow{\textbf{sync\_commit}}$$
-$$\left\{\begin{array}{l} \mathsf{my\_half}(g\_current,range,None)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+  $$\left\{\begin{array}{l} \ghost{\mathsf{own}\ (a1)}{g1}\ *\ \ghost{\mathsf{own}\ (a2)}{g2}\ *\ \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ \\*\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit}}$$
+$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
  \vspace*{-10pt}
-        \begin{lstlisting}[language = C]
+        \begin{lstlisting}[language = C,numbers = none]
       release2(l);
@@
         \vspace*{-10pt}
-         \begin{lstlisting}[language = C]
+         \begin{lstlisting}[language = C,numbers = none]
       return;
-    } else {
-      int y = p->key;
-      if (x<y){
+     else 
         ....
-      } else if (y<x){
-        ....
-      }else {
+       else 
       	p->value=value;
       	\end{lstlisting} 
-$$\left\{\begin{array}{l} \mathsf{my\_half}(g\_current,range,None)*\ R\ np\ *
-\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
- $$\Downarrow{\textbf{sync\_commit}}$$
-$$\left\{\begin{array}{l} \mathsf{my\_half}(g\_current,range,None)*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
-
-      	\begin{lstlisting}[language = C]
+$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *
+\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit}}$$
+$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+         \vspace*{-10pt}
+      	\begin{lstlisting}[language = C, numbers = none]
         release2(l);
@@
         $$\left\{\begin{array}{l}  \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
-        \begin{lstlisting}[language = C] 
+         \vspace*{-10pt}
+        \begin{lstlisting}[language = C, numbers = none] 
       	return;
-      }
-    }
-  }
-
-} 
 \end{lstlisting}
 \end{subfigure}
-\caption{The \lstinline{insert} function annotated with separation logic specification}
+\caption{The $\texttt{insert}$ function annotated with separation logic specification}
 \label{insertproof}
@@
 \subsubsection{lookup}
+
+The code for the $\mathsf{lookup}$ method is shown in \ref{lookup}.
+It takes the location of the root pointer and a key as the arguments.
+A thread spans the tree to find a given key using hand-over-hand locking mechanism.
+Once a thread finds the key in the tree, it gets the value associated with that key, release the current node's lock, and return the value.
+
+The atomic specification for $\mathsf{lookup}$ method can be written as follows:
+\begin{align*} \forall t.\ \langle &\mathsf{nodebox\_rep}(p,g,g\_root)\ |\ \mathsf{tree\_rep}(t,g,g\_root)\rangle \end{align*} 
+$$\mathsf{lookup}(p,k)$$ 
+\begin{align*}\langle\lambda v.\ \mathsf{nodebox\_rep}(p,g,g\_root)\ |\ (v = \mathsf{lookup}(p,k))\ \&\&\ \mathsf{tree\_rep}(t,g,g\_root)\rangle \end{align*}
+
+This specification is almost similar to the specification given to the $\mathsf{insert}$ method, except the content of public part in the postcondition.
+To prove the loop inside the lookup method, we need loop-invariant as follows:
+\begin{align*} &\mathsf{lookup\_inv}(b, x, \mathit{g\_root}, \mathit{range},\mathit{info}) \triangleq\ \exists\ \mathit{lock},\ \mathit{g\_current},\ \mathit{np},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{\mathit{g\_current}}\\& *\ R\ np\ * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \mathsf{nodebox\_rep}(b,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
+
+Where $P_p$ and $Q_p$ are $\mathsf{tree\_rep}(g\_root, t)$ and $(v = \mathsf{lookup}(p,k))\ \&\&\ \mathsf{tree\_rep}(g\_root, t)$ respectively.
+The proof steps for the $\texttt{lookup}$ verification are almost similar to the steps used in $\texttt{insert}$ verification with few differences which we discuss below.
+
+Figure \ref{lookupproof} shows the $\mathsf{lookup}$ function annotated with separation logic specification.
+The proof starts with $\mathsf{nodebox\_rep}$ and $\mathsf{atomic\_shift}$ assertions as the precondition and follows the approach similar to the $\mathsf{insert}$ proof for the verification of loop body.
+Once we find the given key (i.e.
+the last $\mathsf{else}$ clause in C code) in the tree, we need to confirm that the current state of the tree in the global invariant agrees with the fact that the current node, represented by $\texttt{g\_current}$ in the global invariant, actually contains the key-value pair $(k, v)$ in the ghost state represented $\texttt{g\_current}$.
+We accomplish this by opening the global invariant encoded in the $\mathsf{atomic\_shift}$, and proving that lookup satisfy the public postcondition $\mathsf{Q\_p}$ with the help of $\mathsf{sync\_commit_same}$ lemma described in section \ref{atomicity}.
+This is the \emph{linearization point} of the $\mathsf{lookup}$ operation and must be done before releasing the current node's lock.
+In the case of spanning left or right sub-tree, we need to establish $\mathsf{lookup\_inv}$ at the end $\mathsf{if}$ and $\mathsf{else\ if}$ clause.
+Here, we need to show that the key we are searching is still in the bound for left/right node, so we use $\mathsf{sync\_rollback}$ lemma to extract the bound of a left/right node encoded as the ghost state in the public precondition ($\mathsf{tree\_rep}(g\_root, t)$).
+
+A critical difference from $\texttt{insert}$ is that the $\texttt{lookup}$ method does not change the shape the tree.
+With the help of the $\texttt{sync\_commit\_same}$, we need no longer the heavyweight $\emph{extract}$ lemma but instead a simpler $\emph{ramif}$ lemma to locate the current node in the global invariants:
+\begin{verbatim}
+Lemma ghost_tree_rep_public_half_ramif: forall tg g_root r_root g_in,
+Ensembles.In (find_ghost_set tg g_root) g_in -> ghost_tree_rep tg g_root r_root |-- 
+EX r: node_info, !! (range_info_in_tree r r_root tg) && (public_half g_in r * 
+(public_half g_in r -* ghost_tree_rep tg g_root r_root)).
+\end{verbatim}
+
+\begin{figure}[htp]
+\begin{subfigure}[t]{1\textwidth}
+ $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+\begin{lstlisting}[language = C,  numbers = none]
+void *lookup (treebox t, int x) {
+  struct tree *p; void *v;  struct tree_t *tgt;
+  tgt = *t;  void *l = tgt->lock;
+  acquire(l);  p = tgt->t;
+ \end{lstlisting}  
+ $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}((-\infty,+\infty),info)}{g\_root}\ *\ R\ b\ *\ \mathsf{lock\_inv}(l,lsh2,R')\ *\ \\
+ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} lookup\_inv \end{array}\right\}$$ 
+  \begin{lstlisting}[language = C, numbers = none]
+    while (p!=NULL) {
+       \end{lstlisting}   
+   $$\left\{\begin{array}{l} lookup\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\\\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+      \begin{lstlisting}[language = C,  numbers = none]
+    if (x<y){
+      tgt=p->left;
+      ....
+    }else if (y<x){
+      tgt=p->right;
+     ....
+    }else {
+    v = p->value;
+           \end{lstlisting} 
+  $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit\_same}}$$
+$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+        \begin{lstlisting}[language = C,  numbers = none]
+      release2(l);
+         \end{lstlisting}
+       $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+         \begin{lstlisting}[language = C, numbers = none]
+       return v;} }
+  release2(l);  return NULL;  }
+ \end{lstlisting} 
+\end{subfigure}
+\caption{The $\texttt{lookup}$ function annotated with separation logic specification}
+\label{lookupproof}
+\end{figure}
+
 \subsection{Delete}
@@
 \section{Related Work}
+Our work is based on the concurrent separation logic of VST in the style of Iris \cite{10.1145/3022670.2951943} and the logical atomicity from TaDA logic  \cite{tada}.
+The notion of the \emph{lower} and \emph{upper} bound is taken from the flow interface paper \cite{krishna2017flow}.
+Our main technical contributions are the verification of two different implementation of binary search tree (using fine-grained locking and lock-free technique) with respect to the same abstract specification, and first c-level mechanized verification of concurrent search-based data structure (i.e.
+BST).
+Most of the other concurrent separation logic (CSL)-based verification works are on toy languages, while VST lets us use the same logic on real C programs.
+
+Gotsman and Yang \cite{gotsman} is one of the earliest work on concurrent separation logic.
+They introduced the program logic to reason locally about the heap-manipulating program with the notion of dynamic ownership of heap parts by unbounded number of locks and threads, and shown the verification of singly-linked list with fine-grained locking.
+Xiong et al.
+\cite{Xiong2017Abstract} have demonstrated the verification of ConcurrentSkipListMap from java.util.concurrent library using the recent advances in fine-grained concurrency reasoning.
+Their work is mainly based on the abstract atomicity from TaDa logic, and give two modular specifications for concurrent maps: one specification focus on the entire map structure which is suitable for verifying implementation, and another specification focus on the key-value pairs which appropriate for verifying clients.
+We use the same idea of atomicity (though implemented in Iris) in our work.
+
+Krishna et al.
+\cite{krishna2017flow} have presented the proof technique for concurrent search structure templates based on the flow framework and the Iris separation logic, and verified the implementation of concurrent B-tree, hash tables, and linked lists based on the templates.
+Their work is closely related to our work; we took some inspiration from them in building our ghost state.
+But, their proofs are on a toy language rather than real C code.
 \section{Conclusion}
 
+%% Bibliography
+\bibliography{sources}
+
 \end{document}

### 2021-04-13 concurrency/Verification of Concurrent Binary Search Tree.tex
citation updated

@@
 Modern CPU design trends have shifted towards multicore processing over the last two decades.
-This is because the computing power of computers has reached its limit in single-threaded execution, while the quantity of transistors per single-core CPU are increasing as described by Moore's law \cite{591665}.
+This is because the computing power of computers has reached its limit in single-threaded execution, while the quantity of transistors per single-core CPU are increasing as described by Moore's law \cite{moore}.
 CPUs have multiple cores now because that is the only way to get more transistors per chip, and achieve a desirable throughput.
@@
 In this thesis, we describe the formal verification of the correctness of concurrent binary search tree with fine-grained locking.
-We implement the BST algorithms in C and use the Verified Software Toolchain(VST) \cite{10.5555/2670099} to verify with respect to given specification.
+We implement the BST algorithms in C and use the Verified Software Toolchain(VST) \cite{plfcc} to verify with respect to given specification.
 The concurrent separation logic (CSL) is employed to reason about behavior of the program implemented in the C programming language.
-Separation logic (SL) \cite{1029817} is an extension of Hoare logic \cite{10.1145/363235.363259}, way of reasoning about programs using a set of logical rules.
+Separation logic (SL) \cite{seplogic} is an extension of Hoare logic \cite{hoare}, way of reasoning about programs using a set of logical rules.
 SL helps to locally reason about programs that play with pointer data structures, where specifications and proofs of a program component only reflect the chunk of memory that the component uses, not the entire global state of the system.
@@
 The version of CSL formalized in VST is rooted on the work of Hobor et al.
-\cite{10.1007/978-3-540-78739-6_27}, but extended with higher-order ghost state in the style of Iris \cite{10.1145/2951913.2951943}.
+\cite{oraclesematic}, but extended with higher-order ghost state in the style of Iris \cite{higherorderghoststate}.
 \section{Background}
+\label{background}
 \subsection{VST and Iris}
-We use the Verified Software Toolchain \cite{10.5555/2670099}  to prove the correctness of the concurrent binary search tree.
+We use the Verified Software Toolchain \cite{plfcc}  to prove the correctness of the concurrent binary search tree.
 VST is a Coq-based proof system that helps to prove separation logic properties of C programs with the assurance that the properties will be preserved in compiled code.
@@
 VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} (\textsf{gname} in Coq) associated with the value $a$, which may be of any type.
-We will see different kind of ghost states used to verify binary search tree in Chapter \ref{chapter4}.
+We will see different kind of ghost states used to verify binary search tree in Section \ref{correctness}.
 
-The main idea behind the logic from Iris \cite{10.1145/3022670.2951943}, a mechnized higher-order concurrent separation logic framework, is the construction of \emph{global invariant} as \emph{ghost state}.
+The main idea behind the logic from Iris \cite{higherorderghoststate}, a mechnized higher-order concurrent separation logic framework, is the construction of \emph{global invariant} as \emph{ghost state}.
 \emph{Global Invariant} is an invariant on the global (ghost and physical) state of program.
@@
 Thread can use the contents of \emph{global invariant} during atomic operation if it can guarantee that no one will ever see an intermediate state in which invariant does not hold.
-The rules from Jung et al.\cite{10.1145/3022670.2951943} for creating and opening invariants are:
+The rules from Jung et al.\cite{higherorderghoststate} for creating and opening invariants are:
 
@@
 
+ The \emph{lock}, used in the implementation of BST operations, provides \texttt{acquire} and \texttt{release} functions, which allow threads to transfer ownership of resource invariants.
+Their pre- and postconditions are as follows:
+$$\{!!\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{acquire}(\ell)\ \{R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$
+$$\{!!(\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{exclusive}\ R) * R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{release}(\ell)\ \{\mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$
+
+The resource invariant $R$ need to be \emph{exclusive}, i.e, it can only hold once in any given state. This allows us to know that if the current thread holds the invariants, it also hold the lock.
+
 \section{Correctness Proofs}
+\label{correctness}
 In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct.
@@
 So, once we acquire the lock, we get access to $\mathsf{sync\_inv}$.
-The $\mathsf{sync\_commit}$ and $\mathsf{sync\_rollback}$, specialized with ghost state with \emph{reference pattern} ($\mathsf{my\_half}$ and $\mathsf{public\_half}$), are the versions of $\mathsf{atomic\_commit}$ and $\mathsf{atomic\_rollback}$ described in Chapter \ref{chapter2}.
+The $\mathsf{sync\_commit}$ and $\mathsf{sync\_rollback}$, specialized with ghost state with \emph{reference pattern} ($\mathsf{my\_half}$ and $\mathsf{public\_half}$), are the versions of $\mathsf{atomic\_commit}$ and $\mathsf{atomic\_rollback}$ described in Section \ref{background}.
 In a commit, we must provide $\mathsf{sync\_inv}$ with resources $R$ that, in combination with the public precondition, allow us to prove the public postcondition; we then gain access to an assertion Q required by the postcondition of the function, and leftover resources $R'$.
@@
 We are not guaranteed that the state of  a tree at the end of the insertion will be $t[k\mapsto v]$ where $t$ is the tree state when insert is called; rather, we know that p always holds some tree during the function's execution, and at some point the function will take that tree $t$, add $[k\mapsto v]$, and then eventually return, while in the meantime another thread may have modified the state of tree from $t[k\mapsto v]$ to any other arbitrary state that the current thread do not know.
-
-The \emph{lock}, used in the implementation of functions, provides \texttt{acquire} and \texttt{release} functions, which allow threads to transfer ownership of resource invariants.
-Their pre- and postconditions are as follows:
-$$\{!!\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{acquire}(\ell)\ \{R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$
-$$\{!!(\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{exclusive}\ R) * R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{release}(\ell)\ \{\mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$
-
-The resource invariant $R$ need to be \emph{exclusive}, i.e, it can only hold once in any given state. This allows us to know that if the current thread holds the invariants, it also hold the lock.
 
@@
 \section{Related Work}
-Our work is based on the concurrent separation logic of VST in the style of Iris \cite{10.1145/3022670.2951943} and the logical atomicity from TaDA logic  \cite{tada}.
+Our work is based on the concurrent separation logic of VST in the style of Iris \cite{higherorderghoststate} and the logical atomicity from TaDA logic  \cite{tada}.
 The notion of the \emph{lower} and \emph{upper} bound is taken from the flow interface paper \cite{krishna2017flow}.
@@
 \bibliography{sources}
-
 \end{document}

### 2021-04-27 concurrency/Verification of Concurrent Binary Search Tree.tex
Give a draft of the introduction part.

@@
 \newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
-\newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
+\renewcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
 \newcommand{\us}{$\mu$s}
@@
 \section{Introduction}
+\input{intro.tex}
 Modern CPU design trends have shifted towards multicore processing over the last two decades.
@@
 $\texttt{x = 0;}$\\
-$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\hspace{30px}\pi = \pi_1\ .\ \pi_2$\\
+$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\qquad\pi = \pi_1\ .\ \pi_2$\\
 $\begin{array}{l || l}
@@
 
-where invariants are provided in the form of assertion  $\knowInv{i}{P}$, and states that $P$ is maintained as an invariant on the global state with name $i$.
-The operator $\pvs[][]$ is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e.
-$\knowInv{i}{..\knowInv{i}{P}..}$).
+where invariants are provided in the form of assertion  $\knowInv{i}{P}$, and states that $P$ is maintained as an invariant on the global state with name $i$. The operator%%  $\pvs[][]$
+is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e. $\knowInv{i}{..\knowInv{i}{P}..}$).
 \subsection{Atomic Specifications}
@@
 \section{Related Work}
+\label{related}
 Our work is based on the concurrent separation logic of VST in the style of Iris \cite{higherorderghoststate} and the logical atomicity from TaDA logic  \cite{tada}.
@@
 \section{Conclusion}
+\label{conclusion}
 

### 2021-04-27 concurrency/intro.tex
Give a draft of the introduction part.

The binary search tree (BST) is a common implementation of an ordered
map, a widely used data structure. The concurrent version of the
ordered map forms the bedrock of many parallel programs. We formally
verified the functional correctness of two versions of fine-grained
concurrent binary search trees (CBST) written in C. One adopts the
hand-over-hand locking technique \cite{bayer1977}, the other is lock
free by using atomic primitives such as compare and swap (CAS). Both
CBST implementations support insert, lookup, and delete operations;
they also share the ``same'' specifications to some extent in our
verification. All the proof is machine-checked in Coq.

Our specifications use the logical atomicity introduced in the TaDA
logic \cite{tada}, in the form of \emph{atomic Hoare
triples}. Intuitively, an atomic Hoare triple $ \langle
P \rangle\,\mathbb{C}\, \langle P \rangle$ means the program
$\mathbb{C}$ ``atomically updates'' from $P$ to $Q$. The program may
actually take multiple steps, but evey step before the atomic update
(linearization point) must preserve the assertion $P$. Meanwhile, the
concurrent environment may also update the state before the
linearization point, as long as the states satisfies $P$. The
assertion $Q$ must become true at the linearization point, then the
environment can do whatever it likes. There is no guarantee that $Q$
is still preserved when $\mathbb{C}$ returns. For example, the
specification of our \texttt{insert} operation may be explained as
follows: during the execution of \texttt{insert}, there is
always \emph{some} BST, and at some point the \texttt{insert} will
take a BST $t$, insert a value with a certain key, and then eventually
returns (meanwhile other threads may have further modified the
inserted tree).

We employ the Verified Software Toolchain (VST) \cite{plfcc} to verify
the correctness of CBST. Although the concurrent separation logic
(CSL) has been formalized in VST rooted on the work of Hobor et
al.\ \cite{oraclesematic}, we extend it with two descendants of CSL so
as to accomplish the verification. One is the logical atomicity
mentioned above; the other is the higher-order ghost state in the
style of Iris \cite{higherorderghoststate}. The ghost states are used
to contruct both the global invariants and the local state in our
proofs. They will be further discussed in \S\ref{correctness}.

We highlights a few innovative aspects about our verification of CBST:
\begin{description}
\item [Range ghost] We abstract the BST via a pair of values
called \emph{range} which represents the lower-bound and upper-bound
of keys on each node to reason about the BST in the current
settings. We prove that range with the merging operation forms a
partial commutative monoid (PCM) so that it can be encoded in ghost
states.

\item [\textsf{sync\_inv} pattern] It is a particular approach
combining locks with general invariants to solve the dilemma caused by
the fine-grained locking mechanism: we do not have a lock to control
the access of the entire BST nor can we access the state of the BST
via atomic operations.
\end{description}

Our specific contributions are:
\begin{itemize}
\item To the best of our knowledge, this is the first mechanized
verification of an concurrent search-based data structure written in a
real programming language.

\item We illustrate how to incorporate the CSL in VST, the higher-order
ghost state in the style of Iris, and the logical atomicity from the
TaDA logic together to verify the CBST.
\end{itemize}

We introduce the background about the verification of concurrent C
programs in \S\ref{background}, including the tool-chain VST and Iris,
the concept of ghost states, global invariants, and atomic
specifications. The thread-safety proofs of operations on CBST are
first explained in \S\ref{safety}, where we show the \emph{recursive}
lock pattern for hand-over-hand locking mechanism. The functional
correctness proofs are presented in \S\ref{correctness}. We detail the
use of the \textsf{sync\_inv} pattern, the combination of recursive
lock invariants and ghost states together in the atomic
specifications, and the proof skeleton of each operation on CBST. The
related work is discussed in \S\ref{related}. We end with the
conclusion of our work in \S\ref{conclusion}.

### 2021-07-08 concurrency/Verification of Concurrent Binary Search Tree.tex
trying to describe generic atomic specs

@@
 \usepackage{amsthm}
-%\usepackage{amssymb}
+\usepackage{mathpartir}
 \usepackage{semantic}
@@
 \newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
-\renewcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
 \newcommand{\us}{$\mu$s}
+\newcommand{\treerep}{\ensuremath{\mathsf{bst}}}
+\newcommand{\nodeboxrep}{\ensuremath{\mathsf{bst\_ref}}}
 
-%\newcommand{\ignore}[1]{}
+\newcommand{\ignore}[1]{}
 
@@
 
-\title{Verifying a Binary Search Tree with Fine-Grained Locking}
+\title{Verifying Fine-Grained and Lock-Free Binary Search Trees}
 \author{Roshan Sharma}
@@
 \input{intro.tex}
-Modern CPU design trends have shifted towards multicore processing over the last two decades.
-This is because the computing power of computers has reached its limit in single-threaded execution, while the quantity of transistors per single-core CPU are increasing as described by Moore's law \cite{moore}.
-CPUs have multiple cores now because that is the only way to get more transistors per chip, and achieve a desirable throughput.
-However, doubling the quantity of cores on one chip does not immediately imply an increase in the performance by a factor of two.
-Programs that we aim to run in multiple cores have to be carefully designed to efficiently divide the workload into threads, and facilitate non-conflicting access to the shared data by multiple threads in parallel.
-These challenges is resolved by designing the concurrent data structures, algorithms that provide thread-safe access to the shared data by storing and organizing data in a standard way.
-Concurrent data structures are now widely employed in real-world applications like database algorithms, file systems, networking, e-commerce, etc.
-Designing these data structures has been considered as one of the difficult tasks in the software development fields.
-Despite the careful design, implementation, and thorough unit testing of these algorithms, it is very likely to expect subtle bugs and unexpected behaviors of the program.
-This is because threads interact with each other in complex and subtle ways such that it is almost impossible to cover every corner case while testing.
-Therefore, it is important to use a precise way of proving the correctness of concurrent data structures which provides a guarantee that the given algorithm correctly implements the specified concurrent data structures.
 
-Formal verification is an active field of research that aims to prove or disprove the correctness of algorithms with respect to a certain formal specification, using mathematical techniques.
-Model checking and theorem proving  (also called deductive reasoning) are the two main approaches in this field.
-Model checking builds the mathematical model of a system as a collection of states and actions that move from one state to another.
-The model checker then outputs yes if the given model meets the given specifications and produces a counterexample otherwise, after a systematic and exhaustive exploration of a mathematical model.
-Model checking approach only works for finite models and certain infinite models that can be easily abstracted into finite sets of states.
-While it is a simple and automated operation, it does not scale well to large and complex systems where the search spaces are typically too large to be explored in a reasonable amount of time with a reasonable amount of space.
-Theorem proving, on the other hand, is a more powerful technique than model checking which can deal with an infinite state space.
-It works by generating a collection of mathematical proof obligations from the system and its specifications and proving these obligations using either proof assistants (such as Coq, Isabelle, PVS, HOL) or using automatic theorem provers (such as SMT solvers).
-Theorem proving allows us to work on more accurate representations of the system and express any properties, but most proofs have to be done manually which requires time and expertise.
-Since the binary search tree that we target for verification is a concurrent data structure of unbounded key-value pairs, and processes.
-The concurrency considerably increases the state space, and we also want to work with a model that's connected to the actual behavior of C programs, while model checking usually works on more abstract models.
-So, we use the theorem proving approach to formalize the proof of correctness.
-
-In this thesis, we describe the formal verification of the correctness of concurrent binary search tree with fine-grained locking.
-We implement the BST algorithms in C and use the Verified Software Toolchain(VST) \cite{plfcc} to verify with respect to given specification.
-The concurrent separation logic (CSL) is employed to reason about behavior of the program implemented in the C programming language.
-Separation logic (SL) \cite{seplogic} is an extension of Hoare logic \cite{hoare}, way of reasoning about programs using a set of logical rules.
-SL helps to locally reason about programs that play with pointer data structures, where specifications and proofs of a program component only reflect the chunk of memory that the component uses, not the entire global state of the system.
-Concurrent separation logic (CSL) is a variant of the separation logic for reasoning about the concurrent programs.
-The version of CSL formalized in VST is rooted on the work of Hobor et al.
-\cite{oraclesematic}, but extended with higher-order ghost state in the style of Iris \cite{higherorderghoststate}.
 \section{Background}
@@
 
-$$\inference[\textsf{inv\_alloc}]{}{\triangleright P \vdash \pvs[E] \mathsf{EX}\ i : \mathsf{iname}, \knowInv{i}{P}}$$
-$$\inference[\textsf{inv\_open}]{i \in E}{\knowInv{i}{P} \vdash \pvs[E][E \setminus i] \triangleright P * (\triangleright P \wand \pvs[E \setminus i][E] \mathsf{emp})}$$
+\begin{mathpar}
+\inference[\textsf{inv\_alloc}]{}{\triangleright P \vdash \pvs[E] \mathsf{EX}\ i : \mathsf{iname}, \knowInv{i}{P}}
+
+\inference[\textsf{inv\_open}]{i \in E}{\knowInv{i}{P} \vdash \pvs[E][E \setminus i] \triangleright P * (\triangleright P \wand \pvs[E \setminus i][E] \mathsf{emp})}
+\end{mathpar}
 
@@
 is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e. $\knowInv{i}{..\knowInv{i}{P}..}$).
+
 \subsection{Atomic Specifications}
-\emph{Linearizability} is one of the most popular traditional correctness condition for concurrent data structures.
-It is defined in terms of a set of high-level operations, such as insert and remove for a set; if a data structure is linearizable, any operation performed on it tends to happen atomically and in some order that corresponds to the real-time ordering of such operations.
-For instance, if operation $A$ is completed before operation $B$, then $B$ should logically follow $A$.
-We can prove the linearizability property of concurrent data structures using \emph{global invariants}.
-Suppose we have an invariant that said ``there exists a linear history of operations performed on the data structures, stored in some ghost state, and the current  state of the data structure is the one that would result from applying each of those operations in order".
-Then this directly implies that data structure is linearizable.
-We can allow a data structure operation to access the content of global invariants, treating that operation as an atomic operation if we are able prove that that operation \emph{appears to execute atomically} at some linearization point.
-This idea was introduced in the TaDA logic \cite{tada} in the form of \emph{atomic Hoare triples} as shown below, and called as  \emph{Logical atomicity}.
-$$\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$
-
-where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of shared data structure.
-$P_l$ and $Q_l$ are \emph{local} to each thread
-and do not interfere with global state of a data structure, while $P_p$ and $Q_p$ are public to all threads and must satisfy following rule: $P_p$ must be true at every point from the beginning of the function until the linearization point, for some abstract value $a$, and $Q_p$ must be true at the lineariztion point, but may not still true by the end of the function.
-So, at the linearization point the operation atomically transitions from $P_p$ to $Q_p$.
+\label{atomic}
+For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads.
+This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations).
+In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$.
+For instance, the specification $$\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution.
+%maybe something about linking ghost state, which parts are real vs.
+ghost, accessing invariants around atomic triple
 
 VST has encoded an atomic Hoare triple, and provide the way that matches the notation VST uses for normal specifications.
-Such specification is called \emph{atomic specification}, and can be formalized in Coq as shown in  \ref{atomic_spec}.
+Such specification is called \emph{atomic specification}, and can be formalized in Coq as shown in  \ref{atomic-spec}.
 This is how we write pre- and postcondition in VST.
@@
 \caption{Atomic Specification in VST}
-\label{atomic_spec}
+\label{atomic-spec}
 \end{figure}
@@
 
-\section{Safety Proofs}
+\ignore{\section{Safety Proofs} %maybe cut this, or roll it into the next section
 \label{safety}
@@
 \subsection{Specification and Verification}
-We begin the verification by defining a predicate $\mathsf{nodebox\_rep}(p)$ that describes the concrete representation of the tree, where $p$ is the pointer to the root node.
+We begin the verification by defining a predicate $\nodeboxrep(p)$ that describes the concrete representation of the tree, where $p$ is the pointer to the root node.
 The $\texttt{nodebox\_rep}$ predicate ties together $\texttt{lock\_inv}$ assertion for each recursive lock as follows:
 \begin{align*}
- &\mathsf{nodebox\_rep}_{\pi}(p) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \exists\ pa\ pb\ \ tp\mapsto\ (pa * pb)\ *\  \mathsf{nodebox\_rep}_{0.5}(pa)\ *\ \mathsf{nodebox\_rep}_{0.5}(pb)  \end{align*}
+ &\nodeboxrep_{\pi}(p) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \exists\ pa\ pb\ \ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa)\ *\ \nodeboxrep_{0.5}(pb)  \end{align*}
 
@@
 The resource invariant $R$ need to be \emph{exclusive}, i.e, it can only hold once in any given state. This allows us to know that if the current thread holds the invariants, it also hold the lock. 
+}
 
@@
 \label{correctness}
-In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct.
+As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
+\begin{mathpar}
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle
+
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle
+
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ t\ k)\rangle
+\end{mathpar}
+In our case, insert, lookup, and delete are simple Gallina functions on a functional tree data structure.
+The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract tree $t$.
+
+In the following sections, we describe the construction of these predicates, and the proofs of the operations, for BST implementations based on hand-over-hand locking and lock-free atomic operations.
+
+%look at some of the Iris papers in CPP, etc. for how we should describe the proofs
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\ignore{In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct.
 To prove the correctness, our threads need to be able to record the information about the actions they have performed on the shared state, instead of sealing all knowledge of the shared data structure inside the lock invariant.
@@
 In VST, any Coq type can be used as ghost state, as long as we can describe what happens when two elements of that type are joined together.
-We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \emph{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.
+We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \emph{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.}
 
@@
 technical material: definition of sync\_inv, sync\_commit and sync\_rollback
-\subsection{Specification of the algorithm}
-In addition to program state, the verification of concurrent program involves two aspects which do not appear in actual C program: the \emph{shares} on shared resources and \emph{ghost state} described in Section \ref{safety}.
+
+\subsection{Specification of the hand-over-hand BST}
+%In addition to program state, the verification of concurrent program involves two aspects which do not appear in actual C program: the \emph{shares} on shared resources and \emph{ghost state} described in Section \ref{safety}.
 We need to extend our specification in Section\ref{safety} with \emph{ghost state} to prove the functional correctness.
+
+To verify our tree implementation, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call.
+In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure.
+Our overall approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$.
+We also need to make sure that our lock invariants support the hand-over-hand pattern.
+
 \subsubsection{Ghost States and Range}
@@
 We introduce the $range$ in our proof using the ghost variables.
-This $ranges$ along with other information about the node represent the abstract state of the data-structure.
-We have used two kind of ghost state in our proof: \emph{per-node} ghost state and \emph{over-all} ghost state.\\
+This $ranges$ along with other information about the node represent the abstract state of the data structure.
+We have used two kind of ghost state in our proof: \emph{per-node} ghost state and \emph{overall} ghost state.\\
 \textbf{Per-node Ghost State}: The per-node ghost state is the product of the range and other information about node (key, value, and ghost name for its child nodes), created for each node.
@@
 
-\textbf{Over-all Ghost State}: We have constructed the per-node ghost state for each node, using unique ghost variable for each node.
+\textbf{Overall Ghost State}: We have constructed the per-node ghost state for each node, using unique ghost variable for each node.
 But if we are given a ghost variable (\emph{gname}) to access the resources in the node labeled with that variable, we do not any clue about whether or not that variable exists in the tree.
@@
 We use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part of the atomic specification respectively.
-The global invariant, which we introduced in our atomic specification as $\mathsf{tree\_rep}$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
-\begin{align*}\mathsf{tree\_rep}(T,g) \triangleq [\ \forall n.
-\mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
+The global invariant, which we introduced in our atomic specification as $\treerep$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
+\begin{align*}\treerep(T,g) \triangleq [\ \forall n. \mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
  Later while writing atomic specification, we will use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part respectively.
@@
 We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification.
-The private pre- and postcondition of the atomic specification is given by defining the predicate $\mathsf{nodebox\_rep}(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and over-all ghost state is represented by ghost name $g$.
-The $\mathsf{nodebox\_rep}$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
+The private pre- and postcondition of the atomic specification is given by defining the predicate $\nodeboxrep(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and over-all ghost state is represented by ghost name $g$.
+The $\nodeboxrep$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
 \begin{align*}
-    &\mathsf{nodebox\_rep}_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \mathsf{nodebox\_rep}_{0.5}(pa,g,ga) \\& *\ \mathsf{nodebox\_rep}_{0.5}(pb,g,gb))\end{align*}
+    &\nodeboxrep_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa,g,ga) \\& *\ \nodeboxrep_{0.5}(pb,g,gb))\end{align*}
 
-The $\mathsf{nodebox\_rep}$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in Section \ref{safety}.
+The $\nodeboxrep$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in Section \ref{safety}.
 In this case, the lock invariant of recursive lock have, along with the information about the children node,$\mathsf{my\_half}$ predicate which holds the partial information describing the current thread's contribution to the shared state.
 
-The public pre- and postcondition of the atomic specification is given by defining the predicate $\mathsf{tree\_rep}(T,g)$ that describes the abstract state of the binear search tree $T$  in terms of collection of ghost states each describing the information about separate node in the tree..
-The $\mathsf{tree\_rep}$ predicate ties together $\mathsf{public\_half}$ part of per-node ghost states and $\mathsf{ghost\_ref}$ of over-all ghost state as follows: \begin{align*}&\mathsf{ghost\_tree\_rep}(tg,range, g, g\_current) \triangleq\ \exists\ k, v, ga, gb\ \in tg,\ghost{\mathsf{public\_half}\ (range, Some(k,v,ga,gb))}{g\_current}\\& *\ \mathsf{ghost\_tree\_rep}(left(tg),(left(range),k),g,ga)\ *\ \mathsf{ghost\_tree\_rep}(right(tg),(k, right(range)),g,gb) \end{align*}
-\begin{align*}\mathsf{tree\_rep}(T,g,g\_root) \triangleq \ \exists\ (tg:ghost\_tree),\ &\mathsf{pure\_tree}(tg)\ =\ T\ \land\ \mathsf{ghost\_tree\_rep}(tg,(-\infty, +\infty),g,g\_root)&\\ *\ \ghost{\mathsf{ghost\_ref}\ (\mathsf{find\_ghost\_set}(tg))}{g}\end{align*}
+The public pre- and postcondition of the atomic specification is given by defining the predicate $\treerep(T,g)$ that describes the abstract state of the binary search tree $T$  in terms of collection of ghost states each describing the information about separate node in the tree..
+The $\treerep$ predicate ties together $\mathsf{public\_half}$ part of per-node ghost states and $\mathsf{ghost\_ref}$ of over-all ghost state as follows: \begin{align*}&\mathsf{ghost\_tree\_rep}(tg,range, g, g\_current) \triangleq\ \exists\ k, v, ga, gb\ \in tg,\ghost{\mathsf{public\_half}\ (range, Some(k,v,ga,gb))}{g\_current}\\& *\ \mathsf{ghost\_tree\_rep}(left(tg),(left(range),k),g,ga)\ *\ \mathsf{ghost\_tree\_rep}(right(tg),(k, right(range)),g,gb) \end{align*}
+\begin{align*}\treerep(T,g,g\_root) \triangleq \ \exists\ (tg:ghost\_tree),\ &\mathsf{pure\_tree}(tg)\ =\ T\ \land\ \mathsf{ghost\_tree\_rep}(tg,(-\infty, +\infty),g,g\_root)&\\ *\ \ghost{\mathsf{ghost\_ref}\ (\mathsf{find\_ghost\_set}(tg))}{g}\end{align*}
 Where $tg$ is called \emph{ghost tree}, which can be existentially quantified, extend the node in pure tree $T$ with actual pointer and ghost names for child nodes.
@@
 The atomic specification for any binary search tree's operation \texttt{op} can be written as follows:
-\begin{align*}\forall t.\ \langle \mathsf{nodebox\_rep}(p,g,g_{\mathit{root}})\ |\ \mathsf{tree\_rep}(t,g)\rangle\ \texttt{op}(p)\ \langle \mathsf{nodebox\_rep}(p,g,g_{\mathit{root}})\ |\ \mathsf{tree\_rep}(t', g)\rangle
+\begin{align*}\forall t.\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t,g)\rangle\ \texttt{op}(p)\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t', g)\rangle
 \end{align*}
@@
 The atomic specification for insert method can be written as follows:
-$$\forall t.\ \langle \mathsf{nodebox\_rep}(p,g,g\_root)\ |\ \mathsf{tree\_rep}(t,g)\rangle $$ 
+$$\forall t.\ \langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(t,g)\rangle $$ 
 $$\texttt{insert}(p,k,v)$$
-$$\langle \mathsf{nodebox\_rep}(p,g,g\_root)\ |\ \mathsf{tree\_rep}(t[k\mapsto v],g)\rangle $$
+$$\langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(t[k\mapsto v],g)\rangle $$
 We are not guaranteed that the state of  a tree at the end of the insertion will be $t[k\mapsto v]$ where $t$ is the tree state when insert is called; rather, we know that p always holds some tree during the function's execution, and at some point the function will take that tree $t$, add $[k\mapsto v]$, and then eventually return, while in the meantime another thread may have modified the state of tree from $t[k\mapsto v]$ to any other arbitrary state that the current thread do not know.
@@
 We defines the loop invariant for the insert function as follows:
-\begin{align*} &\mathsf{insert\_inv}(b, x, g, g\_root) \triangleq\ \exists\ \mathit{lock},\mathit{g\_current},\mathit{np},\mathit{range},\mathit{info},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{g\_current}\ *\ R\ \mathit{np}\\& * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \mathsf{nodebox\_rep}(b,g,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
+\begin{align*} &\mathsf{insert\_inv}(b, x, g, g\_root) \triangleq\ \exists\ \mathit{lock},\mathit{g\_current},\mathit{np},\mathit{range},\mathit{info},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{g\_current}\ *\ R\ \mathit{np}\\& * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,g,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
 
-Where $\mathsf{P_p}$ and $\mathsf{Q_p}$ are $\mathsf{tree\_rep}(t,g)$ and $\mathsf{tree\_rep}(t[k\mapsto v],g)$ respectively.
+Where $\mathsf{P_p}$ and $\mathsf{Q_p}$ are $\treerep(t,g)$ and $\treerep(t[k\mapsto v],g)$ respectively.
 The resources $\mathsf{R}$, parameterized by a node pointer $np$, describes the concrete information about a node.
@@
 Figure \ref{insertproof} shows the $\texttt{insert}$ function annotated with separation logic specification.
-The proof starts with $\mathsf{nodebox\_rep}$ and $\mathsf{atomic\_shift}$ as the precondition.
+The proof starts with $\nodeboxrep$ and $\mathsf{atomic\_shift}$ as the precondition.
 After acquiring the lock for the root node, the thread accesses the information inside the lock invariant of the root node.
 In order to prove $\mathsf{while}$ loop, we show that the precondition satisfies the $\mathsf{insert\_inv}$, then prove that the loop body preserves the loop invariant ( cases inside $\mathsf{else if}$ clauses at line 15 and 17 in \ref{insert}).
-Once we locate the empty node for inserting new key-value pair (i.e inside the first $\mathsf{if}$ clause), we open the global invariant encoded in the $\mathsf{atomic\_shift}$, create the ghost names $\texttt{g1}$ and $\texttt{g2}$ for two new empty child nodes of current node, and prove that insertion satisfy the public postcondition $\mathsf{tree\_rep}(t[k\mapsto v],g)$ with the help of $\mathsf{sync\_commit}$ rule described in section \ref{atomicity}.
+Once we locate the empty node for inserting new key-value pair (i.e inside the first $\mathsf{if}$ clause), we open the global invariant encoded in the $\mathsf{atomic\_shift}$, create the ghost names $\texttt{g1}$ and $\texttt{g2}$ for two new empty child nodes of current node, and prove that insertion satisfy the public postcondition $\treerep(t[k\mapsto v],g)$ with the help of $\mathsf{sync\_commit}$ rule described in section \ref{atomicity}.
 This is the \emph{linearization point} of the $\mathsf{insert}$ operation and must be done at any point in the proof, but before releasing the current node's lock (i.e in critical section).
-In order to prove the public postcondition $\mathsf{tree\_rep}(t[k\mapsto v],g)$, we need to first locate the current node in $\mathsf{tree\_rep}(t,g)$ using ghost name $g\_current$, and then perform necessary modification such that the final state of the global invariant satisfies $\mathsf{tree\_rep}(t[k\mapsto v],g)$.
+In order to prove the public postcondition $\treerep(t[k\mapsto v],g)$, we need to first locate the current node in $\treerep(t,g)$ using ghost name $g\_current$, and then perform necessary modification such that the final state of the global invariant satisfies $\treerep(t[k\mapsto v],g)$.
 We create and prove separate lemma for this, called $\emph{extract}$ lemma which is depicted in the \ref{extract_insert}.
 
-With the help of $\mathsf{in\_tree}$ assertion, we locate a leaf node denoted by ghost name $g\_current$ in the global state $\mathsf{tree\_rep}(t,g)$.
-Then we prove the \emph{extract} lemma, which says that the state of tree after adding new key-value pair (k,v) will be $\mathsf{tree\_rep}(t[k\mapsto v],g)$, by applying induction in $t$.
+With the help of $\mathsf{in\_tree}$ assertion, we locate a leaf node denoted by ghost name $g\_current$ in the global state $\treerep(t,g)$.
+Then we prove the \emph{extract} lemma, which says that the state of tree after adding new key-value pair (k,v) will be $\treerep(t[k\mapsto v],g)$, by applying induction in $t$.
 
@@
 the following lemma:
-\begin{verbatim}
-Lemma in_tree_left_range:
-  forall (B: Type) (b: tree -> B -> mpred) (Q : B → mpred) (x x0: Z) (g g_root : gname)
-    (inv_names : invG) (v: val) (g_in ga gb: gname) (r a: node_info),
-    check_key_exist' x r.1 = true -> r.2 = Some (Some (x0, v, ga, gb)) -> x < x0 ->
-    atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
-                 b Q * my_half g_in gsh1 r * in_tree g g_in * my_half ga gsh1 a
-    |-- atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
-    b Q * my_half g_in gsh1 r *
-    (EX ba, !! (less_than_equal ba r.1.1 = true /\
-                range_inclusion a.1 (ba, Finite_Integer x0) = true) &&
-            (in_tree g g_in * my_half ga gsh1 (ba, Finite_Integer x0, a.2))).  
-\end{verbatim}
+%\begin{verbatim}
+%Lemma in_tree_left_range:
+%  forall (B: Type) (b: tree -> B -> mpred) (Q : B → mpred) (x x0: Z) (g g_root : gname)
+%    (inv_names : invG) (v: val) (g_in ga gb: gname) (r a: node_info),
+%    check_key_exist' x r.1 = true -> r.2 = Some (Some (x0, v, ga, gb)) -> x < x0 ->
+%    atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
+%                 b Q * my_half g_in gsh1 r * in_tree g g_in * my_half ga gsh1 a
+%    |-- atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
+%    b Q * my_half g_in gsh1 r *
+%    (EX ba, !! (less_than_equal ba r.1.1 = true /\
+%                range_inclusion a.1 (ba, Finite_Integer x0) = true) &&
+%            (in_tree g g_in * my_half ga gsh1 (ba, Finite_Integer x0, a.2))).  
+%\end{verbatim}
 The lemma for the right branch case is similar. Both use the
@@
 \begin{subfigure}[t]{1\textwidth}
- $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+ $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
  \vspace*{-20pt}
@@
  $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}((-\infty,+\infty),info)}{g\_root}\ *\ R\ b\ *\ \mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\
- \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} insert\_inv \end{array}\right\}$$ 
+ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} insert\_inv \end{array}\right\}$$ 
   \begin{lstlisting}[language = C,numbers = none]
@@
        \end{lstlisting}   
-   $$\left\{\begin{array}{l} insert\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ \\*\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+   $$\left\{\begin{array}{l} insert\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ \\*\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
       \begin{lstlisting}[language = C,numbers = none]
@@
            \end{lstlisting} 
-  $$\left\{\begin{array}{l} \ghost{\mathsf{own}\ (a1)}{g1}\ *\ \ghost{\mathsf{own}\ (a2)}{g2}\ *\ \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ \\*\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit}}$$
-$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+  $$\left\{\begin{array}{l} \ghost{\mathsf{own}\ (a1)}{g1}\ *\ \ghost{\mathsf{own}\ (a2)}{g2}\ *\ \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ \\*\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit}}$$
+$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
  \vspace*{-10pt}
@@
          \end{lstlisting}
-       $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+       $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
         \vspace*{-10pt}
@@
 $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *
-\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit}}$$
-$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit}}$$
+$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
          \vspace*{-10pt}
@@
         \end{lstlisting}
-        $$\left\{\begin{array}{l}  \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+        $$\left\{\begin{array}{l}  \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
          \vspace*{-10pt}
@@
 The atomic specification for $\mathsf{lookup}$ method can be written as follows:
-\begin{align*} \forall t.\ \langle &\mathsf{nodebox\_rep}(p,g,g\_root)\ |\ \mathsf{tree\_rep}(t,g,g\_root)\rangle \end{align*} 
+\begin{align*} \forall t.\ \langle &\nodeboxrep(p,g,g\_root)\ |\ \treerep(t,g,g\_root)\rangle \end{align*} 
 $$\mathsf{lookup}(p,k)$$ 
-\begin{align*}\langle\lambda v.\ \mathsf{nodebox\_rep}(p,g,g\_root)\ |\ (v = \mathsf{lookup}(p,k))\ \&\&\ \mathsf{tree\_rep}(t,g,g\_root)\rangle \end{align*}
+\begin{align*}\langle\lambda v.\ \nodeboxrep(p,g,g\_root)\ |\ (v = \mathsf{lookup}(p,k))\ \&\&\ \treerep(t,g,g\_root)\rangle \end{align*}
 
@@
 To prove the loop inside the lookup method, we need loop-invariant as follows:
-\begin{align*} &\mathsf{lookup\_inv}(b, x, \mathit{g\_root}, \mathit{range},\mathit{info}) \triangleq\ \exists\ \mathit{lock},\ \mathit{g\_current},\ \mathit{np},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{\mathit{g\_current}}\\& *\ R\ np\ * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \mathsf{nodebox\_rep}(b,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
+\begin{align*} &\mathsf{lookup\_inv}(b, x, \mathit{g\_root}, \mathit{range},\mathit{info}) \triangleq\ \exists\ \mathit{lock},\ \mathit{g\_current},\ \mathit{np},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{\mathit{g\_current}}\\& *\ R\ np\ * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
 
-Where $P_p$ and $Q_p$ are $\mathsf{tree\_rep}(g\_root, t)$ and $(v = \mathsf{lookup}(p,k))\ \&\&\ \mathsf{tree\_rep}(g\_root, t)$ respectively.
+Where $P_p$ and $Q_p$ are $\treerep(g\_root, t)$ and $(v = \mathsf{lookup}(p,k))\ \&\&\ \treerep(g\_root, t)$ respectively.
 The proof steps for the $\texttt{lookup}$ verification are almost similar to the steps used in $\texttt{insert}$ verification with few differences which we discuss below.
@@
 Figure \ref{lookupproof} shows the $\mathsf{lookup}$ function annotated with separation logic specification.
-The proof starts with $\mathsf{nodebox\_rep}$ and $\mathsf{atomic\_shift}$ assertions as the precondition and follows the approach similar to the $\mathsf{insert}$ proof for the verification of loop body.
+The proof starts with $\nodeboxrep$ and $\mathsf{atomic\_shift}$ assertions as the precondition and follows the approach similar to the $\mathsf{insert}$ proof for the verification of loop body.
 Once we find the given key (i.e.
@@
 In the case of spanning left or right sub-tree, we need to establish $\mathsf{lookup\_inv}$ at the end $\mathsf{if}$ and $\mathsf{else\ if}$ clause.
-Here, we need to show that the key we are searching is still in the bound for left/right node, so we use $\mathsf{sync\_rollback}$ lemma to extract the bound of a left/right node encoded as the ghost state in the public precondition ($\mathsf{tree\_rep}(g\_root, t)$).
+Here, we need to show that the key we are searching is still in the bound for left/right node, so we use $\mathsf{sync\_rollback}$ lemma to extract the bound of a left/right node encoded as the ghost state in the public precondition ($\treerep(g\_root, t)$).
 
@@
 \begin{subfigure}[t]{1\textwidth}
- $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+ $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
 \begin{lstlisting}[language = C,  numbers = none]
@@
  $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}((-\infty,+\infty),info)}{g\_root}\ *\ R\ b\ *\ \mathsf{lock\_inv}(l,lsh2,R')\ *\ \\
- \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} lookup\_inv \end{array}\right\}$$ 
+ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} lookup\_inv \end{array}\right\}$$ 
   \begin{lstlisting}[language = C, numbers = none]
@@
        \end{lstlisting}   
-   $$\left\{\begin{array}{l} lookup\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\\\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+   $$\left\{\begin{array}{l} lookup\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\\\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
       \begin{lstlisting}[language = C,  numbers = none]
@@
            \end{lstlisting} 
-  $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\mathsf{nodebox\_rep}(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit\_same}}$$
-$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+  $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit\_same}}$$
+$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
         \begin{lstlisting}[language = C,  numbers = none]
@@
          \end{lstlisting}
-       $$\left\{\begin{array}{l} \mathsf{nodebox\_rep}(b,g\_root)\ *\ Q\end{array}\right\}$$
+       $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
          \begin{lstlisting}[language = C, numbers = none]

### 2021-07-08 concurrency/intro.tex
trying to describe generic atomic specs

@@
 free by using atomic primitives such as compare and swap (CAS). Both
-CBST implementations support insert, lookup, and delete operations;
+CBST implementations support insert, lookup, and delete operations; %or do we have lock-free without delete?
 they also share the ``same'' specifications to some extent in our
-verification. All the proof is machine-checked in Coq.
+verification. All the proof is machine-checked in Coq. %Say something about existing proofs of fine-grained data structures and how we compare.
 
@@
 verification of an concurrent search-based data structure written in a
-real programming language.
+real programming language. %should probably clarify what we mean by "real"
 

### 2021-07-12 concurrency/Verification of Concurrent Binary Search Tree.tex
doing some editing

@@
 \newcommand{\us}{$\mu$s}
+\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
 \newcommand{\treerep}{\ensuremath{\mathsf{bst}}}
@@
 \author{Shengyi Wang}
+\author{Alexander Oey}
 \author{Lennart Beringer}
@@
 In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together.
-VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} (\textsf{gname} in Coq) associated with the value $a$, which may be of any type.
+VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} ($\gnamety$ in Coq) associated with the value $a$, which may be of any type.
 We will see different kind of ghost states used to verify binary search tree in Section \ref{correctness}.
@@
 
-technical material: definition of sync\_inv, sync\_commit and sync\_rollback
-
 \subsection{Specification of the hand-over-hand BST}
@@
 We have used two kind of ghost state in our proof: \emph{per-node} ghost state and \emph{overall} ghost state.\\
-\textbf{Per-node Ghost State}: The per-node ghost state is the product of the range and other information about node (key, value, and ghost name for its child nodes), created for each node.
-In VST, to use any Coq type as a ghost state, we need to create an instance of \texttt{Ghost} typeclass, \emph{separation algebra} with associated \texttt{join} relation, with an additional \texttt{valid} predicate marking those elements of the algrebra that can be used in ghost assertions.
-\texttt{Ghost} typeclass is related to the PCMs described in \ref{chapter2}.
-For instance, ghost instance for range can be created as follows:
-\begin{verbatim}
- Program Instance range_ghost : Ghost :=
-  { G := (number*number); valid g := True; Join_G a b c := c =  merge_range a b }.
-\end{verbatim}
-Where \texttt{number} $\in (-\infty,+\infty)$ and $\texttt{merge\_range}$ is a function that merges two ranges into one.
-We extend this range information with the other information about the node: key, value and $\texttt{gname}$ for its child.
-We define our total ghost state as follows:
-\begin{verbatim}
-Definition ghost_info : Type := (key * val * gname * gname)%type.
-Instance node_ghost : Ghost := prod_PCM range_ghost (exclusive_PCM (option ghost_info)).
-\end{verbatim}
-The \texttt{exclusive\_PCM} is an exlusive algebra, which means that \texttt{Some} only joins with \texttt{None}, not with another \texttt{Some} (even the same one).
-Now, we use ghost assertion to represent this instance.
-This ghost state follow \emph{reference} pattern: each thread holds partial information describing its contribution to the shared state, and the shared resources hold a "reference" copy that records all of the contributions.
-VST provides \texttt{my\_half} and \texttt{public\_half} assertions for this: contribution element $\mathsf{my\_half}\ \mathit{g}\ \mathit{sh}\ a$ and reference element $\mathsf{public\_half}\ \mathit{g}\ r$.
-To join two elements, we combine the shares and values of the contributions, and require that the elements contain at most one reference between them to ensure the uniqueness of the reference value.
-When \texttt{my\_half} element has the full share, it is guaranteed to be equal to the reference element.
-In our case, \texttt{my\_half} assertion will be local to a thread (that means it will stay in private pre- and postcondition of atomic specification), and  \texttt{public\_half} will be shared with all threads (that means public pre-and postcondition).
+\textbf{Per-node Ghost State}: The per-node ghost state is includes a node's range and other information about the node: its current key, value, and ghost names of its child nodes.
+In VST, we can use a type as ghost state by proving an instance of the \texttt{Ghost} typeclass, analogous to the \emph{resource algebras} of Iris.
+The essential part is a \emph{join} relation describing how two pieces of ghost state combine.
+We define the join of two ranges to be the range that contains both of them, and use an \emph{exclusive} algebra for the rest of the data, so that only one thread knows the current state of a node at a time.
+Ghost state is shared between the per-node lock invariants and the abstract state of the data structure using the \emph{reference} pattern: each thread holds partial information describing its contribution to the shared state, and the shared resources hold a ``reference'' copy that records all of the contributions.
+In VST, this is implemented with assertions $\mathsf{my\_half}\ g\ \mathit{sh}\ a$ and $\mathsf{public\_half}\ g\ r$, where $g$ is the node's identifier ($\gnamety$), $a$/$r$ is the value of the ghost state, and $\mathit{sh}$ is a share describing how much of the reference value is known: if $\mathit{sh}$ is the full share, then we know that $a = r$.
+In the proof, the \texttt{my\_half} assertion for a node will be held by the node's lock, and \texttt{public\_half} will be held by the shared state.
 
-\textbf{Overall Ghost State}: We have constructed the per-node ghost state for each node, using unique ghost variable for each node.
-But if we are given a ghost variable (\emph{gname}) to access the resources in the node labeled with that variable, we do not any clue about whether or not that variable exists in the tree.
-So, we need another kind of ghost state, which represents the a set of all per-node ghost variable.
-Again We can use ``reference" pattern to define this ghost state:
-\begin{verbatim}
+\textbf{Overall Ghost State}: We now have per-node ghost state for each node, indexed by a ghost name $g_n$.
+But if we are given an arbitrary $\gnamety$ for a node, we do not know whether that $\gnamety$ exists in the tree.
+We need another kind of ghost state to hold the set of all nodes in the current tree.
+Again we can use the ``reference'' pattern to define this ghost state, where the abstract state $\mathsf{ghost\_ref}\ s$ holds the full set of nodes, and each node's lock invariant holds an $\mathsf{in\_tree\ g_n}$ assertion guaranteeing that $g_n$ is in $s$.%notation for gnames
+\ignore{\begin{verbatim}
 Definition ghost_ref g r1 := ghost_reference(P := set_PCM) r1 g.
 Definition in_tree g g1 := EX sh: share, ghost_part(P :=set_PCM) sh (Ensembles.Singleton g1) g.
-\end{verbatim}
-We use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part of the atomic specification respectively.
-The global invariant, which we introduced in our atomic specification as $\treerep$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
-\begin{align*}\treerep(T,g) \triangleq [\ \forall n. \mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
+\end{verbatim}}
+\ignore{The global invariant, which we introduced in our atomic specification as the $\treerep$ predicate above, ties together the pieces as follows:
+\begin{align*}\treerep(T,g) \triangleq *_n. \mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
  Later while writing atomic specification, we will use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part respectively.
@@
 Lemma\ \mathbf{\ node\_exist\_in\_tree : }\ \forall\ g\ s\ g\_in,\  \mathbf{in\_tree}\ g\ g\_in  * \mathbf{ghost\_ref}\ g\ s\ \Rightarrow\ g\_in\ \in s .
-$$
-\subsubsection{Combining all pieces in atomic specification}
+$$}
+\subsubsection{Handle and Abstract State}
+Now we are ready to define the $\treerep$ and $\nodeboxrep$ predicates.
+
+describe lock invariant
+
 We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification.
-The private pre- and postcondition of the atomic specification is given by defining the predicate $\nodeboxrep(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and over-all ghost state is represented by ghost name $g$.
+The private pre- and postcondition of the atomic specification is given by defining the predicate $\nodeboxrep(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and overall ghost state is represented by ghost name $g$.
 The $\nodeboxrep$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
@@
     &\nodeboxrep_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa,g,ga) \\& *\ \nodeboxrep_{0.5}(pb,g,gb))\end{align*}
-
-The $\nodeboxrep$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in Section \ref{safety}.
-In this case, the lock invariant of recursive lock have, along with the information about the children node,$\mathsf{my\_half}$ predicate which holds the partial information describing the current thread's contribution to the shared state.
 
@@
 
-The atomic specification for any binary search tree's operation \texttt{op} can be written as follows:
+The atomic specification for any binary search tree's operation \texttt{op} should then be
 \begin{align*}\forall t.\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t,g)\rangle\ \texttt{op}(p)\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t', g)\rangle
 \end{align*}
-
-Where $t$ and $t'$ are the abstract state of a tree before and after the function's execution.
+where $t$ and $t'$ are the abstract state of a tree before and after the function's execution. Now we are prepared to prove that each of our BST functions satisfies its specification.
 

### 2021-07-13 concurrency/Verification of Concurrent Binary Search Tree.tex
a few more edits

@@
 VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} ($\gnamety$ in Coq) associated with the value $a$, which may be of any type.
-We will see different kind of ghost states used to verify binary search tree in Section \ref{correctness}.
+We will see different kind of ghost states used to verify binary search tree in the following sections.
 
@@
 
-\section{Correctness Proofs}
-\label{correctness}
+\section{BST with Hand-over-Hand Locking}
+\label{hand-over-hand}
 As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
@@
 
-%look at some of the Iris papers in CPP, etc. for how we should describe the proofs
+%look at some of the Iris papers in CPP, etc.
+for how we should describe the proofs, e.g.
+section 4 of Formal Verification of a Concurrent Bounded Queue in a Weak Memory Model, Mevel and Jourdan, ICFP 2021
 
@@
 the last $\mathsf{else}$ clause in C code) in the tree, we need to confirm that the current state of the tree in the global invariant agrees with the fact that the current node, represented by $\texttt{g\_current}$ in the global invariant, actually contains the key-value pair $(k, v)$ in the ghost state represented $\texttt{g\_current}$.
-We accomplish this by opening the global invariant encoded in the $\mathsf{atomic\_shift}$, and proving that lookup satisfy the public postcondition $\mathsf{Q\_p}$ with the help of $\mathsf{sync\_commit_same}$ lemma described in section \ref{atomicity}.
+We accomplish this by opening the global invariant encoded in the $\mathsf{atomic\_shift}$, and proving that lookup satisfy the public postcondition $\mathsf{Q\_p}$ with the help of the $\mathsf{sync\_commit\_same}$ lemma described in section \ref{atomicity}.
 This is the \emph{linearization point} of the $\mathsf{lookup}$ operation and must be done before releasing the current node's lock.
@@
 
+\section{Lock-Free BST}
+We want to prove that a lock-free BST implementation satisfies the same specification as our hand-over-hand implementation.
+Unfortunately, provably-correct deletion in a lock-free setting is a research topic in itself (cite?), so we begin with a lock-free BST that only supports insert and lookup.
+Once again, we want to prove
+\begin{mathpar}
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle
+
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle
+\end{mathpar}
+though the precise definitions of $\nodeboxrep$ and $\treerep$ will differ. %something about how the fact that the specs are the same really comes across to a client -- and should we prove a client?
+
 \section{Related Work}
@@
 But, their proofs are on a toy language rather than real C code.
+
 \section{Conclusion}

### 2021-09-07 concurrency/Verification of Concurrent Binary Search Tree.tex
Small modification in the lookup section.

@@
 
-The code for the $\mathsf{lookup}$ method is shown in \ref{lookup}.
-It takes the location of the root pointer and a key as the arguments.
-A thread spans the tree to find a given key using hand-over-hand locking mechanism.
-Once a thread finds the key in the tree, it gets the value associated with that key, release the current node's lock, and return the value.
+The code for the $\mathsf{lookup}$ method is shown in Figure
+\ref{lookupproof}. It takes the location of the root pointer and a key
+as the arguments. A thread spans the tree to find a given key using
+hand-over-hand locking mechanism. Once a thread finds the key in the
+tree, it gets the value associated with that key, release the current
+node's lock, and return the value.
 
-The atomic specification for $\mathsf{lookup}$ method can be written as follows:
-\begin{align*} \forall t.\ \langle &\nodeboxrep(p,g,g\_root)\ |\ \treerep(t,g,g\_root)\rangle \end{align*} 
+The atomic specification for $\mathsf{lookup}$ method can be written
+as follows:
+\begin{align*} \forall M.\ \langle &\nodeboxrep(p,g,g\_root)\ |\ \treerep(M,g,g\_root)\rangle \end{align*} 
 $$\mathsf{lookup}(p,k)$$ 
-\begin{align*}\langle\lambda v.\ \nodeboxrep(p,g,g\_root)\ |\ (v = \mathsf{lookup}(p,k))\ \&\&\ \treerep(t,g,g\_root)\rangle \end{align*}
+\begin{align*}\langle\lambda v.\ \nodeboxrep(p,g,g\_root)\ |\ (v = M[k])\ \wedge\ \treerep(M,g,g\_root)\rangle \end{align*}
 
-This specification is almost similar to the specification given to the $\mathsf{insert}$ method, except the content of public part in the postcondition.
-To prove the loop inside the lookup method, we need loop-invariant as follows:
+This specification is quite similar to the specification given to the
+$\mathsf{insert}$ method, except the content of public part in the
+postcondition. It says that at some point the $\mathsf{lookup}$
+function will take a map $M$, get the value $M[k]$, and then
+eventually return with the value. Again, it is not guaranteed that the
+state of the map at the end of the execution is the same one when the
+$\mathsf{lookup}$ is called. The key part to verify the specification
+is to prove the loop inside the $\mathsf{lookup}$ method. We define
+the loop-invariant as follows:
 \begin{align*} &\mathsf{lookup\_inv}(b, x, \mathit{g\_root}, \mathit{range},\mathit{info}) \triangleq\ \exists\ \mathit{lock},\ \mathit{g\_current},\ \mathit{np},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{\mathit{g\_current}}\\& *\ R\ np\ * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
 
-Where $P_p$ and $Q_p$ are $\treerep(g\_root, t)$ and $(v = \mathsf{lookup}(p,k))\ \&\&\ \treerep(g\_root, t)$ respectively.
-The proof steps for the $\texttt{lookup}$ verification are almost similar to the steps used in $\texttt{insert}$ verification with few differences which we discuss below.
+Where $P_p$ and $Q_p$ are $\treerep(g\_root, t)$ and $(v =
+M[k])\ \wedge\ \treerep(g\_root, t)$ respectively. The proof steps for
+the $\texttt{lookup}$ verification are almost similar to the steps
+used in $\texttt{insert}$ verification with a few differences which we
+discuss below.
 
-Figure \ref{lookupproof} shows the $\mathsf{lookup}$ function annotated with separation logic specification.
-The proof starts with $\nodeboxrep$ and $\mathsf{atomic\_shift}$ assertions as the precondition and follows the approach similar to the $\mathsf{insert}$ proof for the verification of loop body.
-Once we find the given key (i.e.
-the last $\mathsf{else}$ clause in C code) in the tree, we need to confirm that the current state of the tree in the global invariant agrees with the fact that the current node, represented by $\texttt{g\_current}$ in the global invariant, actually contains the key-value pair $(k, v)$ in the ghost state represented $\texttt{g\_current}$.
-We accomplish this by opening the global invariant encoded in the $\mathsf{atomic\_shift}$, and proving that lookup satisfy the public postcondition $\mathsf{Q\_p}$ with the help of the $\mathsf{sync\_commit\_same}$ lemma described in section \ref{atomicity}.
-This is the \emph{linearization point} of the $\mathsf{lookup}$ operation and must be done before releasing the current node's lock.
-In the case of spanning left or right sub-tree, we need to establish $\mathsf{lookup\_inv}$ at the end $\mathsf{if}$ and $\mathsf{else\ if}$ clause.
-Here, we need to show that the key we are searching is still in the bound for left/right node, so we use $\mathsf{sync\_rollback}$ lemma to extract the bound of a left/right node encoded as the ghost state in the public precondition ($\treerep(g\_root, t)$).
+Figure \ref{lookupproof} shows the $\mathsf{lookup}$ function
+annotated with separation logic specification. The proof starts with
+$\nodeboxrep$ and $\mathsf{atomic\_shift}$ assertions as the
+precondition and follows the approach similar to the $\mathsf{insert}$
+proof for the verification of loop body. Once we find the given key
+(i.e. the last $\mathsf{else}$ clause in C code) in the tree, we need
+to confirm that the current state of the tree in the global invariant
+agrees with the fact that the current node, represented by
+$\texttt{g\_current}$ in the global invariant, actually contains the
+key-value pair $(k, v)$ in the ghost state represented
+$\texttt{g\_current}$. We accomplish this by opening the global
+invariant encoded in the $\mathsf{atomic\_shift}$, and proving that
+lookup satisfy the public postcondition $\mathsf{Q\_p}$ with the help
+of the $\mathsf{sync\_commit\_same}$ lemma described in section
+\ref{atomicity}. This is the \emph{linearization point} of the
+$\mathsf{lookup}$ operation and must be done before releasing the
+current node's lock. In the case of spanning left or right sub-tree,
+we need to establish $\mathsf{lookup\_inv}$ at the end $\mathsf{if}$
+and $\mathsf{else\ if}$ clause. Here, we need to show that the key we
+are searching is still in the bound for left/right node, so we use
+$\mathsf{sync\_rollback}$ lemma to extract the bound of a left/right
+node encoded as the ghost state in the public precondition
+($\treerep(M, g, g\_root)$).
 
-A critical difference from $\texttt{insert}$ is that the $\texttt{lookup}$ method does not change the shape the tree.
-With the help of the $\texttt{sync\_commit\_same}$, we need no longer the heavyweight $\emph{extract}$ lemma but instead a simpler $\emph{ramif}$ lemma to locate the current node in the global invariants:
+A critical difference from $\texttt{insert}$ is that the
+$\texttt{lookup}$ method does not change the shape the tree. With the
+help of the $\texttt{sync\_commit\_same}$, we need no longer the
+heavyweight $\emph{extract}$ lemma but instead a simpler
+$\emph{ramif}$ lemma to locate the current node in the global
+invariants:
 \begin{verbatim}

### 2021-12-13 concurrency/Verification of Concurrent Binary Search Tree.tex
started rewriting proofs

--- 2021-07-13 concurrency/Verification of Concurrent Binary Search Tree.tex
@@
 
+\newcommand{\myhalf}[2]{\ensuremath{\mathsf{my\_half}_{#1}(#2)}}
+\newcommand{\publichalf}[1]{\ensuremath{\mathsf{public\_half}(#1)}}
+
 \newcommand{\ignore}[1]{}
@@
 \author{Alexander Oey}
+\author{Anastasiia Evdokimova}
 \author{Lennart Beringer}
@@
 To prove that an operation on a data structure satisfies a logically atomic specification, we must show that there are no visible intermediate states of the operation, i.e., that other threads see the data structure as unchanged until the linearization point at which the operation takes effect.
-This is often implemented with either a lock-free series of atomic memory accesses, in which case all but the last access must make changes that are considered ``invisible'', or a coarse-grained lock, in which case any changes may be made and no other thread can access the data structure until the operation is complete.
-Fine-grained locking presents an interesting middle ground between these two approaches: other threads may continue to access the data structure as long as they do not require a section that is currently locked, and the visible changes may be implemented with multiple non-atomic operations in the locked section.
+In traditional concurrent separation logics such as VST's~\cite{oraclesematic}, this is done by protecting the data structure with a single lock whose \emph{lock invariant} describes consistent states of the data structure.
+In modern separation logics such as Iris~\cite{iris} or FCSL~\cite{fcsl}, we can make a series of atomic modifications to the data structure without continuously holding a lock, as long as one modification can be designated the \emph{linearization point} at which the new state becomes public, and all prior changes are invisible to other threads.
+Fine-grained locking, in which the pieces of a data structure are each protected by a separate lock with a traditional lock invariant, requires us to synthesize these two approaches: we may make a series of non-atomic modifications to a part of the data structure as long as we hold its lock, but when we release the lock we need to ensure that either we have reached the linearization point, or our changes are invisible to other threads.
 
-In an Iris-style separation logic, the lock-free approach means that each atomic operation gets access to the state of the data structure stored in an atomic shift, and must either maintain its previous state or (at the linearization point) move to a new one; in the coarse-grained approach we can store the state of the data structure in a lock invariant, and the function has access to all of it throughout the critical section.
-With fine-grained locking, neither of these approaches quite applies as-is: there are no atomic operations by which we access the state of the data structure, but neither is there a lock whose invariant controls access to the entire structure.
-However, we can take a modified form of the lock-free approach, where we treat each critical section as an atomic operation that can access the atomic shift.
-Typically, we will perform a sequence of non-atomic operations in the body of the critical section without access to the global state of the data structure, and then (before exiting the critical section) access the atomic shift and show that we have either maintained the original state or satisfied the postcondition of the operation.
-In this section, we describe a general approach to associating fine-grained locks with pieces of the abstract state of a data structure, allowing them to interact with atomic shifts and be used to prove atomic specifications.
-(Note that this approach is not novel to us; it is adapted from something Ralf Jung worked on.
-We need to be more clear about the relationship.)
- We use the following rules to associate fine-grained locks with abstract state of a data structure:
-\begin{align*}
-    \textsf{sync\_inv}(g,sh,R) \triangleq  \exists\ a,\ R\ g\ a\ *\ \textsf{my\_half}(g,sh,a)   \end{align*}
-$$\inference[\textsf{sync\_commit}]{\forall a, R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * \exists x_0'\ x_1'.\ (x_0, x_1) \leadsto (x_0', x_1') \land \\(\mathsf{my\_half}(g, \mathit{sh}, x_0') * \mathsf{public\_half}(g, x_1') \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) *  \textsf{sync\_inv}(g,sh,R) \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
-$$\inference[\textsf{sync\_rollback}]{\forall a, R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{public\_half}(g, x_1) \Rrightarrow P_p\ a * R')}{\begin{array}{c}\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{sync\_inv}(g,sh,R)  \Rrightarrow \\\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{sync\_inv}(g,sh,R')\end{array}}$$
+We accomplish this using a technique derived from unpublished work by Jung~\cite{ralf-convo} for treating the critical sections of locks as atomic operations.
+The key is to include in each lock invariant a piece of ghost state that represents the abstract state of the locked section.
+From the outside, the effect of the critical section is to atomically update that piece of ghost state, and all other changes are internal to the lock.
+These pieces of ghost state can then be composed to yield the abstract state of the entire data structure.
+In effect, this construction turns a traditional lock invariant into a \emph{lock variant} $\ell \islock R(a)$, where the value of $a$ can be updated in the critical section as long as it corresponds to a linearization point or invisible change to the abstract state of the data structure as a whole.
+This allows traditional lock invariants to interact with atomic shifts and be used to prove atomic specifications.
 
-The $\mathsf{sync\_inv}$ predicate associates the contribution part of ghost state $a$ for a node with ghost name $g$, along with the other concrete information about a node parameterized on $a$.
-In the atomic specification, we protect the $\mathsf{sync\_inv}$ using lock inside lock invariant.
-So, once we acquire the lock, we get access to $\mathsf{sync\_inv}$.
-The $\mathsf{sync\_commit}$ and $\mathsf{sync\_rollback}$, specialized with ghost state with \emph{reference pattern} ($\mathsf{my\_half}$ and $\mathsf{public\_half}$), are the versions of $\mathsf{atomic\_commit}$ and $\mathsf{atomic\_rollback}$ described in Section \ref{background}.
-In a commit, we must provide $\mathsf{sync\_inv}$ with resources $R$ that, in combination with the public precondition, allow us to prove the public postcondition; we then gain access to an assertion Q required by the postcondition of the function, and leftover resources $R'$.
-The state of the node $g$ in public precondition (described by an assertion $\mathsf{public\_half}$) must be joinable with the state described by an assertion $\mathsf{sync\_inv}$.
-In a rollback, we provide $\mathsf{sync\_inv}$ with resources $R$ that, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as $\mathsf{sync\_inv}$ with the leftover resources $R$.
-We must use these rules before releasing the node's lock (we lose the $\mathsf{sync\_inv}$ after releasing the lock).
+The construction of the lock variants is simple.
+We begin with a \emph{part-reference} ghost state, a common construction for sharing information between an invariant and the threads that act on it, with elements $\publichalf{a}$ (the current state $a$) and $\myhalf{\pi}{p}$ (partial information about the current state, annotated with a share $\pi$).
+When the share $\pi$ is 1, we know that $p = a$.
+(do we want to define this ourselves, or reference it somewhere?
+note also that this is what Iris calls an authoritative algebra) A lock variant with an assertion $R(a)$ is then defined as %notation
+$$\ell \islock R(a) \triangleq \ell \islock \exists a.\ R(a) * \myhalf{\pi}{a}$$ %eliding ghost names here
+The share $\pi$ may be set to 1 if the lock's information about $a$ is always fully up to date, or a smaller share if it can become outdated due to operations on other parts of the data structure.
+Since $a$ is existentially quantified, a thread that changes the value of $a$ forgets it after releasing the lock (as required in a lock invariant), but it must have synchronized with a $\publichalf{a}$ held elsewhere (usually in a global invariant) that will remember the change.
+Thus, the ghost state acts as a bridge between the component lock invariant and the global abstract state of the data structure.
+%smaller example before we get into BST?
+
+From the atomic rules of section~\ref{atomic}, we derive the following rules for accessing and modifying pieces of the global abstract state:
+\begin{mathpar}
+\inference[\textsf{sync\_commit}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * \exists x_0',\,x_1'.\ (x_0, x_1) \leadsto (x_0', x_1')\ \land \\(\mathsf{my\_half}(g, \mathit{sh}, x_0') * \mathsf{public\_half}(g, x_1') \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}
+
+\inference[\textsf{sync\_rollback}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{public\_half}(g, x_1) \Rrightarrow P_p\ a * R')}{\begin{array}{c}\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R  \Rrightarrow \\\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R'\end{array}}
+\end{mathpar}
+A thread that holds $\myhalf{}{a}$ (i.e., is in the critical section of a lock with a lock variant) may interact with the abstract state $P_p$ held in an $\textsf{atomic\_shift}$, as long as the corresponding $\publichalf{}$ is part of the abstract state.
+It may either leave both halves as is and gain updated resources $R'$ (including, e.g., a snapshot of the current abstract state), or update both halves to a new state that satisfies the public postcondition $Q_p$, executing the linearization point and obtaining the final postcondition $Q$.
+Typically we will use one of these rules at the end of each critical section, to show that the changes made in the critical section either are invisible to other threads (\textsf{sync\_rollback}) or have fulfilled the linearization point (\textsf{sync\_commit}).
+We can also derive simpler rules for special cases, such as a version of \textsf{sync\_commit} for read-only operations that do not change the state of any component:
+
+$$\inference[\textsf{sync\_commit\_same}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{my\_half}(g, \mathit{sh}, x_0) * \mathsf{public\_half}(g, x_1) \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
 
 \subsection{Specification of the hand-over-hand BST}
-%In addition to program state, the verification of concurrent program involves two aspects which do not appear in actual C program: the \emph{shares} on shared resources and \emph{ghost state} described in Section \ref{safety}.
-We need to extend our specification in Section\ref{safety} with \emph{ghost state} to prove the functional correctness.
+%rewrite to use the above: this is the global state made up of components, etc.
 
@@
 
-\subsection{Insert and Lookup}
+\subsection{Proofs: Insert and Lookup}
+The basic verification style in VST is to divide proofs into two parts: proofs that the C implementations meet specifications that are Coq-level encodings of their behavior, and proofs that those encodings actually have the desired properties.
+In the single-threaded case, we might prove that some code implements a binary search tree written in Gallina, and then prove that that tree satisfies map axioms.
+Similarly, we divide our concurrency proofs into two parts: code-specific proofs that describe how the C code implements changes to per-node ghost state, and more abstract proofs describing how changes in local ghost state lead to the desired updates to the global state.
 
-\subsubsection{insert}
-The C outline-code for the insert method of binary search tree is shown in \ref{insert}.
-A node has a lock, key, value, and pointers to the left and right child.
-Each leaf node in tree are empty node with lock.
-Whenever a thread try to insert key-value pair in the tree, it first spans the tree to find the right position for new key using hand-over-hand locking mechanism; acquire the lock for child before releasing the lock for node.
-After locating right leaf node to insert new key-value, thread creates two new empty leaf nodes, insert key-value in the current node, and link the newly created leaf nodes to the current node.
-If the key already exist in the tree, then thread simply swaps the old value associated with node with the new value keeping the child pointers as it is.
+\subsubsection{Insert}
+An outline of the C code for the insert function is shown in figure \ref{insert-proof}.
+When a thread inserts a key $k$ into a tree, it first crawls the tree to find the right position for $k$ using the hand-over-hand locking mechanism, acquiring the lock for the next node before releasing the lock for the current node.
+If it encounters a node with key $k$, then it simply changes the value at that node.
+Otherwise, it will eventually reach a leaf node, where it creates two new empty leaf nodes, inserts the key and value into the current node, and links the newly created leaf nodes to the current node.
+The atomic specification for insert method is $$\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle$$
+The \texttt{insert} function should atomically add the key $k$ with value $v$ to the tree at $p$.
 
-The atomic specification for insert method can be written as follows:
-$$\forall t.\ \langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(t,g)\rangle $$ 
-$$\texttt{insert}(p,k,v)$$
-$$\langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(t[k\mapsto v],g)\rangle $$
-We are not guaranteed that the state of  a tree at the end of the insertion will be $t[k\mapsto v]$ where $t$ is the tree state when insert is called; rather, we know that p always holds some tree during the function's execution, and at some point the function will take that tree $t$, add $[k\mapsto v]$, and then eventually return, while in the meantime another thread may have modified the state of tree from $t[k\mapsto v]$ to any other arbitrary state that the current thread do not know.
+The key to the verification of the C code is the loop invariant for the top-level loop. We define the loop invariant for the insert function as follows: 
+\begin{align*} &\mathsf{insert\_inv}(b, x, g, g\_root) \triangleq\ \exists\ \mathit{lock},\mathit{g\_current},\mathit{np},\mathit{range},\mathit{info},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{g\_current}\ *\ R\ \mathit{np}\\& * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,g,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
+where $P_p$ and $Q_p$ are $\treerep(t,g)$ and $\treerep(t[k\mapsto v],g)$ respectively.
+The resources $R$, parameterized by a node pointer $\mathit{np}$, describes the concrete information about a node.
+Also, $x$ is the key to be inserted that must be in the bound $\mathsf{range}$.
+This loop invariant has some existential variables which characterize the state of each iteration of the loop.
 
-Since we have given the specification that is strong enough to specify both the safety and the functional correctness properties of insert method, we can now verify that body of the function satisfies the atomic specification given above using various VST and Iris (encoded in VST) tactics.
-Once we have outermost specification for a function, the another creative part in verification is finding the right loop-invariant.
-We defines the loop invariant for the insert function as follows:
-\begin{align*} &\mathsf{insert\_inv}(b, x, g, g\_root) \triangleq\ \exists\ \mathit{lock},\mathit{g\_current},\mathit{np},\mathit{range},\mathit{info},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{g\_current}\ *\ R\ \mathit{np}\\& * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,g,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
-
-Where $\mathsf{P_p}$ and $\mathsf{Q_p}$ are $\treerep(t,g)$ and $\treerep(t[k\mapsto v],g)$ respectively.
-The resources $\mathsf{R}$, parameterized by a node pointer $np$, describes the concrete information about a node.
-Also, $x$ is the key to be inserted that must be in the bound $\mathsf{range}$.
-This loop invariant has some existential variables which characterized the state of each iteration of loop in the C code.
-
-Figure \ref{insertproof} shows the $\texttt{insert}$ function annotated with separation logic specification.
+Figure \ref{insertproof} shows the $\texttt{insert}$ function with separation logic annotations.
 The proof starts with $\nodeboxrep$ and $\mathsf{atomic\_shift}$ as the precondition.
 After acquiring the lock for the root node, the thread accesses the information inside the lock invariant of the root node.
-In order to prove $\mathsf{while}$ loop, we show that the precondition satisfies the $\mathsf{insert\_inv}$, then prove that the loop body preserves the loop invariant ( cases inside $\mathsf{else if}$ clauses at line 15 and 17 in \ref{insert}).
-Once we locate the empty node for inserting new key-value pair (i.e inside the first $\mathsf{if}$ clause), we open the global invariant encoded in the $\mathsf{atomic\_shift}$, create the ghost names $\texttt{g1}$ and $\texttt{g2}$ for two new empty child nodes of current node, and prove that insertion satisfy the public postcondition $\treerep(t[k\mapsto v],g)$ with the help of $\mathsf{sync\_commit}$ rule described in section \ref{atomicity}.
-This is the \emph{linearization point} of the $\mathsf{insert}$ operation and must be done at any point in the proof, but before releasing the current node's lock (i.e in critical section).
-In order to prove the public postcondition $\treerep(t[k\mapsto v],g)$, we need to first locate the current node in $\treerep(t,g)$ using ghost name $g\_current$, and then perform necessary modification such that the final state of the global invariant satisfies $\treerep(t[k\mapsto v],g)$.
-We create and prove separate lemma for this, called $\emph{extract}$ lemma which is depicted in the \ref{extract_insert}.
+In order to prove $\mathsf{while}$ loop, we show that the precondition satisfies the $\mathsf{insert\_inv}$, then prove that the loop body preserves the loop invariant (cases inside $\mathsf{else if}$ clauses at line 15 and 17 in \ref{insert}).
+Once we locate the empty node for inserting the new key-value pair (inside the first $\mathsf{if}$ clause), we open the global invariant encoded in the $\mathsf{atomic\_shift}$, create the ghost names $\texttt{g1}$ and $\texttt{g2}$ for two new empty child nodes of current node, and use the $\mathsf{sync\_commit}$ rule---we have reached the linearization point and completed the insertion.
+The premise of $\mathsf{sync\_commit}$ requires that we have enough information to know that inserting the key at this point will change the abstract state from some unknown $t$ to $t[k\mapsto v]$, which we show in the second part of the proof.
 
-With the help of $\mathsf{in\_tree}$ assertion, we locate a leaf node denoted by ghost name $g\_current$ in the global state $\treerep(t,g)$.
-Then we prove the \emph{extract} lemma, which says that the state of tree after adding new key-value pair (k,v) will be $\treerep(t[k\mapsto v],g)$, by applying induction in $t$.
-
-Another important step in the $\mathsf{insert}$ proof is the case when we find that the key to be inserted already exists in the tree (i.e.
-the last $\texttt{else}$ clause in C code).
-The modification of the global state of the tree t is shown in \ref{extract_insert2}.
+The other branch of the code-level proof is the case where the key to be inserted already exists in the tree (the last $\texttt{else}$ clause in the code).
+%The modification of the global state of the tree t is shown in \ref{extract_insert2}.
 In that case, the old value $v'$ associated with key is overridden by new value $v$ keeping a key $40$ and child pointers as as they are.
@@
 We augment the same $\emph{extract}$ lemma with the clause that states the scenario in \ref{extract_insert2}, and prove the lemma in the same induction over the tree $t$.
+In this case, we change the value at the node but leave the rest of the tree intact, and this is the linearization point for the insertion.
+In the second part of the proof, we will show that this case also transforms $t$ to $t[k\mapsto v]$.
 
-In the rest two cases, the $\texttt{insert}$ method compares the key
-of the current node with the key $\texttt{x}$ and goes down into the
-corresponding left or right branch. We need to show that the range in
-the ghost state shrinks correctly. To achieve this goal we open the
-global invariant encoded in the $\mathsf{atomic\_shift}$ to retrieve
-the necessary information. But this time these operations maintain the
-original state. For the left branch case, we accomplish the proof by
-the following lemma:
+In the other two cases, we compare the key
+of the current node with the new key \texttt{x} and move to the left or right child accordingly.
+Here we again need to access the atomic shift, to ensure that the range of the node we move to is included in the range of the current node.
+We can use $\mathsf{sync\_rollback}$ to retrieve this information without making any changes to the abstract state of the tree.
+%For the left branch case, we accomplish the proof by the following lemma:
 %\begin{verbatim}
@@
 %\end{verbatim}
-The lemma for the right branch case is similar. Both use the
-$\texttt{sync\_rollback}$ lemma described in section \ref{atomicity}.
+%The lemma for the right branch case is similar. Both use the
+%$\texttt{sync\_rollback}$ lemma described in section \ref{atomicity}.
 
@@
 
-\subsubsection{lookup}
+\paragraph{Changing the Global State} describe ``extract lemmas'' here
 
-The code for the $\mathsf{lookup}$ method is shown in \ref{lookup}.
-It takes the location of the root pointer and a key as the arguments.
-A thread spans the tree to find a given key using hand-over-hand locking mechanism.
-Once a thread finds the key in the tree, it gets the value associated with that key, release the current node's lock, and return the value.
+\subsubsection{Lookup}
 
-The atomic specification for $\mathsf{lookup}$ method can be written as follows:
-\begin{align*} \forall t.\ \langle &\nodeboxrep(p,g,g\_root)\ |\ \treerep(t,g,g\_root)\rangle \end{align*} 
-$$\mathsf{lookup}(p,k)$$ 
-\begin{align*}\langle\lambda v.\ \nodeboxrep(p,g,g\_root)\ |\ (v = \mathsf{lookup}(p,k))\ \&\&\ \treerep(t,g,g\_root)\rangle \end{align*}
+The code for the $\mathsf{lookup}$ method is shown in Figure
+\ref{lookupproof}. It takes the location of the root pointer and a key
+as the arguments. A thread spans the tree to find a given key using
+hand-over-hand locking mechanism. Once a thread finds the key in the
+tree, it gets the value associated with that key, releases the current
+node's lock, and returns the value.
+The atomic specification for $\mathsf{lookup}$ is
+$$\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle$$
+The \texttt{lookup} function should atomically look up the value of the key $k$ in the tree at $p$.
 
-This specification is almost similar to the specification given to the $\mathsf{insert}$ method, except the content of public part in the postcondition.
-To prove the loop inside the lookup method, we need loop-invariant as follows:
+As before, the key to the verification is the invariant for the main loop:
 \begin{align*} &\mathsf{lookup\_inv}(b, x, \mathit{g\_root}, \mathit{range},\mathit{info}) \triangleq\ \exists\ \mathit{lock},\ \mathit{g\_current},\ \mathit{np},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{\mathit{g\_current}}\\& *\ R\ np\ * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
+where $P_p$ and $Q_p$ are $\treerep(g\_root, t)$ and $(v =
+M[k])\ \wedge\ \treerep(g\_root, t)$ respectively. The proof steps for
+the $\texttt{lookup}$ verification are similar to the steps
+used in $\texttt{insert}$ verification, with a few differences which we
+discuss below.
 
-Where $P_p$ and $Q_p$ are $\treerep(g\_root, t)$ and $(v = \mathsf{lookup}(p,k))\ \&\&\ \treerep(g\_root, t)$ respectively.
-The proof steps for the $\texttt{lookup}$ verification are almost similar to the steps used in $\texttt{insert}$ verification with few differences which we discuss below.
+Figure \ref{lookupproof} shows the $\mathsf{lookup}$ function
+with separation logic annotations. The proof starts with
+$\nodeboxrep$ and $\mathsf{atomic\_shift}$ assertions as the
+precondition and follows the same approach as the $\mathsf{insert}$
+proof for the verification of loop body. Once we find the target key
+(the last $\mathsf{else}$ clause in the code), we need
+to confirm that the value present in the tree corresponds with the value we would find if we looked up the key in the current abstract state. We accomplish this by opening the global
+invariant encoded in the $\mathsf{atomic\_shift}$, and proving that
+lookup satisfies the public postcondition $\mathsf{Q\_p}$ with the help
+of the $\mathsf{sync\_commit\_same}$ lemma described in section
+\ref{atomicity}. This is the linearization point of the
+$\mathsf{lookup}$ operation and must be done before releasing the
+current node's lock. When we move into the left or right sub-tree,
+we need to establish $\mathsf{lookup\_inv}$ at the end of the $\mathsf{if}$
+or $\mathsf{else\ if}$ clause. Here, we need to show that the key we
+are searching for is still in the range of our new node, so we use
+$\mathsf{sync\_rollback}$ to extract the bound of the left/right
+node encoded as the ghost state in the public precondition
+($\treerep(M, g, g\_root)$).
 
-Figure \ref{lookupproof} shows the $\mathsf{lookup}$ function annotated with separation logic specification.
-The proof starts with $\nodeboxrep$ and $\mathsf{atomic\_shift}$ assertions as the precondition and follows the approach similar to the $\mathsf{insert}$ proof for the verification of loop body.
-Once we find the given key (i.e.
-the last $\mathsf{else}$ clause in C code) in the tree, we need to confirm that the current state of the tree in the global invariant agrees with the fact that the current node, represented by $\texttt{g\_current}$ in the global invariant, actually contains the key-value pair $(k, v)$ in the ghost state represented $\texttt{g\_current}$.
-We accomplish this by opening the global invariant encoded in the $\mathsf{atomic\_shift}$, and proving that lookup satisfy the public postcondition $\mathsf{Q\_p}$ with the help of the $\mathsf{sync\_commit\_same}$ lemma described in section \ref{atomicity}.
-This is the \emph{linearization point} of the $\mathsf{lookup}$ operation and must be done before releasing the current node's lock.
-In the case of spanning left or right sub-tree, we need to establish $\mathsf{lookup\_inv}$ at the end $\mathsf{if}$ and $\mathsf{else\ if}$ clause.
-Here, we need to show that the key we are searching is still in the bound for left/right node, so we use $\mathsf{sync\_rollback}$ lemma to extract the bound of a left/right node encoded as the ghost state in the public precondition ($\treerep(g\_root, t)$).
-
-A critical difference from $\texttt{insert}$ is that the $\texttt{lookup}$ method does not change the shape the tree.
-With the help of the $\texttt{sync\_commit\_same}$, we need no longer the heavyweight $\emph{extract}$ lemma but instead a simpler $\emph{ramif}$ lemma to locate the current node in the global invariants:
+A critical difference from $\texttt{insert}$ is that the
+$\texttt{lookup}$ method does not change the shape of the tree. With the
+help of $\texttt{sync\_commit\_same}$, we need no longer the
+heavyweight $\emph{extract}$ lemma but instead a simpler
+$\emph{ramif}$ lemma to locate the current node in the global
+invariants:
 \begin{verbatim}

### 2021-12-14 concurrency/Verification of Concurrent Binary Search Tree.tex
Add content for delete in writeup

@@
 %\usepackage{lstlangcoq}
+\usepackage[edges]{forest}
 \renewcommand{\lstlistingname}{Figure}
@@
 
+The code for the delete method is shown in ??. It takes the location of the root pointer
+and a key as the arguments. A thread traverses the tree using hand-over-hand locking,
+until the thread finds the key to be deleted. To remove the key, the tree is changed 
+through an operation called pushdown left such that the node to be deleted points to an empty leaf
+right child. To achieve that, the thread sets the right child's left child as the 
+new right child of the to-be-deleted node and moves the right child's right child as 
+the new parent of the to-be-deleted node. This operation is repeated until the right
+child of the to-be-deleted node is an empty leaf node. Then, the thread will delete
+the node and set the delete node's parent pointer to the left child of the deleted node.
+
+The atomic specification for delete method can be written as follows:
+$$\forall t.\ \langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(t,g)\rangle $$ 
+$$\texttt{delete}(p,k)$$
+$$\langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(\texttt{delete}(t, k),g)\rangle $$
+
+The delete method is unique compared to insert and lookup because the pushdown left 
+operation can change the \textit{public} half range of more than one nodes. In addition,
+the pushdown left operation is one of the main reasons why the ghost tree of the binary
+search tree is not just a ``copy" of the pure tree: the functional model of pushdown\_left
+is unable to account for the change of deleted node's parent's pointer from pointing
+at the deleted node the to the deleted node's right child.
+
+\usetikzlibrary{positioning}
+
+\forestset{
+   lab/.style={
+        label={[anchor=west, scale=0.75] north east:#1},
+    },
+}
+\begin{forest}
+for tree={
+    grow=south,
+    circle, draw, minimum size=3ex, inner sep=1pt,
+    s sep=3mm
+        }
+    % my label/.style={
+        % label={[anchor=south west] north east:#1},
+    % },
+    [, phantom, for children={fit=band}, s sep+=20mm,
+         before drawing tree={%
+          tikz+={%
+            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
+            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
+            \node [anchor=center] at ($(a.east)!1/2!(b.west)$) {$\longrightarrow\quad$};
+          },
+        },
+        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
+            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
+                    [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}]
+                    [55, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,75)$};}
+                        [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(40,55)$};}]
+                        [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
+                    ]
+                ]
+                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
+            ]
+        ]
+        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
+            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                [55, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
+                    [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
+                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(40,35)$};}]
+                        [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,55)$};}]
+                    ]
+                    [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
+                ]
+                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
+            ]
+        ]
+        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
+            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                [55, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
+                    [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                            [,no edge, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
+                ]
+                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
+            ]
+        ]
+        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
+            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                [55, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
+                    [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
+                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                        [,no edge, draw=none]
+                    ]
+                    [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
+                ]
+                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
+            ]
+        ]
+    ]
+\end{forest}
+
+\subsubsection{delete}
+
+Similar with insert, delete starts by acquiring the root node lock, so that the thread
+can access the information inside the lock invariant of the root node. Then, we prove
+the while loop by showing that the precondition satisfies delete\_inv and the 
+loop body preserves the loop invariant. The first two cases are similar to insert and
+lookup in which the thread finds the target node, traversing by hand-over-hand locking.
+Once the target node is located, the thread will call pushdown\_left and return after.
+Using the specification of pushdown\_left, we show that the post-condition of 
+pushdown\_left implies the loop invariant. Note that before going into pushdown\_left,
+the thread is holding the lock to the target node and will return to the delete
+method without holding any.
+
+\subsubsection{pushdown\_left}
+
+The proof to pushdown\_left is the key step of proving the delete method. 
+First, we show that the precondition of pushdown\_left satisfies the precondition
+of the loop invariant, which includes the assertion that the thread is holding the 
+lock to the deleted node (i.e. the information inside the lock invariant is accessible)
+and an atomic\_shift. Then, the thread accesses that information to acquire a second 
+lock to the right child of the deleted node. Due to the hand-over-hand locking 
+mechanism this will not cause a deadlock since the thread has already acquired the 
+lock to the deleted node. Next, we prove that the action of deleting the node,
+with sync\_commit by opening the global invariant and removing the relevant 
+ghost names (the target node and its right sentinel node), satisfies the loop
+invariant and the postcondition of pushdown\_left.
+
+The second part of pushdown\_left is proving that the turn\_left operation satisfies
+the loop invariant. Using a variant of the sync\_rollback rule, we show that the 
+turn\_left operation preserves the BST, apart from changes in the ranges of two nodes,
+the target node and its right child, and information about the nodes in those two nodes.
+To satisfy the structure of the ranges, we only need to prove that the new range for
+the target node is encompasses the old range of both nodes. This will imply that
+the range of the target node's old parent includes the target node's right child 
+new range as the right child becomes the new parent (and the old parent points to 
+the right child).
+
+One key step of the proof is proving that two ghost trees are equivalent if the union
+of its ghost nodes are equal and the ghost tree is sorted. This allows the global invariant
+to be preserved during the turn\_left operation in which no nodes were actually changed
+except for the structure of the tree.
+
 \section{Lock-Free BST}

### 2021-12-29 concurrency/Verification of Concurrent Binary Search Tree.tex
updated todo list
started describing lock invariants and global ghost state

@@
 \newcommand{\islock}{\boxdotright}
+\newcommand{\lockvar}{\islock^{v}}
 \newcommand{\isaex}{\!\mathrel\odot\joinrel\rightarrow\!}
@@
 These pieces of ghost state can then be composed to yield the abstract state of the entire data structure.
-In effect, this construction turns a traditional lock invariant into a \emph{lock variant} $\ell \islock R(a)$, where the value of $a$ can be updated in the critical section as long as it corresponds to a linearization point or invisible change to the abstract state of the data structure as a whole.
+In effect, this construction turns a traditional lock invariant into a \emph{lock variant} $\ell \lockvar R$, where $R$ is parameterized by a value that can be updated in the critical section, as long as it corresponds to a linearization point or invisible change to the abstract state of the data structure as a whole.
 This allows traditional lock invariants to interact with atomic shifts and be used to prove atomic specifications.
@@
 note also that this is what Iris calls an authoritative algebra) A lock variant with an assertion $R(a)$ is then defined as %notation
-$$\ell \islock R(a) \triangleq \ell \islock \exists a.\ R(a) * \myhalf{\pi}{a}$$ %eliding ghost names here
+$$\ell \lockvar_\pi R \triangleq \ell \islock \exists a.\ R(a) * \myhalf{\pi}{a}$$ %eliding ghost names here
 The share $\pi$ may be set to 1 if the lock's information about $a$ is always fully up to date, or a smaller share if it can become outdated due to operations on other parts of the data structure.
@@
 In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure.
-Our overall approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$.
+As described above, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$.
 We also need to make sure that our lock invariants support the hand-over-hand pattern.
 
-\subsubsection{Ghost States and Range}
- The main challenge in proving the properties of concurrent data structure is to abstract the data structure in a way that helps to reason about it in the concurrent setting.
-Krishna et al.
-\cite{krishna2017flow} mention in their paper that we can abstract binary search tree using appropriate bounds in each node: we can use the product of \emph{lower-bound} and \emph{upper-bound} on each node of binary tree by propagating from each node to its child the appropriate bounds on the values presents in the sub-tree.
-The \emph{lower-bound} and \emph{upper-bound} help us to prove that we perform the operation in the right node: for instance, if we $\mathsf{insert}$ a key in the tree, then key should be inside those two bounds.
-We use the term \emph{range} to represent the product of lower- and upper-bound.
-\ref{range_bst} shows an example of BST with each node labeled with appropriate bound propagated all the way from \emph{root} to empty \emph{leaf} nodes.
+\subsubsection{Key Ranges}
+When we traverse the BST looking for a key $k$, we will reach either a node containing $k$, or an empty node where $k$ would appear if it was in the tree.
+To prove correctness of this procedure, we need ghost state that tracks where each key ``should appear'' in the current tree.
+As observed by \citet{krishna2017flow}, this can be done by associating each node with a lower and upper bound on the keys appearing in the subtree rooted at that node.
+We refer to the product of these bounds as a \emph{range}.
+The range of a node is inherited from its parent, based on the parent's key: if node $n$ has range $(l, r)$ and key $k$, then its left child has range $(l, k)$ and its right child has range $(k, r)$.
+Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty \emph{leaf} nodes.
 \begin{figure}[htb]
@@
 \end{figure}
-The range for root node is always $(-\infty,+\infty)$ and then, we can calculate the range for all nodes in the tree by enforcing each node to propagates appropriate bounds.
-For the left child, range will be lower-bound for the node (lower-bound) and the key in the node (upper-bound), while for the right child range will be a key in the node (lower-bound) and upper-bound for the node (upper-bound).
-This properties in each node help us to prove that the thread have changed the state of data-structure to the valid state.
-For instance, we can confirm that a key has been inserted in the right place if a key is inside the range of the node.
-Suppose we want to insert a key $38$ in \ref{range_bst}, then we will locate \emph{right} child of $35$ as the right place for $38$.
-During insertion, we can confirm that the node is indeed right place for $38$ by checking that $38$ is in the range (35,40).
-We introduce the $range$ in our proof using the ghost variables.
-This $ranges$ along with other information about the node represent the abstract state of the data structure.
-We have used two kind of ghost state in our proof: \emph{per-node} ghost state and \emph{overall} ghost state.\\
-\textbf{Per-node Ghost State}: The per-node ghost state is includes a node's range and other information about the node: its current key, value, and ghost names of its child nodes.
-In VST, we can use a type as ghost state by proving an instance of the \texttt{Ghost} typeclass, analogous to the \emph{resource algebras} of Iris.
-The essential part is a \emph{join} relation describing how two pieces of ghost state combine.
-We define the join of two ranges to be the range that contains both of them, and use an \emph{exclusive} algebra for the rest of the data, so that only one thread knows the current state of a node at a time.
-Ghost state is shared between the per-node lock invariants and the abstract state of the data structure using the \emph{reference} pattern: each thread holds partial information describing its contribution to the shared state, and the shared resources hold a ``reference'' copy that records all of the contributions.
-In VST, this is implemented with assertions $\mathsf{my\_half}\ g\ \mathit{sh}\ a$ and $\mathsf{public\_half}\ g\ r$, where $g$ is the node's identifier ($\gnamety$), $a$/$r$ is the value of the ghost state, and $\mathit{sh}$ is a share describing how much of the reference value is known: if $\mathit{sh}$ is the full share, then we know that $a = r$.
-In the proof, the \texttt{my\_half} assertion for a node will be held by the node's lock, and \texttt{public\_half} will be held by the shared state.
+At the leaves, these ranges partition the space of all possible keys.
+If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.
+For example, suppose we want to insert a key $38$ into the tree in Figure~\ref{range_bst}.
+The right child of the node with key 35 is a leaf with range $(35, 40)$, so it is guaranteed to be the right place to insert 38.
 
-\textbf{Overall Ghost State}: We now have per-node ghost state for each node, indexed by a ghost name $g_n$.
-But if we are given an arbitrary $\gnamety$ for a node, we do not know whether that $\gnamety$ exists in the tree.
-We need another kind of ghost state to hold the set of all nodes in the current tree.
-Again we can use the ``reference'' pattern to define this ghost state, where the abstract state $\mathsf{ghost\_ref}\ s$ holds the full set of nodes, and each node's lock invariant holds an $\mathsf{in\_tree\ g_n}$ assertion guaranteeing that $g_n$ is in $s$.%notation for gnames
-\ignore{\begin{verbatim}
-Definition ghost_ref g r1 := ghost_reference(P := set_PCM) r1 g.
-Definition in_tree g g1 := EX sh: share, ghost_part(P :=set_PCM) sh (Ensembles.Singleton g1) g.
-\end{verbatim}}
-\ignore{The global invariant, which we introduced in our atomic specification as the $\treerep$ predicate above, ties together the pieces as follows:
-\begin{align*}\treerep(T,g) \triangleq *_n. \mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
- Later while writing atomic specification, we will use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part respectively.
-Once we open \emph{global invariant}, we get access to $\mathsf{ghost\_ref}$ which will be used with $\mathsf{in\_tree}$ to establish the fact that node actually exists in the tree.
-We proved the following lemma for this:
-$$
-Lemma\ \mathbf{\ node\_exist\_in\_tree : }\ \forall\ g\ s\ g\_in,\  \mathbf{in\_tree}\ g\ g\_in  * \mathbf{ghost\_ref}\ g\ s\ \Rightarrow\ g\_in\ \in s .
-$$}
-\subsubsection{Handle and Abstract State}
-Now we are ready to define the $\treerep$ and $\nodeboxrep$ predicates.
+\subsubsection{Per-Node and Global Ghost State}
+Importantly, the fact that the leaf ranges partition the space of possible keys remains true even if operations in other threads restructure the tree during our traversal.
+Furthermore, operations further up the tree can only \emph{increase} the range of nodes below them, so if we find a node whose range includes our key, we are guaranteed to have found the right place for it.
+This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations.
+More precisely, the abstract state of a node in the tree consists of its range plus a representation of its contents: for a leaf node, $\mathsf{None}$, and for an internal node, $\mathsf{Some}$ of its current key, value, and the identifiers of its left and right children.
+This per-node abstract state serves as a bridge between the per-node locks of the implementation and the abstract state of the entire tree (the $\treerep$ predicate) in the specification.
 
-describe lock invariant
+The abstract state of a tree is a map from keys to values.
+The $\treerep$ predicate then describes the conditions under which a map $m$ is implemented by the ghost state of a tree.
+Roughly speaking, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$.
+We formalize this by first defining the collection of $\mathsf{public\_half}$ assertions---i.e., the authoritative copies of the per-node ghost states---needed to form an abstract tree:
+\begin{align*}
+\mathsf{public}&\mathsf{\_halves}(t, g, (l, r)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
+& |\ \mathsf{Leaf} \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{None})\\
+& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{Some}\ (k, v, g_l, g_r)) * \mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
+\end{align*}
+We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{public\_half}$ assertions of each node into a single assertion.
+We compute the range of each node's children as described above: if the key in the current node is $k$, then the left child has range $(l, k)$ and the right child has range $(k, r)$.
 
+This collection of $\mathsf{public\_half}$ assertions can be combined with per-node lock variants to ensure that the nodes in an implementation actually form a valid tree.
+However, we also need to record the fact that each node in the implementation actually belongs to the tree $t$.
+We do this by augmenting $\mathsf{public\_halves}$ with a piece of ghost state $\mathsf{ghost\_nodes}(S)$ that records the set of all nodes in the tree, constructed so that each node can hold an $\mathsf{in\_tree}\ g$ assertion guaranteeing that $g \in S$.
+Then our final definition of $\treerep$ is:
+$$\treerep(m, g) \triangleq \exists t.\ t \text{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty)) * \mathsf{ghost\_nodes}(\mathsf{node\_names}(t))$$
+The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in the ghost tree $t$.
+Because we quantify over the abstract tree $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node names and key-value pairs.
+(We will take advantage of this property when verifying the \lstinline{delete} function.)
+
+As described in section~\ref{atomicity}, we connect the per-node locks of our C implementation with the global abstract state of $\treerep$ by associating each node's lock with a lock variant assertion $\ell \islock_\pi R$, where $R$ is parameterized by the abstract state of the node (range + possible contents).
+We will fix the share $\pi$ to 1/2 so that the range in a node's lock can be out of date with respect to its $\mathsf{public\_half}$ (in particular, it may have grown due to rearrangements further up the tree).
+We then need to choose a variant $R$ describing the resources held by the lock for a given abstract state.
+The C implementation of a node is:
+\begin{lstlisting}[language = C,numbers = none]
+typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
+typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
+\end{lstlisting}
+The node data of each \lstinline{tree} is wrapped in a \lstinline{tree_t} struct, and is only accessible when the lock in the \lstinline{lock} field is held.
+The lock variant for a node at location $p$ then needs to describe the contents of the \lstinline{t} field for each possible abstract state:
+\begin{align*}
+&\mathsf{match}\ c\ \mathsf{with}\\
+&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
+&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)
+\end{align*}
+Because our implementation is doing hand-over-hand locking, in the internal case, we also need to obtain the lock variant assertions for the child nodes, so the full lock variant is: %this isn't enough; we need to talk about selflock and shares
+\begin{align*}
+\mathsf{node\_inv}((l, r), c) \triangleq \ &\mathsf{match}\ c\ \mathsf{with}\\
+&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
+&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * p_l \lockvar \mathsf{node\_inv} * p_r \lockvar \mathsf{node\_inv}
+\end{align*}
+Because we are using lock variants, we do not need to know the ranges or contents of the child nodes: instead, the variant construction guarantees that they are consistent with the global ghost state.
+
+%%edit from here
 We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification.
@@
 
+\section{Using the Specifications}
+client and client proofs
+
 \section{Related Work}

### 2021-12-29 concurrency/Verification of Concurrent Binary Search Tree.tex
finished first-draft explanation of BST lock invariants and ghost state

@@
 &|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
-&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * p_l \lockvar \mathsf{node\_inv} * p_r \lockvar \mathsf{node\_inv}
+&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * \\&\qquad p_l.\texttt{lock} \lockvar \mathsf{node\_inv} * p_r.\texttt{lock} \lockvar \mathsf{node\_inv}
 \end{align*}
-Because we are using lock variants, we do not need to know the ranges or contents of the child nodes: instead, the variant construction guarantees that they are consistent with the global ghost state.
+While we do not know the ranges or contents of the child nodes, the lock-variant construction guarantees that they are consistent with the $\mathsf{public\_half}$ assertions of the global ghost state.
 
-%%edit from here
-We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification.
-The private pre- and postcondition of the atomic specification is given by defining the predicate $\nodeboxrep(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and overall ghost state is represented by ghost name $g$.
-The $\nodeboxrep$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
-\begin{align*}
-    &\nodeboxrep_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa,g,ga) \\& *\ \nodeboxrep_{0.5}(pb,g,gb))\end{align*}
+Now we have everything we need to specify the tree operations:
+\begin{mathpar}
+\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle
 
-The public pre- and postcondition of the atomic specification is given by defining the predicate $\treerep(T,g)$ that describes the abstract state of the binary search tree $T$  in terms of collection of ghost states each describing the information about separate node in the tree..
-The $\treerep$ predicate ties together $\mathsf{public\_half}$ part of per-node ghost states and $\mathsf{ghost\_ref}$ of over-all ghost state as follows: \begin{align*}&\mathsf{ghost\_tree\_rep}(tg,range, g, g\_current) \triangleq\ \exists\ k, v, ga, gb\ \in tg,\ghost{\mathsf{public\_half}\ (range, Some(k,v,ga,gb))}{g\_current}\\& *\ \mathsf{ghost\_tree\_rep}(left(tg),(left(range),k),g,ga)\ *\ \mathsf{ghost\_tree\_rep}(right(tg),(k, right(range)),g,gb) \end{align*}
-\begin{align*}\treerep(T,g,g\_root) \triangleq \ \exists\ (tg:ghost\_tree),\ &\mathsf{pure\_tree}(tg)\ =\ T\ \land\ \mathsf{ghost\_tree\_rep}(tg,(-\infty, +\infty),g,g\_root)&\\ *\ \ghost{\mathsf{ghost\_ref}\ (\mathsf{find\_ghost\_set}(tg))}{g}\end{align*}
-Where $tg$ is called \emph{ghost tree}, which can be existentially quantified, extend the node in pure tree $T$ with actual pointer and ghost names for child nodes.
+\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{lookup}\ m\ k = v\rangle
 
-The atomic specification for any binary search tree's operation \texttt{op} should then be
-\begin{align*}\forall t.\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t,g)\rangle\ \texttt{op}(p)\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t', g)\rangle
-\end{align*}
-where $t$ and $t'$ are the abstract state of a tree before and after the function's execution. Now we are prepared to prove that each of our BST functions satisfies its specification.
+\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ m\ k)\rangle
+\end{mathpar}
+where $\treerep(m)$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock variant for the root node:
+$$\nodeboxrep(b) \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \lockvar \mathsf{node\_inv}$$
+
+Now we are prepared to prove that each of our BST functions satisfies its specification.
 

### 2022-01-19 concurrency/Verification of Concurrent Binary Search Tree.tex
explained hand-over-hand selflock

@@
 \label{safety}
-\begin{figure}[htp]
-\begin{subfigure}[t]{\textwidth}
-\begin{lstlisting}[language = C]
-typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
-typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
-typedef struct tree_t **treebox;
 
-void insert (treebox t, int x, void *value) {
-  struct tree_t *tgt = *t;
-  struct tree *p;
-  void *l = tgt>lock;
-  acquire(l);
-  for(;;) {
-    p = tgt->t;
-    if (p==NULL) {
-      tree_t *p1 = (struct tree_t *) surely_malloc (sizeof *tgt);
-      tree_t *p2 = (struct tree_t *) surely_malloc (sizeof *tgt);
-      p1 ->t = NULL;
-      p2 ->t = NULL;
-      lock_t *l1 = (lock_t *) surely_malloc(sizeof(lock_t));
-      makelock(l1);
-      p1->lock = l1;
-      release2(l1);
-      lock_t *l2 = (lock_t *) surely_malloc(sizeof(lock_t));
-      makelock(l2);
-      p2->lock = l2;
-      release2(l2);
-      p = (struct tree *) surely_malloc (sizeof *p);
-      tgt->t = p;
-      p->key=x; p->value=value; p->left=p1; p->right=p2;
-      release2(l);
-      return;
-    } else {
-      int y = p->key;
-      if (x<y){
-      	tgt = p->left;
-        void *l_old = l;
-        l = tgt->lock;
-        acquire(l);
-        release2(l_old);
-      } else if (y<x){
-        tgt = p->right;
-        void *l_old = l;
-        l = tgt->lock;
-        acquire(l);
-        release2(l_old);
-      }else {
-      	p->value=value;
-        release2(l);
-      	return;
-      }
-    }
-  }
-} 
-\end{lstlisting} 
-\end{subfigure}
-\caption{Insert Method}
-\label{insert}
-\end{figure}     
-\begin{figure}[htp]
-\begin{subfigure}[t]{\textwidth}
- \begin{lstlisting}
-void *lookup (treebox t, int x) {
-  struct tree *p; void *v;
-  struct tree_t *tgt;
-  tgt = *t;
-  void *l = tgt->lock;
-  acquire(l);
-  p = tgt->t;
-  while (p!=NULL) {
-    int y = p->key;
-    if (x<y){
-      tgt=p->left;
-      void *l_old = l;
-      l = tgt->lock;
-      acquire(l);
-      p=tgt->t;
-      release2(l_old);
-    }else if (y<x){
-      tgt=p->right;
-      void *l_old = l;
-      l = tgt->lock;
-      acquire(l);
-      p=tgt->t;
-      release2(l_old);
-    }else {
-      v = p->value;
-      release2(l);
-      return v;
-    }
-  }
-  release2(l);
-  return NULL;
-}
-\end{lstlisting}
-\end{subfigure}
-\caption{Lookup Method}
-\label{lookup}
-\end{figure}
 In this section, we will focus on safety proofs of concurrent binary search tree.
@@
 The code for $\texttt{insert}$ and $\texttt{lookup}$ methods of concurrent binary search tree that we are interested to verify is presented in Figure 1.
-
-\subsection{Lock invariants for hand-over-hand locking}
-A node of CBST have the pointer to its children, so lock invariant of a node should have the knowledge about its children.   
-Whenever we release the lock for the parent node, we have to recover all resources that the parent node accessed while acquiring the lock.
-By releasing the parent node's lock, we would lose the information about the lock\_inv of the node and we would not be able to release the node's lock later.
-To solve this problem, we use \emph{recursive} lock, one whose invariant includes a share of lock itself so that we can retain lock\_inv assertion for a node even after releasing its parent's lock.
-In VST, we can make such an invariant using $\mathsf{selflock}$ function along with the lemma $\mathsf{selflock\_eq}$: $\forall Q\ \mathit{sh}\ p, \mathsf{selflock}\ Q\ \mathit{sh}\ p = Q * \mathsf{lock\_inv}\ \mathit{sh}\ p\ (\mathsf{selflock}\ Q\ \mathit{sh}\ p)$.
-We can define $\mathsf{lock\_inv}$ predicate for \emph{recursive} lock $\mathit{lock}$ as follows: $\mathsf{lock\_inv}\ \mathit{sh1}\ p\ (\mathsf{selflock}\ P\ \mathit{sh2}\ \mathit{lock})$ where $\mathit{P}$ represents the knowledge about the children of current node, $\mathit{sh1}$ and $\mathit{sh2}$ are the two halves of the writable share.
 
@@
 A program instruction can use the %contents of global invariant if it can guarantee that no one will ever see an intermediate state %in which the invariant does not hold.
+
+\subsection{Lock invariants for hand-over-hand locking}
+\label{selflock}
+\begin{figure}[htp]
+\begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
+typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
+typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
+typedef struct tree_t **treebox;
+\end{lstlisting}
+\begin{subfigure}[t]{0.5\textwidth}
+\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
+void insert (treebox t, int x, void *value) {
+  struct tree_t *tgt = *t;
+  struct tree *p;
+  void *l = tgt->lock;
+  acquire(l);
+  for(;;) {
+    p = tgt->t;
+    if (p==NULL) {
+      tree_t *p1 = malloc(sizeof *tgt);
+      tree_t *p2 = malloc(sizeof *tgt);
+      p1->t = NULL;
+      p2->t = NULL;
+      lock_t *l1 = malloc(sizeof(lock_t));
+      makelock(l1);
+      p1->lock = l1;
+      release2(l1);
+      lock_t *l2 = malloc(sizeof(lock_t));
+      makelock(l2);
+      p2->lock = l2;
+      release(l2);
+      p = malloc(sizeof *p);
+      tgt->t = p;
+      p->key=x; p->value=value; p->left=p1; p->right=p2;
+      release(l);
+      return;
+    } else {
+      int y = p->key;
+      if (x<y){
+      	tgt = p->left;
+        void *l_old = l;
+        l = tgt->lock;
+        acquire(l);
+        release(l_old);
+      } else if (y<x){
+        tgt = p->right;
+        void *l_old = l;
+        l = tgt->lock;
+        acquire(l);
+        release(l_old);
+      } else {
+      	p->value=value;
+        release(l);
+      	return;
+      }
+    }
+  }
+} 
+\end{lstlisting} 
+\end{subfigure}\qquad
+\begin{subfigure}[t]{0.4\textwidth}
+ \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
+void *lookup (treebox t, int x) {
+  struct tree *p; void *v;
+  struct tree_t *tgt;
+  tgt = *t;
+  void *l = tgt->lock;
+  acquire(l);
+  p = tgt->t;
+  while (p != NULL) {
+    int y = p->key;
+    if (x<y){
+      tgt = p->left;
+      void *l_old = l;
+      l = tgt->lock;
+      acquire(l);
+      p = tgt->t;
+      release(l_old);
+    } else if (y<x){
+      tgt = p->right;
+      void *l_old = l;
+      l = tgt->lock;
+      acquire(l);
+      p = tgt->t;
+      release2(l_old);
+    } else {
+      v = p->value;
+      release(l);
+      return v;
+    }
+  }
+  release(l);
+  return NULL;
+}
+\end{lstlisting}
+\end{subfigure}
+\caption{BST with hand-over-hand locking}
+\label{bst-conc}
+\end{figure}
+
+Hand-over-hand locking is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it.
+We can see this pattern in the BST functions in figure~\ref{bst-conc}, for instance at lines 32-33 of the \lstinline{insert} function, where we acquire the new lock \lstinline{l} before releasing the current lock \lstinline{l_old}.
+Hand-over-hand locking patterns have been verified in many concurrent separation logics (see for instance \cite{gotsman}), usually by assigning each node a lock invariant that includes the lock assertion for its child nodes: for a linked list, we might write $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock.
+When the lock invariant assertion $\islock$ is \emph{replicable}, this approach is sufficient to express hand-over-hand locking: by acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.
+
+But if we attempt to account for \emph{ownership} of locks, as is required if we want to free the lock and return its resources, the picture becomes more complicated.
+The usual approach is to annotate the lock assertion with a share $\pi$, and say that the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
+We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$?
+Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
+\begin{align*}
+&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
+&\texttt{acquire}(\ell');\\
+&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
+&\texttt{release}(\ell);\\
+&\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
+\end{align*}
+In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$.
+This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share.
+Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant:
+$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$
+
+Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
+\begin{align*}
+&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
+&\texttt{acquire}(\ell');\\
+&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
+&\texttt{release}(\ell);\\
+&\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
+\end{align*}
+Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
+The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
 
@@
 \end{align*}
-Because our implementation is doing hand-over-hand locking, in the internal case, we also need to obtain the lock variant assertions for the child nodes, so the full lock variant is: %this isn't enough; we need to talk about selflock and shares
+We then combine this with the hand-over-hand invariant pattern of section~\ref{selflock}, giving us a final lock variant of:
 \begin{align*}
-\mathsf{node\_inv}((l, r), c) \triangleq \ &\mathsf{match}\ c\ \mathsf{with}\\
+\mathsf{node\_inv}((l, r), c) \triangleq \ &p.\texttt{lock} \lockvar_{\pi_1} \mathsf{node\_inv} * \mathsf{match}\ c\ \mathsf{with}\\
 &|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
-&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * \\&\qquad p_l.\texttt{lock} \lockvar \mathsf{node\_inv} * p_r.\texttt{lock} \lockvar \mathsf{node\_inv}
+&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * \\&\qquad p_l.\texttt{lock} \lockvar_{\pi_2} \mathsf{node\_inv} * p_r.\texttt{lock} \lockvar_{\pi_2} \mathsf{node\_inv}
 \end{align*}
-While we do not know the ranges or contents of the child nodes, the lock-variant construction guarantees that they are consistent with the $\mathsf{public\_half}$ assertions of the global ghost state.
+While we do not know the ranges or contents of the child nodes, the lock-variant construction guarantees that they are consistent with the $\mathsf{public\_half}$ assertions of the global ghost state, and the hand-over-hand pattern allows us to acquire $p_l.\texttt{lock}$ or $p_r.\texttt{lock}$ and then release $p.\texttt{lock}$ without losing any shares of the lock assertions.
 
@@
 $$\nodeboxrep(b) \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \lockvar \mathsf{node\_inv}$$
-
 Now we are prepared to prove that each of our BST functions satisfies its specification.

### 2022-01-19 concurrency/Verification of Concurrent Binary Search Tree.tex
clightgen'd up to CompCert 3.10
also patched for 64-bit

@@
 \subsubsection{Insert}
-An outline of the C code for the insert function is shown in figure \ref{insert-proof}.
-When a thread inserts a key $k$ into a tree, it first crawls the tree to find the right position for $k$ using the hand-over-hand locking mechanism, acquiring the lock for the next node before releasing the lock for the current node.
+An outline of the C code for the insert function is shown in figure \ref{bst-conc}.
+When a thread inserts a key $k$ into a tree, it first crawls the tree to find the right position for $k$ using hand-over-hand locking.
 If it encounters a node with key $k$, then it simply changes the value at that node.
 Otherwise, it will eventually reach a leaf node, where it creates two new empty leaf nodes, inserts the key and value into the current node, and links the newly created leaf nodes to the current node.
-The atomic specification for insert method is $$\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle$$
-The \texttt{insert} function should atomically add the key $k$ with value $v$ to the tree at $p$.
+The atomic specification for insert method is $$\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle$$
+The \lstinline{insert} function should atomically add the key-value pair $(k, v)$ to the map implemented by the BST at $p$.
 
-The key to the verification of the C code is the loop invariant for the top-level loop. We define the loop invariant for the insert function as follows: 
-\begin{align*} &\mathsf{insert\_inv}(b, x, g, g\_root) \triangleq\ \exists\ \mathit{lock},\mathit{g\_current},\mathit{np},\mathit{range},\mathit{info},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{g\_current}\ *\ R\ \mathit{np}\\& * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,g,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
-where $P_p$ and $Q_p$ are $\treerep(t,g)$ and $\treerep(t[k\mapsto v],g)$ respectively.
-The resources $R$, parameterized by a node pointer $\mathit{np}$, describes the concrete information about a node.
-Also, $x$ is the key to be inserted that must be in the bound $\mathsf{range}$.
-This loop invariant has some existential variables which characterize the state of each iteration of the loop.
+The main loop of the \lstinline{insert} function traverses the tree to find the node where $k$ should be inserted.
+At the beginning of each iteration, we are at some node $n$ and hold the lock for that node, giving us the lock variant $\mathsf{node\_inv}((l, r), c)$ where $(l, r)$ is the range of keys in the subtree rooted at $n$.
+The loop invariant holds that $k \in (l, r)$, that is, the key $k$ always belongs in the subtree of the current node.
+This guarantees us that if we reach an empty leaf, that leaf is exactly the right place to add a node with key $k$.
 
-Figure \ref{insertproof} shows the $\texttt{insert}$ function with separation logic annotations.
-The proof starts with $\nodeboxrep$ and $\mathsf{atomic\_shift}$ as the precondition.
-After acquiring the lock for the root node, the thread accesses the information inside the lock invariant of the root node.
-In order to prove $\mathsf{while}$ loop, we show that the precondition satisfies the $\mathsf{insert\_inv}$, then prove that the loop body preserves the loop invariant (cases inside $\mathsf{else if}$ clauses at line 15 and 17 in \ref{insert}).
-Once we locate the empty node for inserting the new key-value pair (inside the first $\mathsf{if}$ clause), we open the global invariant encoded in the $\mathsf{atomic\_shift}$, create the ghost names $\texttt{g1}$ and $\texttt{g2}$ for two new empty child nodes of current node, and use the $\mathsf{sync\_commit}$ rule---we have reached the linearization point and completed the insertion.
-The premise of $\mathsf{sync\_commit}$ requires that we have enough information to know that inserting the key at this point will change the abstract state from some unknown $t$ to $t[k\mapsto v]$, which we show in the second part of the proof.
-
-The other branch of the code-level proof is the case where the key to be inserted already exists in the tree (the last $\texttt{else}$ clause in the code).
-%The modification of the global state of the tree t is shown in \ref{extract_insert2}.
-In that case, the old value $v'$ associated with key is overridden by new value $v$ keeping a key $40$ and child pointers as as they are.
-This is also the \emph{linearization point} of the $\mathsf{insert}$ operation.
-We augment the same $\emph{extract}$ lemma with the clause that states the scenario in \ref{extract_insert2}, and prove the lemma in the same induction over the tree $t$.
-In this case, we change the value at the node but leave the rest of the tree intact, and this is the linearization point for the insertion.
-In the second part of the proof, we will show that this case also transforms $t$ to $t[k\mapsto v]$.
-
-In the other two cases, we compare the key
-of the current node with the new key \texttt{x} and move to the left or right child accordingly.
-Here we again need to access the atomic shift, to ensure that the range of the node we move to is included in the range of the current node.
-We can use $\mathsf{sync\_rollback}$ to retrieve this information without making any changes to the abstract state of the tree.
-%For the left branch case, we accomplish the proof by the following lemma:
-%\begin{verbatim}
-%Lemma in_tree_left_range:
-%  forall (B: Type) (b: tree -> B -> mpred) (Q : B → mpred) (x x0: Z) (g g_root : gname)
-%    (inv_names : invG) (v: val) (g_in ga gb: gname) (r a: node_info),
-%    check_key_exist' x r.1 = true -> r.2 = Some (Some (x0, v, ga, gb)) -> x < x0 ->
-%    atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
-%                 b Q * my_half g_in gsh1 r * in_tree g g_in * my_half ga gsh1 a
-%    |-- atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
-%    b Q * my_half g_in gsh1 r *
-%    (EX ba, !! (less_than_equal ba r.1.1 = true /\
-%                range_inclusion a.1 (ba, Finite_Integer x0) = true) &&
-%            (in_tree g g_in * my_half ga gsh1 (ba, Finite_Integer x0, a.2))).  
-%\end{verbatim}
-%The lemma for the right branch case is similar. Both use the
-%$\texttt{sync\_rollback}$ lemma described in section \ref{atomicity}.
+The function has two possible linearization points: after line 23, where we finish adding a node $(k, v)$ to the tree, and after line 41, where we change the value of an existing node with key $k$ to $v$.
+In both cases, we use the \textsf{sync\_commit} rule to atomically update the abstract state of the data structure from $\treerep m$ to $\treerep (\mathrm{insert}\ m\ k\ v)$, by showing that the update to the tree implementation corresponds to a local change in the abstract tree that implements the mathematical $\mathsf{insert}$.
+In the first case, the local change is the insertion of $(k, v)$ at an empty leaf whose range includes $k$, and the addition of two new empty leaves as its children; in the second case, we simply change the value of an internal node whose key is already $k$.
+Our higher-level proofs, then, need to show that these two changes are both valid operations on ghost trees that yield the desired insertion in the global state.
 
@@
 \end{figure}
-\begin{figure}[htp]
-\begin{subfigure}[t]{1\textwidth}
- $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
- \vspace*{-20pt}
-\begin{lstlisting}[language = C,numbers = none]
-insert(treebox t, int x, void *value)
-  tree_t *tgt = *t;
-  acquire(tgt->lock)
- \end{lstlisting}  
- $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}((-\infty,+\infty),info)}{g\_root}\ *\ R\ b\ *\ \mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\
- \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} insert\_inv \end{array}\right\}$$ 
-  \begin{lstlisting}[language = C,numbers = none]
-  for(;;) {
-       \end{lstlisting}   
-   $$\left\{\begin{array}{l} insert\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ \\*\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
-      \begin{lstlisting}[language = C,numbers = none]
-    p = tgt->t;
-    if (p==NULL)
-      tree_t *p1, *p2 = malloc();
-      p = malloc(); tgt->t = p;
-      p->key=x; p->value=value; p->left=p1; p->right=p2;
-           \end{lstlisting} 
-  $$\left\{\begin{array}{l} \ghost{\mathsf{own}\ (a1)}{g1}\ *\ \ghost{\mathsf{own}\ (a2)}{g2}\ *\ \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ \\*\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit}}$$
-$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
- \vspace*{-10pt}
-        \begin{lstlisting}[language = C,numbers = none]
-      release2(l);
-         \end{lstlisting}
-       $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
-        \vspace*{-10pt}
-         \begin{lstlisting}[language = C,numbers = none]
-      return;
-     else 
-        ....
-       else 
-      	p->value=value;
-      	\end{lstlisting} 
-$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *
-\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit}}$$
-$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
-         \vspace*{-10pt}
-      	\begin{lstlisting}[language = C, numbers = none]
-        release2(l);
-        \end{lstlisting}
-        $$\left\{\begin{array}{l}  \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
-         \vspace*{-10pt}
-        \begin{lstlisting}[language = C, numbers = none] 
-      	return;
-\end{lstlisting}
-\end{subfigure}
-\caption{The $\texttt{insert}$ function annotated with separation logic specification}
-\label{insertproof}
-\end{figure}
 
-\paragraph{Changing the Global State} describe ``extract lemmas'' here
+\paragraph{Changing the Global State} describe ``extract lemmas'' here---how much detail?
 

### 2022-01-23 concurrency/ITP 2022/Verification of Concurrent Binary Search Tree.tex
ITP 2022 formatting

--- 2022-01-19 concurrency/Verification of Concurrent Binary Search Tree.tex
@@
-% !TEX TS-program = pdflatex
-% !TEX encoding = UTF-8 Unicode
-%% For double-blind review submission
-\documentclass[acmsmall,screen]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission
-%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission
-%\documentclass[acmlarge]{acmart}\settopmatter{}
+\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+%This is a template for producing LIPIcs articles. 
+%See lipics-v2021-authors-guidelines.pdf for further information.
+%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
+%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
+%for section-numbered lemmas etc., use "numberwithinsect"
+%for enabling cleveref support, use "cleveref"
+%for enabling autoref support, use "autoref"
+%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
+%for enabling thm-restate support, use "thm-restate"
+%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
+%for producing a PDF according the PDF/A standard, add "pdfa"
 
-%% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format should change 'acmlarge' to
-%% 'sigplan,10pt'.
+%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
+%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
 
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
+%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+
+\bibliographystyle{plainurl}% the mandatory bibstyle
+
+\usepackage{txfonts}
 \usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                         %% http://ctan.org/pkg/subcaption
-
-\makeatletter\if@ACM@journal\makeatother
-%\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\else\makeatother
-%% Conference information (used by SIGPLAN proceedings format)
-%% Supplied to authors by publisher for camera-ready submission
-%\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\fi
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
 \usepackage{uri}
@@
 \usepackage[utf8]{inputenc}
-\usepackage{geometry}
-\geometry{a4paper}
-
-\usepackage{graphicx} % support the \includegraphics command and options
-\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
-%\usepackage{amssymb}
-\usepackage{url}
-\usepackage{listings}
 
 \title{Verifying Fine-Grained and Lock-Free Binary Search Trees}
-\author{Roshan Sharma}
-\author{Shengyi Wang}
-\author{Alexander Oey}
-\author{Anastasiia Evdokimova}
-\author{Lennart Beringer}
-\author{William Mansky}
+\author{Roshan Sharma}{University of Illinois at Chicago, USA}{}{}{}
+\author{Shengyi Wang}{Princeton University, USA}{}{}{}
+\author{Alexander Oey}{, USA}{}{}{}
+\author{Anastasiia Evdokimova}{University of Illinois at Chicago, USA}{}{}{}
+\author{Lennart Beringer}{Princeton University, USA}{}{}{}
+\author{William Mansky}{University of Illinois at Chicago, USA}{}{}{}
 
-\date{} % Activate to display a given date or no date (if empty),
-         % otherwise the current date is printed
+\authorrunning{R.
+Sharma, S.
+Wang, A.
+Oey, A.
+Evdokimova, L.
+Beringer, and W.
+Mansky} %TODO mandatory.
+First: Use abbreviated first/middle names.
+Second (only in severe cases): Use first author plus 'et al.'
+
+\Copyright{Roshan Sharma, Shengyi Wang, Alexander Oey, Anastasiia Evdokimova, Lennart Beringer, and William Mansky} %TODO mandatory, please use full first names.
+LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+
+\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+
+%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
+Please enter author specific funding statements as fifth argument of the \author macro.
+
+\acknowledgements{I want to thank \dots}%optional
+
+%\nolinenumbers %uncomment to disable line numbering
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{John Q. Open and Joan R. Access}
+\EventNoEds{2}
+\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+\EventShortTitle{CVIT 2016}
+\EventAcronym{CVIT}
+\EventYear{2016}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \begin{document}
-\acmConference{Some conference}
+
 \maketitle
 
-\catcode`\@=11
+\begin{abstract}
+abstract
+\end{abstract}
+
 \section{Introduction}
@@
 To prove correctness of this procedure, we need ghost state that tracks where each key ``should appear'' in the current tree.
-As observed by \citet{krishna2017flow}, this can be done by associating each node with a lower and upper bound on the keys appearing in the subtree rooted at that node.
+As observed by Krishna et al.~\cite{krishna2017flow}, this can be done by associating each node with a lower and upper bound on the keys appearing in the subtree rooted at that node.
 We refer to the product of these bounds as a \emph{range}.
@@
 \centering
-\includegraphics[width=70mm,scale=0.5]{FIG/range_prop.png}
+\includegraphics[width=70mm,scale=0.5]{../FIG/range_prop.png}
 \caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
@@
 \centering
-\includegraphics[width=150mm,scale=0.5]{FIG/extract_insert.png}
+\includegraphics[width=150mm,scale=0.5]{../FIG/extract_insert.png}
 \caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation }
@@
 \centering
-\includegraphics[width=150mm,scale=0.5]{FIG/extract_insert_2.png}
+\includegraphics[width=150mm,scale=0.5]{../FIG/extract_insert_2.png}
 \caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation, where key $k$ already exists in the tree $t$ }
@@
 %% Bibliography
-\bibliography{sources}
+\bibliography{../sources}
 \end{document}

### 2022-01-23 concurrency/ITP 2022/Verification of Concurrent Binary Search Tree.tex
moved hand-over-hand to background, since it's already present in Gotsman et al.'s 2007 paper

@@
 
+% Contributions:
+% verified C implementation of BST -- the first?
+% technique for hand-over-hand locking with share accounting -- slight modification of Gotsman's for logics without held predicates that remember the invariant; amounts to converting Locked_I into another share of the lock assertion
+% technique for connecting locks with abstract state/atomicity
+% VST-style decomposition of proofs into C-to-local-abstract and local-to-global-abstract
+% lock-free implementation?
+
 \section{Introduction}
@@
 \subsection{VST and Iris}
+
 We use the Verified Software Toolchain \cite{plfcc}  to prove the correctness of the concurrent binary search tree.
@@
 
-\subsection{Ghost State and Global Invariants}
+\ignore{\subsection{Ghost State and Global Invariants}
 We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program.
@@
 where invariants are provided in the form of assertion $\knowInv{i}{P}$, and states that $P$ is maintained as an invariant on the global state with name $i$. The operator%%  $\pvs[][]$
-is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e. $\knowInv{i}{..\knowInv{i}{P}..}$).
+is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e.
+$\knowInv{i}{..\knowInv{i}{P}..}$).}
+
+\subsection{Hand-over-Hand Locking}
+\label{selflock}
+
+Hand-over-hand locking (also called lock coupling) is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it.
+Hand-over-hand locking is a classic example of a pattern that requires reference to lock invariants inside other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}.
+The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock.
+By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.
+
+This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources.
+In this case we the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
+We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$?
+Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
+\begin{align*}
+&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
+&\texttt{acquire}(\ell');\\
+&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
+&\texttt{release}(\ell);\\
+&\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
+\end{align*}
+In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$.
+This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share.
+Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al.
+instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
+$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$
+
+Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
+\begin{align*}
+&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
+&\texttt{acquire}(\ell');\\
+&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
+&\texttt{release}(\ell);\\
+&\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
+\end{align*}
+Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
+The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
 
@@
 
-\subsection{Lock invariants for hand-over-hand locking}
+\subsection{BST code}
 \label{selflock}
+
 \begin{figure}[htp]
@@
 \end{figure}
-
-Hand-over-hand locking is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it.
-We can see this pattern in the BST functions in figure~\ref{bst-conc}, for instance at lines 32-33 of the \lstinline{insert} function, where we acquire the new lock \lstinline{l} before releasing the current lock \lstinline{l_old}.
-Hand-over-hand locking patterns have been verified in many concurrent separation logics (see for instance \cite{gotsman}), usually by assigning each node a lock invariant that includes the lock assertion for its child nodes: for a linked list, we might write $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock.
-When the lock invariant assertion $\islock$ is \emph{replicable}, this approach is sufficient to express hand-over-hand locking: by acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.
-
-But if we attempt to account for \emph{ownership} of locks, as is required if we want to free the lock and return its resources, the picture becomes more complicated.
-The usual approach is to annotate the lock assertion with a share $\pi$, and say that the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
-We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$?
-Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
-\begin{align*}
-&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
-&\texttt{acquire}(\ell');\\
-&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
-&\texttt{release}(\ell);\\
-&\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
-\end{align*}
-In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$.
-This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share.
-Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant:
-$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$
-
-Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
-\begin{align*}
-&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
-&\texttt{acquire}(\ell');\\
-&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
-&\texttt{release}(\ell);\\
-&\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
-\end{align*}
-Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
-The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
 

### 2022-01-25 concurrency/ITP 2022/Verification of Concurrent Binary Search Tree.tex
corrected some trivial typos

@@
 This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources.
-In this case we the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
+In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
 We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$?
@@
 ghost state) stored in the public precondition, while the commit is used to prove the public postcondition from the precondition, and obtain an assertion $Q$.
-To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of roolbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.
+To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of rollbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.
 

### 2022-01-25 concurrency/ITP 2022/intro.tex
moved hand-over-hand to background, since it's already present in Gotsman et al.'s 2007 paper
corrected some trivial typos

--- 2021-07-08 concurrency/intro.tex
@@
+% Contributions:
+% verified C implementation of BST -- the first?
+% technique for connecting locks with abstract state/atomicity
+% VST-style decomposition of proofs into C-to-local-abstract and local-to-global-abstract
+% lock-free implementation?
+
 The binary search tree (BST) is a common implementation of an ordered
@@
 triples}. Intuitively, an atomic Hoare triple $ \langle
-P \rangle\,\mathbb{C}\, \langle P \rangle$ means the program
+P \rangle\,\mathbb{C}\, \langle Q \rangle$ means the program
 $\mathbb{C}$ ``atomically updates'' from $P$ to $Q$. The program may
-actually take multiple steps, but evey step before the atomic update
+actually take multiple steps, but every step before the atomic update
 (linearization point) must preserve the assertion $P$. Meanwhile, the
@@
 style of Iris \cite{higherorderghoststate}. The ghost states are used
-to contruct both the global invariants and the local state in our
+to construct both the global invariants and the local state in our
 proofs. They will be further discussed in \S\ref{correctness}.
@@
 \end{description}
+
+%% Do we have contributions, and are they sufficient?
+Range ghost is adapted from flow interfaces.
+Sync_inv pattern isn't totally ours, though it might be instructive w.r.t.
+combining traditional and Iris-style CSLs, and it might be worth explaining in more detail.
+We're working on "actual" code, but missing a few pieces to make our proofs foundational.
+(but maybe it's worth highlighting the extra work we do to work on actual code) We could promote this as the first working example of Iris+VST.
+We don't yet have any demonstrations on larger or interesting data structures, and this alone was quite a bit of work.
+The delete logic is rather interesting.
+Nothing to say about flow interfaces yet, but maybe soon.
 

### 2022-01-28 concurrency/ITP 2022/Verification of Concurrent Binary Search Tree.tex
starting to rewrite ITP submission

@@
 
-\title{Verifying Fine-Grained and Lock-Free Binary Search Trees}
+\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking}
 \author{Roshan Sharma}{University of Illinois at Chicago, USA}{}{}{}
@@
 This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources.
-In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
+In this case we the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
 We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$?
@@
 \label{atomic}
+
 For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads.
@@
 
-\ignore{\section{Safety Proofs} %maybe cut this, or roll it into the next section
-\label{safety}
-
-In this section, we will focus on safety proofs of concurrent binary search tree.
-Here safety means thread-safe.
-An operation of any concurrent data-structure is thread-safe if it functions correctly during simultaneous execution by multiple threads.
-A thread can update the data structure if it holds a sufficiently large share of the data structure so that no other thread can possibly read it.
-If we want to have a data structure that can be modified by multiple threads, we must move shares between threads via locks.
-In VST, we can use $\mathsf{lock\_inv}$ assertion to assert that there exists a lock in memory with a given invariant: $\mathsf{lock\_inv}\ \mathsf{sh}\ p\ R$ means that the current thread owns share $\mathsf{sh}$ of a lock at location $p$ with invariant $R$.
-We use the term \emph{lock invariant} for $R$, which is a predicate representing the resources protected by a lock.
-The $\mathsf{lock\_inv}$ predicate is sufficient to prove the safety properties of the concurrent data-structure.
-
-%We have taken binary search tree with fine-grained locking.
-In hand-over-hand locking mechanism, we acquire the lock for a successor before releasing the lock for a predecessor.
-The code for $\texttt{insert}$ and $\texttt{lookup}$ methods of concurrent binary search tree that we are interested to verify is presented in Figure 1.
-
-\subsection{Specification and Verification}
-We begin the verification by defining a predicate $\nodeboxrep(p)$ that describes the concrete representation of the tree, where $p$ is the pointer to the root node.
-The $\texttt{nodebox\_rep}$ predicate ties together $\texttt{lock\_inv}$ assertion for each recursive lock as follows:
-\begin{align*}
- &\nodeboxrep_{\pi}(p) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \exists\ pa\ pb\ \ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa)\ *\ \nodeboxrep_{0.5}(pb)  \end{align*}
-
- The ownership of each node's lock is divided into two halves: one half is owned by \emph{lock\ invariant} and another half is owned by node itself.
-But, in case of root node, we need to split one half into multiple pieces, and share those pieces to all threads such that all of them initially have the knowledge of root node.
-
- The \emph{lock}, used in the implementation of BST operations, provides \texttt{acquire} and \texttt{release} functions, which allow threads to transfer ownership of resource invariants.
-Their pre- and postconditions are as follows:
-$$\{!!\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{acquire}(\ell)\ \{R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$
-$$\{!!(\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{exclusive}\ R) * R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{release}(\ell)\ \{\mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$
-
-The resource invariant $R$ need to be \emph{exclusive}, i.e, it can only hold once in any given state. This allows us to know that if the current thread holds the invariants, it also hold the lock. 
-}
+\section{Locks and Atomicity}
 
@@
 \end{mathpar}
-where $\treerep(m)$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock variant for the root node:
+where $\treerep\ m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock variant for the root node:
 $$\nodeboxrep(b) \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \lockvar \mathsf{node\_inv}$$
@@
 The function has two possible linearization points: after line 23, where we finish adding a node $(k, v)$ to the tree, and after line 41, where we change the value of an existing node with key $k$ to $v$.
-In both cases, we use the \textsf{sync\_commit} rule to atomically update the abstract state of the data structure from $\treerep m$ to $\treerep (\mathrm{insert}\ m\ k\ v)$, by showing that the update to the tree implementation corresponds to a local change in the abstract tree that implements the mathematical $\mathsf{insert}$.
+In both cases, we use the \textsf{sync\_commit} rule to atomically update the abstract state of the data structure from $\treerep\ m$ to $\treerep (\mathrm{insert}\ m\ k\ v)$, by showing that the update to the tree implementation corresponds to a local change in the abstract tree that implements the mathematical $\mathsf{insert}$.
 In the first case, the local change is the insertion of $(k, v)$ at an empty leaf whose range includes $k$, and the addition of two new empty leaves as its children; in the second case, we simply change the value of an internal node whose key is already $k$.
@@
 Most of the other concurrent separation logic (CSL)-based verification works are on toy languages, while VST lets us use the same logic on real C programs.
+VeriFast~\cite{verifast-conc} also operates on C code, albeit without a full formal semantics of C.
 
@@
 \cite{Xiong2017Abstract} have demonstrated the verification of ConcurrentSkipListMap from java.util.concurrent library using the recent advances in fine-grained concurrency reasoning.
-Their work is mainly based on the abstract atomicity from TaDa logic, and give two modular specifications for concurrent maps: one specification focus on the entire map structure which is suitable for verifying implementation, and another specification focus on the key-value pairs which appropriate for verifying clients.
+Their work is mainly based on the abstract atomicity from TaDA logic, and give two modular specifications for concurrent maps: one specification focus on the entire map structure which is suitable for verifying implementation, and another specification focus on the key-value pairs which appropriate for verifying clients.
 We use the same idea of atomicity (though implemented in Iris) in our work.

### 2022-01-28 concurrency/ITP 2022/intro.tex
starting to rewrite ITP submission

--- 2021-04-27 concurrency/intro.tex
@@
-The binary search tree (BST) is a common implementation of an ordered
-map, a widely used data structure. The concurrent version of the
-ordered map forms the bedrock of many parallel programs. We formally
-verified the functional correctness of two versions of fine-grained
-concurrent binary search trees (CBST) written in C. One adopts the
-hand-over-hand locking technique \cite{bayer1977}, the other is lock
-free by using atomic primitives such as compare and swap (CAS). Both
-CBST implementations support insert, lookup, and delete operations;
-they also share the ``same'' specifications to some extent in our
-verification. All the proof is machine-checked in Coq.
+% Contributions:
+% verified C implementation of BST with atomic specs -- the first? VeriFast did it, but not atomic specs
+% technique for connecting locks with abstract state/atomicity
+% VST-style decomposition of proofs into C-to-local-abstract and local-to-global-abstract
+% delete proofs -- interesting interplay between non-visible range increase and lock-protected range modification
+% lock-free implementation?
 
-Our specifications use the logical atomicity introduced in the TaDA
-logic \cite{tada}, in the form of \emph{atomic Hoare
-triples}. Intuitively, an atomic Hoare triple $ \langle
-P \rangle\,\mathbb{C}\, \langle P \rangle$ means the program
-$\mathbb{C}$ ``atomically updates'' from $P$ to $Q$. The program may
-actually take multiple steps, but evey step before the atomic update
-(linearization point) must preserve the assertion $P$. Meanwhile, the
-concurrent environment may also update the state before the
-linearization point, as long as the states satisfies $P$. The
-assertion $Q$ must become true at the linearization point, then the
-environment can do whatever it likes. There is no guarantee that $Q$
-is still preserved when $\mathbb{C}$ returns. For example, the
-specification of our \texttt{insert} operation may be explained as
-follows: during the execution of \texttt{insert}, there is
-always \emph{some} BST, and at some point the \texttt{insert} will
-take a BST $t$, insert a value with a certain key, and then eventually
-returns (meanwhile other threads may have further modified the
-inserted tree).
+% start with atomic specs for cglock implementation, where the local ghost state *is* the global ghost state!
 
-We employ the Verified Software Toolchain (VST) \cite{plfcc} to verify
-the correctness of CBST. Although the concurrent separation logic
-(CSL) has been formalized in VST rooted on the work of Hobor et
-al.\ \cite{oraclesematic}, we extend it with two descendants of CSL so
-as to accomplish the verification. One is the logical atomicity
-mentioned above; the other is the higher-order ghost state in the
-style of Iris \cite{higherorderghoststate}. The ghost states are used
-to contruct both the global invariants and the local state in our
-proofs. They will be further discussed in \S\ref{correctness}.
+Concurrent separation logic~\cite{csl} is a useful tool for proving correctness of concurrent programs.
+While early iterations focused on data structures implemented with locks~\cite{gotsman, oraclesematic} and specifications involving memory safety and data-race-freedom, more recent logics are able to prove linearizability-style functional correctness~\cite{tada} of lock-free implementations that use low-level atomic memory operations~\cite{iris}, even under weak memory models~\cite{gps, rsl}.
+In particular, \emph{logical atomicity}~\cite{tada} has become the gold standard for concurrent data structure specifications, capturing the idea that data structure operations should appear to take place instantaneously with no externally visible intermediate states.
+The effects of a logically atomic operation become visible at a \emph{linearization point}, a single instruction (usually an atomic memory operation or call to a logically atomic function) that publishes the new state of the data structure.
 
-We highlights a few innovative aspects about our verification of CBST:
-\begin{description}
-\item [Range ghost] We abstract the BST via a pair of values
-called \emph{range} which represents the lower-bound and upper-bound
-of keys on each node to reason about the BST in the current
-settings. We prove that range with the merging operation forms a
-partial commutative monoid (PCM) so that it can be encoded in ghost
-states.
+Intuitively, a critical section in a lock-based implementation serves the same role as an atomic instruction in a lock-free data structure, and it is not hard to identify linearization points in lock-based implementations---the effects of an update become visible when the lock protecting the updated data is released.
+In practice, however, the interaction between traditional lock-based separation logic and atomic specifications has been underexplored.
+In this paper, we demonstrate an approach to proving that lock-based implementations of data structures satisfy logically atomic specifications.
+The approach has two main components.
+First, we associate each lock with a piece of ghost state representing the abstract state of the part of the data structure it protects.
+In a \emph{coarse-grained} implementation, where a single lock protects the entire data structure, this is sufficient (in combination with proof rules for logical atomicity) to prove that operations complete atomically.
+In a \emph{fine-grained} implementation, however, a lock may only protect a small piece of the data structure (a single node in a tree, for instance), and the overall state of the data structure may be composed of a large number of per-lock abstract states.
+For these data structures, our second step is to prove lemmas describing how changes to local abstract state propagate to change (or maintain) the global abstract state.
+This gives us a two-layer verification process: we prove that the code of each function implements some local changes to the abstract state, and separately we prove that these local changes implement a global change.
 
-\item [\textsf{sync\_inv} pattern] It is a particular approach
-combining locks with general invariants to solve the dilemma caused by
-the fine-grained locking mechanism: we do not have a lock to control
-the access of the entire BST nor can we access the state of the BST
-via atomic operations.
-\end{description}
+Our proof technique is implemented in the Verified Software Toolchain~\cite{plfcc}, a system for proving separation logic properties of C programs, so we can prove correctness of C implementations of lock-based concurrent data structures.
+We take advantage of recent extensions to VST to include Iris-style ghost state and logically atomic specifications~\cite{vst-iris}.
+We demonstrate our technique by proving the correctness of a binary search tree with hand-over-hand locking.
+%something about client
 
-Our specific contributions are:
-\begin{itemize}
-\item To the best of our knowledge, this is the first mechanized
-verification of an concurrent search-based data structure written in a
-real programming language.
+%Our specific contributions are:
+%\begin{itemize}
+%\item To the best of our knowledge, this is the first mechanized
+%verification of a concurrent search-based data structure written in a
+%real programming language. %should probably clarify what we mean by "real"
 
-\item We illustrate how to incorporate the CSL in VST, the higher-order
-ghost state in the style of Iris, and the logical atomicity from the
-TaDA logic together to verify the CBST.
-\end{itemize}
+%\item We illustrate how to incorporate the CSL in VST, the higher-order
+%ghost state in the style of Iris, and the logical atomicity from the
+%TaDA logic together to verify the CBST.
+%\end{itemize}
 
-We introduce the background about the verification of concurrent C
-programs in \S\ref{background}, including the tool-chain VST and Iris,
-the concept of ghost states, global invariants, and atomic
-specifications. The thread-safety proofs of operations on CBST are
-first explained in \S\ref{safety}, where we show the \emph{recursive}
-lock pattern for hand-over-hand locking mechanism. The functional
-correctness proofs are presented in \S\ref{correctness}. We detail the
-use of the \textsf{sync\_inv} pattern, the combination of recursive
-lock invariants and ghost states together in the atomic
-specifications, and the proof skeleton of each operation on CBST. The
-related work is discussed in \S\ref{related}. We end with the
-conclusion of our work in \S\ref{conclusion}.
+%We introduce the background about the verification of concurrent C
+%programs in \S\ref{background}, including the tool-chain VST and Iris,
+%the concept of ghost states, global invariants, and atomic
+%specifications. The thread-safety proofs of operations on CBST are
+%first explained in \S\ref{safety}, where we show the \emph{recursive}
+%lock pattern for hand-over-hand locking mechanism. The functional
+%correctness proofs are presented in \S\ref{correctness}. We detail the
+%use of the \textsf{sync\_inv} pattern, the combination of recursive
+%lock invariants and ghost states together in the atomic
+%specifications, and the proof skeleton of each operation on CBST. The
+%related work is discussed in \S\ref{related}. We end with the
+%conclusion of our work in \S\ref{conclusion}.

### 2022-01-28 concurrency/bst_conc_nblocking.c

@@ -175,7 +175,7 @@ int main(void)
   }
   
 
- // printf("I am done to spwan all thread here \n");
+ // printf("I am done spawning all threads here \n");
   /*JOIN */
   for (int i = 0; i < N; i++)
   {

### 2022-01-28 concurrency/intro.tex
started rewriting proofs

@@
 triples}. Intuitively, an atomic Hoare triple $ \langle
-P \rangle\,\mathbb{C}\, \langle P \rangle$ means the program
+P \rangle\,\mathbb{C}\, \langle Q \rangle$ means the program
 $\mathbb{C}$ ``atomically updates'' from $P$ to $Q$. The program may
@@
 
+%% Do we have contributions, and are they sufficient?
+Range ghost is adapted from flow interfaces.
+Sync_inv pattern isn't totally ours, though it might be instructive w.r.t.
+combining traditional and Iris-style CSLs, and it might be worth explaining in more detail.
+We're working on "actual" code, but missing a few pieces to make our proofs foundational.
+(but maybe it's worth highlighting the extra work we do to work on actual code) We could promote this as the first working example of Iris+VST.
+We don't yet have any demonstrations on larger or interesting data structures, and this alone was quite a bit of work.
+The delete logic is rather interesting.
+Nothing to say about flow interfaces yet, but maybe soon.
+
 Our specific contributions are:

### 2022-02-15 concurrency/Verification of Concurrent Binary Search Tree.tex
Updated new diagrams

@@
 Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty \emph{leaf} nodes.
+
 \begin{figure}[htb]
@@
 \end{figure}
+
+\usetikzlibrary{positioning}
+\forestset{
+   lab/.style={
+        label={[anchor=west, scale=0.75] north east:#1},
+    },
+}
+\begin{forest}
+for tree={
+    grow=south,
+    circle, draw, minimum size=3ex, inner sep=1pt,
+    s sep=3mm
+        }
+    % my label/.style={
+        % label={[anchor=south west] north east:#1},
+    % },
+    [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+        [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+        [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+            [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                    [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                            [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                            [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$};}]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+                [,no edge, draw=none]
+            ]
+        ]
+    ]
+\end{forest}
+
 At the leaves, these ranges partition the space of all possible keys.
@@
 \end{figure}
+
+\begin{forest}
+for tree={
+    grow=south,
+    circle, draw, minimum size=3ex, inner sep=1pt,
+    s sep=3mm
+        }
+    [, phantom, for children={fit=band}, s sep+=20mm,
+         before drawing tree={%
+          tikz+={%
+            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
+            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
+            \node [anchor=center] at ($(a.east)!1/2!(b.west)$) {$\longrightarrow\quad$};
+          },
+        },
+        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                                [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$}; \node[left=0pt of .west, scale=0.35] {$\text{g\_current}$};}
+                                ]
+                            ]
+                            [,no edge, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
+        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                                [k, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$}; \node[left=0pt of .west, scale=0.35] {$\text{g\_current}$};}
+                                    [, tikz={\node[left=0pt of .west, scale=0.75]  {$g_1$};}]
+                                    [, tikz={\node[left=0pt of .west, scale=0.75]  {$g_2$};}]
+                                ]
+                            ]
+                            [,no edge, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
+    ]
+\end{forest}
+
 \begin{figure}[htb]
@@
 \end{figure}
+
+\begin{forest}
+for tree={
+    grow=south,
+    circle, draw, minimum size=3ex, inner sep=1pt,
+    s sep=3mm
+        }
+    [, phantom, for children={fit=band}, s sep+=20mm,
+         before drawing tree={%
+          tikz+={%
+            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
+            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
+            \node [anchor=center] at ($(a.east)!1/2!(b.west)$) {$\longrightarrow\quad$};
+          },
+        },
+        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                                [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$}; \node[left=0pt of .west, scale=0.35] {$\text{g\_current}$};}
+                                ]
+                            ]
+                            [$g_2$, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
+        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                                [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$}; \node[left=0pt of .west, scale=0.35] {$\text{g\_current}$};}
+                                ]
+                            ]
+                            [$g_2$, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
+    ]
+\end{forest}
 
@@
 
-\usetikzlibrary{positioning}
-
-\forestset{
-   lab/.style={
-        label={[anchor=west, scale=0.75] north east:#1},
-    },
-}
 \begin{forest}

### 2022-04-06 concurrency/intro.tex
minor corrections/comments

@@
 $\mathbb{C}$ ``atomically updates'' from $P$ to $Q$. The program may
-actually take multiple steps, but evey step before the atomic update
+actually take multiple steps, but every step before the atomic update
 (linearization point) must preserve the assertion $P$. Meanwhile, the
@@
 returns (meanwhile other threads may have further modified the
-inserted tree).
+inserted tree)\footnote{Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t.
+the situation where one of the other threads' action following the linearization point removes the element just inserted.}.
 
@@
 
-We highlights a few innovative aspects about our verification of CBST:
+We highlight a few innovative aspects of our verification of CBST:
 \begin{description}

### 2022-04-07 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
folder for ASL submission

--- 2022-01-28 concurrency/ITP 2022/Verification of Concurrent Binary Search Tree.tex
@@
-\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
-%This is a template for producing LIPIcs articles. 
-%See lipics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
+\documentclass[runningheads]{llncs}
 
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
-\bibliographystyle{plainurl}% the mandatory bibstyle
+\bibliographystyle{splncs04}% the mandatory bibstyle
 
 \usepackage{txfonts}
+\usepackage[T1]{fontenc}
+% T1 fonts will be used to generate the final print and online PDFs,
+% so please use T1 fonts in your manuscript whenever possible.
+% Other font encondings may result in incorrect characters.
+%
+\usepackage{graphicx}
+% Used for displaying a sample figure. If possible, figure files should
+% be included in EPS format.
+%
 \usepackage{subcaption} %% For complex figures with subfigures/subcaptions
@@
 \usepackage{uri}
-\usepackage{mathtools}
-\usepackage{amsthm}
 \usepackage{mathpartir}
 \usepackage{semantic}
-\usepackage{graphicx}
 \usepackage{cases}
 \usepackage{hyperref}
+% If you use the hyperref package, please uncomment the following two lines
+% to display URLs in blue roman font according to Springer's eBook style:
+\usepackage{color}
+\renewcommand\UrlFont{\color{blue}\rmfamily}
 \usepackage{stmaryrd}
 \usepackage{listings}
-\usepackage{iris}
+\usepackage{amssymb}
+\usepackage{iris-llncs}
 %\usepackage{lstlangcoq}
@@
 \newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
-\newcommand{\treerep}{\ensuremath{\mathsf{bst}}}
+\newcommand{\treerep}{\ensuremath{\mathsf{bst\_abs}}}
 \newcommand{\nodeboxrep}{\ensuremath{\mathsf{bst\_ref}}}
@@
 
-\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking}
-\author{Roshan Sharma}{University of Illinois at Chicago, USA}{}{}{}
-\author{Shengyi Wang}{Princeton University, USA}{}{}{}
-\author{Alexander Oey}{, USA}{}{}{}
-\author{Anastasiia Evdokimova}{University of Illinois at Chicago, USA}{}{}{}
-\author{Lennart Beringer}{Princeton University, USA}{}{}{}
-\author{William Mansky}{University of Illinois at Chicago, USA}{}{}{}
+\begin{document}
 
-\authorrunning{R.
-Sharma, S.
-Wang, A.
-Oey, A.
-Evdokimova, L.
-Beringer, and W.
-Mansky} %TODO mandatory.
-First: Use abbreviated first/middle names.
-Second (only in severe cases): Use first author plus 'et al.'
+\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking\thanks{Supported by the NSF (details).}}
+%
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{Roshan Sharma\inst{1} \and
+Shengyi Wang\inst{2}\orcidID{?} \and
+Alexander Oey\inst{3} \and
+Anastasiia Evdokimova\inst{4} \and
+Lennart Beringer\inst{2}\orcidID{?} \and
+William Mansky\inst{4}\orcidID{0000-0002-5351-895X}}
 
-\Copyright{Roshan Sharma, Shengyi Wang, Alexander Oey, Anastasiia Evdokimova, Lennart Beringer, and William Mansky} %TODO mandatory, please use full first names.
-LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\authorrunning{R. Sharma et al.}
 
-\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+\institute{Amazon, USA \and Princeton University, USA \and Cisco, USA \and University of Illinois Chicago, USA} %do we need a note about prior affiliations?
 
-\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-\acknowledgements{I want to thank \dots}%optional
+%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} https://dl.acm.org/ccs/ccs_flat.cfm
 
 %\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
 
@@
 \begin{abstract}
-abstract
+Modern concurrent separation logic techniques have been very effective at proving that lock-free data structures implement atomic specifications: they perform the same operations as sequential data structures, and do not expose any intermediate states to other threads.
+In this paper, we describe a method of proving the same kinds of specifications for lock-based data structures, with both coarse-grained and fine-grained locking patterns.
+We begin from the traditional separation logic specifications of locks and lock invariants, and combine them with a ghost state pattern that connects the critical section of a lock with the linearization point of a data structure operation.
+For fine-grained data structures, this allows us to decompose the proofs into a proof that a lock-based program performs some local changes to ghost state, and a proof that these local changes correctly implement an update to the global state of the data structure.
+Our proof technique is implemented in the Verified Software Toolchain and applied to C implementations of lock-based concurrent data structures.
+
+\keywords{concurrent separation logic, fine-grained locking, logical atomicity, Verified Software Toolchain}
 \end{abstract}
@@
 Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
-\begin{align*}
-&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
-&\texttt{acquire}(\ell');\\
-&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
-&\texttt{release}(\ell);\\
-&\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
-\end{align*}
+$$\begin{array}{l}
+\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
+\texttt{acquire}(\ell');\\
+\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
+\texttt{release}(\ell);\\
+\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
+\end{array}$$
 In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$.
@@
 Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
-\begin{align*}
-&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
-&\texttt{acquire}(\ell');\\
-&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
-&\texttt{release}(\ell);\\
-&\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
-\end{align*}
+$$\begin{array}{l}
+\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
+\texttt{acquire}(\ell');\\
+\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
+\texttt{release}(\ell);\\
+\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
+\end{array}$$
 Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
@@
 ghost, accessing invariants around atomic triple
+%Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t.
+the situation where one of the other threads' action following the linearization point removes the element just inserted.
 
@@
 \section{Locks and Atomicity}
+Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
+\begin{mathpar}
+\{\mathsf{BST}\ p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\ p\ (\mathsf{insert}\ t\ k\ v)\}
+
+\{\mathsf{BST}\ p\ t\}\ \texttt{lookup}(p, k)\ \{v.\ \mathsf{BST}\ p\ t \land \mathsf{lookup}\ t\ k = v\}
+
+\{\mathsf{BST}\ p\ t\}\ \texttt{delete}(p, k)\ \{\mathsf{BST}\ p\ (\mathsf{delete}\ t\ k)\}
+\end{mathpar}
+where $\mathsf{BST}\ p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$.
+For instance, $t$ may be a recursively defined tree type in Coq, and $\mathsf{insert}$ may be a pure Coq function that adds a key-value pair to such a tree; $\mathsf{BST}$ could then be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, and $p$ could be the location of the root node.
+
+A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs:
+\begin{mathpar}
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle
+
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle
+
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ t\ k)\rangle
+\end{mathpar}
+These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call.
+Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: in separation logic terms, the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.
+The easiest way to fix this is to add a lock that each operation acquires at the start and releases at the end, with $\mathsf{BST}\ p\ t$ as its \emph{lock invariant}: the lock then holds full ownership of the data structure, and provides it to whichever thread is currently executing an operation on the tree at $p$.
+
+In this \emph{coarse-grained} locking implementation, any thread with partial ownership of the lock can acquire the lock, perform all the operations of the sequential insertion/lookup/deletion, and then release the lock.
+Naively, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock---but this is not enough to tell us that we have changed the lock's state by the end of the operation, since we merely restore the invariant that there is \emph{some} tree $t$ at $p$.
+This is where ghost state and atomic triples come in: we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure.
+For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$.
 
@@
 We formalize this by first defining the collection of $\mathsf{public\_half}$ assertions---i.e., the authoritative copies of the per-node ghost states---needed to form an abstract tree:
-\begin{align*}
+$$\begin{array}{r l}
 \mathsf{public}&\mathsf{\_halves}(t, g, (l, r)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
@@
 & |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{Some}\ (k, v, g_l, g_r)) * \mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
-\end{align*}
+\end{array}$$
 We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{public\_half}$ assertions of each node into a single assertion.
@@
 Then our final definition of $\treerep$ is:
-$$\treerep(m, g) \triangleq \exists t.\ t \text{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty)) * \mathsf{ghost\_nodes}(\mathsf{node\_names}(t))$$
+$$\treerep(m, g) \triangleq \exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty)) * \mathsf{ghost\_nodes}(\mathsf{node\_names}(t))$$
 The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in the ghost tree $t$.
@@
 The lock variant for a node at location $p$ then needs to describe the contents of the \lstinline{t} field for each possible abstract state:
-\begin{align*}
-&\mathsf{match}\ c\ \mathsf{with}\\
-&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
-&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)
-\end{align*}
+$$\begin{array}{l}
+\mathsf{match}\ c\ \mathsf{with}\\
+|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
+|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)
+\end{array}$$
 We then combine this with the hand-over-hand invariant pattern of section~\ref{selflock}, giving us a final lock variant of:
-\begin{align*}
+$$\begin{array}{r l}
 \mathsf{node\_inv}((l, r), c) \triangleq \ &p.\texttt{lock} \lockvar_{\pi_1} \mathsf{node\_inv} * \mathsf{match}\ c\ \mathsf{with}\\
@@
 &|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * \\&\qquad p_l.\texttt{lock} \lockvar_{\pi_2} \mathsf{node\_inv} * p_r.\texttt{lock} \lockvar_{\pi_2} \mathsf{node\_inv}
-\end{align*}
+\end{array}$$
 While we do not know the ranges or contents of the child nodes, the lock-variant construction guarantees that they are consistent with the $\mathsf{public\_half}$ assertions of the global ghost state, and the hand-over-hand pattern allows us to acquire $p_l.\texttt{lock}$ or $p_r.\texttt{lock}$ and then release $p.\texttt{lock}$ without losing any shares of the lock assertions.
@@
 
+\begin{forest}
+for tree={
+    grow=south,
+    circle, draw, minimum size=3ex, inner sep=1pt,
+    s sep=3mm
+        }
+    [, phantom, for children={fit=band}, s sep+=20mm,
+         before drawing tree={%
+          tikz+={%
+            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
+            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
+            \node [anchor=center] at ($(a.east)!1/2!(b.west)$) {$\longrightarrow\quad$};
+          },
+        },
+        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                                [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$}; \node[left=0pt of .west, scale=0.35] {$\mathrm{g\_current}$};}
+                                ]
+                            ]
+                            [,no edge, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
+        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                                [k, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$}; \node[left=0pt of .west, scale=0.35] {$\mathrm{g\_current}$};}
+                                    [, tikz={\node[left=0pt of .west, scale=0.75]  {$g_1$};}]
+                                    [, tikz={\node[left=0pt of .west, scale=0.75]  {$g_2$};}]
+                                ]
+                            ]
+                            [,no edge, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
+    ]
+\end{forest}
+
 \begin{figure}[htb]
@@
 \end{figure}
+
+\begin{forest}
+for tree={
+    grow=south,
+    circle, draw, minimum size=3ex, inner sep=1pt,
+    s sep=3mm
+        }
+    [, phantom, for children={fit=band}, s sep+=20mm,
+         before drawing tree={%
+          tikz+={%
+            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
+            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
+            \node [anchor=center] at ($(a.east)!1/2!(b.west)$) {$\longrightarrow\quad$};
+          },
+        },
+        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                                [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$}; \node[left=0pt of .west, scale=0.35] {$\mathrm{g\_current}$};}
+                                ]
+                            ]
+                            [$g_2$, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
+        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                                [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$}; \node[left=0pt of .west, scale=0.35] {$\mathrm{g\_current}$};}
+                                ]
+                            ]
+                            [$g_2$, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
+    ]
+\end{forest}
 
@@
 As before, the key to the verification is the invariant for the main loop:
-\begin{align*} &\mathsf{lookup\_inv}(b, x, \mathit{g\_root}, \mathit{range},\mathit{info}) \triangleq\ \exists\ \mathit{lock},\ \mathit{g\_current},\ \mathit{np},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{\mathit{g\_current}}\\& *\ R\ np\ * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
+$$\begin{array}{l} \mathsf{lookup\_inv}(b, x, \mathit{g\_root}, \mathit{range},\mathit{info}) \triangleq\ \exists\ \mathit{lock},\ \mathit{g\_current},\ \mathit{np},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{\mathit{g\_current}}\\ *\ R\ np\ * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{array}$$
 where $P_p$ and $Q_p$ are $\treerep(g\_root, t)$ and $(v =
@@
 
+%TaDA-live: proves atomicity specs of some lock-based data structures incidentally, no mechanization
+%search structure templates: lock coupling data structures with atomic specs specifically, really the toy language is the only disadvantage, unclear whether we'd rather replicate them in whole or in part
+
 \section{Conclusion}
 \label{conclusion}
+
+\subsubsection{Acknowledgements} I want to thank Zhen Zheng and Ralf Jung, ...
 

### 2022-04-07 concurrency/ASL 2022/intro.tex
folder for ASL submission

--- 2022-01-28 concurrency/ITP 2022/intro.tex
@@
 % technique for connecting locks with abstract state/atomicity
-% VST-style decomposition of proofs into C-to-local-abstract and local-to-global-abstract
+	% note that the flow interfaces team sort of does atomicity for locks, but they assume an atomic spec for locks that's also application-specific?
+no, they implement them in the TaDA style and prove an atomic spec.
+% VST-style decomposition of proofs into C-to-local-abstract and local-to-global-abstract -- this corresponds to flow interfaces/CSS templates, basically
 % delete proofs -- interesting interplay between non-visible range increase and lock-protected range modification
@@
 Concurrent separation logic~\cite{csl} is a useful tool for proving correctness of concurrent programs.
-While early iterations focused on data structures implemented with locks~\cite{gotsman, oraclesematic} and specifications involving memory safety and data-race-freedom, more recent logics are able to prove linearizability-style functional correctness~\cite{tada} of lock-free implementations that use low-level atomic memory operations~\cite{iris}, even under weak memory models~\cite{gps, rsl}.
+While early iterations focused on data structures implemented with locks~\cite{gotsman,oraclesematic} and specifications involving memory safety and data-race-freedom, more recent logics are able to prove linearizability-style functional correctness~\cite{tada} of lock-free implementations that use low-level atomic memory operations~\cite{iris}, even under weak memory models~\cite{gps}%update cite
+.
 In particular, \emph{logical atomicity}~\cite{tada} has become the gold standard for concurrent data structure specifications, capturing the idea that data structure operations should appear to take place instantaneously with no externally visible intermediate states.

### 2022-04-08 concurrency/ASL 2022/intro.tex
Rewrote intro about different lock specs

@@
 
-Concurrent separation logic~\cite{csl} is a useful tool for proving correctness of concurrent programs.
+Concurrent separation logic (CSL)~\cite{csl} is a useful tool for proving correctness of concurrent programs.
 While early iterations focused on data structures implemented with locks~\cite{gotsman,oraclesematic} and specifications involving memory safety and data-race-freedom, more recent logics are able to prove linearizability-style functional correctness~\cite{tada} of lock-free implementations that use low-level atomic memory operations~\cite{iris}, even under weak memory models~\cite{gps}%update cite
@@
 Intuitively, a critical section in a lock-based implementation serves the same role as an atomic instruction in a lock-free data structure, and it is not hard to identify linearization points in lock-based implementations---the effects of an update become visible when the lock protecting the updated data is released.
-In practice, however, the interaction between traditional lock-based separation logic and atomic specifications has been underexplored.
+The translation of this intuition into proof, however, depends on the specifications used for lock operations.
+The CSL literature contains two distinct specifications for lock acquire and release:
+\begin{enumerate}
+\item In Gotsman/Hobor-style specs, each lock is associated with a lock invariant $R$, which is gained by \texttt{acquire} operations and restored by \texttt{release} operations.
+\item In TaDA-style specs, \texttt{acquire} atomically moves a lock from the unlocked state to the locked state, and \texttt{release} does the reverse.
+\end{enumerate}
+Style 1 can be derived from style 2, and is also a natural fit for atomicity proofs of larger data structures, so it has been the style of choice for most recent verifications~\cite{tada-live,templates}.
+However, unpublished work by Zhang~\cite{atomic-syncer} shows that style 1 can also be used to prove atomic specifications for data structures.
+
+In this paper, we adapt the latter approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t.
+the CompCert compiler~\cite{compcert}, and has recently been extended with support for general ghost state and atomicity proofs.
+We demonstrate that 1) this approach can be used to prove atomic specs for interesting data structures with fine-grained locking, and 2) these proofs are much more complicated than those built with style 2 lock specs.
+The key difference is that in style 1, we must associate each locked component with some resources and then relate those resources to the state of the data structure as a whole, while in style 2 each lock access can interact directly with the full state of the data structure.
+However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise.
+
+%In practice, however, the interaction between traditional lock-based separation logic and atomic specifications has been underexplored.
 In this paper, we demonstrate an approach to proving that lock-based implementations of data structures satisfy logically atomic specifications.
@@
 
-Our proof technique is implemented in the Verified Software Toolchain~\cite{plfcc}, a system for proving separation logic properties of C programs, so we can prove correctness of C implementations of lock-based concurrent data structures.
-We take advantage of recent extensions to VST to include Iris-style ghost state and logically atomic specifications~\cite{vst-iris}.
-We demonstrate our technique by proving the correctness of a binary search tree with hand-over-hand locking.
-%something about client
+%We demonstrate our technique by proving the correctness of a binary search tree with hand-over-hand locking. %something about client
 

### 2022-04-08 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
restructured paper, updated todo

@@
 \begin{abstract}
-Modern concurrent separation logic techniques have been very effective at proving that lock-free data structures implement atomic specifications: they perform the same operations as sequential data structures, and do not expose any intermediate states to other threads.
-In this paper, we describe a method of proving the same kinds of specifications for lock-based data structures, with both coarse-grained and fine-grained locking patterns.
-We begin from the traditional separation logic specifications of locks and lock invariants, and combine them with a ghost state pattern that connects the critical section of a lock with the linearization point of a data structure operation.
-For fine-grained data structures, this allows us to decompose the proofs into a proof that a lock-based program performs some local changes to ghost state, and a proof that these local changes correctly implement an update to the global state of the data structure.
-Our proof technique is implemented in the Verified Software Toolchain and applied to C implementations of lock-based concurrent data structures.
+Logical atomicity has been widely accepted as a specification format for data structures in concurrent separation logic.
+While both lock-free and lock-based data structures have been verified against logically atomic specifications, most of the latter start with atomic specifications for the locks as well.
+In this paper, we compare this approach with one based on older lock-invariant-based specifications for locks.
+We show that we can still prove logically atomic specifications for data structures with fine-grained locking using these older specs, but the proofs are significantly more complicated than those that use atomic lock specifications.
+Our proof technique is implemented in the Verified Software Toolchain, which relies on older lock specifications for its soundness proof, and applied to C implementations of lock-based concurrent data structures.
 
@@
 \label{background}
+
+\subsection{Two Styles of Lock Specification}
+Soon after the introduction of concurrent separation logic~\cite{csl}, both Gotsman et al.~\cite{gotsman} and Hobor et al.~\cite{oraclesematic} extended it with support for C-style storable locks, where a lock is a special memory location designated as coordinating access to some other resource.
+Their lock specifications were essentially equivalent, and of the form:
+$$\{ \ell \islock R \}\ \texttt{acquire}(\ell)\ \{ R * \ell \islock R \} \qquad \{ R * \ell \islock R \}\ \texttt{release}(\ell)\ \{ \ell \islock R \}$$
+Each lock $\ell$ is associated with a ``lock invariant'' $R$, an arbitrary separation logic predicate representing the protected resource, which is gained by a thread that acquires the lock and must be restored when the lock is released.
+At the time, there was no formal treatment of atomic operations in CSL, so actual lock implementations were beyond the scope of verification: these specs were taken as axioms, or proved through direct appeal to data-race-free guarantees of specific memory models.
+
+Later, the TaDA logic~\cite{tada} presented an alternative set of lock specifications, based on the newly introduced notion of logical atomicity:
+$$\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle$$
+An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return.
+A release simply moves atomically from the locked state to the unlocked state.
+These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(U(\ell) * R) \vee L(\ell)$, they can be used to derive the lock-invariant-based specs.
+It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g.
+compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it.
+
 \subsection{VST and Iris}
 
-We use the Verified Software Toolchain \cite{plfcc}  to prove the correctness of the concurrent binary search tree.
-VST is a Coq-based proof system that helps to prove separation logic properties of C programs with the assurance that the properties will be preserved in compiled code.
-First, we write programs in C programming and using CompCert’s \emph{clightgen}, the ASTs for C programs can be generated in Coq.
-We then state and prove specifications for them using Verifiable C.
-Verifiable C is a higher-order separation logic for reasoning about the functional correctness of C programs and carries a soundness proof that links high-level specifications to assembly code generated by the CompCert Verified compiler.
-Verifiable C supports reasoning about two separate kinds of concurrency:  Pthreads-style concurrency with locks (used by implementing blocking data structures that can be fine-grained or coarse-grained ), and concurrency using C11 atomic operations (i.e.
-lock-free data structures).
+The Verified Software Toolchain (VST)~\cite{plfcc} is a Coq-based system for proving separation logic specifications of C programs.
+It is proved sound against not only the C semantics of the CompCert compiler, but also the compiler correctness theorem~\cite{cpm}, so that verified programs are guaranteed to behave correctly when compiled and run.
+VST directly implements the concurrent separation logic of Hobor et al.~\cite{oracle-sematics}, and the invariant-style lock specifications are incorporated into its soundness proof at a fundamental level.
+
+Iris~\cite{iris} is a Coq-based separation logic framework that synthesizes ideas from a range of CSLs with the idea that ``ghost state is all you need''.
+Iris begins with a small core logic including arbitrary higher-order ghost state, and derives features including invariants and logical atomicity from particular instances of ghost state.
+Iris is designed as a language- and logic-independent framework, and recent versions of VST take advantage of this, incorporating Iris-style ghost state in the foundational model and directly using Iris tactics to reason about invariants and atomicity.
+All proofs described in this paper have been formalized in this VST+Iris setting.
+%What else do we need to say?
 
@@
 
-\subsection{Hand-over-Hand Locking}
-\label{selflock}
-
-Hand-over-hand locking (also called lock coupling) is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it.
-Hand-over-hand locking is a classic example of a pattern that requires reference to lock invariants inside other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}.
-The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock.
-By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.
-
-This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources.
-In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
-We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$?
-Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
-$$\begin{array}{l}
-\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
-\texttt{acquire}(\ell');\\
-\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
-\texttt{release}(\ell);\\
-\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
-\end{array}$$
-In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$.
-This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share.
-Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al.
-instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
-$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$
-
-Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
-$$\begin{array}{l}
-\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
-\texttt{acquire}(\ell');\\
-\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
-\texttt{release}(\ell);\\
-\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
-\end{array}$$
-Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
-The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
-
-\subsection{Atomic Specifications}
+\ignore{\subsection{Atomic Specifications}
 \label{atomic}
@@
 ghost state) stored in the public precondition, while the commit is used to prove the public postcondition from the precondition, and obtain an assertion $Q$.
-To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of rollbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.
+To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of rollbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.}
 
@@
 Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: in separation logic terms, the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.
-The easiest way to fix this is to add a lock that each operation acquires at the start and releases at the end, with $\mathsf{BST}\ p\ t$ as its \emph{lock invariant}: the lock then holds full ownership of the data structure, and provides it to whichever thread is currently executing an operation on the tree at $p$.
-
+
+\subsection{Atomicity with lock invariants}
+The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end.
+With the invariant-style lock specs, we can make $\mathsf{BST}\ p\ t$ the invariant of this lock: the lock then holds full ownership of the data structure, and provides it to whichever thread is currently executing an operation on the tree at $p$.
 In this \emph{coarse-grained} locking implementation, any thread with partial ownership of the lock can acquire the lock, perform all the operations of the sequential insertion/lookup/deletion, and then release the lock.
@@
 For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$.
-
-\section{BST with Hand-over-Hand Locking}
+Then we can prove the atomic triples above, where at the linearization point we access $\mathsf{bst\_abs}$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then satisfy the atomic postcondition with one half of the ghost state and return the other half to the lock invariant.
+The atomic postcondition reflects the fact that the abstract state has been accessed or changed, while the lock invariant maintains the connection between the abstract state and the actual data structure in memory.
+This approach to making changes to a locked data structure visible via ghost state is due to Zhang and Jung's proofs of an atomic syncer in Iris~\cite{atomic-syncer}.%, a higher-order function that wraps any arbitrary functionality in a lock.
+They proved that if a target function $f$ has a spec of the form $\{R * P\}\ f\ \{R * Q\}$, where $R$ is e.g.
+the ownership assertion for a data structure, then the locked version $f'$ of $f$ has spec $\langle g.
+P * \mathsf{ghost}_{.5}\ g\rangle\ f'\ \langle Q * \mathsf{ghost}_{.5}\ g'\rangle$.
+
+\label{atomicity}
+
+In a \emph{fine-grained} implementation, there are multiple locks on different pieces of the data structure, so the ghost state of each lock invariant cannot be the abstract state of the entire data structure.
+Instead, the ghost state represents the abstract state of the locked section alone, and the abstract state of the data structure is derived from the composition of the states of the locked components.
+Each locked component $c$ has a piece of ghost state $\mathsf{ghost}\_c$ that is split between the lock invariant and the top-level abstract state.
+At the end of a critical section for $c$, we atomically access the top-level abstract state, combine the two halves of $\mathsf{ghost}\_c$, update $\mathsf{ghost}\_c$ to reflect any local changes, and then show either that the update to $c$ does not change the top-level state of the data structure, or that the update to $c$ changes the top-level state to one that satisfies the atomic postcondition for the function.
+For instance, when deleting a node from a binary search tree, we rotate nodes to reconstruct a valid tree; at the end of the critical section for each rotation, we show either that the values in the tree are unchanged (and the tree is still well-ordered), or that we have removed the node to be deleted.
+%elaborate on two-level structure
+
+\ignore{The construction of the lock variants is simple.
+We begin with a \emph{part-reference} ghost state, a common construction for sharing information between an invariant and the threads that act on it, with elements $\publichalf{a}$ (the current state $a$) and $\myhalf{\pi}{p}$ (partial information about the current state, annotated with a share $\pi$).
+When the share $\pi$ is 1, we know that $p = a$.
+(do we want to define this ourselves, or reference it somewhere?
+note also that this is what Iris calls an authoritative algebra) A lock variant with an assertion $R(a)$ is then defined as %notation
+$$\ell \lockvar_\pi R \triangleq \ell \islock \exists a.\ R(a) * \myhalf{\pi}{a}$$ %eliding ghost names here
+The share $\pi$ may be set to 1 if the lock's information about $a$ is always fully up to date, or a smaller share if it can become outdated due to operations on other parts of the data structure.
+Since $a$ is existentially quantified, a thread that changes the value of $a$ forgets it after releasing the lock (as required in a lock invariant), but it must have synchronized with a $\publichalf{a}$ held elsewhere (usually in a global invariant) that will remember the change.
+Thus, the ghost state acts as a bridge between the component lock invariant and the global abstract state of the data structure.
+%smaller example before we get into BST?
+}
+
+\ignore{From the atomic rules of section~\ref{atomic}, we derive the following rules for accessing and modifying pieces of the global abstract state:
+\begin{mathpar}
+\inference[\textsf{sync\_commit}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * \exists x_0',\,x_1'.\ (x_0, x_1) \leadsto (x_0', x_1')\ \land \\(\mathsf{my\_half}(g, \mathit{sh}, x_0') * \mathsf{public\_half}(g, x_1') \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}
+
+\inference[\textsf{sync\_rollback}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{public\_half}(g, x_1) \Rrightarrow P_p\ a * R')}{\begin{array}{c}\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R  \Rrightarrow \\\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R'\end{array}}
+\end{mathpar}
+A thread that holds $\myhalf{}{a}$ (i.e., is in the critical section of a lock with a lock variant) may interact with the abstract state $P_p$ held in an $\textsf{atomic\_shift}$, as long as the corresponding $\publichalf{}$ is part of the abstract state.
+It may either leave both halves as is and gain updated resources $R'$ (including, e.g., a snapshot of the current abstract state), or update both halves to a new state that satisfies the public postcondition $Q_p$, executing the linearization point and obtaining the final postcondition $Q$.
+Typically we will use one of these rules at the end of each critical section, to show that the changes made in the critical section either are invisible to other threads (\textsf{sync\_rollback}) or have fulfilled the linearization point (\textsf{sync\_commit}).
+We can also derive simpler rules for special cases, such as a version of \textsf{sync\_commit} for read-only operations that do not change the state of any component:
+
+$$\inference[\textsf{sync\_commit\_same}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{my\_half}(g, \mathit{sh}, x_0) * \mathsf{public\_half}(g, x_1) \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
+}
+
+\subsection{Atomicity with atomic locks}
+Proving atomic specifications for a data structure using atomic specifications for locks is more immediate.
+As an example, consider Patel et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}.
+%outline how this worked for them
+
+In particular, in the fine-grained case, there is no need to create extra ghost state to abstract the state of each locked component, and then synthesize it into a top-level abstract state for the data structure.
+Instead, each lock acquire and release can atomically access the top-level abstract state, and directly make changes there.
+It is even possible for a thread to acquire a lock without owning a share of the lock assertion: the lock assertion can be extracted directly from the top-level state during the atomic \texttt{acquire} operation, and returned when \texttt{acquire} returns.
+This completely circumvents the recursive-lock reasoning needed for hand-over-hand locking with lock-invariant specs.
+%This comes later in the current version.
+
+\section{A Verified Binary Search Tree with Hand-over-Hand Locking}
 \label{hand-over-hand}
+
+We demonstrate our approach by verifying a binary search tree (BST) with fine-grained locking in VST, using the above approach to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
 As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
@@
 
-\subsection{BST code}
+\ignore{\subsection{BST code}
 \label{selflock}
@@
 \end{figure}
-
-\subsection{Fine-grained locking and atomicity}
-\label{atomicity}
-
-To prove that an operation on a data structure satisfies a logically atomic specification, we must show that there are no visible intermediate states of the operation, i.e., that other threads see the data structure as unchanged until the linearization point at which the operation takes effect.
-In traditional concurrent separation logics such as VST's~\cite{oraclesematic}, this is done by protecting the data structure with a single lock whose \emph{lock invariant} describes consistent states of the data structure.
-In modern separation logics such as Iris~\cite{iris} or FCSL~\cite{fcsl}, we can make a series of atomic modifications to the data structure without continuously holding a lock, as long as one modification can be designated the \emph{linearization point} at which the new state becomes public, and all prior changes are invisible to other threads.
-Fine-grained locking, in which the pieces of a data structure are each protected by a separate lock with a traditional lock invariant, requires us to synthesize these two approaches: we may make a series of non-atomic modifications to a part of the data structure as long as we hold its lock, but when we release the lock we need to ensure that either we have reached the linearization point, or our changes are invisible to other threads.
-
-We accomplish this using a technique derived from unpublished work by Jung~\cite{ralf-convo} for treating the critical sections of locks as atomic operations.
-The key is to include in each lock invariant a piece of ghost state that represents the abstract state of the locked section.
-From the outside, the effect of the critical section is to atomically update that piece of ghost state, and all other changes are internal to the lock.
-These pieces of ghost state can then be composed to yield the abstract state of the entire data structure.
-In effect, this construction turns a traditional lock invariant into a \emph{lock variant} $\ell \lockvar R$, where $R$ is parameterized by a value that can be updated in the critical section, as long as it corresponds to a linearization point or invisible change to the abstract state of the data structure as a whole.
-This allows traditional lock invariants to interact with atomic shifts and be used to prove atomic specifications.
-
-The construction of the lock variants is simple.
-We begin with a \emph{part-reference} ghost state, a common construction for sharing information between an invariant and the threads that act on it, with elements $\publichalf{a}$ (the current state $a$) and $\myhalf{\pi}{p}$ (partial information about the current state, annotated with a share $\pi$).
-When the share $\pi$ is 1, we know that $p = a$.
-(do we want to define this ourselves, or reference it somewhere?
-note also that this is what Iris calls an authoritative algebra) A lock variant with an assertion $R(a)$ is then defined as %notation
-$$\ell \lockvar_\pi R \triangleq \ell \islock \exists a.\ R(a) * \myhalf{\pi}{a}$$ %eliding ghost names here
-The share $\pi$ may be set to 1 if the lock's information about $a$ is always fully up to date, or a smaller share if it can become outdated due to operations on other parts of the data structure.
-Since $a$ is existentially quantified, a thread that changes the value of $a$ forgets it after releasing the lock (as required in a lock invariant), but it must have synchronized with a $\publichalf{a}$ held elsewhere (usually in a global invariant) that will remember the change.
-Thus, the ghost state acts as a bridge between the component lock invariant and the global abstract state of the data structure.
-%smaller example before we get into BST?
-
-From the atomic rules of section~\ref{atomic}, we derive the following rules for accessing and modifying pieces of the global abstract state:
-\begin{mathpar}
-\inference[\textsf{sync\_commit}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * \exists x_0',\,x_1'.\ (x_0, x_1) \leadsto (x_0', x_1')\ \land \\(\mathsf{my\_half}(g, \mathit{sh}, x_0') * \mathsf{public\_half}(g, x_1') \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}
-
-\inference[\textsf{sync\_rollback}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{public\_half}(g, x_1) \Rrightarrow P_p\ a * R')}{\begin{array}{c}\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R  \Rrightarrow \\\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R'\end{array}}
-\end{mathpar}
-A thread that holds $\myhalf{}{a}$ (i.e., is in the critical section of a lock with a lock variant) may interact with the abstract state $P_p$ held in an $\textsf{atomic\_shift}$, as long as the corresponding $\publichalf{}$ is part of the abstract state.
-It may either leave both halves as is and gain updated resources $R'$ (including, e.g., a snapshot of the current abstract state), or update both halves to a new state that satisfies the public postcondition $Q_p$, executing the linearization point and obtaining the final postcondition $Q$.
-Typically we will use one of these rules at the end of each critical section, to show that the changes made in the critical section either are invisible to other threads (\textsf{sync\_rollback}) or have fulfilled the linearization point (\textsf{sync\_commit}).
-We can also derive simpler rules for special cases, such as a version of \textsf{sync\_commit} for read-only operations that do not change the state of any component:
-
-$$\inference[\textsf{sync\_commit\_same}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{my\_half}(g, \mathit{sh}, x_0) * \mathsf{public\_half}(g, x_1) \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
+}
+
+\subsection{Hand-over-Hand Locking}
+\label{selflock}
+
+Hand-over-hand locking (also called lock coupling) is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it.
+Hand-over-hand locking is a classic example of a pattern that requires reference to lock invariants inside other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}.
+The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock.
+By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.
+
+This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources.
+In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
+We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$?
+Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
+$$\begin{array}{l}
+\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
+\texttt{acquire}(\ell');\\
+\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
+\texttt{release}(\ell);\\
+\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
+\end{array}$$
+In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$.
+This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share.
+Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al.
+instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
+$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$
+
+Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
+$$\begin{array}{l}
+\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
+\texttt{acquire}(\ell');\\
+\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
+\texttt{release}(\ell);\\
+\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
+\end{array}$$
+Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
+The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
 
@@
 
+\usetikzlibrary{positioning}
+
 \begin{figure}[htb]
 \centering
-\includegraphics[width=70mm,scale=0.5]{../FIG/range_prop.png}
-\caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
-\label{range_bst}
-\end{figure}
-
-\usetikzlibrary{positioning}
 \forestset{
@@
 \end{forest}
+\caption{A BST with each node labeled with its range}
+\label{range_bst}
+\end{figure}
 
@@
 \end{mathpar}
-where $\treerep\ m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock variant for the root node:
+where $\treerep\ m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock variant for the root node: %get rid of lock variant terminology and notation?
 $$\nodeboxrep(b) \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \lockvar \mathsf{node\_inv}$$
@@
 Similarly, we divide our concurrency proofs into two parts: code-specific proofs that describe how the C code implements changes to per-node ghost state, and more abstract proofs describing how changes in local ghost state lead to the desired updates to the global state.
+%talk less about this, since it's already explained above
 
@@
 
-\section{Lock-Free BST}
+\ignore{\section{Lock-Free BST}
 We want to prove that a lock-free BST implementation satisfies the same specification as our hand-over-hand implementation.
@@
 \end{mathpar}
-though the precise definitions of $\nodeboxrep$ and $\treerep$ will differ. %something about how the fact that the specs are the same really comes across to a client -- and should we prove a client?
+though the precise definitions of $\nodeboxrep$ and $\treerep$ will differ. %something about how the fact that the specs are the same really comes across to a client
+}
 
@@
 \label{related}
-Our work is based on the concurrent separation logic of VST in the style of Iris \cite{higherorderghoststate} and the logical atomicity from TaDA logic  \cite{tada}.
+Gotsman et al.~\cite{gotsman}, in the same paper in which they introduced invariant-style lock specs, also verified a linked list with hand-over-hand locking, which became a common example for CSLs that handled fine-grained locking.
+Among the most recent of these is the work of Patel et al.~\cite{templates}, who used atomic lock specs to prove correctness of the same data structure.
+This work updates the work of Gotsman et al.
+with modern specifications, and directly compares the results to those of Patel et al.
+
+%Our work is based on the concurrent separation logic of VST in the style of Iris \cite{higherorderghoststate} and the logical atomicity from TaDA logic  \cite{tada}.
 The notion of the \emph{lower} and \emph{upper} bound is taken from the flow interface paper \cite{krishna2017flow}.
@@
 
-Gotsman and Yang \cite{gotsman} is one of the earliest work on concurrent separation logic.
-They introduced the program logic to reason locally about the heap-manipulating program with the notion of dynamic ownership of heap parts by unbounded number of locks and threads, and shown the verification of singly-linked list with fine-grained locking.
-Xiong et al.
+%Xiong et al.
 \cite{Xiong2017Abstract} have demonstrated the verification of ConcurrentSkipListMap from java.util.concurrent library using the recent advances in fine-grained concurrency reasoning.
@@
 
-Krishna et al.
-\cite{krishna2017flow} have presented the proof technique for concurrent search structure templates based on the flow framework and the Iris separation logic, and verified the implementation of concurrent B-tree, hash tables, and linked lists based on the templates.
-Their work is closely related to our work; we took some inspiration from them in building our ghost state.
-But, their proofs are on a toy language rather than real C code.
-
 %TaDA-live: proves atomicity specs of some lock-based data structures incidentally, no mechanization
@@
 \label{conclusion}
-
-\subsubsection{Acknowledgements} I want to thank Zhen Zheng and Ralf Jung, ...
+While TaDA-style atomic specs for locks have been widely adopted%?
+, the implications of the switch away from invariant-based specs had not been thoroughly examined.
+We have now demonstrated that invariant-based specs can still be used to prove logically atomic specs for larger data structures (even for fine-grained implementations) and that this requires significantly more complex proofs than using atomic lock specs (especially for fine-grained implementations).
+The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the two-level proof structure we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures.
+In all other settings, however, we should strongly prefer to use atomic specifications for locks to simplify our proofs.
+Indeed, we intend to investigate shifting the foundations of VST to use atomic operations and derive atomic specs for locks: while updating the soundness proof may be an enormous effort, we believe we have shown it to be worth the trouble in the long run.
+
+\subsubsection{Acknowledgements} This work benefited greatly from correspondence with Ralf Jung, who pointed out the atomic syncer example and explained the finer points of logical atomicity in Iris.
 

### 2022-04-08 concurrency/ASL 2022/intro.tex
restructured paper, updated todo

@@
 the CompCert compiler~\cite{compcert}, and has recently been extended with support for general ghost state and atomicity proofs.
-We demonstrate that 1) this approach can be used to prove atomic specs for interesting data structures with fine-grained locking, and 2) these proofs are much more complicated than those built with style 2 lock specs.
-The key difference is that in style 1, we must associate each locked component with some resources and then relate those resources to the state of the data structure as a whole, while in style 2 each lock access can interact directly with the full state of the data structure.
+We demonstrate that 1) \textbf{this approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs are much more complicated than those built with style 2 lock specs}.
+The key difference is that in style 1 we must associate each locked component with some resources and then relate those resources to the state of the data structure as a whole, while in style 2 each lock access can interact directly with the full state of the data structure.
 However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise.
+To our knowledge, this is the first formal comparison of the two styles of lock specifications, and our technique may be useful in tools like VST where the older style is deeply integrated into the tool.
 

### 2022-04-25 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
editing up to delete

@@
 
-\newcommand{\myhalf}[2]{\ensuremath{\mathsf{my\_half}_{#1}(#2)}}
-\newcommand{\publichalf}[1]{\ensuremath{\mathsf{public\_half}(#1)}}
+\newcommand{\nodeghost}[2]{\ensuremath{\mathsf{node\_ghost}_{#1}\ (#2)}}
+\newcommand{\myhalf}[2]{\ensuremath{\mathsf{node\_ghost}^{.5}_{#1}\ (#2)}}
+\newcommand{\publichalf}[2]{\ensuremath{\mathsf{node\_ghost}^{.5}_{#1}\ (#2)}}
 
@@
 
-The Verified Software Toolchain (VST)~\cite{plfcc} is a Coq-based system for proving separation logic specifications of C programs.
+Our proof technique is developed in a combination of two Coq-based separation logic verification systems, the Verified Software Toolchain (VST) and Iris.
+VST~\cite{plfcc} is a Coq-based system for proving separation logic specifications of C programs.
 It is proved sound against not only the C semantics of the CompCert compiler, but also the compiler correctness theorem~\cite{cpm}, so that verified programs are guaranteed to behave correctly when compiled and run.
@@
 
-Iris~\cite{iris} is a Coq-based separation logic framework that synthesizes ideas from a range of CSLs with the idea that ``ghost state is all you need''.
+Iris~\cite{iris} is a framework that synthesizes ideas from a range of CSLs with the idea that ``ghost state is all you need''.
 Iris begins with a small core logic including arbitrary higher-order ghost state, and derives features including invariants and logical atomicity from particular instances of ghost state.
@@
 \section{Locks and Atomicity}
+\label{atomic}
 Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
 \begin{mathpar}
-\{\mathsf{BST}\ p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\ p\ (\mathsf{insert}\ t\ k\ v)\}
-
-\{\mathsf{BST}\ p\ t\}\ \texttt{lookup}(p, k)\ \{v.\ \mathsf{BST}\ p\ t \land \mathsf{lookup}\ t\ k = v\}
-
-\{\mathsf{BST}\ p\ t\}\ \texttt{delete}(p, k)\ \{\mathsf{BST}\ p\ (\mathsf{delete}\ t\ k)\}
+\{\mathsf{BST}\ p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\ p\ (\mathrm{abs\_insert}\ t\ k\ v)\}
+
+\{\mathsf{BST}\ p\ t\}\ \texttt{lookup}(p, k)\ \{v.\ \mathsf{BST}\ p\ t \land \mathrm{abs\_lookup}\ t\ k = v\}
+
+\{\mathsf{BST}\ p\ t\}\ \texttt{delete}(p, k)\ \{\mathsf{BST}\ p\ (\mathrm{abs\_delete}\ t\ k)\}
 \end{mathpar}
 where $\mathsf{BST}\ p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$.
-For instance, $t$ may be a recursively defined tree type in Coq, and $\mathsf{insert}$ may be a pure Coq function that adds a key-value pair to such a tree; $\mathsf{BST}$ could then be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, and $p$ could be the location of the root node.
+For instance, $t$ may be a recursively defined tree type in Coq, and $\mathrm{abs\_insert}$ may be a pure Coq function that adds a key-value pair to such a tree; $\mathsf{BST}$ could then be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, and $p$ could be the location of the root node.
 
@@
 \begin{mathpar}
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ t\ k)\rangle
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle
+
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle
+
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
 \end{mathpar}
 These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call.
-Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: in separation logic terms, the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.
+We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource.
+We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access).
+Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.
 
@@
 For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$.
-Then we can prove the atomic triples above, where at the linearization point we access $\mathsf{bst\_abs}$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then satisfy the atomic postcondition with one half of the ghost state and return the other half to the lock invariant.
+Then we can prove the atomic triples above, where at the linearization point we access $\treerep$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then satisfy the atomic postcondition with one half of the ghost state and return the other half to the lock invariant.
 The atomic postcondition reflects the fact that the abstract state has been accessed or changed, while the lock invariant maintains the connection between the abstract state and the actual data structure in memory.
@@
 Proving atomic specifications for a data structure using atomic specifications for locks is more immediate.
-As an example, consider Patel et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}.
+As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}.
 %outline how this worked for them
@@
 \begin{mathpar}
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ t\ k)\rangle
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle
+
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle
+
+\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
 \end{mathpar}
-In our case, insert, lookup, and delete are simple Gallina functions on a functional tree data structure.
-The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract tree $t$.
-
-In the following sections, we describe the construction of these predicates, and the proofs of the operations, for BST implementations based on hand-over-hand locking and lock-free atomic operations.
+where insert, lookup, and delete are simple Gallina functions on a functional tree data structure.
+The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree.
+In this section, we describe the construction of these predicates, and the proofs of the operations, for a BST implementation based on hand-over-hand locking, using lock-invariant-based specs for our locks.
 
@@
 %%%%%%%%%%%%%%%%%%%%%
-
-\ignore{In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct.
-To prove the correctness, our threads need to be able to record the information about the actions they have performed on the shared state, instead of sealing all knowledge of the shared data structure inside the lock invariant.
-We can accomplish this with ghost variables, a simple form of auxiliary state.
-Aside from the \emph{permission} on shared resources held by each thread that we used in the safety proofs, the verification of correctness properties also involves the \emph{ghost state}.
-In VST, any Coq type can be used as ghost state, as long as we can describe what happens when two elements of that type are joined together.
-We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \emph{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.}
-
-%Linearizability is one of the most popular strong correctness condition for the most of %concurrent data-structure, which implies that every operation appears to take place atomically, %in some order, consistent with the real-time ordering of those operations.
-%In order to record the action each thread performed as a linear history, we will define global %invariants using the ghost state, which are similar to lock invariants but are not associated %with any particular memory location.
-Instead, a global invariant is true before and after every %step of a program, acting as a publicly accessible resource.
-A program instruction can use the %contents of global invariant if it can guarantee that no one will ever see an intermediate state %in which the invariant does not hold.
 
@@
 
+%make this more concise, and maybe merge it into a later section
 This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources.
@@
 In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure.
-As described above, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$.
+As described in section \ref{atomicity}, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$.
 We also need to make sure that our lock invariants support the hand-over-hand pattern.
 
-\subsubsection{Key Ranges}
+\subsubsection{Key Ranges}%can cut this down if need be
 When we traverse the BST looking for a key $k$, we will reach either a node containing $k$, or an empty node where $k$ would appear if it was in the tree.
@@
     circle, draw, minimum size=3ex, inner sep=1pt,
-    s sep=3mm
+    s sep=3mm, l sep=0mm
         }
@@
 This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations.
-More precisely, the abstract state of a node in the tree consists of its range plus a representation of its contents: for a leaf node, $\mathsf{None}$, and for an internal node, $\mathsf{Some}$ of its current key, value, and the identifiers of its left and right children.
-This per-node abstract state serves as a bridge between the per-node locks of the implementation and the abstract state of the entire tree (the $\treerep$ predicate) in the specification.
-
-The abstract state of a tree is a map from keys to values.
-The $\treerep$ predicate then describes the conditions under which a map $m$ is implemented by the ghost state of a tree.
-Roughly speaking, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$.
-We formalize this by first defining the collection of $\mathsf{public\_half}$ assertions---i.e., the authoritative copies of the per-node ghost states---needed to form an abstract tree:
+More precisely, the abstract state of a node in the tree is of type $\mathsf{range} \times \mathsf{option\ (Z \times val \times gname \times gname)}$, where the second component is $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, l, r)$ for an internal node with key $k$, value $v$, and left and right children $l$ and $r$.
+The predicate $\nodeghost{g}{r, n}$ asserts that the node with identifier $g$ has range $r$ and contents $n$, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.
+
+The $\treerep$ predicate describes the conditions under which an abstract map $m$ is implemented by the ghost state of a tree.
+Intuitively, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$.
+We formalize this by first defining the collection of per-node ghost state fragments needed to form an abstract tree:
 $$\begin{array}{r l}
 \mathsf{public}&\mathsf{\_halves}(t, g, (l, r)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
-& |\ \mathsf{Leaf} \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{None})\\
-& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{Some}\ (k, v, g_l, g_r)) * \mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
+& |\ \mathsf{Leaf} \Rightarrow \publichalf{g}{(l, r), \mathsf{None}}\\
+& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \publichalf{g}{(l, r), \mathsf{Some}\ (k, v, g_l, g_r)}\ * \\&\qquad\mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
 \end{array}$$
-We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{public\_half}$ assertions of each node into a single assertion.
+We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{node\_ghost}$ assertions of each node into a single assertion.
 We compute the range of each node's children as described above: if the key in the current node is $k$, then the left child has range $(l, k)$ and the right child has range $(k, r)$.
 
-This collection of $\mathsf{public\_half}$ assertions can be combined with per-node lock variants to ensure that the nodes in an implementation actually form a valid tree.
-However, we also need to record the fact that each node in the implementation actually belongs to the tree $t$.
-We do this by augmenting $\mathsf{public\_halves}$ with a piece of ghost state $\mathsf{ghost\_nodes}(S)$ that records the set of all nodes in the tree, constructed so that each node can hold an $\mathsf{in\_tree}\ g$ assertion guaranteeing that $g \in S$.
+This collection of per-node ghost state describes the contents and structure of the tree, but a thread that acquires a node's lock also needs to know that the node it contains actually appears in the tree $t$.
+For this purpose, we add a ghost-state set $\mathsf{ghost\_nodes}(S)$ that records the identifiers of all nodes in the tree, defined so that each node can hold an $\mathsf{in\_tree}\ g$ assertion guaranteeing that $g \in S$.
 Then our final definition of $\treerep$ is:
-$$\treerep(m, g) \triangleq \exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty)) * \mathsf{ghost\_nodes}(\mathsf{node\_names}(t))$$
+$$\begin{array}{r l}\treerep_g\ m, \triangleq &\exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty))\ * \\&\mathsf{ghost\_nodes}(\mathsf{ids}(t))\end{array}$$
 The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in the ghost tree $t$.
-Because we quantify over the abstract tree $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node names and key-value pairs.
-(We will take advantage of this property when verifying the \lstinline{delete} function.)
-
-As described in section~\ref{atomicity}, we connect the per-node locks of our C implementation with the global abstract state of $\treerep$ by associating each node's lock with a lock variant assertion $\ell \islock_\pi R$, where $R$ is parameterized by the abstract state of the node (range + possible contents).
-We will fix the share $\pi$ to 1/2 so that the range in a node's lock can be out of date with respect to its $\mathsf{public\_half}$ (in particular, it may have grown due to rearrangements further up the tree).
-We then need to choose a variant $R$ describing the resources held by the lock for a given abstract state.
+Because we quantify over the abstract tree $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node IDs and key-value pairs.
+We will take advantage of this property when verifying the \lstinline{delete} function.
+
+Next, we connect the per-node locks of our C implementation with the global abstract state of $\treerep$ by associating each node's lock with a lock invariant assertion $\ell \islock R$, where $R$ includes both the concrete points-to assertion for the node and the corresponding abstract $\mathsf{node\_ghost}$.
 The C implementation of a node is:
 \begin{lstlisting}[language = C,numbers = none]
-typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
 typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
 \end{lstlisting}
-The node data of each \lstinline{tree} is wrapped in a \lstinline{tree_t} struct, and is only accessible when the lock in the \lstinline{lock} field is held.
-The lock variant for a node at location $p$ then needs to describe the contents of the \lstinline{t} field for each possible abstract state:
+where the \lstinline{t} field points to a \lstinline{tree} struct containing the key, value, and child pointers, and the \lstinline{lock} field contains the lock protecting the node.
+The lock invariant for a node at location $p$ needs to describe the contents of the \lstinline{t} field for each possible abstract state:
 $$\begin{array}{l}
@@
 $$\begin{array}{r l}
-\mathsf{node\_inv}((l, r), c) \triangleq \ &p.\texttt{lock} \lockvar_{\pi_1} \mathsf{node\_inv} * \mathsf{match}\ c\ \mathsf{with}\\
+\mathsf{node\_inv}_g \triangleq \ &\exists l, r, c.\ \myhalf{g}{((l, r), c)} * p.\texttt{lock} \islock_{\pi_1} \mathsf{node\_inv}_g * \mathsf{match}\ c\ \mathsf{with}\\
 &|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
-&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * \\&\qquad p_l.\texttt{lock} \lockvar_{\pi_2} \mathsf{node\_inv} * p_r.\texttt{lock} \lockvar_{\pi_2} \mathsf{node\_inv}
+&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)\ * \\&\qquad p_l.\texttt{lock} \islock_{\pi_2} \mathsf{node\_inv}_{g_l} * p_r.\texttt{lock} \islock_{\pi_2} \mathsf{node\_inv}_{g_r}
 \end{array}$$
-While we do not know the ranges or contents of the child nodes, the lock-variant construction guarantees that they are consistent with the $\mathsf{public\_half}$ assertions of the global ghost state, and the hand-over-hand pattern allows us to acquire $p_l.\texttt{lock}$ or $p_r.\texttt{lock}$ and then release $p.\texttt{lock}$ without losing any shares of the lock assertions.
+The hand-over-hand pattern allows us to acquire $p_l.\texttt{lock}$ or $p_r.\texttt{lock}$ and then release $p.\texttt{lock}$ without losing any shares of the lock assertions.
 
@@
 \begin{mathpar}
-\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle
-
-\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{lookup}\ m\ k = v\rangle
-
-\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ m\ k)\rangle
+\forall m.\,\langle \nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{insert}\ m\ k\ v)\rangle
+
+\forall m.\,\langle \nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep_g\,p\ |\ \treerep_g\,m \land \mathrm{lookup}\,m\ k = v\rangle
+
+\forall m.\,\langle \nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{delete}\ m\ k)\rangle
 \end{mathpar}
-where $\treerep\ m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock variant for the root node: %get rid of lock variant terminology and notation?
-$$\nodeboxrep(b) \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \lockvar \mathsf{node\_inv}$$
+where $\treerep_g\,m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock for the root node:
+$$\nodeboxrep_g\,b \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \islock \mathsf{node\_inv}_g$$
 Now we are prepared to prove that each of our BST functions satisfies its specification.
@@
 \subsection{Proofs: Insert and Lookup}
-The basic verification style in VST is to divide proofs into two parts: proofs that the C implementations meet specifications that are Coq-level encodings of their behavior, and proofs that those encodings actually have the desired properties.
-In the single-threaded case, we might prove that some code implements a binary search tree written in Gallina, and then prove that that tree satisfies map axioms.
-Similarly, we divide our concurrency proofs into two parts: code-specific proofs that describe how the C code implements changes to per-node ghost state, and more abstract proofs describing how changes in local ghost state lead to the desired updates to the global state.
-%talk less about this, since it's already explained above
+%How much detail do we want here?
 
@@
 The function has two possible linearization points: after line 23, where we finish adding a node $(k, v)$ to the tree, and after line 41, where we change the value of an existing node with key $k$ to $v$.
-In both cases, we use the \textsf{sync\_commit} rule to atomically update the abstract state of the data structure from $\treerep\ m$ to $\treerep (\mathrm{insert}\ m\ k\ v)$, by showing that the update to the tree implementation corresponds to a local change in the abstract tree that implements the mathematical $\mathsf{insert}$.
+In both cases, we atomically update the abstract state of the data structure from $\treerep\ m$ to $\treerep (\mathrm{insert}\ m\ k\ v)$, by showing that the update to the tree implementation corresponds to a local change in the abstract tree that implements the mathematical $\mathsf{insert}$.
+The proofs are split into a proof that the C code implements the local update, and a proof that the local update to a ghost-state tree implements the global update.
 In the first case, the local change is the insertion of $(k, v)$ at an empty leaf whose range includes $k$, and the addition of two new empty leaves as its children; in the second case, we simply change the value of an internal node whose key is already $k$.
-Our higher-level proofs, then, need to show that these two changes are both valid operations on ghost trees that yield the desired insertion in the global state.
-
+
+\begin{figure}[htb]
+\centering
+\begin{forest}
+for tree={
+    grow=south,
+    circle, draw, minimum size=3ex, inner sep=1pt,
+    s sep=3mm, l sep=0mm
+        }
+    [, phantom, for children={fit=band}, s sep+=20mm,
+         before drawing tree={%
+          tikz+={%
+            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
+            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
+            \node [anchor=south] at ($(a.east)!1/2!(b.west)$) {$\Longrightarrow\quad$};
+          },
+        },
+        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                                [, double, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$};} %weirdly lopsided
+                                ]
+                            ]
+                            [,no edge, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
+        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
+            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
+                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
+                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
+                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
+                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                                [k, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$};}
+                                    []
+                                    []
+                                ]
+                            ]
+                            [,no edge, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
+    ]
+\end{forest}
+\caption{Adding a new key (38) at a leaf}
+\label{extract_insert}
+\end{figure}
+
+\ignore{\begin{figure}[htb]
+\centering
 \begin{forest}
@@
             \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
-            \node [anchor=center] at ($(a.east)!1/2!(b.west)$) {$\longrightarrow\quad$};
+            \node [anchor=south] at ($(a.east)!1/2!(b.west)$) {$\Longrightarrow\quad$};
           },
@@
                             ]
-                            [,no edge, draw=none]
-                        ]
-                        [,no edge, draw=none]
-                    ]
-                    [,no edge, draw=none]
-                ]
-            ]
-        ]
-        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
-            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
-            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
-                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
-                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
-                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
-                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
-                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
-                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
-                                [k, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$}; \node[left=0pt of .west, scale=0.35] {$\mathrm{g\_current}$};}
-                                    [, tikz={\node[left=0pt of .west, scale=0.75]  {$g_1$};}]
-                                    [, tikz={\node[left=0pt of .west, scale=0.75]  {$g_2$};}]
-                                ]
-                            ]
-                            [,no edge, draw=none]
-                        ]
-                        [,no edge, draw=none]
-                    ]
-                    [,no edge, draw=none]
-                ]
-            ]
-        ]
-    ]
-\end{forest}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=150mm,scale=0.5]{../FIG/extract_insert.png}
-\caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation }
-\label{extract_insert}
-\end{figure}
-\begin{figure}[htb]
-\centering
-\includegraphics[width=150mm,scale=0.5]{../FIG/extract_insert_2.png}
-\caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation, where key $k$ already exists in the tree $t$ }
-\label{extract_insert2}
-\end{figure}
-
-\begin{forest}
-for tree={
-    grow=south,
-    circle, draw, minimum size=3ex, inner sep=1pt,
-    s sep=3mm
-        }
-    [, phantom, for children={fit=band}, s sep+=20mm,
-         before drawing tree={%
-          tikz+={%
-            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
-            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
-            \node [anchor=center] at ($(a.east)!1/2!(b.west)$) {$\longrightarrow\quad$};
-          },
-        },
+                            [$g_2$, draw=none]
+                        ]
+                        [,no edge, draw=none]
+                    ]
+                    [,no edge, draw=none]
+                ]
+            ]
+        ]
         [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
@@
         ]
-        [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
-            [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
-            [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
-                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
-                [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
-                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
-                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
-                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
-                                [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
-                                [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$}; \node[left=0pt of .west, scale=0.35] {$\mathrm{g\_current}$};}
-                                ]
-                            ]
-                            [$g_2$, draw=none]
-                        ]
-                        [,no edge, draw=none]
-                    ]
-                    [,no edge, draw=none]
-                ]
-            ]
-        ]
     ]
 \end{forest}
+\caption{Updating an existing key with a new value}
+\label{extract_insert2}
+\end{figure}}
 
@@
 
-\subsubsection{Lookup}
+The proof of \lstinline{lookup} is similar---the linearization points are at the lookup of the target key if it exists in the tree, and at the leaf node whose range contains the target key if it is not present in the tree.
+The \lstinline{lookup} function does not change the structure of the tree, so no particular reasoning on the abstract state is required.
+
+\ignore{\subsubsection{Lookup}
 
@@
 \label{lookupproof}
-\end{figure}
+\end{figure} }
 
@@
 
-The code for the delete method is shown in ??. It takes the location of the root pointer
-and a key as the arguments. A thread traverses the tree using hand-over-hand locking,
-until the thread finds the key to be deleted. To remove the key, the tree is changed 
-through an operation called pushdown left such that the node to be deleted points to an empty leaf
-right child. To achieve that, the thread sets the right child's left child as the 
-new right child of the to-be-deleted node and moves the right child's right child as 
-the new parent of the to-be-deleted node. This operation is repeated until the right
-child of the to-be-deleted node is an empty leaf node. Then, the thread will delete
-the node and set the delete node's parent pointer to the left child of the deleted node.
-
-The atomic specification for delete method can be written as follows:
-$$\forall t.\ \langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(t,g)\rangle $$ 
-$$\texttt{delete}(p,k)$$
-$$\langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(\texttt{delete}(t, k),g)\rangle $$
+The code for the delete method is shown in ??.%pseudocode?
+The tree traversal is the same as in \lstinline{insert}, but the behavior upon finding the target node is more complicated.
+First, we call a helper function \lstinline{pushdown_left} that rearranges the nodes in the tree, maintaining the ordering of nodes but moving the node to be deleted down through the tree until its right child is empty.
+This is accomplished by repeatedly rotating sets of three nodes: the target node's right child is moved up to become its parent, with the target node on the left and the right child's right child moved under the target node, as illustrated in Figure~\ref{delete-diagram}.
+Once the node to be deleted has no right child, it is removed from the tree, and its left child becomes the left child of its parent.
+
+The atomic specification for \lstinline{delete} is:
+$$\forall t.\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (\mathrm{abs\_delete}\ m\ k)\rangle$$
 
@@
 
+\begin{figure}[htb]
+\centering
 \forestset{
@@
             \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
-            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
-            \node [anchor=center] at ($(a.east)!1/2!(b.west)$) {$\longrightarrow\quad$};
+            \node (c) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
+            \node (b) [anchor=south] at ($(a.east)!1/2!(c.west)$) {};
+            \node [anchor=south] at ($(a.east)!1/3!(b.west)$) {$\Longrightarrow\quad$};
+            \node [anchor=south] at ($(b.east)!1/2!(c.west)$) {$\Longrightarrow$};
           },
@@
             [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
-                [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
+                [40, very thick, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
                     [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}]
-                    [55, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,75)$};}
+                    [55, very thick, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,75)$};}
                         [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(40,55)$};}]
-                        [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
+                        [60, very thick, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
                     ]
@@
                 [55, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
-                    [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
+                    [40, very thick, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
                         [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(40,35)$};}]
-                        [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,55)$};}]
+                        [50, very thick, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,55)$};}]
                     ]
@@
                     [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
-                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                        [40, very thick, tikz={\node[left=1.8pt of .east, cross out, scale=2, draw]{}; \node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
                             [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
@@
         ]
-        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
+        \ignore{[ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
             [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
@@
             ]
-        ]
+        ]}
     ]
 \end{forest}
+\caption{Deleting node 40 from the tree via \lstinline{pushdown_left}}
+\label{delete-diagram}
+\end{figure}
 
@@
 Gotsman et al.~\cite{gotsman}, in the same paper in which they introduced invariant-style lock specs, also verified a linked list with hand-over-hand locking, which became a common example for CSLs that handled fine-grained locking.
-Among the most recent of these is the work of Patel et al.~\cite{templates}, who used atomic lock specs to prove correctness of the same data structure.
+Among the most recent of these is the work of Krishna et al.~\cite{templates}, who used atomic lock specs to prove correctness of the same data structure.
 This work updates the work of Gotsman et al.
-with modern specifications, and directly compares the results to those of Patel et al.
+with modern specifications, and directly compares the results to those of Krishna et al.
 

### 2022-04-27 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
delete proof edits

@@
 
-\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking\thanks{Supported by the NSF (details).}}
+\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking\thanks{Supported by the National Science Foundation via grant \#1811894.}}
 %
@@
                                 [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
-                                [, double, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$};} %weirdly lopsided
+                                [, very thick, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$};} %weirdly lopsided
                                 ]
@@
                                 [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
-                                [k, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$};}
+                                [38, very thick, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$};}
                                     []
@@
 
-The code for the delete method is shown in ??.%pseudocode?
+The code for the \lstinline{delete} method is shown in ??. %pseudocode?
 The tree traversal is the same as in \lstinline{insert}, but the behavior upon finding the target node is more complicated.
@@
 
-The delete method is unique compared to insert and lookup because the pushdown left 
-operation can change the \textit{public} half range of more than one nodes. In addition,
-the pushdown left operation is one of the main reasons why the ghost tree of the binary
-search tree is not just a ``copy" of the pure tree: the functional model of pushdown\_left
-is unable to account for the change of deleted node's parent's pointer from pointing
-at the deleted node the to the deleted node's right child.
+The \lstinline{delete} function is responsible for most of the complexity of the abstract state.
+It changes the ranges of nodes in the tree in two ways.
+First, the rotation performed by \lstinline{pushdown_left} changes the ranges of the node to be deleted and its right child (e.g., nodes 40 and 55 in the first step of Figure~\ref{delete-diagram}); this is a local change, as the ranges of all other nodes (even the children of the rearranged nodes) are preserved.
+Second, the actual removal of the node increases the ranges of the subtree below it; for instance, in the last step of Figure~\ref{delete-diagram}, the range of node 35 changes from $(30, 40)$ to $(30, 50)$ when the node is removed.
+The local change occurs while \lstinline{pushdown_left} holds the locks of all nodes involved, so no intermediate states are visible to other threads, but the latter change occurs while holding only the lock of the deleted node, so the algebra for ranges must allow the ranges to increase at any time.
+Fortunately, \lstinline{insert} and \lstinline{lookup} only use the ranges to assert that the target key is in the range of the currently examined node, which is still true if the range increases.
+
+The linearization point of \lstinline{pushdown_left} is the removal of the target node.
+The rotation steps must therefore be considered as not changing the abstract state of the tree.
+This is why it is critical that the abstract state is a map from keys to values, rather than a more explicit representation of the tree's structure---other threads must not rely on particular arrangements of nodes in the tree, since any node that is not currently locked by a thread may be rearranged by another  thread.
+%how much more do we want to say here?
 
@@
                     [40, very thick, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
-                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(40,35)$};}]
+                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}]
                         [50, very thick, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,55)$};}]
@@
                     [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
-                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
+                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}]
                         [,no edge, draw=none]
@@
 
-\subsubsection{delete}
+\ignore{\subsubsection{delete}
 
@@
 to be preserved during the turn\_left operation in which no nodes were actually changed
-except for the structure of the tree.
+except for the structure of the tree.}
 

### 2022-04-27 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
clarified what we lose vs. atomic lock proofs

@@
 Lennart Beringer\inst{2}\orcidID{?} \and
-William Mansky\inst{4}\orcidID{0000-0002-5351-895X}}
+William Mansky\inst{4}\orcidID{0000-0002-5351-895X}} %confirm author order
 
@@
 At the time, there was no formal treatment of atomic operations in CSL, so actual lock implementations were beyond the scope of verification: these specs were taken as axioms, or proved through direct appeal to data-race-free guarantees of specific memory models.
+In garbage-collected languages that do not deallocate locks, the assertion $\ell \islock R$ can be freely duplicated and shared among threads accessing $R$; if we want to track ownership and eventually deallocate the lock, we can instead split the assertion into fractional shares such that $\ell \islock_{\pi_1} R * \ell \islock_{\pi_2} R = \ell \islock_{\pi_1 + \pi_2} R$, and then allow deallocation only when we own the full share $\ell \islock_1 R$.
 
@@
 A release simply moves atomically from the locked state to the unlocked state.
-These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(U(\ell) * R) \vee L(\ell)$, they can be used to derive the lock-invariant-based specs.
+These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs.
 It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g.
 compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it.
+In this style, tracking ownership of a lock is orthogonal to the lock assertion itself; a mechanism such as cancelable invariants~\cite{rustbelt-relaxed} can be used to both make the lock assertion $\mathsf{U}$/$\mathsf{L}$ publicly available and deallocate it once all threads are finished.
 
@@
 It is proved sound against not only the C semantics of the CompCert compiler, but also the compiler correctness theorem~\cite{cpm}, so that verified programs are guaranteed to behave correctly when compiled and run.
-VST directly implements the concurrent separation logic of Hobor et al.~\cite{oracle-sematics}, and the invariant-style lock specifications are incorporated into its soundness proof at a fundamental level.
+VST directly implements the concurrent separation logic of Hobor et al.~\cite{oraclesematic}, and the invariant-style lock specifications are incorporated into its soundness proof at a fundamental level.
 
@@
 Iris is designed as a language- and logic-independent framework, and recent versions of VST take advantage of this, incorporating Iris-style ghost state in the foundational model and directly using Iris tactics to reason about invariants and atomicity.
+Iris has also been used as the basis for proofs with atomic lock specifications, such as the work of Krishna et al.~\cite{templates}.
 All proofs described in this paper have been formalized in this VST+Iris setting.
@@
 \label{atomic}
+In this section, we outline our approach to connecting lock invariants and atomic specifications, and contrast it with the atomic-lock-based approach.
+
 Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
@@
 This is where ghost state and atomic triples come in: we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure.
-For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$.
+For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock_\pi R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$.
+%clarify share
 Then we can prove the atomic triples above, where at the linearization point we access $\treerep$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then satisfy the atomic postcondition with one half of the ghost state and return the other half to the lock invariant.
@@
 %elaborate on two-level structure
+As we will see in section~\ref{hand-over-hand}, the lock invariant for each component must also carefully account for the ownership of both that component's lock and the locks of related nodes (e.g., child nodes in a tree).
 
@@
 As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}.
-%outline how this worked for them
-
-In particular, in the fine-grained case, there is no need to create extra ghost state to abstract the state of each locked component, and then synthesize it into a top-level abstract state for the data structure.
-Instead, each lock acquire and release can atomically access the top-level abstract state, and directly make changes there.
-It is even possible for a thread to acquire a lock without owning a share of the lock assertion: the lock assertion can be extracted directly from the top-level state during the atomic \texttt{acquire} operation, and returned when \texttt{acquire} returns.
-This completely circumvents the recursive-lock reasoning needed for hand-over-hand locking with lock-invariant specs.
+The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
+The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.
+
+Note that in this case, the private precondition $\mathsf{inFP}(n)$ is merely the knowledge that $n$ is in the data structure, and does not involve any ownership---in particular, a caller does not need a share of the lock for the root node.
+A thread can acquire the lock for any node $n$ in the data structure by atomically accessing $\mathsf{CSS}$, extracting the sub-assertion corresponding to $n$, and switching it to the locked state.
+More generally, in this approach there is no need to track ownership of the locks of individual nodes, a source of considerable complexity in the lock-invariant-based specs.
 %This comes later in the current version.
@@
 
+This pattern is the main source of complexity in our proofs that is not present in proofs with atomic lock specifications.
+For instance, in the hand-over-hand locking proofs of Krishna et al.~\cite{templates}, the only information a node holds about its children is the fact that they are in the data structure; their lock assertions are held in the global abstract state along with the rest of their resources.
+This significantly simplifies the specification and proof of the hand-over-hand synchronization.
+
 \subsection{Specification of the hand-over-hand BST}
@@
 , the implications of the switch away from invariant-based specs had not been thoroughly examined.
-We have now demonstrated that invariant-based specs can still be used to prove logically atomic specs for larger data structures (even for fine-grained implementations) and that this requires significantly more complex proofs than using atomic lock specs (especially for fine-grained implementations).
-The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the two-level proof structure we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures.
-In all other settings, however, we should strongly prefer to use atomic specifications for locks to simplify our proofs.
-Indeed, we intend to investigate shifting the foundations of VST to use atomic operations and derive atomic specs for locks: while updating the soundness proof may be an enormous effort, we believe we have shown it to be worth the trouble in the long run.
+We have now demonstrated that invariant-based specs can still be used to prove logically atomic specs for larger data structures (even for fine-grained implementations) and that this requires somewhat more complex proofs than using atomic lock specs (especially for fine-grained implementations).
+The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the approach we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures.
+When possible, however, we would prefer to use atomic specifications for locks to simplify our proofs.
+Indeed, we intend to investigate shifting the foundations of VST to use atomic operations and derive atomic specs for locks: while updating the soundness proof will be a considerable effort, using the newer style of lock specification will simplify all future proofs about lock-based data structures.
 

### 2022-04-27 concurrency/ASL 2022/intro.tex
clarified what we lose vs. atomic lock proofs

@@
 
-In this paper, we adapt the latter approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t.
+In this paper, we adapt this approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t.
 the CompCert compiler~\cite{compcert}, and has recently been extended with support for general ghost state and atomicity proofs.
-We demonstrate that 1) \textbf{this approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs are much more complicated than those built with style 2 lock specs}.
-The key difference is that in style 1 we must associate each locked component with some resources and then relate those resources to the state of the data structure as a whole, while in style 2 each lock access can interact directly with the full state of the data structure.
+We demonstrate that 1) \textbf{our approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs require some additional complications that can be avoided with style 2 lock specs}.
+The key difference is that in style 1 a thread must gain ownership of the lock invariant assertion for a component before acquiring the lock and reading/modifying the component, while in style 2 the locks and components alike can be considered part of a single abstract state that is accessed at each lock access.
 However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise.

### 2022-05-04 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
templates

@@
 
-\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking\thanks{Supported by the National Science Foundation via grant \#1811894.}}
+\title{Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}}
 %
@@
 
-\keywords{concurrent separation logic, fine-grained locking, logical atomicity, Verified Software Toolchain}
+\keywords{concurrent separation logic, fine-grained locking, logical atomicity, Verified Software Toolchain, Iris}
 \end{abstract}

### 2022-05-05 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
Tune the layout of the figures.

@@
 \centering
-\forestset{
-   lab/.style={
-        label={[anchor=west, scale=0.75] north east:#1},
-    },
-}
+%% \forestset{
+%%    lab/.style={
+%%         label={[anchor=west, scale=0.75] north east:#1},
+%%     },
+%% }
 \begin{forest}
 for tree={
-    grow=south,
-    circle, draw, minimum size=3ex, inner sep=1pt,
-    s sep=3mm, l sep=0mm
+  calign=fixed edge angles,
+  calign angle=65,
+  circle, draw, minimum size=3ex, inner sep=1pt, l sep=0, l=0,
+%%     s sep=3mm, l sep=0mm
         }
@@
         [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
-            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+            [25, tikz={\node[left=-2pt of .west, scale=0.75]  {$(20,30)$};}]
             [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
-                [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                [45, tikz={\node[right=0pt of .south east, scale=0.75]  {$(30,50)$};}
                     [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
@@
   for tree={
-    grow=south,
+    calign=fixed edge angles,
+    calign angle=50,
     circle, draw, minimum size=3ex, inner sep=1pt,
-    s sep=3mm, l sep=0mm
-        }
-    [, phantom, for children={fit=band}, s sep+=20mm,
+    l=0, l sep=0.1em }
+    [, phantom, for children={fit=band}, s sep+=10mm,
          before drawing tree={%
           tikz+={%
-            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
-            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
-            \node [anchor=south] at ($(a.east)!1/2!(b.west)$) {$\Longrightarrow\quad$};
+            \node (a) [inner sep=0pt, fit=(!1) (!1ll) (!1ll1111)] {};
+            \node (b) [inner sep=0pt, fit=(!l) (!lll111l) (!l1)] {};
+            \node [anchor=south] at ($(a.east)!1/2!(b.west)$) {{\large$\Longrightarrow$}};
           },
@@
             [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
-                [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
+                [25, tikz={\node[left=-2pt of .west, scale=0.75]  {$(20,30)$};}]
                 [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
-                    [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
+                    [45, tikz={\node[right=0pt of .south east, scale=0.75]  {$(30,50)$};}
                         [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
@@
     circle, draw, minimum size=3ex, inner sep=1pt,
-    s sep=3mm
+    s sep=3mm, l sep=0
         }
@@
           tikz+={%
-            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
-            \node (c) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
+            \node (a) [inner sep=0pt, fit=(!1) (!1ll) (!1l1ll)] {};
+            \node (c) [inner sep=0pt, fit=(!l) (!ll) (!ll111)] {};
             \node (b) [anchor=south] at ($(a.east)!1/2!(c.west)$) {};
-            \node [anchor=south] at ($(a.east)!1/3!(b.west)$) {$\Longrightarrow\quad$};
-            \node [anchor=south] at ($(b.east)!1/2!(c.west)$) {$\Longrightarrow$};
+            \node [anchor=south] at ($(a.east)!2/5!(b.west)$) {$\Longrightarrow$};
+            \node [anchor=south] at ($(b.east)!3/5!(c.west)$) {$\Longrightarrow$};
           },

### 2022-05-10 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
small edits

--- 2022-05-04 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
@@
 \label{background}
+
+\subsection{Logically Atomic Specifications}
+\label{atomic}
+
+%revise
+For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads.
+This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations).
+In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+%describe weird binder
+Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$.
+For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution.
+%maybe something about linking ghost state, which parts are real vs.
+ghost, accessing invariants around atomic triple
+%Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t.
+the situation where one of the other threads' action following the linearization point removes the element just inserted.
 
@@
 \begin{mathpar}
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
+\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle
+
+\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle
+
+\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
 \end{mathpar}
@@
 Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.
-
-\subsection{Atomicity with lock invariants}
+%a bit more about how we prove these, and what the next couple sections are doing about them
+
+\subsection{Atomicity with lock invariants}%make clearer the connection to atomic specs
 The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end.
@@
 As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}.
-The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
+The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
 The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.
@@
 \section{A Verified Binary Search Tree with Hand-over-Hand Locking}
-\label{hand-over-hand}
-
-We demonstrate our approach by verifying a binary search tree (BST) with fine-grained locking in VST, using the above approach to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
+\label{hand-over-hand}%show code, somewhere, maybe cut-down pseudocode
+
+We demonstrate our approach to proving logically atomic specifications with invariant-based lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
 As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
 \begin{mathpar}
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
+\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle
+
+\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle
+
+\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
 \end{mathpar}
@@
 In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
-We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$?
-Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
+We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n'$'s %awkward
+ lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
 $$\begin{array}{l}
@@
 \begin{mathpar}
-\forall m.\,\langle \nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{insert}\ m\ k\ v)\rangle
-
-\forall m.\,\langle \nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep_g\,p\ |\ \treerep_g\,m \land \mathrm{lookup}\,m\ k = v\rangle
-
-\forall m.\,\langle \nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{delete}\ m\ k)\rangle
+\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{insert}\ m\ k\ v)\rangle
+
+\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep_g\,p\ |\ \treerep_g\,m \land \mathrm{lookup}\,m\ k = v\rangle
+
+\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{delete}\ m\ k)\rangle
 \end{mathpar}
@@
 Otherwise, it will eventually reach a leaf node, where it creates two new empty leaf nodes, inserts the key and value into the current node, and links the newly created leaf nodes to the current node.
-The atomic specification for insert method is $$\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle$$
+The atomic specification for insert method is $$\langle m.\,\nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle$$
 The \lstinline{insert} function should atomically add the key-value pair $(k, v)$ to the map implemented by the BST at $p$.
@@
 The atomic specification for $\mathsf{lookup}$ is
-$$\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle$$
+$$\langle m.\,\nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{lookup}\ m\ k = v\rangle$$
 The \texttt{lookup} function should atomically look up the value of the key $k$ in the tree at $p$.
@@
 The atomic specification for \lstinline{delete} is:
-$$\forall t.\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (\mathrm{abs\_delete}\ m\ k)\rangle$$
+$$\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (\mathrm{abs\_delete}\ m\ k)\rangle$$
 

### 2022-05-15 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
edits throughout

@@
 
-%revise
 For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads.
-This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations).
-In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
-%describe weird binder
-Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$.
-For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution.
-%maybe something about linking ghost state, which parts are real vs.
-ghost, accessing invariants around atomic triple
-%Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t.
-the situation where one of the other threads' action following the linearization point removes the element just inserted.
+This intuitive condition has been formalized in terms of linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations).
+In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form $$\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$ where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+This triple says that $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$.
+The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained.
+For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$.
+The $\mathsf{stack}$ itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the $\mathsf{is\_stack}$ assertion.
 
@@
 
-Later, the TaDA logic~\cite{tada} presented an alternative set of lock specifications, based on the newly introduced notion of logical atomicity:
+Later, the TaDA logic~\cite{tada}, which codified the notion of logical atomicity, presented an alternative set of lock specifications:
 $$\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle$$
@@
 \begin{mathpar}
-\{\mathsf{BST}\ p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\ p\ (\mathrm{abs\_insert}\ t\ k\ v)\}
-
-\{\mathsf{BST}\ p\ t\}\ \texttt{lookup}(p, k)\ \{v.\ \mathsf{BST}\ p\ t \land \mathrm{abs\_lookup}\ t\ k = v\}
-
-\{\mathsf{BST}\ p\ t\}\ \texttt{delete}(p, k)\ \{\mathsf{BST}\ p\ (\mathrm{abs\_delete}\ t\ k)\}
+\{\mathsf{BST}\:p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\:p\ (t[k \mapsto v])\}
+
+\{\mathsf{BST}\:p\ t\}\ \texttt{lookup}(p, k)\ \{v.\ \mathsf{BST}\:p\ t \land t(k) = v\}
+
+\{\mathsf{BST}\:p\ t\}\ \texttt{delete}(p, k)\ \{\mathsf{BST}\:p\ (t[k \mapsto \_])\}
 \end{mathpar}
-where $\mathsf{BST}\ p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$.
-For instance, $t$ may be a recursively defined tree type in Coq, and $\mathrm{abs\_insert}$ may be a pure Coq function that adds a key-value pair to such a tree; $\mathsf{BST}$ could then be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, and $p$ could be the location of the root node.
+where $\mathsf{BST}\:p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$.
+For instance, $t$ may be a recursively defined tree type in Coq, and $\mathsf{BST}$ would be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, with the root node located at $p$.
 
@@
 \begin{mathpar}
-\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle
-
-\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle
-
-\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
+\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto v])\rangle
+
+\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land m(k) = v\rangle
+
+\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto \_])\rangle
 \end{mathpar}
@@
 Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.
-%a bit more about how we prove these, and what the next couple sections are doing about them
-
-\subsection{Atomicity with lock invariants}%make clearer the connection to atomic specs
+
+To show that an implementation satisfies these atomic specifications, we must show that 1) the implementation can execute safely without owning any piece of the public assertion $\treerep$, only accessing it atomically and restoring it up until the linearization point, and 2) at some linearization point, the implementation atomically transforms the current map $m$ into one that satisfies the public postcondition.
+The linearization point may be a (physically or logically) atomic operation, or it may occur between steps of the program (e.g., if the current thread holds a lock on the modified section of the data structure, so other threads cannot distinguish between the original state and the modified state until the lock is released).
+In the following sections, we discuss the general structure of these proofs for lock-based implementations using each of the two kinds of lock specifications.
+
+\subsection{Atomicity with lock invariants}
 The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end.
-With the invariant-style lock specs, we can make $\mathsf{BST}\ p\ t$ the invariant of this lock: the lock then holds full ownership of the data structure, and provides it to whichever thread is currently executing an operation on the tree at $p$.
-In this \emph{coarse-grained} locking implementation, any thread with partial ownership of the lock can acquire the lock, perform all the operations of the sequential insertion/lookup/deletion, and then release the lock.
-Naively, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock---but this is not enough to tell us that we have changed the lock's state by the end of the operation, since we merely restore the invariant that there is \emph{some} tree $t$ at $p$.
-This is where ghost state and atomic triples come in: we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure.
+These \emph{coarse-grained} locking implementations are easy to verify with invariant-style lock specs: the lock's invariant can contain the whole data structure assertion (e.g.
+$\mathsf{BST}\:p\ t$), so that whichever thread is currently executing an operation on the tree at $p$ has full ownership of the data structure.
+If we only want to prove that the operation is thread-safe, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock.
+On the other hand, if we want to prove specifications that describe changes to the data structure---as in the atomic specification $\langle m.\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \treerep\ (m[k \mapsto v])\rangle$---we need ghost state to track the changes.
+
+Specifically, we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure.
 For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock_\pi R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$.
 %clarify share
-Then we can prove the atomic triples above, where at the linearization point we access $\treerep$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then satisfy the atomic postcondition with one half of the ghost state and return the other half to the lock invariant.
+Then we can prove the atomic triples above, where at the linearization point we access $\treerep$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then fulfill the atomic postcondition with one half of the ghost state and return the other half to the lock invariant.
 The atomic postcondition reflects the fact that the abstract state has been accessed or changed, while the lock invariant maintains the connection between the abstract state and the actual data structure in memory.
@@
 A thread can acquire the lock for any node $n$ in the data structure by atomically accessing $\mathsf{CSS}$, extracting the sub-assertion corresponding to $n$, and switching it to the locked state.
-More generally, in this approach there is no need to track ownership of the locks of individual nodes, a source of considerable complexity in the lock-invariant-based specs.
-%This comes later in the current version.
+More generally, in this approach there is no need to track ownership of the locks of individual nodes, which as we will see in section~\ref{selflock} is a source of considerable complexity in lock-invariant-based proofs.
 
@@
 \begin{mathpar}
-\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle
-
-\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle
-
-\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
+\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto v])\rangle
+
+\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land m(k) = v\rangle
+
+\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto \_])\rangle
 \end{mathpar}
@@
 \ignore{\subsection{BST code}
-\label{selflock}
 
@@
 The atomic specification for \lstinline{delete} is:
-$$\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (\mathrm{abs\_delete}\ m\ k)\rangle$$
+$$\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto \_])\rangle$$
 
@@
 
+VeriFast~\cite{verifast}, a separation-logic-based verifier for C and Java programs, also supports lock-based and atomic concurrency~\cite{verifast-conc}, and has been used to verify a hand-over-hand-locking linked list similar to that of Krishna et al.
+The specifications of both the lock operations and the list itself use a precursor of TaDA's logical atomicity.
+VeriFast is not a foundationally verified system, but its basic logic is verified against the semantics of a toy language in Coq.
+
 %Our work is based on the concurrent separation logic of VST in the style of Iris \cite{higherorderghoststate} and the logical atomicity from TaDA logic  \cite{tada}.
@@
 \label{conclusion}
-While TaDA-style atomic specs for locks have been widely adopted%?
-, the implications of the switch away from invariant-based specs had not been thoroughly examined.
+While most new CSL proofs of lock-based data structures use TaDA-style atomic specs for their locks, the implications of the switch away from invariant-based specs had not been thoroughly examined.
 We have now demonstrated that invariant-based specs can still be used to prove logically atomic specs for larger data structures (even for fine-grained implementations) and that this requires somewhat more complex proofs than using atomic lock specs (especially for fine-grained implementations).

### 2022-05-15 concurrency/ASL 2022/intro.tex
edits throughout

@@
 % Contributions:
-% verified C implementation of BST with atomic specs -- the first? VeriFast did it, but not atomic specs
+% verified C implementation of BST with atomic specs -- the first? VeriFast did it, with a precursor to atomic specs
 % technique for connecting locks with abstract state/atomicity
@@
 Concurrent separation logic (CSL)~\cite{csl} is a useful tool for proving correctness of concurrent programs.
-While early iterations focused on data structures implemented with locks~\cite{gotsman,oraclesematic} and specifications involving memory safety and data-race-freedom, more recent logics are able to prove linearizability-style functional correctness~\cite{tada} of lock-free implementations that use low-level atomic memory operations~\cite{iris}, even under weak memory models~\cite{gps}%update cite
-.
+While early iterations focused on data structures implemented with locks and specifications involving memory safety and data-race-freedom~\cite{gotsman,oraclesematic}, more recent logics are able to prove linearizability-style functional correctness of lock-free implementations that use low-level atomic memory operations~\cite{tada,iris}, even under weak memory models~\cite{rustbelt-relaxed}.
 In particular, \emph{logical atomicity}~\cite{tada} has become the gold standard for concurrent data structure specifications, capturing the idea that data structure operations should appear to take place instantaneously with no externally visible intermediate states.
@@
 \begin{enumerate}
-\item In Gotsman/Hobor-style specs, each lock is associated with a lock invariant $R$, which is gained by \texttt{acquire} operations and restored by \texttt{release} operations.
-\item In TaDA-style specs, \texttt{acquire} atomically moves a lock from the unlocked state to the locked state, and \texttt{release} does the reverse.
+\item In Gotsman/Hobor-style specs~\cite{gotsman,oraclesematic}, each lock is associated with a lock invariant $R$, which is gained by \texttt{acquire} operations and restored by \texttt{release} operations.
+\item In TaDA-style specs~\cite{tada}, \texttt{acquire} atomically moves a lock from the unlocked state to the locked state, and \texttt{release} does the reverse.
 \end{enumerate}
@@
 However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise.
-To our knowledge, this is the first formal comparison of the two styles of lock specifications, and our technique may be useful in tools like VST where the older style is deeply integrated into the tool.
+To our knowledge, this is the first formal comparison of the two styles of lock specifications, and our technique should be useful in tools like VST where the older style is deeply integrated into the tool.
 

### 2022-05-15 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
Comments on current draft

@@
 
+\newcommand{\LB}[1]{\textbf{\textcolor{red}{[ LB: #1]}}}
 \newcommand{\TODO}[1]{\textbf{\textcolor{red}{[ TODO: #1]}}}
@@
 In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form $$\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$ where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
-This triple says that $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$.
-The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained.
+This triple says that $P_l$ and $Q_l$ must be true before and after the call\LB{'Must be true': is there a partial-versus-total distinction here, or a reading where the precondition is an assumption (or implies safety) and the postoncdition is a guarantee?}, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$.
+\LB{This suggests I should interpret the leftmost occurrence of 'a' as a binder whose scope extends all the way to $Q_p(a)$.
+So maybe the binding should happen outside the left pair of angle brackets?
+Also, as decorations 'l' and 'p' are not indices but just markers (they don't range over anything) and both 'private' and 'public' start with the letter 'p', maybe replacing $P_l$, $P_p$, $Q_l$, $Q_p$ with say $P$, $Q$, $R$, $S$, or maybe $P$, $P'$, $Q,$, $Q'$ could help.
+Indeed, the concrete patters below suggest something like $P$, $Q$, $P'$, $Q'$, to emphasize that the two predicates to the left of $\mid$ have something in common, and so do the two predicates to the right of $\mid$.} The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained.
 For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$.
@@
 Iris begins with a small core logic including arbitrary higher-order ghost state, and derives features including invariants and logical atomicity from particular instances of ghost state.
-Iris is designed as a language- and logic-independent framework, and recent versions of VST take advantage of this, incorporating Iris-style ghost state in the foundational model and directly using Iris tactics to reason about invariants and atomicity.
+Iris is designed as a language- and logic-independent framework, and recent versions of VST take advantage of this, incorporating Iris-style ghost state in the foundational model and directly using Iris tactics to reason about invariants and atomicity\LB{of actual C implementations}.
 Iris has also been used as the basis for proofs with atomic lock specifications, such as the work of Krishna et al.~\cite{templates}.
@@
 %What else do we need to say?
+\LB{Maybe swap the last 2 sentences?}
 
@@
 
-Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
+\LB{Suppose versus: we have in VST's library, or ``will introduce below''}Suppose we have a sequential implementation of a binary search tree.
+It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
 \begin{mathpar}
@@
 where $\mathsf{BST}\:p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$.
-For instance, $t$ may be a recursively defined tree type in Coq, and $\mathsf{BST}$ would be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, with the root node located at $p$.
+For instance, $t$ may be a recursively defined tree type in Coq\LB{It seems that $t$ is not a Coq type but a value of such a recursivle defined type.
+Maybe give the inductive type here.}, and $\mathsf{BST}$ would be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, with the root node located at $p$.
 
@@
 We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access).
-Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.
+\LB{Here one wonders how the two predicates bst\_ref and bst\_abs are coordinated/related to each other, since their arguments are disjoint.
+Maybe just say that this wil be explained below, or is the magic of ghost state/names/identifiers?}Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.\LB{How does one create the first, empty tree?
+Maybe add a fourth operation/spec above?}
 
@@
 As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}.
-The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
+The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure\LB{add comma here?} $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
 The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.
@@
 
-We demonstrate our approach to proving logically atomic specifications with invariant-based lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
-As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
+We demonstrate our {\LB It's still not clear which one is `` ours'' and why it is prefereed\ldots} approach to proving logically atomic specifications with invariant-based lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
+As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations\LB{As these specs were already given on page 5 we don't need to repeat (maybe put them in a Figure on page 5 and then refer to the figure here?}:
 \begin{mathpar}
@@
 \end{mathpar}
-where insert, lookup, and delete are simple Gallina functions on a functional tree data structure.
+where insert, lookup, and delete are simple Gallina functions on a functional tree data structure\LB{Really?
+I had interpreted them on page 5 as being the C functions, as they were introduced in the sequential Hoare triples for BST}.
 The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree.
@@
 Hand-over-hand locking is a classic example of a pattern that requires reference to lock invariants inside other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}.
-The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock.
+The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write\LB{label this definition as equation (1)} $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock.
 By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.
@@
 This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources.
-In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
+In this case the lock assertion carries a share $\pi$ \LB{$\pi$ a rational(real?) number in the interval\ldots}, where the full share $\pi = 1$ is required to free the lock, while any \LB{nonzero?} share is sufficient to acquire or release it.
 We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n'$'s %awkward
@@
 \end{array}$$
-In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$.
+\LB{At first, I thought that the $\pi$ in $ \ell'' \islock_\pi R(n'')$ should/could be different from the one in $\ell \islock_\pi R(n)$.
+But the point is that its the same $\pi$, and $\pi$ is a parameter to entire setup.
+Maybe that's worth saying?}  the In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$ \LB{perhaps better:  \ldots we had to givegive up our knowledge of $n'$'s lock $\ell'$ in the second proof step (the call to release) here.}.
 This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share.
 Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al.
-instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
+instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.
+\LB{Do you mean some thing like Gotsman uses Locked(l), defined by $Locked(l) = l \mapsto_{\pi_1}R(n)$?}}\LB{i.e by replacing definition (1) by the following (and label the definition, too).
+And again $pi_1$ and $\pi_2$ are parameters, not chosen new each time.)}:
 $$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$
 
-Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
+Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent\LB{wait, locks have parents, too?
+I though just nodes in the tree}.
+Then the same sequence of operations gives us:
 $$\begin{array}{l}
@@
 Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
-The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
+The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible\LB{what does this mean, i.e.
+when/why might it not be possible?}, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
 
@@
 
-To verify our tree implementation, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call.
+To verify our tree implementation\LB{I think here would be a good place to show the C structs and function prototypes of insert etc.
+(but not the function bodies)}, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call.
 In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure.
@@
 
-At the leaves, these ranges partition the space of all possible keys.
+At the leaves\LB{What's the difference between an empty node (eg the two leaves at the bottom of the figure) and a non-existent (leaf) node, ie one that could be the right child of node 40, for range (40,45)?
+Also, one of the two bounds of a range should be exclussive while the other one should be inclusive.
+Otherwise, eg the ranges of the two empty leaf nodes overlap as they both contain 35.
+(More generally, the ranges of any two children of a an inner node at oresent both contain the node's value, which seems to suggest that value could potentiall be found in three places).}, these ranges partition the space of all possible keys.
 If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.
@@
 This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations.
-More precisely, the abstract state of a node in the tree is of type $\mathsf{range} \times \mathsf{option\ (Z \times val \times gname \times gname)}$, where the second component is $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, l, r)$ for an internal node with key $k$, value $v$, and left and right children $l$ and $r$.
-The predicate $\nodeghost{g}{r, n}$ asserts that the node with identifier $g$ has range $r$ and contents $n$, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.
+More precisely, the abstract state of a node in the tree is of \LB{Coq}type $\mathsf{range} \times \mathsf{option\ (Z \times val \times gname \times gname)}$, where the second component is $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, l, r)$ for an internal node with key $k$, value $v$, and left and right children $l$ and $r$.
+\LB{I think it would help to include the Coq definition of tree here and then decorate $t$, $g$, $l$, $r$ ith their Coq type (also, we use Z=key; not sure what the type of values is).
+And $l$ and $r$ are not children (which I'd interpret as trees) but gnames, a concept that hasn't been introduced yet}.The predicate $\nodeghost{g}{r, n}$ asserts that the node with identifier $g$ has range $r$ \LB{Now $r$ ranges over $Z \times Z)$ whereas the preceding sentence used $r$ to refer to a gname/child.
+And we should likely explain why nodes need identifiers in the first place - (the concept of ghost names is a technique/ artifact of Iris but to what extent is it visible at the top level?
+One wonders whether there's a bijecton between C-nodes in memory and gnames or some other data structure that keeps track of such an association, and how intensional the resulting reasoning style gets (invariance under renaming gnames...) )} and contents $n$ \LB{is ``content'' the same as the value range of the model-level KV store/BST?}, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.
 
@@
 Intuitively, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$.
-We formalize this by first defining the collection of per-node ghost state fragments needed to form an abstract tree:
+We formalize this by first defining the collection of per-node ghost state fragments needed to form an abstract tree\LB{node\_ghost has not yet been inyroduced.
+Is it a SL formula (ie mpred)?
+And now it seems $r$ refers to the upper bound of a range pair\ldots}:
 $$\begin{array}{r l}
@@
 This collection of per-node ghost state describes the contents and structure of the tree, but a thread that acquires a node's lock also needs to know that the node it contains actually appears in the tree $t$.
-For this purpose, we add a ghost-state set $\mathsf{ghost\_nodes}(S)$ that records the identifiers of all nodes in the tree, defined so that each node can hold an $\mathsf{in\_tree}\ g$ assertion guaranteeing that $g \in S$.
+For this purpose, we add a ghost-state set $\mathsf{ghost\_nodes}(S)$ that\LB{Names ``ghost\_node'' versus ``node\_ghost'' are not easily kept apart in a reader's midn.
+Also, what's the type of $S$, and of ``ghost\_node'' (perhaps ``list gname'')?}  records the identifiers of all nodes in the tree, defined so that each node can hold an $\mathsf{in\_tree}\ g$ assertion \LB{in\_tree is also undefined here.
+And so are ``implements'' and ``ids'' in the definition of bst\_abs that follows} guaranteeing that $g \in S$.
 Then our final definition of $\treerep$ is:
@@
 Next, we connect the per-node locks of our C implementation with the global abstract state of $\treerep$ by associating each node's lock with a lock invariant assertion $\ell \islock R$, where $R$ includes both the concrete points-to assertion for the node and the corresponding abstract $\mathsf{node\_ghost}$.
-The C implementation of a node is:
+The C implementation of a node is\LB{move this up?}:
 \begin{lstlisting}[language = C,numbers = none]
@@
 where the \lstinline{t} field points to a \lstinline{tree} struct containing the key, value, and child pointers, and the \lstinline{lock} field contains the lock protecting the node.
-The lock invariant for a node at location $p$ needs to describe the contents of the \lstinline{t} field for each possible abstract state:
+The lock invariant for a node at location $p$ needs to describe the contents of the \lstinline{t} field for each possible abstract state\LB{It's dificult to understand this partial defnition.
+What's c?
+Item {\tt t} seems to be a CompCert val.
+And the definition does not link $g_l$ with $p_l$ or $g_r$ with $p_r$, which looks odd (but there's probably a reason for it)}:
 $$\begin{array}{l}
@@
 
-Now we have everything we need to specify the tree operations:
+Now we have everything we need to specify the tree operations\LB{This is the third time these specs are shown\ldots}:
 \begin{mathpar}
@@
 Otherwise, it will eventually reach a leaf node, where it creates two new empty leaf nodes, inserts the key and value into the current node, and links the newly created leaf nodes to the current node.
-The atomic specification for insert method is $$\langle m.\,\nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle$$
+The atomic specification for insert method is\LB{Oh, a fourth copy of this spec, but now bst\_ref and bst\_abs don't carry an index $g$ -- is that significant?}$$\langle m.\,\nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle$$
 The \lstinline{insert} function should atomically add the key-value pair $(k, v)$ to the map implemented by the BST at $p$.
@@
 
-The function has two possible linearization points: after line 23, where we finish adding a node $(k, v)$ to the tree, and after line 41, where we change the value of an existing node with key $k$ to $v$.
+The function has two possible linearization points: after line\LB{In the previous 2 paragraphs, we can perhaps (but maybe not, in fact) get away without showing the code.
+But not here.} 23, where we finish adding a node $(k, v)$ to the tree, and after line 41, where we change the value of an existing node with key $k$ to $v$.
 In both cases, we atomically update the abstract state of the data structure from $\treerep\ m$ to $\treerep (\mathrm{insert}\ m\ k\ v)$, by showing that the update to the tree implementation corresponds to a local change in the abstract tree that implements the mathematical $\mathsf{insert}$.
@@
 We have now demonstrated that invariant-based specs can still be used to prove logically atomic specs for larger data structures (even for fine-grained implementations) and that this requires somewhat more complex proofs than using atomic lock specs (especially for fine-grained implementations).
-The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the approach we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures.
-When possible, however, we would prefer to use atomic specifications for locks to simplify our proofs.
+The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the approach we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures.\LB{The preceding two sentences contain a kwy the motivation of this paper.
+We should mention the question what modifications should/must be applied to VST (and its CompCert-link in Santiago's thesis) explictly, early in Section 1\ldots} When possible, however, we would prefer to use atomic specifications for locks to simplify our proofs.
 Indeed, we intend to investigate shifting the foundations of VST to use atomic operations and derive atomic specs for locks: while updating the soundness proof will be a considerable effort, using the newer style of lock specification will simplify all future proofs about lock-based data structures.
 
-\subsubsection{Acknowledgements} This work benefited greatly from correspondence with Ralf Jung, who pointed out the atomic syncer example and explained the finer points of logical atomicity in Iris.
+\subsubsection{Acknowledgements} This work benefited greatly from correspondence with Ralf Jung, who pointed out the atomic syncer example\LB{I suspect the syncer example is a toy data structure, far simpler than a BST?}  and explained the finer points of logical atomicity in Iris.
 

### 2022-05-15 concurrency/ASL 2022/intro.tex
Comments on current draft

@@
 \end{enumerate}
-Style 1 can be derived from style 2, and is also a natural fit for atomicity proofs of larger data structures, so it has been the style of choice for most recent verifications~\cite{tada-live,templates}.
+Style 1 can be derived from style 2, and is also a natural fit for atomicity proofs of larger data structures, so it has been the style of choice for most recent verifications~\cite{tada-live,templates}\LB{Does TaDaLive really use Gotsman/Hobor specs rather than TaDa-style specs???}.
 However, unpublished work by Zhang~\cite{atomic-syncer} shows that style 1 can also be used to prove atomic specifications for data structures.
 
-In this paper, we adapt this approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t.
+In this paper, we adapt this \LB{This reads like ``we carry out Zhang's approach in VST''} approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t.
 the CompCert compiler~\cite{compcert}, and has recently been extended with support for general ghost state and atomicity proofs.
-We demonstrate that 1) \textbf{our approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs require some additional complications that can be avoided with style 2 lock specs}.
-The key difference is that in style 1 a thread must gain ownership of the lock invariant assertion for a component before acquiring the lock and reading/modifying the component, while in style 2 the locks and components alike can be considered part of a single abstract state that is accessed at each lock access.
+We demonstrate that 1) \textbf{our\LB{again, replace ``our approach'' by style1 or style 2 or whatever} approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs require some additional complications that can be avoided with style 2 lock specs}.
+The key difference is that in style 1 a thread must gain ownership of the lock invariant assertion for a component before acquiring the lock\LB{I don't understand -- above, you say that acquire RESULTS in this assertion so how can we gain it before acquiringng the lock?} and reading/modifying the component, while in style 2 the locks and components alike can be considered part of a single abstract state that is accessed at each lock access.
 However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise.
@@
 %conclusion of our work in \S\ref{conclusion}.
+
+\LB{In contrast to all previous wroks, we our work is on full C implementations of binary search trees rather than pseudo code.
+This is enabled by recently added incorporation of (slightly limited) Iris-style ghost state and invariants (and Iris prof mode), and of TaDa-style atomic specs in VST (citation to William's technical report?)}

### 2022-05-16 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
dealt with some of Lennart's comments

@@
 
+\newcommand{\wm}[1]{\textbf{\textcolor{blue}{[ WM: #1]}}}
 \newcommand{\LB}[1]{\textbf{\textcolor{red}{[ LB: #1]}}}
@@
 In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form $$\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$ where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
-This triple says that $P_l$ and $Q_l$ must be true before and after the call\LB{'Must be true': is there a partial-versus-total distinction here, or a reading where the precondition is an assumption (or implies safety) and the postoncdition is a guarantee?}, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$.
+This triple says that if $P_l$ is true before a call, and the fact that $P_p$ is true for some value of $a$ is held in shared state (e.g.
+a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ up until the \emph{linearization point} of $c$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $c$ ends).
 \LB{This suggests I should interpret the leftmost occurrence of 'a' as a binder whose scope extends all the way to $Q_p(a)$.
-So maybe the binding should happen outside the left pair of angle brackets?
-Also, as decorations 'l' and 'p' are not indices but just markers (they don't range over anything) and both 'private' and 'public' start with the letter 'p', maybe replacing $P_l$, $P_p$, $Q_l$, $Q_p$ with say $P$, $Q$, $R$, $S$, or maybe $P$, $P'$, $Q,$, $Q'$ could help.
+So maybe the binding should happen outside the left pair of angle brackets?}\wm{In the original TaDA paper they did this with a fancy $\forall$ outside that I can't figure out how to replicate.
+In Iris, they use the notation I'm using here, even though the scope looks wrong.} \LB{Also, as decorations 'l' and 'p' are not indices but just markers (they don't range over anything) and both 'private' and 'public' start with the letter 'p', maybe replacing $P_l$, $P_p$, $Q_l$, $Q_p$ with say $P$, $Q$, $R$, $S$, or maybe $P$, $P'$, $Q,$, $Q'$ could help.
 Indeed, the concrete patters below suggest something like $P$, $Q$, $P'$, $Q'$, to emphasize that the two predicates to the left of $\mid$ have something in common, and so do the two predicates to the right of $\mid$.} The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained.
@@
 Iris begins with a small core logic including arbitrary higher-order ghost state, and derives features including invariants and logical atomicity from particular instances of ghost state.
-Iris is designed as a language- and logic-independent framework, and recent versions of VST take advantage of this, incorporating Iris-style ghost state in the foundational model and directly using Iris tactics to reason about invariants and atomicity\LB{of actual C implementations}.
+Iris is designed as a language- and logic-independent framework, and recent versions of VST take advantage of this, incorporating Iris-style ghost state in the foundational model and directly using Iris tactics to reason about invariants and atomicity of C implementations.
+All proofs described in this paper have been formalized in this VST+Iris setting.
 Iris has also been used as the basis for proofs with atomic lock specifications, such as the work of Krishna et al.~\cite{templates}.
-All proofs described in this paper have been formalized in this VST+Iris setting.
 %What else do we need to say?
-\LB{Maybe swap the last 2 sentences?}
 
@@
 
-\LB{Suppose versus: we have in VST's library, or ``will introduce below''}Suppose we have a sequential implementation of a binary search tree.
-It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
+Consider a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
 \begin{mathpar}
@@
 where $\mathsf{BST}\:p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$.
-For instance, $t$ may be a recursively defined tree type in Coq\LB{It seems that $t$ is not a Coq type but a value of such a recursivle defined type.
-Maybe give the inductive type here.}, and $\mathsf{BST}$ would be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, with the root node located at $p$.
-
-A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs:
+For instance, $t$ could be an element of a recursively defined tree type in Coq, and $\mathsf{BST}$ could be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, with the root node located at $p$.
+
+A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs\footnote{We will probably also want functions to create and deallocate trees, but these functions need not be thread-safe, since the tree will not be shared when created and must not be shared when deallocated.}:
 \begin{mathpar}
-\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto v])\rangle
-
-\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land m(k) = v\rangle
-
-\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto \_])\rangle
+\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto v])\rangle
+
+\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m \land m(k) = v\rangle
+
+\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto \_])\rangle
 \end{mathpar}
@@
 We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource.
+They are connected by reference to a fixed but arbitrary ghost state index $g$, which we will generally omit.
 We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access).
-\LB{Here one wonders how the two predicates bst\_ref and bst\_abs are coordinated/related to each other, since their arguments are disjoint.
-Maybe just say that this wil be explained below, or is the magic of ghost state/names/identifiers?}Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.\LB{How does one create the first, empty tree?
-Maybe add a fourth operation/spec above?}
+Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.
 
@@
 As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}.
-The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure\LB{add comma here?} $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
+The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure, $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
 The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.
@@
 
-We demonstrate our {\LB It's still not clear which one is `` ours'' and why it is prefereed\ldots} approach to proving logically atomic specifications with invariant-based lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
+We demonstrate our approach to proving logically atomic specifications with invariant-based (type 1) lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
 As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations\LB{As these specs were already given on page 5 we don't need to repeat (maybe put them in a Figure on page 5 and then refer to the figure here?}:
@@
 \end{mathpar}
-where insert, lookup, and delete are simple Gallina functions on a functional tree data structure\LB{Really?
-I had interpreted them on page 5 as being the C functions, as they were introduced in the sequential Hoare triples for BST}.
 The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree.
@@
 This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources.
-In this case the lock assertion carries a share $\pi$ \LB{$\pi$ a rational(real?) number in the interval\ldots}, where the full share $\pi = 1$ is required to free the lock, while any \LB{nonzero?} share is sufficient to acquire or release it.
+In this case the lock assertion carries a share $\pi$ \LB{$\pi$ a rational(real?) number in the interval\ldots}, where the full share $\pi = 1$ is required to free the lock, while any nonzero share is sufficient to acquire or release it.
 We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n'$'s %awkward
@@
 Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
-The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible\LB{what does this mean, i.e.
-when/why might it not be possible?}, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
+The mechanism for defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as such a mechanism exists, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
 
@@
 \end{forest}
-\caption{A BST with each node labeled with its range}
+\caption{A BST with each node labeled with its range; we omit the empty children of nodes other than 35}
 \label{range_bst}
@@
 
-At the leaves\LB{What's the difference between an empty node (eg the two leaves at the bottom of the figure) and a non-existent (leaf) node, ie one that could be the right child of node 40, for range (40,45)?
-Also, one of the two bounds of a range should be exclussive while the other one should be inclusive.
-Otherwise, eg the ranges of the two empty leaf nodes overlap as they both contain 35.
-(More generally, the ranges of any two children of a an inner node at oresent both contain the node's value, which seems to suggest that value could potentiall be found in three places).}, these ranges partition the space of all possible keys.
+At the leaves, these ranges partition the space of all keys not in the tree.
 If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.
@@
 This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations.
-More precisely, the abstract state of a node in the tree is of \LB{Coq}type $\mathsf{range} \times \mathsf{option\ (Z \times val \times gname \times gname)}$, where the second component is $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, l, r)$ for an internal node with key $k$, value $v$, and left and right children $l$ and $r$.
-\LB{I think it would help to include the Coq definition of tree here and then decorate $t$, $g$, $l$, $r$ ith their Coq type (also, we use Z=key; not sure what the type of values is).
-And $l$ and $r$ are not children (which I'd interpret as trees) but gnames, a concept that hasn't been introduced yet}.The predicate $\nodeghost{g}{r, n}$ asserts that the node with identifier $g$ has range $r$ \LB{Now $r$ ranges over $Z \times Z)$ whereas the preceding sentence used $r$ to refer to a gname/child.
-And we should likely explain why nodes need identifiers in the first place - (the concept of ghost names is a technique/ artifact of Iris but to what extent is it visible at the top level?
-One wonders whether there's a bijecton between C-nodes in memory and gnames or some other data structure that keeps track of such an association, and how intensional the resulting reasoning style gets (invariance under renaming gnames...) )} and contents $n$ \LB{is ``content'' the same as the value range of the model-level KV store/BST?}, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.
+More precisely, we identify each node with a ghost name $g$, and the abstract state of a node in the tree is a pair of its range and its contents, where the contents are $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, g_l, g_r)$ for an internal node with key $k$, value $v$, and left and right children with identifiers $g_l$ and $g_r$.
+The predicate $\nodeghost{g}{(i, j), n}$ asserts that the node with identifier $g$ has range $(i, j)$ \LB{we should likely explain why nodes need identifiers in the first place - (the concept of ghost names is a technique/ artifact of Iris but to what extent is it visible at the top level?
+One wonders whether there's a bijecton between C-nodes in memory and gnames or some other data structure that keeps track of such an association, and how intensional the resulting reasoning style gets (invariance under renaming gnames...) )} and contents $n$, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.
 
@@
 Intuitively, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$.
-We formalize this by first defining the collection of per-node ghost state fragments needed to form an abstract tree\LB{node\_ghost has not yet been inyroduced.
-Is it a SL formula (ie mpred)?
-And now it seems $r$ refers to the upper bound of a range pair\ldots}:
+We formalize this by first defining the collection of per-node ghost state fragments needed to form an abstract tree:
 $$\begin{array}{r l}
-\mathsf{public}&\mathsf{\_halves}(t, g, (l, r)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
-& |\ \mathsf{Leaf} \Rightarrow \publichalf{g}{(l, r), \mathsf{None}}\\
-& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \publichalf{g}{(l, r), \mathsf{Some}\ (k, v, g_l, g_r)}\ * \\&\qquad\mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
+\mathsf{public}&\mathsf{\_halves}(t, g, (i, j)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
+& |\ \mathsf{Leaf} \Rightarrow \publichalf{g}{(i, j), \mathsf{None}}\\
+& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \publichalf{g}{(i, j), \mathsf{Some}\ (k, v, g_l, g_r)}\ * \\&\qquad\mathsf{public\_halves}(t_l, g_l, (i, k)) * \mathsf{public\_halves}(t_r, g_r, (k, j))
 \end{array}$$
-We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{node\_ghost}$ assertions of each node into a single assertion.
-We compute the range of each node's children as described above: if the key in the current node is $k$, then the left child has range $(l, k)$ and the right child has range $(k, r)$.
+We recursively walk through the tree, starting from a node $g$ with range $(i, j)$, and collect the $\mathsf{node\_ghost}$ assertions of each node into a single assertion.
+We compute the range of each node's children as described above: if the key in the current node is $k$, then the left child has range $(i, k)$ and the right child has range $(k, j)$.
 
 This collection of per-node ghost state describes the contents and structure of the tree, but a thread that acquires a node's lock also needs to know that the node it contains actually appears in the tree $t$.
-For this purpose, we add a ghost-state set $\mathsf{ghost\_nodes}(S)$ that\LB{Names ``ghost\_node'' versus ``node\_ghost'' are not easily kept apart in a reader's midn.
-Also, what's the type of $S$, and of ``ghost\_node'' (perhaps ``list gname'')?}  records the identifiers of all nodes in the tree, defined so that each node can hold an $\mathsf{in\_tree}\ g$ assertion \LB{in\_tree is also undefined here.
+For this purpose, we add a ghost assertion $\mathsf{ghost\_nodes}(S)$ that maintains the set $S$ of identifiers of all nodes in the tree, defined so that each node can hold an $\mathsf{in\_tree}\ g$ assertion \LB{in\_tree is also undefined here.
 And so are ``implements'' and ``ids'' in the definition of bst\_abs that follows} guaranteeing that $g \in S$.
 Then our final definition of $\treerep$ is:
-$$\begin{array}{r l}\treerep_g\ m, \triangleq &\exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty))\ * \\&\mathsf{ghost\_nodes}(\mathsf{ids}(t))\end{array}$$
+$$\begin{array}{r l}\treerep_g\ m \triangleq &\exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty))\ * \\&\mathsf{ghost\_nodes}(\mathrm{ids}(t))\end{array}$$
 The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in the ghost tree $t$.
@@
 where the \lstinline{t} field points to a \lstinline{tree} struct containing the key, value, and child pointers, and the \lstinline{lock} field contains the lock protecting the node.
-The lock invariant for a node at location $p$ needs to describe the contents of the \lstinline{t} field for each possible abstract state\LB{It's dificult to understand this partial defnition.
-What's c?
-Item {\tt t} seems to be a CompCert val.
-And the definition does not link $g_l$ with $p_l$ or $g_r$ with $p_r$, which looks odd (but there's probably a reason for it)}:
+The lock invariant for a node at location $p$ needs to map the abstract contents $c$ to the values present in memory in the \lstinline{t} field of $p$:
 $$\begin{array}{l}
 \mathsf{match}\ c\ \mathsf{with}\\
-|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
-|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)
+|\ \mathsf{None} \Rightarrow p.\texttt{t} = \texttt{NULL}\\
+|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ p\texttt{t} \mapsto (k, v, p_l, p_r)
 \end{array}$$
-We then combine this with the hand-over-hand invariant pattern of section~\ref{selflock}, giving us a final lock variant of:
+We then combine this with the hand-over-hand invariant pattern of section~\ref{selflock}, giving us a final lock invariant of:
 $$\begin{array}{r l}
-\mathsf{node\_inv}_g \triangleq \ &\exists l, r, c.\ \myhalf{g}{((l, r), c)} * p.\texttt{lock} \islock_{\pi_1} \mathsf{node\_inv}_g * \mathsf{match}\ c\ \mathsf{with}\\
-&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
-&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)\ * \\&\qquad p_l.\texttt{lock} \islock_{\pi_2} \mathsf{node\_inv}_{g_l} * p_r.\texttt{lock} \islock_{\pi_2} \mathsf{node\_inv}_{g_r}
+\mathsf{node\_inv}_g(p) \triangleq \ &\exists i\ j\ c.\ \myhalf{g}{(i, j), c} * p.\texttt{lock} \islock_{\pi_1} \mathsf{node\_inv}_g * \\&\mathsf{match}\ c\ \mathsf{with}\\
+&|\ \mathsf{None} \Rightarrow p.\texttt{t} = \texttt{NULL}\\
+&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (i, j) \land \exists p_l, p_r.\ p.\texttt{t} \mapsto (k, v, p_l, p_r)\ * \\&\qquad p_l.\texttt{lock} \islock_{\pi_2} \mathsf{node\_inv}_{g_l} * p_r.\texttt{lock} \islock_{\pi_2} \mathsf{node\_inv}_{g_r}
 \end{array}$$
@@
 where $\treerep_g\,m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock for the root node:
-$$\nodeboxrep_g\,b \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \islock \mathsf{node\_inv}_g$$
+$$\nodeboxrep_g\,b \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \islock \mathsf{node\_inv}_g(p)$$
 Now we are prepared to prove that each of our BST functions satisfies its specification.
@@
 We have now demonstrated that invariant-based specs can still be used to prove logically atomic specs for larger data structures (even for fine-grained implementations) and that this requires somewhat more complex proofs than using atomic lock specs (especially for fine-grained implementations).
-The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the approach we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures.\LB{The preceding two sentences contain a kwy the motivation of this paper.
-We should mention the question what modifications should/must be applied to VST (and its CompCert-link in Santiago's thesis) explictly, early in Section 1\ldots} When possible, however, we would prefer to use atomic specifications for locks to simplify our proofs.
+The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the approach we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures.
+When possible, however, we would prefer to use atomic specifications for locks to simplify our proofs.
 Indeed, we intend to investigate shifting the foundations of VST to use atomic operations and derive atomic specs for locks: while updating the soundness proof will be a considerable effort, using the newer style of lock specification will simplify all future proofs about lock-based data structures.
 
-\subsubsection{Acknowledgements} This work benefited greatly from correspondence with Ralf Jung, who pointed out the atomic syncer example\LB{I suspect the syncer example is a toy data structure, far simpler than a BST?}  and explained the finer points of logical atomicity in Iris.
+\subsubsection{Acknowledgements} This work benefited greatly from correspondence with Ralf Jung, who pointed out the atomic syncer example\LB{I suspect the syncer example is a toy data structure, far simpler than a BST?}\wm{It's a generic higher-order function that wraps the argument in a critical section.
+We're specializing it to a particular data structure here.}  and explained the finer points of logical atomicity in Iris.
 

### 2022-05-16 concurrency/ASL 2022/intro.tex
dealt with some of Lennart's comments

--- 2022-05-15 concurrency/ASL 2022/intro.tex
@@
 \end{enumerate}
-Style 1 can be derived from style 2, and is also a natural fit for atomicity proofs of larger data structures, so it has been the style of choice for most recent verifications~\cite{tada-live,templates}.
+Style 2 specs are known to imply style 1 specs, and have been the style of choice for most recent verifications~\cite{tada-live,templates}.
 However, unpublished work by Zhang~\cite{atomic-syncer} shows that style 1 can also be used to prove atomic specifications for data structures.
 
-In this paper, we adapt this approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t.
-the CompCert compiler~\cite{compcert}, and has recently been extended with support for general ghost state and atomicity proofs.
-We demonstrate that 1) \textbf{our approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs require some additional complications that can be avoided with style 2 lock specs}.
-The key difference is that in style 1 a thread must gain ownership of the lock invariant assertion for a component before acquiring the lock and reading/modifying the component, while in style 2 the locks and components alike can be considered part of a single abstract state that is accessed at each lock access.
+In this paper, we adapt this \LB{This reads like ``we carry out Zhang's approach in VST''}\wm{That's basically true.} approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that has recently been extended with support for general ghost state and atomicity proofs.
+The soundness proof for VST~\cite{cpm} lifts the single-threaded correctness theorem of the CompCert compiler~\cite{compcert} to a concurrent setting, and relies explicitly on style 1 lock specs; modifying this proof to rely on atomic operations instead is both theoretically and practically daunting.
+We demonstrate that 1) \textbf{our approach can be used to prove atomic specs for interesting data structures with fine-grained locking, starting from style 1 lock specs}, and 2) \textbf{these proofs require some additional complications that can be avoided with style 2 lock specs}.
+The key difference is that in style 1 a thread must gain ownership of a handle to a component's lock before acquiring the lock and reading/modifying the component, while in style 2 the locks and components alike can be considered part of a single abstract state that is accessed at each lock access.
 However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise.
+
 To our knowledge, this is the first formal comparison of the two styles of lock specifications, and our technique should be useful in tools like VST where the older style is deeply integrated into the tool.
+This is also the first \emph{foundational} verification of a C implementation against logically atomic specifications; prior tools such as VeriFast~\cite{verifast-conc} have been used to verify C implementations, but are not proved sound against a C semantics and so can avoid the complexities of lock specs w.r.t.
+concurrent soundness.
 
@@
 %conclusion of our work in \S\ref{conclusion}.
+
+\LB{In contrast to all previous wroks, we our work is on full C implementations of binary search trees rather than pseudo code.
+This is enabled by recently added incorporation of (slightly limited) Iris-style ghost state and invariants (and Iris prof mode), and of TaDa-style atomic specs in VST (citation to William's technical report?)}\wm{It's currently a rejected ITP paper rather than a technical report, sadly.}

### 2022-05-16 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
edited proofs

@@
 This intuitive condition has been formalized in terms of linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations).
-In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form $$\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$ where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form \[\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle\] where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
 This triple says that if $P_l$ is true before a call, and the fact that $P_p$ is true for some value of $a$ is held in shared state (e.g.
@@
 Indeed, the concrete patters below suggest something like $P$, $Q$, $P'$, $Q'$, to emphasize that the two predicates to the left of $\mid$ have something in common, and so do the two predicates to the right of $\mid$.} The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained.
-For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$.
+For instance, the specification \[\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle\] expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$.
 The $\mathsf{stack}$ itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the $\mathsf{is\_stack}$ assertion.
@@
 Their lock specifications were essentially equivalent, and of the form:
-$$\{ \ell \islock R \}\ \texttt{acquire}(\ell)\ \{ R * \ell \islock R \} \qquad \{ R * \ell \islock R \}\ \texttt{release}(\ell)\ \{ \ell \islock R \}$$
+\[\{ \ell \islock R \}\ \texttt{acquire}(\ell)\ \{ R * \ell \islock R \} \qquad \{ R * \ell \islock R \}\ \texttt{release}(\ell)\ \{ \ell \islock R \}\]
 Each lock $\ell$ is associated with a ``lock invariant'' $R$, an arbitrary separation logic predicate representing the protected resource, which is gained by a thread that acquires the lock and must be restored when the lock is released.
@@
 Later, the TaDA logic~\cite{tada}, which codified the notion of logical atomicity, presented an alternative set of lock specifications:
-$$\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle$$
+\[\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle\]
 An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return.
@@
 Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$.
-For instance, the specification $$\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution.
+For instance, the specification \[\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle\] expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution.
 %maybe something about linking ghost state, which parts are real vs.
@@
 As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}.
-The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure, $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
+The lock specification used in the paper is \[\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle\] where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure, $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
 The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.
@@
 We demonstrate our approach to proving logically atomic specifications with invariant-based (type 1) lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
-As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations\LB{As these specs were already given on page 5 we don't need to repeat (maybe put them in a Figure on page 5 and then refer to the figure here?}:
-\begin{mathpar}
-\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto v])\rangle
-
-\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land m(k) = v\rangle
-
-\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto \_])\rangle
-\end{mathpar}
-The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree.
-In this section, we describe the construction of these predicates, and the proofs of the operations, for a BST implementation based on hand-over-hand locking, using lock-invariant-based specs for our locks.
+As described in section~\ref{atomic}, we aim to define assertions $\nodeboxrep$ (the per-thread handle to the BST) and $\treerep$ (the shared abstract state) such that the operations of the concurrent BST satisfy atomic triples along the lines of \[\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto v])\rangle\] In this section, we describe the construction of the two predicates and the proofs of the operations for a BST implementation with hand-over-hand locking, using lock-invariant-based specs for our locks.
 
@@
 Hand-over-hand locking (also called lock coupling) is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it.
-Hand-over-hand locking is a classic example of a pattern that requires reference to lock invariants inside other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}.
-The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write\LB{label this definition as equation (1)} $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock.
+Hand-over-hand locking is a classic example of a pattern that requires lock invariants that refer to other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}.
+The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write \begin{equation}\label{inv1}R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')\end{equation} where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock.
 By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.
@@
 This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources.
-In this case the lock assertion carries a share $\pi$ \LB{$\pi$ a rational(real?) number in the interval\ldots}, where the full share $\pi = 1$ is required to free the lock, while any nonzero share is sufficient to acquire or release it.
-We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n'$'s %awkward
- lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
-$$\begin{array}{l}
+In this case the lock assertion carries a share $\pi$ (often represented as a rational number in the range $(0, 1]$), where the full share $\pi = 1$ is required to free the lock while any share is sufficient to acquire or release it.
+We can amend our lock invariant to \begin{equation}R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')\end{equation} for some fixed share $\pi$, but then, what happens to the share of $\ell'$ when we release the lock on $n$?
+Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
+\[\begin{array}{l}
 \{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
@@
 \{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
-\end{array}$$
-\LB{At first, I thought that the $\pi$ in $ \ell'' \islock_\pi R(n'')$ should/could be different from the one in $\ell \islock_\pi R(n)$.
-But the point is that its the same $\pi$, and $\pi$ is a parameter to entire setup.
-Maybe that's worth saying?}  the In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$ \LB{perhaps better:  \ldots we had to givegive up our knowledge of $n'$'s lock $\ell'$ in the second proof step (the call to release) here.}.
-This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share.
-Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al.
-instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.
-\LB{Do you mean some thing like Gotsman uses Locked(l), defined by $Locked(l) = l \mapsto_{\pi_1}R(n)$?}}\LB{i.e by replacing definition (1) by the following (and label the definition, too).
-And again $pi_1$ and $\pi_2$ are parameters, not chosen new each time.)}:
-$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$
-
-Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent\LB{wait, locks have parents, too?
-I though just nodes in the tree}.
-Then the same sequence of operations gives us:
-$$\begin{array}{l}
+\end{array}\]
+In the call to \texttt{release}, we give up resources to reestablish $\ell$'s lock invariant, including the lock assertion for $\ell'$.
+As a result, we are left with a share of $\ell$ (the lock for the node we left behind) and no share of $\ell'$ (the lock for the current node).
+
+Both of these problems can be solved by making the lock \emph{recursive}, including a share of $\ell$ in $\ell$'s own lock invariant\footnote{Gotsman et al.
+instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is functionally equivalent to an extra share of the lock assertion.
+\LB{Do you mean some thing like Gotsman uses Locked(l), defined by $Locked(l) = l \mapsto_{\pi_1}R(n)$?}\wm{It's not defined that way---it's a separate assertion---but it could be implemented that way.}}.
+We split ownership of the lock $\ell$ for a node $n$ into two shares: $\pi_1$ is held by $\ell$ itself, and $\pi_2$ is held by the lock of $n$'s parent.
+Then we amend definition \ref{inv1} to
+\[R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')\]
+Now the same sequence of operations gives us:
+\[\begin{array}{l}
 \{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
@@
 \{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
-\end{array}$$
-Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
+\end{array}\]
+When we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
 The mechanism for defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as such a mechanism exists, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
@@
 \subsection{Specification of the hand-over-hand BST}
-%rewrite to use the above: this is the global state made up of components, etc.
-
-To verify our tree implementation\LB{I think here would be a good place to show the C structs and function prototypes of insert etc.
-(but not the function bodies)}, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call.
-In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure.
-As described in section \ref{atomicity}, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$.
+The C implementation of a node in our BST is:
+\begin{lstlisting}[language = C,numbers = none]
+typedef struct tree {int key; void *value;
+                         struct tree_t *left, *right;} tree;
+typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
+\end{lstlisting}
+Each node (of type \lstinline{tree_t}) has a \lstinline{lock} field that holds the lock protecting the node, and a \lstinline{t} field that is either \lstinline{NULL} (for a leaf, which contains no key or value) or points to a \lstinline{tree} struct containing the node's key, value, and child pointers.
+To verify operations on this structure, we need to relate a \lstinline{tree_t} in memory to an abstract state (a mathematical map from keys to values), and divide ownership of those resources between the assertions $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure).
+As described in section \ref{atomicity}, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$, while $\nodeboxrep$ will be a a pointer to the lock of the root node.
 We also need to make sure that our lock invariants support the hand-over-hand pattern.
@@
 The range of a node is inherited from its parent, based on the parent's key: if node $n$ has range $(l, r)$ and key $k$, then its left child has range $(l, k)$ and its right child has range $(k, r)$.
-Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty \emph{leaf} nodes.
+Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty leaf nodes.
 
@@
 \end{forest}
-\caption{A BST with each node labeled with its range; we omit the empty children of nodes other than 35}
+\caption{A BST with each node labeled with its range; empty children of nodes other than 35 are omitted}
 \label{range_bst}
@@
 If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.
-For example, suppose we want to insert a key $38$ into the tree in Figure~\ref{range_bst}.
+For example, suppose we want to insert a key $38$ into the tree in figure~\ref{range_bst}.
 The right child of the node with key 35 is a leaf with range $(35, 40)$, so it is guaranteed to be the right place to insert 38.
@@
 This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations.
-More precisely, we identify each node with a ghost name $g$, and the abstract state of a node in the tree is a pair of its range and its contents, where the contents are $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, g_l, g_r)$ for an internal node with key $k$, value $v$, and left and right children with identifiers $g_l$ and $g_r$.
-The predicate $\nodeghost{g}{(i, j), n}$ asserts that the node with identifier $g$ has range $(i, j)$ \LB{we should likely explain why nodes need identifiers in the first place - (the concept of ghost names is a technique/ artifact of Iris but to what extent is it visible at the top level?
-One wonders whether there's a bijecton between C-nodes in memory and gnames or some other data structure that keeps track of such an association, and how intensional the resulting reasoning style gets (invariance under renaming gnames...) )} and contents $n$, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.
+More precisely, we identify each node with a ghost name $g$  \LB{we should likely explain why nodes need identifiers in the first place - (the concept of ghost names is a technique/ artifact of Iris but to what extent is it visible at the top level?
+One wonders whether there's a bijecton between C-nodes in memory and gnames or some other data structure that keeps track of such an association, and how intensional the resulting reasoning style gets (invariance under renaming gnames...) )}\wm{Not sure we want to get into this, but yes, there should be a bijection between pointers to \lstinline{tree_t}s and a certain subset of ghost names.}, and the abstract state of a node in the tree is a pair of its range and its contents, where the contents are $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, g_l, g_r)$ for an internal node with key $k$, value $v$, and left and right children with identifiers $g_l$ and $g_r$.
+The predicate $\nodeghost{g}{(i, j), n}$ asserts that the node with identifier $g$ has range $(i, j)$ and contents $n$, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.
 
@@
 We formalize this by first defining the collection of per-node ghost state fragments needed to form an abstract tree:
-$$\begin{array}{r l}
+\[\begin{array}{r l}
 \mathsf{public}&\mathsf{\_halves}(t, g, (i, j)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
@@
 & |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \publichalf{g}{(i, j), \mathsf{Some}\ (k, v, g_l, g_r)}\ * \\&\qquad\mathsf{public\_halves}(t_l, g_l, (i, k)) * \mathsf{public\_halves}(t_r, g_r, (k, j))
-\end{array}$$
+\end{array}\]
 We recursively walk through the tree, starting from a node $g$ with range $(i, j)$, and collect the $\mathsf{node\_ghost}$ assertions of each node into a single assertion.
@@
 This collection of per-node ghost state describes the contents and structure of the tree, but a thread that acquires a node's lock also needs to know that the node it contains actually appears in the tree $t$.
-For this purpose, we add a ghost assertion $\mathsf{ghost\_nodes}(S)$ that maintains the set $S$ of identifiers of all nodes in the tree, defined so that each node can hold an $\mathsf{in\_tree}\ g$ assertion \LB{in\_tree is also undefined here.
-And so are ``implements'' and ``ids'' in the definition of bst\_abs that follows} guaranteeing that $g \in S$.
+For this purpose, we define ghost state assertions $\mathsf{ghost\_nodes}(S)$ and $\mathsf{in\_tree}\ g$, where $S$ is a set of identifiers and $g$ is an identifier, such that $\mathsf{in\_tree}\ g * \mathsf{ghost\_nodes}(S)$ implies that $g \in S$.
 Then our final definition of $\treerep$ is:
-$$\begin{array}{r l}\treerep_g\ m \triangleq &\exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty))\ * \\&\mathsf{ghost\_nodes}(\mathrm{ids}(t))\end{array}$$
-The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in the ghost tree $t$.
-Because we quantify over the abstract tree $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node IDs and key-value pairs.
-We will take advantage of this property when verifying the \lstinline{delete} function.
+\[\begin{array}{r l}\treerep_g\ m \triangleq &\exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty))\ * \\&\mathsf{ghost\_nodes}(\mathrm{ids}(t))\end{array}\]
+where $\mathrm{ids}(t)$ is the set of all node identifiers in the abstract tree $t$, and $t$ implements $m$ iff the key-value pairs in $t$ are exactly those in $m$.
+The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in $t$.
+Thanks to the existential quantification over $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node IDs and key-value pairs.
+We will take advantage of this fact when verifying the \lstinline{delete} function.
 
 Next, we connect the per-node locks of our C implementation with the global abstract state of $\treerep$ by associating each node's lock with a lock invariant assertion $\ell \islock R$, where $R$ includes both the concrete points-to assertion for the node and the corresponding abstract $\mathsf{node\_ghost}$.
-The C implementation of a node is\LB{move this up?}:
-\begin{lstlisting}[language = C,numbers = none]
-typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
-\end{lstlisting}
-where the \lstinline{t} field points to a \lstinline{tree} struct containing the key, value, and child pointers, and the \lstinline{lock} field contains the lock protecting the node.
 The lock invariant for a node at location $p$ needs to map the abstract contents $c$ to the values present in memory in the \lstinline{t} field of $p$:
-$$\begin{array}{l}
+\[\begin{array}{l}
 \mathsf{match}\ c\ \mathsf{with}\\
 |\ \mathsf{None} \Rightarrow p.\texttt{t} = \texttt{NULL}\\
-|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ p\texttt{t} \mapsto (k, v, p_l, p_r)
-\end{array}$$
+|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ p.\texttt{t} \mapsto (k, v, p_l, p_r)
+\end{array}\]
 We then combine this with the hand-over-hand invariant pattern of section~\ref{selflock}, giving us a final lock invariant of:
-$$\begin{array}{r l}
-\mathsf{node\_inv}_g(p) \triangleq \ &\exists i\ j\ c.\ \myhalf{g}{(i, j), c} * p.\texttt{lock} \islock_{\pi_1} \mathsf{node\_inv}_g * \\&\mathsf{match}\ c\ \mathsf{with}\\
+\[\begin{array}{r l}
+\mathsf{node\_inv}_g(p) \triangleq \ &\exists i\,j\ c.\ \myhalf{g}{(i, j), c} * p.\texttt{lock} \islock_{\pi_1} \mathsf{node\_inv}_g * \\&\mathsf{match}\ c\ \mathsf{with}\\
 &|\ \mathsf{None} \Rightarrow p.\texttt{t} = \texttt{NULL}\\
 &|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (i, j) \land \exists p_l, p_r.\ p.\texttt{t} \mapsto (k, v, p_l, p_r)\ * \\&\qquad p_l.\texttt{lock} \islock_{\pi_2} \mathsf{node\_inv}_{g_l} * p_r.\texttt{lock} \islock_{\pi_2} \mathsf{node\_inv}_{g_r}
-\end{array}$$
+\end{array}\]
 The hand-over-hand pattern allows us to acquire $p_l.\texttt{lock}$ or $p_r.\texttt{lock}$ and then release $p.\texttt{lock}$ without losing any shares of the lock assertions.
 
-Now we have everything we need to specify the tree operations\LB{This is the third time these specs are shown\ldots}:
+Now we have everything we need to specify the tree operations\LB{This is the second time these specs are shown\ldots}:
 \begin{mathpar}
@@
 where $\treerep_g\,m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock for the root node:
-$$\nodeboxrep_g\,b \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \islock \mathsf{node\_inv}_g(p)$$
+\[\nodeboxrep_g\,b \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \islock \mathsf{node\_inv}_g(p)\]
 Now we are prepared to prove that each of our BST functions satisfies its specification.
@@
 
-\subsubsection{Insert}
-An outline of the C code for the insert function is shown in figure \ref{bst-conc}.
-When a thread inserts a key $k$ into a tree, it first crawls the tree to find the right position for $k$ using hand-over-hand locking.
-If it encounters a node with key $k$, then it simply changes the value at that node.
-Otherwise, it will eventually reach a leaf node, where it creates two new empty leaf nodes, inserts the key and value into the current node, and links the newly created leaf nodes to the current node.
-The atomic specification for insert method is\LB{Oh, a fourth copy of this spec, but now bst\_ref and bst\_abs don't carry an index $g$ -- is that significant?}$$\langle m.\,\nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle$$
-The \lstinline{insert} function should atomically add the key-value pair $(k, v)$ to the map implemented by the BST at $p$.
-
-The main loop of the \lstinline{insert} function traverses the tree to find the node where $k$ should be inserted.
-At the beginning of each iteration, we are at some node $n$ and hold the lock for that node, giving us the lock variant $\mathsf{node\_inv}((l, r), c)$ where $(l, r)$ is the range of keys in the subtree rooted at $n$.
-The loop invariant holds that $k \in (l, r)$, that is, the key $k$ always belongs in the subtree of the current node.
-This guarantees us that if we reach an empty leaf, that leaf is exactly the right place to add a node with key $k$.
-
-The function has two possible linearization points: after line\LB{In the previous 2 paragraphs, we can perhaps (but maybe not, in fact) get away without showing the code.
-But not here.} 23, where we finish adding a node $(k, v)$ to the tree, and after line 41, where we change the value of an existing node with key $k$ to $v$.
-In both cases, we atomically update the abstract state of the data structure from $\treerep\ m$ to $\treerep (\mathrm{insert}\ m\ k\ v)$, by showing that the update to the tree implementation corresponds to a local change in the abstract tree that implements the mathematical $\mathsf{insert}$.
+\begin{figure}[htp]
+\begin{subfigure}[t]{0.6\textwidth}
+\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
+void insert(tree_t** t, int k, void *v) {
+  tree_t *tgt = *t;
+  acquire(tgt->lock);
+  while(1) {
+    tree* p = tgt->t;
+    if (p == NULL) {
+      // set tgt->t to a new node with k and v
+      release(tgt->lock);
+      return;
+    }
+    if (k < p->key){
+      void *l_old = tgt->lock;
+      tgt = p->left;
+      acquire(tgt->lock);
+      release(l_old);
+    } else if (p->key < k){
+      void *l_old = tgt->lock;
+      tgt = p->right;
+      acquire(tgt->lock);
+      release(l_old);
+    } else {
+    	p->value = v;
+        release(tgt->lock);
+    	return;
+    }
+  }
+} 
+\end{lstlisting} 
+\end{subfigure}\qquad
+\begin{subfigure}[t]{0.4\textwidth}
+ \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
+void *lookup(tree_t** t, int k) {
+  tree_t *tgt = *t;
+  acquire(tgt->lock);
+  while(1) {
+    tree *p = tgt->t;
+    if (p == NULL) {
+
+      release(tgt->lock);
+      return NULL;
+    }
+    if (k < p->key){
+      void *l_old = tgt->lock;
+      tgt = p->left;
+      acquire(tgt->lock);
+      release(l_old);
+    } else if (p->key < k){
+      void *l_old = tgt->lock;
+      tgt = p->right;
+      acquire(tgt->lock);
+      release(l_old);
+    } else {
+      void* v = p->value;
+      release(tgt->lock);
+      return v;
+    }
+  }
+}
+\end{lstlisting}
+\end{subfigure}
+\caption{Code outlines for \lstinline{insert} and \lstinline{lookup}}
+\label{bst-conc}
+\end{figure}
+
+The code for the insert and lookup functions is shown in figure \ref{bst-conc}.
+The traversal process is the same in each function: we access the current node's key, compare it to the target key \lstinline{k}, and move to the left child if \lstinline{k} is less than the current key and right if it is greater, using hand-over-hand locking.
+Thus, both loops have the same invariant.
+At the beginning of each iteration, we are at a node at pointer $p$ with ghost identifier $g$ and hold the lock for that node, giving us the lock invariant $\mathsf{node\_inv}_g(p)$ with some range $(i, j)$.
+The loop invariant requires that $\texttt{k} \in (i, j)$, that is, the target key always belongs in the subtree of the current node.
+This guarantees that if we reach an empty leaf, that leaf is the place where key \lstinline{k} should be held.
+The actual logic of each operation happens when we either reach an empty leaf node (line 7) or find a node with \lstinline{k} as its key (line 22); there are also the linearization points of the two functions.
+
+When an insertion encounters a node with key \lstinline{k}, it simply changes the value at that node to \lstinline{value}.
+If it reaches a leaf node, it allocates a new node with key \lstinline{k} and value \lstinline{v}, and inserts it at that position.
+In both cases, we must show that we can atomically update the abstract state of the tree from $\treerep\ m$ to $\treerep (m[\texttt{k}\mapsto \texttt{v}])$, by showing that the update to the tree in memory corresponds to a local change in the abstract tree that inserts the new key-value pair.
 The proofs are split into a proof that the C code implements the local update, and a proof that the local update to a ghost-state tree implements the global update.
-In the first case, the local change is the insertion of $(k, v)$ at an empty leaf whose range includes $k$, and the addition of two new empty leaves as its children; in the second case, we simply change the value of an internal node whose key is already $k$.
-
-\begin{figure}[htb]
+In the first case, the local change is the insertion of $(\texttt{k}, \texttt{v})$ at an empty leaf whose range includes \lstinline{k}, and the addition of two new empty leaves as its children; in the second case, we simply change the value of an internal node whose key is already \lstinline{k}.
+
+\ignore{\begin{figure}[htb]
 \centering
@@
 \label{extract_insert}
-\end{figure}
+\end{figure}}
 
@@
 
-\paragraph{Changing the Global State} describe ``extract lemmas'' here---how much detail?
-
-The proof of \lstinline{lookup} is similar---the linearization points are at the lookup of the target key if it exists in the tree, and at the leaf node whose range contains the target key if it is not present in the tree.
-The \lstinline{lookup} function does not change the structure of the tree, so no particular reasoning on the abstract state is required.
+%\paragraph{Changing the Global State} describe ``extract lemmas'' here---how much detail?
+
+The proof of \lstinline{lookup} is similar, but simpler.
+We must show that if the key is found in the tree, then the same key-value pair exists in the abstract state, and that if the key is not found then it is not present in the abstract state.
+The former follows from the lock invariant $\mathsf{node\_inv}$, which guarantees that the key and value in memory match the key and value in the ghost node; the latter follows from the ranges, since if the abstract tree contains a leaf node with range $(i, j)$ such that $\texttt{k} \in (i, j)$, it cannot contain a node with key \lstinline{k}.
+In either case, the abstract state remains unchanged.
 
@@
 
-The code for the $\mathsf{lookup}$ method is shown in Figure
+The code for the $\mathsf{lookup}$ method is shown in figure
 \ref{lookupproof}. It takes the location of the root pointer and a key
@@
 The atomic specification for $\mathsf{lookup}$ is
-$$\langle m.\,\nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{lookup}\ m\ k = v\rangle$$
+\[\langle m.\,\nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{lookup}\ m\ k = v\rangle\]
 The \texttt{lookup} function should atomically look up the value of the key $k$ in the tree at $p$.
@@
 As before, the key to the verification is the invariant for the main loop:
-$$\begin{array}{l} \mathsf{lookup\_inv}(b, x, \mathit{g\_root}, \mathit{range},\mathit{info}) \triangleq\ \exists\ \mathit{lock},\ \mathit{g\_current},\ \mathit{np},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{\mathit{g\_current}}\\ *\ R\ np\ * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{array}$$
+\[\begin{array}{l} \mathsf{lookup\_inv}(b, x, \mathit{g\_root}, \mathit{range},\mathit{info}) \triangleq\ \exists\ \mathit{lock},\ \mathit{g\_current},\ \mathit{np},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{\mathit{g\_current}}\\ *\ R\ np\ * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{array}\]
 where $P_p$ and $Q_p$ are $\treerep(g\_root, t)$ and $(v =
@@
 \begin{subfigure}[t]{1\textwidth}
- $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
+ \[\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}\]
 \begin{lstlisting}[language = C,  numbers = none]
@@
 
-The code for the \lstinline{delete} method is shown in ??. %pseudocode?
-The tree traversal is the same as in \lstinline{insert}, but the behavior upon finding the target node is more complicated.
-First, we call a helper function \lstinline{pushdown_left} that rearranges the nodes in the tree, maintaining the ordering of nodes but moving the node to be deleted down through the tree until its right child is empty.
-This is accomplished by repeatedly rotating sets of three nodes: the target node's right child is moved up to become its parent, with the target node on the left and the right child's right child moved under the target node, as illustrated in Figure~\ref{delete-diagram}.
-Once the node to be deleted has no right child, it is removed from the tree, and its left child becomes the left child of its parent.
-
-The atomic specification for \lstinline{delete} is:
-$$\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto \_])\rangle$$
-
-The \lstinline{delete} function is responsible for most of the complexity of the abstract state.
-It changes the ranges of nodes in the tree in two ways.
-First, the rotation performed by \lstinline{pushdown_left} changes the ranges of the node to be deleted and its right child (e.g., nodes 40 and 55 in the first step of Figure~\ref{delete-diagram}); this is a local change, as the ranges of all other nodes (even the children of the rearranged nodes) are preserved.
-Second, the actual removal of the node increases the ranges of the subtree below it; for instance, in the last step of Figure~\ref{delete-diagram}, the range of node 35 changes from $(30, 40)$ to $(30, 50)$ when the node is removed.
-The local change occurs while \lstinline{pushdown_left} holds the locks of all nodes involved, so no intermediate states are visible to other threads, but the latter change occurs while holding only the lock of the deleted node, so the algebra for ranges must allow the ranges to increase at any time.
-Fortunately, \lstinline{insert} and \lstinline{lookup} only use the ranges to assert that the target key is in the range of the currently examined node, which is still true if the range increases.
-
-The linearization point of \lstinline{pushdown_left} is the removal of the target node.
-The rotation steps must therefore be considered as not changing the abstract state of the tree.
-This is why it is critical that the abstract state is a map from keys to values, rather than a more explicit representation of the tree's structure---other threads must not rely on particular arrangements of nodes in the tree, since any node that is not currently locked by a thread may be rearranged by another  thread.
-%how much more do we want to say here?
+\begin{figure}[htp]
+\begin{subfigure}[t]{0.6\textwidth}
+\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
+void pushdown_left (tree_t *tgt) {
+  while(1) {
+    tree *p = tgt->t;
+    tree_t *rc = p->right;
+    void *lq = rc->lock;
+    acquire(rc->lock);
+    tree *q = rc->t;
+    if (q == NULL) {
+      tree_t *lc = p->left;
+      // move lc->t to tgt; deallocate p, lc, and rc
+      release(tgt->lock);
+      return;
+    } else {
+      turn_left(tgt, rc);
+      tgt = q->left;
+      release(tgt->lock);
+    }
+  }
+}
+\end{lstlisting} 
+\end{subfigure}\qquad
+\begin{subfigure}[t]{0.4\textwidth}
+ \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
+
+\end{lstlisting}
+\end{subfigure}
+\caption{Code outline for \lstinline{pushdown_left}}
+\label{pushdown-left}
+\end{figure}
+
+The \lstinline{delete} function uses the same traversal logic as \lstinline{insert} and \lstinline{lookup}, but when it finds the node with the target key, it calls a helper function \lstinline{pushdown_left}, outlined in figure~\ref{pushdown-left}.
+This function rearranges the nodes in the tree, moving the node to be deleted down through the tree until its right child is empty while maintaining the binary search property.
+This is accomplished by repeatedly rotating sets of three nodes: the target node's right child \lstinline{rc} is moved up to become its parent, with the target node on the left and the right child's right child moved under the target node, as illustrated in figure~\ref{delete-diagram}.
+Once the node to be deleted has no right child, its contents are replaced with those of its left child, and its children are removed from the tree.
+
+The \lstinline{pushdown_left} function changes the structure of the tree in two ways.
+First, the rotation performed by \lstinline{turn_left} (line 14) changes the ranges of the node to be deleted and its right child (e.g., nodes 40 and 55 in the first step of figure~\ref{delete-diagram}); this is a local change, as the ranges of all other nodes (even the children of the rearranged nodes) are preserved.
+This is not a linearization point, so the rotation must be considered not to change the abstract state of the tree.
+We accounted for this by quantifying over the abstract tree $t$ in the definition of $\treerep$; rearranging nodes without changing the key-value map still implements the same abstract state, and other threads cannot assume that the tree structure they have observed is still the current structure of the tree.
+Second, the final removal of the node (line 10) increases the ranges of the subtree below it; for example, in the last step of figure~\ref{delete-diagram}, the range of node 35 changes from $(30, 40)$ to $(30, 50)$ when the node is removed.
+This is the linearization point of the function.
+The change occurs while \lstinline{pushdown_left} holds only the lock of the deleted node, so the algebra for range ghost state must allow a node's range to increase at any time, even when the node's lock is held.
+Fortunately, \lstinline{insert} and \lstinline{lookup} only use ranges to assert that the target key is in the range of the current node, which is still true if the range increases.
 
@@
 
-\section{Using the Specifications}
-client and client proofs
+\ignore{\section{Using the Specifications}
+}
 

### 2022-05-17 concurrency/bst_client.c
Client proof upload

@@ -1,0 +1,72 @@
+#include "./bst_conc.h"
+
+int a, b, c, d;
+
+void insert_int(treebox t, int key, int value){
+   void* p_value = surely_malloc(sizeof(int));
+   *((int*)p_value) = value;
+   insert(t, key, (void*) &value);
+}
+
+void* update_tree(void* t){
+  a = 1;
+  b = 2;
+  c = 3; 
+  d = 4;
+  // insert_int(t, 2, 20);
+  // insert_int(t, 1, 1);
+  // insert_int(t, 1, 100);
+  // insert_int(t, 2, 1);
+  // int m = 1;
+  // int n = 2;
+  // int k = 2;
+  // int j = 3;
+  // void* pm = &m;
+  // void* pn = &n;
+  // void* pk = &k;
+  // void* pj = &k;
+  insert(t, 1, &a);
+  insert(t, 2, &b);
+  insert(t, 1, &c);
+  insert(t, 2, &d);
+  return (void*)NULL;
+}
+
+int retrieve_value(treebox t){
+  void* searchResult;
+  int y = -1;
+  do {
+     searchResult = lookup (t,  2);
+     if (searchResult != NULL) {
+        y = *((int *) searchResult);
+        //printf("Key 1 returns %d\n", y);
+     }
+  } while (y != 4);
+   searchResult = lookup (t,  1);
+   return *((int *) searchResult);
+}
+
+
+
+int main(){
+  treebox t;
+  t = treebox_new();  
+
+   
+ int* result;
+
+
+  //  makelock((void*)t_lock);
+  // /* Spawn */
+  spawn((void *)&update_tree, t);
+
+  //printf("Key 1 returns %d\n", retrieve_value(t));
+  int retVal =  retrieve_value(t);
+  
+  // /*JOIN */
+  //  acquire((void*)t_lock);  
+  //  freelock2((void*)t_lock);
+  
+  treebox_free(t);
+  return 0;
+}

### 2022-05-17 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
Added and edited Anastasiia's notes on the client proof

@@
 \[\nodeboxrep_g\,b \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \islock \mathsf{node\_inv}_g(p)\]
-Now we are prepared to prove that each of our BST functions satisfies its specification.
+Next, we will prove that each of our BST functions satisfies its specification.
 
@@
 
-\ignore{\section{Using the Specifications}
-}
+\section{Using the Specifications}
+
+\begin{figure}[htp]
+\begin{subfigure}[t]{0.4\textwidth}
+\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
+void* update_tree(void* t){
+  a = 1;
+  b = 2;
+  c = 3; 
+  d = 4;
+  insert(t, 1, &a);
+  insert(t, 2, &b);
+  insert(t, 1, &c);
+  insert(t, 2, &d);
+  return NULL;
+}
+\end{lstlisting} 
+\end{subfigure}\qquad
+\begin{subfigure}[t]{0.5\textwidth}
+ \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
+int retrieve_value(treebox t){
+  void* searchResult;
+  int y = -1;
+  do {
+     searchResult = lookup (t,  2);
+     if (searchResult != NULL) {
+        y = *((int *) searchResult);
+     }
+  } while (y != 4);
+   searchResult = lookup (t,  1);
+   return *((int *) searchResult);
+}
+\end{lstlisting}
+\end{subfigure}
+\caption{Client code}
+\label{bst-conc-client}
+\end{figure}
+
+To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}.
+A producer thread runs the \lstinline{update_tree} function to populate a tree, while a consumer thread runs \lstinline{retrieve_value} to wait until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1.
+This is a data-structure-level version of the message-passing idiom, and demonstrates one of the advantages of atomic specifications: we can use them to synchronize threads and pass ownership of resources via data structure operations.
+The invariant for the BST \lstinline{t} is a simple state machine with three states, depending on the contents of the abstract map $m$: 1) $m(2) = \texttt{NULL}$, 2) $m(2) \mapsto i$ for some integer $i \neq 4$; and 3) $m(2) \mapsto 4 \land m(1) \mapsto 3$.
+In this final state, the invariant also holds the resources to be transferred from the producer to the consumer.
+We also include ghost state to reflect the fact that there is exactly one producer thread (so the producer always knows exactly which values are in the tree) and one consumer thread (so it can retrieve the resources without worrying about interference).
+The proof of the client uses only the atomic specifications of the BST operations, and is not affected by the style of lock specification.
+
+%The consumer thread repeatedly tries to read the value stored at the key of 2 to see if it is equal to 4.
+Once 4 can be observed, it returns the value stored with the key of 1.
+Since all the inserts by the producer thread are performed in a specific order, the consumer thread should always return 3.
+Since the fact that, once 4 is returned by the lookup of 2, 3 is already stored with the key of 1 was a part of the invariant, the only thing that this part of the proof required was showing that, once the consumer breaks out of the loop, the tree has already reached its final state.
+Remembering this fact between the two lookup calls – the one that happens in the loop and the one that happens right after – required the addition of the ghost snap, another special ghost variable which could only change its value once.
+Consumer also owns a unique $\mathsf{reader\_token}$, which is used to replace the resources in the invariant once they can be retrieved.
+Because this token is unique, as long as the consumer thread owns it, the fact that the resources have not been replaced with it yet can be proven.
 

### 2022-05-17 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
client proof edits

@@
 \end{subfigure}\qquad
-\begin{subfigure}[t]{0.5\textwidth}
+\begin{subfigure}[t]{0.55\textwidth}
  \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
 int retrieve_value(treebox t){
-  void* searchResult;
   int y = -1;
@@
   } while (y != 4);
-   searchResult = lookup (t,  1);
+  void* searchResult = lookup (t,  1);
   return *((int *) searchResult);
@@
 To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}.
-A producer thread runs the \lstinline{update_tree} function to populate a tree, while a consumer thread runs \lstinline{retrieve_value} to wait until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1.
+A producer thread running the \lstinline{update_tree} function populates a tree, while a consumer thread running \lstinline{retrieve_value} waits until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1.
 This is a data-structure-level version of the message-passing idiom, and demonstrates one of the advantages of atomic specifications: we can use them to synchronize threads and pass ownership of resources via data structure operations.

### 2022-05-18 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
code tweaks, check in bib

@@
 \author{Roshan Sharma\inst{1} \and
-Shengyi Wang\inst{2}\orcidID{?} \and
+Shengyi Wang\inst{2}\orcidID{0000-0002-2286-8703} \and
 Alexander Oey\inst{3} \and
@@
 This intuitive condition has been formalized in terms of linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations).
-In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form \[\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle\] where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form \[\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle\] where $P_l$ and $Q_l$ are \emph{local} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
 This triple says that if $P_l$ is true before a call, and the fact that $P_p$ is true for some value of $a$ is held in shared state (e.g.
@@
 
-\begin{figure}[htp]
+\begin{figure}[ht]
 \begin{subfigure}[t]{0.6\textwidth}
@@
     if (p == NULL) {
-      // set tgt->t to a new node with k and v
+      ... // set tgt->t to a new node with k, v
       release(tgt->lock);
@@
     } else if (p->key < k){
-      void *l_old = tgt->lock;
-      tgt = p->right;
-      acquire(tgt->lock);
-      release(l_old);
+      ... // move to p->right instead
     } else {
@@
       tree_t *lc = p->left;
-      // move lc->t to tgt; deallocate p, lc, and rc
+      ... // move lc->t to tgt; deallocate p, lc, and rc
       release(tgt->lock);

### 2022-05-18 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
edits and space hacks

@@
 Anastasiia Evdokimova\inst{4} \and
-Lennart Beringer\inst{2}\orcidID{?} \and
+Lennart Beringer\inst{2}\orcidID{0000-0002-1570-3492} \and
 William Mansky\inst{4}\orcidID{0000-0002-5351-895X}} %confirm author order
@@
 \subsection{Logically Atomic Specifications}
-\label{atomic}
+\label{atomic-triples}
 
@@
 a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ up until the \emph{linearization point} of $c$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $c$ ends).
-\LB{This suggests I should interpret the leftmost occurrence of 'a' as a binder whose scope extends all the way to $Q_p(a)$.
+%\LB{This suggests I should interpret the leftmost occurrence of 'a' as a binder whose scope extends all the way to $Q_p(a)$.
 So maybe the binding should happen outside the left pair of angle brackets?}\wm{In the original TaDA paper they did this with a fancy $\forall$ outside that I can't figure out how to replicate.
 In Iris, they use the notation I'm using here, even though the scope looks wrong.} \LB{Also, as decorations 'l' and 'p' are not indices but just markers (they don't range over anything) and both 'private' and 'public' start with the letter 'p', maybe replacing $P_l$, $P_p$, $Q_l$, $Q_p$ with say $P$, $Q$, $R$, $S$, or maybe $P$, $P'$, $Q,$, $Q'$ could help.
-Indeed, the concrete patters below suggest something like $P$, $Q$, $P'$, $Q'$, to emphasize that the two predicates to the left of $\mid$ have something in common, and so do the two predicates to the right of $\mid$.} The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained.
+Indeed, the concrete patters below suggest something like $P$, $Q$, $P'$, $Q'$, to emphasize that the two predicates to the left of $\mid$ have something in common, and so do the two predicates to the right of $\mid$.}
+The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained.
 For instance, the specification \[\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle\] expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$.
@@
 This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations).
-In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{local} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
 Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$.
@@
 
-A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs\footnote{We will probably also want functions to create and deallocate trees, but these functions need not be thread-safe, since the tree will not be shared when created and must not be shared when deallocated.}:
+A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs\footnote{We will also have functions to create and deallocate trees, but these functions need not be thread-safe, since the tree will not be shared when created and must not be shared when deallocated.}:
 \begin{mathpar}
@@
 We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource.
-They are connected by reference to a fixed but arbitrary ghost state index $g$, which we will generally omit.
+They are connected by reference to a fixed but arbitrary ghost state identifier $g$, which we will generally omit.
 We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access).
@@
 As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}.
-The lock specification used in the paper is \[\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle\] where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure, $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
+The lock specification used\footnote{In fact, this specification is derived from the basic TaDA-style atomic lock specificaton.} is \[\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle\] where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure, $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state).
 The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.
 
-Note that in this case, the private precondition $\mathsf{inFP}(n)$ is merely the knowledge that $n$ is in the data structure, and does not involve any ownership---in particular, a caller does not need a share of the lock for the root node.
+Note that in this case, the local precondition $\mathsf{inFP}(n)$ is merely the knowledge that $n$ is in the data structure, and does not involve any ownership---in particular, a caller does not need a share of the lock for the root node.
 A thread can acquire the lock for any node $n$ in the data structure by atomically accessing $\mathsf{CSS}$, extracting the sub-assertion corresponding to $n$, and switching it to the locked state.
@@
 Both of these problems can be solved by making the lock \emph{recursive}, including a share of $\ell$ in $\ell$'s own lock invariant\footnote{Gotsman et al.
-instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is functionally equivalent to an extra share of the lock assertion.
-\LB{Do you mean some thing like Gotsman uses Locked(l), defined by $Locked(l) = l \mapsto_{\pi_1}R(n)$?}\wm{It's not defined that way---it's a separate assertion---but it could be implemented that way.}}.
+instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is functionally equivalent to an extra share of the lock assertion.}.
 We split ownership of the lock $\ell$ for a node $n$ into two shares: $\pi_1$ is held by $\ell$ itself, and $\pi_2$ is held by the lock of $n$'s parent.
@@
 \texttt{acquire}(\ell');\\
-\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
+\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')\ * \\\ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
 \texttt{release}(\ell);\\
@@
 This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations.
-More precisely, we identify each node with a ghost name $g$  \LB{we should likely explain why nodes need identifiers in the first place - (the concept of ghost names is a technique/ artifact of Iris but to what extent is it visible at the top level?
-One wonders whether there's a bijecton between C-nodes in memory and gnames or some other data structure that keeps track of such an association, and how intensional the resulting reasoning style gets (invariance under renaming gnames...) )}\wm{Not sure we want to get into this, but yes, there should be a bijection between pointers to \lstinline{tree_t}s and a certain subset of ghost names.}, and the abstract state of a node in the tree is a pair of its range and its contents, where the contents are $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, g_l, g_r)$ for an internal node with key $k$, value $v$, and left and right children with identifiers $g_l$ and $g_r$.
+More precisely, each node's ghost state, identified by a ghost name $g$, contains a pair of its range and its contents, where the contents are $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, g_l, g_r)$ for an internal node with key $k$, value $v$, and left and right children with identifiers $g_l$ and $g_r$.
 The predicate $\nodeghost{g}{(i, j), n}$ asserts that the node with identifier $g$ has range $(i, j)$ and contents $n$, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.
@@
 
-This collection of per-node ghost state describes the contents and structure of the tree, but a thread that acquires a node's lock also needs to know that the node it contains actually appears in the tree $t$.
+This collection of per-node ghost state describes the contents and structure of the tree, but a thread that acquires a node's lock also needs to know that the node actually appears in the tree $t$.
 For this purpose, we define ghost state assertions $\mathsf{ghost\_nodes}(S)$ and $\mathsf{in\_tree}\ g$, where $S$ is a set of identifiers and $g$ is an identifier, such that $\mathsf{in\_tree}\ g * \mathsf{ghost\_nodes}(S)$ implies that $g \in S$.
-Then our final definition of $\treerep$ is:
+Then the full definition of $\treerep$ is:
 \[\begin{array}{r l}\treerep_g\ m \triangleq &\exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty))\ * \\&\mathsf{ghost\_nodes}(\mathrm{ids}(t))\end{array}\]
@@
 
-Now we have everything we need to specify the tree operations\LB{This is the second time these specs are shown\ldots}:
+Now we can fill in the pieces of the specifications outlined in section \ref{atomic}:
 \begin{mathpar}
-\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{insert}\ m\ k\ v)\rangle
-
-\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep_g\,p\ |\ \treerep_g\,m \land \mathrm{lookup}\,m\ k = v\rangle
-
-\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{delete}\ m\ k)\rangle
+\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(m[k \mapsto v])\rangle
+
+\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep_g\,p\ |\ \treerep_g\,m \land m(k) = v\rangle
+
+\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(m[k\mapsto \_])\rangle
 \end{mathpar}
@@
 \begin{figure}[ht]
-\begin{subfigure}[t]{0.6\textwidth}
+\begin{subfigure}[t]{0.52\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
-void insert(tree_t** t, int k, void *v) {
+void insert(tree_t** t, int k, void *v){
   tree_t *tgt = *t;
@@
     if (p == NULL) {
-      ... // set tgt->t to a new node with k, v
+      ... // make a new node with k, v
       release(tgt->lock);
@@
  \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
-void *lookup(tree_t** t, int k) {
+void *lookup(tree_t** t, int k){
   tree_t *tgt = *t;
@@
 This guarantees that if we reach an empty leaf, that leaf is the place where key \lstinline{k} should be held.
-The actual logic of each operation happens when we either reach an empty leaf node (line 7) or find a node with \lstinline{k} as its key (line 22); there are also the linearization points of the two functions.
+The actual logic of each operation happens when we either reach an empty leaf node (line 7) or find a node with \lstinline{k} as its key (line 19); there are also the linearization points of the two functions.
 
@@
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
-void pushdown_left (tree_t *tgt) {
+void pushdown_left (tree_t *tgt){
   while(1) {
@@
 
-\subsubsection{Acknowledgements} This work benefited greatly from correspondence with Ralf Jung, who pointed out the atomic syncer example\LB{I suspect the syncer example is a toy data structure, far simpler than a BST?}\wm{It's a generic higher-order function that wraps the argument in a critical section.
-We're specializing it to a particular data structure here.}  and explained the finer points of logical atomicity in Iris.
+\subsubsection{Acknowledgements} This work benefited greatly from correspondence with Ralf Jung, who pointed out Zhang's syncer approach and explained the finer points of logical atomicity in Iris.
 

### 2022-05-18 concurrency/ASL 2022/intro.tex
edits and space hacks

@@
 
-In this paper, we adapt this \LB{This reads like ``we carry out Zhang's approach in VST''}\wm{That's basically true.} approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that has recently been extended with support for general ghost state and atomicity proofs.
+In this paper, we adapt this approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that has recently been extended with support for general ghost state and atomicity proofs.
 The soundness proof for VST~\cite{cpm} lifts the single-threaded correctness theorem of the CompCert compiler~\cite{compcert} to a concurrent setting, and relies explicitly on style 1 lock specs; modifying this proof to rely on atomic operations instead is both theoretically and practically daunting.
@@
 %conclusion of our work in \S\ref{conclusion}.
-
-\LB{In contrast to all previous wroks, we our work is on full C implementations of binary search trees rather than pseudo code.
-This is enabled by recently added incorporation of (slightly limited) Iris-style ghost state and invariants (and Iris prof mode), and of TaDa-style atomic specs in VST (citation to William's technical report?)}\wm{It's currently a rejected ITP paper rather than a technical report, sadly.}

### 2022-05-18 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
Some fiddling to make the paper fit into 15 pages

@@
 A release simply moves atomically from the locked state to the unlocked state.
-These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs.
+These operations make no mention of a protected resource \LB{maybe add this:\emph{, i.e.~they obey the pattern of atomic specs for $P_l = Q_l = True$,} ?}, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs.
 It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g.
@@
 
-Our proof technique is developed in a combination of two Coq-based separation logic verification systems, the Verified Software Toolchain (VST) and Iris.
+We employ a combination of two Coq-based verification systems, the Verified Software Toolchain (VST) and Iris.
 VST~\cite{plfcc} is a Coq-based system for proving separation logic specifications of C programs.
@@
 
-A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs\footnote{We will also have functions to create and deallocate trees, but these functions need not be thread-safe, since the tree will not be shared when created and must not be shared when deallocated.}:
+A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs\footnote{We also have functions to create and deallocate trees, but these functions need not be thread-safe, since the tree will not be shared when created and must not be shared when deallocated.}:
 \begin{mathpar}
@@
 \subsection{Atomicity with lock invariants}
-The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end.
-These \emph{coarse-grained} locking implementations are easy to verify with invariant-style lock specs: the lock's invariant can contain the whole data structure assertion (e.g.
-$\mathsf{BST}\:p\ t$), so that whichever thread is currently executing an operation on the tree at $p$ has full ownership of the data structure.
-If we only want to prove that the operation is thread-safe, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock.
-On the other hand, if we want to prove specifications that describe changes to the data structure---as in the atomic specification $\langle m.\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \treerep\ (m[k \mapsto v])\rangle$---we need ghost state to track the changes.
+The easiest way to synchronize a data structure is to add a lock that
+each operation acquires at the start and releases at the end. These
+\emph{coarse-grained} locking implementations are easy to verify with
+invariant-style lock specs: the lock's invariant can contain the whole
+data structure assertion (e.g. $\mathsf{BST}\:p\ t$), so 
+%that whichever thread is currently executing an operation on the tree at
+%$p$ has full ownership of the data structure. 
+a thread always has full ownership of the data structure when
+performing an operation on $p$.  If we only want to prove that the operation
+is thread-safe, we might choose the lock invariant $\exists t.\
+\mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree
+in some state $t$ every time we acquire the lock. On the other hand,
+if we want to prove specifications that describe changes to the data
+structure---as in the atomic specification $\langle m.\ \treerep\
+m\rangle\ \texttt{insert}(p, k, v)\ \langle \treerep\ (m[k \mapsto
+v])\rangle$---we need ghost state to track the changes.
 
@@
 
-When an insertion encounters a node with key \lstinline{k}, it simply changes the value at that node to \lstinline{value}.
-If it reaches a leaf node, it allocates a new node with key \lstinline{k} and value \lstinline{v}, and inserts it at that position.
+When insertion encounters a node with key \lstinline{k}, it simply changes that node's value to \lstinline{value}.
+If it reaches a leaf, it allocates a new node with key \lstinline{k} and value \lstinline{v}, and inserts it at that position.
 In both cases, we must show that we can atomically update the abstract state of the tree from $\treerep\ m$ to $\treerep (m[\texttt{k}\mapsto \texttt{v}])$, by showing that the update to the tree in memory corresponds to a local change in the abstract tree that inserts the new key-value pair.
@@
 
-\begin{figure}[htp]
+\begin{figure}[t]%[htp]
 \begin{subfigure}[t]{0.6\textwidth}
@@
 
-The \lstinline{pushdown_left} function changes the structure of the tree in two ways.
-First, the rotation performed by \lstinline{turn_left} (line 14) changes the ranges of the node to be deleted and its right child (e.g., nodes 40 and 55 in the first step of figure~\ref{delete-diagram}); this is a local change, as the ranges of all other nodes (even the children of the rearranged nodes) are preserved.
+Function \lstinline{pushdown_left} changes the structure of the tree in two ways.
+First, the rotation \lstinline{turn_left} (line 14) changes the ranges of the node to be deleted and its right child (e.g., nodes 40 and 55 in the first step of figure~\ref{delete-diagram}); this is a local change, as the ranges of all other nodes (even the children of the rearranged nodes) are preserved.
 This is not a linearization point, so the rotation must be considered not to change the abstract state of the tree.
@@
 
-\begin{figure}[htb]
+\begin{figure}[t]%[htb]
 \centering
@@
 \section{Using the Specifications}
-
-\begin{figure}[htp]
+\begin{figure}[t]%[htp]
 \begin{subfigure}[t]{0.4\textwidth}
@@
 \end{figure}
-
 To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}.
@@
 
-VeriFast~\cite{verifast}, a separation-logic-based verifier for C and Java programs, also supports lock-based and atomic concurrency~\cite{verifast-conc}, and has been used to verify a hand-over-hand-locking linked list similar to that of Krishna et al.
-The specifications of both the lock operations and the list itself use a precursor of TaDA's logical atomicity.
-VeriFast is not a foundationally verified system, but its basic logic is verified against the semantics of a toy language in Coq.
+VeriFast~\cite{verifast}, a separation-logic-based verifier for C and
+Java, supports lock-based and atomic
+concurrency~\cite{verifast-conc}, and has been used to verify a
+hand-over-hand-locking linked list similar to that of Krishna et al.
+The specifications of the lock operations and the list itself use
+a precursor of TaDA's logical atomicity. VeriFast is not 
+foundational, but its basic logic is verified
+against the semantics of a toy language in Coq.
 

### 2022-05-18 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
Adjust the layout of one formula

--- 2022-05-18 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
@@
 \[\begin{array}{r l}
-\mathsf{public}&\mathsf{\_halves}(t, g, (i, j)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
-& |\ \mathsf{Leaf} \Rightarrow \publichalf{g}{(i, j), \mathsf{None}}\\
-& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \publichalf{g}{(i, j), \mathsf{Some}\ (k, v, g_l, g_r)}\ * \\&\qquad\mathsf{public\_halves}(t_l, g_l, (i, k)) * \mathsf{public\_halves}(t_r, g_r, (k, j))
+& \mathsf{public\_halves}(t, g, (i, j)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
+& \;\qquad |\ \mathsf{Leaf} \Rightarrow \publichalf{g}{(i, j), \mathsf{None}}\\
+& \;\qquad |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \publichalf{g}{(i, j), \mathsf{Some}\ (k, v, g_l, g_r)}\ * \\
+& \;\qquad \qquad\mathsf{public\_halves}(t_l, g_l, (i, k)) * \mathsf{public\_halves}(t_r, g_r, (k, j))
 \end{array}\]

### 2022-05-18 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
slight edits

@@
 A release simply moves atomically from the locked state to the unlocked state.
-These operations make no mention of a protected resource \LB{maybe add this:\emph{, i.e.~they obey the pattern of atomic specs for $P_l = Q_l = True$,} ?}, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs.
+These operations do not specify what is protected by the lock, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs.
 It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g.
@@
 %$p$ has full ownership of the data structure. 
-a thread always has full ownership of the data structure when
-performing an operation on $p$.  If we only want to prove that the operation
+that a thread always has full ownership of the data structure when
+performing an operation.  If we only want to prove that the operation
 is thread-safe, we might choose the lock invariant $\exists t.\
@@
 
-Function \lstinline{pushdown_left} changes the structure of the tree in two ways.
+The \lstinline{pushdown_left} function changes the structure of the tree in two ways.
 First, the rotation \lstinline{turn_left} (line 14) changes the ranges of the node to be deleted and its right child (e.g., nodes 40 and 55 in the first step of figure~\ref{delete-diagram}); this is a local change, as the ranges of all other nodes (even the children of the rearranged nodes) are preserved.
@@
 \section{Using the Specifications}
-\begin{figure}[t]%[htp]
+To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}.
+A producer thread running the \lstinline{update_tree} function populates a tree, while a consumer thread running \lstinline{retrieve_value} waits until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1.
+This is a data-structure-level version of the message-passing idiom, and demonstrates one of the advantages of atomic specifications: we can use them to synchronize threads and pass ownership of resources via data structure operations.
+The invariant for the BST \lstinline{t} is a simple state machine with three states, depending on the contents of the abstract map $m$: 1) $m(2) = \texttt{NULL}$, 2) $m(2) \mapsto i$ for some integer $i \neq 4$; and 3) $m(2) \mapsto 4 \land m(1) \mapsto 3$.
+In this final state, the invariant also holds the resources to be transferred from the producer to the consumer.
+We also include ghost state to reflect the fact that there is exactly one producer thread (so the producer always knows exactly which values are in the tree) and one consumer thread (so it can retrieve the resources without worrying about interference).
+The proof of the client uses only the atomic specifications of the BST operations, and is not affected by the style of lock specification.
+
+\begin{figure}[htb]
 \begin{subfigure}[t]{0.4\textwidth}
@@
 \end{figure}
-To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}.
-A producer thread running the \lstinline{update_tree} function populates a tree, while a consumer thread running \lstinline{retrieve_value} waits until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1.
-This is a data-structure-level version of the message-passing idiom, and demonstrates one of the advantages of atomic specifications: we can use them to synchronize threads and pass ownership of resources via data structure operations.
-The invariant for the BST \lstinline{t} is a simple state machine with three states, depending on the contents of the abstract map $m$: 1) $m(2) = \texttt{NULL}$, 2) $m(2) \mapsto i$ for some integer $i \neq 4$; and 3) $m(2) \mapsto 4 \land m(1) \mapsto 3$.
-In this final state, the invariant also holds the resources to be transferred from the producer to the consumer.
-We also include ghost state to reflect the fact that there is exactly one producer thread (so the producer always knows exactly which values are in the tree) and one consumer thread (so it can retrieve the resources without worrying about interference).
-The proof of the client uses only the atomic specifications of the BST operations, and is not affected by the style of lock specification.
 

### 2022-06-28 concurrency/bst_template.c
Checking in Than's template proofs

@@ -1,0 +1,316 @@
+//
+//  main.c
+//  BST
+//
+//  Created by Duc Than Nguyen on 1/2/22.
+//
+
+
+#include <stddef.h>
+#include <stdio.h>
+/* #include <stdlib.h> */
+#include "threads.h"
+
+extern void *malloc (size_t n);
+extern void free(void *p);
+extern void exit(int code);
+
+typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
+typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
+
+typedef struct tree_t **treebox;
+lock_t thread_lock;
+treebox tb;
+
+
+void *surely_malloc (size_t n) {
+    void *p = malloc(n);
+    if (!p) exit(1);
+    return p;
+}
+
+treebox treebox_new(void) {
+    treebox p = (treebox) surely_malloc(sizeof (*p));
+    tree_t *newt = (tree_t *) surely_malloc(sizeof(tree_t));
+    *p = newt;
+    lock_t *l = (lock_t *) surely_malloc(sizeof(lock_t));
+    makelock(l);
+    newt->lock = l;
+    newt->t = NULL;
+    release2(l);
+    return p;
+}
+
+
+void tree_free(struct tree_t *tgp) {
+    struct tree_t *pa, *pb;
+    tree* p;
+    void *l = tgp->lock;
+    acquire(l);
+    p = tgp->t;
+    if (p!=NULL) {
+        pa=p->left;
+        pb=p->right;
+        free(p);
+        tree_free(pa);
+        tree_free(pb);
+    }
+    freelock2(l);
+    free(l);
+    free(tgp);
+}
+
+void treebox_free(treebox b) {
+    struct tree_t *t = *b;
+    tree_free(t);
+    free(b);
+}
+
+//Orginal Code
+void insert (treebox t, int x, void *value) {
+    struct tree_t *tgt = *t;
+    struct tree *p;
+    void *l = tgt->lock;
+    acquire(l);
+    for(;;) {
+        p = tgt->t;
+        if (p==NULL) {
+            tree_t *p1 = (struct tree_t *) surely_malloc (sizeof *tgt);
+            tree_t *p2 = (struct tree_t *) surely_malloc (sizeof *tgt);
+            p1 ->t = NULL;
+            p2 ->t = NULL;
+            lock_t *l1 = (lock_t *) surely_malloc(sizeof(lock_t));
+            makelock(l1);
+            p1->lock = l1;
+            release2(l1);
[…]

### 2023-03-30 concurrency/CONCUR 2023/paper.tex
initial writing for CONCUR 2023 - wish us luck

--- 2022-01-28 concurrency/ITP 2022/Verification of Concurrent Binary Search Tree.tex
@@
-\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
 %This is a template for producing LIPIcs articles. 
@@
 \bibliographystyle{plainurl}% the mandatory bibstyle
-
 \usepackage{txfonts}
@@
 
-\hyphenation{Comp-Cert}
+\title{Verifying Concurrent Search Structure Templates for C Programs using VST} %TODO Please add
 
-\usepackage[utf8]{inputenc}
+%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line
 
-\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking}
-\author{Roshan Sharma}{University of Illinois at Chicago, USA}{}{}{}
-\author{Shengyi Wang}{Princeton University, USA}{}{}{}
-\author{Alexander Oey}{, USA}{}{}{}
-\author{Anastasiia Evdokimova}{University of Illinois at Chicago, USA}{}{}{}
-\author{Lennart Beringer}{Princeton University, USA}{}{}{}
-\author{William Mansky}{University of Illinois at Chicago, USA}{}{}{}
+\author{Duc Than Nguyen}{University of Illinois at Chicago, USA \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+Please provide at least the name of the affiliation and the country.
+The full address is optional.
+Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.
 
-\authorrunning{R.
-Sharma, S.
-Wang, A.
-Oey, A.
-Evdokimova, L.
-Beringer, and W.
-Mansky} %TODO mandatory.
-First: Use abbreviated first/middle names.
-Second (only in severe cases): Use first author plus 'et al.'
+\author{William Mansky}{University of Illinois at Chicago, USA}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
 
-\Copyright{Roshan Sharma, Shengyi Wang, Alexander Oey, Anastasiia Evdokimova, Lennart Beringer, and William Mansky} %TODO mandatory, please use full first names.
-LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\authorrunning{Duc Than Nguyen and William Mansky} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+
+\Copyright{Duc Than Nguyen and William Mansky} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 
-\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+\keywords{...} %TODO mandatory; please add comma-separated list of keywords
+
+\category{} %optional, e.g. invited paper
+
+\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+%\relatedversiondetails[linktext={opt.
+text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g.
+Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional
+
+%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+%\supplementdetails[linktext={opt.
+text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
+Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
 
@@
 
+%TODO mandatory: add short abstract of the document
 \begin{abstract}
-abstract
+... 
 \end{abstract}
 
-% Contributions:
-% verified C implementation of BST -- the first?
-% technique for hand-over-hand locking with share accounting -- slight modification of Gotsman's for logics without held predicates that remember the invariant; amounts to converting Locked_I into another share of the lock assertion
-% technique for connecting locks with abstract state/atomicity
-% VST-style decomposition of proofs into C-to-local-abstract and local-to-global-abstract
-% lock-free implementation?
-
 \section{Introduction}
-\input{intro.tex}
+\label{sec:introduction}
+Krishna et al.
+proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (optimistic concurrency, hand-over-hand locking, internal links) from the proof of correctness of the underlying data structure (linked list, hashtable, B-tree).
+Ideally, it should be possible to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns and obtain $n * m$ verified concurrent data structures.
+In practice, the story is more complicated: certain patterns only work for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
+In this paper, we apply the template approach to concurrent data structure implementations in C, verified using the Verified Software Toolchain (VST)~\cite{plfcc}, and report on its effectiveness and the challenges we encountered.
+The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}; our proofs take advantage of recent work integrating Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
 
 \section{Background}
-\label{background}
-\subsection{VST and Iris}
+\subsection{Concurrent Search Structure Templates}
 
-We use the Verified Software Toolchain \cite{plfcc}  to prove the correctness of the concurrent binary search tree.
-VST is a Coq-based proof system that helps to prove separation logic properties of C programs with the assurance that the properties will be preserved in compiled code.
-First, we write programs in C programming and using CompCert’s \emph{clightgen}, the ASTs for C programs can be generated in Coq.
-We then state and prove specifications for them using Verifiable C.
-Verifiable C is a higher-order separation logic for reasoning about the functional correctness of C programs and carries a soundness proof that links high-level specifications to assembly code generated by the CompCert Verified compiler.
-Verifiable C supports reasoning about two separate kinds of concurrency:  Pthreads-style concurrency with locks (used by implementing blocking data structures that can be fine-grained or coarse-grained ), and concurrency using C11 atomic operations (i.e.
-lock-free data structures).
+\subsection{Iris and VST}
 
-\ignore{\subsection{Ghost State and Global Invariants}
-We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program.
-But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness.
-For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads.
-But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.
-\begin{figure}[htb]
-\centering
-$\texttt{x = 0;}$\\
-$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\qquad\pi = \pi_1\ .\ \pi_2$\\
-$\begin{array}{l || l}
-\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
-\texttt{acquire(l);} & \texttt{acquire(l);}\\
-\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
-\texttt{x++;} & \texttt{x++;}\\
-\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
-\texttt{release(l);} & \texttt{release(l);}\\
-\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
-\end{array}$\\
-$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
-\caption{The increment example annotated with separation logic assertion}
-\label{figure1}
-\end{figure}
-We must preserve the connection between the value of shared resources and the work performed by each thread.
-A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread.
-They do not appear in the original program.
-Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}.
-In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together.
-VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} ($\gnamety$ in Coq) associated with the value $a$, which may be of any type.
-We will see different kind of ghost states used to verify binary search tree in the following sections.
+\section{Search Structure Templates}
+\subsection{Lock Coupling}
 
-The main idea behind the logic from Iris \cite{higherorderghoststate}, a mechnized higher-order concurrent separation logic framework, is the construction of \emph{global invariant} as \emph{ghost state}.
-\emph{Global Invariant} is an invariant on the global (ghost and physical) state of program.
-We can open any invariant but must close it again before taking any steps of execution unless those execution are atomic.
-Thread can use the contents of \emph{global invariant} during atomic operation if it can guarantee that no one will ever see an intermediate state in which invariant does not hold.
-The rules from Jung et al.\cite{higherorderghoststate} for creating and opening invariants are:
+\subsection{Give-Up}
 
-\begin{mathpar}
-\inference[\textsf{inv\_alloc}]{}{\triangleright P \vdash \pvs[E] \mathsf{EX}\ i : \mathsf{iname}, \knowInv{i}{P}}
+\subsection{Templates vs. Internal Reorganization}
 
-\inference[\textsf{inv\_open}]{i \in E}{\knowInv{i}{P} \vdash \pvs[E][E \setminus i] \triangleright P * (\triangleright P \wand \pvs[E \setminus i][E] \mathsf{emp})}
-\end{mathpar}
+\section{Verified Data Structures}
+\subsection{Binary Search Tree}
 
-where invariants are provided in the form of assertion $\knowInv{i}{P}$, and states that $P$ is maintained as an invariant on the global state with name $i$. The operator%%  $\pvs[][]$
-is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e.
-$\knowInv{i}{..\knowInv{i}{P}..}$).}
+%%
+%% Bibliography
+%%
 
-\subsection{Hand-over-Hand Locking}
-\label{selflock}
+%% Please use bibtex,
 
-Hand-over-hand locking (also called lock coupling) is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it.
-Hand-over-hand locking is a classic example of a pattern that requires reference to lock invariants inside other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}.
-The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock.
-By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.
+\bibliography{lipics}
 
-This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources.
-In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it.
-We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$?
-Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
-\begin{align*}
-&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
-&\texttt{acquire}(\ell');\\
-&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
-&\texttt{release}(\ell);\\
-&\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
-\end{align*}
-In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$.
-This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share.
-Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al.
-instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
-$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$
+\appendix
 
-Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
-\begin{align*}
-&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
-&\texttt{acquire}(\ell');\\
-&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
-&\texttt{release}(\ell);\\
-&\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
-\end{align*}
-Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later.
-The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.
+\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}
 
-\subsection{Atomic Specifications}
-\label{atomic}
+List of different predefined enumeration styles:
 
-For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads.
-This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations).
-In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure.
-Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$.
-For instance, the specification $$\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution.
-%maybe something about linking ghost state, which parts are real vs.
-ghost, accessing invariants around atomic triple
-
-VST has encoded an atomic Hoare triple, and provide the way that matches the notation VST uses for normal specifications.
-Such specification is called \emph{atomic specification}, and can be formalized in Coq as shown in  \ref{atomic-spec}.
-This is how we write pre- and postcondition in VST.
-\begin{figure}[htb]
-\centering
-\begin{verbatim}
-Program Definition insert_spec :=
-DECLARE _insert
-ATOMIC TYPE W OBJ a INVS Ei Eo
-WITH ...
-PRE [ ... ]
-  PROP (...)
-  LOCAL (...)
-  SEP (P_l) | (P_p)
-POST [ ... ]
-  PROP ()
-  LOCAL ()
-  SEP (Q_l) | (Q_p)
-\end{verbatim}
-\caption{Atomic Specification in VST}
-\label{atomic-spec}
-\end{figure}
-\texttt{W} is the \textsf{TypeTree} representing the type of arguments passed with \texttt{WITH} clause; \texttt{a} is the abstract state for the triple; \texttt{Ei} and \texttt{Eo} are the sets of invariants names inside and outside the triple.
-The \texttt{PROP} clause describes things that that are true independent of program state, the \texttt{LOCAL} clause describes the values contained in C local variables, and the \texttt{SEP} clause represents the \emph{separating conjunction} (*) of \emph{spatial predicates}, predicates on some part of the memory.
-In VST, while proving that a function implements an atomic specification, the precondition will contains an $\mathsf{atomic\_shift}$ assertion with the public pre- and postcondition, and the masks inside it.
-This atomic shift can be accessed through following two rules:
-$$\inference[\textsf{atomic\_commit}]{\forall a, R * P_p\ a \Rrightarrow \mathsf{EX}\ y,\ Q_p\ a\ y * R'\ y}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
-$$\inference[\textsf{atomic\_rollback}]{\forall a, R * P_p\ a \Rrightarrow P_p\ a * R'}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R'}$$
-During commit, we must provide resources $R$ which, in combination with the public precondition $P_p$, allow us to prove the public postcondition $Q_p$.
-Then, we gain access to an assertion $Q$ required by the postcondition of the function, and leftover resources $R'$.
-During rollback, we provide resources $R$ which, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as leftover resources $R'$.
-The rollback is specially used to learn some relationship between pieces of information (e.g.
-ghost state) stored in the public precondition, while the commit is used to prove the public postcondition from the precondition, and obtain an assertion $Q$.
-To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of rollbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.
-
-\section{Locks and Atomicity}
-
-\section{BST with Hand-over-Hand Locking}
-\label{hand-over-hand}
-As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
-\begin{mathpar}
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ t\ k)\rangle
-\end{mathpar}
-In our case, insert, lookup, and delete are simple Gallina functions on a functional tree data structure.
-The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract tree $t$.
-
-In the following sections, we describe the construction of these predicates, and the proofs of the operations, for BST implementations based on hand-over-hand locking and lock-free atomic operations.
-
-%look at some of the Iris papers in CPP, etc.
-for how we should describe the proofs, e.g.
-section 4 of Formal Verification of a Concurrent Bounded Queue in a Weak Memory Model, Mevel and Jourdan, ICFP 2021
-
-%%%%%%%%%%%%%%%%%%%%%
-
-\ignore{In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct.
-To prove the correctness, our threads need to be able to record the information about the actions they have performed on the shared state, instead of sealing all knowledge of the shared data structure inside the lock invariant.
-We can accomplish this with ghost variables, a simple form of auxiliary state.
-Aside from the \emph{permission} on shared resources held by each thread that we used in the safety proofs, the verification of correctness properties also involves the \emph{ghost state}.
-In VST, any Coq type can be used as ghost state, as long as we can describe what happens when two elements of that type are joined together.
-We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \emph{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.}
-
-%Linearizability is one of the most popular strong correctness condition for the most of %concurrent data-structure, which implies that every operation appears to take place atomically, %in some order, consistent with the real-time ordering of those operations.
-%In order to record the action each thread performed as a linear history, we will define global %invariants using the ghost state, which are similar to lock invariants but are not associated %with any particular memory location.
-Instead, a global invariant is true before and after every %step of a program, acting as a publicly accessible resource.
-A program instruction can use the %contents of global invariant if it can guarantee that no one will ever see an intermediate state %in which the invariant does not hold.
-
-\subsection{BST code}
-\label{selflock}
-
-\begin{figure}[htp]
-\begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
-typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
-typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
-typedef struct tree_t **treebox;
-\end{lstlisting}
-\begin{subfigure}[t]{0.5\textwidth}
-\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
-void insert (treebox t, int x, void *value) {
-  struct tree_t *tgt = *t;
-  struct tree *p;
-  void *l = tgt->lock;
-  acquire(l);
-  for(;;) {
-    p = tgt->t;
-    if (p==NULL) {
-      tree_t *p1 = malloc(sizeof *tgt);
-      tree_t *p2 = malloc(sizeof *tgt);
-      p1->t = NULL;
-      p2->t = NULL;
-      lock_t *l1 = malloc(sizeof(lock_t));
-      makelock(l1);
-      p1->lock = l1;
-      release2(l1);
-      lock_t *l2 = malloc(sizeof(lock_t));
-      makelock(l2);
-      p2->lock = l2;
-      release(l2);
-      p = malloc(sizeof *p);
-      tgt->t = p;
-      p->key=x; p->value=value; p->left=p1; p->right=p2;
-      release(l);
-      return;
-    } else {
-      int y = p->key;
-      if (x<y){
-      	tgt = p->left;
-        void *l_old = l;
-        l = tgt->lock;
-        acquire(l);
-        release(l_old);
-      } else if (y<x){
-        tgt = p->right;
-        void *l_old = l;
-        l = tgt->lock;
-        acquire(l);
-        release(l_old);
-      } else {
-      	p->value=value;
-        release(l);
-      	return;
-      }
-    }
-  }
-} 
-\end{lstlisting} 
-\end{subfigure}\qquad
-\begin{subfigure}[t]{0.4\textwidth}
- \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
-void *lookup (treebox t, int x) {
-  struct tree *p; void *v;
-  struct tree_t *tgt;
-  tgt = *t;
-  void *l = tgt->lock;
-  acquire(l);
-  p = tgt->t;
-  while (p != NULL) {
-    int y = p->key;
-    if (x<y){
-      tgt = p->left;
-      void *l_old = l;
-      l = tgt->lock;
-      acquire(l);
-      p = tgt->t;
-      release(l_old);
-    } else if (y<x){
-      tgt = p->right;
-      void *l_old = l;
-      l = tgt->lock;
-      acquire(l);
-      p = tgt->t;
-      release2(l_old);
-    } else {
-      v = p->value;
-      release(l);
-      return v;
-    }
-  }
-  release(l);
-  return NULL;
-}
-\end{lstlisting}
-\end{subfigure}
-\caption{BST with hand-over-hand locking}
-\label{bst-conc}
-\end{figure}
-
-\subsection{Fine-grained locking and atomicity}
-\label{atomicity}
-
-To prove that an operation on a data structure satisfies a logically atomic specification, we must show that there are no visible intermediate states of the operation, i.e., that other threads see the data structure as unchanged until the linearization point at which the operation takes effect.
-In traditional concurrent separation logics such as VST's~\cite{oraclesematic}, this is done by protecting the data structure with a single lock whose \emph{lock invariant} describes consistent states of the data structure.
-In modern separation logics such as Iris~\cite{iris} or FCSL~\cite{fcsl}, we can make a series of atomic modifications to the data structure without continuously holding a lock, as long as one modification can be designated the \emph{linearization point} at which the new state becomes public, and all prior changes are invisible to other threads.
-Fine-grained locking, in which the pieces of a data structure are each protected by a separate lock with a traditional lock invariant, requires us to synthesize these two approaches: we may make a series of non-atomic modifications to a part of the data structure as long as we hold its lock, but when we release the lock we need to ensure that either we have reached the linearization point, or our changes are invisible to other threads.
-
-We accomplish this using a technique derived from unpublished work by Jung~\cite{ralf-convo} for treating the critical sections of locks as atomic operations.
-The key is to include in each lock invariant a piece of ghost state that represents the abstract state of the locked section.
-From the outside, the effect of the critical section is to atomically update that piece of ghost state, and all other changes are internal to the lock.
-These pieces of ghost state can then be composed to yield the abstract state of the entire data structure.
-In effect, this construction turns a traditional lock invariant into a \emph{lock variant} $\ell \lockvar R$, where $R$ is parameterized by a value that can be updated in the critical section, as long as it corresponds to a linearization point or invisible change to the abstract state of the data structure as a whole.
-This allows traditional lock invariants to interact with atomic shifts and be used to prove atomic specifications.
-
-The construction of the lock variants is simple.
-We begin with a \emph{part-reference} ghost state, a common construction for sharing information between an invariant and the threads that act on it, with elements $\publichalf{a}$ (the current state $a$) and $\myhalf{\pi}{p}$ (partial information about the current state, annotated with a share $\pi$).
-When the share $\pi$ is 1, we know that $p = a$.
-(do we want to define this ourselves, or reference it somewhere?
-note also that this is what Iris calls an authoritative algebra) A lock variant with an assertion $R(a)$ is then defined as %notation
-$$\ell \lockvar_\pi R \triangleq \ell \islock \exists a.\ R(a) * \myhalf{\pi}{a}$$ %eliding ghost names here
-The share $\pi$ may be set to 1 if the lock's information about $a$ is always fully up to date, or a smaller share if it can become outdated due to operations on other parts of the data structure.
-Since $a$ is existentially quantified, a thread that changes the value of $a$ forgets it after releasing the lock (as required in a lock invariant), but it must have synchronized with a $\publichalf{a}$ held elsewhere (usually in a global invariant) that will remember the change.
-Thus, the ghost state acts as a bridge between the component lock invariant and the global abstract state of the data structure.
-%smaller example before we get into BST?
-
-From the atomic rules of section~\ref{atomic}, we derive the following rules for accessing and modifying pieces of the global abstract state:
-\begin{mathpar}
-\inference[\textsf{sync\_commit}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * \exists x_0',\,x_1'.\ (x_0, x_1) \leadsto (x_0', x_1')\ \land \\(\mathsf{my\_half}(g, \mathit{sh}, x_0') * \mathsf{public\_half}(g, x_1') \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}
-
-\inference[\textsf{sync\_rollback}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{public\_half}(g, x_1) \Rrightarrow P_p\ a * R')}{\begin{array}{c}\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R  \Rrightarrow \\\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R'\end{array}}
-\end{mathpar}
-A thread that holds $\myhalf{}{a}$ (i.e., is in the critical section of a lock with a lock variant) may interact with the abstract state $P_p$ held in an $\textsf{atomic\_shift}$, as long as the corresponding $\publichalf{}$ is part of the abstract state.
-It may either leave both halves as is and gain updated resources $R'$ (including, e.g., a snapshot of the current abstract state), or update both halves to a new state that satisfies the public postcondition $Q_p$, executing the linearization point and obtaining the final postcondition $Q$.
-Typically we will use one of these rules at the end of each critical section, to show that the changes made in the critical section either are invisible to other threads (\textsf{sync\_rollback}) or have fulfilled the linearization point (\textsf{sync\_commit}).
-We can also derive simpler rules for special cases, such as a version of \textsf{sync\_commit} for read-only operations that do not change the state of any component:
-
-$$\inference[\textsf{sync\_commit\_same}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{my\_half}(g, \mathit{sh}, x_0) * \mathsf{public\_half}(g, x_1) \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
-
-\subsection{Specification of the hand-over-hand BST}
-%rewrite to use the above: this is the global state made up of components, etc.
-
-To verify our tree implementation, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call.
-In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure.
-As described above, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$.
-We also need to make sure that our lock invariants support the hand-over-hand pattern.
-
-\subsubsection{Key Ranges}
-When we traverse the BST looking for a key $k$, we will reach either a node containing $k$, or an empty node where $k$ would appear if it was in the tree.
-To prove correctness of this procedure, we need ghost state that tracks where each key ``should appear'' in the current tree.
-As observed by Krishna et al.~\cite{krishna2017flow}, this can be done by associating each node with a lower and upper bound on the keys appearing in the subtree rooted at that node.
-We refer to the product of these bounds as a \emph{range}.
-The range of a node is inherited from its parent, based on the parent's key: if node $n$ has range $(l, r)$ and key $k$, then its left child has range $(l, k)$ and its right child has range $(k, r)$.
-Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty \emph{leaf} nodes.
-\begin{figure}[htb]
-\centering
-\includegraphics[width=70mm,scale=0.5]{../FIG/range_prop.png}
-\caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
-\label{range_bst}
-\end{figure}
-At the leaves, these ranges partition the space of all possible keys.
-If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.
-For example, suppose we want to insert a key $38$ into the tree in Figure~\ref{range_bst}.
-The right child of the node with key 35 is a leaf with range $(35, 40)$, so it is guaranteed to be the right place to insert 38.
-
-\subsubsection{Per-Node and Global Ghost State}
-Importantly, the fact that the leaf ranges partition the space of possible keys remains true even if operations in other threads restructure the tree during our traversal.
-Furthermore, operations further up the tree can only \emph{increase} the range of nodes below them, so if we find a node whose range includes our key, we are guaranteed to have found the right place for it.
-This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations.
-More precisely, the abstract state of a node in the tree consists of its range plus a representation of its contents: for a leaf node, $\mathsf{None}$, and for an internal node, $\mathsf{Some}$ of its current key, value, and the identifiers of its left and right children.
-This per-node abstract state serves as a bridge between the per-node locks of the implementation and the abstract state of the entire tree (the $\treerep$ predicate) in the specification.
-
-The abstract state of a tree is a map from keys to values.
-The $\treerep$ predicate then describes the conditions under which a map $m$ is implemented by the ghost state of a tree.
-Roughly speaking, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$.
-We formalize this by first defining the collection of $\mathsf{public\_half}$ assertions---i.e., the authoritative copies of the per-node ghost states---needed to form an abstract tree:
-\begin{align*}
-\mathsf{public}&\mathsf{\_halves}(t, g, (l, r)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
-& |\ \mathsf{Leaf} \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{None})\\
-& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{Some}\ (k, v, g_l, g_r)) * \mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
-\end{align*}
-We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{public\_half}$ assertions of each node into a single assertion.
-We compute the range of each node's children as described above: if the key in the current node is $k$, then the left child has range $(l, k)$ and the right child has range $(k, r)$.
-
-This collection of $\mathsf{public\_half}$ assertions can be combined with per-node lock variants to ensure that the nodes in an implementation actually form a valid tree.
-However, we also need to record the fact that each node in the implementation actually belongs to the tree $t$.
-We do this by augmenting $\mathsf{public\_halves}$ with a piece of ghost state $\mathsf{ghost\_nodes}(S)$ that records the set of all nodes in the tree, constructed so that each node can hold an $\mathsf{in\_tree}\ g$ assertion guaranteeing that $g \in S$.
-Then our final definition of $\treerep$ is:
-$$\treerep(m, g) \triangleq \exists t.\ t \text{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty)) * \mathsf{ghost\_nodes}(\mathsf{node\_names}(t))$$
-The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in the ghost tree $t$.
-Because we quantify over the abstract tree $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node names and key-value pairs.
-(We will take advantage of this property when verifying the \lstinline{delete} function.)
-
-As described in section~\ref{atomicity}, we connect the per-node locks of our C implementation with the global abstract state of $\treerep$ by associating each node's lock with a lock variant assertion $\ell \islock_\pi R$, where $R$ is parameterized by the abstract state of the node (range + possible contents).
-We will fix the share $\pi$ to 1/2 so that the range in a node's lock can be out of date with respect to its $\mathsf{public\_half}$ (in particular, it may have grown due to rearrangements further up the tree).
-We then need to choose a variant $R$ describing the resources held by the lock for a given abstract state.
-The C implementation of a node is:
-\begin{lstlisting}[language = C,numbers = none]
-typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
-typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
-\end{lstlisting}
-The node data of each \lstinline{tree} is wrapped in a \lstinline{tree_t} struct, and is only accessible when the lock in the \lstinline{lock} field is held.
-The lock variant for a node at location $p$ then needs to describe the contents of the \lstinline{t} field for each possible abstract state:
-\begin{align*}
-&\mathsf{match}\ c\ \mathsf{with}\\
-&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
-&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)
-\end{align*}
-We then combine this with the hand-over-hand invariant pattern of section~\ref{selflock}, giving us a final lock variant of:
-\begin{align*}
-\mathsf{node\_inv}((l, r), c) \triangleq \ &p.\texttt{lock} \lockvar_{\pi_1} \mathsf{node\_inv} * \mathsf{match}\ c\ \mathsf{with}\\
-&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
-&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * \\&\qquad p_l.\texttt{lock} \lockvar_{\pi_2} \mathsf{node\_inv} * p_r.\texttt{lock} \lockvar_{\pi_2} \mathsf{node\_inv}
-\end{align*}
-While we do not know the ranges or contents of the child nodes, the lock-variant construction guarantees that they are consistent with the $\mathsf{public\_half}$ assertions of the global ghost state, and the hand-over-hand pattern allows us to acquire $p_l.\texttt{lock}$ or $p_r.\texttt{lock}$ and then release $p.\texttt{lock}$ without losing any shares of the lock assertions.
-
-Now we have everything we need to specify the tree operations:
-\begin{mathpar}
-\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle
-
-\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{lookup}\ m\ k = v\rangle
-
-\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ m\ k)\rangle
-\end{mathpar}
-where $\treerep\ m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock variant for the root node:
-$$\nodeboxrep(b) \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \lockvar \mathsf{node\_inv}$$
-Now we are prepared to prove that each of our BST functions satisfies its specification.
-
-\subsection{Proofs: Insert and Lookup}
-The basic verification style in VST is to divide proofs into two parts: proofs that the C implementations meet specifications that are Coq-level encodings of their behavior, and proofs that those encodings actually have the desired properties.
-In the single-threaded case, we might prove that some code implements a binary search tree written in Gallina, and then prove that that tree satisfies map axioms.
-Similarly, we divide our concurrency proofs into two parts: code-specific proofs that describe how the C code implements changes to per-node ghost state, and more abstract proofs describing how changes in local ghost state lead to the desired updates to the global state.
-
-\subsubsection{Insert}
-An outline of the C code for the insert function is shown in figure \ref{bst-conc}.
-When a thread inserts a key $k$ into a tree, it first crawls the tree to find the right position for $k$ using hand-over-hand locking.
-If it encounters a node with key $k$, then it simply changes the value at that node.
-Otherwise, it will eventually reach a leaf node, where it creates two new empty leaf nodes, inserts the key and value into the current node, and links the newly created leaf nodes to the current node.
-The atomic specification for insert method is $$\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle$$
-The \lstinline{insert} function should atomically add the key-value pair $(k, v)$ to the map implemented by the BST at $p$.
-
-The main loop of the \lstinline{insert} function traverses the tree to find the node where $k$ should be inserted.
-At the beginning of each iteration, we are at some node $n$ and hold the lock for that node, giving us the lock variant $\mathsf{node\_inv}((l, r), c)$ where $(l, r)$ is the range of keys in the subtree rooted at $n$.
-The loop invariant holds that $k \in (l, r)$, that is, the key $k$ always belongs in the subtree of the current node.
-This guarantees us that if we reach an empty leaf, that leaf is exactly the right place to add a node with key $k$.
-
-The function has two possible linearization points: after line 23, where we finish adding a node $(k, v)$ to the tree, and after line 41, where we change the value of an existing node with key $k$ to $v$.
-In both cases, we use the \textsf{sync\_commit} rule to atomically update the abstract state of the data structure from $\treerep\ m$ to $\treerep (\mathrm{insert}\ m\ k\ v)$, by showing that the update to the tree implementation corresponds to a local change in the abstract tree that implements the mathematical $\mathsf{insert}$.
-In the first case, the local change is the insertion of $(k, v)$ at an empty leaf whose range includes $k$, and the addition of two new empty leaves as its children; in the second case, we simply change the value of an internal node whose key is already $k$.
-Our higher-level proofs, then, need to show that these two changes are both valid operations on ghost trees that yield the desired insertion in the global state.
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=150mm,scale=0.5]{../FIG/extract_insert.png}
-\caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation }
-\label{extract_insert}
-\end{figure}
-\begin{figure}[htb]
-\centering
-\includegraphics[width=150mm,scale=0.5]{../FIG/extract_insert_2.png}
-\caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation, where key $k$ already exists in the tree $t$ }
-\label{extract_insert2}
-\end{figure}
-
-\paragraph{Changing the Global State} describe ``extract lemmas'' here---how much detail?
-
-\subsubsection{Lookup}
-
-The code for the $\mathsf{lookup}$ method is shown in Figure
-\ref{lookupproof}. It takes the location of the root pointer and a key
-as the arguments. A thread spans the tree to find a given key using
-hand-over-hand locking mechanism. Once a thread finds the key in the
-tree, it gets the value associated with that key, releases the current
-node's lock, and returns the value.
-The atomic specification for $\mathsf{lookup}$ is
-$$\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle$$
-The \texttt{lookup} function should atomically look up the value of the key $k$ in the tree at $p$.
-
-As before, the key to the verification is the invariant for the main loop:
-\begin{align*} &\mathsf{lookup\_inv}(b, x, \mathit{g\_root}, \mathit{range},\mathit{info}) \triangleq\ \exists\ \mathit{lock},\ \mathit{g\_current},\ \mathit{np},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{\mathit{g\_current}}\\& *\ R\ np\ * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}
-where $P_p$ and $Q_p$ are $\treerep(g\_root, t)$ and $(v =
-M[k])\ \wedge\ \treerep(g\_root, t)$ respectively. The proof steps for
-the $\texttt{lookup}$ verification are similar to the steps
-used in $\texttt{insert}$ verification, with a few differences which we
-discuss below.
-
-Figure \ref{lookupproof} shows the $\mathsf{lookup}$ function
-with separation logic annotations. The proof starts with
-$\nodeboxrep$ and $\mathsf{atomic\_shift}$ assertions as the
-precondition and follows the same approach as the $\mathsf{insert}$
-proof for the verification of loop body. Once we find the target key
-(the last $\mathsf{else}$ clause in the code), we need
-to confirm that the value present in the tree corresponds with the value we would find if we looked up the key in the current abstract state. We accomplish this by opening the global
-invariant encoded in the $\mathsf{atomic\_shift}$, and proving that
-lookup satisfies the public postcondition $\mathsf{Q\_p}$ with the help
-of the $\mathsf{sync\_commit\_same}$ lemma described in section
-\ref{atomicity}. This is the linearization point of the
-$\mathsf{lookup}$ operation and must be done before releasing the
-current node's lock. When we move into the left or right sub-tree,
-we need to establish $\mathsf{lookup\_inv}$ at the end of the $\mathsf{if}$
-or $\mathsf{else\ if}$ clause. Here, we need to show that the key we
-are searching for is still in the range of our new node, so we use
-$\mathsf{sync\_rollback}$ to extract the bound of the left/right
-node encoded as the ghost state in the public precondition
-($\treerep(M, g, g\_root)$).
-
-A critical difference from $\texttt{insert}$ is that the
-$\texttt{lookup}$ method does not change the shape of the tree. With the
-help of $\texttt{sync\_commit\_same}$, we need no longer the
-heavyweight $\emph{extract}$ lemma but instead a simpler
-$\emph{ramif}$ lemma to locate the current node in the global
-invariants:
-\begin{verbatim}
-Lemma ghost_tree_rep_public_half_ramif: forall tg g_root r_root g_in,
-Ensembles.In (find_ghost_set tg g_root) g_in -> ghost_tree_rep tg g_root r_root |-- 
-EX r: node_info, !! (range_info_in_tree r r_root tg) && (public_half g_in r * 
-(public_half g_in r -* ghost_tree_rep tg g_root r_root)).
-\end{verbatim}
-
-\begin{figure}[htp]
-\begin{subfigure}[t]{1\textwidth}
- $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
-\begin{lstlisting}[language = C,  numbers = none]
-void *lookup (treebox t, int x) {
-  struct tree *p; void *v;  struct tree_t *tgt;
-  tgt = *t;  void *l = tgt->lock;
-  acquire(l);  p = tgt->t;
- \end{lstlisting}  
- $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}((-\infty,+\infty),info)}{g\_root}\ *\ R\ b\ *\ \mathsf{lock\_inv}(l,lsh2,R')\ *\ \\
- \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} lookup\_inv \end{array}\right\}$$ 
-  \begin{lstlisting}[language = C, numbers = none]
-    while (p!=NULL) {
-       \end{lstlisting}   
-   $$\left\{\begin{array}{l} lookup\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\\\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
-      \begin{lstlisting}[language = C,  numbers = none]
-    if (x<y){
-      tgt=p->left;
-      ....
-    }else if (y<x){
-      tgt=p->right;
-     ....
-    }else {
-    v = p->value;
-           \end{lstlisting} 
-  $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit\_same}}$$
-$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
-        \begin{lstlisting}[language = C,  numbers = none]
-      release2(l);
-         \end{lstlisting}
-       $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
-         \begin{lstlisting}[language = C, numbers = none]
-       return v;} }
-  release2(l);  return NULL;  }
- \end{lstlisting} 
-\end{subfigure}
-\caption{The $\texttt{lookup}$ function annotated with separation logic specification}
-\label{lookupproof}
-\end{figure}
-
-\subsection{Delete}
-
-The code for the delete method is shown in ??. It takes the location of the root pointer
-and a key as the arguments. A thread traverses the tree using hand-over-hand locking,
-until the thread finds the key to be deleted. To remove the key, the tree is changed 
-through an operation called pushdown left such that the node to be deleted points to an empty leaf
-right child. To achieve that, the thread sets the right child's left child as the 
-new right child of the to-be-deleted node and moves the right child's right child as 
-the new parent of the to-be-deleted node. This operation is repeated until the right
-child of the to-be-deleted node is an empty leaf node. Then, the thread will delete
-the node and set the delete node's parent pointer to the left child of the deleted node.
-
-The atomic specification for delete method can be written as follows:
-$$\forall t.\ \langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(t,g)\rangle $$ 
-$$\texttt{delete}(p,k)$$
-$$\langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(\texttt{delete}(t, k),g)\rangle $$
-
-The delete method is unique compared to insert and lookup because the pushdown left 
-operation can change the \textit{public} half range of more than one nodes. In addition,
-the pushdown left operation is one of the main reasons why the ghost tree of the binary
-search tree is not just a ``copy" of the pure tree: the functional model of pushdown\_left
-is unable to account for the change of deleted node's parent's pointer from pointing
-at the deleted node the to the deleted node's right child.
-
-\usetikzlibrary{positioning}
-
-\forestset{
-   lab/.style={
-        label={[anchor=west, scale=0.75] north east:#1},
-    },
-}
-\begin{forest}
-for tree={
-    grow=south,
-    circle, draw, minimum size=3ex, inner sep=1pt,
-    s sep=3mm
-        }
-    % my label/.style={
-        % label={[anchor=south west] north east:#1},
-    % },
-    [, phantom, for children={fit=band}, s sep+=20mm,
-         before drawing tree={%
-          tikz+={%
-            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
-            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
-            \node [anchor=center] at ($(a.east)!1/2!(b.west)$) {$\longrightarrow\quad$};
-          },
-        },
-        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
-            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
-            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
-                [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
-                    [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}]
-                    [55, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,75)$};}
-                        [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(40,55)$};}]
-                        [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
-                    ]
-                ]
-                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
-            ]
-        ]
-        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
-            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
-            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
-                [55, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
-                    [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
-                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(40,35)$};}]
-                        [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,55)$};}]
-                    ]
-                    [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
-                ]
-                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
-            ]
-        ]
-        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
-            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
-            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
-                [55, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
-                    [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
-                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
-                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
-                            [,no edge, draw=none]
-                        ]
-                        [,no edge, draw=none]
-                    ]
-                    [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
-                ]
-                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
-            ]
-        ]
-        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
-            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
-            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
-                [55, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
-                    [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
-                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
-                        [,no edge, draw=none]
-                    ]
-                    [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
-                ]
-                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
-            ]
-        ]
-    ]
-\end{forest}
-
-\subsubsection{delete}
-
-Similar with insert, delete starts by acquiring the root node lock, so that the thread
-can access the information inside the lock invariant of the root node. Then, we prove
-the while loop by showing that the precondition satisfies delete\_inv and the 
-loop body preserves the loop invariant. The first two cases are similar to insert and
-lookup in which the thread finds the target node, traversing by hand-over-hand locking.
-Once the target node is located, the thread will call pushdown\_left and return after.
-Using the specification of pushdown\_left, we show that the post-condition of 
-pushdown\_left implies the loop invariant. Note that before going into pushdown\_left,
-the thread is holding the lock to the target node and will return to the delete
-method without holding any.
-
-\subsubsection{pushdown\_left}
-
-The proof to pushdown\_left is the key step of proving the delete method. 
-First, we show that the precondition of pushdown\_left satisfies the precondition
-of the loop invariant, which includes the assertion that the thread is holding the 
-lock to the deleted node (i.e. the information inside the lock invariant is accessible)
-and an atomic\_shift. Then, the thread accesses that information to acquire a second 
-lock to the right child of the deleted node. Due to the hand-over-hand locking 
-mechanism this will not cause a deadlock since the thread has already acquired the 
-lock to the deleted node. Next, we prove that the action of deleting the node,
-with sync\_commit by opening the global invariant and removing the relevant 
-ghost names (the target node and its right sentinel node), satisfies the loop
-invariant and the postcondition of pushdown\_left.
-
-The second part of pushdown\_left is proving that the turn\_left operation satisfies
-the loop invariant. Using a variant of the sync\_rollback rule, we show that the 
-turn\_left operation preserves the BST, apart from changes in the ranges of two nodes,
-the target node and its right child, and information about the nodes in those two nodes.
-To satisfy the structure of the ranges, we only need to prove that the new range for
-the target node is encompasses the old range of both nodes. This will imply that
-the range of the target node's old parent includes the target node's right child 
-new range as the right child becomes the new parent (and the old parent points to 
-the right child).
-
-One key step of the proof is proving that two ghost trees are equivalent if the union
-of its ghost nodes are equal and the ghost tree is sorted. This allows the global invariant
-to be preserved during the turn\_left operation in which no nodes were actually changed
-except for the structure of the tree.
-
-\section{Lock-Free BST}
-We want to prove that a lock-free BST implementation satisfies the same specification as our hand-over-hand implementation.
-Unfortunately, provably-correct deletion in a lock-free setting is a research topic in itself (cite?), so we begin with a lock-free BST that only supports insert and lookup.
-Once again, we want to prove
-\begin{mathpar}
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle
-
-\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle
-\end{mathpar}
-though the precise definitions of $\nodeboxrep$ and $\treerep$ will differ. %something about how the fact that the specs are the same really comes across to a client -- and should we prove a client?
-
-\section{Using the Specifications}
-client and client proofs
-
-\section{Related Work}
-\label{related}
-Our work is based on the concurrent separation logic of VST in the style of Iris \cite{higherorderghoststate} and the logical atomicity from TaDA logic  \cite{tada}.
-The notion of the \emph{lower} and \emph{upper} bound is taken from the flow interface paper \cite{krishna2017flow}.
-Our main technical contributions are the verification of two different implementation of binary search tree (using fine-grained locking and lock-free technique) with respect to the same abstract specification, and first c-level mechanized verification of concurrent search-based data structure (i.e.
-BST).
-Most of the other concurrent separation logic (CSL)-based verification works are on toy languages, while VST lets us use the same logic on real C programs.
-VeriFast~\cite{verifast-conc} also operates on C code, albeit without a full formal semantics of C.
-
-Gotsman and Yang \cite{gotsman} is one of the earliest work on concurrent separation logic.
-They introduced the program logic to reason locally about the heap-manipulating program with the notion of dynamic ownership of heap parts by unbounded number of locks and threads, and shown the verification of singly-linked list with fine-grained locking.
-Xiong et al.
-\cite{Xiong2017Abstract} have demonstrated the verification of ConcurrentSkipListMap from java.util.concurrent library using the recent advances in fine-grained concurrency reasoning.
-Their work is mainly based on the abstract atomicity from TaDA logic, and give two modular specifications for concurrent maps: one specification focus on the entire map structure which is suitable for verifying implementation, and another specification focus on the key-value pairs which appropriate for verifying clients.
-We use the same idea of atomicity (though implemented in Iris) in our work.
-
-Krishna et al.
-\cite{krishna2017flow} have presented the proof technique for concurrent search structure templates based on the flow framework and the Iris separation logic, and verified the implementation of concurrent B-tree, hash tables, and linked lists based on the templates.
-Their work is closely related to our work; we took some inspiration from them in building our ghost state.
-But, their proofs are on a toy language rather than real C code.
-
-\section{Conclusion}
-\label{conclusion}
-
-%% Bibliography
-\bibliography{../sources}
 \end{document}

### 2023-03-30 concurrency/templates/bst_template_coupling.c
initial pushing for template style

@@ -1,0 +1,308 @@
+//
+//  main.c
+//  BST
+//
+//  Created by Duc-Than Nguyen on 1/2/22.
+//
+
+
+#include <stdio.h>
+/* #include <stdlib.h> */
+#include "threads.h"
+
+
+extern void *malloc (size_t n);
+extern void free(void *p);
+extern void exit(int code);
+
+typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
+typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
+
+typedef struct tree_t **treebox;
+treebox tb;
+
+void *surely_malloc (size_t n) {
+    void *p = malloc(n);
+    if (!p) exit(1);
+    return p;
+}
+
+treebox treebox_new(void) {
+    treebox p = (treebox) surely_malloc(sizeof (*p));
+    tree_t *newt = (tree_t *) surely_malloc(sizeof(tree_t));
+    *p = newt;
+    lock_t *l;
+    l = makelock();
+    newt->lock = l;
+    newt->t = NULL;
+    release(l);
+    return p;
+}
+
+void tree_free(struct tree_t *tgp) {
+    struct tree_t *pa, *pb;
+    tree* p;
+    void *l = tgp->lock;
+    acquire(l);
+    p = tgp->t;
+    if (p!=NULL) {
+        pa=p->left;
+        pb=p->right;
+        free(p);
+        tree_free(pa);
+        tree_free(pb);
+    }
+    freelock(l);
+    free(tgp);
+}
+
+void treebox_free(treebox b) {
+    struct tree_t *t = *b;
+    tree_free(t);
+    free(b);
+}
+
+//Template style
+
+typedef struct pn {
+    struct tree_t *p;
+    struct tree_t *n;
+} pn;
+
+int findnext (pn *pn, int x, void *value){
+    int y = pn->p->t->key;
+    
+    if (x < y){
+        pn->n = pn->p->t->left;
+        return 1;
+    }
+    else if (x > y){
+        pn->n = pn->p->t->right;
+        return 1;
+    }
+    else{
+        return 0;
+    }
+}
+
+int traverse(pn *pn, int x, void *value) {
+    //printf("Traversing...\n");
+    //fflush(stdout);
+    int flag = 1;
+    for( ; ; ){
+        pn->p = pn->n;
+        if (pn->p->t == NULL){
+            break;
+        } else {
[…]

### 2023-03-30 concurrency/templates/bst_template_giveup.c

@@ -1,0 +1,316 @@
+//
+//  main.c
+//  BST
+//
+//  Created by Duc-Than Nguyen on 1/2/22.
+//
+
+
+#include <stdio.h>
+/* #include <stdlib.h> */
+#include <limits.h>
+#include "threads.h"
+
+
+extern void *malloc (size_t n);
+extern void free(void *p);
+extern void exit(int code);
+
+typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
+typedef struct tree_t {tree *t; lock_t *lock; int min; int max; } tree_t;
+
+typedef struct tree_t **treebox;
+treebox tb;
+
+void *surely_malloc (size_t n) {
+    void *p = malloc(n);
+    if (!p) exit(1);
+    return p;
+}
+
+treebox treebox_new(void) {
+    treebox p = (treebox) surely_malloc(sizeof (*p));
+    tree_t *newt = (tree_t *) surely_malloc(sizeof(tree_t));
+    *p = newt;
+    lock_t *l;
+    l = makelock();
+    newt->lock = l;
+    newt->t = NULL;
+    newt->min = INT_MIN;
+    newt->max = INT_MAX;
+    release(l);
+    return p;
+}
+
+void tree_free(struct tree_t *tgp) {
+    struct tree_t *pa, *pb;
+    tree* p;
+    void *l = tgp->lock;
+    acquire(l);
+    p = tgp->t;
+    if (p!=NULL) {
+        pa=p->left;
+        pb=p->right;
+        free(p);
+        tree_free(pa);
+        tree_free(pb);
+    }
+    freelock(l);
+    free(tgp);
+}
+
+void treebox_free(treebox b) {
+    struct tree_t *t = *b;
+    tree_free(t);
+    free(b);
+}
+
+//Template style
+
+typedef struct pn {
+    struct tree_t *p;
+    struct tree_t *n;
+} pn;
+
+int findnext (pn *pn, int x, void *value){
+    int y = pn->p->t->key;
+    
+    if (x < y){
+        pn->n = pn->p->t->left;
+        return 1;
+    }
+    else if (x > y){
+        pn->n = pn->p->t->right;
+        return 1;
+    }
+    else{
+        return 0;
+    }
+}
+
+int inrange(pn *pn, int x){
+    if (x > pn->p->min && x < pn->p->max)
+        return 1;
+    else
+        return 0;
+}
+
+int traverse(pn *pn, int x, void *value) {
[…]

### 2023-03-30 concurrency/bst_template.c
fix breakage

@@ -1,308 +1,0 @@
-//
-//  main.c
-//  BST
-//
-//  Created by Duc-Than Nguyen on 1/2/22.
-//
-
-
-#include <stdio.h>
-/* #include <stdlib.h> */
-#include "threads.h"
-
-
-extern void *malloc (size_t n);
-extern void free(void *p);
-extern void exit(int code);
-
-typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
-typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
-
-typedef struct tree_t **treebox;
-treebox tb;
-
-void *surely_malloc (size_t n) {
-    void *p = malloc(n);
-    if (!p) exit(1);
-    return p;
-}
-
-treebox treebox_new(void) {
-    treebox p = (treebox) surely_malloc(sizeof (*p));
-    tree_t *newt = (tree_t *) surely_malloc(sizeof(tree_t));
-    *p = newt;
-    lock_t *l;
-    l = makelock();
-    newt->lock = l;
-    newt->t = NULL;
-    release(l);
-    return p;
-}
-
-void tree_free(struct tree_t *tgp) {
-    struct tree_t *pa, *pb;
-    tree* p;
-    void *l = tgp->lock;
-    acquire(l);
-    p = tgp->t;
-    if (p!=NULL) {
-        pa=p->left;
-        pb=p->right;
-        free(p);
-        tree_free(pa);
-        tree_free(pb);
-    }
-    freelock(l);
-    free(tgp);
-}
-
-void treebox_free(treebox b) {
-    struct tree_t *t = *b;
-    tree_free(t);
-    free(b);
-}
-
-//Template style
-
-typedef struct pn {
-    struct tree_t *p;
-    struct tree_t *n;
-} pn;
-
-int findnext (pn *pn, int x, void *value){
-    int y = pn->p->t->key;
-    
-    if (x < y){
-        pn->n = pn->p->t->left;
-        return 1;
-    }
-    else if (x > y){
-        pn->n = pn->p->t->right;
-        return 1;
-    }
-    else{
-        return 0;
-    }
-}
-
-int traverse(pn *pn, int x, void *value) {
-    //printf("Traversing...\n");
-    //fflush(stdout);
-    int flag = 1;
-    for( ; ; ){
-        pn->p = pn->n;
-        if (pn->p->t == NULL){
-            break;
-        } else {
[…]

### 2023-03-31 concurrency/CONCUR 2023/paper.tex
WIP
tried adding code and say something about give-up template

@@
 \usepackage{listings}
+\usepackage{parcolumns}
 \usepackage{iris}
@@
 
-\lstset{language=C,basicstyle=\ttfamily,mathescape=true,columns=fullflexible}
+\lstset{language=C,basicstyle=\ttfamily,
+	xleftmargin=\dimexpr\fboxsep-\fboxrule,
+	mathescape=true,columns=fullflexible}
 
@@
 
+\definecolor{codegreen}{rgb}{0,0.6,0}
+\definecolor{codegray}{rgb}{0.5,0.5,0.5}
+\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
+
+\lstdefinestyle{myStyle}{
+	backgroundcolor=\color{white},   
+	commentstyle=\color{codegreen},
+	basicstyle=\ttfamily\footnotesize,
+	breakatwhitespace=false,         
+	breaklines=true,                 
+	keepspaces=true,                 
+	numbers=left,       
+	numbersep=5pt,                  
+	showspaces=false,                
+	showstringspaces=false,
+	showtabs=false,                  
+	tabsize=1,
+}
+
 \title{Verifying Concurrent Search Structure Templates for C Programs using VST} %TODO Please add
@@
 
-\keywords{...} %TODO mandatory; please add comma-separated list of keywords
+\keywords{concurrent separation logic, fine-grained locking, logical atomicity,
+	Verified Software Toolchain, Iris} %TODO mandatory; please add comma-separated list of keywords
 
@@
 Krishna et al.
-proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (optimistic concurrency, hand-over-hand locking, internal links) from the proof of correctness of the underlying data structure (linked list, hashtable, B-tree).
+proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (optimistic concurrency, hand-over-hand locking, and internal links) from the proof of correctness of the underlying data structure (linked list, hashtable, and B-tree).
 Ideally, it should be possible to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns and obtain $n * m$ verified concurrent data structures.
@@
 In this paper, we apply the template approach to concurrent data structure implementations in C, verified using the Verified Software Toolchain (VST)~\cite{plfcc}, and report on its effectiveness and the challenges we encountered.
-The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}; our proofs take advantage of recent work integrating Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
+The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}; our proofs take advantage of recent work integrating Iris-style logical atomicity into VST~\cite{mansky2022bringing}.
 
@@
 \subsection{Lock Coupling}
+\begin{figure}[!ht]
+	\begin{subfigure}[t]{0.48\textwidth}
+		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
+	\end{subfigure}\qquad
+		\begin{subfigure}[t]{0.49\textwidth}
+			\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
+	\end{subfigure}
+	\caption{Code outlines for \lstinline{traverse} and \lstinline{insert}}
+	\label{bst-conc}
+\end{figure}
+\newpage
+\subsection{Give-Up}
 
-\subsection{Give-Up}
+The give-up template does not maintain locks during traversal between nodes, and instead only acquire them when loading or storing to a node.
+The range field, which is stored in each node, serves as a reference point for threads to determine if they are still on the correct path.
+If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again.
+This behavior is in line with the give-up template algorithm's approach, where threads terminate their search if they encounter obstacles or deviate from the expected path.
 
@@
 \section{Verified Data Structures}
-\subsection{Binary Search Tree}
+
+\section{Related Work}
+
+\section{Conclusion}
 
@@
 
-\bibliography{lipics}
+\bibliography{../sources}
 
-\appendix
+%\appendix
 
-\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}
+%\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}
 
-List of different predefined enumeration styles:
+%List of different predefined enumeration styles:
 

### 2023-03-31 concurrency/CONCUR 2023/lock_insert.c

@@ -1,0 +1,13 @@
+void insert (treebox t, int x, void *value){
+  struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+  pn->n = *t;
+  acquire(pn->n->lock);
+  if (traverse(pn, x, value) == 0){
+        pn->p->t->value = value;
+  }
+  else{
+    insertOp(pn, x, value);
+  }
+  release(pn->n->lock);
+  free(pn);
+}

### 2023-03-31 concurrency/CONCUR 2023/lock_traverse.c

@@ -1,0 +1,20 @@
+int traverse(pn *pn, int x, void *value){
+  int flag = 1;
+  for( ; ; ){
+    pn->p = pn->n;
+    if (pn->p->t == NULL)
+      break;
+    else{
+      int b = findnext(pn, x, value);
+      if (b == 0){
+        flag = 0;
+        break;
+      }
+      else{
+        acquire(pn->n->lock);
+        release(pn->p->lock);
+      }
+    }
+  }
+  return flag;
+}

### 2023-04-03 concurrency/CONCUR 2023/paper.tex
working on section concurrent search structure templates

@@
 \subsection{Concurrent Search Structure Templates}
+A search structure is a data structure that is designed to efficiently store and retrieve data based on some search criteria.
+It is used to organize data in a way that enables fast and efficient search, insertion, deletion, and traversal operations.
+Concurrent search structures are data structures that can be accessed and modified concurrently by multiple threads or processes.
+Designing concurrent search structures presents several key challenges.
+The key challenges of concurrent search structures include ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, maintaining performance and efficiency while managing synchronization and memory, and ensuring compatibility with different hardware and software platforms.
+The complex interaction of concurrency and shared memory in concurrent search structures makes formal verification challenging.
+Modularity is an important aspect of simplifying formal proofs for concurrent search structures, as it has been for the design and maintenance of large systems.
+Breaking down complex systems into smaller, more manageable parts through modularity is important not only in designing and maintaining large systems, but also in simplifying formal proofs for concurrent search structures.
+Modularity enables separation of concerns and minimizes interactions between different parts of the system, which in turn reduces the complexity of formal proofs.
+Additionally, modularity allows for verified components to be reused, further simplifying the verification process.
+{\color{red}{(Say something to link to template algorithms)}}
 
@@
 \section{Search Structure Templates}
+The C implementation of a node in our template style is:
+\begin{lstlisting}[language = C, backgroundcolor=\color{white},   
+	commentstyle=\color{codegreen},
+	basicstyle=\ttfamily\footnotesize,
+	breakatwhitespace=false,  breaklines=true,  keepspaces=true,                                
+	showspaces=false,  showstringspaces=false,
+	showtabs=false, tabsize=1]
+typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
+typedef struct tree_t {tree *t; lock_t *lock; int min; int max;} tree_t;
+typedef struct pn {struct tree_t *p; struct tree_t *n;} pn;
+\end{lstlisting}
+Each node (of type \lstinline{tree_t}) has a lock field that holds the lock protecting the node, and a \lstinline{t} field that is either \lstinline{NULL}, indicating that the node is a leaf node and does not contain any key or value, or points to a \lstinline{tree} struct containing the node's key, value, and child pointers.
+
 \subsection{Lock Coupling}
@@
 	\begin{subfigure}[t]{0.48\textwidth}
+		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
+		\caption{The \lstinline{traverse} method of the give-up template algorithm}
+	\end{subfigure}\qquad
+	\begin{subfigure}[t]{0.48\textwidth}
 		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
-	\end{subfigure}\qquad
-		\begin{subfigure}[t]{0.49\textwidth}
-			\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm}	
 	\end{subfigure}
-	\caption{Code outlines for \lstinline{traverse} and \lstinline{insert}}
-	\label{bst-conc}
+	\caption{The \lstinline{traverse} method of the give-up and lock-coupling template algorithms}
+	\label{traverse}
 \end{figure}
-\newpage
+
 \subsection{Give-Up}

### 2023-04-03 concurrency/CONCUR 2023/giveup_traverse.c

@@ -1,0 +1,26 @@
+int traverse(pn *pn, int x, void *value){
+  tree_t *p = (pn->n); 
+  int flag = 1;
+  for( ; ; ){
+    acquire(pn->n->lock); 
+    pn->p = pn->n;
+    if (inrange(pn, x) == 1){
+      if (pn->p->t == NULL)
+        break;
+      else{
+        int b = findnext(pn, x, value);
+        if (b == 0){
+          flag = 0;
+          break;
+        }
+        else
+          release(pn->p->lock);
+      }
+    }
+    else{
+      release(pn->p->lock);
+      pn->n = p;
+    }
+  }
+  return flag;
+}

### 2023-04-03 concurrency/CONCUR 2023/lock_insert.c

@@ -1,13 +1,0 @@
-void insert (treebox t, int x, void *value){
-  struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
-  pn->n = *t;
-  acquire(pn->n->lock);
-  if (traverse(pn, x, value) == 0){
-        pn->p->t->value = value;
-  }
-  else{
-    insertOp(pn, x, value);
-  }
-  release(pn->n->lock);
-  free(pn);
-}

### 2023-04-03 concurrency/CONCUR 2023/struct.c

@@ -1,0 +1,7 @@
+typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
+typedef struct tree_t {tree *t; lock_t *lock; int min; int max; } tree_t;
+
+typedef struct pn {
+    struct tree_t *p;
+    struct tree_t *n;
+} pn;

### 2023-04-03 concurrency/CONCUR 2023/paper.tex
working on section SS templates

@@
 \subsection{Lock Coupling}
+The lock-coupling template utilizes the hand-over-hand locking scheme to preclude interference from other threads during their traversal.
+Each thread continuously maintains at least one lock during its traversal between nodes.
+This approach ensures that other threads cannot interfere with the traversal or make any changes that would invalidate the ongoing search.
+In Figure \ref{traverse_lock}, two versions of \lstinline{traverse} are presented; one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}).
+
+\begin{figure}[!ht]
+	\begin{subfigure}[t]{0.48\textwidth}
+		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
+		\caption{The \lstinline{traverse} method of the lock-coupling template give-up template algorithm written in C}
+		\label{traverse_lock_a}
+	\end{subfigure}\qquad
+	\begin{subfigure}[t]{0.45\textwidth}
+		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language}
+		\label{traverse_lock_b}	
+	\end{subfigure}
+	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
+	\label{traverse_lock}
+\end{figure}
+
 \begin{figure}[!ht]
@@
 		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
-		\caption{The \lstinline{traverse} method of the give-up template algorithm}
+		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
 	\end{subfigure}\qquad
-	\begin{subfigure}[t]{0.48\textwidth}
-		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
+	\begin{subfigure}[t]{0.45\textwidth}
+		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
 		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm}	
 	\end{subfigure}
-	\caption{The \lstinline{traverse} method of the give-up and lock-coupling template algorithms}
+	\caption{The \lstinline{traverse} method of the give-up template algorithms}
 	\label{traverse}

### 2023-04-03 concurrency/CONCUR 2023/struct.c

@@ -1,7 +1,0 @@
-typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
-typedef struct tree_t {tree *t; lock_t *lock; int min; int max; } tree_t;
-
-typedef struct pn {
-    struct tree_t *p;
-    struct tree_t *n;
-} pn;

### 2023-04-03 concurrency/CONCUR 2023/paper.tex
working on section SS templates

--- 2023-03-31 concurrency/CONCUR 2023/paper.tex
@@
 Ideally, it should be possible to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns and obtain $n * m$ verified concurrent data structures.
-In practice, the story is more complicated: certain patterns only work for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
-In this paper, we apply the template approach to concurrent data structure implementations in C, verified using the Verified Software Toolchain (VST)~\cite{plfcc}, and report on its effectiveness and the challenges we encountered.
+However, in practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
+In this paper, we apply the template approach to concurrent data structure implementations in C, which we verify using the Verified Software Toolchain (VST)~\cite{plfcc}, and report on its effectiveness and the challenges we encountered.
 The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}; our proofs take advantage of recent work integrating Iris-style logical atomicity into VST~\cite{mansky2022bringing}.
@@
 \subsection{Concurrent Search Structure Templates}
+A search structure is a type of data structure that is designed to facilitate the efficient storage and retrieval of data based on specific search criteria.
+These structures organize data in a manner that enables quick and efficient search, insertion, deletion, and traversal operations.
+Concurrent search structures, in turn, are data structures that can be accessed and modified concurrently by multiple threads or processes.
+The design of concurrent search structures poses several significant challenges, including the need to ensure correctness and consistency under concurrent access, achieve scalability by minimizing contention and maximizing parallelism, maintain performance and efficiency while managing synchronization and memory, and ensure compatibility with diverse hardware and software platforms.
+The complex interplay between concurrency and shared memory in concurrent search structures makes formal verification a challenging task.
+Modularity is a crucial aspect of simplifying formal proofs for concurrent search structures, as it is for designing and maintaining large systems.
+By decomposing complex systems into smaller, more manageable components, modularity facilitates the separation of concerns and minimizes interactions between different parts of the system, thus reducing the complexity of formal proofs.
+Moreover, modularity permits the reuse of verified components, which further simplifies the verification process.
+{\color{red}{(Say something to link to template algorithms)}}
 
@@
 \section{Search Structure Templates}
+The C implementation of a node in our template style is:
+\begin{lstlisting}[language = C, backgroundcolor=\color{white},   
+	commentstyle=\color{codegreen},
+	basicstyle=\ttfamily\footnotesize,
+	breakatwhitespace=false,  breaklines=true,  keepspaces=true,                                
+	showspaces=false,  showstringspaces=false,
+	showtabs=false, tabsize=1]
+typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
+typedef struct tree_t {tree *t; lock_t *lock; int min; int max;} tree_t;
+typedef struct pn {struct tree_t *p; struct tree_t *n;} pn;
+\end{lstlisting}
+Each node (of type \lstinline{tree_t}) contains a \lstinline{lock} field that protects the node and a \lstinline{t} field.
+The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers.
+
 \subsection{Lock Coupling}
+The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal.
+Each thread continuously maintains at least one lock during traversal between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
+Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}).
+
 \begin{figure}[!ht]
@@
 		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
+		\caption{The \lstinline{traverse} method of the lock-coupling template give-up template algorithm written in C}
+		\label{traverse_lock_a}
 	\end{subfigure}\qquad
-		\begin{subfigure}[t]{0.49\textwidth}
-			\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
+	\begin{subfigure}[t]{0.45\textwidth}
+		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language}
+		\label{traverse_lock_b}	
 	\end{subfigure}
-	\caption{Code outlines for \lstinline{traverse} and \lstinline{insert}}
-	\label{bst-conc}
+	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
+	\label{traverse_lock}
 \end{figure}
-\newpage
+
+\begin{figure}[!ht]
+	\begin{subfigure}[t]{0.48\textwidth}
+		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
+		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
+	\end{subfigure}\qquad
+	\begin{subfigure}[t]{0.45\textwidth}
+		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm}	
+	\end{subfigure}
+	\caption{The \lstinline{traverse} method of the give-up template algorithms}
+	\label{traverse}
+\end{figure}
+
 \subsection{Give-Up}
 
-The give-up template does not maintain locks during traversal between nodes, and instead only acquire them when loading or storing to a node.
-The range field, which is stored in each node, serves as a reference point for threads to determine if they are still on the correct path.
+The give-up template does not maintain locks during traversal between nodes; instead, it acquires them only when loading or storing to a node.
+In each node, the range field serves as a reference point for threads to determine if they are still on the correct path.
 If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again.
-This behavior is in line with the give-up template algorithm's approach, where threads terminate their search if they encounter obstacles or deviate from the expected path.
+This behavior aligns with the give-up template algorithm's approach, wherein threads terminate their search upon encountering obstacles or deviating from the expected path.
 

### 2023-04-05 concurrency/CONCUR 2023/paper.tex
write more lock-coupling and give-up templates

@@
 proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (optimistic concurrency, hand-over-hand locking, and internal links) from the proof of correctness of the underlying data structure (linked list, hashtable, and B-tree).
-Ideally, it should be possible to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns and obtain $n * m$ verified concurrent data structures.
+Ideally, it should be possible to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns and obtain $n \times m$ verified concurrent data structures.
 However, in practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
@@
 typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
-typedef struct tree_t {tree *t; lock_t *lock; int min; int max;} tree_t;
+typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
 typedef struct pn {struct tree_t *p; struct tree_t *n;} pn;
@@
 The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers.
+To implement hand-over-hand locking, which guarantees that at least one lock is maintained during traversal between nodes (i.e., the lock for the next node is obtained before releasing the lock for the current node), the \lstinline{pn} structure must contain two nodes of type \lstinline{tree_t}: the \lstinline{p} field, representing the current node, and the \lstinline{n} field, representing the next node.
 
@@
 Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}).
+The lock-coupling template algorithm, described in an ML-like language, is presented in Figure \ref{traverse_lock_b} \cite{shasha_1988}.
+The algorithm operates on the assumption that certain helper function(s) must be provided by the implementation, which satisfy certain criteria.
+Such a function, namely \lstinline{findNext}, is utilized by the algorithm.
+The \lstinline{findNext} function determines the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.
 
@@
 
-\begin{figure}[!ht]
-	\begin{subfigure}[t]{0.48\textwidth}
-		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
-		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
-	\end{subfigure}\qquad
-	\begin{subfigure}[t]{0.45\textwidth}
-		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
-		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm}	
-	\end{subfigure}
-	\caption{The \lstinline{traverse} method of the give-up template algorithms}
-	\label{traverse}
-\end{figure}
-
 \subsection{Give-Up}
@@
 This behavior aligns with the give-up template algorithm's approach, wherein threads terminate their search upon encountering obstacles or deviating from the expected path.
+
+\begin{figure}[!ht]
+	\begin{subfigure}[t]{0.48\textwidth}
+		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
+		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
+		\label{traverse_giveup_a}
+	\end{subfigure}\qquad
+	\begin{subfigure}[t]{0.45\textwidth}
+		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm}
+		\label{traverse_giveup_b}	
+	\end{subfigure}
+	\caption{The \lstinline{traverse} method of the give-up template algorithms}
+	\label{traverse_giveup}
+\end{figure}
+
+Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{shasha_1988}. 
+Similar to the lock-coupling template algorithm, the give-up template algorithm utilizes a helper function named \lstinline{inRange} in addition to the \lstinline{findNext n k} function discussed earlier in the lock-coupling section.
+The role of the \lstinline{inRange} function is to check whether the key value $k$ belongs to the range of node $n$.
+If the key value is outside the node's range, the search is abandoned, and the thread is restarted.
 

### 2023-04-05 concurrency/CONCUR 2023/paper.tex
write more lock-coupling and give-up templates

@@
 The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers.
-To implement hand-over-hand locking, which guarantees that at least one lock is maintained during traversal between nodes (i.e., the lock for the next node is obtained before releasing the lock for the current node), the \lstinline{pn} structure must contain two nodes of type \lstinline{tree_t}: the \lstinline{p} field, representing the current node, and the \lstinline{n} field, representing the next node.
+To implement hand-over-hand locking, which guarantees that at least one lock is maintained during traversal between nodes (i.e., the lock for the next node is obtained before releasing the lock for the current node), the \lstinline{pn} struct contain two nodes of type \lstinline{tree_t}: the \lstinline{p} field, representing the current node, and the \lstinline{n} field, representing the next node.
 
@@
 The \lstinline{findNext} function determines the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.
+To utilize the template methodology for concurrent data structure implementations in C, we implement the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
+The \lstinline{pn} struct include two fields: \lstinline{p} and \lstinline{n}, which represent the current and next node, respectively.
+Additionally, we require a target key \lstinline{k}, and a dummy node's value \lstinline{value} (which we call dummy because it is not actually used in \lstinline{traverse}, but we include it to maintain consistent function signatures with other implementations).
 
@@
 		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
-		\caption{The \lstinline{traverse} method of the lock-coupling template give-up template algorithm written in C}
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
 		\label{traverse_lock_a}

### 2023-04-05 concurrency/CONCUR 2023/paper.tex
write more lock-coupling

@@
 Moreover, modularity permits the reuse of verified components, which further simplifies the verification process.
-{\color{red}{(Say something to link to template algorithms)}}
+{\color{red}{(Say something to connect to template algorithms)}}
 
@@
 The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers.
-To implement hand-over-hand locking, which guarantees that at least one lock is maintained during traversal between nodes (i.e., the lock for the next node is obtained before releasing the lock for the current node), the \lstinline{pn} struct contain two nodes of type \lstinline{tree_t}: the \lstinline{p} field, representing the current node, and the \lstinline{n} field, representing the next node.
+To implement hand-over-hand locking, which guarantees that at least one lock is maintained during traversal between nodes (i.e., the lock for the next node is obtained before releasing the lock for the current node), the \lstinline{pn} struct contain two nodes of type \lstinline{tree_t}: the \lstinline{p} field, denoting the current node, and the \lstinline{n} field, representing the next node.
 
@@
 The \lstinline{findNext} function determines the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.
-To utilize the template methodology for concurrent data structure implementations in C, we implement the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
-The \lstinline{pn} struct include two fields: \lstinline{p} and \lstinline{n}, which represent the current and next node, respectively.
-Additionally, we require a target key \lstinline{k}, and a dummy node's value \lstinline{value} (which we call dummy because it is not actually used in \lstinline{traverse}, but we include it to maintain consistent function signatures with other implementations).
 
-\begin{figure}[!ht]
+\begin{figure}[ht]
 	\begin{subfigure}[t]{0.48\textwidth}
@@
 \end{figure}
+
+To utilize the template methodology for concurrent data structure implementations in C, we implement the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
+The \lstinline{pn} struct include two fields: \lstinline{p} and \lstinline{n}, which represent the current and next node, respectively.
+Additionally, we require a target key \lstinline{k}, and a dummy node's value \lstinline{value} (which we call dummy because it is not actually used in \lstinline{traverse}, but we include it to maintain consistent function signatures with other implementations).
+Our version of the \lstinline{traverse} function is similar to the one written in an ML-like language, as it also relies on a helper function called \lstinline{findnext}.
+The purpose of \lstinline{findnext} is to determine the next node to be visited during the traversal.
+In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}.
+Alternatively, it terminates when \lstinline{findnext} fails to find a next node that the key value $k$ contains in the set of key values.
+The \lstinline{traverse} function returns a boolean value that indicates whether it has successfully found the next node during traversal.
+If it fails to do so, it returns false, allowing other functions to make decisions based on this result.
+For example, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of all values less than 20, and a right child with one greater than 20.
+If we want to insert a new node with the target key $k$ into the tree, we can use the \lstinline{traverse} function to navigate the tree.
+When \lstinline{traverse} fails to find the next node (either to the left or right), it means that it has encountered a node with a target key $k$ that is the same as the value of the node.
+At this point, we simply need to update the value of that node to \lstinline{value}.
+When \lstinline{findnext} reaches a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking.
+It then releases the lock for the current node, ensuring that the link between the two nodes is not removed or rearranged while we traverse it.
 
@@
 Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{shasha_1988}. 
-Similar to the lock-coupling template algorithm, the give-up template algorithm utilizes a helper function named \lstinline{inRange} in addition to the \lstinline{findNext n k} function discussed earlier in the lock-coupling section.
+Similar to the lock-coupling template algorithm, the give-up template algorithm utilizes a helper function named \lstinline{inRange} in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
 The role of the \lstinline{inRange} function is to check whether the key value $k$ belongs to the range of node $n$.

### 2023-04-05 concurrency/CONCUR 2023/paper.tex
WIP

--- 2023-04-05 concurrency/CONCUR 2023/paper.tex
@@
 However, in practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
-In this paper, we apply the template approach to concurrent data structure implementations in C, which we verify using the Verified Software Toolchain (VST)~\cite{plfcc}, and report on its effectiveness and the challenges we encountered.
-The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}; our proofs take advantage of recent work integrating Iris-style logical atomicity into VST~\cite{mansky2022bringing}.
+
+In this paper, we apply the template approach to concurrent data structure implementations in C.
+We verify them using the Verified Software Toolchain (VST)~\cite{plfcc}, and report on its effectiveness and the challenges we encountered.
+The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}.
+Our proofs make use of recent work that has integrated Iris-style logical atomicity into VST~\cite{mansky2022bringing}.
 
@@
 \subsection{Concurrent Search Structure Templates}
-A search structure is a type of data structure that is designed to facilitate the efficient storage and retrieval of data based on specific search criteria.
+A search structure is a type of data structure that is designed to efficiently store and retrieve data based on specific search criteria.
 These structures organize data in a manner that enables quick and efficient search, insertion, deletion, and traversal operations.
 Concurrent search structures, in turn, are data structures that can be accessed and modified concurrently by multiple threads or processes.
-The design of concurrent search structures poses several significant challenges, including the need to ensure correctness and consistency under concurrent access, achieve scalability by minimizing contention and maximizing parallelism, maintain performance and efficiency while managing synchronization and memory, and ensure compatibility with diverse hardware and software platforms.
+However, the design of concurrent search structures poses several significant challenges, such as ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory.
+Additionally, these structures must ensure compatibility with diverse hardware and software platforms.
+
 The complex interplay between concurrency and shared memory in concurrent search structures makes formal verification a challenging task.
@@
 Moreover, modularity permits the reuse of verified components, which further simplifies the verification process.
-{\color{red}{(Say something to link to template algorithms)}}
+Concurrent search structures such as lock-free hash tables, concurrent tries, and B-trees are frequently employed.
+These structures allow developers to create systems that are both efficient and scalable, while also ensuring thread safety.
+
+{\color{red}{(Say something to connect to template algorithms)}}
 
@@
 \end{lstlisting}
-Each node (of type \lstinline{tree_t}) contains a \lstinline{lock} field that protects the node and a \lstinline{t} field.
+Each node of type \lstinline{tree_t} has a \lstinline{lock} field that protects the node and a \lstinline{t} field that points to a \lstinline{tree} struct.
 The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers.
-To implement hand-over-hand locking, which guarantees that at least one lock is maintained during traversal between nodes (i.e., the lock for the next node is obtained before releasing the lock for the current node), the \lstinline{pn} structure must contain two nodes of type \lstinline{tree_t}: the \lstinline{p} field, representing the current node, and the \lstinline{n} field, representing the next node.
+The \lstinline{pn} struct contains two nodes of type \lstinline{tree_t}: the \lstinline{p}  field, which denotes the current node, and the \lstinline{n} field, which represents the next node.
+To traverse the tree using hand-over-hand locking, we require at least one lock is maintained during traversal between nodes, meaning that the lock for the next node is obtained before releasing the lock for the current node.
 
@@
 The lock-coupling template algorithm, described in an ML-like language, is presented in Figure \ref{traverse_lock_b} \cite{shasha_1988}.
-The algorithm operates on the assumption that certain helper function(s) must be provided by the implementation, which satisfy certain criteria.
-Such a function, namely \lstinline{findNext}, is utilized by the algorithm.
-The \lstinline{findNext} function determines the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.
+The algorithm assumes that certain helper function(s) must be provided by the implementation that satisfy certain criteria.
+One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.
 
-\begin{figure}[!ht]
+\begin{figure}[ht]
 	\begin{subfigure}[t]{0.48\textwidth}
 		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
-		\caption{The \lstinline{traverse} method of the lock-coupling template give-up template algorithm written in C}
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
 		\label{traverse_lock_a}
@@
 
+To utilize the template methodology for concurrent data structure implementations in C, we have implemented the the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
+The \lstinline{pn} struct includes two fields, namely \lstinline{p} and \lstinline{n}, which respectively represent the current and next node.
+Furthermore, we require a target key \lstinline{k}, and a dummy node's value \lstinline{value} (referred to as 'dummy' because it is not actuallly used in \lstinline{traverse}).
+We include this dummy node to maintain consistency with other implementations.
+Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language since it relies on a helper function called \lstinline{findnext}.
+The purpose of \lstinline{findnext} is to determine the next node to be visited during the traversal.
+In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}.
+Alternatively, it terminates if \lstinline{findnext} fails to find a next node that includes the key value $k$ in the set of key values.
+The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal.
+If it fails to do so, it returns false, which other functions can use to make decisions.
+For instance, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of values less than 20, and a right child with one greater than 20.
+If we want to insert a new node with the target key $k$ into the tree, we can utilize the \lstinline{traverse} function to traverse the tree.
+When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $k$ that equals the node's value.
+At this point, we just have to modify the value of that node to \lstinline{value}.
+When \lstinline{findnext} arrives at a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking.
+It then releases the lock for the current node, ensuring that the link between the two nodes is not removed or rearranged while traversing it.
+
 \subsection{Give-Up}
-
-The give-up template does not maintain locks during traversal between nodes; instead, it acquires them only when loading or storing to a node.
-In each node, the range field serves as a reference point for threads to determine if they are still on the correct path.
+The give-up template is another algorithm for implementing concurrent search in a tree data structure.
+Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires locks when loading or storing to a node.
+This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios.
+Each node has the range field that serves as a reference point for threads to determine if they are still on the correct path.
 If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again.
@@
 Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{shasha_1988}. 
-Similar to the lock-coupling template algorithm, the give-up template algorithm utilizes a helper function named \lstinline{inRange} in addition to the \lstinline{findNext n k} function discussed earlier in the lock-coupling section.
-The role of the \lstinline{inRange} function is to check whether the key value $k$ belongs to the range of node $n$.
-If the key value is outside the node's range, the search is abandoned, and the thread is restarted.
+The give-up template algorithm uses a helper function called \lstinline{inRange} in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
+The purpose of the \lstinline{inRange} function is to determine if the key value $k$ is within the range of node $n$.
+If the key value is outside the node's range, the search is terminated, and the thread is restarted.
 

### 2023-04-05 concurrency/CONCUR 2023/paper.tex
add more about give-up

@@
 
+\begin{lstlisting}[language = C, backgroundcolor=\color{white},   
+	commentstyle=\color{codegreen},
+	basicstyle=\ttfamily\footnotesize,
+	breakatwhitespace=false,  breaklines=true,  keepspaces=true,                                
+	showspaces=false,  showstringspaces=false,
+	showtabs=false, tabsize=1]
+	typedef struct tree_t {tree *t; lock_t *lock; int min; int max;} tree_t;
+\end{lstlisting}
+
+%% C implementation 
+To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node.
+Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{tree_t} in our C implementation.
+These fields respectively store the lower and upper bounds of the keys.
+When a thread arrives at a node in search of $x$, it check if $x$ falls within the range of that node.
+
 \subsection{Templates vs. Internal Reorganization}

### 2023-04-05 concurrency/CONCUR 2023/paper.tex
fix breakage

@@
 		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
-		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm}
+		\caption{The \lstinline{traverse} method of the give-up template algorithm written in ML-like language}
 		\label{traverse_giveup_b}	
@@
 These fields respectively store the lower and upper bounds of the keys.
-When a thread arrives at a node in search of $x$, it check if $x$ falls within the range of that node.
+When a thread arrives at a node in search of $k$, it check if $k$ falls within the range of that node.
 

### 2023-04-05 concurrency/CONCUR 2023/giveup_traverse.c

@@ -1,14 +1,14 @@
-int traverse(pn *pn, int x, void *value){
+int traverse(pn *pn, int k, void *value){
   tree_t *p = (pn->n); 
   int flag = 1;
   for( ; ; ){
     acquire(pn->n->lock); 
     pn->p = pn->n;
-    if (inrange(pn, x) == 1){
+    if (inrange(pn, k) == 1){
       if (pn->p->t == NULL)
         break;
       else{
-        int b = findnext(pn, x, value);
+        int b = findnext(pn, k, value);
         if (b == 0){
           flag = 0;
           break;

### 2023-04-05 concurrency/CONCUR 2023/lock_traverse.c

@@ -1,11 +1,11 @@
-int traverse(pn *pn, int x, void *value){
+int traverse(pn *pn, int k, void *value){
   int flag = 1;
   for( ; ; ){
     pn->p = pn->n;
     if (pn->p->t == NULL)
       break;
     else{
-      int b = findnext(pn, x, value);
+      int b = findnext(pn, k, value);
       if (b == 0){
         flag = 0;
         break;

### 2023-04-07 concurrency/CONCUR 2023/paper.tex

@@
 Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}).
-The lock-coupling template algorithm, described in an ML-like language, is presented in Figure \ref{traverse_lock_b} \cite{shasha_1988}.
 The algorithm assumes that certain helper function(s) must be provided by the implementation that satisfy certain criteria.
@@
 		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
-		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language}
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language \cite{krishna2021automated}}
 		\label{traverse_lock_b}	
@@
 
-Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{shasha_1988}. 
+Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{krishna2021automated}. 
 The give-up template algorithm uses a helper function called \lstinline{inRange} in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
@@
 Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{tree_t} in our C implementation.
-These fields respectively store the lower and upper bounds of the keys.
-When a thread arrives at a node in search of $k$, it check if $k$ falls within the range of that node.
+These fields store the lower and upper bounds of the keys, respectively.
+When a thread arrives at a node in search of the key $k$, it checks whether $k$ falls within the range of that node.
+If $k$ is within the range, we continue by calling \lstinline{findnext} to find the new node.
+If $k$ is not within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (line 21-22 Figure \ref{traverse_giveup_a})
 

### 2023-04-08 concurrency/CONCUR 2023/paper.tex
initial writing for verified data structures

@@
 \usepackage{mathtools}
+\usepackage{amsmath}
 \usepackage{amsthm}
@@
 \newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
-\newcommand{\treerep}{\ensuremath{\mathsf{bst}}}
-\newcommand{\nodeboxrep}{\ensuremath{\mathsf{bst\_ref}}}
+\newcommand{\treerep}{\ensuremath{\mathsf{Node}}}
+\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Node\_ref}}}
+\newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
 
@@
 \newcommand{\publichalf}[1]{\ensuremath{\mathsf{public\_half}(#1)}}
+
+% comments from authors 
+\newcommand{\than}[1]{\textbf{\textcolor{blue}{[ Than: #1]}}}
 
@@
 The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}.
-Our proofs make use of recent work that has integrated Iris-style logical atomicity into VST~\cite{mansky2022bringing}.
+Our proofs make use of recent work that has integrated Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
+
+Our specific contributions are:
+\begin{itemize}
+\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real programming language.
+\item 
+\end{itemize}
 
@@
 \subsection{Concurrent Search Structure Templates}
-A search structure is a type of data structure that is designed to efficiently store and retrieve data based on specific search criteria.
-These structures organize data in a manner that enables quick and efficient search, insertion, deletion, and traversal operations.
-Concurrent search structures, in turn, are data structures that can be accessed and modified concurrently by multiple threads or processes.
-However, the design of concurrent search structures poses several significant challenges, such as ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory.
+A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria.
+These structures organize data to enable quick and efficient search, insertion, deletion, and traversal operations.
+Concurrent search structures, on the other hand, are data structures accessible and modifiable by multiple threads or processes concurrently.
+However, designing concurrent search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory.
 Additionally, these structures must ensure compatibility with diverse hardware and software platforms.
@@
 The complex interplay between concurrency and shared memory in concurrent search structures makes formal verification a challenging task.
-Modularity is a crucial aspect of simplifying formal proofs for concurrent search structures, as it is for designing and maintaining large systems.
-By decomposing complex systems into smaller, more manageable components, modularity facilitates the separation of concerns and minimizes interactions between different parts of the system, thus reducing the complexity of formal proofs.
-Moreover, modularity permits the reuse of verified components, which further simplifies the verification process.
-Concurrent search structures such as lock-free hash tables, concurrent tries, and B-trees are frequently employed.
-These structures allow developers to create systems that are both efficient and scalable, while also ensuring thread safety.
+Modularity is crucial in simplifying formal proofs for concurrent search structures, as it is for designing and maintaining large systems.
+By decomposing complex systems into smaller, more manageable components, modularity facilitates the separation of concerns and minimizes interactions between different parts of the system, reducing the complexity of formal proofs.
+Moreover, modularity permits the reuse of verified components, further simplifying the verification process.
+Lock-free hash tables, concurrent tries, and B-trees are frequently used concurrent search structures.
+These structures allow developers to create efficient, scalable systems while ensuring thread safety.
 
-{\color{red}{(Say something to connect to template algorithms)}}
+\than{Need to say the connection to template algorithms here.}
 
@@
 The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers.
+
 The \lstinline{pn} struct contains two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node.
-To traverse the tree using hand-over-hand locking, we require at least one lock is maintained during traversal between nodes, meaning that the lock for the next node is obtained before releasing the lock for the current node.
+To traverse the tree using hand-over-hand locking, we must maintain at least one lock during traversal between nodes.
+This means that the lock for the next node is obtained before releasing the lock for the current node.
 
-\subsection{Lock Coupling}
+\subsection{Lock-coupling Template}
 The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal.
 Each thread continuously maintains at least one lock during traversal between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
+
 Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}).
-The algorithm assumes that certain helper function(s) must be provided by the implementation that satisfy certain criteria.
+The algorithm assumes that certain helper functions must be provided by the implementation that satisfy certain criteria.
 One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.
+
+To ensure thread safety, it is essential to acquire and release locks in a specific order during traversal.
+The lock-coupling scheme achieves this by requiring threads to acquire locks in increasing order of node addresses.
+The thread then releases the lock for the previous node before acquiring the lock for the next node in the traversal sequence.
+This guarantees that the locks are acquired and released in the same order across all threads and prevents deadlocks from occurring.
 
@@
 
-To utilize the template methodology for concurrent data structure implementations in C, we have implemented the the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
+To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
 The \lstinline{pn} struct includes two fields, namely \lstinline{p} and \lstinline{n}, which respectively represent the current and next node.
-Furthermore, we require a target key \lstinline{k}, and a dummy node's value \lstinline{value} (referred to as 'dummy' because it is not actuallly used in \lstinline{traverse}).
+Additionally, we require a target key \lstinline{k} and a dummy node value \lstinline{value} (referred to as 'dummy' because it is not actually used in \lstinline{traverse}).
 We include this dummy node to maintain consistency with other implementations.
-Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language since it relies on a helper function called \lstinline{findnext}.
+Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findnext}.
 The purpose of \lstinline{findnext} is to determine the next node to be visited during the traversal.
@@
 If it fails to do so, it returns false, which other functions can use to make decisions.
-For instance, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of values less than 20, and a right child with one greater than 20.
-If we want to insert a new node with the target key $k$ into the tree, we can utilize the \lstinline{traverse} function to traverse the tree.
+
+For instance, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of values less than 20 and a right child with one greater than 20.
+If we want to insert a new node with the target key $k$ into the tree, we can use the \lstinline{traverse} function to traverse the tree.
 When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $k$ that equals the node's value.
-At this point, we just have to modify the value of that node to \lstinline{value}.
+At this point, we only have to modify the value of that node to \lstinline{value}.
+
 When \lstinline{findnext} arrives at a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking.
 It then releases the lock for the current node, ensuring that the link between the two nodes is not removed or rearranged while traversing it.
+To utilize the template methodology for concurrent data structure implementations, it is crucial to keep track of the locks acquired for each node during the traversal.
 
 \subsection{Give-Up}
-The give-up template is another algorithm for implementing concurrent search in a tree data structure.
+The give-up template is an algorithm for implementing concurrent search in a tree data structure.
 Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires locks when loading or storing to a node.
 This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios.
-Each node has the range field that serves as a reference point for threads to determine if they are still on the correct path.
+Each node has a range field that serves as a reference point for threads to determine if they are still on the correct path.
 If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again.
-This behavior aligns with the give-up template algorithm's approach, wherein threads terminate their search upon encountering obstacles or deviating from the expected path.
+This behavior aligns with the give-up template algorithm's approach, in which threads terminate their search upon encountering obstacles or deviating from the expected path.
 
@@
 Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{krishna2021automated}.
-The give-up template algorithm uses a helper function called \lstinline{inRange} in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
-The purpose of the \lstinline{inRange} function is to determine if the key value $k$ is within the range of node $n$.
+The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
+The purpose of the \lstinline{inRange} function is to determine whether the key value $k$ falls within the range of node $n$.
 If the key value is outside the node's range, the search is terminated, and the thread is restarted.
 
-\begin{lstlisting}[language = C, backgroundcolor=\color{white},   
-	commentstyle=\color{codegreen},
-	basicstyle=\ttfamily\footnotesize,
-	breakatwhitespace=false,  breaklines=true,  keepspaces=true,                                
-	showspaces=false,  showstringspaces=false,
-	showtabs=false, tabsize=1]
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, commentstyle=\color{codegreen}, basicstyle=\ttfamily\footnotesize,
+breakatwhitespace=false, breaklines=true, keepspaces=true,                              showspaces=false, showstringspaces=false, showtabs=false, tabsize=1]
 	typedef struct tree_t {tree *t; lock_t *lock; int min; int max;} tree_t;
@@
 When a thread arrives at a node in search of the key $k$, it checks whether $k$ falls within the range of that node.
-If $k$ is within the range, we continue by calling \lstinline{findnext} to find the new node.
-If $k$ is not within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (line 21-22 Figure \ref{traverse_giveup_a})
+If $k$ falls within the range, the thread continues by calling \lstinline{findnext} to find the new node.
+If $k$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).
 
@@
 \section{Verified Data Structures}
+
+In this section, we verify \lstinline{traverse} for both lock-coupling and give-up templates.
+
+For many concurrent data structures, the ideal correctness condition requires that the data structure behaves identically to a sequential implementation, even when accessed concurrently by multiple threads.
+The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}.
+These take the form of $\forall a.\ \left\langle P_l\ |\ P_p(a) \right\rangle\ c\ \left\langle Q_l\ |\ Q_p(a)\right\rangle$, where $P_l$ and $Q_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $P_p$ and $Q_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+This triple suggests that if $P_l$ holds true before a call and $P_p$ is true for some value of $a$ in a shared state (e.g.
+a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $c$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $c$ ends).
+For instance, the specification
+$\forall s.\ \left\langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\right\rangle\ \texttt{push}(v)\ \left\langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\right\rangle$
+expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution.
+
+\subsection{Lock-coupling Template}
+Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template:
+
+\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}
+
+$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->p->lock, R'}) \ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
+$$\texttt{traverse(pn, k)}$$
+$$\left\langle v.\ \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->n->lock, R')} \ast \texttt{R(pn->p)} \ast (\texttt{k} \in \texttt{range})
+\ |\ \treerep\ \texttt{t} \right\rangle$$
+
+We need instantiate $\nodeboxrep$ (the resources associated with each client thread) and $\treerep$ (the abstract representation of the data structure), and show how they interact to access and update the state of the data structure for each call.
+In a lock-based implementation, $\nodeboxrep$ serves as a pointer to the root node's lock, and acquiring this lock allows the thread to access more elements within the data structure.
+As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node.
+The resources $\texttt{R}$, parameterized by a node pointer, describes the concrete information about a node.
+The $\lockinv$ asserts that there exists a lock in memory with a given invariant.
+
+\subsection{Give-up Template}
 
@@
 %% Bibliography
+%% 
 %%

### 2023-04-09 concurrency/CONCUR 2023/paper.tex
WIP on verified

@@
 The algorithm assumes that certain helper functions must be provided by the implementation that satisfy certain criteria.
-One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.
+One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$ belonging to the set of key values.
 
@@
 		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
-		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language \cite{krishna2021automated}}
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language \cite{krishna2019compositional}}
 		\label{traverse_lock_b}	
@@
 In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}.
-Alternatively, it terminates if \lstinline{findnext} fails to find a next node that includes the key value $k$ in the set of key values.
+Alternatively, it terminates if \lstinline{findnext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values.
 The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal.
@@
 For instance, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of values less than 20 and a right child with one greater than 20.
-If we want to insert a new node with the target key $k$ into the tree, we can use the \lstinline{traverse} function to traverse the tree.
-When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $k$ that equals the node's value.
+If we want to insert a new node with the target key $\texttt{k}$ into the tree, we can use the \lstinline{traverse} function to traverse the tree.
+When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $\texttt{k}$ that equals the node's value.
 At this point, we only have to modify the value of that node to \lstinline{value}.
@@
 
-Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{krishna2021automated}.
+Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{krishna2019compositional}.
 The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
@@
 These fields store the lower and upper bounds of the keys, respectively.
-When a thread arrives at a node in search of the key $k$, it checks whether $k$ falls within the range of that node.
-If $k$ falls within the range, the thread continues by calling \lstinline{findnext} to find the new node.
-If $k$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).
+When a thread arrives at a node in search of the key $\texttt{k}$, it checks whether $\texttt{k}$ falls within the range of that node.
+If $\texttt{k}$ falls within the range, the thread continues by calling \lstinline{findnext} to find the new node.
+If $\texttt{k}$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).
 
@@
 The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}.
-These take the form of $\forall a.\ \left\langle P_l\ |\ P_p(a) \right\rangle\ c\ \left\langle Q_l\ |\ Q_p(a)\right\rangle$, where $P_l$ and $Q_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $P_p$ and $Q_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+These take the form of $\forall a.\ \left\langle P_l\ |\ P_p(a) \right\rangle\ \texttt{c}\ \left\langle Q_l\ |\ Q_p(a)\right\rangle$, where $P_l$ and $Q_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $P_p$ and $Q_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
 This triple suggests that if $P_l$ holds true before a call and $P_p$ is true for some value of $a$ in a shared state (e.g.
-a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $c$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $c$ ends).
+a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $\texttt{c}$ ends).
 For instance, the specification
-$\forall s.\ \left\langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\right\rangle\ \texttt{push}(v)\ \left\langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\right\rangle$
+$\forall s.\ \left\langle \texttt{is\_stack}\ p\ |\ \texttt{stack}\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}\ p\ |\ \texttt{stack}\ (v::s)\right\rangle$
 expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution.
@@
 
-$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->p->lock, R'}) \ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
+$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->p->lock,R'}) \ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
 $$\texttt{traverse(pn, k)}$$
-$$\left\langle v.\ \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->n->lock, R')} \ast \texttt{R(pn->p)} \ast (\texttt{k} \in \texttt{range})
+$$\left\langle \texttt{n'}.\ \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->n'->lock,R')} \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
 \ |\ \treerep\ \texttt{t} \right\rangle$$
 
-We need instantiate $\nodeboxrep$ (the resources associated with each client thread) and $\treerep$ (the abstract representation of the data structure), and show how they interact to access and update the state of the data structure for each call.
+We need to instantiate $\nodeboxrep$ (the resources associated with each client thread) and $\treerep$ (the abstract representation of the data structure), and show how they interact to access and update the state of the data structure for each call.
 In a lock-based implementation, $\nodeboxrep$ serves as a pointer to the root node's lock, and acquiring this lock allows the thread to access more elements within the data structure.
@@
 The resources $\texttt{R}$, parameterized by a node pointer, describes the concrete information about a node.
-The $\lockinv$ asserts that there exists a lock in memory with a given invariant.
+The lock invariant $\lockinv$ predicate asserts that there exists a lock that protects the resources.
 

### 2023-04-10 concurrency/CONCUR 2023/paper.tex
add  command for Lennart

@@
 \newcommand{\than}[1]{\textbf{\textcolor{blue}{[ Than: #1]}}}
+\newcommand{\lb}[1]{\textbf{\textcolor{red}{[ Lennart: #1]}}}
 
@@
 However, in practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
+
+\lb{say something}
 
@@
 
+$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} 
+%\ast \lockinv \texttt{(pn->p->lock,R'}) 
+\ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
+$$\texttt{traverse(pn, k)}$$
+$$\left\langle \texttt{n'}.\ \nodeboxrep \texttt{(pn)} \
+%ast \lockinv \texttt{(pn->n'->lock,R')} 
+\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
+\ |\ \treerep\ \texttt{t} \right\rangle$$
+
+We need to instantiate $\nodeboxrep$ (the resources associated with each client thread) and $\treerep$ (the abstract representation of the data structure), and demonstrate how they interact to access and update the state of the data structure for each call.
+In a lock-based implementation, $\nodeboxrep$ serves as a pointer to the root node's lock, and acquiring this lock allows the thread to access more elements within the data structure.
+As previously mentioned, the \lstinline{pn} struct consists of two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node.
+The resources $\texttt{R}$, parameterized by a node pointer, describe the concrete information about a node.
+The lock invariant $\lockinv$ predicate asserts that a lock exists to protect the resources.
+
+\subsection{Give-up Template}
+Moving forward with the proofs, we define the following specification for the traverse function within the give-up template:
+
 $$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->p->lock,R'}) \ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
@@
 \ |\ \treerep\ \texttt{t} \right\rangle$$
-
-We need to instantiate $\nodeboxrep$ (the resources associated with each client thread) and $\treerep$ (the abstract representation of the data structure), and show how they interact to access and update the state of the data structure for each call.
-In a lock-based implementation, $\nodeboxrep$ serves as a pointer to the root node's lock, and acquiring this lock allows the thread to access more elements within the data structure.
-As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node.
-The resources $\texttt{R}$, parameterized by a node pointer, describes the concrete information about a node.
-The lock invariant $\lockinv$ predicate asserts that there exists a lock that protects the resources.
-
-\subsection{Give-up Template}
 

### 2023-04-11 concurrency/CONCUR 2023/paper.tex
outline notes

--- 2023-04-08 concurrency/CONCUR 2023/paper.tex
@@
 \usepackage{mathtools}
-\usepackage{amsmath}
 \usepackage{amsthm}
@@
 \newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
-\newcommand{\treerep}{\ensuremath{\mathsf{Node}}}
-\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Node\_ref}}}
-\newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
+\newcommand{\treerep}{\ensuremath{\mathsf{bst}}}
+\newcommand{\nodeboxrep}{\ensuremath{\mathsf{bst\_ref}}}
 
@@
 \newcommand{\publichalf}[1]{\ensuremath{\mathsf{public\_half}(#1)}}
-
-% comments from authors 
-\newcommand{\than}[1]{\textbf{\textcolor{blue}{[ Than: #1]}}}
 
@@
 The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}.
-Our proofs make use of recent work that has integrated Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
-
-Our specific contributions are:
-\begin{itemize}
-\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real programming language.
-\item 
-\end{itemize}
+Our proofs make use of recent work that has integrated Iris-style logical atomicity into VST~\cite{mansky2022bringing}.
 
@@
 \subsection{Concurrent Search Structure Templates}
-A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria.
-These structures organize data to enable quick and efficient search, insertion, deletion, and traversal operations.
-Concurrent search structures, on the other hand, are data structures accessible and modifiable by multiple threads or processes concurrently.
-However, designing concurrent search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory.
+A search structure is a type of data structure that is designed to efficiently store and retrieve data based on specific search criteria.
+These structures organize data in a manner that enables quick and efficient search, insertion, deletion, and traversal operations.
+Concurrent search structures, in turn, are data structures that can be accessed and modified concurrently by multiple threads or processes.
+However, the design of concurrent search structures poses several significant challenges, such as ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory.
 Additionally, these structures must ensure compatibility with diverse hardware and software platforms.
@@
 The complex interplay between concurrency and shared memory in concurrent search structures makes formal verification a challenging task.
-Modularity is crucial in simplifying formal proofs for concurrent search structures, as it is for designing and maintaining large systems.
-By decomposing complex systems into smaller, more manageable components, modularity facilitates the separation of concerns and minimizes interactions between different parts of the system, reducing the complexity of formal proofs.
-Moreover, modularity permits the reuse of verified components, further simplifying the verification process.
-Lock-free hash tables, concurrent tries, and B-trees are frequently used concurrent search structures.
-These structures allow developers to create efficient, scalable systems while ensuring thread safety.
+Modularity is a crucial aspect of simplifying formal proofs for concurrent search structures, as it is for designing and maintaining large systems.
+By decomposing complex systems into smaller, more manageable components, modularity facilitates the separation of concerns and minimizes interactions between different parts of the system, thus reducing the complexity of formal proofs.
+Moreover, modularity permits the reuse of verified components, which further simplifies the verification process.
+Concurrent search structures such as lock-free hash tables, concurrent tries, and B-trees are frequently employed.
+These structures allow developers to create systems that are both efficient and scalable, while also ensuring thread safety.
 
-\than{Need to say the connection to template algorithms here.}
+{\color{red}{(Say something to connect to template algorithms)}}
+
+%point 1: explain the top-level goal and approach of templates, how it relates to modularity
+
+%point 2: lay out the template approach, what a template is, what a data structure is, how their code and their specifications interface
 
@@
 The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers.
+The \lstinline{pn} struct contains two nodes of type \lstinline{tree_t}: the \lstinline{p}  field, which denotes the current node, and the \lstinline{n} field, which represents the next node.
+To traverse the tree using hand-over-hand locking, we require at least one lock is maintained during traversal between nodes, meaning that the lock for the next node is obtained before releasing the lock for the current node.
 
-The \lstinline{pn} struct contains two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node.
-To traverse the tree using hand-over-hand locking, we must maintain at least one lock during traversal between nodes.
-This means that the lock for the next node is obtained before releasing the lock for the current node.
-
-\subsection{Lock-coupling Template}
+\subsection{Lock Coupling}
 The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal.
 Each thread continuously maintains at least one lock during traversal between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
-
 Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}).
-The algorithm assumes that certain helper functions must be provided by the implementation that satisfy certain criteria.
+The algorithm assumes that certain helper function(s) must be provided by the implementation that satisfy certain criteria.
 One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.
-
-To ensure thread safety, it is essential to acquire and release locks in a specific order during traversal.
-The lock-coupling scheme achieves this by requiring threads to acquire locks in increasing order of node addresses.
-The thread then releases the lock for the previous node before acquiring the lock for the next node in the traversal sequence.
-This guarantees that the locks are acquired and released in the same order across all threads and prevents deadlocks from occurring.
 
@@
 
-To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
+To utilize the template methodology for concurrent data structure implementations in C, we have implemented the the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
 The \lstinline{pn} struct includes two fields, namely \lstinline{p} and \lstinline{n}, which respectively represent the current and next node.
-Additionally, we require a target key \lstinline{k} and a dummy node value \lstinline{value} (referred to as 'dummy' because it is not actually used in \lstinline{traverse}).
+Furthermore, we require a target key \lstinline{k}, and a dummy node's value \lstinline{value} (referred to as 'dummy' because it is not actuallly used in \lstinline{traverse}).
 We include this dummy node to maintain consistency with other implementations.
-Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findnext}.
+Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language since it relies on a helper function called \lstinline{findnext}.
 The purpose of \lstinline{findnext} is to determine the next node to be visited during the traversal.
@@
 If it fails to do so, it returns false, which other functions can use to make decisions.
-
-For instance, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of values less than 20 and a right child with one greater than 20.
-If we want to insert a new node with the target key $k$ into the tree, we can use the \lstinline{traverse} function to traverse the tree.
+For instance, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of values less than 20, and a right child with one greater than 20.
+If we want to insert a new node with the target key $k$ into the tree, we can utilize the \lstinline{traverse} function to traverse the tree.
 When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $k$ that equals the node's value.
-At this point, we only have to modify the value of that node to \lstinline{value}.
-
+At this point, we just have to modify the value of that node to \lstinline{value}.
 When \lstinline{findnext} arrives at a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking.
 It then releases the lock for the current node, ensuring that the link between the two nodes is not removed or rearranged while traversing it.
-To utilize the template methodology for concurrent data structure implementations, it is crucial to keep track of the locks acquired for each node during the traversal.
 
 \subsection{Give-Up}
-The give-up template is an algorithm for implementing concurrent search in a tree data structure.
+The give-up template is another algorithm for implementing concurrent search in a tree data structure.
 Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires locks when loading or storing to a node.
 This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios.
-Each node has a range field that serves as a reference point for threads to determine if they are still on the correct path.
+Each node has the range field that serves as a reference point for threads to determine if they are still on the correct path.
 If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again.
-This behavior aligns with the give-up template algorithm's approach, in which threads terminate their search upon encountering obstacles or deviating from the expected path.
+This behavior aligns with the give-up template algorithm's approach, wherein threads terminate their search upon encountering obstacles or deviating from the expected path.
 
@@
 Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{krishna2021automated}. 
-The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
-The purpose of the \lstinline{inRange} function is to determine whether the key value $k$ falls within the range of node $n$.
+The give-up template algorithm uses a helper function called \lstinline{inRange} in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
+The purpose of the \lstinline{inRange} function is to determine if the key value $k$ is within the range of node $n$.
 If the key value is outside the node's range, the search is terminated, and the thread is restarted.
 
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, commentstyle=\color{codegreen}, basicstyle=\ttfamily\footnotesize,
-breakatwhitespace=false, breaklines=true, keepspaces=true,                              showspaces=false, showstringspaces=false, showtabs=false, tabsize=1]
+\begin{lstlisting}[language = C, backgroundcolor=\color{white},   
+	commentstyle=\color{codegreen},
+	basicstyle=\ttfamily\footnotesize,
+	breakatwhitespace=false,  breaklines=true,  keepspaces=true,                                
+	showspaces=false,  showstringspaces=false,
+	showtabs=false, tabsize=1]
 	typedef struct tree_t {tree *t; lock_t *lock; int min; int max;} tree_t;
@@
 When a thread arrives at a node in search of the key $k$, it checks whether $k$ falls within the range of that node.
-If $k$ falls within the range, the thread continues by calling \lstinline{findnext} to find the new node.
-If $k$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).
+If $k$ is within the range, we continue by calling \lstinline{findnext} to find the new node.
+If $k$ is not within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (line 21-22 Figure \ref{traverse_giveup_a})
 
-\subsection{Templates vs. Internal Reorganization}
+\subsection{Templates vs. Internal Reorganization} % move to section 4?
+In the binary search tree, \texttt{pushdown\_left} involves both the structure of the tree and the concurrency protocol.
+There's no good way to disentangle this into a template and a concurrency-free data structure operation.
 
 \section{Verified Data Structures}
-
-In this section, we verify \lstinline{traverse} for both lock-coupling and give-up templates.
-
-For many concurrent data structures, the ideal correctness condition requires that the data structure behaves identically to a sequential implementation, even when accessed concurrently by multiple threads.
-The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}.
-These take the form of $\forall a.\ \left\langle P_l\ |\ P_p(a) \right\rangle\ c\ \left\langle Q_l\ |\ Q_p(a)\right\rangle$, where $P_l$ and $Q_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $P_p$ and $Q_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
-This triple suggests that if $P_l$ holds true before a call and $P_p$ is true for some value of $a$ in a shared state (e.g.
-a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $c$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $c$ ends).
-For instance, the specification
-$\forall s.\ \left\langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\right\rangle\ \texttt{push}(v)\ \left\langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\right\rangle$
-expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution.
-
-\subsection{Lock-coupling Template}
-Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template:
-
-\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}
-
-$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->p->lock, R'}) \ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
-$$\texttt{traverse(pn, k)}$$
-$$\left\langle v.\ \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->n->lock, R')} \ast \texttt{R(pn->p)} \ast (\texttt{k} \in \texttt{range})
-\ |\ \treerep\ \texttt{t} \right\rangle$$
-
-We need instantiate $\nodeboxrep$ (the resources associated with each client thread) and $\treerep$ (the abstract representation of the data structure), and show how they interact to access and update the state of the data structure for each call.
-In a lock-based implementation, $\nodeboxrep$ serves as a pointer to the root node's lock, and acquiring this lock allows the thread to access more elements within the data structure.
-As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node.
-The resources $\texttt{R}$, parameterized by a node pointer, describes the concrete information about a node.
-The $\lockinv$ asserts that there exists a lock in memory with a given invariant.
-
-\subsection{Give-up Template}
 
@@
 %% Bibliography
-%% 
 %%

### 2023-04-11 concurrency/CONCUR 2023/giveup_traverse.c
WIP

@@ -4,11 +4,11 @@ int traverse(pn *pn, int k, void *value){
   for( ; ; ){
     acquire(pn->n->lock); 
     pn->p = pn->n;
-    if (inrange(pn, k) == 1){
+    if (inRange(pn, k) == 1){
       if (pn->p->t == NULL)
         break;
       else{
-        int b = findnext(pn, k, value);
+        int b = findNext(pn, k, value);
         if (b == 0){
           flag = 0;
           break;

### 2023-04-11 concurrency/CONCUR 2023/lock_traverse.c

@@ -5,7 +5,7 @@ int traverse(pn *pn, int k, void *value){
     if (pn->p->t == NULL)
       break;
     else{
-      int b = findnext(pn, k, value);
+      int b = findNext(pn, k, value);
       if (b == 0){
         flag = 0;
         break;

### 2023-04-11 concurrency/CONCUR 2023/paper.tex
fix breakage

--- 2023-04-10 concurrency/CONCUR 2023/paper.tex
@@
 % comments from authors 
-\newcommand{\than}[1]{\textbf{\textcolor{blue}{[ Than: #1]}}}
-\newcommand{\lb}[1]{\textbf{\textcolor{red}{[ Lennart: #1]}}}
-
+\newcommand{\than}[1]{\textbf{\textcolor{blue}{[Than: #1]}}}
+\newcommand{\lb}[1]{\textbf{\textcolor{red}{[Lennart: #1]}}}
+\newcommand{\wm}[1]{\textbf{\textcolor{violet}{[Willam: #1]}}}
 \newcommand{\ignore}[1]{}
@@
 
+\wm{point 1 - explain the top-level goal and approach of templates, how it relates to modularity}
+
+\wm{point 2 - lay out the template approach, what a template is, what a data structure is, how their code and their specifications interface}
+
 \subsection{Iris and VST}
@@
 The C implementation of a node in our template style is:
-\begin{lstlisting}[language = C, backgroundcolor=\color{white},   
-	commentstyle=\color{codegreen},
-	basicstyle=\ttfamily\footnotesize,
-	breakatwhitespace=false,  breaklines=true,  keepspaces=true,                                
-	showspaces=false,  showstringspaces=false,
-	showtabs=false, tabsize=1]
-typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
-typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
-typedef struct pn {struct tree_t *p; struct tree_t *n;} pn;
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+typedef struct node_t {node *t; lock_t *lock;} node_t;
+typedef struct pn {struct node_t *p; struct node_t *n;} pn;
 \end{lstlisting}
-Each node of type \lstinline{tree_t} has a \lstinline{lock} field that protects the node and a \lstinline{t} field that points to a \lstinline{tree} struct.
-The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers.
+\than{I changed \lstinline{tree_t} into \lstinline{node_t}, and \lstinline{tree} into \lstinline{node}, since concurrent search structure will not depend on particular data structures.}
 
-The \lstinline{pn} struct contains two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node.
-To traverse the tree using hand-over-hand locking, we must maintain at least one lock during traversal between nodes.
-This means that the lock for the next node is obtained before releasing the lock for the current node.
+Each node of type \lstinline{node_t} has a \lstinline{lock} field that protects the node and a \lstinline{t} field that points to a \lstinline{node} struct.
+Depending on the data structure, the \lstinline{t} field can have different interpretations.
+In a binary search tree, if the \lstinline{t} field is \lstinline{NULL}, it indicates a leaf node with no key or value; otherwise, it points to a \lstinline{node} struct containing the node's key, value, and child pointers.
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+typedef struct node {int key; void *value; struct node_t *left, *right;} node;
+\end{lstlisting}
+
+For a linked list, a \lstinline{t} field with \lstinline{NULL} represents the last node in the list, while a non-null \lstinline{t} field points to a \lstinline{node} struct holding the node's key, value, and a pointer to the next item in the list.
+
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+typedef struct node {int key; void *value; struct node_t *next;} node;
+\end{lstlisting}
+
+For the purpose of reasoning about lock-coupling style (also known as hand-over-hand locking), we specifically consider traversing the tree using this locking method.
+To accomplish this, we must maintain at least one lock during traversal between nodes.
+This requires obtaining the lock for the next node before releasing the lock for the current node.
+Therefore, we utilize the \lstinline{pn} struct containing two \lstinline{node_t} type nodes: the \lstinline{p} field representing the current node, and the \lstinline{n} field denoting the next node.
 
@@
 We include this dummy node to maintain consistency with other implementations.
-Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findnext}.
-The purpose of \lstinline{findnext} is to determine the next node to be visited during the traversal.
+Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findNext}.
+The purpose of \lstinline{findNext} is to determine the next node to be visited during the traversal.
 In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}.
-Alternatively, it terminates if \lstinline{findnext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values.
+Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values.
 The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal.
@@
 
-When \lstinline{findnext} arrives at a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking.
+When \lstinline{findNext} arrives at a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking.
 It then releases the lock for the current node, ensuring that the link between the two nodes is not removed or rearranged while traversing it.
@@
 When a thread arrives at a node in search of the key $\texttt{k}$, it checks whether $\texttt{k}$ falls within the range of that node.
-If $\texttt{k}$ falls within the range, the thread continues by calling \lstinline{findnext} to find the new node.
+If $\texttt{k}$ falls within the range, the thread continues by calling \lstinline{findNext} to find the new node.
 If $\texttt{k}$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).
 
-\subsection{Templates vs. Internal Reorganization}
+\subsection{Templates vs. Internal Reorganization} % move to section 4?
+In the binary search tree, \texttt{pushdown\_left} involves both the structure of the tree and the concurrency protocol.
+There's no good way to disentangle this into a template and a concurrency-free data structure operation.
 
@@
 $$\texttt{traverse(pn, k)}$$
-$$\left\langle \texttt{n'}.\ \nodeboxrep \texttt{(pn)} \
+$$\left\langle \texttt{res}, \texttt{n'}. \ \nodeboxrep \texttt{(pn)}
 %ast \lockinv \texttt{(pn->n'->lock,R')} 
-\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
+\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range}) \ast (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))
 \   |\ \treerep\ \texttt{t} \right\rangle$$
@@
 \ |\ \treerep\ \texttt{t} \right\rangle$$
+
+\section{Proof Mechanization}
 

### 2023-04-12 concurrency/CONCUR 2023/paper.tex
WIP verified lock-coupling

@@
 The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
-The purpose of the \lstinline{inRange} function is to determine whether the key value $k$ falls within the range of node $n$.
+The purpose of the \lstinline{inRange} function is to determine whether the key value $\texttt{k}$ falls within the range of node $\texttt{n}$.
 If the key value is outside the node's range, the search is terminated, and the thread is restarted.
-
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, commentstyle=\color{codegreen}, basicstyle=\ttfamily\footnotesize,
-breakatwhitespace=false, breaklines=true, keepspaces=true,                              showspaces=false, showstringspaces=false, showtabs=false, tabsize=1]
-	typedef struct tree_t {tree *t; lock_t *lock; int min; int max;} tree_t;
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
 \end{lstlisting}
@@
 There's no good way to disentangle this into a template and a concurrency-free data structure operation.
+
+\than{Or can we move this section to Proof mechanization, where we talk on BST speficially.}
 
@@
 
-We need to instantiate $\nodeboxrep$ (the resources associated with each client thread) and $\treerep$ (the abstract representation of the data structure), and demonstrate how they interact to access and update the state of the data structure for each call.
-In a lock-based implementation, $\nodeboxrep$ serves as a pointer to the root node's lock, and acquiring this lock allows the thread to access more elements within the data structure.
-As previously mentioned, the \lstinline{pn} struct consists of two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node.
-The resources $\texttt{R}$, parameterized by a node pointer, describe the concrete information about a node.
-The lock invariant $\lockinv$ predicate asserts that a lock exists to protect the resources.
+The particulars of $\nodeboxrep$ and $\treerep$ might differ between implementations, but at their core, $\nodeboxrep$ encompasses the resources tied to each client thread, while $\treerep$ acts as an abstract representation containing ghost state that connects the data structure's concrete state to its abstract counterpart.
+Specifically, $\nodeboxrep(pn)$ functions as a pointer to both the node's lock and lock-invariant assertions, representing a protected resource through an arbitrary separation logic predicate.
+Acquiring this lock enables a thread to access further elements within the data structure.
+As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{node_t}: the \lstinline{p} field, which signifies the current node, and the \lstinline{n} field, representing the next node.
+Remember that in the lock-coupling style, each thread consistently maintains at least one lock during traversal between nodes, ensuring other threads cannot disrupt the traversal or invalidate the ongoing search.
+Thus, before invoking the \texttt{traverse} function, the thread must acquire a lock for the current node \texttt{pn->p} and obtain the resource protected by the lock.
+This resource, denoted as \texttt{R(pn->p)}, is parameterized by a node pointer \texttt{pn->p} and describes the concrete information about the node.
+
+%The lock invariant $\lockinv$ predicate asserts that a lock exists to protect the resources.
+
+The specification says ....
+
+The key to the verification of \texttt{traverse} function is the loop invariant for the top-level loop. We define the loop invariant for \texttt{traverse} function of lock-coupling style as follows:
+
+$$
+\texttt{traverse\_inv} 
+$$
+
+Since the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}.
+Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values.
+Therefore, we also define loop invariants for such these cases as follows:
 
@@
 \section{Proof Mechanization}
+We have veried the give-up and lock-coupling template algorithms.
+We describe the mechanized formal verification of the correctness of such these templates, implemented in C and verified with the VST, using concurrent separation
+logic (CSL) to reason about the behavior of the C code, and integrated Iris-style logical atomicity into VST. All proofs are formalized in Coq using VST, and can be found online at ...
 

### 2023-04-12 concurrency/CONCUR 2023/paper.tex
WIP verified lock-coupling

@@
 $$\texttt{traverse(pn, k)}$$
-$$\left\langle \texttt{res}, \texttt{n'}. \ \nodeboxrep \texttt{(pn)}
+$$\left\langle \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn)}
 %ast \lockinv \texttt{(pn->n'->lock,R')} 
-\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range}) \ast (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))
+\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
 \   |\ \treerep\ \texttt{t} \right\rangle$$
@@
 
-The specification says ....
+The specification states that if the node \texttt{p} is within the footprint of the search structure, then the \texttt{traverse} function operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.
 
 The key to the verification of \texttt{traverse} function is the loop invariant for the top-level loop. We define the loop invariant for \texttt{traverse} function of lock-coupling style as follows: 
-
-$$
-\texttt{traverse\_inv} 
-$$
+\begin{align*} &\mathsf{traverse\_inv}() \triangleq\ \exists\ (x\in \mathit{range})\land \texttt{R(pn->p)} \ast  \nodeboxrep(b,\mathit{g\_root})\ \ast \cdots 
+\end{align*}
 

### 2023-04-12 concurrency/CONCUR 2023/paper.tex
transcribed Lennart's comments and added some of my own

@@
 \begin{abstract}
-... 
+Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort.
+In this paper, we implement the template approach in the Verified Software Toolchain, and use it to prove correctness of C implementations of concurrent data structures.
+%say more
 \end{abstract}
@@
 Krishna et al.
-proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (optimistic concurrency, hand-over-hand locking, and internal links) from the proof of correctness of the underlying data structure (linked list, hashtable, and B-tree).
-Ideally, it should be possible to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns and obtain $n \times m$ verified concurrent data structures.
-However, in practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
-
-\lb{say something}
+proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (e.g.
+optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of correctness of the underlying data structure (e.g.
+linked list, hashtable, B-tree).
+The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency.
+In theory, this could allow us to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns, and immediately obtain $n \times m$ verified concurrent data structures.
+In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
 
@@
 \subsection{Concurrent Search Structure Templates}
+\wm{This is probably more background than we need on search structures.}
 A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria.
@@
 Concurrent search structures, on the other hand, are data structures accessible and modifiable by multiple threads or processes concurrently.
-However, designing concurrent search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory.
-Additionally, these structures must ensure compatibility with diverse hardware and software platforms.
+However, designing concurrent search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory, including cache behavior.
+\wm{cite Masstree?} Additionally, these structures must ensure compatibility with diverse hardware and software platforms.
 
@@
 
-\section{Search Structure Templates}
+\section{Search Structure Templates in VST}
 The C implementation of a node in our template style is:
@@
 typedef struct node_t {node *t; lock_t *lock;} node_t;
-typedef struct pn {struct node_t *p; struct node_t *n;} pn;
 \end{lstlisting}
@@
 
-Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}).
+Figure \ref{traverse_lock} shows the \lstinline{traverse} function as presented by Krishna et al.
+(Figure \ref{traverse_lock_a}) and the corresponding C implementation (Figure \ref{traverse_lock_b}).
+The C implementation uses a struct
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+typedef struct pn {struct node_t *p; struct node_t *n;} pn;
+\end{lstlisting}
+to mimic the pair of nodes \lstinline{(p, n)} returned by the functional implementation, where \lstinline{n} is the current node and \lstinline{p} is its parent.
 The algorithm assumes that certain helper functions must be provided by the implementation that satisfy certain criteria.
-One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$ belonging to the set of key values.
+One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$.
+In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a boolean and, if a next node is found, modifies \lstinline{pn->n}.
 
@@
 \begin{figure}[ht]
+	\begin{subfigure}[t]{0.45\textwidth}
+		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional} \wm{Shouldn't it be lock and then unlock?}}
+		\label{traverse_lock_a}	
+	\end{subfigure}\qquad
 	\begin{subfigure}[t]{0.48\textwidth}
@@
 		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
-		\label{traverse_lock_a}
-	\end{subfigure}\qquad
-	\begin{subfigure}[t]{0.45\textwidth}
-		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
-		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language \cite{krishna2019compositional}}
 		\label{traverse_lock_b}
@@
 To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
-The \lstinline{pn} struct includes two fields, namely \lstinline{p} and \lstinline{n}, which respectively represent the current and next node.
 Additionally, we require a target key \lstinline{k} and a dummy node value \lstinline{value} (referred to as 'dummy' because it is not actually used in \lstinline{traverse}).
-We include this dummy node to maintain consistency with other implementations.
+We include this dummy value to maintain consistency with other implementations \wm{What other implementations?}.
 Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findNext}.
@@
 The give-up template is an algorithm for implementing concurrent search in a tree data structure.
-Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires locks when loading or storing to a node.
+Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template takes an \emph{optimistic concurrency control} approach \lb{c.f.
+RCU}, only acquiring locks when loading or storing to a node.
 This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios.
@@
 \begin{figure}[!ht]
+	\begin{subfigure}[t]{0.45\textwidth}
+		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
+		\caption{The \lstinline{traverse} method of the give-up template algorithm written in ML-like language}
+		\label{traverse_giveup_a}	
+	\end{subfigure}\qquad
 	\begin{subfigure}[t]{0.48\textwidth}
@@
 		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
-		\label{traverse_giveup_a}
-	\end{subfigure}\qquad
-	\begin{subfigure}[t]{0.45\textwidth}
-		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
-		\caption{The \lstinline{traverse} method of the give-up template algorithm written in ML-like language}
 		\label{traverse_giveup_b}
@@
 
-Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{krishna2019compositional}.
+Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented.
 The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
@@
 If the key value is outside the node's range, the search is terminated, and the thread is restarted.
+
+\wm{Describe why we need extra pieces in the struct, and how this relates to the other version.}
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
@@
 \section{Verified Data Structures}
+
+\wm{I think this is the wrong organization.
+The proofs of traverse should be presented in the previous section, where we describe traverse.
+This section should show findNext and insert/lookup for specific data structure(s), and then describe how we combine them into top-level specs.}
 

### 2023-04-12 concurrency/CONCUR 2023/paper.tex
resolve William's comment

@@
 		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional} \wm{Shouldn't it be lock and then unlock?}}
+		\than{I copied exactly the same from thesis of Krishna. However, in the their of proofs, they had lock before unlock like you said. Do we need to changed to fit with our context and their code?} 
 		\label{traverse_lock_a}	
@@
 To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node.
-Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{tree_t} in our C implementation.
+Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{node_t} in our C implementation.
 These fields store the lower and upper bounds of the keys, respectively.
@@
 The key to the verification of \texttt{traverse} function is the loop invariant for the top-level loop. We define the loop invariant for \texttt{traverse} function of lock-coupling style as follows: 
-\begin{align*} &\mathsf{traverse\_inv}() \triangleq\ \exists\ (x\in \mathit{range})\land \texttt{R(pn->p)} \ast  \nodeboxrep(b,\mathit{g\_root})\ \ast \cdots 
+\begin{align*} &\mathsf{traverse\_inv}() \triangleq\ \exists\ (\mathsf{k} \in \mathsf{range})\land \mathsf{R(pn->p)} \ast  \nodeboxrep(\mathsf{pn})\ \ast \cdots 
 \end{align*}
@@
 Therefore, we also define loop invariants for such these cases as follows:
+
+\begin{align*} 
+	&\mathsf{traverse\_inv\_1}() \triangleq\ \exists\ (\mathsf{k} \in \mathsf{range})\land \mathsf{R(pn->p)} \ast  \nodeboxrep(\mathsf{pn})\ \ast \cdots 
+	\\
+	& \mathsf{traverse\_inv\_2}() \triangleq\ 
+\end{align*}
 

### 2023-04-13 concurrency/CONCUR 2023/paper.tex
WIP

@@
 
-For instance, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of values less than 20 and a right child with one greater than 20.
+For instance, in a binary search tree, a node with a value of 42 would have a left child with a set of key values consisting of values less than 42 and a right child with one greater than 42.
 If we want to insert a new node with the target key $\texttt{k}$ into the tree, we can use the \lstinline{traverse} function to traverse the tree.
@@
 The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}.
-These take the form of $\forall a.\ \left\langle P_l\ |\ P_p(a) \right\rangle\ \texttt{c}\ \left\langle Q_l\ |\ Q_p(a)\right\rangle$, where $P_l$ and $Q_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $P_p$ and $Q_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
-This triple suggests that if $P_l$ holds true before a call and $P_p$ is true for some value of $a$ in a shared state (e.g.
-a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $\texttt{c}$ ends).
+These take the form of $\forall a.\ \left\langle \mathsf{P}_l\ |\ \mathsf{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \mathsf{Q}_l\ |\ \mathsf{Q}_p(a)\right\rangle$, where $\mathsf{P}_l$ and $\mathsf{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\mathsf{P}_p$ and $\mathsf{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+This triple suggests that if $\mathsf{P}_l$ holds true before a call and $\mathsf{P}_p$ is true for some value of $a$ in a shared state (e.g.
+a global invariant), then $\mathsf{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\mathsf{Q}_p$ will become true atomically for the same value $a$ (and $\mathsf{Q}_l$ will be true after $\texttt{c}$ ends).
 For instance, the specification
-$\forall s.\ \left\langle \texttt{is\_stack}\ p\ |\ \texttt{stack}\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}\ p\ |\ \texttt{stack}\ (v::s)\right\rangle$
+$\forall s.\ \left\langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\right\rangle\ \mathsf{push}(v)\ \left\langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v::s)\right\rangle$
 expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution.
@@
 
-$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} 
+$$\forall \mathsf{t}.\ \left\langle \nodeboxrep \mathsf{(pn->p)} 
 %\ast \lockinv \texttt{(pn->p->lock,R'}) 
-\ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
+\ast \mathsf{R(pn->p)} \ | \ \treerep\ \mathsf{t} \right\rangle $$
 $$\texttt{traverse(pn, k)}$$
-$$\left\langle \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn)}
+$$\left\langle \ \mathsf{res} \ \mathsf{n'}. \ (\mathsf{res} \leftrightarrow (\mathsf{pn->n'->t} = \mathsf{NULL})) \ast \nodeboxrep \mathsf{(pn->n')}
 %ast \lockinv \texttt{(pn->n'->lock,R')} 
-\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
-\   |\ \treerep\ \texttt{t} \right\rangle$$
+\ast \mathsf{R(pn->n')} \ast (\mathsf{k} \in \mathsf{range})
+\   |\ \treerep\ \mathsf{t} \right\rangle$$
+
+\than{Actually, $\nodeboxrep \mathsf{(pn->p)}$ and $\mathsf{data\_at \ p \ pn} \ast \mathsf{my\_half \ g\_root} (-\inf, +\inf)$ is equivalent to $\mathsf{tree\_rep}$.
+But now we can ignore the information of $\mathsf{my\_half}$.
+}
 
 The particulars of $\nodeboxrep$ and $\treerep$ might differ between implementations, but at their core, $\nodeboxrep$ encompasses the resources tied to each client thread, while $\treerep$ acts as an abstract representation containing ghost state that connects the data structure's concrete state to its abstract counterpart.
-Specifically, $\nodeboxrep(pn)$ functions as a pointer to both the node's lock and lock-invariant assertions, representing a protected resource through an arbitrary separation logic predicate.
+Specifically, $\nodeboxrep(\mathsf{pn})$ functions as a pointer to both the node's lock and lock-invariant assertions, representing a protected resource through an arbitrary separation logic predicate.
 Acquiring this lock enables a thread to access further elements within the data structure.
-As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{node_t}: the \lstinline{p} field, which signifies the current node, and the \lstinline{n} field, representing the next node.
+As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{node_t}: the $\mathsf{p}$ field, which signifies the parent node, and the $\mathsf{n}$ field, representing the current node.
 Remember that in the lock-coupling style, each thread consistently maintains at least one lock during traversal between nodes, ensuring other threads cannot disrupt the traversal or invalidate the ongoing search.
-Thus, before invoking the \texttt{traverse} function, the thread must acquire a lock for the current node \texttt{pn->p} and obtain the resource protected by the lock.
-This resource, denoted as \texttt{R(pn->p)}, is parameterized by a node pointer \texttt{pn->p} and describes the concrete information about the node.
+Thus, before invoking the \texttt{traverse} function, the thread must acquire a lock for the parent node $\mathsf{pn->p}$ and obtain the resource protected by the lock.
+This resource, denoted as $\mathsf{R(pn->p)}$, is parameterized by a node pointer $\mathsf{pn->p}$ and describes the concrete information about the node.
 
@@
 The key to the verification of \texttt{traverse} function is the loop invariant for the top-level loop. We define the loop invariant for \texttt{traverse} function of lock-coupling style as follows: 
-\begin{align*} &\mathsf{traverse\_inv}() \triangleq\ \exists\ (\mathsf{k} \in \mathsf{range})\land \mathsf{R(pn->p)} \ast  \nodeboxrep(\mathsf{pn})\ \ast \cdots 
+\begin{align*} &\mathsf{traverse\_inv (pn)} \triangleq\ \exists \mathsf{n'}.\ (\mathsf{k} \in \mathsf{range})\land \nodeboxrep(\mathsf{pn->n'})  \ast  \mathsf{R(pn->n')}   
 \end{align*}
@@
 Therefore, we also define loop invariants for such these cases as follows:
+\begin{align*} 
+	\mathsf{traverse\_inv\_1 (pn)} \triangleq\ \exists \mathsf{n'}.\ (\mathsf{k} \in \mathsf{range})\land \nodeboxrep(\mathsf{pn->n'})  \ast \mathsf{R(pn->n')} \ast (\mathsf{pn->n'->t} = \mathsf{NULL})
+	\\
+	\mathsf{traverse\_inv\_2 (pn)} \triangleq\  \exists \mathsf{n'}.\ (\mathsf{k} \in \mathsf{range})\land \nodeboxrep(\mathsf{pn->n'})  \ast \mathsf{R(pn->n')} \ast (\mathsf{pn->n'->t} \neq \mathsf{NULL})
+\end{align*}
 
-\begin{align*} 
-	&\mathsf{traverse\_inv\_1}() \triangleq\ \exists\ (\mathsf{k} \in \mathsf{range})\land \mathsf{R(pn->p)} \ast  \nodeboxrep(\mathsf{pn})\ \ast \cdots 
-	\\
-	& \mathsf{traverse\_inv\_2}() \triangleq\ 
-\end{align*}
+More specifically, $\mathsf{traverse\_inv\_1 (pn)}$ is true when the \lstinline{traverse} function successfully locates the subsequent node during traversal, as indicated by the condition $\mathsf{pn->n'->t} = \mathsf{NULL}$.
+In the context of a linked-list data structure, a \lstinline{t} field with a value of \lstinline{NULL} signals that the \lstinline{traverse} function has arrived at the list's final node.
+
+In contrast, $\mathsf{traverse\_inv\_2 (pn)}$ is true when the \lstinline{traverse} function fails to locate the next node during traversal, which means $\mathsf{pn->n'->t} \neq \mathsf{NULL}$.
+For instance, when trying to insert a new node with a specific key \lstinline{k} into a linked list, if the \lstinline{traverse} function cannot identify the next node, it implies that the current node has a target key equal to its value.
+Since the node holds the key, the \lstinline{t} field should not be \lstinline{NULL} .
+
+%The loop invariants mentioned above include specific existential variables (such as $\mathsf{n'}$) that define the state of each iteration throughout the loop.
 

### 2023-04-14 concurrency/CONCUR 2023/paper.tex
add comment
modify verified lock-coupling section

--- 2023-04-12 concurrency/CONCUR 2023/paper.tex
@@
 The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}.
-These take the form of $\forall a.\ \left\langle P_l\ |\ P_p(a) \right\rangle\ \texttt{c}\ \left\langle Q_l\ |\ Q_p(a)\right\rangle$, where $P_l$ and $Q_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $P_p$ and $Q_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
-This triple suggests that if $P_l$ holds true before a call and $P_p$ is true for some value of $a$ in a shared state (e.g.
-a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $\texttt{c}$ ends).
+These take the form of $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+This triple suggests that if $\texttt{P}_l$ holds true before a call and $\texttt{P}_p$ is true for some value of $a$ in a shared state (e.g.
+a global invariant), then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).
 For instance, the specification
@@
 
-$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} 
+$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn->n)} 
 %\ast \lockinv \texttt{(pn->p->lock,R'}) 
-\ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
+\ast \texttt{R(pn->n)} \ | \ \treerep\ \texttt{t} \right\rangle $$
 $$\texttt{traverse(pn, k)}$$
-$$\left\langle \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn)}
+$$\left\langle \ \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn->n')}
 %ast \lockinv \texttt{(pn->n'->lock,R')} 
@@
 
+\than{Actually, $\nodeboxrep \texttt{(pn->n)}$ and $\texttt{data\_at  n pn} \ast \texttt{my\_half g\_root} (-\inf, +\inf)$ is equivalent to $\texttt{tree\_rep}$.
+But now we can ignore the information of $\texttt{my\_half}$.
+}
+
 The particulars of $\nodeboxrep$ and $\treerep$ might differ between implementations, but at their core, $\nodeboxrep$ encompasses the resources tied to each client thread, while $\treerep$ acts as an abstract representation containing ghost state that connects the data structure's concrete state to its abstract counterpart.
-Specifically, $\nodeboxrep(pn)$ functions as a pointer to both the node's lock and lock-invariant assertions, representing a protected resource through an arbitrary separation logic predicate.
+Specifically, $\nodeboxrep(\texttt{pn->n})$ functions as a pointer to both the node's lock and lock-invariant assertions, representing a protected resource through an arbitrary separation logic predicate.
 Acquiring this lock enables a thread to access further elements within the data structure.
-As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{node_t}: the \lstinline{p} field, which signifies the current node, and the \lstinline{n} field, representing the next node.
+As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{node_t}: the $\texttt{p}$ field, which signifies the parent node, and the $\texttt{n}$ field, representing the current node.
 Remember that in the lock-coupling style, each thread consistently maintains at least one lock during traversal between nodes, ensuring other threads cannot disrupt the traversal or invalidate the ongoing search.
-Thus, before invoking the \texttt{traverse} function, the thread must acquire a lock for the current node \texttt{pn->p} and obtain the resource protected by the lock.
-This resource, denoted as \texttt{R(pn->p)}, is parameterized by a node pointer \texttt{pn->p} and describes the concrete information about the node.
+Thus, before invoking the \texttt{traverse} function, the thread must acquire a lock for the current node $\texttt{pn->n}$ and obtain the resource protected by the lock.
+This resource, denoted as $\texttt{R(pn->n)}$, is parameterized by a node pointer $\texttt{pn->n}$ and describes the concrete information about the node.
 
@@
 
-The specification states that if the node \texttt{p} is within the footprint of the search structure, then the \texttt{traverse} function operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.
+The specification states that if the node \texttt{n} is within the footprint of the search structure, then the \texttt{traverse} function operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.
 
 The key to the verification of \texttt{traverse} function is the loop invariant for the top-level loop. We define the loop invariant for \texttt{traverse} function of lock-coupling style as follows: 
-\begin{align*} &\mathsf{traverse\_inv}() \triangleq\ \exists\ (\mathsf{k} \in \mathsf{range})\land \mathsf{R(pn->p)} \ast  \nodeboxrep(\mathsf{pn})\ \ast \cdots 
+\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast  \texttt{R(pn->n')}   
 \end{align*}
@@
 Therefore, we also define loop invariants for such these cases as follows:
+\begin{align*} 
+	\texttt{traverse\_inv\_1(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast \texttt{R(pn->n')} \ast (\texttt{pn->n'->t} = \texttt{NULL})
+	\\
+	\texttt{traverse\_inv\_2(pn)} \triangleq\  \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast \texttt{R(pn->n')} \ast (\texttt{pn->n'->t} \neq \texttt{NULL})
+\end{align*}
 
-\begin{align*} 
-	&\mathsf{traverse\_inv\_1}() \triangleq\ \exists\ (\mathsf{k} \in \mathsf{range})\land \mathsf{R(pn->p)} \ast  \nodeboxrep(\mathsf{pn})\ \ast \cdots 
-	\\
-	& \mathsf{traverse\_inv\_2}() \triangleq\ 
-\end{align*}
+More specifically, $\texttt{traverse\_inv\_1 (pn)}$ is true when the \lstinline{traverse} function successfully locates the subsequent node during traversal, as indicated by the condition $\texttt{pn->n'->t} = \texttt{NULL}$.
+In the context of a linked-list data structure, a \lstinline{t} field with a value of \lstinline{NULL} signals that the \lstinline{traverse} function has arrived at the list's final node.
+
+In contrast, $\texttt{traverse\_inv\_2 (pn)}$ is true when the \lstinline{traverse} function fails to locate the next node during traversal, which means $\texttt{pn->n'->t} \neq \texttt{NULL}$.
+For instance, when trying to insert a new node with a specific key \lstinline{k} into a linked list, if the \lstinline{traverse} function cannot identify the next node, it implies that the current node has a target key equal to its value.
+Since the node holds the key, the \lstinline{t} field should not be \lstinline{NULL} .
+
+\than{FIX ME: These things above didn't represent hand-over-hand locking instance, since I tried to avoid talking about lock invariant.}
+%The loop invariants mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop.
+% CHANGE pn-> p to pn->n
+
+The proof starts with $\nodeboxrep \texttt{(pn->n)}$, which includes a pointer to the node's lock, the lock-invariant assertion of $\texttt{(pn->n)}$, and the resource $\texttt{R(pn->n)}$ obtained prior to calling the \lstinline{traverse} function.
+When \lstinline{findNext} successfully identifies the next node (for example, \lstinline{n'}) to visit, the lock invariant within $\nodeboxrep$ enables us to access the resource $\texttt{R(pn->n')}$ upon reaching line 14 in the C implementation of the lock-coupling template (Figure \ref{traverse_lock_b}).
+Following this, we release the lock of the current node (line 15 in Figure \ref{traverse_lock_b}), illustrating that we consistently maintain a single lock during node traversal.
+\than{I feel somewhat perplexed while attempting to articulate the situation mentioned above. Can someone help me fix or explain it more?}
 

### 2023-04-14 concurrency/CONCUR 2023/paper.tex
fixed breakage

@@
 
-The proof starts with $\nodeboxrep \texttt{(pn->n)}$, which includes a pointer to the node's lock, the lock-invariant assertion of $\texttt{(pn->n)}$, and the resource $\texttt{R(pn->n)}$ obtained prior to calling the \lstinline{traverse} function.
-When \lstinline{findNext} successfully identifies the next node (for example, \lstinline{n'}) to visit, the lock invariant within $\nodeboxrep$ enables us to access the resource $\texttt{R(pn->n')}$ upon reaching line 14 in the C implementation of the lock-coupling template (Figure \ref{traverse_lock_b}).
-Following this, we release the lock of the current node (line 15 in Figure \ref{traverse_lock_b}), illustrating that we consistently maintain a single lock during node traversal.
+The proof begins with $\nodeboxrep \texttt{(pn->n)}$, which comprises a pointer to the node's lock, the lock-invariant assertion of $\texttt{(pn->n)}$, and the resource $\texttt{R(pn->n)}$ acquired before invoking the \lstinline{traverse} function.
+When \lstinline{findNext} successfully detects the next node (e.g., \lstinline{n'}) to visit, the lock invariant within $\nodeboxrep$ allows us to access the resource $\texttt{R(pn->n')}$ at line 14 in the C implementation of the lock-coupling template (Figure \ref{traverse_lock_b}).
+Subsequently, we release the lock of the current node (line 15 in Figure \ref{traverse_lock_b}), demonstrating our consistent approach to maintaining a single lock during node traversal.
+The proof concludes by verifying that the \lstinline{traverse} function satisfies the post-condition of the specification mentioned earlier, owing to loop invariants \texttt{traverse\_inv\_1} and \texttt{traverse\_inv\_2}.
+
 \than{I feel somewhat perplexed while attempting to articulate the situation mentioned above. Can someone help me fix or explain it more?}

### 2023-04-14 concurrency/CONCUR 2023/paper.tex
initial writing proof of give-up traverse

@@
 \newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
+\newcommand{\infp}{\ensuremath{\mathsf{InFP}}}
 
@@
 
-The specification states that if the node \texttt{n} is within the footprint of the search structure, then the \texttt{traverse} function operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.
+%The specification states that if the node \texttt{n} is within the footprint of the search structure, then the \texttt{traverse} function operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.
 
-The key to the verification of \texttt{traverse} function is the loop invariant for the top-level loop. We define the loop invariant for \texttt{traverse} function of lock-coupling style as follows: 
+The crux of verifying the \texttt{traverse} function lies in the loop invariant for the top-level loop. We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
 \begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast  \texttt{R(pn->n')}   
 \end{align*}
-
-Since the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}.
-Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values.
-Therefore, we also define loop invariants for such these cases as follows:
+The \lstinline{traverse} function stops when it reaches the base case, which happens when the current node is \lstinline{NULL}.
+Alternatively, the function terminates if \lstinline{findNext} fails to identify a subsequent node with the key value $\texttt{k}$ within its set of key values.
+In addition to the \texttt{traverse\_inv} we defined earlier, we also need to establish loop invariants for these particular situations as follows:
 \begin{align*} 
@@
 
-More specifically, $\texttt{traverse\_inv\_1 (pn)}$ is true when the \lstinline{traverse} function successfully locates the subsequent node during traversal, as indicated by the condition $\texttt{pn->n'->t} = \texttt{NULL}$.
+More specifically, $\texttt{traverse\_inv\_1}$ is true when the \lstinline{traverse} function successfully locates the subsequent node during traversal, as indicated by the condition $\texttt{pn->n'->t} = \texttt{NULL}$.
 In the context of a linked-list data structure, a \lstinline{t} field with a value of \lstinline{NULL} signals that the \lstinline{traverse} function has arrived at the list's final node.
-
-In contrast, $\texttt{traverse\_inv\_2 (pn)}$ is true when the \lstinline{traverse} function fails to locate the next node during traversal, which means $\texttt{pn->n'->t} \neq \texttt{NULL}$.
+In contrast, $\texttt{traverse\_inv\_2}$ is true when the \lstinline{traverse} function fails to locate the next node during traversal, which means $\texttt{pn->n'->t} \neq \texttt{NULL}$.
 For instance, when trying to insert a new node with a specific key \lstinline{k} into a linked list, if the \lstinline{traverse} function cannot identify the next node, it implies that the current node has a target key equal to its value.
@@
 
-$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->p->lock,R'}) \ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
+$$\forall \texttt{t}.\ \left\langle  \infp (\texttt{pn->n}) \ | \ \treerep\ \texttt{t} \right\rangle$$
 $$\texttt{traverse(pn, k)}$$
-$$\left\langle \texttt{n'}.\ \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->n'->lock,R')} \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
+$$\left\langle \texttt{res} \ \texttt{n'}.
+\ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast  \infp (\texttt{pn->n'}) \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
 \ |\ \treerep\ \texttt{t} \right\rangle$$
+
+Compared to the more complex specification of the lock-coupling style \texttt{traverse} function, the specification of the give-up template \texttt{traverse} function is simpler.
+The local pre-condition for the specification only includes $\infp (\texttt{pn->n})$, a footprint predicate indicating that node \texttt{n} is within the search structure's footprint.
+The \texttt{traverse} function then operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.
+Consequently, the local post-condition is expected to assert that \texttt{n'} is also situated within the footprint predicate and that the resource \texttt{R(pn->n')} contains the concrete information about node \texttt{n'}.
 
@@
 We describe the mechanized formal verification of the correctness of such these templates, implemented in C and verified with the VST, using concurrent separation
-logic (CSL) to reason about the behavior of the C code, and integrated Iris-style logical atomicity into VST. All proofs are formalized in Coq using VST, and can be found online at ...
+logic (CSL) to reason about the behavior of the C code, and integrated Iris-style logical atomicity into VST. All proofs are formalized in Coq using VST, and can be found online at
+
+\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}
 

### 2023-04-14 concurrency/CONCUR 2023/paper.tex
writing proof of give-up traverse

@@
 $$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn->n)} 
-%\ast \lockinv \texttt{(pn->p->lock,R'}) 
 \ast \texttt{R(pn->n)} \ | \ \treerep\ \texttt{t} \right\rangle $$
@@
 $$\left\langle \ \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn->n')}
-%ast \lockinv \texttt{(pn->n'->lock,R')} 
 \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
@@
 This resource, denoted as $\texttt{R(pn->n)}$, is parameterized by a node pointer $\texttt{pn->n}$ and describes the concrete information about the node.
+In the specification, the local pre-condition encompasses $\nodeboxrep(\texttt{pn->n})$ and resource $\texttt{R(pn->n)}$.
+Upon operating on the entire search structure, the \texttt{traverse} function yields a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.
+The local post-condition is expected to contain $\nodeboxrep(\texttt{pn->n'})$, resource $\texttt{R(pn->n)}$, and a Boolean \texttt{res} that indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
 
@@
 
-Compared to the more complex specification of the lock-coupling style \texttt{traverse} function, the specification of the give-up template \texttt{traverse} function is simpler.
-The local pre-condition for the specification only includes $\infp (\texttt{pn->n})$, a footprint predicate indicating that node \texttt{n} is within the search structure's footprint.
-The \texttt{traverse} function then operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.
-Consequently, the local post-condition is expected to assert that \texttt{n'} is also situated within the footprint predicate and that the resource \texttt{R(pn->n')} contains the concrete information about node \texttt{n'}.
+In contrast to the more intricate specification of the lock-coupling style \texttt{traverse} function, the give-up template \texttt{traverse} function's specification is comparatively simpler.
+The local pre-condition for the specification solely comprises $\infp (\texttt{pn->n})$, a footprint predicate signifying that node \texttt{n} resides within the search structure's footprint.
+Subsequently, the \texttt{traverse} function operates on the entire search structure, returning a subsequent node \texttt{n'}, where \texttt{k} lies within the node \texttt{n'}'s range.
+As a result, the local post-condition is anticipated to assert that \texttt{n'} is also positioned within the footprint predicate, the resource \texttt{R(pn->n')} embodies the concrete information concerning node \texttt{n'}, and a Boolean \texttt{res} indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
 

### 2023-04-14 concurrency/CONCUR 2023/paper.tex
re-structured paper

@@
 \section{Search Structure Templates in VST}
+
+\subsection{Search Structure Templates}
 The C implementation of a node in our template style is:
@@
 
-\subsection{Lock-coupling Template}
+\subsubsection{Lock-coupling Template}
 The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal.
@@
 		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
-		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional} \wm{Shouldn't it be lock and then unlock?}}
-		\than{I copied exactly the same from thesis of Krishna. However, in the their of proofs, they had lock before unlock like you said. Do we need to changed to fit with our context and their code?} 
+		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
+			%\wm{Shouldn't it be lock and then unlock?}} \than{I copied exactly the same from thesis of Krishna.
+However, in the their of proofs, they had lock before unlock like you said.
+Do we need to changed to fit with our context and their code?}
 		\label{traverse_lock_a}	
@@
 
-\subsection{Give-Up}
+\subsubsection{Give-Up Template}
 The give-up template is an algorithm for implementing concurrent search in a tree data structure.
@@
 
-\subsection{Templates vs. Internal Reorganization} % move to section 4?
-In the binary search tree, \texttt{pushdown\_left} involves both the structure of the tree and the concurrency protocol.
-There's no good way to disentangle this into a template and a concurrency-free data structure operation.
-
-\than{Or can we move this section to Proof mechanization, where we talk on BST speficially.}
-
-\section{Verified Data Structures}
-
-\wm{I think this is the wrong organization.
-The proofs of traverse should be presented in the previous section, where we describe traverse.
-This section should show findNext and insert/lookup for specific data structure(s), and then describe how we combine them into top-level specs.}
+\subsection{Proofs of Templates} \than{Can someone suggest another title?}
 
@@
 
-\subsection{Lock-coupling Template}
+\subsubsection{Lock-coupling Template}
 Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template:
@@
 
-\subsection{Give-up Template}
+\subsubsection{Give-up Template}
 Moving forward with the proofs, we define the following specification for the traverse function within the give-up template:
@@
 As a result, the local post-condition is anticipated to assert that \texttt{n'} is also positioned within the footprint predicate, the resource \texttt{R(pn->n')} embodies the concrete information concerning node \texttt{n'}, and a Boolean \texttt{res} indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
+
+\subsection{Templates vs. Internal Reorganization} % move to section 4?
+In the binary search tree, \texttt{pushdown\_left} involves both the structure of the tree and the concurrency protocol.
+There's no good way to disentangle this into a template and a concurrency-free data structure operation.
+
+\than{Or can we move this section to Proof mechanization, where we talk on BST speficially.}
+
+\section{Verified Data Structures}
+
+\wm{I think this is the wrong organization.
+The proofs of traverse should be presented in the previous section, where we describe traverse.
+This section should show findNext and insert/lookup for specific data structure(s), and then describe how we combine them into top-level specs.}
 

### 2023-04-15 concurrency/CONCUR 2023/paper.tex
fixed proof of lock-coupling traverse

@@
 \newcommand{\islock}{\boxdotright}
-\newcommand{\lockvar}{\islock^{v}}
+\newcommand{\lockvar}{\islock}
 \newcommand{\isaex}{\!\mathrel\odot\joinrel\rightarrow\!}
@@
 Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template:
-
-\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}
-
+%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
 $$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn->n)} 
@@
 $$\left\langle \ \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn->n')}
-\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
+\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
 \   |\ \treerep\ \texttt{t} \right\rangle$$
@@
 As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{node_t}: the $\texttt{p}$ field, which signifies the parent node, and the $\texttt{n}$ field, representing the current node.
+
+We use the assertion $\ell \lockvar \texttt{R}$ to associate a resource invariant \texttt{R} with a lock location $\ell$.
+Thus, formally, we can write $\nodeboxrep \texttt{(pn->n)} = \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$.
 Remember that in the lock-coupling style, each thread consistently maintains at least one lock during traversal between nodes, ensuring other threads cannot disrupt the traversal or invalidate the ongoing search.
@@
 This resource, denoted as $\texttt{R(pn->n)}$, is parameterized by a node pointer $\texttt{pn->n}$ and describes the concrete information about the node.
+
 In the specification, the local pre-condition encompasses $\nodeboxrep(\texttt{pn->n})$ and resource $\texttt{R(pn->n)}$.
@@
 \end{align*}
-The \lstinline{traverse} function stops when it reaches the base case, which happens when the current node is \lstinline{NULL}.
-Alternatively, the function terminates if \lstinline{findNext} fails to identify a subsequent node with the key value $\texttt{k}$ within its set of key values.
-In addition to the \texttt{traverse\_inv} we defined earlier, we also need to establish loop invariants for these particular situations as follows:
-\begin{align*} 
-	\texttt{traverse\_inv\_1(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast \texttt{R(pn->n')} \ast (\texttt{pn->n'->t} = \texttt{NULL})
-	\\
-	\texttt{traverse\_inv\_2(pn)} \triangleq\  \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast \texttt{R(pn->n')} \ast (\texttt{pn->n'->t} \neq \texttt{NULL})
-\end{align*}
+The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop.
+The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 6 in Figure \ref{traverse_lock_b}).
+At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition.
+On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (refer to line 11 in Figure \ref{traverse_lock_b}).
+In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.
 
-More specifically, $\texttt{traverse\_inv\_1}$ is true when the \lstinline{traverse} function successfully locates the subsequent node during traversal, as indicated by the condition $\texttt{pn->n'->t} = \texttt{NULL}$.
-In the context of a linked-list data structure, a \lstinline{t} field with a value of \lstinline{NULL} signals that the \lstinline{traverse} function has arrived at the list's final node.
-In contrast, $\texttt{traverse\_inv\_2}$ is true when the \lstinline{traverse} function fails to locate the next node during traversal, which means $\texttt{pn->n'->t} \neq \texttt{NULL}$.
-For instance, when trying to insert a new node with a specific key \lstinline{k} into a linked list, if the \lstinline{traverse} function cannot identify the next node, it implies that the current node has a target key equal to its value.
-Since the node holds the key, the \lstinline{t} field should not be \lstinline{NULL} .
-
-\than{FIX ME: These things above didn't represent hand-over-hand locking instance, since I tried to avoid talking about lock invariant.}
 %The loop invariants mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop.
@@
 
-The proof begins with $\nodeboxrep \texttt{(pn->n)}$, which comprises a pointer to the node's lock, the lock-invariant assertion of $\texttt{(pn->n)}$, and the resource $\texttt{R(pn->n)}$ acquired before invoking the \lstinline{traverse} function.
-When \lstinline{findNext} successfully detects the next node (e.g., \lstinline{n'}) to visit, the lock invariant within $\nodeboxrep$ allows us to access the resource $\texttt{R(pn->n')}$ at line 14 in the C implementation of the lock-coupling template (Figure \ref{traverse_lock_b}).
-Subsequently, we release the lock of the current node (line 15 in Figure \ref{traverse_lock_b}), demonstrating our consistent approach to maintaining a single lock during node traversal.
-The proof concludes by verifying that the \lstinline{traverse} function satisfies the post-condition of the specification mentioned earlier, owing to loop invariants \texttt{traverse\_inv\_1} and \texttt{traverse\_inv\_2}.
+The proof begins with $\nodeboxrep \texttt{(pn->n)}$, which comprises a lock invariant $\texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$, and a resource $\texttt{R(pn->n)}$ acquired before invoking the \lstinline{traverse} function.
+After \lstinline{findNext} successfully detects the next node (for example, \texttt{n'}) to visit, we acquire the lock of node \texttt{n'} (see line 14 in Figure \ref{traverse_lock_b}).
+It means we can get the resource $\texttt{R(pn->n')}$ from $\texttt{pn->n'->lock} \lockvar \texttt{R(pn->n')}$.
+Subsequently, we release the lock of the current node \texttt{p}.
+It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
+By acquiring the lock for the next node \texttt{n'} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it (hand-over-hand locking).
+The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
 
+\than{FIX ME: These things above didn't represent hand-over-hand locking instance, since I tried to avoid talking about lock invariant.}
 \than{I feel somewhat perplexed while attempting to articulate the situation mentioned above. Can someone help me fix or explain it more?}

### 2023-04-15 concurrency/CONCUR 2023/paper.tex
WIP

@@
 
-For the purpose of reasoning about lock-coupling style (also known as hand-over-hand locking), we specifically consider traversing the tree using this locking method.
+%For the purpose of reasoning about lock-coupling style (also known as hand-over-hand locking), we specifically consider traversing the tree using this locking method.
 To accomplish this, we must maintain at least one lock during traversal between nodes.
@@
 expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution.
+The stack itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the \texttt{is\_stack} assertion.
+
+To demonstrate the proof of an implementation that meets the atomic specification, let's take the example of the stack specification mentioned above.
+This specification assert that the implementation \texttt{push} appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states.
+The local preconditions and postconditions, \texttt{is\_stack}, contain the points-to and lock-invariant assertions necessary for a thread, while the public preconditions and postconditions, \texttt{stack}, involve the ghost state that defines the overall state of the stack.
+Ghost state is a type of program state that aids in the verification process without affecting the program's runtime behavior, and by referring to an arbitrary ghost state identifier, it is possible to connect local and public conditions.
+To show that the implementation of \texttt{push}(v) satisfies these atomic specifications, two things must be demonstrated.
+Firstly, the implementation can execute safely without owning any piece of public assertion $\texttt{stack}$, only accessing it automatically and restoring it up until the linearization point.
+Secondly, at some linearization point, the implementation transforms the current stack $s$ into $v::s$ atomically, which satisfies the public postcondition $\texttt{stack}(v::s)$.
 
@@
 
-The particulars of $\nodeboxrep$ and $\treerep$ might differ between implementations, but at their core, $\nodeboxrep$ encompasses the resources tied to each client thread, while $\treerep$ acts as an abstract representation containing ghost state that connects the data structure's concrete state to its abstract counterpart.
+The particulars of $\nodeboxrep$ and $\treerep$ might differ between implementations, but at their core, $\nodeboxrep$ encompasses the resources tied to each client thread, while $\treerep$ acts as an abstract representation containing ghost state that connects the data structure's concrete state to its abstract counterpart and is treated as an atomically accessed shared resource.
 Specifically, $\nodeboxrep(\texttt{pn->n})$ functions as a pointer to both the node's lock and lock-invariant assertions, representing a protected resource through an arbitrary separation logic predicate.
@@
 
-\than{FIX ME: These things above didn't represent hand-over-hand locking instance, since I tried to avoid talking about lock invariant.}
+%\than{FIX ME: These things above didn't represent hand-over-hand locking instance, since I tried to avoid talking about lock invariant.}
 \than{I feel somewhat perplexed while attempting to articulate the situation mentioned above. Can someone help me fix or explain it more?}
@@
 \section{Verified Data Structures}
+\subsection{Lock-coupling Template}
 
@@
 This section should show findNext and insert/lookup for specific data structure(s), and then describe how we combine them into top-level specs.}
+\subsection{Give-up Template}
 

### 2023-04-15 concurrency/CONCUR 2023/paper.tex
add insert of give-up and lock-coupling to the verified data structures sections

@@
 To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
-Additionally, we require a target key \lstinline{k} and a dummy node value \lstinline{value} (referred to as 'dummy' because it is not actually used in \lstinline{traverse}).
-We include this dummy value to maintain consistency with other implementations \wm{What other implementations?}.
 Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findNext}.
@@
 
-The particulars of $\nodeboxrep$ and $\treerep$ might differ between implementations, but at their core, $\nodeboxrep$ encompasses the resources tied to each client thread, while $\treerep$ acts as an abstract representation containing ghost state that connects the data structure's concrete state to its abstract counterpart and is treated as an atomically accessed shared resource.
-Specifically, $\nodeboxrep(\texttt{pn->n})$ functions as a pointer to both the node's lock and lock-invariant assertions, representing a protected resource through an arbitrary separation logic predicate.
-Acquiring this lock enables a thread to access further elements within the data structure.
-As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{node_t}: the $\texttt{p}$ field, which signifies the parent node, and the $\texttt{n}$ field, representing the current node.
+The specifics of $\nodeboxrep$ and $\treerep$ may vary across implementations, but fundamentally, $\nodeboxrep$ encompasses resources associated with each client thread, while $\treerep$ serves as an abstract representation containing ghost state that links the data structure's concrete state to its abstract equivalent, treated as an atomically accessed shared resource.
+In particular, $\nodeboxrep(\texttt{pn->n})$ acts as a pointer to both the node's lock and lock-invariant assertions, representing a protected resource through an arbitrary separation logic predicate.
+Acquiring this lock allows a thread to access additional elements within the data structure.
 
+As previously mentioned, the \lstinline{pn} struct includes two nodes of type \lstinline{node_t}: the $\texttt{p}$ field, signifying the parent node, and the $\texttt{n}$ field, representing the current node.
 We use the assertion $\ell \lockvar \texttt{R}$ to associate a resource invariant \texttt{R} with a lock location $\ell$.
-Thus, formally, we can write $\nodeboxrep \texttt{(pn->n)} = \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$.
-Remember that in the lock-coupling style, each thread consistently maintains at least one lock during traversal between nodes, ensuring other threads cannot disrupt the traversal or invalidate the ongoing search.
-Thus, before invoking the \texttt{traverse} function, the thread must acquire a lock for the current node $\texttt{pn->n}$ and obtain the resource protected by the lock.
-This resource, denoted as $\texttt{R(pn->n)}$, is parameterized by a node pointer $\texttt{pn->n}$ and describes the concrete information about the node.
+Thus, formally, we can express $\nodeboxrep \texttt{(pn->n)} = \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$.
 
-In the specification, the local pre-condition encompasses $\nodeboxrep(\texttt{pn->n})$ and resource $\texttt{R(pn->n)}$.
-Upon operating on the entire search structure, the \texttt{traverse} function yields a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.
-The local post-condition is expected to contain $\nodeboxrep(\texttt{pn->n'})$, resource $\texttt{R(pn->n)}$, and a Boolean \texttt{res} that indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
+Keep in mind that in the lock-coupling style, each thread consistently retains at least one lock while traversing between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
+Consequently, before calling the \texttt{traverse} function, the thread must acquire a lock for the current node $\texttt{pn->n}$ and obtain the resource safeguarded by the lock.
+This resource, represented as $\texttt{R(pn->n)}$, is parameterized by a node pointer $\texttt{pn->n}$ and describes the node's concrete information.
 
-%The lock invariant $\lockinv$ predicate asserts that a lock exists to protect the resources.
+As mentioned earlier, we can define a piece of ghost state, $\texttt{ghost\_node t}$, connecting the node in memory to the public abstract state of the data structure.
+Specifically, we can have one half of the ghost state contained in the resource $\texttt{R(pn->n)}$, and the other half in the abstract state $\texttt{Node t}$.
+When proving the atomic triple above, at the linearization point where we access $\texttt{Node}$, we combine both halves of the ghost state, update them to reflect the new state of the concrete data structure, and then fulfill the atomic postcondition with one half of the ghost state, returning the other half to the lock invariant.
 
-%The specification states that if the node \texttt{n} is within the footprint of the search structure, then the \texttt{traverse} function operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.
+In the specification, the local pre-condition includes both the representation of the node $\nodeboxrep(\texttt{pn->n})$ and the resource $\texttt{R(pn->n)}$.
+When the \texttt{traverse} function operates on the entire search structure, it generates a subsequent node, \texttt{n'}, such that the key \texttt{k} falls within the range of node \texttt{n'}.
+The local post-condition is expected to encompass the representation of the new node $\nodeboxrep(\texttt{pn->n'})$, the resource $\texttt{R(pn->n)}$, and a Boolean variable, \texttt{res}, which indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
+The key to verifying the \texttt{traverse} function is identifying the loop invariant for the top-level loop.
+We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
 
-The crux of verifying the \texttt{traverse} function lies in the loop invariant for the top-level loop. We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
 \begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast  \texttt{R(pn->n')}   
@@
 \section{Verified Data Structures}
+In this section, we show how to birng together all into top-level specifications in order to verify two template algorithms, lock-coupling and give-up, in real C programming language.
+
 \subsection{Lock-coupling Template}
+\begin{figure}[ht]
 
-\wm{I think this is the wrong organization.
+	\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
+	\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
+	\label{insert_lock}	
+\end{figure}
+
+%\wm{I think this is the wrong organization.
 The proofs of traverse should be presented in the previous section, where we describe traverse.
@@
 \subsection{Give-up Template}
+
+\begin{figure}[ht]
+	\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
+	\caption{The \lstinline{insert} method of the give-up template algorithm}
+	\label{insert_giveup}
+\end{figure}
 

### 2023-04-15 concurrency/CONCUR 2023/giveup_insert.c

@@ -1,0 +1,12 @@
+void insert (node_t **t, int k, void *value){
+  struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+  pn->n = *t;
+  if (traverse(pn, k) == 0){
+    pn->p->t->value = value;
+  }
+  else{
+    insertOp(pn, k, value);
+  }
+  release(pn->n->lock);
+  free(pn);
+}

### 2023-04-15 concurrency/CONCUR 2023/giveup_traverse.c

@@ -1,4 +1,4 @@
-int traverse(pn *pn, int k, void *value){
+int traverse(pn *pn, int k){
   tree_t *p = (pn->n); 
   int flag = 1;
   for( ; ; ){
@@ -8,7 +8,7 @@ int traverse(pn *pn, int k, void *value){
       if (pn->p->t == NULL)
         break;
       else{
-        int b = findNext(pn, k, value);
+        int b = findNext(pn, k);
         if (b == 0){
           flag = 0;
           break;

### 2023-04-15 concurrency/CONCUR 2023/lock_insert.c

@@ -1,0 +1,13 @@
+void insert (node_t **t, int k, void *value){
+  struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+  pn->n = *t;
+  acquire(pn->n->lock);  
+  if (traverse(pn, k) == 0){
+    pn->p->t->value = value;
+  }
+  else{
+    insertOp(pn, k, value);
+  }
+  release(pn->n->lock);
+  free(pn);
+}

### 2023-04-15 concurrency/CONCUR 2023/lock_traverse.c

@@ -1,11 +1,11 @@
-int traverse(pn *pn, int k, void *value){
+int traverse(pn *pn, int k){
   int flag = 1;
   for( ; ; ){
     pn->p = pn->n;
     if (pn->p->t == NULL)
       break;
     else{
-      int b = findNext(pn, k, value);
+      int b = findNext(pn, k);
       if (b == 0){
         flag = 0;
         break;

### 2023-04-16 concurrency/CONCUR 2023/paper.tex
WIP give-up template

@@
 $$\left\langle \texttt{res} \ \texttt{n'}.
-\ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast  \infp (\texttt{pn->n'}) \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
+\ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast  \infp (\texttt{pn->n'}) \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
 \ |\ \treerep\ \texttt{t} \right\rangle$$
@@
 In contrast to the more intricate specification of the lock-coupling style \texttt{traverse} function, the give-up template \texttt{traverse} function's specification is comparatively simpler.
-The local pre-condition for the specification solely comprises $\infp (\texttt{pn->n})$, a footprint predicate signifying that node \texttt{n} resides within the search structure's footprint.
+The local pre-condition for the specification solely comprises $\infp(\texttt{pn->n})$, a footprint predicate signifying that node \texttt{n} reside within the search structure's footprint.
 Subsequently, the \texttt{traverse} function operates on the entire search structure, returning a subsequent node \texttt{n'}, where \texttt{k} lies within the node \texttt{n'}'s range.
 As a result, the local post-condition is anticipated to assert that \texttt{n'} is also positioned within the footprint predicate, the resource \texttt{R(pn->n')} embodies the concrete information concerning node \texttt{n'}, and a Boolean \texttt{res} indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
+
+We define the loop invariant for the give-up template of \texttt{traverse} function as follows: 
+\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
+\end{align*}
+
+The proof starts with $\infp (\texttt{pn->n})$ predicate from the local pre-condition.
+We get the resource $\texttt{R(pn->n)}$ into the thread-local state after acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}).
+In the \texttt{then} of \texttt{inRange}, this give us the additional predicate $\texttt{k} \in \texttt{range(n)}$.
+Similar to proof of \texttt{traverse} of lock-coupling template, it stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 8 in Figure \ref{traverse_giveup_b}).
+At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition.
+On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values.
+Otherwise, the \lstinline{findNext} successfully detects the next node (for example, \texttt{n'}) to visit, then it give a predicate that asserts \texttt{n'} is in the footprint of structure, $\infp(\texttt{pn->n})$.
+From that, after releasing the lock of node the current node \texttt{p} (since \texttt{p} and  \texttt{n} point to the same memory address - line 6 in Figure \ref{traverse_giveup_b}), we can prove it satisfies the loop invariant $\texttt{traverse\_inv(pn)}$ and complete this branch.
+
+In the \texttt{else} branch of the call to \texttt{inRange}, since \texttt{k} is not in the range of the current node \texttt{p}, it gives up, and relinguishes the lock on \texttt{p}, the goes back to the root of data structure to re-try.
 

### 2023-04-17 concurrency/CONCUR 2023/paper.tex
WIP

--- 2023-04-15 concurrency/CONCUR 2023/paper.tex
@@
 %\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
-$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn->n)} 
-\ast \texttt{R(pn->n)} \ | \ \treerep\ \texttt{t} \right\rangle $$
-$$\texttt{traverse(pn, k)}$$
-$$\left\langle \ \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn->n')}
-\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
-\   |\ \treerep\ \texttt{t} \right\rangle$$
+\begin{mathpar}
+	{\color{blue} {\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn->n)} 
+	\ast \texttt{R(pn->n)} \ | \ \treerep\ \texttt{t} \right\rangle}}\\ 
+	\texttt{traverse(pn, k)} \\ 
+	\color{blue} {\langle \ \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn->n')} \ast }
+	\\ 
+	\color{blue} {\texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
+	\   |\ \treerep\ \texttt{t} \rangle}
+\end{mathpar}
 
-\than{Actually, $\nodeboxrep \texttt{(pn->n)}$ and $\texttt{data\_at  n pn} \ast \texttt{my\_half g\_root} (-\inf, +\inf)$ is equivalent to $\texttt{tree\_rep}$.
+%\than{Actually, $\nodeboxrep \texttt{(pn->n)}$ and $\texttt{data\_at  n pn} \ast \texttt{my\_half g\_root} (-\inf, +\inf)$ is equivalent to $\texttt{tree\_rep}$.
 But now we can ignore the information of $\texttt{my\_half}$.
@@
 
-$$\forall \texttt{t}.\ \left\langle  \infp (\texttt{pn->n}) \ | \ \treerep\ \texttt{t} \right\rangle$$ 
-$$\texttt{traverse(pn, k)}$$
-$$\left\langle \texttt{res} \ \texttt{n'}.
-\ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast  \infp (\texttt{pn->n'}) \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
-\ |\ \treerep\ \texttt{t} \right\rangle$$
+\begin{mathpar}
+	{\color{blue}{\forall \texttt{t}.\ \left\langle \infp (\texttt{pn->n}) \ | \ \treerep\ \texttt{t} \right\rangle}} \\ 
+	\texttt{traverse(pn, k)} \\ 
+	\color{blue}{\langle \texttt{res} \ \texttt{n'}.
+\ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast  \infp (\texttt{pn->n'}) \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
+	\ |\ \treerep\ \texttt{t}\rangle}
+\end{mathpar}
 
-In contrast to the more intricate specification of the lock-coupling style \texttt{traverse} function, the give-up template \texttt{traverse} function's specification is comparatively simpler.
-The local pre-condition for the specification solely comprises $\infp (\texttt{pn->n})$, a footprint predicate signifying that node \texttt{n} resides within the search structure's footprint.
-Subsequently, the \texttt{traverse} function operates on the entire search structure, returning a subsequent node \texttt{n'}, where \texttt{k} lies within the node \texttt{n'}'s range.
-As a result, the local post-condition is anticipated to assert that \texttt{n'} is also positioned within the footprint predicate, the resource \texttt{R(pn->n')} embodies the concrete information concerning node \texttt{n'}, and a Boolean \texttt{res} indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
+In contrast to the more complex specification of the lock-coupling style \texttt{traverse} function, the give-up template's \texttt{traverse} function has a comparatively simpler specification.
+The local pre-condition for the specification solely comprises $\infp(\texttt{pn->n})$, which is a footprint predicate signifying that node \texttt{n} resides within the search structure's footprint.
+Subsequently, the \texttt{traverse} function operates on the entire search structure, returning a subsequent node \texttt{n'} where \texttt{k} lies within the node \texttt{n'}'s range.
+Therefore, the local post-condition is expected to assert that \texttt{n'} is also positioned within the footprint predicate, that the resource \texttt{R(pn->n')} embodies the concrete information concerning node \texttt{n'}, and that a Boolean \texttt{res} indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
+
+We have defined the loop invariant for the given template of the \texttt{traverse} function as follows: 
+\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
+\end{align*}
+
+To begin the proof, we start with the $\infp (\texttt{pn->n})$ predicate from the local pre-condition.
+After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\texttt{R(pn->n)}$ and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}.
+Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 8 in Figure \ref{traverse_giveup_b}).
+At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.
+
+If \lstinline{findNext} cannot find a subsequent node with the key value $\texttt{k}$ within its set of key values, the function terminates.
+Otherwise, it successfully detects the next node (for example, \texttt{n'}) to visit, giving a predicate that asserts \texttt{n'} is in the footprint of the structure, $\infp(\texttt{pn->n})$.
+After releasing the lock of the current node \texttt{p} (since \texttt{p} and \texttt{n} point to the same memory address - line 6 in Figure \ref{traverse_giveup_b}), we can prove that it satisfies the loop invariant $\texttt{traverse\_inv(pn)}$ and complete this branch of the proof.
+In the \texttt{else} branch of the call to \texttt{inRange}, since \texttt{k} is not in the range of the current node \texttt{p}, the function gives up and relinquishes the lock on \texttt{p}, then goes back to the root of the data structure to retry.
 
@@
 \section{Verified Data Structures}
-In this section, we show how to birng together all into top-level specifications in order to verify two template algorithms, lock-coupling and give-up, in real C programming language.
+In this section, we demonstrate how to integrate all components into high-level specifications for the purpose of verifying two template algorithms, lock-coupling and give-up, in the real C programming language.
+
+\begin{mathpar}
+	\langle m.\ \nodeboxrep\ pn \ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto v])\rangle
+
+	\langle m.\ \nodeboxrep\ pn\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land m(k) = v\rangle
+
+	\langle m.\ \nodeboxrep\ pn \ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto \_])\rangle
+\end{mathpar}
 
@@
 \section{Proof Mechanization}
-We have veried the give-up and lock-coupling template algorithms.
-We describe the mechanized formal verification of the correctness of such these templates, implemented in C and verified with the VST, using concurrent separation
-logic (CSL) to reason about the behavior of the C code, and integrated Iris-style logical atomicity into VST. All proofs are formalized in Coq using VST, and can be found online at
+We describe the mechanized formal verification of the correctness of such these templates, implemented in C and verified with the VST, using concurrent separation logic (CSL) to reason about the behavior of the C code, and integrated Iris-style logical atomicity into VST.
+All proofs are formalized in Coq using VST, and can be found online at
 

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
WIP
initial writing for insert of lock-coupling

@@
 If it fails to do so, it returns false, which other functions can use to make decisions.
-
 For instance, in a binary search tree, a node with a value of 42 would have a left child with a set of key values consisting of values less than 42 and a right child with one greater than 42.
@@
 When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $\texttt{k}$ that equals the node's value.
-At this point, we only have to modify the value of that node to \lstinline{value}.
 
@@
 The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
+\than{Could you write more things here to explain why we need inRange.}
 The purpose of the \lstinline{inRange} function is to determine whether the key value $\texttt{k}$ falls within the range of node $\texttt{n}$.
 If the key value is outside the node's range, the search is terminated, and the thread is restarted.
+To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node.
+Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{node_t} in our C implementation.
+These fields store the lower and upper bounds of the keys, respectively.
+In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys.
+This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes.
+If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
 
-\wm{Describe why we need extra pieces in the struct, and how this relates to the other version.}
+%\wm{Describe why we need extra pieces in the struct, and how this relates to the other version.}
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
@@
 %% C implementation
-To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node.
-Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{node_t} in our C implementation.
-These fields store the lower and upper bounds of the keys, respectively.
+
 When a thread arrives at a node in search of the key $\texttt{k}$, it checks whether $\texttt{k}$ falls within the range of that node.
@@
 \subsubsection{Lock-coupling Template}
+\label{traverse_proof_lock}
 Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template:
@@
 \begin{mathpar}
-	\langle m.\ \nodeboxrep\ pn \ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto v])\rangle
+	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle\ \texttt{insert(p, k, v)}\ \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}\rangle
 
-	\langle m.\ \nodeboxrep\ pn\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land m(k) = v\rangle
+	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \rangle\ \texttt{lookup(p, k)}\ \langle \texttt{v}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \land \texttt{m(k) = v}\rangle
 
-	\langle m.\ \nodeboxrep\ pn \ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto \_])\rangle
+	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle\ \texttt{delete(p, k)}\ \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{(m[k} \mapsto \_ \texttt{])}\rangle
 \end{mathpar}
@@
 \end{figure}
+
+The \texttt{insert} method in Figure \ref{insert_lock} is the main method of lock-coupling template.
+It makes use of an auxiliary \texttt{traverse}  mentioned in Figure \ref{traverse_lock_b}.
+If \texttt{traverse} fails to find the next node (either to left or right), it implies that it has reached a node with a target key \texttt{k} that equals the node's value.
+At this point, we only have to modify the value of that to \texttt{value} (lines 4-5 in Figure \ref{insert_lock}).
+Otherwise, if \texttt{traverse} reaches a leaf, it allocates a new node with key \texttt{k} and value \texttt{v} by calling \texttt{insertOp(pn, k, v)} (line 9 in Figure \ref{insert_lock}).
+
+Before moving on to the proof of \texttt{insert} operations for binary search tree, we present the specifications of helper functions, \texttt{findNext} is used in \texttt{traverse} function, and \texttt{insertOp} is used in \texttt{insert} function.
+They are all Hoare triples that operate on the local $\nodeboxrep$ predicate, meaning their proofs need not reason about concurrency.
+
+\begin{mathpar}
+	% spec of findNext
+	{\color{blue}\{ \  \texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} \ \}}
+	\\ 
+	\texttt{findNext(pn, k)} 
+	\\
+	{\color{blue} \{ \texttt{res n'.} \ \texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)}  \ast }
+	\\ 
+	{\color{blue} \texttt{if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} 
+	\texttt{ else } \texttt{(n' = n} \land \texttt{k = k')}  \}}
+	\\
+	\\
+	%spec of insertOp
+	{\color{blue} \left\{ \texttt{pn->n} \ast \texttt{n->t}  \right\}  } 
+	\\
+	\texttt{insertOp(pn, k, v)}
+	\\
+	{\color{blue} \{\texttt{t' l r lk1 lk2}. \ \ \texttt{pn->n} \ast \texttt{n->t'} \ast \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ast }
+	\\
+	{\color{blue} \texttt{l} \mapsto \texttt{(NULL, lk1)} \ast \texttt{r} \mapsto \texttt{(NULL, lk2)}
+	\} }
+\end{mathpar}
+
+The specification of \texttt{findNext} states that a non-null \texttt{t} field points to a \texttt{node} struct containing the current node's key  \texttt{k'}, the node's value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
+When \texttt{findNext} sucessfully locates the next node \texttt{n'}, it sets \texttt{n'} as the left child node \texttt{l} if the provided key \texttt{k} is less than the current node's key \texttt{k'}.
+Conversely, if the given key \texttt{k} is greater than the current node's key \texttt{k'}, the next node \texttt{n'} is set as the right child node \texttt{r}.
+In case where \texttt{findNext} is unable to identify the next node \texttt{n'}, it retains the current node \texttt{n} as \texttt{n'},  indicating no change and that the provided key \texttt{k} is  identical to the current node's key \texttt{k'}.
+
+The \texttt{insertOp} specification indicates that once \texttt{traverse} successfully reaches a leaf node (either the left or right node), the leaf node will have its \texttt{t} field set to \texttt{NULL}.
+This signifies that the leaf node currently has no key or value.
+At this location, a new node with the specified key \texttt{k} and value \texttt{v} should be inserted.
+Upon locating the appropriate leaf node for adding the new key-value pair \texttt{(k, v)}, the process generates two new empty child nodes with corresponding locks, \texttt{(NULL, lk1)} and \texttt{(NULL, lk2)}, which are pointed to by the \texttt{l} and \texttt{r} of the current node.
+
+The atomic specification for \texttt{insert} method is
+\begin{mathpar}
+	{\color{blue}\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle } \ \ \texttt{insert(p, k, v)}\ 
+	{\color{blue}\langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}\rangle }
+\end{mathpar}
+
+Similar to the \texttt{traverse} specification, the \texttt{insert} function's specification is composed of two components: local and public pre- and postconditions.
+The local pre- and postconditions, containing $\nodeboxrep \ \texttt{p}$, describe the binary search tree's concrete state with the root node $\texttt{p}$, which points to the node's lock and lock invariant assertion.
+On the other hand, the public pre- and postconditions are represented by the predicate $\treerep \ \texttt{m}$, illustrating the binary search tree's abstract state as a collection of ghost states, with each state providing information about an individual node in the tree.
+Here, $\texttt{m}$ and $\texttt{m[k} \mapsto \texttt{v]}$ denote the tree's abstract states before and after the function's execution, respectively.
+
+The proof for \texttt{insert} begins with preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$.
+We set \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_lock}).
+After the lock for the root node is acquired (line 4 in Figure \ref{insert_lock}), the thread accesses information within the root node's lock invariant, \texttt{R(pn->n)}.
+As seen in Section \ref{traverse_proof_lock}, we supply the same precondition for the \texttt{traverse} function, which includes local precondition $\nodeboxrep \texttt{(pn->n)} \ast \texttt{R(pn->n)}$ and public precondition $\treerep\ \texttt{m}$.
+We then apply the \texttt{traverse} specification from Section \ref{traverse_proof_lock}.
+
+In the \texttt{then} branch, when the \texttt{traverse} fails to find the next node, resulting in $\texttt{res} = \texttt{false}$, it implies that $\texttt{pn->n'->t} \neq \texttt{NULL}$.
+This indicates that the key to be inserted already exists in the tree.
+In this scenario, we update the node's value while maintaining the rest of the tree structure (line 6 in Figure \ref{insert_lock}); this acts as the linearization point for the insertion.
+Conversely, when the \texttt{traverse} reaches a leaf and produces $\texttt{res} = \texttt{true}$, it implies $\texttt{pn->n'->t} = \texttt{NULL}$ (line 9 in Figure \ref{insert_lock}).
+We then utilize the \texttt{insertOp} specification, and with enough information, we can conclude that inserting the key at this location will alter the abstract state from an unknown $\texttt{m}$ to $\texttt{m[k} \mapsto \texttt{v]}$.
+In both instances, we must prove that we can atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, by demonstrating that the in-memory tree update corresponds to a local modification in the abstract tree, which inserts the new key-value pair.
 

### 2023-04-18 concurrency/CONCUR 2023/giveup_insert.c

@@ -1,11 +1,11 @@
-void insert (node_t **t, int k, void *value){
+void insert (node_t **p, int k, void *v){
   struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
-  pn->n = *t;
+  pn->n = *p;
   if (traverse(pn, k) == 0){
-    pn->p->t->value = value;
+    pn->p->t->value = v;
   }
   else{
-    insertOp(pn, k, value);
+    insertOp(pn, k, v);
   }
   release(pn->n->lock);
   free(pn);

### 2023-04-18 concurrency/CONCUR 2023/lock_insert.c

@@ -1,12 +1,12 @@
-void insert (node_t **t, int k, void *value){
+void insert (node_t **p, int k, void *v){
   struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
-  pn->n = *t;
+  pn->n = *p;
   acquire(pn->n->lock);  
   if (traverse(pn, k) == 0){
-    pn->p->t->value = value;
+    pn->p->t->value = v;
   }
   else{
-    insertOp(pn, k, value);
+    insertOp(pn, k, v);
   }
   release(pn->n->lock);
   free(pn);

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
fix breakage

@@
 For instance, the specification
-$\forall s.\ \left\langle \texttt{is\_stack}\ p\ |\ \texttt{stack}\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}\ p\ |\ \texttt{stack}\ (v::s)\right\rangle$
+$\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$
 expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution.
@@
 The local preconditions and postconditions, \texttt{is\_stack}, contain the points-to and lock-invariant assertions necessary for a thread, while the public preconditions and postconditions, \texttt{stack}, involve the ghost state that defines the overall state of the stack.
-Ghost state is a type of program state that aids in the verification process without affecting the program's runtime behavior, and by referring to an arbitrary ghost state identifier, it is possible to connect local and public conditions.
+Ghost state is a type of program state that aids in the verification process without affecting the program's runtime behavior, and by referring to an arbitrary ghost state identifier $g$ (which we will generally omit), it is possible to connect local and public conditions.
 To show that the implementation of \texttt{push}(v) satisfies these atomic specifications, two things must be demonstrated.
@@
 
+\begin{figure}
+	\centering
 	\begin{mathpar}
-	% spec of findNext
-	{\color{blue}\{ \  \texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} \ \}}
+		% spec of inRange
+		{\color{blue}
+			\left\{ 
+			\begin{array}{c}
+				\texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} 
+			\end{array}
+			\right\}
+		}
 		\\ 
@@
 		\\
-	{\color{blue} \{ \texttt{res n'.} \ \texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)}  \ast }
-	\\ 
-	{\color{blue} \texttt{if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} 
-	\texttt{ else } \texttt{(n' = n} \land \texttt{k = k')}  \}}
-	\\
-	\\
-	%spec of insertOp
-	{\color{blue} \left\{ \texttt{pn->n} \ast \texttt{n->t}  \right\}  } 
+		{\color{blue}
+			\left\{\exists \  \texttt{res n'.}
+			\begin{array}{c}
+				\ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{t} \mapsto \texttt{(k', v', l, r)} \  \ast \\ 
+				\ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n' = n} \land \texttt{k = k'))}
+			\end{array}
+			\right\}
+		}
+	\end{mathpar}
+	\caption{Specification of \texttt{findNext} for the lock-coupling and give-up templates}
+	\label{fig:findNext_lock}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\begin{mathpar}
+		% spec of inRange
+		{\color{blue}
+			\left\{ 
+			\begin{array}{c}
+				\ \texttt{pn->n} \ \ast \ \texttt{n->t} 
+			\end{array}
+			\right\}
+		}
 		\\ 
@@
 		\\
-	{\color{blue} \{\texttt{t' l r lk1 lk2}. \ \ \texttt{pn->n} \ast \texttt{n->t'} \ast \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ast }
-	\\
-	{\color{blue} \texttt{l} \mapsto \texttt{(NULL, lk1)} \ast \texttt{r} \mapsto \texttt{(NULL, lk2)}
-	\} }
+		{\color{blue}
+			\left\{\exists \  \texttt{t' l r lk1 lk2}.
+			\begin{array}{c}
+				\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
+				\ \texttt{l} \mapsto \texttt{(NULL, lk1)}\  \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2)}
+			\end{array}
+			\right\}
+		}
 	\end{mathpar}
+	\caption{Specification of \texttt{insertOp} for the lock-coupling template}
+	\label{fig:insertOp_lock}
+\end{figure}
 
-The specification of \texttt{findNext} states that a non-null \texttt{t} field points to a \texttt{node} struct containing the current node's key  \texttt{k'}, the node's value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
+The specification of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that a non-null \texttt{t} field points to a \texttt{node} struct containing the current node's key  \texttt{k'}, the node's value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
 When \texttt{findNext} sucessfully locates the next node \texttt{n'}, it sets \texttt{n'} as the left child node \texttt{l} if the provided key \texttt{k} is less than the current node's key \texttt{k'}.
@@
 
-The \texttt{insertOp} specification indicates that once \texttt{traverse} successfully reaches a leaf node (either the left or right node), the leaf node will have its \texttt{t} field set to \texttt{NULL}.
+The \texttt{insertOp} specification (Figure \ref{fig:insertOp_lock}) indicates that once \texttt{traverse} successfully reaches a leaf node (either the left or right node), the leaf node will have its \texttt{t} field set to \texttt{NULL}.
 This signifies that the leaf node currently has no key or value.
@@
 
+\begin{figure}
+	\centering
+\begin{mathpar}
+	% spec of inRange
+	{\color{blue}
+		\left\{ 
+		\begin{array}{c}
+			\texttt{pn->n} \ \ast \texttt{n->t} \ \ast \ \texttt{n->min} \ \ast \ \texttt{n->max}
+		\end{array}
+		\right\}
+	}
+	\\ 
+	\texttt{inRange(pn, k)} 
+	\\
+	{\color{blue}
+		\left\{\exists \  \texttt{res.}
+		\begin{array}{c}
+			 \ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{n->min} \ \ast \ \texttt{n->max}\  \ast \\ 
+			\ \texttt{(if res then } \texttt{(min < x} \land \texttt{x < max)} \land \texttt{(x} \leq \texttt{min} \lor \texttt{x} \geq \texttt{max))}
+		\end{array}
+		\right\}
+	}
+\end{mathpar}
+\caption{Specification of \texttt{inRange} for the give-up template}
+\label{fig:inRange_giveup}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\begin{mathpar}
+	%spec of insertOp
+	{\color{blue}
+		\left\{ 
+		\begin{array}{c}
+			\texttt{pn->n} \ \ast \texttt{n->t} \  \ast \texttt{n->min} \ \ast \texttt{n->max}
+		\end{array}
+		\right\}
+	}
+	\\
+	\texttt{insertOp(pn, k, v)}
+	\\
+	{\color{blue}
+	\left\{\exists \ \texttt{t' l r lk1 lk2}. 
+	\begin{array}{l}
+		\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\
+		\ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, x))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (x, max))}
+	\end{array}
+	\right\}
+	}
+\end{mathpar}
+\caption{Specification of \texttt{insertOp} for the give-up template}
+\label{fig:insertOp_giveup}
+\end{figure}
+
 \section{Proof Mechanization}

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
modified specs

@@
 \label{traverse_proof_lock}
-Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template:
+Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template, as illustrated in Figure \ref{fig:traverse_lock}.
 %\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}
+
+\begin{figure}[h]
+	\centering
 	\begin{mathpar}
-	{\color{blue} {\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn->n)} 
-	\ast \texttt{R(pn->n)} \ | \ \treerep\ \texttt{t} \right\rangle}}\\ 
-	\texttt{traverse(pn, k)} \\ 
-	\color{blue} {\langle \ \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn->n')} \ast }
+		% spec of inRange
+		{\color{blue}
+			 \forall \  \texttt{t}. \left\langle
+			\begin{array}{c}
+				\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{t}
+			\end{array}
+			\right\rangle
+		}
 		\\ 
-	\color{blue} {\texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
-	\   |\ \treerep\ \texttt{t} \rangle}
+		\texttt{insertOp(pn, k, v)} 
+		\\
+		{\color{blue}
+			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
+			\begin{array}{c}
+				\ \texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
+				\ast \ \texttt{R(pn->n')} \ast \ \texttt{(k} \in \texttt{range(n'))}
+			\end{array}
+			\ \Bigg| \ \treerep\ \texttt{t}
+			\right\rangle
+		}
 	\end{mathpar}
-
+	\caption{Specification of \texttt{traverse} for the lock-coupling template}
+	\label{fig:traverse_lock}
+\end{figure}
 %\than{Actually, $\nodeboxrep \texttt{(pn->n)}$ and $\texttt{data\_at  n pn} \ast \texttt{my\_half g\_root} (-\inf, +\inf)$ is equivalent to $\texttt{tree\_rep}$.
@@
 We use the assertion $\ell \lockvar \texttt{R}$ to associate a resource invariant \texttt{R} with a lock location $\ell$.
-Thus, formally, we can express $\nodeboxrep \texttt{(pn->n)} = \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$.
+Thus, formally, we can express $\nodeboxrep \texttt{(pn->n)} \triangleq\ \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$.
 
@@
 
-\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast  \texttt{R(pn->n')}   
+\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
 \end{align*}
+
 The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop.
@@
 \subsubsection{Give-up Template}
-Moving forward with the proofs, we define the following specification for the traverse function within the give-up template:
+Moving forward with the proofs, we define the following specification for the traverse function within the give-up template, as depicted in Figure \ref{fig:traverse_giveup}.
 
+\begin{figure}[h]
+	\centering
 	\begin{mathpar}
-	{\color{blue}{\forall \texttt{t}.\ \left\langle \infp (\texttt{pn->n}) \ | \ \treerep\ \texttt{t} \right\rangle}} \\ 
-	\texttt{traverse(pn, k)} \\ 
-	\color{blue}{\langle \texttt{res} \ \texttt{n'}.
-\ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast  \infp (\texttt{pn->n'}) \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
-	\ |\ \treerep\ \texttt{t}\rangle}
+		% spec of inRange
+		{\color{blue}
+			\forall \  \texttt{t}. \left\langle
+			\begin{array}{c}
+				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{t}
+			\end{array}
+			\right\rangle
+		}
+		\\ 
+		\texttt{insertOp(pn, k, v)} 
+		\\
+		{\color{blue}
+			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
+			\begin{array}{c}
+				\ \texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
+				\\ 
+				\ast \ \texttt{R(pn->n')} \ \ast \ (\texttt{k} \in \texttt{range(n'))}
+			\end{array}
+			\ \Bigg| \ \treerep\ \texttt{t}
+			\right\rangle
+		}
 	\end{mathpar}
+	\caption{Specification of \texttt{traverse} for the give-up template}
+	\label{fig:traverse_giveup}
+\end{figure}
 
@@
 We have defined the loop invariant for the given template of the \texttt{traverse} function as follows: 
-\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
+\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
 \end{align*}
@@
 
-\begin{figure}
+\begin{figure}[h]
 	\centering

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
modified specs

@@
 		{\color{blue}
-			 \forall \  \texttt{t}. \left\langle
-			\begin{array}{c}
-				\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{t}
-			\end{array}
+			 \forall \  \texttt{m}. \left\langle
+				\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{m} \
 			\right\rangle
@@
 			\end{array}
-			\ \Bigg| \ \treerep\ \texttt{t}
+			\ \Bigg| \ \treerep\ \texttt{m} \
 			\right\rangle
@@
 \begin{mathpar}
-	{\color{blue}\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle } \ \ \texttt{insert(p, k, v)}\ 
-	{\color{blue}\langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}\rangle }
+	{\color{blue}
+		\texttt{m.}\left\langle 
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m}
+		\right\rangle
+	}
+	\ \texttt{insert(p, k, v)}\ 
+	{\color{blue}
+		\left\langle 
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}
+		\right\rangle
+	}
 \end{mathpar}
@@
 		\begin{array}{c}
-			 \ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{n->min} \ \ast \ \texttt{n->max}\  \ast \\ 
-			\ \texttt{(if res then } \texttt{(min < x} \land \texttt{x < max)} \land \texttt{(x} \leq \texttt{min} \lor \texttt{x} \geq \texttt{max))}
+			 \ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{n->min} \ \ast \ \texttt{n->max}\  \\ 
+			 \ \ast  \ \texttt{(if res then } \texttt{(min < x} \land \texttt{x < max)} \land \texttt{(x} \leq \texttt{min} \lor \texttt{x} \geq \texttt{max))}
 		\end{array}
@@
 	\left\{\exists \ \texttt{t' l r lk1 lk2}. 
-	\begin{array}{l}
-		\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\
-		\ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, x))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (x, max))}
+	\begin{array}{c}
+		\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \\ 
+		\ \ast \ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, x))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (x, max))}
 	\end{array}

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
modified specs

@@
 We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
-
 \begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
 \end{align*}
-
 The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop.
@@
 		{\color{blue}
-			\forall \  \texttt{t}. \left\langle
+			\forall \  \texttt{m}. \left\langle
 			\begin{array}{c}
-				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{t}
+				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{m}
 			\end{array}
@@
 			\begin{array}{c}
-				\ \texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
+				\texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
 				\\ 
@@
 			\begin{array}{c}
-				\texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} 
+				\texttt{pn}\mapsto\texttt{n} \ast \texttt{n}\mapsto\texttt{t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} 
 			\end{array}
@@
 			\begin{array}{c}
-				\ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{t} \mapsto \texttt{(k', v', l, r)} \  \ast \\ 
-				\ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n' = n} \land \texttt{k = k'))}
+				\texttt{pn}\mapsto\texttt{n} \ast \texttt{n}\mapsto\texttt{t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} \\\  
+				\ast \ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n' = n} \land \texttt{k = k'))}
 			\end{array}
@@
 			\begin{array}{c}
-				\ \texttt{pn->n} \ \ast \ \texttt{n->t} 
+				\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} 
 			\end{array}
@@
 			\begin{array}{c}
-				\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
+				\ \texttt{pn} \mapsto \texttt{n} \ \ast \ \texttt{n}\mapsto \texttt{t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
 				\ \texttt{l} \mapsto \texttt{(NULL, lk1)}\  \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2)}
@@
 		\begin{array}{c}
-			\texttt{pn->n} \ \ast \texttt{n->t} \ \ast \ \texttt{n->min} \ \ast \ \texttt{n->max}
+			\texttt{pn}\mapsto\texttt{n} \ \ast \texttt{n}\mapsto\texttt{t} \ \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n->max}
 		\end{array}
@@
 		\begin{array}{c}
-			 \ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{n->min} \ \ast \ \texttt{n->max}\  \\ 
+			 \ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} \ \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n}\mapsto\texttt{max}\  \\ 
 			 \ \ast  \ \texttt{(if res then } \texttt{(min < x} \land \texttt{x < max)} \land \texttt{(x} \leq \texttt{min} \lor \texttt{x} \geq \texttt{max))}
@@
 		\begin{array}{c}
-			\texttt{pn->n} \ \ast \texttt{n->t} \  \ast \texttt{n->min} \ \ast \texttt{n->max}
+			\texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} \  \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n}\mapsto\texttt{max}
 		\end{array}
@@
 	\begin{array}{c}
-		\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \\ 
+		\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \\ 
 		\ \ast \ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, x))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (x, max))}

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
fix breakage

@@
 				\texttt{pn}\mapsto\texttt{n} \ast \texttt{n}\mapsto\texttt{t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} \\\  
-				\ast \ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n' = n} \land \texttt{k = k'))}
+				\ast \ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n = n'} \land \texttt{k = k'))}
 			\end{array}
@@
 		\begin{array}{c}
-			\texttt{pn}\mapsto\texttt{n} \ \ast \texttt{n}\mapsto\texttt{t} \ \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n->max}
+			\texttt{pn}\mapsto\texttt{n} \ \ast \texttt{n}\mapsto\texttt{(t, lock, min, max)} 
 		\end{array}
@@
 		\begin{array}{c}
-			 \ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} \ \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n}\mapsto\texttt{max}\  \\ 
-			 \ \ast  \ \texttt{(if res then } \texttt{(min < x} \land \texttt{x < max)} \land \texttt{(x} \leq \texttt{min} \lor \texttt{x} \geq \texttt{max))}
+			 \ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{(t, lock, min, max)} \  \\ 
+			 \ \ast  \ \texttt{(if res then } \texttt{(min < k} \land \texttt{k < max)} \land \texttt{(k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max))}
 		\end{array}
@@
 		\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \\ 
-		\ \ast \ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, x))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (x, max))}
+		\ \ast \ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, k))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (k, max))}
 	\end{array}

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
filled in some explanations about templates

@@
 \subsection{Concurrent Search Structure Templates}
-\wm{This is probably more background than we need on search structures.}
+
 A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria.
 These structures organize data to enable quick and efficient search, insertion, deletion, and traversal operations.
-Concurrent search structures, on the other hand, are data structures accessible and modifiable by multiple threads or processes concurrently.
-However, designing concurrent search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory, including cache behavior.
-\wm{cite Masstree?} Additionally, these structures must ensure compatibility with diverse hardware and software platforms.
+For efficiency, many such structures are designed to allow concurrent access and modification, often employing fine-grained or lock-free concurrency to allow as many threads as possible to operate on separate parts of the data structure.
+Designing these search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory, including cache behavior~\cite{masstree}.
+The complex interplay between concurrency and data structure design in concurrent search structures makes them challenging targets for formal verification.
 
-The complex interplay between concurrency and shared memory in concurrent search structures makes formal verification a challenging task.
-Modularity is crucial in simplifying formal proofs for concurrent search structures, as it is for designing and maintaining large systems.
-By decomposing complex systems into smaller, more manageable components, modularity facilitates the separation of concerns and minimizes interactions between different parts of the system, reducing the complexity of formal proofs.
-Moreover, modularity permits the reuse of verified components, further simplifying the verification process.
-Lock-free hash tables, concurrent tries, and B-trees are frequently used concurrent search structures.
-These structures allow developers to create efficient, scalable systems while ensuring thread safety.
+The Concurrent Search Structure Template approach~\cite{templates} aims make the problem tractable by separating verification of concurrency control patterns from verification of the underlying data structure.
+Addressing the proof of each component separately makes the individual proofs easier, and also offers the possibility of proof reuse: each verified concurrency pattern (``template'') can be applied to many different data structures, and each verified sequential data structure can be outfitted with many different templates.
+In theory, verifying $n$ templates and $m$ data structures might yield $n \times m$ verified concurrent data structures; in practice, as we will see, both templates and data structures make assumptions that may invalidate certain combinations.
+%how much more can we say about this?
 
-\than{Need to say the connection to template algorithms here.}
-
-\wm{point 1 - explain the top-level goal and approach of templates, how it relates to modularity}
-
-\wm{point 2 - lay out the template approach, what a template is, what a data structure is, how their code and their specifications interface}
+We will discuss several templates in section~\ref{templates}, but broadly, a search structure template consists of a \lstinline{traverse} function that travels through a data structure in search of a particular key, using the appropriate concurrency control mechanism when moving from node to node.
+The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine when the current node does not contain the target key.
+\wm{Show an example template here?} Once the appropriate node for the key has been found, the template returns it to a top-level function that calls out to the data structure to perform the appropriate operation (e.g.
+insert, lookup, delete) on that node.
+Thus, the \lstinline{traverse} function can be verified without knowing anything about the target data structure other than its synchronization mechanism, and without knowing anything about the operation to be performed other than that it can be performed on a single node (or sometimes a node and its parent).
+The job of the data structure is to implement the \lstinline{findNext} function and the specific operations; the combination of the template and the data structure is then a fully operational concurrent data structure.
+%what more do we want to say here?
 
@@
 \subsection{Search Structure Templates}
+\label{templates}
 The C implementation of a node in our template style is:
@@
 If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again.
-This behavior aligns with the give-up template algorithm's approach, in which threads terminate their search upon encountering obstacles or deviating from the expected path.
+This happens when another concurrent thread restructures the data structure, invalidating the state of the current thread in between releasing one lock and acquiring another.
+For instance, the node we are traveling to may be moved to another part of the data structure intended to hold a different set of keys, or even deleted, before we acquire its lock.
 
@@
 Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented.
-The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section.
-\than{Could you write more things here to explain why we need inRange.}
-The purpose of the \lstinline{inRange} function is to determine whether the key value $\texttt{k}$ falls within the range of node $\texttt{n}$.
-If the key value is outside the node's range, the search is terminated, and the thread is restarted.
+The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed in the previous section.
+The purpose of the \lstinline{inRange} function is to determine whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors.
+If the key value is outside the node's range (e.g.
+because the node has been relocated), the search is terminated, and the thread is restarted.
 To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node.

### 2023-04-18 concurrency/templates/bst_template_coupling.c
fix breakage in code C

@@ -63,7 +63,6 @@ void treebox_free(treebox b) {
 }
 
 //Template style
-
 typedef struct pn {
     struct tree_t *p;
     struct tree_t *n;
@@ -86,8 +85,6 @@ int findnext (pn *pn, int x, void *value){
 }
 
 int traverse(pn *pn, int x, void *value) {
-    //printf("Traversing...\n");
-    //fflush(stdout);
     int flag = 1;
     for( ; ; ){
         pn->p = pn->n;
@@ -105,8 +102,6 @@ int traverse(pn *pn, int x, void *value) {
             }
         }
     }
-    //printf("Done traversing\n");
-    //fflush(stdout);
     return flag;
 }
 
@@ -147,7 +142,6 @@ void insert (treebox t, int x, void *value) {
 }
 
 void *lookup (treebox t, int x) {
-    //printf("Looking up %d\n", x);
     struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
     void *v;
     pn->n = *t;
@@ -163,8 +157,6 @@ void *lookup (treebox t, int x) {
 
     release(pn->n->lock);
     free(pn);
-    //printf("Done lookup\n");
-    //fflush(stdout);
     return v;
 }
 
@@ -227,7 +219,6 @@ void delete (treebox t, int x) {
 }
 
 //Traverse
-
 void Inorder(struct tree_t *p){
     if (p->t != NULL){
         Inorder(p->t->left);
@@ -242,9 +233,6 @@ void traverseInorder (treebox t){
     Inorder(tgt);
 }
 
-
-
-
 void *thread_func(void *args) {
     lock_t *l = (lock_t*)args;
     
@@ -257,7 +245,6 @@ void *thread_func(void *args) {
     return (void *)NULL;
 }
 
-
 int main (void) {
     tb = treebox_new();
     lock_t *t_lock ;
@@ -268,30 +255,11 @@ int main (void) {
     t_lock = makelock();
     /* Spawn */
     spawn((void *)&thread_func, (void *)t_lock);
-    
-    //insert at key 1
-    //insert(tb,1,"ONE");
-    //insert(tb,4,"four");
-    
     /*JOIN */
     acquire((void*)t_lock);
     freelock((void*)t_lock);
-    
-    // printf ("%d", sizeof (pthread_mutex_t));
-    
-/*    printf("%s\n", lookup(tb, 4));
-    fflush(stdout);
[…]

### 2023-04-18 concurrency/templates/bst_template_giveup.c

@@ -66,7 +66,6 @@ void treebox_free(treebox b) {
 }
 
 //Template style
-
 typedef struct pn {
     struct tree_t *p;
     struct tree_t *n;
@@ -276,27 +275,9 @@ int main (void) {
     t_lock = makelock();
     /* Spawn */
     spawn((void *)&thread_func, (void *)t_lock);
-    
-    //insert at key 1
-    //insert(tb,1,"ONE");
-    //insert(tb,4,"four");
-    
     /*JOIN */
     acquire((void*)t_lock);
     freelock((void*)t_lock);
-    
-    // printf ("%d", sizeof (pthread_mutex_t));
-    
-/*    printf("%s\n", lookup(tb, 4));
-    fflush(stdout);
-    printf("That's one...\n");
-    fflush(stdout);
-    printf("%s\n", lookup(tb, 5));
-    fflush(stdout);
-    printf("%s\n", lookup(tb, 6));*/
-    
-    //printf("%s\n", lookup2(tb, 5));
-    
     printf ("\nTraverse\n");
     traverseInorder(tb);
     

### 2023-04-19 concurrency/CONCUR 2023/paper.tex
write more about proving insert of give-up template

@@
 \subsubsection{Lock-coupling Template}
+\label{lock-coupling-algo}
 The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal.
@@
 \subsubsection{Give-Up Template}
+\label{give-up-algo}
 The give-up template is an algorithm for implementing concurrent search in a tree data structure.
@@
 		\\ 
-		\texttt{insertOp(pn, k, v)} 
+		\texttt{traverse(pn, k)} 
 		\\
@@
 
-We have defined the loop invariant for the given template of the \texttt{traverse} function as follows: 
+In contrast to the lock-coupling style, where each lock is associated with a lock-invariant—an arbitrary separation logic predicate representing the protected resource acquired by a thread when the lock is obtained and must be restored upon lock release—the give-up template has a different approach.
+In this method, every node possesses a lock bit at location $\ell$, which is set to \texttt{True} if and only if the node is locked.
+This protocol is encoded into the definition of the abstract state $\treerep$.
+The lock protects resources denoted by \texttt{R}, which can be removed from $\treerep$ by threads when the node is locked.
+
+Before proving the \texttt{traverse} function, let's define the loop invariant for the given template of the \texttt{traverse} function as follows:
 \begin{align*} &\texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
@@
 In both instances, we must prove that we can atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, by demonstrating that the in-memory tree update corresponds to a local modification in the abstract tree, which inserts the new key-value pair.
+Next, we release (line 11) the node \texttt{n} and return it to the shared state. Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
 
@@
 \end{figure}
+
+Using the give-up template, the \texttt{insert} method in Figure \ref{insert_giveup} is not significantly different from the \texttt{insert} method of the lock-coupling template.
+The primary distinction is that it does not require acquiring a lock before calling the \texttt{traverse} function.
+Furthermore, it utilizes an auxiliary \texttt{traverse} function, as illustrated in Figure \ref{fig:traverse_giveup}.
+This function shares the same objective as the previously discussed \texttt{traverse} function in the lock-coupling approach.
+
+The give-up template retains the specification of the \texttt{findNext} helper function from the lock-coupling template while introducing an extra helper function called \texttt{inRange}.
+Additionally, the \texttt{insertOp} specification exhibits subtle differences from that of the lock-coupling template, as it incorporates two new fields,  \texttt{min} and \texttt{max},  into each \texttt{node\_t} type node, as showcased in Section  \ref{give-up-algo}.
+
+The \texttt{inRange} specification (Figure \ref{fig:inRange_giveup}) simply necessitates that \texttt{inRange(pn, k)}  yields a boolean value, \texttt{res}.
+When \texttt{res} is \texttt{True}, it signifies that \texttt{k} is within the range of \texttt{(min, max)} associated with the node \texttt{n}.
+Conversely, \texttt{res} is \texttt{False}, it indicates that \texttt{k} is outside the boundaries of \texttt{(min, max)}.
+
+As previously mentioned, the \texttt{insertOp} specification (Figure \ref{fig:insertOp_giveup}) exhibits slight differences from that of the lock-coupling template, primarily due to the incorporation of two extra fields.
+As a result, node \texttt{n} needs point-to predicates for these new fields.
+Moreover, the current node's \texttt{l} and \texttt{r} point to two new empty child nodes with associated locks and ranges: \texttt{(NULL, lk1, (min, k))} and \texttt{(NULL, lk2, (k, max))}.
+In the context of binary search trees, a node inherits its range from its parent based on the parent's key; thus, if node \texttt{n} has range \texttt{(min, max)} and key \texttt{k}, its left child will have the range \texttt{(min, k)} and its right child the range \texttt{(k, max)}.
 

### 2023-04-20 concurrency/CONCUR 2023/paper.tex
combine insert function of giveup and lock to save more space
write more for give-up

@@
 \end{figure}
-%\than{Actually, $\nodeboxrep \texttt{(pn->n)}$ and $\texttt{data\_at  n pn} \ast \texttt{my\_half g\_root} (-\inf, +\inf)$ is equivalent to $\texttt{tree\_rep}$.
-But now we can ignore the information of $\texttt{my\_half}$.
-}
 
@@
 
-%The loop invariants mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop.
-% CHANGE pn-> p to pn->n
-
 The proof begins with $\nodeboxrep \texttt{(pn->n)}$, which comprises a lock invariant $\texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$, and a resource $\texttt{R(pn->n)}$ acquired before invoking the \lstinline{traverse} function.
@@
 
-%\than{FIX ME: These things above didn't represent hand-over-hand locking instance, since I tried to avoid talking about lock invariant.}
 \than{I feel somewhat perplexed while attempting to articulate the situation mentioned above. Can someone help me fix or explain it more?}
@@
 \subsubsection{Give-up Template}
+\label{traverse_proof_giveup}
 Moving forward with the proofs, we define the following specification for the traverse function within the give-up template, as depicted in Figure \ref{fig:traverse_giveup}.
@@
 \begin{figure}[ht]
-
+	\begin{subfigure}[t]{0.48\textwidth}
 	\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
@@
 	\label{insert_lock}	
+	\end{subfigure}\qquad
+    \begin{subfigure}[t]{0.48\textwidth}
+    \lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
+    \caption{The \lstinline{insert} method of the give-up template algorithm}
+    \label{insert_giveup}
+    \end{subfigure} 
+\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
+\label{insert_lock_giveup} 
 \end{figure}
@@
 	\begin{mathpar}
-		% spec of inRange
 		{\color{blue}
@@
 
-\begin{figure}[ht]
-	\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
-	\caption{The \lstinline{insert} method of the give-up template algorithm}
-	\label{insert_giveup}
-\end{figure}
-
 Using the give-up template, the \texttt{insert} method in Figure \ref{insert_giveup} is not significantly different from the \texttt{insert} method of the lock-coupling template.
@@
 \begin{mathpar}
-	% spec of inRange
 	{\color{blue}
@@
 	\begin{mathpar}
-	%spec of insertOp
 	{\color{blue}
@@
 \end{figure}
+
+While the \texttt{insert} specification interface for the give-up template is identical to that of the lock-coupling template, a few adjustments need to be made within $\nodeboxrep\ \texttt{p}$ and $\treerep \ \texttt{m}$.
+Essentially, the \texttt{insert} function's specification comprises two main components: local and public preconditions and postconditions, as previously mentioned.
+Moreover, we must demonstrate that it is possible to atomically update the tree's abstract state from an unknown \texttt{m} to $\texttt{m[k} \mapsto \texttt{v])}$.
+
+The local preconditions and postconditions, containing $\nodeboxrep \ \texttt{p}$, depict the binary search tree's concrete state, where the root node $\texttt{p}$ points to the node's lock and footprint predicate, signifying that the root node is located within the search structure's footprint.
+In contrast, the public preconditions and postconditions are denoted by the predicate  $\treerep \ \texttt{m}$, which is identical to the public predicates of the \texttt{traverse} function in the give-up template.
+
+The proof for the \texttt{insert} operation begins with the preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$.
+We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (refer to line 3 in Figure \ref{insert_giveup}).
+In Section \ref{traverse_proof_giveup}, we discuss providing the identical preconditions for the \texttt{traverse} function, which encompasses the local precondition $\infp(\texttt{pn->n})$ and the public precondition $\treerep\ \texttt{m}$.
+Subsequently, we apply the \texttt{traverse} specification from Section \ref{traverse_proof_giveup}.
+
+Analogous to the lock-coupling template's \texttt{insert} operation, there are two possible branches representing the outcome following the execution of \texttt{traverse}.
+Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.
 

### 2023-04-20 concurrency/CONCUR 2023/paper.tex
combine two figure into one

@@
 \end{lstlisting}
-\than{I changed \lstinline{tree_t} into \lstinline{node_t}, and \lstinline{tree} into \lstinline{node}, since concurrent search structure will not depend on particular data structures.}
 
@@
 		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
-			%\wm{Shouldn't it be lock and then unlock?}} \than{I copied exactly the same from thesis of Krishna.
-However, in the their of proofs, they had lock before unlock like you said.
-Do we need to changed to fit with our context and their code?}
 		\label{traverse_lock_a}	
@@
 
-The \texttt{inRange} specification (Figure \ref{fig:inRange_giveup}) simply necessitates that \texttt{inRange(pn, k)}  yields a boolean value, \texttt{res}.
-When \texttt{res} is \texttt{True}, it signifies that \texttt{k} is within the range of \texttt{(min, max)} associated with the node \texttt{n}.
-Conversely, \texttt{res} is \texttt{False}, it indicates that \texttt{k} is outside the boundaries of \texttt{(min, max)}.
-
-As previously mentioned, the \texttt{insertOp} specification (Figure \ref{fig:insertOp_giveup}) exhibits slight differences from that of the lock-coupling template, primarily due to the incorporation of two extra fields.
-As a result, node \texttt{n} needs point-to predicates for these new fields.
-Moreover, the current node's \texttt{l} and \texttt{r} point to two new empty child nodes with associated locks and ranges: \texttt{(NULL, lk1, (min, k))} and \texttt{(NULL, lk2, (k, max))}.
-In the context of binary search trees, a node inherits its range from its parent based on the parent's key; thus, if node \texttt{n} has range \texttt{(min, max)} and key \texttt{k}, its left child will have the range \texttt{(min, k)} and its right child the range \texttt{(k, max)}.
-
-\begin{figure}
+\begin{figure}[h]
 	\centering
 	\begin{mathpar}
+		% inRange for give-up
 		{\color{blue}
@@
 		}
-\end{mathpar}
-\caption{Specification of \texttt{inRange} for the give-up template}
-\label{fig:inRange_giveup}
-\end{figure}
-
-\begin{figure}
-	\centering
-	\begin{mathpar}
+		\\
+		\\
+		%insertOp for give-up
 		{\color{blue}
@@
 	\end{mathpar}
-\caption{Specification of \texttt{insertOp} for the give-up template}
-\label{fig:insertOp_giveup}
+	\caption{Specifications of \texttt{inRange} and \texttt{insertOp} for the give-up template}
+	\label{fig:inRange_insertOp_giveup}
 \end{figure}
+
+The \texttt{inRange} specification (Figure \ref{fig:inRange_insertOp_giveup}) simply necessitates that \texttt{inRange(pn, k)}  yields a boolean value, \texttt{res}.
+When \texttt{res} is \texttt{True}, it signifies that \texttt{k} is within the range of \texttt{(min, max)} associated with the node \texttt{n}.
+Conversely, \texttt{res} is \texttt{False}, it indicates that \texttt{k} is outside the boundaries of \texttt{(min, max)}.
+
+As previously mentioned, the \texttt{insertOp} specification (Figure \ref{fig:inRange_insertOp_giveup}) exhibits slight differences from that of the lock-coupling template, primarily due to the incorporation of two extra fields.
+As a result, node \texttt{n} needs point-to predicates for these new fields.
+Moreover, the current node's \texttt{l} and \texttt{r} point to two new empty child nodes with associated locks and ranges: \texttt{(NULL, lk1, (min, k))} and \texttt{(NULL, lk2, (k, max))}.
+In the context of binary search trees, a node inherits its range from its parent based on the parent's key; thus, if node \texttt{n} has range \texttt{(min, max)} and key \texttt{k}, its left child will have the range \texttt{(min, k)} and its right child the range \texttt{(k, max)}.
 

### 2023-04-20 concurrency/CONCUR 2023/paper.tex
fix breakage

@@
 \section{Proof Mechanization}
-We describe the mechanized formal verification of the correctness of such these templates, implemented in C and verified with the VST, using concurrent separation logic (CSL) to reason about the behavior of the C code, and integrated Iris-style logical atomicity into VST.
-All proofs are formalized in Coq using VST, and can be found online at
+We present the mechanized formal verification of the correctness of these templates, which are implemented in C and verified with the Verified Software Toolchain (VST).
+We employ concurrent separation logic (CSL) to analyze the behavior of the C code and incorporate Iris-style logical atomicity within VST.
+
+By utilizing VST, we can create programs in standard C, automatically generate abstract syntax trees (ASTs) for them in Coq, and subsequently state and interactively prove specifications for these programs with the support of automated tactics.
+All proofs are formalized in Coq using VST and can be found online at
 

### 2023-04-21 concurrency/CONCUR 2023/paper.tex
first pass on abstract

@@
 Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort.
-In this paper, we implement the template approach in the Verified Software Toolchain, and use it to prove correctness of C implementations of concurrent data structures.
-%say more
+In this paper, we implement the template approach in the Verified Software Toolchain, and use it to prove correctness of C implementations of fine-grained concurrent data structures.
+In the process, we explore some of the practical requirements and limitations of the template approach on real code.
+%What's our conclusion?
+Are the template proofs easier/fewer lines than the proofs without the templates?
+We find that the template approach is generally effective on real C implementations, although some templates require modifications to the target data structure, and some data structures have operations that are not easily decomposed.
+This work represents a step towards practical compositional verification of real-world concurrent data structures.
 \end{abstract}
@@
 \end{itemize}
+\wm{This will be useful for contributions: Than, what would you say the hardest part was in completing the proofs?}
 

### 2023-04-22 concurrency/CONCUR 2023/paper.tex
answer William's question

@@
 \wm{This will be useful for contributions: Than, what would you say the hardest part was in completing the proofs?}
+
+\than{The most challenging aspect of the project was the proofs involving traversal.
+It took me a considerable amount of time to understand the pre- and post-conditions, as well as determining the appropriate loop invariants.
+I found myself frequently switching between proving the \texttt{insertOp} and \texttt{traverse} functions in order to combine all elements.
+However, once I had completed the \texttt{traverse} proofs, everything became much easier.}
 

### 2023-04-23 concurrency/CONCUR 2023/paper.tex
write more proof mechanization

@@
 
-\author{Duc Than Nguyen}{University of Illinois at Chicago, USA \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{Duc Than Nguyen}{University of Illinois at Chicago, USA \and \url{http://dnguye96.people.uic.edu} }{dnguye96@uic.edu}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.
@@
 
-\author{William Mansky}{University of Illinois at Chicago, USA}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
+\author{Lennart Beringer}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~eberinge} }{eberinge@cs.princeton.edu}{https://orcid.org/0000-0002-1570-3492}{(Optional) author-specific funding acknowledgements}
+
+\author{William Mansky}{University of Illinois at Chicago, USA\and \url{https://mansky.lab.uic.edu/} }{mansky1@uic.edu}{https://orcid.org/0000-0002-5351-895X}{(Optional) author-specific funding acknowledgements}
+
+\author{Shengyi Wang}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~shengyiw} }{shengyiw@cs.princeton.edu}{https://orcid.org/[0000-0002-2286-8703}{(Optional) author-specific funding acknowledgements}
 
@@
 However, once I had completed the \texttt{traverse} proofs, everything became much easier.}
+
+\paragraph*{Related Work}
+\than{I think we will write related work inside introduction part}
 
@@
 We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
-\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
-\end{align*}
+\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
+\end{mathpar}
 The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop.
-The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 6 in Figure \ref{traverse_lock_b}).
+The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 6 in Figure \ref{traverse_lock_b}).
 At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition.
-On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (refer to line 11 in Figure \ref{traverse_lock_b}).
+On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (line 11 in Figure \ref{traverse_lock_b}).
 In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.
@@
 The proof begins with $\nodeboxrep \texttt{(pn->n)}$, which comprises a lock invariant $\texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$, and a resource $\texttt{R(pn->n)}$ acquired before invoking the \lstinline{traverse} function.
-After \lstinline{findNext} successfully detects the next node (for example, \texttt{n'}) to visit, we acquire the lock of node \texttt{n'} (see line 14 in Figure \ref{traverse_lock_b}).
+After \lstinline{findNext} successfully detects the next node (for example, \texttt{n'}) to visit, we acquire the lock of node \texttt{n'} (line 14 in Figure \ref{traverse_lock_b}).
 It means we can get the resource $\texttt{R(pn->n')}$ from $\texttt{pn->n'->lock} \lockvar \texttt{R(pn->n')}$.
@@
 Before proving the \texttt{traverse} function, let's define the loop invariant for the given template of the \texttt{traverse} function as follows:
-\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
-\end{align*}
-
+\begin{mathpar} \texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
+\end{mathpar}
 To begin the proof, we start with the $\infp (\texttt{pn->n})$ predicate from the local pre-condition.
 After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\texttt{R(pn->n)}$ and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}.
-Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 8 in Figure \ref{traverse_giveup_b}).
+Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 8 in Figure \ref{traverse_giveup_b}).
 At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.
@@
 	\begin{subfigure}[t]{0.48\textwidth}
-	\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
+	\lstinputlisting[language=C, style=myStyle, escapechar=|]{lock_insert.c} 
 	\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
@@
 In both instances, we must prove that we can atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, by demonstrating that the in-memory tree update corresponds to a local modification in the abstract tree, which inserts the new key-value pair.
-Next, we release (line 11) the node \texttt{n} and return it to the shared state. Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
+Next, we release (line 11 in Figure \ref{insert_lock}) the node \texttt{n} and return it to the shared state.
+Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
 
@@
 The proof for the \texttt{insert} operation begins with the preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$.
-We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (refer to line 3 in Figure \ref{insert_giveup}).
+We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_giveup}).
 In Section \ref{traverse_proof_giveup}, we discuss providing the identical preconditions for the \texttt{traverse} function, which encompasses the local precondition $\infp(\texttt{pn->n})$ and the public precondition $\treerep\ \texttt{m}$.
@@
 \section{Proof Mechanization}
-We present the mechanized formal verification of the correctness of these templates, which are implemented in C and verified with the Verified Software Toolchain (VST).
-We employ concurrent separation logic (CSL) to analyze the behavior of the C code and incorporate Iris-style logical atomicity within VST.
+We present the mechanized formal verification of the correctness of these templates, which are implemented in C and verified with the Verified Software Toolchain.
+We employ concurrent separation logic to analyze the behavior of the C code and incorporate Iris-style logical atomicity within VST.
+
+\begin{table}[h]
+	\begin{center}
+		\begin{tabular}{l | c || l | c } 
+			\hline
+			\textbf{Lock-coupling template} & \textbf{Line of proofs} & \textbf{Give-up  template} & \textbf{Line of proofs} \\
+			\hline
+			Template specfications & 190 & Template specfications & 138 \\
+			\texttt{findNext} proof 	& 27  & \texttt{findNext} proof & 27 \\
+			\texttt{insertOp} proof 	& 32  & \texttt{insertOp} proof & 31 \\
+			& & \texttt{inRange} proof & 30 \\
+			\texttt{traverse} proof	& 360  & \texttt{traverse} proof & 336 \\ 
+			\texttt{insert} proof 	   		& 450  & \texttt{insert} proof & 284 \\
+			\texttt{lookup} proof 	   	& 235  &  \texttt{lookup} proof & 304 \\ 
+			\texttt{pushdown\_left} proof 	   	& 282 & & \\	
+			\texttt{delete} proof 		& 182	 &   & 	 \\
+			supporting proofs    & 	1909	 & supporting proofs 	& 1876 \\
+			\hline
+			\textbf{Total} & \textbf{3667} & & \textbf{3026}  
+		\end{tabular}
+		\caption{Size of Coq definitions and proofs, by topic.}
+		\label{table_coq}
+	\end{center}
+\end{table}
 
 By utilizing VST, we can create programs in standard C, automatically generate abstract syntax trees (ASTs) for them in Coq, and subsequently state and interactively prove specifications for these programs with the support of automated tactics.
+Table \ref{table_coq} presents a summary of our Coq development.
+We formalized all previously described proofs in Coq, taking advantage of VST's features for verifying separation logic specifications of C functions.
+This encompasses supplementary proofs, template specifications (encompassing all helper functions such as \texttt{findNext}, \texttt{insertOp}, and others, in addition to \texttt{traverse}, \texttt{insert}, \texttt{lookup}, and \texttt{delete}), as well as proofs against with lock-coupling and give-up template specifications.
 All proofs are formalized in Coq using VST and can be found online at
@@
 \href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}
-
-\section{Related Work}
 
@@
 %% Please use bibtex, 
-
+\newpage
 \bibliography{../sources}

### 2023-04-23 concurrency/CONCUR 2023/paper.tex
fix breakage and restructured
fix some breakage

@@
 \subsection{Iris and VST}
+\than{write more here ... }
 
@@
 In contrast to the lock-coupling style, where each lock is associated with a lock-invariant—an arbitrary separation logic predicate representing the protected resource acquired by a thread when the lock is obtained and must be restored upon lock release—the give-up template has a different approach.
-In this method, every node possesses a lock bit at location $\ell$, which is set to \texttt{True} if and only if the node is locked.
+In this method, every node possesses a lock bit at location $\ell$, which is set to \texttt{true} if and only if the node is locked.
 This protocol is encoded into the definition of the abstract state $\treerep$.
-The lock protects resources denoted by \texttt{R}, which can be removed from $\treerep$ by threads when the node is locked.
+The protocol is formally defined as $\texttt{inv\_for\_lock} \ \ell \ \texttt{R} \triangleq \exists b.
+\ \ell \mapsto b \ \ast \ \texttt{if } b \ \texttt{then emp else R}$, where a lock at location $\ell$ safeguards resources denoted by \texttt{R}.
+When the lock is held, meaning the node is locked ($\ell \mapsto \texttt{true}$), threads can remove resources \texttt{R} from $\treerep$.
+On the other hand, when the lock is not held ($\ell \mapsto \texttt{false}$), resources \texttt{R} are required to be part of the invariant.
 
@@
 
-\subsection{Templates vs. Internal Reorganization} % move to section 4?
-In the binary search tree, \texttt{pushdown\_left} involves both the structure of the tree and the concurrency protocol.
-There's no good way to disentangle this into a template and a concurrency-free data structure operation.
-
-\than{Or can we move this section to Proof mechanization, where we talk on BST speficially.}
-
 \section{Verified Data Structures}
 In this section, we demonstrate how to integrate all components into high-level specifications for the purpose of verifying two template algorithms, lock-coupling and give-up, in the real C programming language.
+
+\than{Write more here ... }
 
@@
 \subsection{Lock-coupling Template}
-\begin{figure}[ht]
+\begin{figure}[h]
 	\begin{subfigure}[t]{0.48\textwidth}
@@
 The \texttt{inRange} specification (Figure \ref{fig:inRange_insertOp_giveup}) simply necessitates that \texttt{inRange(pn, k)}  yields a boolean value, \texttt{res}.
-When \texttt{res} is \texttt{True}, it signifies that \texttt{k} is within the range of \texttt{(min, max)} associated with the node \texttt{n}.
-Conversely, \texttt{res} is \texttt{False}, it indicates that \texttt{k} is outside the boundaries of \texttt{(min, max)}.
+When \texttt{res} is \texttt{true}, it signifies that \texttt{k} is within the range of \texttt{(min, max)} associated with the node \texttt{n}.
+Conversely, \texttt{res} is \texttt{false}, it indicates that \texttt{k} is outside the boundaries of \texttt{(min, max)}.
 
@@
 Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.
+
+\subsection{Templates vs. Internal Reorganization} % move to section 4?
+In the binary search tree, \texttt{pushdown\_left} involves both the structure of the tree and the concurrency protocol.
+There's no good way to disentangle this into a template and a concurrency-free data structure operation.
+
+\than{Write more ...}
 

### 2023-04-24 concurrency/CONCUR 2023/paper.tex
todos

--- 2023-04-22 concurrency/CONCUR 2023/paper.tex
@@
 
-\than{The most challenging aspect of the project was the proofs involving traversal.
-It took me a considerable amount of time to understand the pre- and post-conditions, as well as determining the appropriate loop invariants.
-I found myself frequently switching between proving the \texttt{insertOp} and \texttt{traverse} functions in order to combine all elements.
-However, once I had completed the \texttt{traverse} proofs, everything became much easier.}
-
 \section{Background}
@@
 The job of the data structure is to implement the \lstinline{findNext} function and the specific operations; the combination of the template and the data structure is then a fully operational concurrent data structure.
+Krishna et al presented templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures, but many common search structures are not graphs; in this paper, we present a version of templates that is independent of flow interfaces.
+%is this a contribution in itself?
 %what more do we want to say here?

### 2023-04-25 concurrency/CONCUR 2023/paper.tex
filling in some of my pieces

--- 2023-04-23 concurrency/CONCUR 2023/paper.tex
@@
 
-The Concurrent Search Structure Template approach~\cite{templates} aims make the problem tractable by separating verification of concurrency control patterns from verification of the underlying data structure.
+The Concurrent Search Structure Template approach of Krishna et al.~\cite{templates} aims make the problem tractable by separating verification of concurrency control patterns from verification of the underlying data structure.
 Addressing the proof of each component separately makes the individual proofs easier, and also offers the possibility of proof reuse: each verified concurrency pattern (``template'') can be applied to many different data structures, and each verified sequential data structure can be outfitted with many different templates.
@@
 \subsection{Iris and VST}
-\than{write more here ... }
+The search structure template approach is built on \emph{concurrent separation logic}, which is used to specify and prove pre- and postconditions for the template and data structure functions.
+Krishna et al.
+implemented their framework in Iris~\cite{iris}, a language-independent separation logic framework built in the Coq proof assistant~\cite{coq}, with flexible support for ghost state, invariants, and atomic specifications.
+This makes it easy to describe the effects of concurrency control functions independently of the underlying data structure, e.g., as ``this function atomically finds a node that contains the target key.'' %forward reference to where we show the specs?
+or show them here?
+The existing implementation is written in Iris's HeapLang, a simple functional programming language with shared-memory concurrency.
+
+To apply the template approach to real-world code, we instead use the Verified Software Toolchain (VST)~\cite{plcc}, a separation-logic-based verifier for C programs.
+VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, giving it a realistic semantics and a guarantee that proved properties will hold on compiled code.
+Recent work on VST~\cite{vst-iris} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications.
+Building on this, we were able to reconstruct the template approach in VST and apply it to real C programs.
+%clean this up a little
 
@@
 In this section, we demonstrate how to integrate all components into high-level specifications for the purpose of verifying two template algorithms, lock-coupling and give-up, in the real C programming language.
-
-\than{Write more here ... }
+The target specifications for our search structure(s) are shown below: each operation should atomically have its effect on the data structure, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
 
@@
 
-\subsection{Templates vs. Internal Reorganization} % move to section 4?
-In the binary search tree, \texttt{pushdown\_left} involves both the structure of the tree and the concurrency protocol.
-There's no good way to disentangle this into a template and a concurrency-free data structure operation.
+\subsection{Templates vs. Internal Reorganization}
+Our original goal was to verify an existing concurrent BST implementation with the template approach, refactoring as necessary but otherwise leaving the code unchanged.
+However, there is a fundamental mismatch between the template approach and the standard approach to deletion in a BST.
+The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert_op} or \lstinline{lookup_op}).
+When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown} operation, %include figure?
+rotating three-node sections of the tree until the node to be deleted is at a leaf.
+This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them).
+In fact, most of the existing template proofs either do not include deletion at all, or do a very lightweight form of deletion in which nodes are marked as ``deleted'' but are never actually removed from the data structure.
 
-\than{Write more ...}
+We believe this represents a fundamental limitation in the template approach.
+Operations in the style of \lstinline{pushdown} are ubiquitous in tree structures (e.g., the rebalancing operations of AVL and red-black trees), and always weave together synchronization and restructuring in a way that cannot easily be decomposed.
+This is a serious impediment to the goal of verifying concurrency control separately from the data structure: even if we have a verified template and a verified single-threaded tree, we still need to verify a complex data-structure-specific concurrent function.
+%in fact, usually the most complicated part!
+%what's the conclusion here? can we see a way through?
 

### 2023-04-25 concurrency/CONCUR 2023/paper.tex
a few more edits and todos

@@
 This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them).
-In fact, most of the existing template proofs either do not include deletion at all, or do a very lightweight form of deletion in which nodes are marked as ``deleted'' but are never actually removed from the data structure.
+This contrasts with linked-list deletion, which only involves the target node and its immediate neighbors.
+In practice, the existing template proofs for tree structures either do not include deletion at all, or use a very lightweight form of deletion in which nodes are marked as ``deleted'' but are never actually removed from the data structure.
 

### 2023-04-26 concurrency/CONCUR 2023/paper.tex
fix some breakage

@@
 
-\paragraph*{Related Work}
+\subsection*{Related Work}
 \than{I think we will write related work inside introduction part}
+
+\than{Write more here}
 
@@
 
-\subsection{Proofs of Templates} \than{Can someone suggest another title?}
-
+\subsection{Proofs of Templates} \
+%than{Can someone suggest another title?}
 In this section, we verify \lstinline{traverse} for both lock-coupling and give-up templates.
@@
 \end{mathpar}
+
+Our primary focus is on verification the correctness of \texttt{insert} operations in binary search trees using lock-coupling and give-up templates.
+The \texttt{lookup} operations closely resemble the \texttt{insert} operations, sharing a significant portion of their proof.
+As for the \texttt{delete} operations, we have supplied proof solely for the lock-coupling template, which is discussed in Section \ref{internal_reorganize}.
+
+The complete proof can be found online in our publicly accessible repository of machine-checked proofs (refer to Section \ref{proof_mech}).
+This repository also contains the other proofs featured in our monograph.
 
@@
 \subsection{Templates vs. Internal Reorganization}
+\label{internal_reorganize}
 Our original goal was to verify an existing concurrent BST implementation with the template approach, refactoring as necessary but otherwise leaving the code unchanged.
 However, there is a fundamental mismatch between the template approach and the standard approach to deletion in a BST.
-The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert_op} or \lstinline{lookup_op}).
+The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}).
 When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown} operation, %include figure?
@@
 We believe this represents a fundamental limitation in the template approach.
-Operations in the style of \lstinline{pushdown} are ubiquitous in tree structures (e.g., the rebalancing operations of AVL and red-black trees), and always weave together synchronization and restructuring in a way that cannot easily be decomposed.
+Operations in the style of \lstinline{pushdown_left} are ubiquitous in tree structures (e.g., the rebalancing operations of AVL and red-black trees), and always weave together synchronization and restructuring in a way that cannot easily be decomposed.
 This is a serious impediment to the goal of verifying concurrency control separately from the data structure: even if we have a verified template and a verified single-threaded tree, we still need to verify a complex data-structure-specific concurrent function.
@@
 \section{Proof Mechanization}
+\label{proof_mech}
 We present the mechanized formal verification of the correctness of these templates, which are implemented in C and verified with the Verified Software Toolchain.

### 2023-04-26 concurrency/CONCUR 2023/paper.tex
add a reference
fix breakage

@@
 
-\than{Write more here}
+\than{William, I've already added the reference of a paper we discussed before}
+
+Proving highly-concurrent traversals correct \cite{feldman2020proving}
 
@@
 \begin{mathpar}
-	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle\ \texttt{insert(p, k, v)}\ \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}\rangle
+	\centering
+{\color{blue} \langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle }
 
-	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \rangle\ \texttt{lookup(p, k)}\ \langle \texttt{v}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \land \texttt{m(k) = v}\rangle
+\texttt{insert(p, k, v)}
 
-	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle\ \texttt{delete(p, k)}\ \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{(m[k} \mapsto \_ \texttt{])}\rangle
+{\color{blue} \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}\rangle}
+
+{\color{blue}	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \rangle }
+
+ \texttt{lookup(p, k)}  \
+
+{\color{blue}\langle \texttt{v}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \land \texttt{m(k) = v}\rangle }
+
+{\color{blue}	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle }
+
+\texttt{delete(p, k)} \ \ \ \ \ \ \
+
+{\color{blue} \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{(m[k} \mapsto \_ \texttt{])}\rangle }
 \end{mathpar}
@@
 The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}).
-When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown} operation, %include figure?
+When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure?
 rotating three-node sections of the tree until the node to be deleted is at a leaf.

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
fix breakage

@@
 
-\subsection{Proofs of Templates} \
-%than{Can someone suggest another title?}
-In this section, we verify \lstinline{traverse} for both lock-coupling and give-up templates.
+\subsection{Proofs of Templates} 
+In this section, we present our verification methodology, termed \lstinline{traverse}, which applies to both lock-coupling and give-up templates.
+The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}.
+Before we delve into the intricacies of these verifications, it is essential to familiarize ourselves with the concept of \emph{atomic triples}.
 
@@
 			\begin{array}{c}
-				\ \texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
+				\ \texttt{(res} \leftrightarrow \texttt{(pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
 				\ast \ \texttt{R(pn->n')} \ast \ \texttt{(k} \in \texttt{range(n'))}
@@
 We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
-\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
+\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ \texttt{(k} \in \texttt{range)}\ \land \ \nodeboxrep\texttt{(pn->n')}  \ \ast \  \texttt{R(pn->n')}   
 \end{mathpar}
@@
 			\begin{array}{c}
-				\texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
+				\texttt{(res} \leftrightarrow \texttt{(pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
 				\\ 
@@
 \begin{mathpar}
-	\centering
-{\color{blue} \langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle }
-
-\texttt{insert(p, k, v)}
-
-{\color{blue} \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}\rangle}
-
-{\color{blue}	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \rangle }
-
- \texttt{lookup(p, k)}  \
-
-{\color{blue}\langle \texttt{v}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \land \texttt{m(k) = v}\rangle }
-
-{\color{blue}	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle }
-
-\texttt{delete(p, k)} \ \ \ \ \ \ \
-
-{\color{blue} \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{(m[k} \mapsto \_ \texttt{])}\rangle }
+	{\color{blue}
+		\texttt{m.}\left\langle 
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m}
+		\right\rangle
+	}
+	\ \texttt{insert(p, k, v)}\ 
+	{\color{blue}
+		\left\langle 
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}
+		\right\rangle
+	}
+\\
+	{\color{blue}
+		\texttt{m.}\left\langle 
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m}
+		\right\rangle
+	}
+	\ \texttt{lookup(p, k)}\ 
+	{\color{blue}
+		\left\langle \texttt{v}. \ \
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m} \land \texttt{m(k) = v}
+		\right\rangle
+	}
+\\
+	{\color{blue}
+		\texttt{m.}\left\langle 
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m}
+		\right\rangle
+	}
+	\ \texttt{delete(p, k)}\ 
+	{\color{blue}
+		\left\langle 
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{(m[k} \mapsto \_ \texttt{])}
+		\right\rangle
+	}
 \end{mathpar}
@@
 We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_giveup}).
-In Section \ref{traverse_proof_giveup}, we discuss providing the identical preconditions for the \texttt{traverse} function, which encompasses the local precondition $\infp(\texttt{pn->n})$ and the public precondition $\treerep\ \texttt{m}$.
+In Section \ref{traverse_proof_giveup}, we discuss providing the identical preconditions for the \texttt{traverse} function, which encompasses the local precondition $\infp\texttt{(pn->n)}$ and the public precondition $\treerep\ \texttt{m}$.
 Subsequently, we apply the \texttt{traverse} specification from Section \ref{traverse_proof_giveup}.
@@
 Analogous to the lock-coupling template's \texttt{insert} operation, there are two possible branches representing the outcome following the execution of \texttt{traverse}.
-Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.
+Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ \texttt{(m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.
 

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
working on intro

@@
 
-In this paper, we apply the template approach to concurrent data structure implementations in C.
-We verify them using the Verified Software Toolchain (VST)~\cite{plfcc}, and report on its effectiveness and the challenges we encountered.
+%give a feel for the templates/the traverse function here, so people can understand why the reimplementation was hard/interesting
+%mention separation logic, logical atomicity
+
+The template approach was originally implemented on top of flow interfaces~\cite{krishna2017flow}, a framework for specifying and verifying graph-style data structures, in a combination of two verifiers: the templates were verified in the interactive Iris prover~\cite{iris}, while the data structure implementations were verified with the automated Grasshopper tool~\cite{grasshopper}.
+The target data structures were written in HeapLang, a simple functional programming language with shared-memory concurrency.
+In this paper, we reimplement the approach in the Verified Software Toolchain (VST)~\cite{plcc}, an interactive system for proving correctness of C programs based on a detailed semantics of the C language.
+This allows us to apply the template approach to concurrent data structure implementations in C.
 The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}.
 Our proofs make use of recent work that has integrated Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
+%clean up the phrasing a little
 
@@
 \begin{itemize}
+\item We reimplement the template approach without flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/subtree''.
+\item We implement the template approach in VST, allowing us to apply it to real C programs and obtain end-to-end correctness proofs in a single verification system.
 \item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real programming language.
-\item 
+\item In the process, we encounter some of the limitations of the template approach on tree data structures. \wm{It would be nice if we had a possible solution!}
 \end{itemize}
-\wm{This will be useful for contributions: Than, what would you say the hardest part was in completing the proofs?}
+%\wm{This will be useful for contributions: Than, what would you say the hardest part was in completing the proofs?}
 
-\than{The most challenging aspect of the project was the proofs involving traversal.
+%\than{The most challenging aspect of the project was the proofs involving traversal.
 It took me a considerable amount of time to understand the pre- and post-conditions, as well as determining the appropriate loop invariants.
@@
 \subsection*{Related Work}
-\than{I think we will write related work inside introduction part}
+CSL provers for fine-grained concurrent data structures: Iris~\cite{iris,iris-folly}, FCSL~\cite{fcsl}, TaDA~\cite{tada,Xiong2017Abstract}, VeriFast~\cite{verifast,verifast-conc}
 
-\than{William, I've already added the reference of a paper we discussed before}
+%Gotsman et al.~\cite{gotsman}, in the same paper in which they introduced invariant-style lock specs, also verified a linked list with hand-over-hand locking, which became a common example for verifiers that handled fine-grained locking.
+VeriFast~\cite{verifast}, a separation-logic-based verifier for C and
+%Java, supports lock-based and atomic
+%concurrency~\cite{verifast-conc}, and has been used to verify a
+%hand-over-hand-locking linked list similar to that of Krishna et al.
+%The specifications of the lock operations and the list itself use
+%a precursor of TaDA's logical atomicity. VeriFast is not 
+%foundational, but its basic logic is verified
+%against the semantics of a toy language in Coq.
+
+%Xiong et al.
+\cite{Xiong2017Abstract} have demonstrated the verification of ConcurrentSkipListMap from java.util.concurrent library using the recent advances in fine-grained concurrency reasoning.
+Their work is mainly based on the abstract atomicity from TaDA logic, and give two modular specifications for concurrent maps: one specification focus on the entire map structure which is suitable for verifying implementation, and another specification focus on the key-value pairs which appropriate for verifying clients.
+We use the same idea of atomicity (though implemented in Iris) in our work.
+
+separating concurrency from functionality: linearizability~\cite{linearizability} (and refinement, e.g. Civl~\cite{civl}), logical atomicity~\cite{tada}
 
 Proving highly-concurrent traversals correct \cite{feldman2020proving}
+
+multicopy search structures~\cite{template-multi}
 
@@
 
-%% Please use bibtex, 
 \newpage

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
citations

@@
 
-\title{Verifying Concurrent Search Structure Templates for C Programs using VST} %TODO Please add
+\title{Compositional Verification of Concurrent C Programs with Search Structure Templates} %keep working on this
 
@@
 \begin{abstract}
+%a bit more of a lead-in?
 Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort.
@@
 \label{sec:introduction}
-Krishna et al.
-proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (e.g.
-optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of correctness of the underlying data structure (e.g.
-linked list, hashtable, B-tree).
+Krishna et al.~proposed concurrent search structure templates~\cite{templates} as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree).
 The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency.
@@
 
-The template approach was originally implemented on top of flow interfaces~\cite{krishna2017flow}, a framework for specifying and verifying graph-style data structures, in a combination of two verifiers: the templates were verified in the interactive Iris prover~\cite{iris}, while the data structure implementations were verified with the automated Grasshopper tool~\cite{grasshopper}.
+The template approach was originally implemented on top of flow interfaces~\cite{krishna2017flow}, a framework for specifying and verifying graph-style data structures, in a combination of two verifiers: the templates were verified in the interactive Iris prover~\cite{iris}, while the data structure implementations were verified with the automated GRASShopper tool~\cite{grasshopper}.
 The target data structures were written in HeapLang, a simple functional programming language with shared-memory concurrency.
@@
 The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}.
-Our proofs make use of recent work that has integrated Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
+Our proofs make use of recent work integrating Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
 %clean up the phrasing a little
@@
 A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria.
-These structures organize data to enable quick and efficient search, insertion, deletion, and traversal operations.
+In the abstract, a search structure implements a map from keys to values, and provides operations such as search, insertion, deletion, and traversal on that map.
 For efficiency, many such structures are designed to allow concurrent access and modification, often employing fine-grained or lock-free concurrency to allow as many threads as possible to operate on separate parts of the data structure.
@@
 
-The Concurrent Search Structure Template approach of Krishna et al.~\cite{templates} aims make the problem tractable by separating verification of concurrency control patterns from verification of the underlying data structure.
+The Concurrent Search Structure Template approach of Krishna et al.~\cite{templates} aims to make the problem tractable by separating verification of concurrency control patterns from verification of the underlying data structure.
 Addressing the proof of each component separately makes the individual proofs easier, and also offers the possibility of proof reuse: each verified concurrency pattern (``template'') can be applied to many different data structures, and each verified sequential data structure can be outfitted with many different templates.
@@
 The job of the data structure is to implement the \lstinline{findNext} function and the specific operations; the combination of the template and the data structure is then a fully operational concurrent data structure.
-Krishna et al presented templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures, but many common search structures are not graphs; in this paper, we present a version of templates that is independent of flow interfaces.
+Krishna et al.~presented templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures, but many common search structures are not graphs; in this paper, we present a version of templates that is independent of flow interfaces.
 %is this a contribution in itself?
@@
 \subsection{Iris and VST}
-The search structure template approach is built on \emph{concurrent separation logic}, which is used to specify and prove pre- and postconditions for the template and data structure functions.
-Krishna et al.
-implemented their framework in Iris~\cite{iris}, a language-independent separation logic framework built in the Coq proof assistant~\cite{coq}, with flexible support for ghost state, invariants, and atomic specifications.
+The search structure template approach uses \emph{concurrent separation logic} to specify and prove pre- and postconditions for the template and data structure functions.
+Krishna et al.~implemented their framework in Iris~\cite{iris}, a language-independent separation logic framework built in the Coq proof assistant~\cite{coq}, with flexible support for ghost state, invariants, and atomic specifications.
 This makes it easy to describe the effects of concurrency control functions independently of the underlying data structure, e.g., as ``this function atomically finds a node that contains the target key.'' %forward reference to where we show the specs?
@@
 
-Figure \ref{traverse_lock} shows the \lstinline{traverse} function as presented by Krishna et al.
-(Figure \ref{traverse_lock_a}) and the corresponding C implementation (Figure \ref{traverse_lock_b}).
+Figure \ref{traverse_lock} shows the \lstinline{traverse} function as presented by Krishna et al.~(Figure \ref{traverse_lock_a}) and the corresponding C implementation (Figure \ref{traverse_lock_b}).
 The C implementation uses a struct

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
clarified the interface mismatch around inRange

@@
 The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine when the current node does not contain the target key.
-\wm{Show an example template here?} Once the appropriate node for the key has been found, the template returns it to a top-level function that calls out to the data structure to perform the appropriate operation (e.g.
+\wm{Show an example template here, and enough of the specification to explain atomicity} Once the appropriate node for the key has been found, the template returns it to a top-level function that calls out to the data structure to perform the appropriate operation (e.g.
 insert, lookup, delete) on that node.
@@
 VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, giving it a realistic semantics and a guarantee that proved properties will hold on compiled code.
-Recent work on VST~\cite{vst-iris} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications.
-Building on this, we were able to reconstruct the template approach in VST and apply it to real C programs.
+Recent work on VST~\cite{iris-vst-arxiv} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications.
+In this paper, we use these extensions to reconstruct the template approach in VST and apply it to real C programs.
 %clean this up a little
@@
 
-\subsection{Search Structure Templates}
+\subsection{Defining the Templates}
 \label{templates}
-The C implementation of a node in our template style is:
+
+The core of each search structure template is a \lstinline{traverse} function that navigates a data structure using the target synchronization mechanism.
+Templates abstract away from the data structure being traversed, operating on a generic \lstinline{node} type and taking data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters.
+We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
@@
 \end{lstlisting}
-
-Each node of type \lstinline{node_t} has a \lstinline{lock} field that protects the node and a \lstinline{t} field that points to a \lstinline{node} struct.
-Depending on the data structure, the \lstinline{t} field can have different interpretations.
-In a binary search tree, if the \lstinline{t} field is \lstinline{NULL}, it indicates a leaf node with no key or value; otherwise, it points to a \lstinline{node} struct containing the node's key, value, and child pointers.
+There are two things worth noting about this data structure.
+First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding as few locks as possible to allow other parts of the data structure to be modified concurrently.
+Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
@@
 \end{lstlisting}
-
-For a linked list, a \lstinline{t} field with \lstinline{NULL} represents the last node in the list, while a non-null \lstinline{t} field points to a \lstinline{node} struct holding the node's key, value, and a pointer to the next item in the list.
-
+while a linked list might use
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
@@
 \end{lstlisting}
+This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component.
+Now, we use \lstinline{node_t} to define \lstinline{traverse} functions for specific concurrency patterns.
 
-%For the purpose of reasoning about lock-coupling style (also known as hand-over-hand locking), we specifically consider traversing the tree using this locking method.
-To accomplish this, we must maintain at least one lock during traversal between nodes.
-This requires obtaining the lock for the next node before releasing the lock for the current node.
-Therefore, we utilize the \lstinline{pn} struct containing two \lstinline{node_t} type nodes: the \lstinline{p} field representing the current node, and the \lstinline{n} field denoting the next node.
+\subsubsection{Lock-Coupling Template}
+\label{lock-coupling-algo}
 
-\subsubsection{Lock-coupling Template}
-\label{lock-coupling-algo}
-The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal.
-Each thread continuously maintains at least one lock during traversal between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
-
-Figure \ref{traverse_lock} shows the \lstinline{traverse} function as presented by Krishna et al.~(Figure \ref{traverse_lock_a}) and the corresponding C implementation (Figure \ref{traverse_lock_b}).
+The first template we consider is lock coupling (also called hand-over-hand locking), in which threads use the locks on each node to prevent interference from other threads during traversal.
+Each thread always holds at least one lock, and acquires the lock on the next node before releasing its current lock, ensuring that other threads cannot invalidate the ongoing search.
+Figure \ref{traverse_lock} shows the lock-coupling \lstinline{traverse} function as presented by Krishna et al.~(Figure \ref{traverse_lock_a}) and our corresponding C implementation (Figure \ref{traverse_lock_b}).
 The C implementation uses a struct
@@
 to mimic the pair of nodes \lstinline{(p, n)} returned by the functional implementation, where \lstinline{n} is the current node and \lstinline{p} is its parent.
-The algorithm assumes that certain helper functions must be provided by the implementation that satisfy certain criteria.
-One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$.
+The template relies on one function provided by the underlying data structure, namely \lstinline{findNext}, which is used to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$.
 In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a boolean and, if a next node is found, modifies \lstinline{pn->n}.
 
-To ensure thread safety, it is essential to acquire and release locks in a specific order during traversal.
+%To ensure thread safety, it is essential to acquire and release locks in a specific order during traversal.
 The lock-coupling scheme achieves this by requiring threads to acquire locks in increasing order of node addresses.
@@
 
-To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.
-Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findNext}.
-The purpose of \lstinline{findNext} is to determine the next node to be visited during the traversal.
-In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}.
-Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values.
-The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal.
-If it fails to do so, it returns false, which other functions can use to make decisions.
-For instance, in a binary search tree, a node with a value of 42 would have a left child with a set of key values consisting of values less than 42 and a right child with one greater than 42.
-If we want to insert a new node with the target key $\texttt{k}$ into the tree, we can use the \lstinline{traverse} function to traverse the tree.
-When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $\texttt{k}$ that equals the node's value.
+Our implementation of \lstinline{traverse} translates the functional implementation into more idiomatic C code.
+The lock-coupling pattern can be seen on lines 14 and 15, where \lstinline{traverse} acquires the next node's lock and then releases the current node's lock.
+The function stops when it reaches an empty leaf node (\lstinline{pn->p->t = NULL}), or when \lstinline{findNext} returns 0, indicating that the target key \lstinline{k} is in the current node.
+The \lstinline{traverse} function returns an integer flag that is 1 when we reach a leaf node and 0 when we find \lstinline{k} in an existing node; the behavior of operations that call \lstinline{traverse} can vary depending on this flag.
+For instance, an \lstinline{insert} operation may create a new node when the flag is 1 and modify an existing node when the flag is 0, while a \lstinline{lookup} operation may fail when the flag is 1 and return the value in the current node when the flag is 0.
 
-When \lstinline{findNext} arrives at a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking.
-It then releases the lock for the current node, ensuring that the link between the two nodes is not removed or rearranged while traversing it.
+%When \lstinline{traverse} arrives at a new node, it acquires the lock for that node and then releases the lock for the current node, according to the lock-coupling pattern---this ensures that the link between the two nodes is not removed or rearranged while traversing it.
 To utilize the template methodology for concurrent data structure implementations, it is crucial to keep track of the locks acquired for each node during the traversal.
@@
 \label{give-up-algo}
-The give-up template is an algorithm for implementing concurrent search in a tree data structure.
-Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template takes an \emph{optimistic concurrency control} approach \lb{c.f.
-RCU}, only acquiring locks when loading or storing to a node.
-This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios.
-Each node has a range field that serves as a reference point for threads to determine if they are still on the correct path.
-If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again.
-This happens when another concurrent thread restructures the data structure, invalidating the state of the current thread in between releasing one lock and acquiring another.
-For instance, the node we are traveling to may be moved to another part of the data structure intended to hold a different set of keys, or even deleted, before we acquire its lock.
+
+We next consider the give-up template, which uses an \emph{optimistic concurrency control} approach, acquiring fewer locks at the cost of sometimes having to recover from synchronization errors.
+Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires a lock just before operating on a node, and holds at most one lock at any time.
+This means that a conflicting operation may invalidate our traversal, for instance by moving the next node to another part of the data structure before we acquire its lock.
+To guard against this, the give-up template maintains a \emph{range} field on each node indicating the range of keys that should be stored in this node and its successors, and \lstinline{traverse} must check the range field at each node to ensure that it is still on the right path.
+If a check fails, we give up and start the traversal over from the root node.
+The give-up template performs well in scenarios where operations generally do not conflict, either because they are mostly to independent parts of the data structure or because they mostly do not delete or relocate nodes.
 
@@
 		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
-		\caption{The \lstinline{traverse} method of the give-up template algorithm written in ML-like language}
+		\caption{The \lstinline{traverse} method of the give-up template algorithm written in an ML-like language}
 		\label{traverse_giveup_a}	
@@
 Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented.
-The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed in the previous section.
-The purpose of the \lstinline{inRange} function is to determine whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors.
-If the key value is outside the node's range (e.g.
-because the node has been relocated), the search is terminated, and the thread is restarted.
-To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node.
-Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{node_t} in our C implementation.
-These fields store the lower and upper bounds of the keys, respectively.
-In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys.
-This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes.
-If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
+In addition to \lstinline{findNext}, the \lstinline{traverse} function uses a helper function called \lstinline{inRange}, which determines whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors.
+If \lstinline{k} is outside the node's range (e.g.
+because the node has been relocated), the search is restarted from the root node \lstinline{r}.
 
-%\wm{Describe why we need extra pieces in the struct, and how this relates to the other version.}
+The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure?
+Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations.
+However, in most sequential settings the key ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself.
+The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to as many target data structures as possible without modifying them.
+Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t} in our C implementation:
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
@@
 \end{lstlisting}
+These fields store lower and upper bounds on the keys reachable from the current node.
+%In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys.
+This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes.
+If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
+We can then define \lstinline{inRange} as a helper function in the template, rather than requiring data structures to provide it.
+\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
 

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
working through template proofs

@@
 because the node has been relocated), the search is restarted from the root node \lstinline{r}.
+As in the previous section, we implement this in C with a loop: in each iteration, we acquire the lock on the current node, check that \lstinline{k} is in range, and then use \lstinline{findNext} as above, releasing the lock before we move to the next node.
+If the \lstinline{inRange} call fails, we release our current lock and return to the root node by setting the current node to the root pointer \lstinline{p} (lines 21-22 in Figure \ref{traverse_giveup_b}).
 
@@
 
-%% C implementation
+\subsection{Template Proofs} 
+In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification.
+%The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}.
+But first, we need to understand how to specify the correctness of \lstinline{traverse} independently of the data structure operations (insert, lookup, etc.) it will be used to implement.
+The key is the concept of \emph{logical atomicity}~\cite{tada}, a highly compositional approach to specifying concurrent data structure operations.
+\wm{Should this go in background instead?}
 
-When a thread arrives at a node in search of the key $\texttt{k}$, it checks whether $\texttt{k}$ falls within the range of that node.
-If $\texttt{k}$ falls within the range, the thread continues by calling \lstinline{findNext} to find the new node.
-If $\texttt{k}$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).
-
-\subsection{Proofs of Templates} 
-In this section, we present our verification methodology, termed \lstinline{traverse}, which applies to both lock-coupling and give-up templates.
-The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}.
-Before we delve into the intricacies of these verifications, it is essential to familiarize ourselves with the concept of \emph{atomic triples}.
-
-For many concurrent data structures, the ideal correctness condition requires that the data structure behaves identically to a sequential implementation, even when accessed concurrently by multiple threads.
-The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}.
-These take the form of $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
-This triple suggests that if $\texttt{P}_l$ holds true before a call and $\texttt{P}_p$ is true for some value of $a$ in a shared state (e.g.
-a global invariant), then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).
+Logical atomicity is a CSL technique for concisely specifying the behavior of a concurrent operation.
+A logically atomic triple has the form $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread.
+More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).
 For instance, the specification
 $\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$
-expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution.
-The stack itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the \texttt{is\_stack} assertion.
+expresses the fact that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution.
+The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion.
+The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context.
 
-To demonstrate the proof of an implementation that meets the atomic specification, let's take the example of the stack specification mentioned above.
+%To demonstrate the proof of an implementation that meets the atomic specification, let's take the example of the stack specification mentioned above.
 This specification assert that the implementation \texttt{push} appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states.
@@
 
+Logical atomicity is key to the template approach: each template's \lstinline{traverse} function is proved to satisfy a logically atomic specification that says roughly ``this function finds the node where key \lstinline{k} belongs''.
+The traverse specifications can then be used to prove atomic specifications for the individual data structure operations, lifting the sequential specifications for insert, lookup, etc.
+to the concurrent setting.
+
 \subsubsection{Lock-coupling Template}
 \label{traverse_proof_lock}
-Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template, as illustrated in Figure \ref{fig:traverse_lock}.
+
+The logically atomic specification for the lock-coupling template's \lstinline{traverse} function is shown in Figure \ref{fig:traverse_lock}.
+\wm{For the template approach to really work, it seems like all the \lstinline{traverse} functions should have the same specification.
+Can we make that happen?}
 %\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}
 
+%maybe inline this instead of making it a figure?
 \begin{figure}[h]
@@
 
-The specifics of $\nodeboxrep$ and $\treerep$ may vary across implementations, but fundamentally, $\nodeboxrep$ encompasses resources associated with each client thread, while $\treerep$ serves as an abstract representation containing ghost state that links the data structure's concrete state to its abstract equivalent, treated as an atomically accessed shared resource.
-In particular, $\nodeboxrep(\texttt{pn->n})$ acts as a pointer to both the node's lock and lock-invariant assertions, representing a protected resource through an arbitrary separation logic predicate.
-Acquiring this lock allows a thread to access additional elements within the data structure.
+The definitions of $\nodeboxrep$ and $\treerep$ may vary depending on the data structure, but fundamentally, $\nodeboxrep$ holds the resources owned by each client thread, while $\treerep$ holds the abstract representation of the data structure, treated as an atomically accessed shared resource.
+In particular, holding $\nodeboxrep(\texttt{pn->n})$ should allow us to acquire the lock on node \texttt{pn->n}, which in turn gives us access to the contents of the node.
+Formally, we write
+\[\nodeboxrep \texttt{(pn->n)} \triangleq\ \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}\]
+where the \emph{lock invariant assertion} $\ell \lockvar \texttt{R}$ expresses that the lock $\ell$ protects resources \texttt{R}.
+In this case, $R$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{pn->n}).
 
-As previously mentioned, the \lstinline{pn} struct includes two nodes of type \lstinline{node_t}: the $\texttt{p}$ field, signifying the parent node, and the $\texttt{n}$ field, representing the current node.
-We use the assertion $\ell \lockvar \texttt{R}$ to associate a resource invariant \texttt{R} with a lock location $\ell$.
-Thus, formally, we can express $\nodeboxrep \texttt{(pn->n)} \triangleq\ \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$.
-
-Keep in mind that in the lock-coupling style, each thread consistently retains at least one lock while traversing between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
+Recall that in the lock-coupling pattern, each thread consistently retains at least one lock while traversing between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
 Consequently, before calling the \texttt{traverse} function, the thread must acquire a lock for the current node $\texttt{pn->n}$ and obtain the resource safeguarded by the lock.

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
section 3 edits

@@
 \label{traverse_proof_lock}
+\wm{Does this organization work, or would it be better to prove each template as we introduce it?}
 
@@
 
-Recall that in the lock-coupling pattern, each thread consistently retains at least one lock while traversing between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
-Consequently, before calling the \texttt{traverse} function, the thread must acquire a lock for the current node $\texttt{pn->n}$ and obtain the resource safeguarded by the lock.
-This resource, represented as $\texttt{R(pn->n)}$, is parameterized by a node pointer $\texttt{pn->n}$ and describes the node's concrete information.
+Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
+Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{pn->n}$ and obtain the resources $\texttt{R(pn->n)}$ guarded by the lock.
+These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $R$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node.
+\wm{Should we replace this with a concrete definition of the invariant?}
 
-As mentioned earlier, we can define a piece of ghost state, $\texttt{ghost\_node t}$, connecting the node in memory to the public abstract state of the data structure.
+%As mentioned earlier, we can define a piece of ghost state, $\texttt{ghost\_node t}$, connecting the node in memory to the public abstract state of the data structure.
 Specifically, we can have one half of the ghost state contained in the resource $\texttt{R(pn->n)}$, and the other half in the abstract state $\texttt{Node t}$.
@@
 
-In the specification, the local pre-condition includes both the representation of the node $\nodeboxrep(\texttt{pn->n})$ and the resource $\texttt{R(pn->n)}$.
-When the \texttt{traverse} function operates on the entire search structure, it generates a subsequent node, \texttt{n'}, such that the key \texttt{k} falls within the range of node \texttt{n'}.
-The local post-condition is expected to encompass the representation of the new node $\nodeboxrep(\texttt{pn->n'})$, the resource $\texttt{R(pn->n)}$, and a Boolean variable, \texttt{res}, which indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
-The key to verifying the \texttt{traverse} function is identifying the loop invariant for the top-level loop.
-We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
+In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{pn->n})$ and the contents $\texttt{R(pn->n)}$.
+The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}.
+The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{pn->n'})$ \wm{This looks weird: \lstinline{pn} doesn't have an \lstinline{n'} field}, its contents $\texttt{R(pn->n')}$, and a Boolean variable \texttt{res} indicating whether \texttt{pn->n'->t} is \texttt{NULL} or not \wm{Indicate the relationship between \lstinline{res} and the \lstinline{flag} return value}.
+The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
 \begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ \texttt{(k} \in \texttt{range)}\ \land \ \nodeboxrep\texttt{(pn->n')}  \ \ast \  \texttt{R(pn->n')}   
-\end{mathpar}
-The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop.
-The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 6 in Figure \ref{traverse_lock_b}).
-At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition.
-On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (line 11 in Figure \ref{traverse_lock_b}).
-In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.
+\end{mathpar}\wm{Where does \lstinline{range} come from here?}
 
-The proof begins with $\nodeboxrep \texttt{(pn->n)}$, which comprises a lock invariant $\texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$, and a resource $\texttt{R(pn->n)}$ acquired before invoking the \lstinline{traverse} function.
-After \lstinline{findNext} successfully detects the next node (for example, \texttt{n'}) to visit, we acquire the lock of node \texttt{n'} (line 14 in Figure \ref{traverse_lock_b}).
-It means we can get the resource $\texttt{R(pn->n')}$ from $\texttt{pn->n'->lock} \lockvar \texttt{R(pn->n')}$.
-Subsequently, we release the lock of the current node \texttt{p}.
-It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
-By acquiring the lock for the next node \texttt{n'} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it (hand-over-hand locking).
-The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
-
-\than{I feel somewhat perplexed while attempting to articulate the situation mentioned above. Can someone help me fix or explain it more?}
+Given this invariant, the proof proceeds as follows.
+\wm{Should we consider giving annotated code after all?} We begin by checking whether the current node is null (line 5 of Figure~\ref{traverse_lock_b}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with \lstinline{res = true}.
+Otherwise, we use \lstinline{findNext} to find a new node \lstinline{n'} to visit (line 8), which is stored in \lstinline{pn->n}.
+If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with \lstinline{res = false}.
+Otherwise, we acquire \lstinline{pn->n->lock}, gaining access to the resources $\texttt{R(pn->n)}$, and then release the lock of the current node \texttt{pn->p} and return its resources.
+%It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
+By acquiring the lock for the next node \texttt{n} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish \lstinline{traverse_inv} for the new values in \lstinline{pn}.
+%The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
 
@@
 \label{traverse_proof_giveup}
-Moving forward with the proofs, we define the following specification for the traverse function within the give-up template, as depicted in Figure \ref{fig:traverse_giveup}.
+
+The atomic specification for the give-up template's \lstinline{traverse} is shown in Figure \ref{fig:traverse_giveup}.
 
@@
 
-In contrast to the more complex specification of the lock-coupling style \texttt{traverse} function, the give-up template's \texttt{traverse} function has a comparatively simpler specification.
-The local pre-condition for the specification solely comprises $\infp(\texttt{pn->n})$, which is a footprint predicate signifying that node \texttt{n} resides within the search structure's footprint.
-Subsequently, the \texttt{traverse} function operates on the entire search structure, returning a subsequent node \texttt{n'} where \texttt{k} lies within the node \texttt{n'}'s range.
-Therefore, the local post-condition is expected to assert that \texttt{n'} is also positioned within the footprint predicate, that the resource \texttt{R(pn->n')} embodies the concrete information concerning node \texttt{n'}, and that a Boolean \texttt{res} indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
+The give-up template's \texttt{traverse} specification is simpler than that of the lock-coupling template.
+The local precondition include any locks, and consists only of $\infp(\texttt{pn->n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
+As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents \texttt{R(pn->n')}.
+Also as before, the Boolean \texttt{res} indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
 
-In contrast to the lock-coupling style, where each lock is associated with a lock-invariant—an arbitrary separation logic predicate representing the protected resource acquired by a thread when the lock is obtained and must be restored upon lock release—the give-up template has a different approach.
-In this method, every node possesses a lock bit at location $\ell$, which is set to \texttt{true} if and only if the node is locked.
-This protocol is encoded into the definition of the abstract state $\treerep$.
-The protocol is formally defined as $\texttt{inv\_for\_lock} \ \ell \ \texttt{R} \triangleq \exists b.
-\ \ell \mapsto b \ \ast \ \texttt{if } b \ \texttt{then emp else R}$, where a lock at location $\ell$ safeguards resources denoted by \texttt{R}.
+\wm{We really need to explain why we have this difference, or get rid of it.} In contrast to the lock-coupling style, where each lock is associated with a lock-invariant, in the give-up template we treat locks more abstractly.
+For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held.
+Formally, we write \[\texttt{inv\_for\_lock} \ \ell \ \texttt{R} \triangleq \exists b.
+\ \ell \mapsto b \ \ast \ \texttt{if } b \ \texttt{then emp else R}\]\wm{Consider using mathsf for logic and reserving texttt for C code.}
+%where a lock at location $\ell$ safeguards resources denoted by \texttt{R}.
 When the lock is held, meaning the node is locked ($\ell \mapsto \texttt{true}$), threads can remove resources \texttt{R} from $\treerep$.
 On the other hand, when the lock is not held ($\ell \mapsto \texttt{false}$), resources \texttt{R} are required to be part of the invariant.
+and $\treerep$ contains an \texttt{inv\_for\_lock} assertion for each node in the data structure.
+As before, $R$ contains the concrete contents of each node, but also includes the \lstinline{min} and \lstinline{max} fields that we use to implement the \lstinline{inRange} check.
 
-Before proving the \texttt{traverse} function, let's define the loop invariant for the given template of the \texttt{traverse} function as follows:
+Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
 \begin{mathpar} \texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
 \end{mathpar}
-To begin the proof, we start with the $\infp (\texttt{pn->n})$ predicate from the local pre-condition.
-After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\texttt{R(pn->n)}$ and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}.
+With this invariant in hand, the proof of \lstinline{traverse} proceeds as follows.
+We start with the $\infp (\texttt{pn->n})$ predicate from the local precondition.
+After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\texttt{R(pn->n)}$, which allows us to perform the \lstinline{inRange} check on line 7.
+If the check fails (lines 20-23), we immediately release the lock and start over from the beginning.
+Otherwise, we proceed as before: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \lstinline{k} belongs (and can satisfy the postcondition); if \lstinline{findNext} returns 0, we have found the node containing \lstinline{k} (and can satisfy the postcondition); otherwise, we release our lock and re-establish \lstinline{traverse_inv} for the new node indicated by \lstinline{findNext}.
+In this case, we do not hold the lock on the new node, but only the $\infp$ assertion indicating that the new node is in the data structure.
+% and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}.
 Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 8 in Figure \ref{traverse_giveup_b}).
@@
 
-If \lstinline{findNext} cannot find a subsequent node with the key value $\texttt{k}$ within its set of key values, the function terminates.
+%If \lstinline{findNext} cannot find a subsequent node with the key value $\texttt{k}$ within its set of key values, the function terminates.
 Otherwise, it successfully detects the next node (for example, \texttt{n'}) to visit, giving a predicate that asserts \texttt{n'} is in the footprint of the structure, $\infp(\texttt{pn->n})$.

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
finished editing pass

@@
 \section{Verified Data Structures}
-In this section, we demonstrate how to integrate all components into high-level specifications for the purpose of verifying two template algorithms, lock-coupling and give-up, in the real C programming language.
-The target specifications for our search structure(s) are shown below: each operation should atomically have its effect on the data structure, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
+In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified concurrent data structures in C.
+The target specifications for our search structure operations are shown below: each operation should atomically have its effect on the data structure, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
 
@@
 \end{mathpar}
+The intention of the template approach is that we should be able to derive implementations from these operations from \lstinline{traverse} instantiated with an appropriate \lstinline{findNext} function, and prove them correct using the \lstinline{traverse} specs of the previous section, without any further reasoning about specific concurrency patterns.
 
-Our primary focus is on verification the correctness of \texttt{insert} operations in binary search trees using lock-coupling and give-up templates.
-The \texttt{lookup} operations closely resemble the \texttt{insert} operations, sharing a significant portion of their proof.
-As for the \texttt{delete} operations, we have supplied proof solely for the lock-coupling template, which is discussed in Section \ref{internal_reorganize}.
+Our target data structure is a binary search tree (BST) implemented in C.
+We discuss the verification of \texttt{insert} in detail for each template; the \texttt{lookup} proofs follow similar reasoning.
+As for \texttt{delete}, there is a mismatch between the template approach and BST deletion, which we discuss in Section \ref{internal_reorganize}.
+\wm{Should we bother presenting the delete proof for lock-coupling?
+Does it use templates at all?} The complete proofs in VST can be found online at \href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}; we discuss the overall development in Section \ref{proof_mech}.
 
-The complete proof can be found online in our publicly accessible repository of machine-checked proofs (refer to Section \ref{proof_mech}).
-This repository also contains the other proofs featured in our monograph.
+\wm{Should we have a separate section where we introduce findNext and insertOp before we get into specific templates?}
 
-\subsection{Lock-coupling Template}
+\subsection{Lock-coupling BST}
 \begin{figure}[h]
@@
 
-The \texttt{insert} method in Figure \ref{insert_lock} is the main method of lock-coupling template.
-It makes use of an auxiliary \texttt{traverse}  mentioned in Figure \ref{traverse_lock_b}.
-If \texttt{traverse} fails to find the next node (either to left or right), it implies that it has reached a node with a target key \texttt{k} that equals the node's value.
-At this point, we only have to modify the value of that to \texttt{value} (lines 4-5 in Figure \ref{insert_lock}).
-Otherwise, if \texttt{traverse} reaches a leaf, it allocates a new node with key \texttt{k} and value \texttt{v} by calling \texttt{insertOp(pn, k, v)} (line 9 in Figure \ref{insert_lock}).
+Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling BST.
+\wm{Actually, is this \lstinline{insert} function part of the template?
+If so, should we discuss it in section 3?} It makes use of the \texttt{traverse} function of Section \ref{lock-coupling-algo} to find the node at which to insert the key \lstinline{k}.
+If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}).
+Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the data-structure-specific \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
 
-Before moving on to the proof of \texttt{insert} operations for binary search tree, we present the specifications of helper functions, \texttt{findNext} is used in \texttt{traverse} function, and \texttt{insertOp} is used in \texttt{insert} function.
-They are all Hoare triples that operate on the local $\nodeboxrep$ predicate, meaning their proofs need not reason about concurrency.
+We instantiate the template with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}).
+Their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples, and their proofs do not require any concurrency reasoning.
 
@@
 	\centering
+	\wm{I think this should be \lstinline{pn->n->t} $\mapsto$ \lstinline{(k',v', l, r)}, but I'm not sure.}
 	\begin{mathpar}
@@
 	\end{mathpar}
-	\caption{Specification of \texttt{findNext} for the lock-coupling and give-up templates}
+	\caption{Specification of \texttt{findNext} for the binary search tree (all templates) \wm{Shouldn't pn have two fields, p and n?}}
 	\label{fig:findNext_lock}
@@
 	\end{mathpar}
-	\caption{Specification of \texttt{insertOp} for the lock-coupling template}
+	\caption{Specification of \texttt{insertOp} for the lock-coupling template \wm{Is this different for the give-up template? Why?}}
 	\label{fig:insertOp_lock}
@@
 
-The specification of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that a non-null \texttt{t} field points to a \texttt{node} struct containing the current node's key  \texttt{k'}, the node's value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
-When \texttt{findNext} sucessfully locates the next node \texttt{n'}, it sets \texttt{n'} as the left child node \texttt{l} if the provided key \texttt{k} is less than the current node's key \texttt{k'}.
-Conversely, if the given key \texttt{k} is greater than the current node's key \texttt{k'}, the next node \texttt{n'} is set as the right child node \texttt{r}.
-In case where \texttt{findNext} is unable to identify the next node \texttt{n'}, it retains the current node \texttt{n} as \texttt{n'},  indicating no change and that the provided key \texttt{k} is  identical to the current node's key \texttt{k'}.
+\wm{Should we recap the definition of the \lstinline{node} struct here?} The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with the current node's key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
+The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the current node's key \texttt{k'}, and the next node \lstinline{n'} is the left child \lstinline{l}; \lstinline{k} is greater than \lstinline{k'} and \lstinline{n'} is the right child \lstinline{r}; or \lstinline{k} is exactly \lstinline{k'} and \lstinline{n'} is the original node \lstinline{n}, in which case the return value is 0.
 
-The \texttt{insertOp} specification (Figure \ref{fig:insertOp_lock}) indicates that once \texttt{traverse} successfully reaches a leaf node (either the left or right node), the leaf node will have its \texttt{t} field set to \texttt{NULL}.
-This signifies that the leaf node currently has no key or value.
-At this location, a new node with the specified key \texttt{k} and value \texttt{v} should be inserted.
-Upon locating the appropriate leaf node for adding the new key-value pair \texttt{(k, v)}, the process generates two new empty child nodes with corresponding locks, \texttt{(NULL, lk1)} and \texttt{(NULL, lk2)}, which are pointed to by the \texttt{l} and \texttt{r} of the current node.
+The \texttt{insertOp} specification (Figure \ref{fig:insertOp_lock}) takes as input a node whose \lstinline{t} field is \lstinline{NULL}, i.e., an empty leaf node.
+It inserts a new node with the specified key \texttt{k} and value \texttt{v} at that leaf, complete with two new empty child nodes with corresponding locks, \texttt{(NULL, lk1)} and \texttt{(NULL, lk2)}, which are pointed to by the \texttt{l} and \texttt{r} of the current node.
 
@@
 		\left\langle 
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (\texttt{m[k} \mapsto \texttt{v]})
 		\right\rangle
@@
 \end{mathpar}
-
-Similar to the \texttt{traverse} specification, the \texttt{insert} function's specification is composed of two components: local and public pre- and postconditions.
+\wm{Consider skipping this paragraph if we need space.} Similar to the \texttt{traverse} specification, the \texttt{insert} function's specification is composed of two components: local and public pre- and postconditions.
 The local pre- and postconditions, containing $\nodeboxrep \ \texttt{p}$, describe the binary search tree's concrete state with the root node $\texttt{p}$, which points to the node's lock and lock invariant assertion.
@@
 
-The proof for \texttt{insert} begins with preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$.
-We set \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_lock}).
-After the lock for the root node is acquired (line 4 in Figure \ref{insert_lock}), the thread accesses information within the root node's lock invariant, \texttt{R(pn->n)}.
-As seen in Section \ref{traverse_proof_lock}, we supply the same precondition for the \texttt{traverse} function, which includes local precondition $\nodeboxrep \texttt{(pn->n)} \ast \texttt{R(pn->n)}$ and public precondition $\treerep\ \texttt{m}$.
-We then apply the \texttt{traverse} specification from Section \ref{traverse_proof_lock}.
+The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{p}$.
+We set \texttt{pn->n} to point to the root node \texttt{*p} (line 3 in Figure \ref{insert_lock}).
+After the lock for the root node is acquired (line 4 in Figure \ref{insert_lock}), the thread acquires the contents of the root node, \texttt{R(pn->n)}.
+We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
 
-In the \texttt{then} branch, when the \texttt{traverse} fails to find the next node, resulting in $\texttt{res} = \texttt{false}$, it implies that $\texttt{pn->n'->t} \neq \texttt{NULL}$.
-This indicates that the key to be inserted already exists in the tree.
-In this scenario, we update the node's value while maintaining the rest of the tree structure (line 6 in Figure \ref{insert_lock}); this acts as the linearization point for the insertion.
-Conversely, when the \texttt{traverse} reaches a leaf and produces $\texttt{res} = \texttt{true}$, it implies $\texttt{pn->n'->t} = \texttt{NULL}$ (line 9 in Figure \ref{insert_lock}).
-We then utilize the \texttt{insertOp} specification, and with enough information, we can conclude that inserting the key at this location will alter the abstract state from an unknown $\texttt{m}$ to $\texttt{m[k} \mapsto \texttt{v]}$.
-In both instances, we must prove that we can atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, by demonstrating that the in-memory tree update corresponds to a local modification in the abstract tree, which inserts the new key-value pair.
-Next, we release (line 11 in Figure \ref{insert_lock}) the node \texttt{n} and return it to the shared state.
-Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
+In the \texttt{then} branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
+we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree.
+In this scenario, we update the node's value while maintaining the rest of the tree structure (line 6 in Figure \ref{insert_lock}).
+Otherwise, when \texttt{traverse} reaches a leaf and produces $\texttt{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 9 in Figure \ref{insert_lock}).
+We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree.
+In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $\texttt{m}$ to $\texttt{m[k} \mapsto \texttt{v]}$, thereby satisfying the public postcondition of \lstinline{insert}.
+We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
+We can then release the lock (line 11 in Figure \ref{insert_lock}) return the node's contents to the shared state.
+%Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
+\wm{Do we ever define $\treerep$?}
 
-%\wm{I think this is the wrong organization.
-The proofs of traverse should be presented in the previous section, where we describe traverse.
-This section should show findNext and insert/lookup for specific data structure(s), and then describe how we combine them into top-level specs.}
-\subsection{Give-up Template}
+\subsection{Give-up BST}
 
-Using the give-up template, the \texttt{insert} method in Figure \ref{insert_giveup} is not significantly different from the \texttt{insert} method of the lock-coupling template.
-The primary distinction is that it does not require acquiring a lock before calling the \texttt{traverse} function.
-Furthermore, it utilizes an auxiliary \texttt{traverse} function, as illustrated in Figure \ref{fig:traverse_giveup}.
-This function shares the same objective as the previously discussed \texttt{traverse} function in the lock-coupling approach.
+The give-up version of the \texttt{insert} method (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version.
+The one difference is that it does not require acquiring a lock before calling its \texttt{traverse} function.
 
-The give-up template retains the specification of the \texttt{findNext} helper function from the lock-coupling template while introducing an extra helper function called \texttt{inRange}.
-Additionally, the \texttt{insertOp} specification exhibits subtle differences from that of the lock-coupling template, as it incorporates two new fields,  \texttt{min} and \texttt{max},  into each \texttt{node\_t} type node, as showcased in Section  \ref{give-up-algo}.
+The give-up BST uses the same definition and specification of \texttt{findNext}, while introducing an extra helper function called \texttt{inRange} \wm{If \lstinline{inRange} is part of the template, I think we should discuss it in the previous section}.
+The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.
+\wm{It would be nice if \lstinline{insertOp} was the same.
+Maybe the template should include an \lstinline{alloc_node_t} function that \lstinline{insertOp} calls?}
 
@@
 				\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{(t, lock, min, max)} \  \\ 
-				\ \ast  \ \texttt{(if res then } \texttt{(min < k} \land \texttt{k < max)} \land \texttt{(k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max))}
+				\ \ast  \ \texttt{(if res then } \texttt{(min < k < max)} \texttt{ else } \texttt{(k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max))}
 			\end{array}
@@
 
-The \texttt{inRange} specification (Figure \ref{fig:inRange_insertOp_giveup}) simply necessitates that \texttt{inRange(pn, k)}  yields a boolean value, \texttt{res}.
-When \texttt{res} is \texttt{true}, it signifies that \texttt{k} is within the range of \texttt{(min, max)} associated with the node \texttt{n}.
-Conversely, \texttt{res} is \texttt{false}, it indicates that \texttt{k} is outside the boundaries of \texttt{(min, max)}.
+The \texttt{inRange} function (Figure \ref{fig:inRange_insertOp_giveup}) simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
 
-As previously mentioned, the \texttt{insertOp} specification (Figure \ref{fig:inRange_insertOp_giveup}) exhibits slight differences from that of the lock-coupling template, primarily due to the incorporation of two extra fields.
-As a result, node \texttt{n} needs point-to predicates for these new fields.
-Moreover, the current node's \texttt{l} and \texttt{r} point to two new empty child nodes with associated locks and ranges: \texttt{(NULL, lk1, (min, k))} and \texttt{(NULL, lk2, (k, max))}.
-In the context of binary search trees, a node inherits its range from its parent based on the parent's key; thus, if node \texttt{n} has range \texttt{(min, max)} and key \texttt{k}, its left child will have the range \texttt{(min, k)} and its right child the range \texttt{(k, max)}.
+As previously mentioned, the \texttt{insertOp} specification (Figure \ref{fig:inRange_insertOp_giveup}) is slightly different from the lock-coupling version, to account for the added \lstinline{min} and \lstinline{max} fields.
+In a BST, if the current node's range is \lstinline{(min, max)} and its key is \lstinline{k}, its left child will have range \texttt{(min, k)} and its right child will have range \texttt{(k, max)}.
+Otherwise, \lstinline{insertOp} behaves identically, storing \lstinline{k} and \lstinline{v} in the current node and creating two empty child nodes.
 
-While the \texttt{insert} specification interface for the give-up template is identical to that of the lock-coupling template, a few adjustments need to be made within $\nodeboxrep\ \texttt{p}$ and $\treerep \ \texttt{m}$.
-Essentially, the \texttt{insert} function's specification comprises two main components: local and public preconditions and postconditions, as previously mentioned.
+The \texttt{insert} specification for the give-up template is identical to that of the lock-coupling template, up to changes in the definitions of $\nodeboxrep\ \texttt{p}$ and $\treerep \ \texttt{m}$.
+%Essentially, the \texttt{insert} function's specification comprises two main components: local and public preconditions and postconditions, as previously mentioned.
 Moreover, we must demonstrate that it is possible to atomically update the tree's abstract state from an unknown \texttt{m} to $\texttt{m[k} \mapsto \texttt{v])}$.
-
-The local preconditions and postconditions, containing $\nodeboxrep \ \texttt{p}$, depict the binary search tree's concrete state, where the root node $\texttt{p}$ points to the node's lock and footprint predicate, signifying that the root node is located within the search structure's footprint.
+\wm{I think we should try to explain what those changes actually are.}
+%The local preconditions and postconditions, containing $\nodeboxrep \ \texttt{p}$, depict the binary search tree's concrete state, where the root node $\texttt{p}$ points to the node's lock and footprint predicate, signifying that the root node is located within the search structure's footprint.
 In contrast, the public preconditions and postconditions are denoted by the predicate  $\treerep \ \texttt{m}$, which is identical to the public predicates of the \texttt{traverse} function in the give-up template.
 
-The proof for the \texttt{insert} operation begins with the preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$.
+%The proof for the \texttt{insert} operation begins with the preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$.
 We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_giveup}).
@@
 
-Analogous to the lock-coupling template's \texttt{insert} operation, there are two possible branches representing the outcome following the execution of \texttt{traverse}.
+%Analogous to the lock-coupling template's \texttt{insert} operation, there are two possible branches representing the outcome following the execution of \texttt{traverse}.
 Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ \texttt{(m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.
+\wm{Is there any difference between the \lstinline{insert} proofs for lock-coupling and give-up, or are all the differences in the \lstinline{traverse}?
+I don't see any difference in the proof outline here.}
 
@@
 \label{internal_reorganize}
-Our original goal was to verify an existing concurrent BST implementation with the template approach, refactoring as necessary but otherwise leaving the code unchanged.
-However, there is a fundamental mismatch between the template approach and the standard approach to deletion in a BST.
+The third BST operation, \lstinline{delete}, presents problems in applying the template approach.
 The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}).
@@
 This contrasts with linked-list deletion, which only involves the target node and its immediate neighbors.
-In practice, the existing template proofs for tree structures either do not include deletion at all, or use a very lightweight form of deletion in which nodes are marked as ``deleted'' but are never actually removed from the data structure.
+In practice, Krishna et al.'s proofs for tree structures~\cite{templates} either do not include deletion at all, or use a very lightweight form of deletion in which nodes are marked as ``deleted'' but are never actually removed from the data structure.
 
@@
 \label{proof_mech}
-We present the mechanized formal verification of the correctness of these templates, which are implemented in C and verified with the Verified Software Toolchain.
-We employ concurrent separation logic to analyze the behavior of the C code and incorporate Iris-style logical atomicity within VST.
+All the proofs described above have been mechanized in VST, using its extensions for logical atomicity~\cite{iris-vst-arxiv}. Statistics on the verification effort are shown in Table~\ref{table_coq}
 
@@
 
-By utilizing VST, we can create programs in standard C, automatically generate abstract syntax trees (ASTs) for them in Coq, and subsequently state and interactively prove specifications for these programs with the support of automated tactics.
+%By utilizing VST, we can create programs in standard C, automatically generate abstract syntax trees (ASTs) for them in Coq, and subsequently state and interactively prove specifications for these programs with the support of automated tactics.
 Table \ref{table_coq} presents a summary of our Coq development.
@@
 This encompasses supplementary proofs, template specifications (encompassing all helper functions such as \texttt{findNext}, \texttt{insertOp}, and others, in addition to \texttt{traverse}, \texttt{insert}, \texttt{lookup}, and \texttt{delete}), as well as proofs against with lock-coupling and give-up template specifications.
-All proofs are formalized in Coq using VST and can be found online at
+\wm{Rather than re-stating what we proved, this is the place to discuss the development: e.g., which parts took the most effort and LoC, which parts we expect to be reusable and in what way, how much bigger or smaller the template proofs were than non-template versions, etc.}
+
+The proofs can be found online at
 
@@
 \section{Conclusion}
+Concurrent search structure templates are a promising approach for proving the correctness of concurrency patterns and data structures separately, and then combining them to obtain verified concurrent data structures.
+The approach translates smoothly to real C programs in VST, with appropriate imperative versions of the key \lstinline{traverse} functions that define the templates.
+However, the translation also exposed some ambiguities and limitations in the approach: for instance, the \lstinline{inRange} function of the give-up template must either be assumed to exist in the data structure or incorporated into the template, and the \lstinline{delete} operation of binary search trees does not decompose naturally into a concurrent part and a data-structure-specific part.
+\wm{Do we have future work planned?
+How do we want to fix this?}
+Ultimately, we hope to use the template approach to verify data structures such as Masstree~\cite{masstree} and Wormhole~\cite{wormhole}, complicated real-world search structures that combine multiple concurrency patterns for maximum performance on multicore architectures.
 

### 2023-04-28 concurrency/CONCUR 2023/paper.tex
updating comments

@@
 \newcommand{\lb}[1]{\textbf{\textcolor{red}{[Lennart: #1]}}}
-\newcommand{\wm}[1]{\textbf{\textcolor{violet}{[Willam: #1]}}}
+\newcommand{\wm}[1]{\textbf{\textcolor{violet}{[William: #1]}}}
 \newcommand{\ignore}[1]{}
@@
 
-%TODO mandatory: add short abstract of the document
+%don't say ``real'' so much
 \begin{abstract}
@@
 In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
+\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}
 
@@
 The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine when the current node does not contain the target key.
-\wm{Show an example template here, and enough of the specification to explain atomicity} Once the appropriate node for the key has been found, the template returns it to a top-level function that calls out to the data structure to perform the appropriate operation (e.g.
+\wm{Show an example template here, and enough of the specification to explain atomicity -- one findNext, one traverse, one top-level insert -- see fig 3 in template paper} Once the appropriate node for the key has been found, the template returns it to a top-level function that calls out to the data structure to perform the appropriate operation (e.g.
 insert, lookup, delete) on that node.
@@
 The core of each search structure template is a \lstinline{traverse} function that navigates a data structure using the target synchronization mechanism.
-Templates abstract away from the data structure being traversed, operating on a generic \lstinline{node} type and taking data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters.
+\wm{\lstinline{traverse} is the core, but it's not the whole template!
+The data structure never sees traverse, but rather a wrapper around it that also acquires locks, etc.
+as necessary.} Templates abstract away from the data structure being traversed, operating on a generic \lstinline{node} type and taking data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters.
 We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
@@
 The logically atomic specification for the lock-coupling template's \lstinline{traverse} function is shown in Figure \ref{fig:traverse_lock}.
-\wm{For the template approach to really work, it seems like all the \lstinline{traverse} functions should have the same specification.
-Can we make that happen?}
+\wm{Note that this isn't the same as give-up, and that reflects the original paper.
+We might be able to put a wrapper function that acquires the initial lock for lock-coupling?}
 %\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}
@@
 These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $R$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node.
-\wm{Should we replace this with a concrete definition of the invariant?}
+\wm{Should we replace this with a concrete definition of the invariant?
+e.g.
+$R = ghost_node(k, v, range) * ...$}
 
@@
 As for \texttt{delete}, there is a mismatch between the template approach and BST deletion, which we discuss in Section \ref{internal_reorganize}.
-\wm{Should we bother presenting the delete proof for lock-coupling?
-Does it use templates at all?} The complete proofs in VST can be found online at \href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}; we discuss the overall development in Section \ref{proof_mech}.
+The complete proofs in VST can be found online at \href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}; we discuss the overall development in Section \ref{proof_mech}.
 
@@
 Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ \texttt{(m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.
-\wm{Is there any difference between the \lstinline{insert} proofs for lock-coupling and give-up, or are all the differences in the \lstinline{traverse}?
-I don't see any difference in the proof outline here.}
+\wm{The proof for give-up is slightly different from the lock-coupling version, because the lock-coupling one needs to acquire an extra lock to satisfy the precondition of its traverse.}
 
@@
 
+%future work: how generically can we implement the templates, and then be able to apply them to real existing data structure implementations?
+%say something about the fact that BST existed before this work, rather than being written to the template?
+
 %%

### 2023-04-28 concurrency/CONCUR 2023/paper.tex
intro edits

@@
 Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort.
-In this paper, we implement the template approach in the Verified Software Toolchain, and use it to prove correctness of C implementations of fine-grained concurrent data structures.
-In the process, we explore some of the practical requirements and limitations of the template approach on real code.
-%What's our conclusion?
-Are the template proofs easier/fewer lines than the proofs without the templates?
-We find that the template approach is generally effective on real C implementations, although some templates require modifications to the target data structure, and some data structures have operations that are not easily decomposed.
-This work represents a step towards practical compositional verification of real-world concurrent data structures.
+In this paper, we implement the template approach in the Verified Software Toolchain (VST), and use it to prove correctness of C implementations of fine-grained concurrent data structures.
+This involves translating code, specifications, and proofs to the idiom of C and VST, and gives us another look at the requirements and limitations of the template approach.
+We encounter several questions about the boundaries between template and data structure, as well as some common data structure operations that cannot naturally be decomposed into templates.
+Nonetheless, the approach appears promising for modular verification of real-world concurrent data structures.
 \end{abstract}
@@
 In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
-\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}
+%\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}
 
@@
 \item We reimplement the template approach without flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/subtree''.
-\item We implement the template approach in VST, allowing us to apply it to real C programs and obtain end-to-end correctness proofs in a single verification system.
-\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real programming language.
-\item In the process, we encounter some of the limitations of the template approach on tree data structures. \wm{It would be nice if we had a possible solution!}
+\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
+\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real programming language, and its first application to data structures not written specifically as case studies.
+\item We highlight several questions about the boundaries of the template/data structure division, and describe a limitation of the template approach on tree data structures.
 \end{itemize}
-%\wm{This will be useful for contributions: Than, what would you say the hardest part was in completing the proofs?}
-
-%\than{The most challenging aspect of the project was the proofs involving traversal.
-It took me a considerable amount of time to understand the pre- and post-conditions, as well as determining the appropriate loop invariants.
-I found myself frequently switching between proving the \texttt{insertOp} and \texttt{traverse} functions in order to combine all elements.
-However, once I had completed the \texttt{traverse} proofs, everything became much easier.}
 

### 2023-04-28 concurrency/CONCUR 2023/paper.tex
more comment updates

@@
 \item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
-\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real programming language, and its first application to data structures not written specifically as case studies.
+\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
 \item We highlight several questions about the boundaries of the template/data structure division, and describe a limitation of the template approach on tree data structures.
@@
 To apply the template approach to real-world code, we instead use the Verified Software Toolchain (VST)~\cite{plcc}, a separation-logic-based verifier for C programs.
-VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, giving it a realistic semantics and a guarantee that proved properties will hold on compiled code.
+VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, allowing it to guarantee that proved properties will hold on compiled code.
 Recent work on VST~\cite{iris-vst-arxiv} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications.
@@
 \wm{Note that this isn't the same as give-up, and that reflects the original paper.
-We might be able to put a wrapper function that acquires the initial lock for lock-coupling?}
+It's the cssOp/insert function that has the same spec in both.}
 %\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}
@@
 e.g.
-$R = ghost_node(k, v, range) * ...$}
+$R = ghost\_node(k, v, range) * ...$}
 
@@
 
-Our target data structure is a binary search tree (BST) implemented in C.
-We discuss the verification of \texttt{insert} in detail for each template; the \texttt{lookup} proofs follow similar reasoning.
+Our target data structure is a binary search tree (BST) implemented in C, a data structure that was already defined and verified as a demonstration of VST's concurrency capabilities.
+We factor out \lstinline{findNext} and \lstinline{traverse} functions as described above, and verify lock-coupling and give-up versions of the BST using the templates.
+We discuss the verification of \texttt{insert} for each template in detail; the \texttt{lookup} proofs follow similar reasoning.
 As for \texttt{delete}, there is a mismatch between the template approach and BST deletion, which we discuss in Section \ref{internal_reorganize}.
@@
 
-\wm{Should we have a separate section where we introduce findNext and insertOp before we get into specific templates?}
+\wm{Should we have a separate section where we introduce findNext (and maybe insertOp) before we get into specific templates?}
 
@@
 	\end{mathpar}
-	\caption{Specification of \texttt{insertOp} for the lock-coupling template \wm{Is this different for the give-up template? Why?}}
+	\caption{Specification of \texttt{insertOp} for the lock-coupling template}
 	\label{fig:insertOp_lock}
@@
 The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.
-\wm{It would be nice if \lstinline{insertOp} was the same.
-Maybe the template should include an \lstinline{alloc_node_t} function that \lstinline{insertOp} calls?}
 
@@
 
-The \texttt{insert} specification for the give-up template is identical to that of the lock-coupling template, up to changes in the definitions of $\nodeboxrep\ \texttt{p}$ and $\treerep \ \texttt{m}$.
+%The \texttt{insert} specification for the give-up template is identical to that of the lock-coupling template, up to changes in the definitions of $\nodeboxrep\ \texttt{p}$ and $\treerep \ \texttt{m}$.
 %Essentially, the \texttt{insert} function's specification comprises two main components: local and public preconditions and postconditions, as previously mentioned.
 Moreover, we must demonstrate that it is possible to atomically update the tree's abstract state from an unknown \texttt{m} to $\texttt{m[k} \mapsto \texttt{v])}$.
-\wm{I think we should try to explain what those changes actually are.}
 %The local preconditions and postconditions, containing $\nodeboxrep \ \texttt{p}$, depict the binary search tree's concrete state, where the root node $\texttt{p}$ points to the node's lock and footprint predicate, signifying that the root node is located within the search structure's footprint.
@@
 Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ \texttt{(m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.
-\wm{The proof for give-up is slightly different from the lock-coupling version, because the lock-coupling one needs to acquire an extra lock to satisfy the precondition of its traverse.}
+
+The proof of the \lstinline{insert} function is similar to that of the previous function: the only difference is that the give-up version of \lstinline{insert} does not acquire a lock before calling \lstinline{traverse}, reflecting the difference in the locking patterns and the preconditions of \lstinline{traverse} between the two templates.
 
@@
 
-\section{Conclusion}
+\section{Conclusion and Future Work}
 Concurrent search structure templates are a promising approach for proving the correctness of concurrency patterns and data structures separately, and then combining them to obtain verified concurrent data structures.
-The approach translates smoothly to real C programs in VST, with appropriate imperative versions of the key \lstinline{traverse} functions that define the templates.
+We have translated the approach to apply to C programs using VST, with appropriate imperative versions of the key \lstinline{traverse} functions that define the templates, and used it to prove the correctness of an (appropriately refactored) existing concurrent data structure.
 However, the translation also exposed some ambiguities and limitations in the approach: for instance, the \lstinline{inRange} function of the give-up template must either be assumed to exist in the data structure or incorporated into the template, and the \lstinline{delete} operation of binary search trees does not decompose naturally into a concurrent part and a data-structure-specific part.
-\wm{Do we have future work planned?
-How do we want to fix this?}
+\wm{Do we have an idea of how to fix this?}
+The next step is to implement templates as a truly generic framework, with proofs of \lstinline{traverse}, etc.
+that can be freely combined with verified sequential data structures via a clear interface that defines the functions a data structure must provide and the assumptions the template makes on them.
 Ultimately, we hope to use the template approach to verify data structures such as Masstree~\cite{masstree} and Wormhole~\cite{wormhole}, complicated real-world search structures that combine multiple concurrency patterns for maximum performance on multicore architectures.
-
-%future work: how generically can we implement the templates, and then be able to apply them to real existing data structure implementations?
-%say something about the fact that BST existed before this work, rather than being written to the template?
 

### 2023-04-28 concurrency/CONCUR 2023/paper.tex
resolved some William's comments

@@
 		{\color{blue}
-			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
+			\left\langle \texttt{res}. \ \exists \  \texttt{n'}.
 			\begin{array}{c}
@@
 The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ \texttt{(k} \in \texttt{range)}\ \land \ \nodeboxrep\texttt{(pn->n')}  \ \ast \  \texttt{R(pn->n')}   
+\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ \texttt{(k} \in \texttt{range(n'))}\ \land \ \nodeboxrep\texttt{(pn->n')}  \ \ast \  \texttt{R(pn->n')}   
 \end{mathpar}\wm{Where does \lstinline{range} come from here?}
@@
 		{\color{blue}
-			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
+			\left\langle \texttt{res}. \ \exists \ \texttt{n'}.
 			\begin{array}{c}
@@
 
-\wm{We really need to explain why we have this difference, or get rid of it.} In contrast to the lock-coupling style, where each lock is associated with a lock-invariant, in the give-up template we treat locks more abstractly.
+%\wm{We really need to explain why we have this difference, or get rid of it.} In contrast to the lock-coupling style, where each lock is associated with a lock-invariant, in the give-up template we treat locks more abstractly.
 For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held.
@@
 	\centering
-	\wm{I think this should be \lstinline{pn->n->t} $\mapsto$ \lstinline{(k',v', l, r)}, but I'm not sure.}
+	%\wm{I think this should be \lstinline{pn->n->t} $\mapsto$ \lstinline{(k',v', l, r)}, but I'm not sure.}
 	\begin{mathpar}
@@
 			\begin{array}{c}
-				\texttt{pn}\mapsto\texttt{n} \ast \texttt{n}\mapsto\texttt{t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} 
+				\texttt{pn->n->t}\mapsto \texttt{(k', v', l, r)} 
 			\end{array}
@@
 		{\color{blue}
-			\left\{\exists \  \texttt{res n'.}
+			\left\{\texttt{res}. \ \exists \  \texttt{n'.}
 			\begin{array}{c}
-				\texttt{pn}\mapsto\texttt{n} \ast \texttt{n}\mapsto\texttt{t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} \\\  
+				\texttt{pn->n->t}\mapsto \texttt{(k', v', l, r)}  \\\  
 				\ast \ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n = n'} \land \texttt{k = k'))}
@@
 	\end{mathpar}
-	\caption{Specification of \texttt{findNext} for the binary search tree (all templates) \wm{Shouldn't pn have two fields, p and n?}}
+	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)} %\wm{Shouldn't pn have two fields, p and n?}
 	\label{fig:findNext_lock}
@@
 			\begin{array}{c}
-				\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} 
+				\ \texttt{pn->n->t}  
 			\end{array}
@@
 			\begin{array}{c}
-				\ \texttt{pn} \mapsto \texttt{n} \ \ast \ \texttt{n}\mapsto \texttt{t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
+				\ \texttt{pn->n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
 				\ \texttt{l} \mapsto \texttt{(NULL, lk1)}\  \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2)}
@@
 			\begin{array}{c}
-				\texttt{pn}\mapsto\texttt{n} \ \ast \texttt{n}\mapsto\texttt{(t, lock, min, max)} 
+				\texttt{pn->n}\mapsto\texttt{(t, lock, min, max)} 
 			\end{array}
@@
 			\begin{array}{c}
-				\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{(t, lock, min, max)} \  \\ 
+				\ \texttt{pn->n}\mapsto\texttt{(t, lock, min, max)} \  \\ 
 				\ \ast  \ \texttt{(if res then } \texttt{(min < k < max)} \texttt{ else } \texttt{(k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max))}
@@
 However, the translation also exposed some ambiguities and limitations in the approach: for instance, the \lstinline{inRange} function of the give-up template must either be assumed to exist in the data structure or incorporated into the template, and the \lstinline{delete} operation of binary search trees does not decompose naturally into a concurrent part and a data-structure-specific part.
-\wm{Do we have an idea of how to fix this?}
+%\wm{Do we have an idea of how to fix this?}
 The next step is to implement templates as a truly generic framework, with proofs of \lstinline{traverse}, etc.

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
restructured some sections

@@
 
+\subsubsection{Logical atomicity} 
+%In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification.
+%The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}.
+
+%But first, we need to understand how to specify the correctness of \lstinline{traverse} independently of the data structure operations (insert, lookup, etc.) it will be used to implement.
+The key is the concept of \emph{logical atomicity}~\cite{tada}, a highly compositional approach to specifying concurrent data structure operations.
+\wm{Should this go in background instead?}
+
+Logical atomicity is a concurrent separation logic (CSL) technique for concisely specifying the behavior of a concurrent operation.
+A logically atomic triple has the form $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread.
+More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).
+For instance, the specification
+$\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$
+expresses the fact that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution.
+The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion.
+The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context.
+
+%To demonstrate the proof of an implementation that meets the atomic specification, let's take the example of the stack specification mentioned above.
+This specification assert that the implementation \texttt{push} appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states.
+The local preconditions and postconditions, \texttt{is\_stack}, contain the points-to and lock-invariant assertions necessary for a thread, while the public preconditions and postconditions, \texttt{stack}, involve the ghost state that defines the overall state of the stack.
+Ghost state is a type of program state that aids in the verification process without affecting the program's runtime behavior, and by referring to an arbitrary ghost state identifier $g$ (which we will generally omit), it is possible to connect local and public conditions.
+To show that the implementation of \texttt{push}(v) satisfies these atomic specifications, two things must be demonstrated.
+Firstly, the implementation can execute safely without owning any piece of public assertion $\texttt{stack}$, only accessing it automatically and restoring it up until the linearization point.
+Secondly, at some linearization point, the implementation transforms the current stack $s$ into $v::s$ atomically, which satisfies the public postcondition $\texttt{stack}(v::s)$.
+
+Logical atomicity is key to the template approach: each template's \lstinline{traverse} function is proved to satisfy a logically atomic specification that says roughly ``this function finds the node where key \lstinline{k} belongs''.
+The traverse specifications can then be used to prove atomic specifications for the individual data structure operations, lifting the sequential specifications for insert, lookup, etc.
+to the concurrent setting.
+
 \subsection{Iris and VST}
@@
 
-\subsubsection{Lock-Coupling Template}
+\subsection{Lock-Coupling Template}
 \label{lock-coupling-algo}
@@
 
-\begin{figure}[ht]
+\begin{figure}[h]
 	\begin{subfigure}[t]{0.45\textwidth}
@@
 
-\subsubsection{Give-Up Template}
-\label{give-up-algo}
-
-We next consider the give-up template, which uses an \emph{optimistic concurrency control} approach, acquiring fewer locks at the cost of sometimes having to recover from synchronization errors.
-Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires a lock just before operating on a node, and holds at most one lock at any time.
-This means that a conflicting operation may invalidate our traversal, for instance by moving the next node to another part of the data structure before we acquire its lock.
-To guard against this, the give-up template maintains a \emph{range} field on each node indicating the range of keys that should be stored in this node and its successors, and \lstinline{traverse} must check the range field at each node to ensure that it is still on the right path.
-If a check fails, we give up and start the traversal over from the root node.
-The give-up template performs well in scenarios where operations generally do not conflict, either because they are mostly to independent parts of the data structure or because they mostly do not delete or relocate nodes.
-
-\begin{figure}[!ht]
-	\begin{subfigure}[t]{0.45\textwidth}
-		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
-		\caption{The \lstinline{traverse} method of the give-up template algorithm written in an ML-like language}
-		\label{traverse_giveup_a}	
-	\end{subfigure}\qquad
-	\begin{subfigure}[t]{0.48\textwidth}
-		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
-		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
-		\label{traverse_giveup_b}
-	\end{subfigure}
-	\caption{The \lstinline{traverse} method of the give-up template algorithms}
-	\label{traverse_giveup}
-\end{figure}
-
-Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented.
-In addition to \lstinline{findNext}, the \lstinline{traverse} function uses a helper function called \lstinline{inRange}, which determines whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors.
-If \lstinline{k} is outside the node's range (e.g.
-because the node has been relocated), the search is restarted from the root node \lstinline{r}.
-As in the previous section, we implement this in C with a loop: in each iteration, we acquire the lock on the current node, check that \lstinline{k} is in range, and then use \lstinline{findNext} as above, releasing the lock before we move to the next node.
-If the \lstinline{inRange} call fails, we release our current lock and return to the root node by setting the current node to the root pointer \lstinline{p} (lines 21-22 in Figure \ref{traverse_giveup_b}).
-
-The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure?
-Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations.
-However, in most sequential settings the key ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself.
-The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to as many target data structures as possible without modifying them.
-Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t} in our C implementation:
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
-	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
-\end{lstlisting}
-These fields store lower and upper bounds on the keys reachable from the current node.
-%In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys.
-This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes.
-If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
-We can then define \lstinline{inRange} as a helper function in the template, rather than requiring data structures to provide it.
-\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
-
-\subsection{Template Proofs} 
-In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification.
-%The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}.
-But first, we need to understand how to specify the correctness of \lstinline{traverse} independently of the data structure operations (insert, lookup, etc.) it will be used to implement.
-The key is the concept of \emph{logical atomicity}~\cite{tada}, a highly compositional approach to specifying concurrent data structure operations.
-\wm{Should this go in background instead?}
-
-Logical atomicity is a CSL technique for concisely specifying the behavior of a concurrent operation.
-A logically atomic triple has the form $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
-Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread.
-More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).
-For instance, the specification
-$\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$
-expresses the fact that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution.
-The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion.
-The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context.
-
-%To demonstrate the proof of an implementation that meets the atomic specification, let's take the example of the stack specification mentioned above.
-This specification assert that the implementation \texttt{push} appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states.
-The local preconditions and postconditions, \texttt{is\_stack}, contain the points-to and lock-invariant assertions necessary for a thread, while the public preconditions and postconditions, \texttt{stack}, involve the ghost state that defines the overall state of the stack.
-Ghost state is a type of program state that aids in the verification process without affecting the program's runtime behavior, and by referring to an arbitrary ghost state identifier $g$ (which we will generally omit), it is possible to connect local and public conditions.
-To show that the implementation of \texttt{push}(v) satisfies these atomic specifications, two things must be demonstrated.
-Firstly, the implementation can execute safely without owning any piece of public assertion $\texttt{stack}$, only accessing it automatically and restoring it up until the linearization point.
-Secondly, at some linearization point, the implementation transforms the current stack $s$ into $v::s$ atomically, which satisfies the public postcondition $\texttt{stack}(v::s)$.
-
-Logical atomicity is key to the template approach: each template's \lstinline{traverse} function is proved to satisfy a logically atomic specification that says roughly ``this function finds the node where key \lstinline{k} belongs''.
-The traverse specifications can then be used to prove atomic specifications for the individual data structure operations, lifting the sequential specifications for insert, lookup, etc.
-to the concurrent setting.
-
-\subsubsection{Lock-coupling Template}
+\subsubsection{Lock-coupling Proof}
 \label{traverse_proof_lock}
-\wm{Does this organization work, or would it be better to prove each template as we introduce it?}
+%\wm{Does this organization work, or would it be better to prove each template as we introduce it?}
 
@@
 %maybe inline this instead of making it a figure?
-\begin{figure}[h]
+\begin{figure}[ht]
 	\centering
@@
 %The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
+
+\subsection{Give-Up Template}
+\label{give-up-algo}
+
+We next consider the give-up template, which uses an \emph{optimistic concurrency control} approach, acquiring fewer locks at the cost of sometimes having to recover from synchronization errors.
+Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires a lock just before operating on a node, and holds at most one lock at any time.
+This means that a conflicting operation may invalidate our traversal, for instance by moving the next node to another part of the data structure before we acquire its lock.
+To guard against this, the give-up template maintains a \emph{range} field on each node indicating the range of keys that should be stored in this node and its successors, and \lstinline{traverse} must check the range field at each node to ensure that it is still on the right path.
+If a check fails, we give up and start the traversal over from the root node.
+The give-up template performs well in scenarios where operations generally do not conflict, either because they are mostly to independent parts of the data structure or because they mostly do not delete or relocate nodes.
+
+\begin{figure}[!ht]
+	\begin{subfigure}[t]{0.45\textwidth}
+		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
+		\caption{The \lstinline{traverse} method of the give-up template algorithm written in an ML-like language}
+		\label{traverse_giveup_a}	
+	\end{subfigure}\qquad
+	\begin{subfigure}[t]{0.48\textwidth}
+		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
+		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
+		\label{traverse_giveup_b}
+	\end{subfigure}
+	\caption{The \lstinline{traverse} method of the give-up template algorithms}
+	\label{traverse_giveup}
+\end{figure}
+
+Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented.
+In addition to \lstinline{findNext}, the \lstinline{traverse} function uses a helper function called \lstinline{inRange}, which determines whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors.
+If \lstinline{k} is outside the node's range (e.g.
+because the node has been relocated), the search is restarted from the root node \lstinline{r}.
+As in the previous section, we implement this in C with a loop: in each iteration, we acquire the lock on the current node, check that \lstinline{k} is in range, and then use \lstinline{findNext} as above, releasing the lock before we move to the next node.
+If the \lstinline{inRange} call fails, we release our current lock and return to the root node by setting the current node to the root pointer \lstinline{p} (lines 21-22 in Figure \ref{traverse_giveup_b}).
+
+The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure?
+Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations.
+However, in most sequential settings the key ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself.
+The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to as many target data structures as possible without modifying them.
+Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t} in our C implementation:
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
+\end{lstlisting}
+These fields store lower and upper bounds on the keys reachable from the current node.
+%In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys.
+This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes.
+If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
+We can then define \lstinline{inRange} as a helper function in the template, rather than requiring data structures to provide it.
+\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
 

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
continue to restructured sections

@@
 
-\authorrunning{Duc Than Nguyen and William Mansky} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{Duc Than Nguyen et al.} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
@@
 
-\subsection{Lock-coupling BST}
-\begin{figure}[h]
-	\begin{subfigure}[t]{0.48\textwidth}
-	\lstinputlisting[language=C, style=myStyle, escapechar=|]{lock_insert.c} 
-	\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
-	\label{insert_lock}	
-	\end{subfigure}\qquad
-    \begin{subfigure}[t]{0.48\textwidth}
-    \lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
-    \caption{The \lstinline{insert} method of the give-up template algorithm}
-    \label{insert_giveup}
-    \end{subfigure} 
-\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
-\label{insert_lock_giveup} 
-\end{figure}
-
-Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling BST.
-\wm{Actually, is this \lstinline{insert} function part of the template?
-If so, should we discuss it in section 3?} It makes use of the \texttt{traverse} function of Section \ref{lock-coupling-algo} to find the node at which to insert the key \lstinline{k}.
-If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}).
-Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the data-structure-specific \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
-
 We instantiate the template with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}).
@@
 
+\than{Talking about different of \texttt{insertOp} specs here.}
+
+\begin{figure}[h]
+	\centering
+	\begin{mathpar}
+		%insertOp for give-up
+		{\color{blue}
+			\left\{ 
+			\begin{array}{c}
+				\texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} \  \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n}\mapsto\texttt{max}
+			\end{array}
+			\right\}
+		}
+		\\
+		\texttt{insertOp(pn, k, v)}
+		\\
+		{\color{blue}
+			\left\{\exists \ \texttt{t' l r lk1 lk2}. 
+			\begin{array}{c}
+				\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \\ 
+				\ \ast \ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, k))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (k, max))}
+			\end{array}
+			\right\}
+		}
+	\end{mathpar}
+	\caption{Specifications of \texttt{inRange} and \texttt{insertOp} for the give-up template}
+	\label{fig:insertOp_giveup}
+\end{figure}
+
+As previously mentioned, the \texttt{insertOp} specification (Figure \ref{fig:insertOp_giveup}) is slightly different from the lock-coupling version, to account for the added \lstinline{min} and \lstinline{max} fields.
+In a BST, if the current node's range is \lstinline{(min, max)} and its key is \lstinline{k}, its left child will have range \texttt{(min, k)} and its right child will have range \texttt{(k, max)}.
+Otherwise, \lstinline{insertOp} behaves identically, storing \lstinline{k} and \lstinline{v} in the current node and creating two empty child nodes.
+
+\subsection{Lock-coupling BST}
+\begin{figure}[h]
+	\begin{subfigure}[t]{0.48\textwidth}
+	\lstinputlisting[language=C, style=myStyle, escapechar=|]{lock_insert.c} 
+	\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
+	\label{insert_lock}	
+	\end{subfigure}\qquad
+    \begin{subfigure}[t]{0.48\textwidth}
+    \lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
+    \caption{The \lstinline{insert} method of the give-up template algorithm}
+    \label{insert_giveup}
+    \end{subfigure} 
+\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
+\label{insert_lock_giveup} 
+\end{figure}
+
+Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling BST.
+\wm{Actually, is this \lstinline{insert} function part of the template?
+If so, should we discuss it in section 3?} It makes use of the \texttt{traverse} function of Section \ref{lock-coupling-algo} to find the node at which to insert the key \lstinline{k}.
+If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}).
+Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the data-structure-specific \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
+
 The atomic specification for \texttt{insert} method is
@@
 		{\color{blue}
-			\left\{\exists \  \texttt{res.}
+			\left\{\texttt{res.} \
 			\begin{array}{c}
@@
 		}
-		\\
-		\\
-		%insertOp for give-up
-		{\color{blue}
-			\left\{ 
-			\begin{array}{c}
-				\texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} \  \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n}\mapsto\texttt{max}
-			\end{array}
-			\right\}
-		}
-		\\
-		\texttt{insertOp(pn, k, v)}
-		\\
-		{\color{blue}
-			\left\{\exists \ \texttt{t' l r lk1 lk2}. 
-			\begin{array}{c}
-				\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \\ 
-				\ \ast \ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, k))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (k, max))}
-			\end{array}
-			\right\}
-		}
 	\end{mathpar}
 	\caption{Specifications of \texttt{inRange} and \texttt{insertOp} for the give-up template}
-	\label{fig:inRange_insertOp_giveup}
+	\label{fig:inRange_giveup}
 \end{figure}
 
-The \texttt{inRange} function (Figure \ref{fig:inRange_insertOp_giveup}) simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
-
-As previously mentioned, the \texttt{insertOp} specification (Figure \ref{fig:inRange_insertOp_giveup}) is slightly different from the lock-coupling version, to account for the added \lstinline{min} and \lstinline{max} fields.
-In a BST, if the current node's range is \lstinline{(min, max)} and its key is \lstinline{k}, its left child will have range \texttt{(min, k)} and its right child will have range \texttt{(k, max)}.
-Otherwise, \lstinline{insertOp} behaves identically, storing \lstinline{k} and \lstinline{v} in the current node and creating two empty child nodes.
+The \texttt{inRange} function (Figure \ref{fig:inRange_giveup}) simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
 

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
try with annotated code

@@
 
-\authorrunning{Duc Than Nguyen et al.} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{Duc Than Nguyen, Lennart Beringer,  William Mansky, and Shengyi Wang} %TODO mandatory.
+First: Use abbreviated first/middle names.
+Second (only in severe cases): Use first author plus 'et al.'
 
@@
 
+\newcommand\lstsetnumber[1]{\gdef\thelstnumber{#1}}
+\begin{figure}[ht]
+	$\color{blue}\left\langle\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\rangle$
+	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
+		int traverse(pn *pn, int k){
+			  int flag = 1; 			$\left\{\begin{array}{l} 		\ghost{\mathsf{my\_half}((-\infty,+\infty),info)}{g\_root}\ *\ R\ b\ *\ \mathsf{lock\_inv}(l,lsh2,R')\ *\ \\
+			  	\nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} lookup\_inv \end{array}\right\}$
+		  		for( ; ; ){ 			$\left\{\begin{array}{l} lookup\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\\\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$
+		  			pn->p = pn->n;
+		  			if (pn->p->t == NULL)
+		  					break;
+		  			else{
+		  					int b = findNext(pn, k);
+		  					if (b == 0){
+		  							flag = 0;
+		  							break;
+		  					}
+		  					else{
+		  						acquire(pn->n->lock);
+		  						release(pn->p->lock);
+		  					}
+		  				}
+	  				}
+  				return flag;
+  		}  $\lstsetnumber{}$\end{lstlisting}
+$\color{blue}\left\langle\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\rangle$
+\caption{proof of \texttt{traverse}}
+\end{figure}
+
 Given this invariant, the proof proceeds as follows.
@@
 \begin{figure}[h]
+	\begin{subfigure}{\textwidth}
 	\centering
@@
 			\label{fig:insertOp_lock}
-\end{figure}
-
-\wm{Should we recap the definition of the \lstinline{node} struct here?} The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with the current node's key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
-The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the current node's key \texttt{k'}, and the next node \lstinline{n'} is the left child \lstinline{l}; \lstinline{k} is greater than \lstinline{k'} and \lstinline{n'} is the right child \lstinline{r}; or \lstinline{k} is exactly \lstinline{k'} and \lstinline{n'} is the original node \lstinline{n}, in which case the return value is 0.
-
-The \texttt{insertOp} specification (Figure \ref{fig:insertOp_lock}) takes as input a node whose \lstinline{t} field is \lstinline{NULL}, i.e., an empty leaf node.
-It inserts a new node with the specified key \texttt{k} and value \texttt{v} at that leaf, complete with two new empty child nodes with corresponding locks, \texttt{(NULL, lk1)} and \texttt{(NULL, lk2)}, which are pointed to by the \texttt{l} and \texttt{r} of the current node.
-
-\than{Talking about different of \texttt{insertOp} specs here.}
-
-\begin{figure}[h]
-	\centering
+	\end{subfigure}
+	%insertOp for give-up of give-up
+	\begin{subfigure}{\textwidth}
 		\begin{mathpar}
-		%insertOp for give-up
 	{\color{blue}
@@
 	\end{mathpar}
-	\caption{Specifications of \texttt{inRange} and \texttt{insertOp} for the give-up template}
+	\caption{Specification of \texttt{insertOp} for the give-up template}
 	\label{fig:insertOp_giveup}
+\end{subfigure}
+	\caption{Specification of \texttt{insertOp} for the lock-coupling template}
+	\label{fig:insertOp}
 \end{figure}
+
+\wm{Should we recap the definition of the \lstinline{node} struct here?} The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with the current node's key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
+The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the current node's key \texttt{k'}, and the next node \lstinline{n'} is the left child \lstinline{l}; \lstinline{k} is greater than \lstinline{k'} and \lstinline{n'} is the right child \lstinline{r}; or \lstinline{k} is exactly \lstinline{k'} and \lstinline{n'} is the original node \lstinline{n}, in which case the return value is 0.
+
+The \texttt{insertOp} specification (Figure \ref{fig:insertOp_lock}) takes as input a node whose \lstinline{t} field is \lstinline{NULL}, i.e., an empty leaf node.
+It inserts a new node with the specified key \texttt{k} and value \texttt{v} at that leaf, complete with two new empty child nodes with corresponding locks, \texttt{(NULL, lk1)} and \texttt{(NULL, lk2)}, which are pointed to by the \texttt{l} and \texttt{r} of the current node.
+
+\than{Talking about different of \texttt{insertOp} specs here.}
 
@@
 			\begin{array}{c}
-				\ \texttt{pn->n}\mapsto\texttt{(t, lock, min, max)} \  \\ 
-				\ \ast  \ \texttt{(if res then } \texttt{(min < k < max)} \texttt{ else } \texttt{(k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max))}
+				\ \texttt{pn->n}\mapsto\texttt{(t, lock, min, max)} \  \ast  \ 
+				\\ \texttt{(if res then } \texttt{(min < k < max)} \texttt{ else } \texttt{(k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max))}
 			\end{array}

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
Use mathsf for all predicates.
The notations after the beginning of section 4 is not modified yet.

@@
 		{\color{blue}
-			\forall \  \texttt{m}. \left\langle
-			\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{m} \
+			\forall \  m. \left\langle
+			\nodeboxrep(\texttt{pn->n})  \ \ast \ \mathsf{R}(\texttt{pn->n}) \ \big| \ \treerep\ m \
 			\right\rangle
@@
 		{\color{blue}
-			\left\langle \texttt{res}. \ \exists \  \texttt{n'}.
+			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
 			\begin{array}{c}
-				\ \texttt{(res} \leftrightarrow \texttt{(pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
-				\ast \ \texttt{R(pn->n')} \ast \ \texttt{(k} \in \texttt{range(n'))}
+				\ (\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ \nodeboxrep(\texttt{pn->n'}) \\  
+				\ast \ \mathsf{R}(\texttt{pn->n'})\ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))
 			\end{array}
-			\ \Bigg| \ \treerep\ \texttt{m} \
+			\ \Bigg| \ \treerep\ m \
 			\right\rangle
@@
 Formally, we write
-\[\nodeboxrep \texttt{(pn->n)} \triangleq\ \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}\]
-where the \emph{lock invariant assertion} $\ell \lockvar \texttt{R}$ expresses that the lock $\ell$ protects resources \texttt{R}.
-In this case, $R$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{pn->n}).
+\[\nodeboxrep(\texttt{pn->n}) \triangleq\ \texttt{pn->n->lock} \lockvar \mathsf{R}(\texttt{pn->n})\]
+where the \emph{lock invariant assertion} $\ell \lockvar \mathsf{R}$ expresses that the lock $\ell$ protects resources $\mathsf{R}$.
+In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{pn->n}).
 
 Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
-Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{pn->n}$ and obtain the resources $\texttt{R(pn->n)}$ guarded by the lock.
-These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $R$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node.
+Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{pn->n}$ and obtain the resources $\mathsf{R}(\texttt{pn->n})$ guarded by the lock.
+These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node.
 \wm{Should we replace this with a concrete definition of the invariant?
@@
 
-In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{pn->n})$ and the contents $\texttt{R(pn->n)}$.
+In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{pn->n})$ and the contents $\mathsf{R}(\texttt{pn->n})$.
 The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}.
-The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{pn->n'})$ \wm{This looks weird: \lstinline{pn} doesn't have an \lstinline{n'} field}, its contents $\texttt{R(pn->n')}$, and a Boolean variable \texttt{res} indicating whether \texttt{pn->n'->t} is \texttt{NULL} or not \wm{Indicate the relationship between \lstinline{res} and the \lstinline{flag} return value}.
+The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{pn->n'})$ \wm{This looks weird: \lstinline{pn} doesn't have an \lstinline{n'} field}, its contents $\mathsf{R}(\texttt{pn->n'})$, and a Boolean variable $\mathit{res}$ indicating whether \texttt{pn->n'->t} is \texttt{NULL} or not \wm{Indicate the relationship between \lstinline{res} and the \lstinline{flag} return value}.
 The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ \texttt{(k} \in \texttt{range(n'))}\ \land \ \nodeboxrep\texttt{(pn->n')}  \ \ast \  \texttt{R(pn->n')}   
+\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{n'}.\ \texttt{k} \in \mathsf{range}(\texttt{n'})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \mathsf{R}(\texttt{pn->n'})
 \end{mathpar}\wm{Where does \lstinline{range} come from here?}
@@
 Given this invariant, the proof proceeds as follows.
-\wm{Should we consider giving annotated code after all?} We begin by checking whether the current node is null (line 5 of Figure~\ref{traverse_lock_b}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with \lstinline{res = true}.
+\wm{Should we consider giving annotated code after all?} We begin by checking whether the current node is null (line 5 of Figure~\ref{traverse_lock_b}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$.
 Otherwise, we use \lstinline{findNext} to find a new node \lstinline{n'} to visit (line 8), which is stored in \lstinline{pn->n}.
-If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with \lstinline{res = false}.
-Otherwise, we acquire \lstinline{pn->n->lock}, gaining access to the resources $\texttt{R(pn->n)}$, and then release the lock of the current node \texttt{pn->p} and return its resources.
+If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$.
+Otherwise, we acquire \lstinline{pn->n->lock}, gaining access to the resources $\mathsf{R}(\texttt{pn->n})$, and then release the lock of the current node \texttt{pn->p} and return its resources.
 %It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
-By acquiring the lock for the next node \texttt{n} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish \lstinline{traverse_inv} for the new values in \lstinline{pn}.
+By acquiring the lock for the next node \texttt{n} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish $\mathsf{traverse\_inv}$ for the new values in \lstinline{pn}.
 %The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
@@
 		{\color{blue}
-			\forall \  \texttt{m}. \left\langle
+			\forall \  m. \left\langle
 			\begin{array}{c}
-				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{m}
+				\infp (\texttt{pn->n}) \ \big| \ \treerep\ m
 			\end{array}
@@
 		{\color{blue}
-			\left\langle \texttt{res}. \ \exists \ \texttt{n'}.
+			\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
 			\begin{array}{c}
-				\texttt{(res} \leftrightarrow \texttt{(pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
+				(\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ \infp(\texttt{pn->n'}) 
 				\\ 
-				\ast \ \texttt{R(pn->n')} \ \ast \ (\texttt{k} \in \texttt{range(n'))}
+				\ast \ \mathsf{R}(\texttt{pn->n'}) \ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))
 			\end{array}
-			\ \Bigg| \ \treerep\ \texttt{m} \
+			\ \Bigg| \ \treerep\ m \
 			\right\rangle
@@
 The local precondition include any locks, and consists only of $\infp(\texttt{pn->n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
-As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents \texttt{R(pn->n')}.
-Also as before, the Boolean \texttt{res} indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
+As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents $\mathsf{R}(\texttt{pn->n'})$.
+Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
 
@@
 For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held.
-Formally, we write \[\texttt{inv\_for\_lock} \ \ell \ \texttt{R} \triangleq \exists b.
-\ \ell \mapsto b \ \ast \ \texttt{if } b \ \texttt{then emp else R}\]\wm{Consider using mathsf for logic and reserving texttt for C code.}
+Formally, we write \[\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists b.
+\ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}\]\wm{Consider using mathsf for logic and reserving texttt for C code.}
 %where a lock at location $\ell$ safeguards resources denoted by \texttt{R}.
@@
 On the other hand, when the lock is not held ($\ell \mapsto \texttt{false}$), resources \texttt{R} are required to be part of the invariant.
-and $\treerep$ contains an \texttt{inv\_for\_lock} assertion for each node in the data structure.
-As before, $R$ contains the concrete contents of each node, but also includes the \lstinline{min} and \lstinline{max} fields that we use to implement the \lstinline{inRange} check.
+and $\treerep$ contains an $\mathsf{inv\_for\_lock}$ assertion for each node in the data structure.
+As before, $\mathsf{R}$ contains the concrete contents of each node, but also includes the \lstinline{min} and \lstinline{max} fields that we use to implement the \lstinline{inRange} check.
 
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{mathpar} \texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
+\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
 \end{mathpar}
@@
 We start with the $\infp (\texttt{pn->n})$ predicate from the local precondition.
-After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\texttt{R(pn->n)}$, which allows us to perform the \lstinline{inRange} check on line 7.
+After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\mathsf{R}(\texttt{pn->n})$, which allows us to perform the \lstinline{inRange} check on line 7.
 If the check fails (lines 20-23), we immediately release the lock and start over from the beginning.
-Otherwise, we proceed as before: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \lstinline{k} belongs (and can satisfy the postcondition); if \lstinline{findNext} returns 0, we have found the node containing \lstinline{k} (and can satisfy the postcondition); otherwise, we release our lock and re-establish \lstinline{traverse_inv} for the new node indicated by \lstinline{findNext}.
+Otherwise, we proceed as before: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \lstinline{k} belongs (and can satisfy the postcondition); if \lstinline{findNext} returns 0, we have found the node containing \lstinline{k} (and can satisfy the postcondition); otherwise, we release our lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \lstinline{findNext}.
 In this case, we do not hold the lock on the new node, but only the $\infp$ assertion indicating that the new node is in the data structure.
@@
 	{\color{blue}
-		\texttt{m.}\left\langle 
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m}
+		m.\left\langle 
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ m
 		\right\rangle
@@
 		\left\langle 
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
 		\right\rangle
@@
 		\left\langle \texttt{v}. \ \
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m} \land \texttt{m(k) = v}
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
 		\right\rangle
@@
 		\left\langle 
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{(m[k} \mapsto \_ \texttt{])}
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
 		\right\rangle

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
sketch proof of lock-coupling traverse

@@
 		{\color{blue}
-			\forall \  m. \left\langle
-			\nodeboxrep(\texttt{pn->n})  \ \ast \ \mathsf{R}(\texttt{pn->n}) \ \big| \ \treerep\ m \
+			\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
+			\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
 			\right\rangle
@@
 			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-			\begin{array}{c}
+			\begin{array}{c} \texttt{pn} \mapsto (\texttt{p, n}) \ \ast 
 				\ (\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ \nodeboxrep(\texttt{pn->n'}) \\  
@@
 The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{n'}.\ \texttt{k} \in \mathsf{range}(\texttt{n'})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \mathsf{R}(\texttt{pn->n'})
+\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{n'}.\ \texttt{k} \in \mathsf{range}(\texttt{n'})\ \land \ \texttt{pn} \mapsto (\texttt{p, n}) \ast \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'})
 \end{mathpar}\wm{Where does \lstinline{range} come from here?}
 
-\newcommand\lstsetnumber[1]{\gdef\thelstnumber{#1}}
-\begin{figure}[ht]
-	$\color{blue}\left\langle\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\rangle$
+\begin{figure}[h]
+	$\color{blue}
+	\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
+	\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
+	\right\rangle$
 	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
 		int traverse(pn *pn, int k){
-			  int flag = 1; 			$\left\{\begin{array}{l} 		\ghost{\mathsf{my\_half}((-\infty,+\infty),info)}{g\_root}\ *\ R\ b\ *\ \mathsf{lock\_inv}(l,lsh2,R')\ *\ \\
-			  	\nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} lookup\_inv \end{array}\right\}$
-		  		for( ; ; ){ 			$\left\{\begin{array}{l} lookup\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ *\\\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$
+			int flag = 1; 			$\left\{\begin{array}{l}  ...		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+			for( ; ; ){ 			$\left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} ... \end{array}\right\}$
 				pn->p = pn->n;
@@
 			return flag;
-  		}  $\lstsetnumber{}$\end{lstlisting}
-$\color{blue}\left\langle\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\rangle$
-\caption{proof of \texttt{traverse}}
+	}  \end{lstlisting}
+	$\color{blue}
+	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
+	\begin{array}{c} \texttt{pn} \mapsto (\texttt{p, n'}) \ \ast 
+		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \nodeboxrep(\texttt{n'}) \ \ast \
+		\\ \mathsf{R}(\texttt{n'})\ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))
+	\end{array}
+	\ \Bigg| \ \treerep\ m \
+	\right\rangle$
+	\caption{The \texttt{traverse} function in lock-coupling template annotated with separation logic specification}
 \end{figure}

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
Finished the mathsf unification.

@@
 		{\color{blue}
-			\left\{\texttt{res}. \ \exists \  \texttt{n'.}
+			\left\{\mathit{res}. \ \exists \  \texttt{n'.}
 			\begin{array}{c}
 				\texttt{pn->n->t}\mapsto \texttt{(k', v', l, r)}  \\\  
-				\ast \ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n = n'} \land \texttt{k = k'))}
+				\ast \ (\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{l} = \texttt{n'} \land \texttt{k} < \texttt{k'}) \lor (\texttt{r} = \texttt{n'} \land \texttt{k} > \texttt{k'})\ \mathsf{else}\ (\texttt{n} = \texttt{n'} \land \texttt{k} = \texttt{k'}))
 			\end{array}
@@
 		\left\langle 
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (\texttt{m[k} \mapsto \texttt{v]})
+		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
 		\right\rangle
@@
 The local pre- and postconditions, containing $\nodeboxrep \ \texttt{p}$, describe the binary search tree's concrete state with the root node $\texttt{p}$, which points to the node's lock and lock invariant assertion.
-On the other hand, the public pre- and postconditions are represented by the predicate $\treerep \ \texttt{m}$, illustrating the binary search tree's abstract state as a collection of ghost states, with each state providing information about an individual node in the tree.
-Here, $\texttt{m}$ and $\texttt{m[k} \mapsto \texttt{v]}$ denote the tree's abstract states before and after the function's execution, respectively.
+On the other hand, the public pre- and postconditions are represented by the predicate $\treerep \ m$, illustrating the binary search tree's abstract state as a collection of ghost states, with each state providing information about an individual node in the tree.
+Here, $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively.
 
@@
 We set \texttt{pn->n} to point to the root node \texttt{*p} (line 3 in Figure \ref{insert_lock}).
-After the lock for the root node is acquired (line 4 in Figure \ref{insert_lock}), the thread acquires the contents of the root node, \texttt{R(pn->n)}.
+After the lock for the root node is acquired (line 4 in Figure \ref{insert_lock}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{pn->n})$.
 We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
 
-In the \texttt{then} branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
+In the $\mathsf{then}$ branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
 we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree.
 In this scenario, we update the node's value while maintaining the rest of the tree structure (line 6 in Figure \ref{insert_lock}).
-Otherwise, when \texttt{traverse} reaches a leaf and produces $\texttt{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 9 in Figure \ref{insert_lock}).
+Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 9 in Figure \ref{insert_lock}).
 We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree.
-In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $\texttt{m}$ to $\texttt{m[k} \mapsto \texttt{v]}$, thereby satisfying the public postcondition of \lstinline{insert}.
+In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}.
 We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
@@
 		{\color{blue}
-			\left\{\texttt{res.} \
+			\left\{\mathit{res.} \
 			\begin{array}{c}
 				\ \texttt{pn->n}\mapsto\texttt{(t, lock, min, max)} \  \ast  \ 
-				\\ \texttt{(if res then } \texttt{(min < k < max)} \texttt{ else } \texttt{(k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max))}
+				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{min} < \texttt{k} < \texttt{max})\ \mathsf{else}\ (\texttt{k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max}))
 			\end{array}

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
finished proof of lock-coupling traverse

@@
 The definitions of $\nodeboxrep$ and $\treerep$ may vary depending on the data structure, but fundamentally, $\nodeboxrep$ holds the resources owned by each client thread, while $\treerep$ holds the abstract representation of the data structure, treated as an atomically accessed shared resource.
-In particular, holding $\nodeboxrep(\texttt{pn->n})$ should allow us to acquire the lock on node \texttt{pn->n}, which in turn gives us access to the contents of the node.
+In particular, holding $\nodeboxrep(\texttt{n})$ should allow us to acquire the lock on node \texttt{n}, which in turn gives us access to the contents of the node.
 Formally, we write
-\[\nodeboxrep(\texttt{pn->n}) \triangleq\ \texttt{pn->n->lock} \lockvar \mathsf{R}(\texttt{pn->n})\]
+\[\nodeboxrep(\texttt{n}) \triangleq\ \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n})\]
 where the \emph{lock invariant assertion} $\ell \lockvar \mathsf{R}$ expresses that the lock $\ell$ protects resources $\mathsf{R}$.
@@
 Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
-Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{pn->n}$ and obtain the resources $\mathsf{R}(\texttt{pn->n})$ guarded by the lock.
+Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{n}$ and obtain the resources $\mathsf{R}(\texttt{n})$ guarded by the lock.
 These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node.
-\wm{Should we replace this with a concrete definition of the invariant?
-e.g.
-$R = ghost\_node(k, v, range) * ...$}
+Formally, the resource  $\mathsf{R}(\texttt{n}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{DS\_specific(\texttt{n})}$.
 
@@
 
-In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{pn->n})$ and the contents $\mathsf{R}(\texttt{pn->n})$.
+In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{n})$ and the contents $\mathsf{R}(\texttt{n})$.
 The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}.
-The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{pn->n'})$ \wm{This looks weird: \lstinline{pn} doesn't have an \lstinline{n'} field}, its contents $\mathsf{R}(\texttt{pn->n'})$, and a Boolean variable $\mathit{res}$ indicating whether \texttt{pn->n'->t} is \texttt{NULL} or not \wm{Indicate the relationship between \lstinline{res} and the \lstinline{flag} return value}.
+The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not.
+The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its boolean state.
 The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{n'}.\ \texttt{k} \in \mathsf{range}(\texttt{n'})\ \land \ \texttt{pn} \mapsto (\texttt{p, n}) \ast \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'})
-\end{mathpar}\wm{Where does \lstinline{range} come from here?}
+\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'})\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'}) \ \ast \ \treerep\ m
+\end{mathpar} %\wm{Where does \lstinline{range} come from here?}
 
@@
 		int traverse(pn *pn, int k){
-			int flag = 1; 			$\left\{\begin{array}{l}  ...		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-			for( ; ; ){ 			$\left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} ... \end{array}\right\}$
+			int flag = 1; 			$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
+				\nodeboxrep(\texttt{n})  \ast \mathsf{R}(\texttt{n})  \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+			for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} ... \end{array}\right\}$
 				pn->p = pn->n;
 				if (pn->p->t == NULL)
+				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
 				break;
@@
 					int b = findNext(pn, k);
+					$\color{blue} \left\{\begin{array}{l} \exists \  \texttt{n'}. \ \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \\ 
+						(\mathit{res'} = \mathsf{false} \ast \texttt{n'} = \texttt{n} \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots )\  \lor \\ 
+						(\mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \cdots)   \end{array}\right\}$
 					if (b == 0){
 						flag = 0;
+						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \texttt{n'} = \texttt{n} \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots    \end{array}\right\}$
 						break;
@@
 					else{
+						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \cdots    \end{array}\right\}$
 						acquire(pn->n->lock);
+						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \\ \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \mathsf{R}(\texttt{n'}) \ast \cdots    \end{array}\right\}$
 						release(pn->p->lock);
+						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \\ \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \mathsf{R}(\texttt{n'}) \ast \cdots    \end{array}\right\}$
 					}
@@
 			return flag;
+			$\color{blue} \left\{\begin{array}{l} (\mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL} \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots) \  \lor \\
+				(\mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast (\texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)
+			\end{array}\right\}$
 	}  \end{lstlisting}
@@
 	\caption{The \texttt{traverse} function in lock-coupling template annotated with separation logic specification}
+	\label{proof_lock_traverse}
 \end{figure}
 
-Given this invariant, the proof proceeds as follows.
-\wm{Should we consider giving annotated code after all?} We begin by checking whether the current node is null (line 5 of Figure~\ref{traverse_lock_b}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$.
-Otherwise, we use \lstinline{findNext} to find a new node \lstinline{n'} to visit (line 8), which is stored in \lstinline{pn->n}.
+Given this invariant, the proof proceeds as follows. %\wm{Should we consider giving annotated code after all?} 
+We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$.
+Otherwise, we use \lstinline{findNext} to find a new node \lstinline{n'} to visit (line 8), which is stored in \lstinline{n}.
 If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$.
-Otherwise, we acquire \lstinline{pn->n->lock}, gaining access to the resources $\mathsf{R}(\texttt{pn->n})$, and then release the lock of the current node \texttt{pn->p} and return its resources.
+Otherwise, we acquire \lstinline{n->lock}, gaining access to the resources $\mathsf{R}(\texttt{n})$, and then release the lock of the current node \texttt{p} and return its resources.
 %It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
@@
 
-\subsubsection{Give-up Template}
+\subsubsection{Give-up Proof}
 \label{traverse_proof_giveup}
-
-The atomic specification for the give-up template's \lstinline{traverse} is shown in Figure \ref{fig:traverse_giveup}.
 
@@
 			\begin{array}{c}
-				\infp (\texttt{pn->n}) \ \big| \ \treerep\ m
+				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
 			\end{array}
@@
 			\begin{array}{c}
-				(\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ \infp(\texttt{pn->n'}) 
-				\\ 
-				\ast \ \mathsf{R}(\texttt{pn->n'}) \ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))
+				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ 
+				\\ \infp(\texttt{pn->n'}) \ \ast \ \mathsf{R}(\texttt{pn->n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
 			\end{array}
@@
 
-The give-up template's \texttt{traverse} specification is simpler than that of the lock-coupling template.
+The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template.
 The local precondition include any locks, and consists only of $\infp(\texttt{pn->n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
@@
 Formally, we write \[\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists b.
-\ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}\]\wm{Consider using mathsf for logic and reserving texttt for C code.}
+\ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}\]
+%\wm{Consider using mathsf for logic and reserving texttt for C code.}
 %where a lock at location $\ell$ safeguards resources denoted by \texttt{R}.

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
finished proof of give-up traverse abd added in appendix

@@
 			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-			\begin{array}{c} \texttt{pn} \mapsto (\texttt{p, n}) \ \ast 
-				\ (\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ \nodeboxrep(\texttt{pn->n'}) \\  
-				\ast \ \mathsf{R}(\texttt{pn->n'})\ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))
+			\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
+				\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
 			\end{array}
@@
 The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'})\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'}) \ \ast \ \treerep\ m
+\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'})\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'}) \ \ast \ \treerep\ m
 \end{mathpar} %\wm{Where does \lstinline{range} come from here?}
@@
 				if (pn->p->t == NULL)
-				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
 						break;
+						$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
 				else{
@@
 						flag = 0;
-						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \texttt{n'} = \texttt{n} \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots    \end{array}\right\}$
 						break;
+						$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \texttt{n'} = \texttt{n} \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots    \end{array}\right\}$
 				}
@@
 		$\color{blue} \left\{\begin{array}{l} (\mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL} \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots) \  \lor \\
-				(\mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast (\texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)
+		(\mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)
 		\end{array}\right\}$
@@
 	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-	\begin{array}{c} \texttt{pn} \mapsto (\texttt{p, n'}) \ \ast 
-		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \nodeboxrep(\texttt{n'}) \ \ast \
-		\\ \mathsf{R}(\texttt{n'})\ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))
+	\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
+		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
 	\end{array}
@@
 			\begin{array}{c}
-				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ 
-				\\ \infp(\texttt{pn->n'}) \ \ast \ \mathsf{R}(\texttt{pn->n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
+				\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
+				\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
 			\end{array}
@@
 The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template.
-The local precondition include any locks, and consists only of $\infp(\texttt{pn->n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
-As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents $\mathsf{R}(\texttt{pn->n'})$.
+The local precondition include any locks, and consists only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
+As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$.
 Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
@@
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
+\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{p q n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \ \infp (\texttt{q}) \ast \infp (\texttt{n'})  \ \ast \ \treerep\ m
 \end{mathpar}
 With this invariant in hand, the proof of \lstinline{traverse} proceeds as follows.
-We start with the $\infp (\texttt{pn->n})$ predicate from the local precondition.
-After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\mathsf{R}(\texttt{pn->n})$, which allows us to perform the \lstinline{inRange} check on line 7.
+We start with the $\infp (\texttt{n})$ predicate from the local precondition.
+After acquiring the lock of the node $\texttt{n}$ (line 5 in Figure \ref{proof_giveup_traverse} in Appendix \ref{sec:apd_giveup_proof}), we obtain the resource $\mathsf{R}(\texttt{n})$, which allows us to perform the \lstinline{inRange} check on line 7.
 If the check fails (lines 20-23), we immediately release the lock and start over from the beginning.
@@
 
-\newpage
 \bibliography{../sources}
 
-%\appendix
+\appendix
 
-%\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}
+\section{Give-up Proof}
+\label{sec:apd_giveup_proof}
+
+\begin{figure}[h]
+	$\color{blue}
+	\forall \  m. \left\langle
+	\begin{array}{c}
+		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
+	\end{array}
+	\right\rangle$
+	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
+		int traverse(pn *pn, int k){
+			node_t *r = (pn->n);
+			int flag = 1; 			$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
+				\infp (\texttt{n})  \ast \infp (\texttt{r}) \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+			for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} ... \end{array}\right\}$
+				acquire(pn->n->lock);
+				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
+				pn->p = pn->n; $\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
+				if (inRange(pn, k) == 1){
+					$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast  \cdots \end{array}\right\}$
+					if (pn->p->t == NULL)
+					break;
+					$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
+					else{
+						int b = findNext(pn, k);
+						$\color{blue} \left\{\begin{array}{l} \exists \  \texttt{n'}. \ \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \\ 
+							(\mathit{res'} = \mathsf{false} \ast \texttt{n'} = \texttt{n} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots )\  \lor \\ 
+							(\mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \infp (\texttt{n'}) \ast \cdots)   \end{array}\right\}$
+						if (b == 0){
+							flag = 0;
+							break;
+							$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL}  \ast \texttt{n'} = \texttt{n} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots   \end{array}\right\}$
+						}
+						else
+						$\color{blue} \left\{\begin{array}{l} \mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n->t} \neq \texttt{NULL} \ast \infp (\texttt{n'}) \ast \cdots    \end{array}\right\}$
+						release(pn->p->lock);
+						$\color{blue} \left\{\begin{array}{l} \mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n->t} \neq \texttt{NULL} \ast \infp (\texttt{n'}) \ast \cdots    \end{array}\right\}$
+					}
+				}
+				else{
+					$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast  \cdots \end{array}\right\}$
+					release(pn->p->lock);
+					$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast  \cdots \end{array}\right\}$
+					pn->n = r;
+					$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast  \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{r, r}) \ast \infp (\texttt{r}) \ast \cdots    \end{array}\right\}$
+				}
+			}
+			return flag;
+			$\color{blue} \left\{\begin{array}{l} (\mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots) \  \lor \\
+				(\mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)
+			\end{array}\right\}$
+	}  \end{lstlisting}
+	$\color{blue}
+	\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
+	\begin{array}{c}
+		\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
+		\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
+	\end{array}
+	\ \Bigg| \ \treerep\ m \
+	\right\rangle$
+	\caption{The \texttt{traverse} function in give-up template annotated with separation logic specification}
+	\label{proof_giveup_traverse}
+\end{figure}
 

### 2023-04-29 concurrency/CONCUR 2023/giveup_traverse.c

@@ -1,5 +1,5 @@
 int traverse(pn *pn, int k){
-  tree_t *p = (pn->n); 
+  node_t *r = (pn->n); 
   int flag = 1;
   for( ; ; ){
     acquire(pn->n->lock); 
@@ -19,7 +19,7 @@ int traverse(pn *pn, int k){
     }
     else{
       release(pn->p->lock);
-      pn->n = p;
+      pn->n = r;
     }
   }
   return flag;

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
finished proof mechanization part

@@
 These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node.
-Formally, the resource  $\mathsf{R}(\texttt{n}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{DS\_specific(\texttt{n})}$.
+Formally, the resource  $\mathsf{R}(\texttt{n}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{DataStruct\_specific(\texttt{n})}$.
 
@@
 \label{proof_mech}
-All the proofs described above have been mechanized in VST, using its extensions for logical atomicity~\cite{iris-vst-arxiv}. Statistics on the verification effort are shown in Table~\ref{table_coq}
+All the proofs described above have been mechanized in VST, using its extensions for logical atomicity~\cite{iris-vst-arxiv}. Statistics on the verification effort are shown in Table~\ref{table_coq}.
 
@@
 This encompasses supplementary proofs, template specifications (encompassing all helper functions such as \texttt{findNext}, \texttt{insertOp}, and others, in addition to \texttt{traverse}, \texttt{insert}, \texttt{lookup}, and \texttt{delete}), as well as proofs against with lock-coupling and give-up template specifications.
-\wm{Rather than re-stating what we proved, this is the place to discuss the development: e.g., which parts took the most effort and LoC, which parts we expect to be reusable and in what way, how much bigger or smaller the template proofs were than non-template versions, etc.}
-
-The proofs can be found online at
-
-\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}
+%\wm{Rather than re-stating what we proved, this is the place to discuss the development: e.g., which parts took the most effort and LoC, which parts we expect to be reusable and in what way, how much bigger or smaller the template proofs were than non-template versions, etc.}
+The construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, demanded substantial effort and lines of code (LoC).
+However, we expect that functions such as \texttt{traverse}, \texttt{inRange}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.
+The methods \texttt{findNext} and \texttt{insertOp} may also be adapted to other data structures, such as linked lists, with some additional effort.
+Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext} and \texttt{insertOp} to restructure the \texttt{node} struct.
+While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future.
+The full development of our mechanization effort is available online
+\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.
 

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
fixed proof mechanization part

@@
 %\wm{Rather than re-stating what we proved, this is the place to discuss the development: e.g., which parts took the most effort and LoC, which parts we expect to be reusable and in what way, how much bigger or smaller the template proofs were than non-template versions, etc.}
-The construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, demanded substantial effort and lines of code (LoC).
+
+Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process.
+Otherwise, the construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, necessitated a considerable amount of time and effort.
 However, we expect that functions such as \texttt{traverse}, \texttt{inRange}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
fixed breakage

@@
 
-\author{Duc Than Nguyen}{University of Illinois at Chicago, USA \and \url{http://dnguye96.people.uic.edu} }{dnguye96@uic.edu}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{Duc Than Nguyen}{University of Illinois at Chicago, USA \and \url{http://dnguye96.people.uic.edu} }{dnguye96@uic.edu}{https://orcid.org/0000-0002-6810-897X}{%(Optional) author-specific funding acknowledgements
+}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.
@@
 
-\author{Lennart Beringer}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~eberinge} }{eberinge@cs.princeton.edu}{https://orcid.org/0000-0002-1570-3492}{(Optional) author-specific funding acknowledgements}
+\author{Lennart Beringer}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~eberinge} }{eberinge@cs.princeton.edu}{https://orcid.org/0000-0002-1570-3492}{}
 
-\author{William Mansky}{University of Illinois at Chicago, USA\and \url{https://mansky.lab.uic.edu/} }{mansky1@uic.edu}{https://orcid.org/0000-0002-5351-895X}{(Optional) author-specific funding acknowledgements}
+\author{William Mansky}{University of Illinois at Chicago, USA\and \url{https://mansky.lab.uic.edu/} }{mansky1@uic.edu}{https://orcid.org/0000-0002-5351-895X}{}
 
-\author{Shengyi Wang}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~shengyiw} }{shengyiw@cs.princeton.edu}{https://orcid.org/[0000-0002-2286-8703}{(Optional) author-specific funding acknowledgements}
+\author{Shengyi Wang}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~shengyiw} }{shengyiw@cs.princeton.edu}{https://orcid.org/[0000-0002-2286-8703}{}
 
@@
 
-\Copyright{Duc Than Nguyen and William Mansky} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{Duc Than Nguyen, Lennart Beringer, William Mansky, and Shengyi Wang} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 
-\keywords{concurrent separation logic, fine-grained locking, logical atomicity,
+\keywords{concurrent separation logic, fine-grained locking, logical atomicity, theorem proving,
 	Verified Software Toolchain, Iris} %TODO mandatory; please add comma-separated list of keywords

### 2023-04-30 concurrency/CONCUR 2023/paper.tex
reformatted give-up proof
finished annotated of insert functions for both template and add them in appendix

@@
 		m.\left\langle 
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ m
-		\right\rangle
-	}
-	\ \texttt{insert(p, k, v)}\ 
+		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
+		\right\rangle
+	}
+	\ \texttt{insert(r, k, v)}\ 
 	{\color{blue}
 		\left\langle 
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
+		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
 		\right\rangle
@@
 		m.\left\langle 
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ m
-		\right\rangle
-	}
-	\ \texttt{lookup(p, k)}\ 
+		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
+		\right\rangle
+	}
+	\ \texttt{lookup(r, k)}\ 
 	{\color{blue}
 		\left\langle \texttt{v}. \ \
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
+		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
 		\right\rangle
@@
 		m.\left\langle 
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ m
-		\right\rangle
-	}
-	\ \texttt{delete(p, k)}\ 
+		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
+		\right\rangle
+	}
+	\ \texttt{delete(r, k)}\ 
 	{\color{blue}
 		\left\langle 
-		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
+		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
 		\right\rangle
@@
 			\begin{array}{c}
-				\ \texttt{pn->n->t}  
+				\ \texttt{pn->n} \mapsto \texttt{NULL}  
 			\end{array}
@@
 		{\color{blue}
-			\left\{\exists \  \texttt{t' l r lk1 lk2}.
-			\begin{array}{c}
-				\ \texttt{pn->n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
+			\left\{\exists \  \texttt{t l r lk1 lk2}.
+			\begin{array}{c}
+				\ \texttt{pn->n->t} \mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
 				\ \texttt{l} \mapsto \texttt{(NULL, lk1)}\  \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2)}
@@
 		\begin{array}{c}
-			\texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} \  \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n}\mapsto\texttt{max}
+			\texttt{pn->n}\mapsto \texttt{NULL} \  \ast \ \texttt{pn->n}\mapsto \texttt{min} \ \ast \ \texttt{pn->n}\mapsto\texttt{max}
 		\end{array}
@@
 	{\color{blue}
-		\left\{\exists \ \texttt{t' l r lk1 lk2}. 
-		\begin{array}{c}
-			\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \\ 
-			\ \ast \ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, k))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (k, max))}
+		\left\{\exists \ \texttt{t l r lk1 lk2}. 
+		\begin{array}{c}
+			\texttt{pn->n->t}\mapsto \texttt{(k, v, l, r)} \  \ast \ \texttt{pn->n}\mapsto \texttt{min} \ \ast \ \texttt{pn->n}\mapsto\texttt{max}  \ \ast \ 
+			\\ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, k))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (k, max))}
 		\end{array}
@@
 \end{subfigure}
-	\caption{Specification of \texttt{insertOp} for the lock-coupling template}
+	\caption{Specification of \texttt{insertOp} for the lock-coupling and give-up templates}
 	\label{fig:insertOp}
@@
 \wm{Consider skipping this paragraph if we need space.} Similar to the \texttt{traverse} specification, the \texttt{insert} function's specification is composed of two components: local and public pre- and postconditions.
-The local pre- and postconditions, containing $\nodeboxrep \ \texttt{p}$, describe the binary search tree's concrete state with the root node $\texttt{p}$, which points to the node's lock and lock invariant assertion.
+The local pre- and postconditions, containing $\nodeboxrep \ \texttt{r}$, describe the binary search tree's concrete state with the root node $\texttt{r}$, which points to the node's lock and lock invariant assertion.
 On the other hand, the public pre- and postconditions are represented by the predicate $\treerep \ m$, illustrating the binary search tree's abstract state as a collection of ghost states, with each state providing information about an individual node in the tree.
@@
 The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{p}$.
-We set \texttt{pn->n} to point to the root node \texttt{*p} (line 3 in Figure \ref{insert_lock}).
-After the lock for the root node is acquired (line 4 in Figure \ref{insert_lock}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{pn->n})$.
+We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{insert_lock}).
+After the lock for the root node is acquired (line 4 in Figure \ref{insert_lock}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$.
 We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
@@
 			\begin{array}{c}
-				\texttt{pn->n}\mapsto\texttt{(t, lock, min, max)} 
+				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} 
 			\end{array}
@@
 			\begin{array}{c}
-				\ \texttt{pn->n}\mapsto\texttt{(t, lock, min, max)} \  \ast  \ 
+				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} \  \ast  \ 
 				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{min} < \texttt{k} < \texttt{max})\ \mathsf{else}\ (\texttt{k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max}))
@@
 	\end{mathpar}
-	\caption{Specifications of \texttt{inRange} and \texttt{insertOp} for the give-up template}
+	\caption{Specifications of \texttt{inRange} for the give-up template}
 	\label{fig:inRange_giveup}
@@
 			\hline
-			\textbf{Lock-coupling template} & \textbf{Line of proofs} & \textbf{Give-up  template} & \textbf{Line of proofs} \\
+			\textbf{Lock-coupling template} & \textbf{LoC} & \textbf{Give-up  template} & \textbf{LoC} \\
 			\hline
@@
 
-\section{Give-up Proof}
-\label{sec:apd_giveup_proof}
-
-\begin{figure}[h]
+\section{Proofs for Section \ref{} and \ref{label}}
+\label{sec:apd_proof}
+Below are partially completed proofs of \texttt{traverse} for the give-up template (Figure \ref{proof_giveup_traverse}), as well as proofs of \texttt{insert} for the lock-coupling and give-up templates, annotated with separation logic specifications.
+The complete proofs can be found in our Coq mechanization, which is mentioned in Section \ref{proof_mech} for reference.
+
+\begin{figure}[!ht]
 	$\color{blue}
@@
 		$\color{blue} \left\{\begin{array}{l} (\mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots) \  \lor \\
-				(\mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)
-			\end{array}\right\}$
+		(\mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)\end{array}\right\}$
 }  \end{lstlisting}
@@
 
-%List of different predefined enumeration styles:
+\begin{figure}[!ht]
+	$\color{blue}
+	m.\left\langle 
+	\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ m
+	\right\rangle$
+	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
+void insert (node_t **r, int k, void *v){
+		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+		pn->n = *r;
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{p, r}) \ast  \texttt{r->lock} \lockvar \mathsf{R}(\texttt{r}) \ast \treerep\ m \end{array}\right\}$
+		acquire(pn->n->lock);
+		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p, r}) \ast  \texttt{r->lock} \lockvar \mathsf{R}(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \ast \cdots \end{array}\right\}$
+		if (traverse(pn, k) == 0){
+		$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL} \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+			pn->p->t->value = v;
+			$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \texttt{p'->t->value} \mapsto \texttt{v} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$
+			//Linearization point
+			$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+		}
+		else{
+			$\color{blue} \left\{\begin{array}{l}  \mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+			insertOp(pn, k, v);
+			$\color{blue} \left\{\begin{array}{l}  \mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast  \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \\ \texttt{p'->t}\mapsto (\texttt{k, v', l, r})  \ast \texttt{l} \mapsto \texttt{NULL} \ast  \texttt{r} \mapsto \texttt{NULL} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$
+			//Linearization point
+			$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+		}
+		release(pn->n->lock);
+		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+		free(pn);
+		$\color{blue} \left\{\begin{array}{l}  \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+} \end{lstlisting}
+	$\color{blue}
+	\left\langle 
+	\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
+	\right\rangle$
+	\caption{The \texttt{insert} function in lock-coupling template annotated with separation logic specification}
+	\label{proof_lock_insert}
+\end{figure}
+
+\begin{figure}[!ht]
+	$\color{blue}
+	m.\left\langle 
+	\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ m
+	\right\rangle$
+	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
+void insert (node_t **r, int k, void *v){
+		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+		pn->n = *r;
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{p, r}) \ast  \infp(\texttt{r}) \ast \treerep\ m \end{array}\right\}$
+		if (traverse(pn, k) == 0){
+				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+				pn->p->t->value = v;
+				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \texttt{p'->t->value} \mapsto \texttt{v} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$
+				//Linearization point
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+		}
+		else{
+				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+				insertOp(pn, k, v);
+				$\color{blue} \left\{\begin{array}{l}  \mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast  \infp (\texttt{p'}) \ast \\ \texttt{p'->t}\mapsto (\texttt{k, v', l, r})  \ast \texttt{l} \mapsto \texttt{NULL} \ast  \texttt{r} \mapsto \texttt{NULL} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$
+				//Linearization point
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+		}
+		release(pn->n->lock);
+		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \cdots \end{array}\right\}$
+		free(pn);
+		$\color{blue} \left\{\begin{array}{l} \infp (\texttt{p'}) \ast \cdots \end{array}\right\}$
+	} \end{lstlisting}
+	$\color{blue}
+	\left\langle 
+	\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
+	\right\rangle$
+	\caption{The \texttt{insert} function in give-up template annotated with separation logic specification}
+	\label{proof_giveup_insert}
+\end{figure}
 

### 2023-04-30 concurrency/CONCUR 2023/giveup_insert.c

@@ -1,6 +1,6 @@
-void insert (node_t **p, int k, void *v){
+void insert (node_t **r, int k, void *v){
   struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
-  pn->n = *p;
+  pn->n = *r;
   if (traverse(pn, k) == 0){
     pn->p->t->value = v;
   }

### 2023-04-30 concurrency/CONCUR 2023/lock_insert.c

@@ -1,6 +1,6 @@
-void insert (node_t **p, int k, void *v){
+void insert (node_t **r, int k, void *v){
   struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
-  pn->n = *p;
+  pn->n = *r;
   acquire(pn->n->lock);  
   if (traverse(pn, k) == 0){
     pn->p->t->value = v;

### 2023-04-30 concurrency/CONCUR 2023/paper.tex
fixed breakage

@@
 If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
-We can then define \lstinline{inRange} as a helper function in the template, rather than requiring data structures to provide it.
+We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it.
 \wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
@@
 The local precondition include any locks, and consists only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
-As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$.
-Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.
+As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$.
+Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{n'->t} is \texttt{NULL} or not.
 
@@
 For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held.
-Formally, we write \[\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists b.
+Formally, we write \[\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists \ b.
 \ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}\]
@@
 and $\treerep$ contains an $\mathsf{inv\_for\_lock}$ assertion for each node in the data structure.
-As before, $\mathsf{R}$ contains the concrete contents of each node, but also includes the \lstinline{min} and \lstinline{max} fields that we use to implement the \lstinline{inRange} check.
+As before, $\mathsf{R}$ contains the concrete contents of each node, but also includes the \texttt{min} and \texttt{max} fields that we use to implement the \texttt{inRange} check.
 
@@
 \end{mathpar}
-With this invariant in hand, the proof of \lstinline{traverse} proceeds as follows.
+With this invariant in hand, the proof of \texttt{traverse} proceeds as follows.
 We start with the $\infp (\texttt{n})$ predicate from the local precondition.
-After acquiring the lock of the node $\texttt{n}$ (line 5 in Figure \ref{proof_giveup_traverse} in Appendix \ref{sec:apd_giveup_proof}), we obtain the resource $\mathsf{R}(\texttt{n})$, which allows us to perform the \lstinline{inRange} check on line 7.
+After acquiring the lock of the node $\texttt{n}$ (line 5 in Figure \ref{proof_giveup_traverse} in Appendix \ref{sec:apd_proof}), we obtain the resource $\mathsf{R}(\texttt{n})$, which allows us to perform the \texttt{inRange} check on line 7.
 If the check fails (lines 20-23), we immediately release the lock and start over from the beginning.
-Otherwise, we proceed as before: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \lstinline{k} belongs (and can satisfy the postcondition); if \lstinline{findNext} returns 0, we have found the node containing \lstinline{k} (and can satisfy the postcondition); otherwise, we release our lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \lstinline{findNext}.
+Otherwise, we proceed as before: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release our lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}.
 In this case, we do not hold the lock on the new node, but only the $\infp$ assertion indicating that the new node is in the data structure.
 % and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}.
-Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 8 in Figure \ref{traverse_giveup_b}).
+Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \texttt{NULL} (line 8 in Figure \ref{traverse_giveup_b}).
 At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.
@@
 \section{Verified Data Structures}
+\label{BST_proof}
 In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified concurrent data structures in C.
@@
 \subsection{Lock-coupling BST}
+\label{BST_lock_insert}
 \begin{figure}[h]
@@
 \subsection{Give-up BST}
-
+\label{BST_giveup_insert}
 The give-up version of the \texttt{insert} method (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version.
@@
 
-\section{Proofs for Section \ref{} and \ref{label}}
+\section{Proofs for Sections \ref{traverse_proof_giveup}, \ref{BST_lock_insert}, and \ref{BST_giveup_insert}}
 \label{sec:apd_proof}
-Below are partially completed proofs of \texttt{traverse} for the give-up template (Figure \ref{proof_giveup_traverse}), as well as proofs of \texttt{insert} for the lock-coupling and give-up templates, annotated with separation logic specifications.
+Below are partially completed proofs of \texttt{traverse} for the give-up template (Figure \ref{proof_giveup_traverse}), as well as proofs of \texttt{insert} for the lock-coupling (Figure \ref{proof_lock_insert}) and give-up (Figure \ref{proof_giveup_insert}) templates, annotated with separation logic specifications.
 The complete proofs can be found in our Coq mechanization, which is mentioned in Section \ref{proof_mech} for reference.
@@
 	$\color{blue}
-	m.\left\langle 
-	\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ m
+	\forall m.\left\langle 
+	\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
 	\right\rangle$
@@
 	\left\langle 
-	\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
+	\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
 	\right\rangle$

### 2023-04-30 concurrency/CONCUR 2023/paper.tex
fixed breakage

@@
 The template relies on one function provided by the underlying data structure, namely \lstinline{findNext}, which is used to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$.
-In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a boolean and, if a next node is found, modifies \lstinline{pn->n}.
+In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a Boolean and, if a next node is found, modifies \lstinline{pn->n}.
 
@@
 The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}.
-The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not.
-The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its boolean state.
+The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not.
+The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its Boolean state.
 The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
@@
 				\nodeboxrep(\texttt{n})  \ast \mathsf{R}(\texttt{n})  \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} ... \end{array}\right\}$
+		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
+			\texttt{pn} \mapsto (\texttt{p, n}) \ast  \nodeboxrep(\texttt{n}) \ast  \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \treerep\ m
+		 \end{array}\right\}$
 				pn->p = pn->n;
@@
 		\infp (\texttt{n})  \ast \infp (\texttt{r}) \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} ... \end{array}\right\}$
+		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
+			\texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \infp (\texttt{r}) \ast \treerep\ m
+		 \end{array}\right\}$
 				acquire(pn->n->lock);

### 2023-04-30 concurrency/CONCUR 2023/paper.tex
filled some Willam's comment

@@
 %\wm{Does this organization work, or would it be better to prove each template as we introduce it?}
-
+In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification.
 The logically atomic specification for the lock-coupling template's \lstinline{traverse} function is shown in Figure \ref{fig:traverse_lock}.
@@
 
-\begin{figure}[h]
+\begin{figure}[!ht]
 	$\color{blue}
@@
 
-Given this invariant, the proof proceeds as follows. %\wm{Should we consider giving annotated code after all?} 
+Figure \ref{proof_lock_traverse} shows the \texttt{traverse} function in lock-coupling template with separation logic annotations.
+Given this invariant, the proof proceeds as follows.
+%\wm{Should we consider giving annotated code after all?}
 We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$.
-Otherwise, we use \lstinline{findNext} to find a new node \lstinline{n'} to visit (line 8), which is stored in \lstinline{n}.
+Otherwise, we use \lstinline{findNext} to find a new node \lstinline{n'} to visit (line 9), which is stored in \lstinline{n}.
 If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$.
@@
 For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held.
-Formally, we write \[\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists \ b.
-\ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}\]
+Formally, we write $$\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists \ b.
+\ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}$$
 %\wm{Consider using mathsf for logic and reserving texttt for C code.}
@@
 \end{mathpar}
+The \texttt{traverse} function, as illustrated within the context of the give-up template and annotated using separation logic, can be found in Figure \ref{proof_giveup_traverse} of Appendix \ref{sec:apd_proof}.
 With this invariant in hand, the proof of \texttt{traverse} proceeds as follows.
 We start with the $\infp (\texttt{n})$ predicate from the local precondition.
-After acquiring the lock of the node $\texttt{n}$ (line 5 in Figure \ref{proof_giveup_traverse} in Appendix \ref{sec:apd_proof}), we obtain the resource $\mathsf{R}(\texttt{n})$, which allows us to perform the \texttt{inRange} check on line 7.
-If the check fails (lines 20-23), we immediately release the lock and start over from the beginning.
+After acquiring the lock of the node $\texttt{n}$ (line 5 in Figure \ref{proof_giveup_traverse} in Appendix \ref{sec:apd_proof}), we obtain the resource $\mathsf{R}(\texttt{n})$, which allows us to perform the \texttt{inRange} check on line 8.
+If the check fails (lines 28-32), we immediately release the lock and start over from the beginning.
 Otherwise, we proceed as before: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release our lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}.
@@
 The target specifications for our search structure operations are shown below: each operation should atomically have its effect on the data structure, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
-
 \begin{mathpar}
 	{\color{blue}
-		m.\left\langle 
+		\forall m.\left\langle 
 		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
@@
 	{\color{blue}
-		m.\left\langle 
+		\forall m.\left\langle 
 		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
@@
 
-\wm{Should we have a separate section where we introduce findNext (and maybe insertOp) before we get into specific templates?}
+%\wm{Should we have a separate section where we introduce findNext (and maybe insertOp) before we get into specific templates?}
 
@@
 
-\wm{Should we recap the definition of the \lstinline{node} struct here?} The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with the current node's key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
+%\wm{Should we recap the definition of the \lstinline{node} struct here?} 
+Let's recall the \texttt{node} struct for a binary search tree. 
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+	typedef struct node {int key; void *value; struct node_t *left, *right;} node;
+\end{lstlisting}
+The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with the current node's key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
 The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the current node's key \texttt{k'}, and the next node \lstinline{n'} is the left child \lstinline{l}; \lstinline{k} is greater than \lstinline{k'} and \lstinline{n'} is the right child \lstinline{r}; or \lstinline{k} is exactly \lstinline{k'} and \lstinline{n'} is the original node \lstinline{n}, in which case the return value is 0.
@@
 It inserts a new node with the specified key \texttt{k} and value \texttt{v} at that leaf, complete with two new empty child nodes with corresponding locks, \texttt{(NULL, lk1)} and \texttt{(NULL, lk2)}, which are pointed to by the \texttt{l} and \texttt{r} of the current node.
-
-\than{Talking about different of \texttt{insertOp} specs here.}
 
@@
 
+Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template for BST, annotated using separation logic.
 The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{p}$.
-We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{insert_lock}).
-After the lock for the root node is acquired (line 4 in Figure \ref{insert_lock}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$.
+We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{proof_lock_insert}).
+After the lock for the root node is acquired (line 5 in Figure \ref{proof_lock_insert}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$.
 We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
-
+\begin{mathpar}
+	% spec of inRange
+	{\color{blue}
+		\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
+		\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
+		\right\rangle
+	}
+	\\ 
+	\texttt{traverse(pn, k)} 
+	\\
+	{\color{blue}
+		\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
+		\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
+			\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
+		\end{array}
+		\ \Bigg| \ \treerep\ m \
+		\right\rangle
+	}
+\end{mathpar}
 In the $\mathsf{then}$ branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
 we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree.
-In this scenario, we update the node's value while maintaining the rest of the tree structure (line 6 in Figure \ref{insert_lock}).
-Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 9 in Figure \ref{insert_lock}).
+In this scenario, we update the node's value while maintaining the rest of the tree structure (line 9 in Figure \ref{proof_lock_insert}).
+Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 14-15 in Figure \ref{proof_lock_insert}).
 We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree.
@@
 We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
-We can then release the lock (line 11 in Figure \ref{insert_lock}) return the node's contents to the shared state.
+We can then release the lock (line 21 in Figure \ref{proof_lock_insert}) return the node's contents to the shared state.
 %Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
-\wm{Do we ever define $\treerep$?}
+%\wm{Do we ever define $\treerep$?}
 
@@
 The give-up BST uses the same definition and specification of \texttt{findNext}, while introducing an extra helper function called \texttt{inRange} \wm{If \lstinline{inRange} is part of the template, I think we should discuss it in the previous section}.
-The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.
+\than{If we move it, this section is very short.} The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.
 
@@
 The proof of the \lstinline{insert} function is similar to that of the previous function: the only difference is that the give-up version of \lstinline{insert} does not acquire a lock before calling \lstinline{traverse}, reflecting the difference in the locking patterns and the preconditions of \lstinline{traverse} between the two templates.
+In Appendix \ref{sec:apd_proof}, Figure \ref{proof_giveup_insert} presents an annotation that uses separation logic for the \texttt{insert} function within the give-up template for BST.
 
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!h]
 	$\color{blue}

### 2023-05-01 concurrency/CONCUR 2023/paper.tex
first pass on related work

@@
 \subsection*{Related Work}
-CSL provers for fine-grained concurrent data structures: Iris~\cite{iris,iris-folly}, FCSL~\cite{fcsl}, TaDA~\cite{tada,Xiong2017Abstract}, VeriFast~\cite{verifast,verifast-conc}
+Recent years have seen major advances in concurrent separation logics for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}.
+The innovations of Iris (custom ghost state and invariants) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach.
+Some of the more complex data structures verified include the \lstinline{ConcurrentSkipListMap} from \lstinline{java.util.concurrent}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}.
+Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeted at an implementation language (C) rather than a toy language or algorithm-level representation.
+%do we want to say this part?
 
@@
 
-%Xiong et al.
-\cite{Xiong2017Abstract} have demonstrated the verification of ConcurrentSkipListMap from java.util.concurrent library using the recent advances in fine-grained concurrency reasoning.
-Their work is mainly based on the abstract atomicity from TaDA logic, and give two modular specifications for concurrent maps: one specification focus on the entire map structure which is suitable for verifying implementation, and another specification focus on the key-value pairs which appropriate for verifying clients.
-We use the same idea of atomicity (though implemented in Iris) in our work.
-
-separating concurrency from functionality: linearizability~\cite{linearizability} (and refinement, e.g. Civl~\cite{civl}), logical atomicity~\cite{tada}
-
-Proving highly-concurrent traversals correct \cite{feldman2020proving}
-
-multicopy search structures~\cite{template-multi}
+Separating concurrency reasoning from data structure reasoning has long been an appealing target.
+Linearizability~\cite{linearizability}, the most common correctness condition for concurrent data structures, was a first step in this direction, describing the conditions under which a concurrent implementation can replace a sequential one in all possible contexts.
+Logical atomicity can be seen as a compositional analogue to linearizability, where the proof that each operation implements its sequential counterpart can be carried out independently.
+Concurrent templates can be seen as the next step of compositionality, allowing the concurrent-to-sequential proof to be decoupled from the specific data structure being verified.
+
+%something about refinement, e.g. Civl~\cite{civl}
+
+It is also worth mentioning recent work that extends the reach of template-style reasoning.
+Later work on search structure templates by Patel et al.~\cite{template-multi} applies the template approach to multicopy search structures, where there may be more than one node containing the target key.
+Feldman et al.~\cite{feldman2020proving} take an approach similar to templates to verify search structures with highly optimistic concurrency patterns, where the data structure may be restructured by other threads during traversal.
+Their technique could potentially be applied to prove correctness of much more complex \lstinline{traverse} operations than those we describe.
 

### 2023-05-01 concurrency/CONCUR 2023/paper.tex
simple example template

@@
 
-We will discuss several templates in section~\ref{templates}, but broadly, a search structure template consists of a \lstinline{traverse} function that travels through a data structure in search of a particular key, using the appropriate concurrency control mechanism when moving from node to node.
+\begin{figure}[h]
+	\begin{subfigure}[t]{0.45\textwidth}
+		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
+%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
+%		\label{traverse_lock_a}	
+	\end{subfigure}\qquad
+	\begin{subfigure}[t]{0.48\textwidth}
+	\begin{lstlisting}[language=caml, style=myStyle]
+	let insert r k =
+	  lockNode r;
+	  let n = traverse r k in
+	  let res = insertOp n k in
+	  unlockNode n; res
+	\end{lstlisting}
+%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
+%		\label{traverse_lock_b}
+	\end{subfigure}
+	\caption{An example search structure template (lock coupling)}
+	\label{template-ex}
+\end{figure}
+
+Figure~\ref{template-ex} shows an example search structure template.
+The core of the template is the \lstinline{traverse} function, which uses a particular concurrency control mechanism to travel through a data structure in search of a particular key.
+The mechanism in this example is \emph{lock coupling}, where we acquire the lock on the next node before releasing the lock on the current node.
 The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine when the current node does not contain the target key.
-\wm{Show an example template here, and enough of the specification to explain atomicity -- one findNext, one traverse, one top-level insert -- see fig 3 in template paper} Once the appropriate node for the key has been found, the template returns it to a top-level function that calls out to the data structure to perform the appropriate operation (e.g.
-insert, lookup, delete) on that node.
-Thus, the \lstinline{traverse} function can be verified without knowing anything about the target data structure other than its synchronization mechanism, and without knowing anything about the operation to be performed other than that it can be performed on a single node (or sometimes a node and its parent).
-The job of the data structure is to implement the \lstinline{findNext} function and the specific operations; the combination of the template and the data structure is then a fully operational concurrent data structure.
-Krishna et al.~presented templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures, but many common search structures are not graphs; in this paper, we present a version of templates that is independent of flow interfaces.
+Once the appropriate node for the key has been found, the template returns it to a top-level function such as \lstinline{insert} that calls out to the data structure to perform the actual insertion on the node.
+Thus, the \lstinline{traverse} and \lstinline{insert} functions can be verified without knowing anything about the target data structure other than its synchronization mechanism, as long as the data structure implements \lstinline{findNext} and \lstinline{insertOp} operations with the required semantics.
+The concurrency functions in the template and the sequential \lstinline{findNext} and \lstinline{insertOp} functions provided by the data structure combine into a fully operational concurrent data structure.
+Krishna et al.~specify templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures, but many common search structures are not graphs; in this paper, we present a version of templates that is independent of flow interfaces.
 %is this a contribution in itself?

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
represented annotated proofs as files rather than showing them fully in tex

@@
 Logical atomicity is key to the template approach: each template's \lstinline{traverse} function is proved to satisfy a logically atomic specification that says roughly ``this function finds the node where key \lstinline{k} belongs''.
-The traverse specifications can then be used to prove atomic specifications for the individual data structure operations, lifting the sequential specifications for insert, lookup, etc.
+The \lstinline{traverse} specification can then be used to prove atomic specifications for the individual data structure operations, lifting the sequential specifications for insert, lookup, etc.
 to the concurrent setting.
@@
 These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node.
-Formally, the resource  $\mathsf{R}(\texttt{n}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{DataStruct\_specific(\texttt{n})}$.
+In formal terms, the resource is represented as $\mathsf{R}(\texttt{n}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{DataStruct\_specific(\texttt{n})}$.
 
@@
 The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its Boolean state.
-The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
+The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
 \begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'})\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'}) \ \ast \ \treerep\ m
@@
 The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template.
-The local precondition include any locks, and consists only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
+The local preconditions include any locks, and consist only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
 As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$.
@@
 
-As previously mentioned, the \texttt{insertOp} specification (Figure \ref{fig:insertOp_giveup}) is slightly different from the lock-coupling version, to account for the added \lstinline{min} and \lstinline{max} fields.
+The \texttt{insertOp} specification (Figure \ref{fig:insertOp_giveup}) differs slightly from the lock-coupling version, to account for the added \lstinline{min} and \lstinline{max} fields.
 In a BST, if the current node's range is \lstinline{(min, max)} and its key is \lstinline{k}, its left child will have range \texttt{(min, k)} and its right child will have range \texttt{(k, max)}.
@@
 	\right\rangle$
-	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
-int traverse(pn *pn, int k){
-		node_t *r = (pn->n);
-		int flag = 1; 			$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
-		\infp (\texttt{n})  \ast \infp (\texttt{r}) \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
-			\texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \infp (\texttt{r}) \ast \treerep\ m
-		 \end{array}\right\}$
-				acquire(pn->n->lock);
-				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
-				pn->p = pn->n; $\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
-				if (inRange(pn, k) == 1){
-				$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast  \cdots \end{array}\right\}$
-						if (pn->p->t == NULL)
-								break;
-								$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
-						else{
-								int b = findNext(pn, k);
-								$\color{blue} \left\{\begin{array}{l} \exists \  \texttt{n'}. \ \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \\ 
-								(\mathit{res'} = \mathsf{false} \ast \texttt{n'} = \texttt{n} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots )\  \lor \\ 
-								(\mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \infp (\texttt{n'}) \ast \cdots)   \end{array}\right\}$
-								if (b == 0){
-										flag = 0;
-										break;
-										$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL}  \ast \texttt{n'} = \texttt{n} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots   \end{array}\right\}$
-								}
-								else
-										$\color{blue} \left\{\begin{array}{l} \mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n->t} \neq \texttt{NULL} \ast \infp (\texttt{n'}) \ast \cdots    \end{array}\right\}$
-										release(pn->p->lock);
-										$\color{blue} \left\{\begin{array}{l} \mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n->t} \neq \texttt{NULL} \ast \infp (\texttt{n'}) \ast \cdots    \end{array}\right\}$
-						}
-				}
-				else{
-						$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast  \cdots \end{array}\right\}$
-						release(pn->p->lock);
-						$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast  \cdots \end{array}\right\}$
-						pn->n = r;
-						$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast  \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{r, r}) \ast \infp (\texttt{r}) \ast \cdots    \end{array}\right\}$
-				}
-		}
-		return flag;
-		$\color{blue} \left\{\begin{array}{l} (\mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots) \  \lor \\
-		(\mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)\end{array}\right\}$
-}  \end{lstlisting}
+		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
 	$\color{blue}
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
+	\begin{subfigure}{\textwidth}
 		$\color{blue}
@@
 		\right\rangle$
-	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
-void insert (node_t **r, int k, void *v){
-		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
-		pn->n = *r;
-		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{p, r}) \ast  \texttt{r->lock} \lockvar \mathsf{R}(\texttt{r}) \ast \treerep\ m \end{array}\right\}$
-		acquire(pn->n->lock);
-		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p, r}) \ast  \texttt{r->lock} \lockvar \mathsf{R}(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \ast \cdots \end{array}\right\}$
-		if (traverse(pn, k) == 0){
-		$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL} \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
-				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \texttt{p'->t->value} \mapsto \texttt{v} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$
-				//Linearization point
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
-		}
-		else{
-				$\color{blue} \left\{\begin{array}{l}  \mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
-				insertOp(pn, k, v);
-				$\color{blue} \left\{\begin{array}{l}  \mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast  \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \\ \texttt{p'->t}\mapsto (\texttt{k, v', l, r})  \ast \texttt{l} \mapsto \texttt{NULL} \ast  \texttt{r} \mapsto \texttt{NULL} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$
-				//Linearization point
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
-		}
-		release(pn->n->lock);
-		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
-		free(pn);
-		$\color{blue} \left\{\begin{array}{l}  \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
-} \end{lstlisting}
+		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
 		$\color{blue}
@@
 		\right\rangle$
-	\caption{The \texttt{insert} function in lock-coupling template annotated with separation logic specification}
+		\caption{The \texttt{insert} function in lock-coupling template annotated with separation logic specification. \\\\}
 		\label{proof_lock_insert}	
-\end{figure}
-
-\begin{figure}[!h]
+	\end{subfigure}\qquad
+	\begin{subfigure}{\textwidth}
 		$\color{blue}
@@
 		\right\rangle$
-	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
-void insert (node_t **r, int k, void *v){
-		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
-		pn->n = *r;
-		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{p, r}) \ast  \infp(\texttt{r}) \ast \treerep\ m \end{array}\right\}$
-		if (traverse(pn, k) == 0){
-				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
-				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \texttt{p'->t->value} \mapsto \texttt{v} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$
-				//Linearization point
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
-		}
-		else{
-				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
-				insertOp(pn, k, v);
-				$\color{blue} \left\{\begin{array}{l}  \mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast  \infp (\texttt{p'}) \ast \\ \texttt{p'->t}\mapsto (\texttt{k, v', l, r})  \ast \texttt{l} \mapsto \texttt{NULL} \ast  \texttt{r} \mapsto \texttt{NULL} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$
-				//Linearization point
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
-		}
-		release(pn->n->lock);
-		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \cdots \end{array}\right\}$
-		free(pn);
-		$\color{blue} \left\{\begin{array}{l} \infp (\texttt{p'}) \ast \cdots \end{array}\right\}$
-	} \end{lstlisting}
+		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
 		$\color{blue}
@@
 		\right\rangle$
-	\caption{The \texttt{insert} function in give-up template annotated with separation logic specification}
+		\caption{The \texttt{insert} function in give-up template annotated with separation logic specification. \\\\}
 		\label{proof_giveup_insert}
+	\end{subfigure}
+	\caption{The \texttt{insert} function in lock-coupling and give-up templates annotated with separation logic specification.}
+	\label{proof_lock_giveup_insert}
 \end{figure}

### 2023-05-02 concurrency/CONCUR 2023/proof_giveup_insert.c

@@ -1,0 +1,23 @@
+void insert (node_t **r, int k, void *v){
+		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+		pn->n = *r;
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{p, r}) \ast  \infp(\texttt{r}) \ast \treerep\ m \end{array}\right\}$
+		if (traverse(pn, k) == 0){
+				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+				pn->p->t->value = v;
+				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \texttt{p'->t->value} \mapsto \texttt{v} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
+				//Linearization point
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+		}
+		else{
+				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+				insertOp(pn, k, v);
[…]

### 2023-05-02 concurrency/CONCUR 2023/proof_giveup_traverse.c

@@ -1,0 +1,43 @@
+int traverse(pn *pn, int k){
+		node_t *r = (pn->n);
+		int flag = 1; 			$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
+		\infp (\texttt{n})  \ast \infp (\texttt{r}) \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
+			\texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \infp (\texttt{r}) \ast \treerep\ m
+		 \end{array}\right\}$
+				acquire(pn->n->lock);
+				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
+				pn->p = pn->n; $\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
+				if (inRange(pn, k) == 1){
+				$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast  \cdots \end{array}\right\}$
+						if (pn->p->t == NULL)
+								break;
[…]

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_insert.c

@@ -1,0 +1,25 @@
+void insert (node_t **r, int k, void *v){
+		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+		pn->n = *r;
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{p, r}) \ast  \texttt{r->lock} \lockvar \mathsf{R}(\texttt{r}) \ast \treerep\ m \end{array}\right\}$
+		acquire(pn->n->lock);
+		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p, r}) \ast  \texttt{r->lock} \lockvar \mathsf{R}(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \ast \cdots \end{array}\right\}$
+		if (traverse(pn, k) == 0){
+		$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL} \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+				pn->p->t->value = v;
+				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \texttt{p'->t->value} \mapsto \texttt{v} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
+				//Linearization point
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+		}
+		else{
[…]

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
represented annotated proofs as files rather than showing them fully in tex

@@
 	\right\rangle$
-\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
-int traverse(pn *pn, int k){
-		int flag = 1; 			$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
-				\nodeboxrep(\texttt{n})  \ast \mathsf{R}(\texttt{n})  \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
-			\texttt{pn} \mapsto (\texttt{p, n}) \ast  \nodeboxrep(\texttt{n}) \ast  \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \treerep\ m
-		 \end{array}\right\}$
-				pn->p = pn->n;
-				if (pn->p->t == NULL)
-						break;
-						$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
-				else{
-						int b = findNext(pn, k);
-						$\color{blue} \left\{\begin{array}{l} \exists \  \texttt{n'}. \ \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \\ 
-						(\mathit{res'} = \mathsf{false} \ast \texttt{n'} = \texttt{n} \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots )\  \lor \\ 
-						(\mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \cdots)   \end{array}\right\}$
-				if (b == 0){
-						flag = 0;
-						break;
-						$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \texttt{n'} = \texttt{n} \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots    \end{array}\right\}$
-				}
-				else{
-				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \cdots    \end{array}\right\}$
-						acquire(pn->n->lock);
-						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \\ \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \mathsf{R}(\texttt{n'}) \ast \cdots    \end{array}\right\}$
-						release(pn->p->lock);
-						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \\ \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \mathsf{R}(\texttt{n'}) \ast \cdots    \end{array}\right\}$
-					}
-				}
-			}
-		return flag;
-		$\color{blue} \left\{\begin{array}{l} (\mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL} \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots) \  \lor \\
-		(\mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)
-		\end{array}\right\}$
-}  \end{lstlisting}
+		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
 	$\color{blue}
@@
 \bibliography{../sources}
-
+\newpage
 \appendix
@@
 		\right\rangle$
-		\caption{The \texttt{insert} function in lock-coupling template annotated with separation logic specification. \\\\}
+		\caption{The \texttt{insert} function in lock-coupling template annotated with separation logic specification \\\\}
 		\label{proof_lock_insert}	
@@
 		\right\rangle$
-		\caption{The \texttt{insert} function in give-up template annotated with separation logic specification. \\\\}
+		\caption{The \texttt{insert} function in give-up template annotated with separation logic specification \\\\}
 		\label{proof_giveup_insert}
 	\end{subfigure}
-	\caption{The \texttt{insert} function in lock-coupling and give-up templates annotated with separation logic specification.}
+	\caption{The \texttt{insert} function in lock-coupling and give-up templates annotated with separation logic specification}
 	\label{proof_lock_giveup_insert}

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_traverse.c

@@ -1,0 +1,34 @@
+int traverse(pn *pn, int k){
+		int flag = 1; 			$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
+				\nodeboxrep(\texttt{n})  \ast \mathsf{R}(\texttt{n})  \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
+			\texttt{pn} \mapsto (\texttt{p, n}) \ast  \nodeboxrep(\texttt{n}) \ast  \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \treerep\ m
+		 \end{array}\right\}$
+				pn->p = pn->n;
+				if (pn->p->t == NULL)
+						break;
+						$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
+				else{
+						int b = findNext(pn, k);
+						$\color{blue} \left\{\begin{array}{l} \exists \  \texttt{n'}. \ \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \\ 
+						(\mathit{res'} = \mathsf{false} \ast \texttt{n'} = \texttt{n} \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots )\  \lor \\ 
[…]

### 2023-05-02 concurrency/CONCUR 2023/proof_giveup_traverse.c
fix breakage of annotations

@@ -16,17 +16,19 @@ int traverse(pn *pn, int k){
 						else{
 								int b = findNext(pn, k);
 								$\color{blue} \left\{\begin{array}{l} \exists \  \texttt{n'}. \ \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \\ 
-								(\mathit{res'} = \mathsf{false} \ast \texttt{n'} = \texttt{n} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots )\  \lor \\ 
-								(\mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \infp (\texttt{n'}) \ast \cdots)   \end{array}\right\}$
+								((\mathit{res'} = \mathsf{false} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots )\  \lor \\ 
+								\ (\mathit{res'} = \mathsf{true}  \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \infp (\texttt{n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'}) \ast \cdots)   \end{array}\right\}$
 								if (b == 0){
+										$\color{blue} \left\{\begin{array}{l} \mathit{res'} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
 										flag = 0;
[…]

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_traverse.c

@@ -10,23 +10,28 @@ int traverse(pn *pn, int k){
 						$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
 				else{
 						int b = findNext(pn, k);
-						$\color{blue} \left\{\begin{array}{l} \exists \  \texttt{n'}. \ \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \\ 
-						(\mathit{res'} = \mathsf{false} \ast \texttt{n'} = \texttt{n} \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots )\  \lor \\ 
-						(\mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \cdots)   \end{array}\right\}$
-				if (b == 0){
-						flag = 0;
-						break;
-						$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \texttt{n'} = \texttt{n} \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots    \end{array}\right\}$
[…]

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_traverse.c
fix breakage of annotations

@@ -5,6 +5,7 @@ int traverse(pn *pn, int k){
 			\texttt{pn} \mapsto (\texttt{p, n}) \ast  \nodeboxrep(\texttt{n}) \ast  \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \treerep\ m
 		 \end{array}\right\}$
 				pn->p = pn->n;
+				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n, n}) \ast  \nodeboxrep(\texttt{n}) \ast  \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \treerep\ m\end{array}\right\}$
 				if (pn->p->t == NULL)
 						break;
 						$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
fix breakage of annotations
fix breakage

@@
 	  let res = insertOp n k in
-	  unlockNode n; res
-	\end{lstlisting}
+	  unlockNode n; 
+	  res\end{lstlisting}
 %		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
@@
 			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-			\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
+			\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
 				\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
@@
 	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-	\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
+	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
 		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
@@
 			\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
-			\begin{array}{c}
+			\begin{array}{l}
 				\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
@@
 		\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-		\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
+		\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
 			\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
@@
 			\left\{\mathit{res.} \
-			\begin{array}{c}
+			\begin{array}{l}
 				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} \  \ast  \ 
@@
 	\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
-	\begin{array}{c}
+	\begin{array}{l}
 		\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
move insert to section templates

@@
 where the \emph{lock invariant assertion} $\ell \lockvar \mathsf{R}$ expresses that the lock $\ell$ protects resources $\mathsf{R}$.
-In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{pn->n}).
+In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).
 
@@
 %The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
+
+\than{I moved \texttt{insert} from Section 4 to here.
+William: Could you please write something here to have a smooth transition from \texttt{traverse} to \texttt{insert}, or do you think you need to have subsections for \texttt{traverse} and \texttt{insert}?
+}
+
+\begin{figure}[h]
+	\begin{subfigure}[t]{0.48\textwidth}
+		\lstinputlisting[language=C, style=myStyle, escapechar=|]{lock_insert.c} 
+		\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
+		\label{insert_lock}	
+	\end{subfigure}\qquad
+	\begin{subfigure}[t]{0.48\textwidth}
+		\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
+		\caption{The \lstinline{insert} method of the give-up template algorithm}
+		\label{insert_giveup}
+	\end{subfigure} 
+	\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
+	\label{insert_lock_giveup} 
+\end{figure}
+
+\than{Should we move the code to template section?}
+
+Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling template.
+%\wm{Actually, is this \lstinline{insert} function part of the template?
+If so, should we discuss it in section 3?}
+It makes use of the \texttt{traverse} function %of Section \ref{lock-coupling-algo} 
+to find the node at which to insert the key \lstinline{k}.
+If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}).
+Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the data-structure-specific \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
+
+The atomic specification for \texttt{insert} method is
+\begin{mathpar}
+	{\color{blue}
+		m.\left\langle 
+		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
+		\right\rangle
+	}
+	\ \texttt{insert(r, k, v)}\ 
+	{\color{blue}
+		\left\langle 
+		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
+		\right\rangle
+	}
+\end{mathpar}
+\wm{Consider skipping this paragraph if we need space.} Similar to the \texttt{traverse} specification, the \texttt{insert} function's specification is composed of two components: local and public pre- and postconditions.
+The local pre- and postconditions, containing $\nodeboxrep \ \texttt{r}$, describe the binary search tree's concrete state with the root node $\texttt{r}$, which points to the node's lock and lock invariant assertion.
+On the other hand, the public pre- and postconditions are represented by the predicate $\treerep \ m$, illustrating the binary search tree's abstract state as a collection of ghost states, with each state providing information about an individual node in the tree.
+Here, $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively.
+
+Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template for BST, annotated using separation logic.
+The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{p}$.
+We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{proof_lock_insert}).
+After the lock for the root node is acquired (line 5 in Figure \ref{proof_lock_insert}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$.
+We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
+\begin{mathpar}
+	% spec of inRange
+	{\color{blue}
+		\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
+		\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
+		\right\rangle
+	}
+	\\ 
+	\texttt{traverse(pn, k)} 
+	\\
+	{\color{blue}
+		\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
+		\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
+			\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
+		\end{array}
+		\ \Bigg| \ \treerep\ m \
+		\right\rangle
+	}
+\end{mathpar}
+In the $\mathsf{then}$ branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
+we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree.
+In this scenario, we update the node's value while maintaining the rest of the tree structure (line 9 in Figure \ref{proof_lock_insert}).
+Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 14-15 in Figure \ref{proof_lock_insert}).
+We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree.
+In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}.
+We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
+We can then release the lock (line 21 in Figure \ref{proof_lock_insert}) return the node's contents to the shared state.
+%Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
+%\wm{Do we ever define $\treerep$?}
 
@@
 
+\than{I moved \texttt{insert} from Section 4 to here.
+William: Could you please write something here to have a smooth transition from \texttt{traverse} to \texttt{insert}, or do you think you need to have subsections for \texttt{traverse} and \texttt{insert}?
+}
+
+The give-up version of the \texttt{insert} method (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version.
+The one difference is that it does not require acquiring a lock before calling its \texttt{traverse} function.
+The proof of the \lstinline{insert} function is similar to that of the previous function: the only difference is that the give-up version of \lstinline{insert} does not acquire a lock before calling \lstinline{traverse}, reflecting the difference in the locking patterns and the preconditions of \lstinline{traverse} between the two templates.
+In Appendix \ref{sec:apd_proof}, Figure \ref{proof_giveup_insert} presents an annotation that uses separation logic for the \texttt{insert} function within the give-up template for BST.
+
 \section{Verified Data Structures}
@@
 
-\subsection{Lock-coupling BST}
-\label{BST_lock_insert}
-\begin{figure}[h]
-	\begin{subfigure}[t]{0.48\textwidth}
-	\lstinputlisting[language=C, style=myStyle, escapechar=|]{lock_insert.c} 
-	\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
-	\label{insert_lock}	
-	\end{subfigure}\qquad
-    \begin{subfigure}[t]{0.48\textwidth}
-    \lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
-    \caption{The \lstinline{insert} method of the give-up template algorithm}
-    \label{insert_giveup}
-    \end{subfigure} 
-\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
-\label{insert_lock_giveup} 
-\end{figure}
-
-Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling BST.
-\wm{Actually, is this \lstinline{insert} function part of the template?
-If so, should we discuss it in section 3?} It makes use of the \texttt{traverse} function of Section \ref{lock-coupling-algo} to find the node at which to insert the key \lstinline{k}.
-If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}).
-Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the data-structure-specific \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
-
-The atomic specification for \texttt{insert} method is
-\begin{mathpar}
-	{\color{blue}
-		m.\left\langle 
-		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
-		\right\rangle
-	}
-	\ \texttt{insert(r, k, v)}\ 
-	{\color{blue}
-		\left\langle 
-		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
-		\right\rangle
-	}
-\end{mathpar}
-\wm{Consider skipping this paragraph if we need space.} Similar to the \texttt{traverse} specification, the \texttt{insert} function's specification is composed of two components: local and public pre- and postconditions.
-The local pre- and postconditions, containing $\nodeboxrep \ \texttt{r}$, describe the binary search tree's concrete state with the root node $\texttt{r}$, which points to the node's lock and lock invariant assertion.
-On the other hand, the public pre- and postconditions are represented by the predicate $\treerep \ m$, illustrating the binary search tree's abstract state as a collection of ghost states, with each state providing information about an individual node in the tree.
-Here, $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively.
-
-Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template for BST, annotated using separation logic.
-The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{p}$.
-We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{proof_lock_insert}).
-After the lock for the root node is acquired (line 5 in Figure \ref{proof_lock_insert}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$.
-We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
-\begin{mathpar}
-	% spec of inRange
-	{\color{blue}
-		\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
-		\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
-		\right\rangle
-	}
-	\\ 
-	\texttt{traverse(pn, k)} 
-	\\
-	{\color{blue}
-		\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-		\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
-			\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
-		\end{array}
-		\ \Bigg| \ \treerep\ m \
-		\right\rangle
-	}
-\end{mathpar}
-In the $\mathsf{then}$ branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
-we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree.
-In this scenario, we update the node's value while maintaining the rest of the tree structure (line 9 in Figure \ref{proof_lock_insert}).
-Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 14-15 in Figure \ref{proof_lock_insert}).
-We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree.
-In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}.
-We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
-We can then release the lock (line 21 in Figure \ref{proof_lock_insert}) return the node's contents to the shared state.
-%Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
-%\wm{Do we ever define $\treerep$?}
-
-\subsection{Give-up BST}
-\label{BST_giveup_insert}
-The give-up version of the \texttt{insert} method (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version.
-The one difference is that it does not require acquiring a lock before calling its \texttt{traverse} function.
-
 The give-up BST uses the same definition and specification of \texttt{findNext}, while introducing an extra helper function called \texttt{inRange} \wm{If \lstinline{inRange} is part of the template, I think we should discuss it in the previous section}.
-\than{If we move it, this section is very short.} The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.
+The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.
 
@@
 The \texttt{inRange} function (Figure \ref{fig:inRange_giveup}) simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
+
+\subsection{Lock-coupling BST}
+\label{BST_lock_insert}
+\than{What do we need to say here? (1) - still keep subsection and filled \texttt{insertOp} and the way we prove \texttt{insert} here, or (2) - get rid of this subsection.}
+\subsection{Give-up BST}
+\label{BST_giveup_insert}
+\than{What do we need to say here? (1) - still keep subsection and filled \texttt{insertOp}, and \texttt{inRange} and the way we prove \texttt{insert} here, or (2) - get rid of this subsection.}
 
@@
 Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ \texttt{(m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.
-
-The proof of the \lstinline{insert} function is similar to that of the previous function: the only difference is that the give-up version of \lstinline{insert} does not acquire a lock before calling \lstinline{traverse}, reflecting the difference in the locking patterns and the preconditions of \lstinline{traverse} between the two templates.
-In Appendix \ref{sec:apd_proof}, Figure \ref{proof_giveup_insert} presents an annotation that uses separation logic for the \texttt{insert} function within the give-up template for BST.
 

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
move insert to section templates

@@
 \begin{figure}[h]
-	\begin{subfigure}[t]{0.45\textwidth}
+	\begin{subfigure}[t]{0.48\textwidth}
+		\begin{subfigure}[t]{\textwidth}
 			\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
@@
 			\label{traverse_lock_a}	
+		\end{subfigure}
+		\\ \\ 
+		\begin{subfigure}[t]{\textwidth}
+			\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
+			\caption{The \lstinline{insert} method of the lock-coupling template algorithm written in C}
+			\label{insert_lock}	
+		\end{subfigure}
 	\end{subfigure}\qquad
@@
 To utilize the template methodology for concurrent data structure implementations, it is crucial to keep track of the locks acquired for each node during the traversal.
+
+\than{I moved \texttt{insert} code of lock-coupling here}
+
+Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling template.
+%\wm{Actually, is this \lstinline{insert} function part of the template?
+If so, should we discuss it in section 3?}
+It makes use of the \texttt{traverse} function %of Section \ref{lock-coupling-algo} 
+to find the node at which to insert the key \lstinline{k}.
+If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}).
+Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the data-structure-specific \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
+\than{...
+that we will discuss in Section Verified DS.}
 
@@
 \than{I moved \texttt{insert} from Section 4 to here.
-William: Could you please write something here to have a smooth transition from \texttt{traverse} to \texttt{insert}, or do you think you need to have subsections for \texttt{traverse} and \texttt{insert}?
-}
-
-\begin{figure}[h]
-	\begin{subfigure}[t]{0.48\textwidth}
-		\lstinputlisting[language=C, style=myStyle, escapechar=|]{lock_insert.c} 
-		\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
-		\label{insert_lock}	
-	\end{subfigure}\qquad
-	\begin{subfigure}[t]{0.48\textwidth}
-		\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
-		\caption{The \lstinline{insert} method of the give-up template algorithm}
-		\label{insert_giveup}
-	\end{subfigure} 
-	\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
-	\label{insert_lock_giveup} 
-\end{figure}
-
-\than{Should we move the code to template section?}
-
-Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling template.
-%\wm{Actually, is this \lstinline{insert} function part of the template?
-If so, should we discuss it in section 3?}
-It makes use of the \texttt{traverse} function %of Section \ref{lock-coupling-algo} 
-to find the node at which to insert the key \lstinline{k}.
-If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}).
-Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the data-structure-specific \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
+William, could you please write something to ensure a smooth transition between \texttt{traverse} and \texttt{insert}, or do you think it would be better to have subsections for \texttt{traverse} and \texttt{insert}?
+}
+
+%\begin{figure}[h]
+%	\begin{subfigure}[t]{0.48\textwidth}
+%		\lstinputlisting[language=C, style=myStyle, escapechar=|]{lock_insert.c} 
+%		\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
+%		\label{insert_lock}	
+%	\end{subfigure}\qquad
+%	\begin{subfigure}[t]{0.48\textwidth}
+%		\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
+%		\caption{The \lstinline{insert} method of the give-up template algorithm}
+%		\label{insert_giveup}
+%	\end{subfigure} 
+%	\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
+%	\label{insert_lock_giveup} 
+%\end{figure}
+%
+%\than{Should we move the code to template section?}
 
@@
 
-Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template for BST, annotated using separation logic.
-The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{p}$.
+Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template, annotated using separation logic.
+The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{r}$.
 We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{proof_lock_insert}).
@@
 
-\begin{figure}[!ht]
-	\begin{subfigure}[t]{0.45\textwidth}
+%\begin{figure}[!ht]
+%	\begin{subfigure}[t]{0.45\textwidth}
+%		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
+%		\caption{The \lstinline{traverse} method of the give-up template algorithm written in an ML-like language}
+%		\label{traverse_giveup_a}	
+%	\end{subfigure}\qquad
+%	\begin{subfigure}[t]{0.48\textwidth}
+%		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
+%		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
+%		\label{traverse_giveup_b}
+%	\end{subfigure}
+%	\caption{The \lstinline{traverse} method of the give-up template algorithms}
+%	\label{traverse_giveup}
+%\end{figure}
+
+\begin{figure}[h]
+	\begin{subfigure}[t]{0.48\textwidth}
+		\begin{subfigure}[t]{\textwidth}
 			\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
@@
 			\label{traverse_giveup_a}	
+		\end{subfigure}
+		\\ \\ 
+		\begin{subfigure}[t]{\textwidth}
+			\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
+			\caption{The \lstinline{insert} method of the give-up template algorithm written in C}
+			\label{insert_giveup}	
+		\end{subfigure}
 	\end{subfigure}\qquad
@@
 	\end{subfigure}
-	\caption{The \lstinline{traverse} method of the give-up template algorithms}
+	\caption{The \lstinline{traverse} and \lstinline{insert} methods of the give-up template algorithms}
 	\label{traverse_giveup}
@@
 
+\than{I moved the code of \texttt{insert} in give-up template here}
+
+The give-up version of the \texttt{insert} method (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version.
+The one difference is that it does not require acquiring a lock before calling its \texttt{traverse} function.
+
 \subsubsection{Give-up Proof}
@@
 \than{I moved \texttt{insert} from Section 4 to here.
-William: Could you please write something here to have a smooth transition from \texttt{traverse} to \texttt{insert}, or do you think you need to have subsections for \texttt{traverse} and \texttt{insert}?
-}
-
-The give-up version of the \texttt{insert} method (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version.
-The one difference is that it does not require acquiring a lock before calling its \texttt{traverse} function.
+William, could you please write something to ensure a smooth transition between \texttt{traverse} and \texttt{insert}, or do you think it would be better to have subsections for \texttt{traverse} and \texttt{insert}?
+}
+
  The proof of the \lstinline{insert} function is similar to that of the previous function: the only difference is that the give-up version of \lstinline{insert} does not acquire a lock before calling \lstinline{traverse}, reflecting the difference in the locking patterns and the preconditions of \lstinline{traverse} between the two templates.
-In Appendix \ref{sec:apd_proof}, Figure \ref{proof_giveup_insert} presents an annotation that uses separation logic for the \texttt{insert} function within the give-up template for BST.
+In Appendix \ref{sec:apd_proof}, Figure \ref{proof_giveup_insert} presents an annotation that uses separation logic for the \texttt{insert} function within the give-up template.
 
@@
 
-The give-up BST uses the same definition and specification of \texttt{findNext}, while introducing an extra helper function called \texttt{inRange} \wm{If \lstinline{inRange} is part of the template, I think we should discuss it in the previous section}.
-The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.
-
 \begin{figure}[h]
@@
 \end{figure}
-
+The give-up BST uses the same definition and specification of \texttt{findNext}, while introducing an extra helper function called \texttt{inRange}.
+%\wm{If \lstinline{inRange} is part of the template, I think we should discuss it in the previous section}.
+%The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.
 The \texttt{inRange} function (Figure \ref{fig:inRange_giveup}) simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
@@
 \label{BST_lock_insert}
-\than{What do we need to say here? (1) - still keep subsection and filled \texttt{insertOp} and the way we prove \texttt{insert} here, or (2) - get rid of this subsection.}
+\than{What do we need to say here?
+Which option should we choose here?
+(1) - Maintain the subsection and fill in \texttt{insertOp}, demonstrating how we prove \texttt{insert}, or (2) - Remove this subsection altogether.}
 \subsection{Give-up BST}
 \label{BST_giveup_insert}
-\than{What do we need to say here? (1) - still keep subsection and filled \texttt{insertOp}, and \texttt{inRange} and the way we prove \texttt{insert} here, or (2) - get rid of this subsection.}
+\than{What do we need to say here?
+Which option should we choose here?
+(1) - Maintain the subsection and fill in \texttt{insertOp} and \texttt{inRange}, demonstrating how we prove \texttt{insert}, or (2) - Remove this subsection altogether.}
 
@@
 Otherwise, the construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, necessitated a considerable amount of time and effort.
-However, we expect that functions such as \texttt{traverse}, \texttt{inRange}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.
+However, we expect that functions such as \texttt{traverse}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.
 The methods \texttt{findNext} and \texttt{insertOp} may also be adapted to other data structures, such as linked lists, with some additional effort.
-Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext} and \texttt{insertOp} to restructure the \texttt{node} struct.
+Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext}, \texttt{inRange}, and \texttt{insertOp} to restructure the \texttt{node} struct.
 While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future.

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
working on Lennart's comments

@@
 
-\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10011074.10011099.10011692</concept_id>
+<concept_desc>Software and its engineering~Formal software verification</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Formal software verification}
 
@@
 
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
+\funding{This work was funded in part by NSF Grant \#1811894.}%optional, to capture a funding statement, which applies to all authors.
 Please enter author specific funding statements as fifth argument of the \author macro.
 
-\acknowledgements{I want to thank \dots}%optional
+%\acknowledgements{I want to thank \dots}%optional
 
@@
 
+In this paper, we apply search structure templates to the problem of verifying C implementations of concurrent search structures.
 The template approach was originally implemented on top of flow interfaces~\cite{krishna2017flow}, a framework for specifying and verifying graph-style data structures, in a combination of two verifiers: the templates were verified in the interactive Iris prover~\cite{iris}, while the data structure implementations were verified with the automated GRASShopper tool~\cite{grasshopper}.
 The target data structures were written in HeapLang, a simple functional programming language with shared-memory concurrency.
-In this paper, we reimplement the approach in the Verified Software Toolchain (VST)~\cite{plcc}, an interactive system for proving correctness of C programs based on a detailed semantics of the C language.
-This allows us to apply the template approach to concurrent data structure implementations in C.
-The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}.
-Our proofs make use of recent work integrating Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
+We reimplement the approach in the Verified Software Toolchain (VST)~\cite{plcc}, an interactive system for proving correctness of C programs based on a detailed semantics of the C language, and apply it to an existing data structure implemented in C.
+The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}, and our proofs make use of recent work integrating Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
 %clean up the phrasing a little
@@
 \subsection*{Related Work}
-Recent years have seen major advances in concurrent separation logics for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}.
+Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}.
 The innovations of Iris (custom ghost state and invariants) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach.
@@
 %how much more can we say about this?
+As a side note, the example templates verified by Krishna et al.~and the ones we present here focus on synchronization via locks, but the approach also applies to lock-free synchronization (as long as it is sequentially consistent; weak-memory template reasoning has not yet been investigated).
 
@@
 The concurrency functions in the template and the sequential \lstinline{findNext} and \lstinline{insertOp} functions provided by the data structure combine into a fully operational concurrent data structure.
-Krishna et al.~specify templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures, but many common search structures are not graphs; in this paper, we present a version of templates that is independent of flow interfaces.
+Krishna et al.~specify templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures (i.e., those in which a node may be reachable along multiple paths), but many common search structures do not require flow-style reasoning; in this paper, we present a version of templates that is independent of flow interfaces.
 %is this a contribution in itself?
@@
 
-Logical atomicity is a concurrent separation logic (CSL) technique for concisely specifying the behavior of a concurrent operation.
+Logical atomicity is a separation logic technique for concisely specifying the behavior of a concurrent operation.
 A logically atomic triple has the form $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
@@
 More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).
-For instance, the specification
+
+As an example, a sequential stack \lstinline{push} operation could conventionally be specified as $\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$.
+Its concurrent counterpart could be specified as
 $\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$
-expresses the fact that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution.
+indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution.
 The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion.
 The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context.
+Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.
 

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
Lennart's comments, second half

@@
 
-Our implementation of \lstinline{traverse} translates the functional implementation into more idiomatic C code.
+Our implementation of \lstinline{traverse} translates the functional implementation into idiomatic C code.
 The lock-coupling pattern can be seen on lines 14 and 15, where \lstinline{traverse} acquires the next node's lock and then releases the current node's lock.
@@
 %\wm{Does this organization work, or would it be better to prove each template as we introduce it?}
-In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification.
-The logically atomic specification for the lock-coupling template's \lstinline{traverse} function is shown in Figure \ref{fig:traverse_lock}.
+We verify the template by proving that \lstinline{traverse} function meets a logically atomic specification, shown in Figure \ref{fig:traverse_lock}.
 \wm{Note that this isn't the same as give-up, and that reflects the original paper.
@@
 The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template.
-The local preconditions include any locks, and consist only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
+The local precondition does not hold any locks, consisting only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
 As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$.
@@
 
-We instantiate the template with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}).
+We instantiate the templates with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}).
 Their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples, and their proofs do not require any concurrency reasoning.

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
cleaning up traverse proof outline

@@
 \subsection{Iris and VST}
-The search structure template approach uses \emph{concurrent separation logic} to specify and prove pre- and postconditions for the template and data structure functions.
-Krishna et al.~implemented their framework in Iris~\cite{iris}, a language-independent separation logic framework built in the Coq proof assistant~\cite{coq}, with flexible support for ghost state, invariants, and atomic specifications.
+The search structure template approach uses concurrent separation logic to specify and prove pre- and postconditions for the template and data structure functions.
+Krishna et al.~implemented their framework in Iris~\cite{iris}, a language-independent CSL framework built in the Coq proof assistant~\cite{coq}, with flexible support for ghost state, invariants, and atomic specifications.
 This makes it easy to describe the effects of concurrency control functions independently of the underlying data structure, e.g., as ``this function atomically finds a node that contains the target key.'' %forward reference to where we show the specs?
@@
 
-\subsection{Defining the Templates}
+\subsection{Defining a Template}
 \label{templates}
@@
 The core of each search structure template is a \lstinline{traverse} function that navigates a data structure using the target synchronization mechanism.
-\wm{\lstinline{traverse} is the core, but it's not the whole template!
-The data structure never sees traverse, but rather a wrapper around it that also acquires locks, etc.
-as necessary.} Templates abstract away from the data structure being traversed, operating on a generic \lstinline{node} type and taking data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters.
+A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well.
+The defining feature of all these functions is that they abstract away from the data structure being traversed: they operate on a generic \lstinline{node} type and take data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters.
 We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
@@
 There are two things worth noting about this data structure.
-First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding as few locks as possible to allow other parts of the data structure to be modified concurrently.
+First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding locks on as few nodes as possible to allow other parts of the data structure to be modified concurrently.
 Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as
@@
 This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component.
-Now, we use \lstinline{node_t} to define \lstinline{traverse} functions for specific concurrency patterns.
+Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns.
 
@@
 	\end{subfigure}
-	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
+	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms \wm{I like the layout here, but it would make more sense if \lstinline{traverse} was b and \lstinline{insert} was c}}
 	\label{traverse_lock}
@@
 Our implementation of \lstinline{traverse} translates the functional implementation into idiomatic C code.
-The lock-coupling pattern can be seen on lines 14 and 15, where \lstinline{traverse} acquires the next node's lock and then releases the current node's lock.
+The lock-coupling pattern can be seen on lines 12-13, where \lstinline{traverse} acquires the next node's lock and then releases the current node's lock.
 The function stops when it reaches an empty leaf node (\lstinline{pn->p->t = NULL}), or when \lstinline{findNext} returns 0, indicating that the target key \lstinline{k} is in the current node.
-The \lstinline{traverse} function returns an integer flag that is 1 when we reach a leaf node and 0 when we find \lstinline{k} in an existing node; the behavior of operations that call \lstinline{traverse} can vary depending on this flag.
-For instance, an \lstinline{insert} operation may create a new node when the flag is 1 and modify an existing node when the flag is 0, while a \lstinline{lookup} operation may fail when the flag is 1 and return the value in the current node when the flag is 0.
+The \lstinline{traverse} function returns 1 when we reach a leaf node and 0 when we find \lstinline{k} in an existing node; the behavior of operations that call \lstinline{traverse} can vary depending on this return value.
+For instance, an \lstinline{insert} operation may create a new node when \lstinline{traverse} returns 1 and modify an existing node when \lstinline{traverse} returns 0, while a \lstinline{lookup} operation may fail on 1 and return the value in the current node on 0.
 
@@
 
-\than{I moved \texttt{insert} code of lock-coupling here}
-
 Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling template.
-%\wm{Actually, is this \lstinline{insert} function part of the template?
-If so, should we discuss it in section 3?}
-It makes use of the \texttt{traverse} function %of Section \ref{lock-coupling-algo} 
-to find the node at which to insert the key \lstinline{k}.
+It uses \texttt{traverse} to find the node at which to insert the key \lstinline{k}.
 If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}).
-Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the data-structure-specific \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
-\than{...
-that we will discuss in Section Verified DS.}
-
-\subsubsection{Lock-coupling Proof}
+Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the underlying data structure's \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
+
+\subsubsection{Verifying the Lock-Coupling Template}
 \label{traverse_proof_lock}
-%\wm{Does this organization work, or would it be better to prove each template as we introduce it?}
+
+\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
 We verify the template by proving that \lstinline{traverse} function meets a logically atomic specification, shown in Figure \ref{fig:traverse_lock}.
-\wm{Note that this isn't the same as give-up, and that reflects the original paper.
-It's the cssOp/insert function that has the same spec in both.}
 %\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}
@@
 	\end{mathpar}
-	\caption{Specification of \texttt{traverse} for the lock-coupling template}
+	\caption{Specification of \texttt{traverse} for the lock-coupling template\wm{Not sure how we want to write this, but when $\mathit{res}$ is false, \lstinline{k} should be the key of \lstinline{n'}, right?}}
 	\label{fig:traverse_lock}
@@
 These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node.
-In formal terms, the resource is represented as $\mathsf{R}(\texttt{n}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{DataStruct\_specific(\texttt{n})}$.
+Formally, the lock invariant is defined by
+\begin{align*}
+&\mathsf{node\_contents}(\texttt{n}, \mathsf{k}, \mathsf{v}, \mathsf{range}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{node\_data}(\texttt{n}, \mathsf{k}, \mathsf{v})\\
+&\mathsf{R}(\texttt{n}) \triangleq \exists\ \mathsf{k}, \mathsf{v}, \mathsf{range}.\ \mathsf{node\_contents}(\texttt{n}, \mathsf{k}, \mathsf{v}, \mathsf{range})\end{align*}
+where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.
 
@@
 
-In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{n})$ and the contents $\mathsf{R}(\texttt{n})$.
+The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(\texttt{n})$ and its contents $\mathsf{R}(\texttt{n})$.
 The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}.
@@
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'})\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'}) \ \ast \ \treerep\ m
-\end{mathpar} %\wm{Where does \lstinline{range} come from here?}
+\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p}, \texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'}.\ &\texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range'}\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \\  &\mathsf{node\_contents}(\texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'})
+\end{align*}
 
@@
 
-Figure \ref{proof_lock_traverse} shows the \texttt{traverse} function in lock-coupling template with separation logic annotations.
-Given this invariant, the proof proceeds as follows.
-%\wm{Should we consider giving annotated code after all?}
+Figure \ref{proof_lock_traverse} shows the proof outline of the \texttt{traverse} function.
 We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$.
-Otherwise, we use \lstinline{findNext} to find a new node \lstinline{n'} to visit (line 9), which is stored in \lstinline{n}.
+Otherwise, we pass the $\mathsf{node\_data}$ from the lock invariant to \lstinline{findNext} (line 9), which returns a new node \lstinline{n''} to visit, stored in \lstinline{pn-> n}.
 If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$.
@@
 %The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
+The two \lstinline{return}s are also the two possible linearization points of the function.
 

### 2023-05-02 concurrency/CONCUR 2023/lock_traverse.c

@@ -1,14 +1,12 @@
 int traverse(pn *pn, int k){
-  int flag = 1;
   for( ; ; ){
     pn->p = pn->n;
     if (pn->p->t == NULL)
-      break;
+      return 1;
     else{
       int b = findNext(pn, k);
       if (b == 0){
-        flag = 0;
-        break;
+        return 0;
       }
       else{
         acquire(pn->n->lock);
@@ -16,5 +14,4 @@ int traverse(pn *pn, int k){
       }
     }
   }
-  return flag;
 }

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_traverse.c

@@ -1,40 +1,29 @@
 int traverse(pn *pn, int k){
-		int flag = 1; 			$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
-				\nodeboxrep(\texttt{n})  \ast \mathsf{R}(\texttt{n})  \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
-			\texttt{pn} \mapsto (\texttt{p, n}) \ast  \nodeboxrep(\texttt{n}) \ast  \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \treerep\ m
-		 \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l}  \texttt{flag} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
+				\nodeboxrep(\texttt{n})  \ast \mathsf{R}(\texttt{n}) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
 				pn->p = pn->n;
-				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n, n}) \ast  \nodeboxrep(\texttt{n}) \ast  \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \treerep\ m\end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ast  \nodeboxrep(\texttt{n'}) \ast  \texttt{k} \in \mathsf{range'} \ast \mathsf{node\_contents}(\mathsf{k'}, \mathsf{v'}, \mathsf{range'})\end{array}\right\}$
 				if (pn->p->t == NULL)
-						break;
[…]

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
finished editing 3.2

@@
 
-\than{I moved \texttt{insert} from Section 4 to here.
-William, could you please write something to ensure a smooth transition between \texttt{traverse} and \texttt{insert}, or do you think it would be better to have subsections for \texttt{traverse} and \texttt{insert}?
-}
-
 %\begin{figure}[h]
@@
 
-The atomic specification for \texttt{insert} method is
+The insert function uses this specification of \lstinline{traverse} to update the state of the data structure. The atomic specification for \texttt{insert} is
 \begin{mathpar}
@@
 \end{mathpar}
-\wm{Consider skipping this paragraph if we need space.} Similar to the \texttt{traverse} specification, the \texttt{insert} function's specification is composed of two components: local and public pre- and postconditions.
-The local pre- and postconditions, containing $\nodeboxrep \ \texttt{r}$, describe the binary search tree's concrete state with the root node $\texttt{r}$, which points to the node's lock and lock invariant assertion.
-On the other hand, the public pre- and postconditions are represented by the predicate $\treerep \ m$, illustrating the binary search tree's abstract state as a collection of ghost states, with each state providing information about an individual node in the tree.
-Here, $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively.
-
-Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template, annotated using separation logic.
-The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{r}$.
-We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{proof_lock_insert}).
-After the lock for the root node is acquired (line 5 in Figure \ref{proof_lock_insert}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$.
-We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
-\begin{mathpar}
-	% spec of inRange
-	{\color{blue}
-		\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
-		\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
-		\right\rangle
-	}
-	\\ 
-	\texttt{traverse(pn, k)} 
-	\\
-	{\color{blue}
-		\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-		\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
-			\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
-		\end{array}
-		\ \Bigg| \ \treerep\ m \
-		\right\rangle
-	}
-\end{mathpar}
-In the $\mathsf{then}$ branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
-we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree.
-In this scenario, we update the node's value while maintaining the rest of the tree structure (line 9 in Figure \ref{proof_lock_insert}).
-Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 14-15 in Figure \ref{proof_lock_insert}).
-We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree.
+where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively.
+The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple.
+We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}.
+If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}).
+Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range.
+We can then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node.
 In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}.
-We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
-We can then release the lock (line 21 in Figure \ref{proof_lock_insert}) return the node's contents to the shared state.
-%Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
-%\wm{Do we ever define $\treerep$?}
+We do this by demonstrating that in both cases, the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
+We can then release the lock held by \lstinline{traverse} and deallocate the \lstinline{pn} structure.
 

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
section 4 edited

@@
 This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component.
-Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns.
+Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns, and then prove (assuming appropriate specifications for data-structure-specific functions) that they correctly implement the data structure operations.
 
@@
 
-\subsubsection{Give-up Proof}
+\subsubsection{Verifying the Give-Up Template}
 \label{traverse_proof_giveup}
@@
 
-\section{Verified Data Structures}
+\section{Verifying Binary Search Trees using Templates}
 \label{BST_proof}
-In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified concurrent data structures in C.
-The target specifications for our search structure operations are shown below: each operation should atomically have its effect on the data structure, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
+In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified C implementations of concurrent data structures.
+The top-level specifications we obtain for search structure operations are shown below: each operation takes effect on the data structure atomically, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
 \begin{mathpar}
@@
 \end{mathpar}
-The intention of the template approach is that we should be able to derive implementations from these operations from \lstinline{traverse} instantiated with an appropriate \lstinline{findNext} function, and prove them correct using the \lstinline{traverse} specs of the previous section, without any further reasoning about specific concurrency patterns.
-
-Our target data structure is a binary search tree (BST) implemented in C, a data structure that was already defined and verified as a demonstration of VST's concurrency capabilities.
-We factor out \lstinline{findNext} and \lstinline{traverse} functions as described above, and verify lock-coupling and give-up versions of the BST using the templates.
-We discuss the verification of \texttt{insert} for each template in detail; the \texttt{lookup} proofs follow similar reasoning.
-As for \texttt{delete}, there is a mismatch between the template approach and BST deletion, which we discuss in Section \ref{internal_reorganize}.
-The complete proofs in VST can be found online at \href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}; we discuss the overall development in Section \ref{proof_mech}.
-
-%\wm{Should we have a separate section where we introduce findNext (and maybe insertOp) before we get into specific templates?}
-
+As we saw in the previous section, these top-level specifications have already been verified at the template level; all we need to do now is provide implementations of data-structure-specific functions (e.g.
+\lstinline{findNext}) that satisfy the right specifications to instantiate the template proofs.
+Once we have chosen a template and a data structure, we should not need to do any further reasoning about specific concurrency patterns.
+
+Our target data structure is a binary search tree (BST) implemented in C.
+A lock-coupling BST was already written verified as a demonstration of VST's concurrency capabilities~\cite{bst-conc}.
+We refactored it into \lstinline{findNext}, \lstinline{traverse}, \lstinline{insertOp}, etc.
+as required by the template approach, and then re-verified it using the template; we then replaced \lstinline{traverse} with the give-up version and verified the resulting give-up BST as well.
+We were able to obtain verified \lstinline{insert} and \lstinline{lookup} operations using this technique; for \texttt{delete}, we found a mismatch between the template approach and BST deletion, which we discuss in Section \ref{internal_reorganize}.
+%The complete proofs in VST can be found online at \href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}; we discuss the overall development in Section \ref{proof_mech}.
+
+\subsection{Instantiating \lstinline{findNext} and \lstinline{insertOp}}
 We instantiate the templates with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}).
-Their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples, and their proofs do not require any concurrency reasoning.
+As expected, their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples (see Figures \ref{fig:findNext_lock} and \ref{fig:insertOp}), and their proofs do not require any concurrency reasoning.
 
@@
 %\wm{Should we recap the definition of the \lstinline{node} struct here?} 
-Let's recall the \texttt{node} struct for a binary search tree. 
+The contents of a BST node are defined in C as:
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
@@
 \end{lstlisting}
-The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with the current node's key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
-The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the current node's key \texttt{k'}, and the next node \lstinline{n'} is the left child \lstinline{l}; \lstinline{k} is greater than \lstinline{k'} and \lstinline{n'} is the right child \lstinline{r}; or \lstinline{k} is exactly \lstinline{k'} and \lstinline{n'} is the original node \lstinline{n}, in which case the return value is 0.
-
+The BST's \lstinline{findNext} function indicates whether the target key is in the current node, its left subtree, or its right subtree.
+The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
+The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the node's key \texttt{k'}, and the next node \lstinline{n'} is the left child \lstinline{l}; \lstinline{k} is greater than \lstinline{k'} and \lstinline{n'} is the right child \lstinline{r}; or \lstinline{k} is exactly \lstinline{k'} and \lstinline{n'} is the original node \lstinline{n}, in which case the return value is 0.
+The proof is straightforward.
+\wm{We should show the actual code for \lstinline{findNext} here!}
+
+The BST's \lstinline{insertOp} function allocates a new \lstinline{node} struct at an empty leaf, as well as two new empty leaves to serve as its children.
 The \texttt{insertOp} specification (Figure \ref{fig:insertOp_lock}) takes as input a node whose \lstinline{t} field is \lstinline{NULL}, i.e., an empty leaf node.
@@
 
-The \texttt{insertOp} specification (Figure \ref{fig:insertOp_giveup}) differs slightly from the lock-coupling version, to account for the added \lstinline{min} and \lstinline{max} fields.
+Unfortunately, while this specification suffices for the lock-coupling template, it cannot be used as is for the give-up template: because we allocate new empty leaf nodes, we must allocate \lstinline{node_t} structures, whose implementation in the give-up template has extra fields that do not appear in the lock-coupling template.
+As shown in Figure \ref{fig:insertOp_giveup}, a give-up template node (even an empty one!) has additional \lstinline{min} and \lstinline{max} fields, and these fields are needed to compute the bounds of the new child nodes.
 In a BST, if the current node's range is \lstinline{(min, max)} and its key is \lstinline{k}, its left child will have range \texttt{(min, k)} and its right child will have range \texttt{(k, max)}.
-Otherwise, \lstinline{insertOp} behaves identically, storing \lstinline{k} and \lstinline{v} in the current node and creating two empty child nodes.
+It may be possible to remedy this discrepancy by adding an ``allocate empty node'' function to the templates, which is then called by \lstinline{insertOp}, but this would still complicate the approach: originally, we expected that templates would be parameterized by data structure implementations but not vice versa.
+However, this two-way dependency seems fundamental when the layout of the node structure depends on the concurrency control mechanism we choose.
 
@@
 The \texttt{inRange} function (Figure \ref{fig:inRange_giveup}) simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
-
-\subsection{Lock-coupling BST}
-\label{BST_lock_insert}
-\than{What do we need to say here?
-Which option should we choose here?
-(1) - Maintain the subsection and fill in \texttt{insertOp}, demonstrating how we prove \texttt{insert}, or (2) - Remove this subsection altogether.}
-\subsection{Give-up BST}
-\label{BST_giveup_insert}
-\than{What do we need to say here?
-Which option should we choose here?
-(1) - Maintain the subsection and fill in \texttt{insertOp} and \texttt{inRange}, demonstrating how we prove \texttt{insert}, or (2) - Remove this subsection altogether.}
+\wm{This should be moved to the template section or removed.}
 
@@
 \label{internal_reorganize}
-The third BST operation, \lstinline{delete}, presents problems in applying the template approach.
+The \lstinline{delete} operation on binary search trees presents difficulties in applying the template approach.
 The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}).
@@
 In practice, Krishna et al.'s proofs for tree structures~\cite{templates} either do not include deletion at all, or use a very lightweight form of deletion in which nodes are marked as ``deleted'' but are never actually removed from the data structure.
+\wm{compare what they say and do for split in a B+-tree}
 
@@
 
-Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process.
+The total lines of code required were about the same as for the original VST verification of the BST.
+The vast majority of the proof effort was in the templates, in both \lstinline{traverse} and the top-level data structure operations.
+This is fairly encouraging, since this proof effort should be reusable for other data structures that use the same templates.
+%Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process.
 Otherwise, the construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, necessitated a considerable amount of time and effort.
-However, we expect that functions such as \texttt{traverse}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.
+%However, we expect that functions such as \texttt{traverse}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.
 The methods \texttt{findNext} and \texttt{insertOp} may also be adapted to other data structures, such as linked lists, with some additional effort.
-Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext}, \texttt{inRange}, and \texttt{insertOp} to restructure the \texttt{node} struct.
+Proving the correctness of e.g.
+a lock-coupling linked list should be as simple as defining a new \lstinline{node} struct, and then implementing and verifying \lstinline{findNext} and \lstinline{insertOp}.
+%Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext}, \texttt{inRange}, and \texttt{insertOp} to restructure the \texttt{node} struct.
 While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future.
-The full development of our mechanization effort is available online
+The full development of our mechanization effort is available online at
 \href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
quick appendix edit

@@
 
-\section{Proofs for Sections \ref{traverse_proof_giveup}, \ref{BST_lock_insert}, and \ref{BST_giveup_insert}}
+\section{Proof Outlines for Templates (Sections \ref{traverse_proof_lock} and \ref{traverse_proof_giveup})}
 \label{sec:apd_proof}
-Below are partially completed proofs of \texttt{traverse} for the give-up template (Figure \ref{proof_giveup_traverse}), as well as proofs of \texttt{insert} for the lock-coupling (Figure \ref{proof_lock_insert}) and give-up (Figure \ref{proof_giveup_insert}) templates, annotated with separation logic specifications.
-The complete proofs can be found in our Coq mechanization, which is mentioned in Section \ref{proof_mech} for reference.
+Below are proof outlines for the \texttt{traverse} function for the give-up template (Figure \ref{proof_giveup_traverse}) as well as the \texttt{insert} function for both templates.
+The complete proofs can be found in our Coq mechanization (see Section \ref{proof_mech}).
+\wm{I should edit these if I have time.}
 

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
edited give-up template

@@
 This means that a conflicting operation may invalidate our traversal, for instance by moving the next node to another part of the data structure before we acquire its lock.
-To guard against this, the give-up template maintains a \emph{range} field on each node indicating the range of keys that should be stored in this node and its successors, and \lstinline{traverse} must check the range field at each node to ensure that it is still on the right path.
+To guard against this, the \lstinline{traverse} function must explicitly check whether the target key is in the range of the current node to ensure that it is still on the right path.
 If a check fails, we give up and start the traversal over from the root node.
@@
 In addition to \lstinline{findNext}, the \lstinline{traverse} function uses a helper function called \lstinline{inRange}, which determines whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors.
+Logically, this range is the same as the range in the lock-coupling template, but that range was a ghost-state construct that only appeared in the proofs; in the give-up template, the range must be stored in memory and checked in the code.
 If \lstinline{k} is outside the node's range (e.g.
@@
 If the \lstinline{inRange} call fails, we release our current lock and return to the root node by setting the current node to the root pointer \lstinline{p} (lines 21-22 in Figure \ref{traverse_giveup_b}).
+The \texttt{insert} operation (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version, with the one difference that it does not need to acquire a lock before calling \texttt{traverse}.
 
@@
 We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it.
-\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
-
-\than{I moved the code of \texttt{insert} in give-up template here}
-
-The give-up version of the \texttt{insert} method (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version.
-The one difference is that it does not require acquiring a lock before calling its \texttt{traverse} function.
+%\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
+The specification of \lstinline{inRange} is shown in Figure \ref{fig:inRange_giveup}; it simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
+
+\begin{figure}[h]
+	\centering
+	\begin{mathpar}
+		% inRange for give-up
+		{\color{blue}
+			\left\{ 
+			\begin{array}{c}
+				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} 
+			\end{array}
+			\right\}
+		}
+		\\ 
+		\texttt{inRange(pn, k)} 
+		\\
+		{\color{blue}
+			\left\{\mathit{res.} \
+			\begin{array}{l}
+				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} \  \ast  \ 
+				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{min} < \texttt{k} < \texttt{max})\ \mathsf{else}\ (\texttt{k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max}))
+			\end{array}
+			\right\}
+		}
+	\end{mathpar}
+	\caption{Specification of \texttt{inRange} for the give-up template}
+	\label{fig:inRange_giveup}
+\end{figure}
 
@@
 
-\than{I moved \texttt{insert} from Section 4 to here.
-William, could you please write something to ensure a smooth transition between \texttt{traverse} and \texttt{insert}, or do you think it would be better to have subsections for \texttt{traverse} and \texttt{insert}?
-}
-
- The proof of the \lstinline{insert} function is similar to that of the previous function: the only difference is that the give-up version of \lstinline{insert} does not acquire a lock before calling \lstinline{traverse}, reflecting the difference in the locking patterns and the preconditions of \lstinline{traverse} between the two templates.
-In Appendix \ref{sec:apd_proof}, Figure \ref{proof_giveup_insert} presents an annotation that uses separation logic for the \texttt{insert} function within the give-up template.
+The specification of the \lstinline{insert} function is the same as for the lock-coupling template, and its proof is quite similar as well.
+The fact that the give-up version of \lstinline{insert} does not acquire a lock before calling \lstinline{traverse} is reflected in the difference in the precondition of \lstinline{traverse} between the two templates.
+The proof outlines for give-up template functions are shown in Appendix \ref{sec:apd_proof}.
 
@@
 However, this two-way dependency seems fundamental when the layout of the node structure depends on the concurrency control mechanism we choose.
-
-\begin{figure}[h]
-	\centering
-	\begin{mathpar}
-		% inRange for give-up
-		{\color{blue}
-			\left\{ 
-			\begin{array}{c}
-				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} 
-			\end{array}
-			\right\}
-		}
-		\\ 
-		\texttt{inRange(pn, k)} 
-		\\
-		{\color{blue}
-			\left\{\mathit{res.} \
-			\begin{array}{l}
-				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} \  \ast  \ 
-				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{min} < \texttt{k} < \texttt{max})\ \mathsf{else}\ (\texttt{k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max}))
-			\end{array}
-			\right\}
-		}
-	\end{mathpar}
-	\caption{Specifications of \texttt{inRange} for the give-up template}
-	\label{fig:inRange_giveup}
-\end{figure}
-The give-up BST uses the same definition and specification of \texttt{findNext}, while introducing an extra helper function called \texttt{inRange}.
-%\wm{If \lstinline{inRange} is part of the template, I think we should discuss it in the previous section}.
-%The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.
-The \texttt{inRange} function (Figure \ref{fig:inRange_giveup}) simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
-\wm{This should be moved to the template section or removed.}
 

### 2023-05-02 concurrency/CONCUR 2023/giveup_traverse.c

@@ -1,17 +1,15 @@
 int traverse(pn *pn, int k){
   node_t *r = (pn->n); 
-  int flag = 1;
   for( ; ; ){
     acquire(pn->n->lock); 
     pn->p = pn->n;
     if (inRange(pn, k) == 1){
       if (pn->p->t == NULL)
-        break;
+        return 1;
       else{
         int b = findNext(pn, k);
         if (b == 0){
-          flag = 0;
-          break;
+          return 0;
         }
         else
           release(pn->p->lock);
@@ -22,5 +20,4 @@ int traverse(pn *pn, int k){
       pn->n = r;
     }
   }
-  return flag;
 }

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
edited giveup traverse proof

@@
 The local precondition does not hold any locks, consisting only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
-As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$.
+\wm{Should we unify $\infp$ and $\nodeboxrep$?} As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$.
 Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{n'->t} is \texttt{NULL} or not.

### 2023-05-02 concurrency/CONCUR 2023/proof_giveup_traverse.c

@@ -1,45 +1,39 @@
 int traverse(pn *pn, int k){
-		node_t *r = (pn->n);
-		int flag = 1; 			$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
-		\infp (\texttt{n})  \ast \infp (\texttt{r}) \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+		node_t *r = (pn->n); $\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  
+\infp (\texttt{r}) \end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
 		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
-			\texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \infp (\texttt{r}) \ast \treerep\ m
+			\exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{\_, n'}) \ast \infp (\texttt{n'}) \ast \infp (\texttt{r})
 		 \end{array}\right\}$
 				acquire(pn->n->lock);
-				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
-				pn->p = pn->n; $\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_, n'}) \ast \infp (\texttt{n'}) \ast \mathsf{R}(\texttt{n'}) \ast \infp (\texttt{r}) \end{array}\right\}$
[…]

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
fixed caption label in subfigs
Fix a typo in orcid.

@@
 
-\author{Shengyi Wang}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~shengyiw} }{shengyiw@cs.princeton.edu}{https://orcid.org/[0000-0002-2286-8703}{}
+\author{Shengyi Wang}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~shengyiw} }{shengyiw@cs.princeton.edu}{https://orcid.org/0000-0002-2286-8703}{}
 
@@
 		\\ \\ 
+		\renewcommand{\thesubfigure}{c}% New fixed/manual numbering
 		\begin{subfigure}[t]{\textwidth}
@@
 	\end{subfigure}\qquad
+	\renewcommand{\thesubfigure}{b}% New fixed/manual numbering
 	\begin{subfigure}[t]{0.48\textwidth}
@@
 	\end{subfigure}
-	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms \wm{I like the layout here, but it would make more sense if \lstinline{traverse} was b and \lstinline{insert} was c}}
+	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms 
+		%\wm{I like the layout here, but it would make more sense if \lstinline{traverse} was b and \lstinline{insert} was c}
+	}
 	\label{traverse_lock}

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
add findnext code

@@
 \end{lstlisting}
-The BST's \lstinline{findNext} function indicates whether the target key is in the current node, its left subtree, or its right subtree.
+The BST's \lstinline{findNext} function (Figure \ref{findnext}) indicates whether the target key is in the current node, its left subtree, or its right subtree.
 The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
@@
 The proof is straightforward.
-\wm{We should show the actual code for \lstinline{findNext} here!}
+%\wm{We should show the actual code for \lstinline{findNext} here!}
+
+\begin{figure}[ht]
+	\lstinputlisting[language=C, style=myStyle]{findnext.c} 
+	\caption{The \lstinline{findNext} method for binary search tree (all templates) written in C}
+	\label{findnext}
+\end{figure}
 

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
fixed line error

@@
 
-As an example, a sequential stack \lstinline{push} operation could conventionally be specified as $\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$.
+As an example, a sequential stack \lstinline{push} operation could conventionally be specified as \\$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$.
 Its concurrent counterpart could be specified as
-$\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$
+$$\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$$
 indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution.

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
unify infp and node_ref

@@
 			\begin{array}{c}
-				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
+				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \nodeboxrep(\texttt{n}) \ \big| \ \treerep\ m
 			\end{array}
@@
 				\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
-				\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
+				\\ \nodeboxrep(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
 			\end{array}
@@
 The local precondition does not hold any locks, consisting only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
-\wm{Should we unify $\infp$ and $\nodeboxrep$?} As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$.
+It is written formally as
+$\nodeboxrep(\texttt{n}) \triangleq\ \infp(\texttt{n})$.
+%\wm{Should we unify $\infp$ and $\nodeboxrep$?} 
+As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$.
 Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{n'->t} is \texttt{NULL} or not.
@@
 	\begin{array}{c}
-		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
+		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \nodeboxrep(\texttt{n}) \ \big| \ \treerep\ m
 	\end{array}
@@
 		\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
-		\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
+		\\ \nodeboxrep(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
 	\end{array}

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_insert.c
lock-coupling insert proof

@@ -1,25 +1,25 @@
 void insert (node_t **r, int k, void *v){
 		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
 		pn->n = *r;
-		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{p, r}) \ast  \texttt{r->lock} \lockvar \mathsf{R}(\texttt{r}) \ast \treerep\ m \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \nodeboxrep\ \texttt{r} \end{array}\right\}$
 		acquire(pn->n->lock);
-		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p, r}) \ast  \texttt{r->lock} \lockvar \mathsf{R}(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \ast \cdots \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \nodeboxrep\ \texttt{r} \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-		$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL} \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep\ \texttt{n'} \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \mathsf{v'}, \mathsf{range'}) \end{array}\right\}$
 				pn->p->t->value = v;
[…]

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
edit giveup proofs

@@
 	\right\rangle$
-	\caption{The \texttt{traverse} function in lock-coupling template annotated with separation logic specification}
+	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
 	\label{proof_lock_traverse}
@@
 
+\wm{Should we get rid of inFP and just talk about Node\_ref?}
 The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template.
@@
 %\wm{Should we unify $\infp$ and $\nodeboxrep$?} 
-As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$.
+As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$ \wm{or node\_contents?}.
 Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{n'->t} is \texttt{NULL} or not.
@@
 For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held.
-Formally, we write $$\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists \ b.
+%Formally, we write $$\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists \ b.
 \ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}$$
@@
 On the other hand, when the lock is not held ($\ell \mapsto \texttt{false}$), resources \texttt{R} are required to be part of the invariant.
-and $\treerep$ contains an $\mathsf{inv\_for\_lock}$ assertion for each node in the data structure.
+%and $\treerep$ contains an $\mathsf{inv\_for\_lock}$ assertion for each node in the data structure.
 As before, $\mathsf{R}$ contains the concrete contents of each node, but also includes the \texttt{min} and \texttt{max} fields that we use to implement the \texttt{inRange} check.
@@
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{p q n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \ \infp (\texttt{q}) \ast \infp (\texttt{n'})  \ \ast \ \treerep\ m
+\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{p r n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \ \infp (\texttt{r}) \ast \infp (\texttt{n'})
 \end{mathpar}
-The \texttt{traverse} function, as illustrated within the context of the give-up template and annotated using separation logic, can be found in Figure \ref{proof_giveup_traverse} of Appendix \ref{sec:apd_proof}.
+Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}).
+Instead, we keep a reference $\infp(\texttt{r})$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.
+
 With this invariant in hand, the proof of \texttt{traverse} proceeds as follows.
-We start with the $\infp (\texttt{n})$ predicate from the local precondition.
-After acquiring the lock of the node $\texttt{n}$ (line 5 in Figure \ref{proof_giveup_traverse} in Appendix \ref{sec:apd_proof}), we obtain the resource $\mathsf{R}(\texttt{n})$, which allows us to perform the \texttt{inRange} check on line 8.
-If the check fails (lines 28-32), we immediately release the lock and start over from the beginning.
-Otherwise, we proceed as before: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release our lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}.
-In this case, we do not hold the lock on the new node, but only the $\infp$ assertion indicating that the new node is in the data structure.
+We begin by using the $\infp$ predicate to acquire the lock on \lstinline{pn->n} (line 4 in Figure \ref{traverse_giveup_b}).
+We then call \lstinline{inRange} to check whether we are still on the path to a node that can hold \lstinline{k}.
+If the check fails (lines 18-20), we immediately release the lock and start over from the root node \lstinline{r}.
+Otherwise, we proceed as in the lock-coupling template: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release the lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}.
+In this case, we once again hold no locks, and have only the $\infp$ assertion indicating that the new node is in the data structure.
 % and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}.
@@
 The fact that the give-up version of \lstinline{insert} does not acquire a lock before calling \lstinline{traverse} is reflected in the difference in the precondition of \lstinline{traverse} between the two templates.
-The proof outlines for give-up template functions are shown in Appendix \ref{sec:apd_proof}.
+Proof outlines for the give-up template functions can be found in Appendix \ref{sec:apd_proof}.
 
@@
 The complete proofs can be found in our Coq mechanization (see Section \ref{proof_mech}).
-\wm{I should edit these if I have time.}
 
@@
 	\right\rangle$
-	\caption{The \texttt{traverse} function in give-up template annotated with separation logic specification}
+	\caption{Proof outline of the give-up \texttt{traverse} function}
 	\label{proof_giveup_traverse}
@@
 		\right\rangle$
-		\caption{The \texttt{insert} function in lock-coupling template annotated with separation logic specification \\\\}
+		\caption{Proof outline of the lock-coupling \texttt{insert} function \\\\}
 		\label{proof_lock_insert}	
@@
 		\right\rangle$
-		\caption{The \texttt{insert} function in give-up template annotated with separation logic specification \\\\}
+		\caption{Proof outline of the give-up \texttt{insert} function \\\\}
 		\label{proof_giveup_insert}
 	\end{subfigure}
-	\caption{The \texttt{insert} function in lock-coupling and give-up templates annotated with separation logic specification}
+	\caption{Proof outlines for the \texttt{insert} function}
 	\label{proof_lock_giveup_insert}

### 2023-05-02 concurrency/CONCUR 2023/proof_giveup_insert.c

@@ -1,23 +1,23 @@
 void insert (node_t **r, int k, void *v){
 		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
 		pn->n = *r;
-		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{p, r}) \ast  \infp(\texttt{r}) \ast \treerep\ m \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \infp(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \infp(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \mathsf{v'}, \mathsf{range'}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \texttt{p'->t->value} \mapsto \texttt{v} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep\ \texttt{n'} \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \texttt{v}, \mathsf{range'}) \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
 				//Linearization point
[…]

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
modified pointers to min and max

@@
 			\begin{array}{c}
-				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} 
+				\texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2}  
 			\end{array}
@@
 			\begin{array}{l}
-				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} \  \ast  \ 
-				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{min} < \texttt{k} < \texttt{max})\ \mathsf{else}\ (\texttt{k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max}))
+				\texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2} \ \ast  \ 
+				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\mathsf{n_1} < \texttt{k} < \mathsf{n_2})\ \mathsf{else}\ (\texttt{k} \leq \mathsf{n_1} \lor \texttt{k} \geq \mathsf{n_2}))
 			\end{array}
@@
 		{\color{blue}
-			\left\{\exists \  \texttt{t l r lk1 lk2}.
-			\begin{array}{c}
+			\left\{\exists \  \texttt{l r lk1 lk2}.
+			\begin{array}{l}
 				\ \texttt{pn->n->t} \mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
@@
 		\begin{array}{c}
-			\texttt{pn->n}\mapsto \texttt{NULL} \  \ast \ \texttt{pn->n}\mapsto \texttt{min} \ \ast \ \texttt{pn->n}\mapsto\texttt{max}
+			\texttt{pn->n}\mapsto \texttt{NULL} \  \ast \ \texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2}
 		\end{array}
@@
 	{\color{blue}
-		\left\{\exists \ \texttt{t l r lk1 lk2}. 
+		\left\{\exists \ \texttt{l r lk1 lk2}. 
 		\begin{array}{c}
-			\texttt{pn->n->t}\mapsto \texttt{(k, v, l, r)} \  \ast \ \texttt{pn->n}\mapsto \texttt{min} \ \ast \ \texttt{pn->n}\mapsto\texttt{max}  \ \ast \ 
-			\\ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, k))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (k, max))}
+			\texttt{pn->n->t}\mapsto \texttt{(k, v, l, r)} \  \ast \ \texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2}  \ \ast \ 
+			\\ \texttt{l} \mapsto (\texttt{NULL}, \texttt{lk1}, (\mathsf{n_1}, \mathsf{k})) \ \ast \ \texttt{r} \mapsto (\texttt{NULL}, \texttt{lk2}, (\mathsf{k}, \mathsf{n_2}))
 		\end{array}

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
modified postcondition of traverse

@@
 			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-			\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
-				\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
+			\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
+				\\ 
+				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
+				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
 			\end{array}
@@
 	\end{mathpar}
-	\caption{Specification of \texttt{traverse} for the lock-coupling template\wm{Not sure how we want to write this, but when $\mathit{res}$ is false, \lstinline{k} should be the key of \lstinline{n'}, right?}}
+	\caption{Specification of \texttt{traverse} for the lock-coupling template%\wm{Not sure how we want to write this, but when $\mathit{res}$ is false, \lstinline{k} should be the key of \lstinline{n'}, right?}
+	}
 	\label{fig:traverse_lock}
@@
 \begin{align*}
-&\mathsf{node\_contents}(\texttt{n}, \mathsf{k}, \mathsf{v}, \mathsf{range}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{node\_data}(\texttt{n}, \mathsf{k}, \mathsf{v})\\
-&\mathsf{R}(\texttt{n}) \triangleq \exists\ \mathsf{k}, \mathsf{v}, \mathsf{range}.\ \mathsf{node\_contents}(\texttt{n}, \mathsf{k}, \mathsf{v}, \mathsf{range})\end{align*}
+	&\mathsf{node\_contents}(\texttt{n}, (\mathsf{k}, \mathsf{v}), \mathsf{range}) \triangleq \mathsf{ghost\_node(\texttt{n}, (\mathsf{k}, \mathsf{v}), range)} \ \ast \ \mathsf{node\_data}(\texttt{n}, \mathsf{k}, \mathsf{v})\\
+	&\mathsf{R}(\texttt{n}) \triangleq \exists\ \mathsf{k}, \mathsf{v}, \mathsf{range}.\ \mathsf{node\_contents}(\texttt{n}, (\mathsf{k}, \mathsf{v}), \mathsf{range})\end{align*}
 where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.
@@
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p}, \texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'}.\ &\texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range'}\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \\  &\mathsf{node\_contents}(\texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'})
+\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p}, \texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'}.\ &\texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range'}\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \\  &\mathsf{node\_contents}(\texttt{n'}, (\mathsf{k'}, \mathsf{v'}), \mathsf{range'})
 \end{align*}
@@
 	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
-		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
+	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
+		\\ 
+		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
 	\end{array}
@@
 			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-			\begin{array}{l}
-				\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
-				\\ \nodeboxrep(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
+			\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
+				\\ 
+				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
+				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
 			\end{array}

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
modified postcondition of traverse
rewrote the deletion problem

--- 2023-05-02 concurrency/CONCUR 2023/paper.tex
@@
 The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}).
-When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure?
+Likewise, in the data structures verified by Krishna et al., deletion can be performed by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation.
+When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
 rotating three-node sections of the tree until the node to be deleted is at a leaf.
 This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them).
-This contrasts with linked-list deletion, which only involves the target node and its immediate neighbors.
-In practice, Krishna et al.'s proofs for tree structures~\cite{templates} either do not include deletion at all, or use a very lightweight form of deletion in which nodes are marked as ``deleted'' but are never actually removed from the data structure.
-\wm{compare what they say and do for split in a B+-tree}
+
+It is instructive to compare BST deletion with Krishna et al.'s discussion of splitting in B-link trees.
+They assume the existence of a maintenance thread that constantly crawls the tree searching for full nodes and splitting them.
+The splitting can be defined as a sequential, data-structure-specific operation on a node and its parent; the crawl itself is a variant of the \lstinline{traverse} operation, and while it is not verified in their existing proofs, we expect that it could be defined as a template function.
+We might similarly imagine refactoring a lock-coupling \lstinline{pushdown_left} into a template-level traversal and a data-structure-level rotation operation.
+However, in the BST, rotation involves three nodes (the node to be deleted and both of its children)---and, crucially, \emph{we only know that it involves three nodes once we know the target data structure is a BST}.
+Thus, there is no clear way to define a template function that acquires enough locks (or otherwise performs enough synchronization) to call out to a sequential rotation operation without making the template specific to BSTs.
 
@@
 Operations in the style of \lstinline{pushdown_left} are ubiquitous in tree structures (e.g., the rebalancing operations of AVL and red-black trees), and always weave together synchronization and restructuring in a way that cannot easily be decomposed.
-This is a serious impediment to the goal of verifying concurrency control separately from the data structure: even if we have a verified template and a verified single-threaded tree, we still need to verify a complex data-structure-specific concurrent function.
-%in fact, usually the most complicated part!
+This is a serious impediment to the goal of verifying concurrency control separately from the data structure: even if we have a verified template and a verified single-threaded tree, we still need to verify a complex data-structure-specific concurrent function to support deletion, which is often the most complicated part of a concurrent search structure.
 %what's the conclusion here? can we see a way through?

### 2023-05-02 concurrency/CONCUR 2023/paper.tex

--- 2023-05-02 concurrency/CONCUR 2023/paper.tex
@@
 	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-	\begin{array}{l}
-		\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
-		\\ \nodeboxrep(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
+	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
+		\\ 
+		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
 	\end{array}

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
explain for node_contents

@@
 where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.
+
+\than{I put an explanation for $\mathsf{node\_contents}(\texttt{n}, \cdot , \mathsf{range})$ here }
+
+The expression $\mathsf{node\_contents}(\texttt{n}, \cdot , \mathsf{range})$  signifies that \texttt{n} points are being directed towards an empty leaf node.
+On the other hand, the expression $\mathsf{node\_contents}(\texttt{n}, (\mathsf{k}, \mathsf{v}), \mathsf{range})$ denotes that \texttt{n} points are directed towards a node, which houses the key $\mathsf{k}$ and its corresponding value $\mathsf{v}$.
 

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
unified inFP and Node_ref

--- 2023-05-02 concurrency/CONCUR 2023/paper.tex
@@
 %\newcommand{\boxdotright}{\!\mathrel\boxdot\joinrel\rightarrow\!}
-\newcommand{\islock}{\boxdotright}
+%\newcommand{\islock}{\boxdotright}
 \newcommand{\lockvar}{\islock}
@@
 \newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
-\newcommand{\infp}{\ensuremath{\mathsf{InFP}}}
+\newcommand{\infp}{\ensuremath{\mathsf{Node\_ref}}}
 
@@
 In particular, holding $\nodeboxrep(\texttt{n})$ should allow us to acquire the lock on node \texttt{n}, which in turn gives us access to the contents of the node.
-Formally, we write
-\[\nodeboxrep(\texttt{n}) \triangleq\ \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n})\]
-where the \emph{lock invariant assertion} $\ell \lockvar \mathsf{R}$ expresses that the lock $\ell$ protects resources $\mathsf{R}$.
-In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).
+Formally, we need to know that the triple \[\left\langle \nodeboxrep(n) \ |\ \treerep\ m \right\rangle\ \texttt{acquire}(n\texttt{->lock})\ \left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle\] holds, i.e., $\nodeboxrep(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant}.
+%In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).
 
@@
 
-The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(\texttt{n})$ and its contents $\mathsf{R}(\texttt{n})$.
+The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(\texttt{n})$ and its contents $\mathsf{R}(\texttt{n})$, indicating that a thread should already hold \lstinline{n}'s lock before calling \lstinline{traverse}.
 The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}.
@@
 \wm{Should we get rid of inFP and just talk about Node\_ref?}
-The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template.
-The local precondition does not hold any locks, consisting only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$.
-It is written formally as
-$\nodeboxrep(\texttt{n}) \triangleq\ \infp(\texttt{n})$.
-%\wm{Should we unify $\infp$ and $\nodeboxrep$?} 
+The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup}, is slightly simpler than that of the lock-coupling template: a thread does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition.
 As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$ \wm{or node\_contents?}.

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_traverse.c

@@ -9,7 +9,7 @@ int traverse(pn *pn, int k){
 				else{
 						int b = findNext(pn, k);
 						$\color{blue} \left\{\begin{array}{l} \exists \  \texttt{n''}. \ \texttt{pn} \mapsto (\texttt{n', n''}) \ast \nodeboxrep(\texttt{n'}) \ast  \texttt{k} \in \mathsf{range'} \ast \mathsf{node\_contents}(\mathsf{k'}, \mathsf{v'}, \mathsf{range'})\ \ast\\ 
-						((\texttt{b} = 0 \ast \mathsf{k'} = \mathsf{k} \ast \texttt{n''} = \texttt{n'})  \lor 						 (\texttt{b} = 1  \ast \texttt{k} \in \mathsf{range}(\texttt{n''}) \ast \texttt{n''->lock} \islock \mathsf{R}(\texttt{n''})))   \end{array}\right\}$
+						((\texttt{b} = 0 \ast \mathsf{k'} = \mathsf{k} \ast \texttt{n''} = \texttt{n'})  \lor 						 (\texttt{b} = 1  \ast \texttt{k} \in \mathsf{range}(\texttt{n''}) \ast \nodeboxrep(\texttt{n''})))   \end{array}\right\}$
 						if (b == 0){
 								$\color{blue} \left\{\begin{array}{l} 
 								\texttt{pn} \mapsto (\texttt{n', n'}) \ast \mathsf{k' = \texttt{k}} \ast \cdots    \end{array}\right\}$
@@ -18,11 +18,11 @@ int traverse(pn *pn, int k){
 								\texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast  \texttt{k} \in \mathsf{range'} \ast \mathsf{node\_contents}(\texttt{k}, \mathsf{v'}, \mathsf{range'})    \end{array}\right\}$
 						}
 						else{
-								$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n''}) \ast \nodeboxrep(\texttt{n'}) \ast  \texttt{k} \in \mathsf{range}(\texttt{n''}) \ast \mathsf{R}(\texttt{n'}) \ast \texttt{n''->lock} \islock \mathsf{R}(\texttt{n''})    \end{array}\right\}$
[…]

### 2023-05-02 concurrency/CONCUR 2023/proof_giveup_insert.c
regularized parens

@@ -5,7 +5,7 @@ void insert (node_t **r, int k, void *v){
 		if (traverse(pn, k) == 0){
 				$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \infp(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \mathsf{v'}, \mathsf{range'}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep\ \texttt{n'} \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \texttt{v}, \mathsf{range'}) \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \texttt{v}, \mathsf{range'}) \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
 				//Linearization point
 		}
 		else{

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_insert.c

@@ -1,25 +1,25 @@
 void insert (node_t **r, int k, void *v){
 		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
 		pn->n = *r;
-		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \nodeboxrep\ \texttt{r} \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \nodeboxrep(\texttt{r}) \end{array}\right\}$
 		acquire(pn->n->lock);
-		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \nodeboxrep\ \texttt{r} \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \nodeboxrep(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-		$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep\ \texttt{n'} \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \mathsf{v'}, \mathsf{range'}) \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \mathsf{v'}, \mathsf{range'}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep\ \texttt{n'} \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \texttt{v}, \mathsf{range'}) \end{array}\right\}$ 
[…]

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
regularized parens
node_contents

--- 2023-05-02 concurrency/CONCUR 2023/paper.tex
@@
 \begin{align*}
-	&\mathsf{node\_contents}(\texttt{n}, (\mathsf{k}, \mathsf{v}), \mathsf{range}) \triangleq \mathsf{ghost\_node(\texttt{n}, (\mathsf{k}, \mathsf{v}), range)} \ \ast \ \mathsf{node\_data}(\texttt{n}, \mathsf{k}, \mathsf{v})\\
-	&\mathsf{R}(\texttt{n}) \triangleq \exists\ \mathsf{k}, \mathsf{v}, \mathsf{range}.\ \mathsf{node\_contents}(\texttt{n}, (\mathsf{k}, \mathsf{v}), \mathsf{range})\end{align*}
+	&\mathsf{node\_contents}(\texttt{n}, \mathsf{c}, \mathsf{range}) \triangleq \mathsf{ghost\_node(\texttt{n}, \mathsf{c}, range)} \ \ast \ \mathsf{node\_data}(\texttt{n}, \mathsf{c})\\
+	&\mathsf{R}(\texttt{n}) \triangleq \exists\ \mathsf{c}, \mathsf{range}.\ \mathsf{node\_contents}(\texttt{n}, \mathsf{c}, \mathsf{range})\end{align*}
 where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.
+The contents $\mathsf{c}$ of a node can be either a key-value pair $(\mathsf{k}, \mathsf{v})$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).
 
@@
 The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}.
-The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not.
-The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its Boolean state.
+The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p}, \texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'}.\ &\texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range'}\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \\  &\mathsf{node\_contents}(\texttt{n'}, (\mathsf{k'}, \mathsf{v'}), \mathsf{range'})
+\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p}, \texttt{n}, \mathsf{c}, \mathsf{range}.\ &\texttt{pn} \mapsto (\texttt{p, n}) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(\texttt{n})  \ \ast \\  &\mathsf{node\_contents}(\texttt{n}, \mathsf{c}, \mathsf{range})
 \end{align*}
@@
 We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$.
-Otherwise, we pass the $\mathsf{node\_data}$ from the lock invariant to \lstinline{findNext} (line 9), which returns a new node \lstinline{n''} to visit, stored in \lstinline{pn-> n}.
+Otherwise, we pass the $\mathsf{node\_data}$ from the lock invariant to \lstinline{findNext} (line 11), which returns a new node \lstinline{n''} to visit, stored in \lstinline{pn-> n}.
 If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$.
@@
 		m.\left\langle 
-		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
 		\right\rangle
@@
 		\left\langle 
-		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
 		\right\rangle
@@
 
-\wm{Should we get rid of inFP and just talk about Node\_ref?}
 The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup}, is slightly simpler than that of the lock-coupling template: a thread does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition.
-As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$ \wm{or node\_contents?}.
-Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{n'->t} is \texttt{NULL} or not.
+As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents.
+Also as before, the Boolean $\mathit{res}$ indicates whether \lstinline{n'} is empty or contains the key \lstinline{k}.
 
 %\wm{We really need to explain why we have this difference, or get rid of it.} In contrast to the lock-coupling style, where each lock is associated with a lock-invariant, in the give-up template we treat locks more abstractly.
-For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held.
+%For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held.
 %Formally, we write $$\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists \ b.
@@
 %and $\treerep$ contains an $\mathsf{inv\_for\_lock}$ assertion for each node in the data structure.
-As before, $\mathsf{R}$ contains the concrete contents of each node, but also includes the \texttt{min} and \texttt{max} fields that we use to implement the \texttt{inRange} check.
+
+%As before, $\mathsf{R}$ contains the concrete contents of each node, but also includes the \texttt{min} and \texttt{max} fields that we use to implement the \texttt{inRange} check.
 
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{p r n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \ \infp (\texttt{r}) \ast \infp (\texttt{n'})
+\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{p}, \texttt{r}, \texttt{n}.\ \texttt{pn} \mapsto (\texttt{p, n}) \ast \ \infp (\texttt{r}) \ast \infp (\texttt{n})
 \end{mathpar}
@@
 		\forall m.\left\langle 
-		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
 		\right\rangle
@@
 		\left\langle \texttt{v}. \ \
-		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
 		\right\rangle
@@
 		\left\langle 
-		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
 		\right\rangle
@@
 		\forall m.\left\langle 
-		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
 		\right\rangle$
@@
 		\left\langle 
-		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
 		\right\rangle$

### 2023-05-02 concurrency/CONCUR 2023/proof_giveup_insert.c

@@ -3,9 +3,9 @@ void insert (node_t **r, int k, void *v){
 		pn->n = *r;
 		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \infp(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-				$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \infp(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \mathsf{v'}, \mathsf{range'}) \end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \infp(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, (\texttt{k}, \mathsf{v'}), \mathsf{range'}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \texttt{v}, \mathsf{range'}) \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, (\texttt{k}, \texttt{v}), \mathsf{range'}) \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
 				//Linearization point
 		}
 		else{
@@ -13,7 +13,7 @@ void insert (node_t **r, int k, void *v){
 				insertOp(pn, k, v);
[…]

### 2023-05-02 concurrency/CONCUR 2023/proof_giveup_traverse.c

@@ -8,22 +8,23 @@ int traverse(pn *pn, int k){
 				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_, n'}) \ast \infp (\texttt{n'}) \ast \mathsf{R}(\texttt{n'}) \ast \infp (\texttt{r}) \end{array}\right\}$
 				pn->p = pn->n; $\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ast \infp (\texttt{n'}) \ast \mathsf{R}(\texttt{n'}) \ast \infp (\texttt{r}) \end{array}\right\}$
 				if (inRange(pn, k) == 1){
-				$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range'} \ast \texttt{pn} \mapsto (\texttt{n', n'}) \ast \infp (\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'}) \ast \infp (\texttt{r}) \end{array}\right\}$
+						$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (\texttt{n', n'}) \ast \infp (\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \mathsf{c}, \mathsf{range}) \ast \infp (\texttt{r}) \end{array}\right\}$
 						if (pn->p->t == NULL)
 								return 1;
-								$\color{violet} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range'} \ast \texttt{pn} \mapsto (\texttt{n', n'}) \ast \texttt{n'->t} = \texttt{NULL} \ast \infp (\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'}) \end{array}\right\}$
+								$\color{violet} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (\texttt{n', n'}) \ast \infp (\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \end{array}\right\}$
 						else{
[…]

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_insert.c

@@ -5,9 +5,9 @@ void insert (node_t **r, int k, void *v){
 		acquire(pn->n->lock);
 		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \nodeboxrep(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-		$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \mathsf{v'}, \mathsf{range'}) \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, (\texttt{k}, \mathsf{v'}), \mathsf{range'}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, \texttt{k}, \texttt{v}, \mathsf{range'}) \end{array}\right\}$ 
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, (\texttt{k}, \texttt{v}), \mathsf{range'}) \end{array}\right\}$ 
 				//Linearization point
 		}
 		else{
@@ -15,7 +15,7 @@ void insert (node_t **r, int k, void *v){
 				insertOp(pn, k, v);
[…]

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_traverse.c

@@ -3,19 +3,21 @@ int traverse(pn *pn, int k){
 				\nodeboxrep(\texttt{n})  \ast \mathsf{R}(\texttt{n}) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
 		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
 				pn->p = pn->n;
-				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ast  \nodeboxrep(\texttt{n'}) \ast  \texttt{k} \in \mathsf{range'} \ast \mathsf{node\_contents}(\mathsf{k'}, \mathsf{v'}, \mathsf{range'})\end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ast  \nodeboxrep(\texttt{n'}) \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(\texttt{n'}, \mathsf{c}, \mathsf{range})\end{array}\right\}$
 				if (pn->p->t == NULL)
-						return 1;						$\color{violet} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ast \texttt{n'->t} = \texttt{NULL}  \ast \nodeboxrep(\texttt{n'}) \ast  \texttt{k} \in \mathsf{range'} \ast \mathsf{node\_contents}(\mathsf{k'}, \mathsf{v'}, \mathsf{range'}) \end{array}\right\}$
+						return 1;						
+						$\color{violet} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \end{array}\right\}$
 				else{
[…]

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
lift William's comment

@@
 
-\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
+%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
 We verify the template by proving that \lstinline{traverse} function meets a logically atomic specification, shown in Figure \ref{fig:traverse_lock}.

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
a few more words about deletion

--- 2023-05-02 concurrency/CONCUR 2023/paper.tex
@@
 This is fairly encouraging, since this proof effort should be reusable for other data structures that use the same templates.
+For lock-coupling deletion, we retained the original (non-templated) proofs; we did not verify give-up deletion, since optimistic deletion is considerably harder to reason about and did not stand to benefit from the template approach.
 %Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process.
@@
 Proving the correctness of e.g.
-a lock-coupling linked list should be as simple as defining a new \lstinline{node} struct, and then implementing and verifying \lstinline{findNext} and \lstinline{insertOp}.
+a lock-coupling linked list should be as simple as defining a new \lstinline{node} struct, and then implementing and verifying \lstinline{findNext} and \lstinline{insertOp} (plus the appropriate functions for deletion).
 %Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext}, \texttt{inRange}, and \texttt{insertOp} to restructure the \texttt{node} struct.

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
spacing

@@
 We reimplement the approach in the Verified Software Toolchain (VST)~\cite{plcc}, an interactive system for proving correctness of C programs based on a detailed semantics of the C language, and apply it to an existing data structure implemented in C.
-The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}, and our proofs make use of recent work integrating Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
+The template approach depends crucially on the idea of \emph{logical atomicity} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}, and our proofs make use of recent work integrating Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.
 %clean up the phrasing a little
@@
 
+We prove correctness of the template by showing that \lstinline{traverse}, \lstinline{insert}, etc.
+meet logically atomic specifications describing their effects on the data structure.
+These specifications are defined in terms of assertions $\nodeboxrep$, representing a client thread's handle to the data structure, and $\treerep$, representing the data structure's abstract state.
+Holding $\nodeboxrep(\texttt{n})$ for a node \texttt{n} should allow us to acquire the lock on \texttt{n}, which in turn gives us access to the contents of the node.
+Formally, we need to know that the triple \[{\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m \right\rangle}\ \texttt{acquire}(n\texttt{->lock})\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}\] holds, i.e., $\nodeboxrep(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant}.
+%In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).
+
 %\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
-We verify the template by proving that \lstinline{traverse} function meets a logically atomic specification, shown in Figure \ref{fig:traverse_lock}.
+The \lstinline{traverse} function can then be specified as follows:
 %\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
-
-%maybe inline this instead of making it a figure?
-\begin{figure}[ht]
-	\centering
 	\begin{mathpar}
@@
 			\right\rangle
-		}
-		\\ 
-		\texttt{traverse(pn, k)} 
-		\\
+		}\\
+		\texttt{traverse(pn, k)} \\
 		{\color{blue}
@@
 	\end{mathpar}
-	\caption{Specification of \texttt{traverse} for the lock-coupling template%\wm{Not sure how we want to write this, but when $\mathit{res}$ is false, \lstinline{k} should be the key of \lstinline{n'}, right?}
-	}
-	\label{fig:traverse_lock}
-\end{figure}
-
-The definitions of $\nodeboxrep$ and $\treerep$ may vary depending on the data structure, but fundamentally, $\nodeboxrep$ holds the resources owned by each client thread, while $\treerep$ holds the abstract representation of the data structure, treated as an atomically accessed shared resource.
-In particular, holding $\nodeboxrep(\texttt{n})$ should allow us to acquire the lock on node \texttt{n}, which in turn gives us access to the contents of the node.
-Formally, we need to know that the triple \[\left\langle \nodeboxrep(n) \ |\ \treerep\ m \right\rangle\ \texttt{acquire}(n\texttt{->lock})\ \left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle\] holds, i.e., $\nodeboxrep(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant}.
-%In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).
-
-Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.
-Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{n}$ and obtain the resources $\mathsf{R}(\texttt{n})$ guarded by the lock.
-These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node.
+The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(\texttt{n})$ and its contents $\mathsf{R}(\texttt{n})$, indicating that a thread should already hold \lstinline{n}'s lock before calling \lstinline{traverse}.
+The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}.
+The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.
+
+The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node.
 Formally, the lock invariant is defined by
@@
 
-The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(\texttt{n})$ and its contents $\mathsf{R}(\texttt{n})$, indicating that a thread should already hold \lstinline{n}'s lock before calling \lstinline{traverse}.
-The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}.
-The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
@@
 
-The insert function uses this specification of \lstinline{traverse} to update the state of the data structure. The atomic specification for \texttt{insert} is
+The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The specification of \texttt{insert} is:
 \begin{mathpar}
@@
 Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range.
-We can then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node.
+We then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node.
 In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}.
-We do this by demonstrating that in both cases, the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
-We can then release the lock held by \lstinline{traverse} and deallocate the \lstinline{pn} structure.
+We do this by demonstrating that in both cases, the update to the concrete data structure corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the data structure.
+Finally, we release the lock acquired by \lstinline{traverse} and deallocate the \lstinline{pn} structure.
 
@@
 %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
-The specification of \lstinline{inRange} is shown in Figure \ref{fig:inRange_giveup}; it simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
-
-\begin{figure}[h]
-	\centering
+The specification of \lstinline{inRange} is:
 	\begin{mathpar}
@@
 	\end{mathpar}
-	\caption{Specification of \texttt{inRange} for the give-up template}
-	\label{fig:inRange_giveup}
-\end{figure}
+It simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
 
@@
 
-\begin{figure}[h]
-	\centering
+The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
 	\begin{mathpar}
@@
 	\end{mathpar}
-	\caption{Specification of \texttt{traverse} for the give-up template}
-	\label{fig:traverse_giveup}
-\end{figure}
-
-The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup}, is slightly simpler than that of the lock-coupling template: a thread does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition.
 As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents.

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
Finish the changing of font before 3.3.

@@
 These specifications are defined in terms of assertions $\nodeboxrep$, representing a client thread's handle to the data structure, and $\treerep$, representing the data structure's abstract state.
-Holding $\nodeboxrep(\texttt{n})$ for a node \texttt{n} should allow us to acquire the lock on \texttt{n}, which in turn gives us access to the contents of the node.
+Holding $\nodeboxrep(n)$ for a node $n$ should allow us to acquire the lock on $n$, which in turn gives us access to the contents of the node.
 Formally, we need to know that the triple \[{\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m \right\rangle}\ \texttt{acquire}(n\texttt{->lock})\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}\] holds, i.e., $\nodeboxrep(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant}.
@@
 		{\color{blue}
-			\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
-			\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
+			\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
+			\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
 			\right\rangle
@@
 		{\color{blue}
-			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-			\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
-				\\ 
-				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
-				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
+			\left\langle \mathit{res}. \ \exists \  n'.
+			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ 
+				\\ 
+				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range})
 			\end{array}
@@
 	\end{mathpar}
-The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(\texttt{n})$ and its contents $\mathsf{R}(\texttt{n})$, indicating that a thread should already hold \lstinline{n}'s lock before calling \lstinline{traverse}.
-The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}.
-The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.
+The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}.
+The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$.
+The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.
 
@@
 \begin{align*}
-	&\mathsf{node\_contents}(\texttt{n}, \mathsf{c}, \mathsf{range}) \triangleq \mathsf{ghost\_node(\texttt{n}, \mathsf{c}, range)} \ \ast \ \mathsf{node\_data}(\texttt{n}, \mathsf{c})\\
-	&\mathsf{R}(\texttt{n}) \triangleq \exists\ \mathsf{c}, \mathsf{range}.\ \mathsf{node\_contents}(\texttt{n}, \mathsf{c}, \mathsf{range})\end{align*}
+	&\mathsf{node\_contents}(n, \mathsf{c}, \mathsf{range}) \triangleq \mathsf{ghost\_node(n, \mathsf{c}, range)} \ \ast \ \mathsf{node\_data}(n, \mathsf{c})\\
+	&\mathsf{R}(n) \triangleq \exists\ \mathsf{c}, \mathsf{range}.\ \mathsf{node\_contents}(n, \mathsf{c}, \mathsf{range})\end{align*}
 where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.
-The contents $\mathsf{c}$ of a node can be either a key-value pair $(\mathsf{k}, \mathsf{v})$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).
+The contents $\mathsf{c}$ of a node can be either a key-value pair $(\texttt{k}, \mathsf{v})$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).
 
@@
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p}, \texttt{n}, \mathsf{c}, \mathsf{range}.\ &\texttt{pn} \mapsto (\texttt{p, n}) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(\texttt{n})  \ \ast \\  &\mathsf{node\_contents}(\texttt{n}, \mathsf{c}, \mathsf{range})
+\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ p, n, \mathsf{c}, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(n)  \ \ast \\  &\mathsf{node\_contents}(n, \mathsf{c}, \mathsf{range})
 \end{align*}
@@
 	$\color{blue}
-	\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
-	\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
+	\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
+	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
 	\right\rangle$
@@
 	$\color{blue}
-	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
-		\\ 
-		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
+	\left\langle \mathit{res}. \ \exists \  n'.
+	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ 
+		\\ 
+		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range})
 	\end{array}
@@
 We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$.
-Otherwise, we pass the $\mathsf{node\_data}$ from the lock invariant to \lstinline{findNext} (line 11), which returns a new node \lstinline{n''} to visit, stored in \lstinline{pn-> n}.
+Otherwise, we pass the $\mathsf{node\_data}$ from the lock invariant to \lstinline{findNext} (line 11), which returns a new node $n''$ to visit, stored in \lstinline{pn->n}.
 If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$.
-Otherwise, we acquire \lstinline{n->lock}, gaining access to the resources $\mathsf{R}(\texttt{n})$, and then release the lock of the current node \texttt{p} and return its resources.
+Otherwise, we acquire $n\texttt{->lock}$, gaining access to the resources $\mathsf{R}(n)$, and then release the lock of the current node $p$ and return its resources.
 %It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
-By acquiring the lock for the next node \texttt{n} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish $\mathsf{traverse\_inv}$ for the new values in \lstinline{pn}.
+By acquiring the lock for the next node $n$ before releasing the lock for the current node $p$, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish $\mathsf{traverse\_inv}$ for the new values in \lstinline{pn}.
 %The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_traverse.c

@@ -1,30 +1,30 @@
 int traverse(pn *pn, int k){
-		$\color{blue} \left\{\begin{array}{l}  \texttt{flag} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
-				\nodeboxrep(\texttt{n})  \ast \mathsf{R}(\texttt{n}) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l}  \mathit{flag} = \mathsf{true} \ast  \texttt{pn} \mapsto (p, n) \ast  
+				\nodeboxrep(n)  \ast \mathsf{R}(n) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
 		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
 				pn->p = pn->n;
-				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ast  \nodeboxrep(\texttt{n'}) \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(\texttt{n'}, \mathsf{c}, \mathsf{range})\end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \mathsf{c}, \mathsf{range})\end{array}\right\}$
 				if (pn->p->t == NULL)
 						return 1;						
-						$\color{violet} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \end{array}\right\}$
+						$\color{violet} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
[…]

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
Changed the font before 4.1.

@@
 			\begin{array}{c}
-				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \nodeboxrep(\texttt{n}) \ \big| \ \treerep\ m
+				\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
 			\end{array}
@@
 	\end{mathpar}
-As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents.
-Also as before, the Boolean $\mathit{res}$ indicates whether \lstinline{n'} is empty or contains the key \lstinline{k}.
+As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents.
+Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}.
 
@@
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{p}, \texttt{r}, \texttt{n}.\ \texttt{pn} \mapsto (\texttt{p, n}) \ast \ \infp (\texttt{r}) \ast \infp (\texttt{n})
+\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
 \end{mathpar}
 Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}).
-Instead, we keep a reference $\infp(\texttt{r})$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.
+Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.
 
@@
 We then call \lstinline{inRange} to check whether we are still on the path to a node that can hold \lstinline{k}.
-If the check fails (lines 18-20), we immediately release the lock and start over from the root node \lstinline{r}.
+If the check fails (lines 18-20), we immediately release the lock and start over from the root node $r$.
 Otherwise, we proceed as in the lock-coupling template: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release the lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}.

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
fixed bad spacing

@@
 Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}.
-The innovations of Iris (custom ghost state and invariants) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach.
+The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach.
 Some of the more complex data structures verified include the \lstinline{ConcurrentSkipListMap} from \lstinline{java.util.concurrent}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}.
@@
 Linearizability~\cite{linearizability}, the most common correctness condition for concurrent data structures, was a first step in this direction, describing the conditions under which a concurrent implementation can replace a sequential one in all possible contexts.
-Logical atomicity can be seen as a compositional analogue to linearizability, where the proof that each operation implements its sequential counterpart can be carried out independently.
-Concurrent templates can be seen as the next step of compositionality, allowing the concurrent-to-sequential proof to be decoupled from the specific data structure being verified.
+Logical atomicity has been shown to be a compositional analogue to linearizability~\cite{la-lin}, where the proof that each operation implements its sequential counterpart can be carried out independently.
+Concurrency templates can be seen as the next step towards compositionality, allowing us to prove linearizability/atomicity of concurrency patterns independently of specific data structures.
 
@@
 A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria.
-In the abstract, a search structure implements a map from keys to values, and provides operations such as search, insertion, deletion, and traversal on that map.
-For efficiency, many such structures are designed to allow concurrent access and modification, often employing fine-grained or lock-free concurrency to allow as many threads as possible to operate on separate parts of the data structure.
+In the abstract, a search structure implements a map from keys to values, providing operations such as search, insertion, deletion, and traversal on that map.
+For efficiency, many search structures allow concurrent access and modification, often employing fine-grained or lock-free concurrency to allow as many threads as possible to operate on separate parts of the data structure.
 Designing these search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory, including cache behavior~\cite{masstree}.
@@
 Figure~\ref{template-ex} shows an example search structure template.
-The core of the template is the \lstinline{traverse} function, which uses a particular concurrency control mechanism to travel through a data structure in search of a particular key.
+The core of the template is the \lstinline{traverse} function, which uses a specific concurrency control mechanism to travel through a data structure in search of the requested key.
 The mechanism in this example is \emph{lock coupling}, where we acquire the lock on the next node before releasing the lock on the current node.
-The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine when the current node does not contain the target key.
+The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine.
 Once the appropriate node for the key has been found, the template returns it to a top-level function such as \lstinline{insert} that calls out to the data structure to perform the actual insertion on the node.
@@
 Its concurrent counterpart could be specified as
-$$\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$$
+$$\forall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle$$
 indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution.
@@
 or show them here?
-The existing implementation is written in Iris's HeapLang, a simple functional programming language with shared-memory concurrency.
+Algorithms are verified in Iris's HeapLang, a simple functional programming language with shared-memory concurrency.
 

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
playing with formatting

@@
 
-The core of each search structure template is a \lstinline{traverse} function that navigates a data structure using the target synchronization mechanism.
+The core of a search structure template is a \lstinline{traverse} function that navigates a data structure using the chosen synchronization mechanism.
 A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well.
@@
 The lock-coupling pattern can be seen on lines 12-13, where \lstinline{traverse} acquires the next node's lock and then releases the current node's lock.
-The function stops when it reaches an empty leaf node (\lstinline{pn->p->t = NULL}), or when \lstinline{findNext} returns 0, indicating that the target key \lstinline{k} is in the current node.
-The \lstinline{traverse} function returns 1 when we reach a leaf node and 0 when we find \lstinline{k} in an existing node; the behavior of operations that call \lstinline{traverse} can vary depending on this return value.
+The function stops when it reaches an empty node (\lstinline{pn->p->t == NULL}), or when \lstinline{findNext} returns 0, indicating that the target key \lstinline{k} is in the current node.
+The \lstinline{traverse} function returns 1 when we reach an empty node and 0 when we find \lstinline{k} in an existing node; the behavior of operations that call \lstinline{traverse} can vary depending on this return value.
 For instance, an \lstinline{insert} operation may create a new node when \lstinline{traverse} returns 1 and modify an existing node when \lstinline{traverse} returns 0, while a \lstinline{lookup} operation may fail on 1 and return the value in the current node on 0.
@@
 If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}).
-Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the underlying data structure's \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
+Otherwise, \texttt{traverse} has reached an empty node that can hold key \texttt{k}, so it calls the underlying data structure's \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).
 
@@
 			\left\langle \mathit{res}. \ \exists \  n'.
-			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ 
+			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
 				\\ 
 				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range})
+				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
 			\end{array}
@@
 \begin{align*}
-	&\mathsf{node\_contents}(n, \mathsf{c}, \mathsf{range}) \triangleq \mathsf{ghost\_node(n, \mathsf{c}, range)} \ \ast \ \mathsf{node\_data}(n, \mathsf{c})\\
-	&\mathsf{R}(n) \triangleq \exists\ \mathsf{c}, \mathsf{range}.\ \mathsf{node\_contents}(n, \mathsf{c}, \mathsf{range})\end{align*}
+	&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c)\\
+	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
 where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.
-The contents $\mathsf{c}$ of a node can be either a key-value pair $(\texttt{k}, \mathsf{v})$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).
+The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).
 
@@
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ p, n, \mathsf{c}, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(n)  \ \ast \\  &\mathsf{node\_contents}(n, \mathsf{c}, \mathsf{range})
+\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(n)  \ \ast \\  &\mathsf{node\_contents}(n, c, \mathsf{range})
 \end{align*}
@@
 	\left\langle \mathit{res}. \ \exists \  n'.
-	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ 
+	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
 		\\ 
 		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range})
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
 	\end{array}
@@
 Figure \ref{proof_lock_traverse} shows the proof outline of the \texttt{traverse} function.
-We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$.
+We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the empty node where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \mathsf{true}$.
 Otherwise, we pass the $\mathsf{node\_data}$ from the lock invariant to \lstinline{findNext} (line 11), which returns a new node $n''$ to visit, stored in \lstinline{pn->n}.
-If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$.
+If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathsf{false}$.
 Otherwise, we acquire $n\texttt{->lock}$, gaining access to the resources $\mathsf{R}(n)$, and then release the lock of the current node $p$ and return its resources.
@@
 			\left\langle \mathit{res}. \ \exists \  n'.
-			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ 
+			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast\ 
 				\\ 
 				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range})
+				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
 			\end{array}
@@
 If the check fails (lines 18-20), we immediately release the lock and start over from the root node $r$.
-Otherwise, we proceed as in the lock-coupling template: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release the lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}.
+Otherwise, we proceed as in the lock-coupling template: if the current node's contents are \texttt{NULL}, we have found the empty node where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release the lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}.
 In this case, we once again hold no locks, and have only the $\infp$ assertion indicating that the new node is in the data structure.
@@
 	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
+	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
 		\\ 
 		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, v), \mathsf{range})
 	\end{array}

### 2023-05-02 concurrency/CONCUR 2023/proof_lock_traverse.c

@@ -1,28 +1,28 @@
 int traverse(pn *pn, int k){
-		$\color{blue} \left\{\begin{array}{l}  \mathit{flag} = \mathsf{true} \ast  \texttt{pn} \mapsto (p, n) \ast  
+		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (p, n) \ast  
 				\nodeboxrep(n)  \ast \mathsf{R}(n) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
 		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
 				pn->p = pn->n;
-				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \mathsf{c}, \mathsf{range})\end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', c, \mathsf{range})\end{array}\right\}$
 				if (pn->p->t == NULL)
 						return 1;						
 						$\color{violet} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
 				else{
-						$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', (\mathsf{k'}, \mathsf{v'}), \mathsf{range})\end{array}\right\}$
+						$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range})\end{array}\right\}$
 						int b = findNext(pn, k);
[…]

### 2023-05-03 concurrency/CONCUR 2023/paper.tex
release spec

@@
 Holding $\nodeboxrep(n)$ for a node $n$ should allow us to acquire the lock on $n$, which in turn gives us access to the contents of the node.
-Formally, we need to know that the triple \[{\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m \right\rangle}\ \texttt{acquire}(n\texttt{->lock})\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}\] holds, i.e., $\nodeboxrep(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant}.
+Formally, we need to know that the following triples hold:
+\begin{align*}
+&{\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m \right\rangle}\ \texttt{acquire}(n\texttt{->lock})\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}\\
+&{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m \right\rangle}\ \texttt{release}(n\texttt{->lock})\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m\right\rangle}
+\end{align*}
+In other words, $\nodeboxrep(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant} for the node.
 %In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).

### 2023-05-03 concurrency/CONCUR 2023/paper.tex
more spacing fixes

@@
 			\right\rangle
-		}\\
-		\texttt{traverse(pn, k)} \\
+		}\vspace{-6pt}\\
+		\vspace{-6pt}\texttt{traverse(pn, k)} \\
 		{\color{blue}
@@
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(n)  \ \ast \\  &\mathsf{node\_contents}(n, c, \mathsf{range})
+\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(n)  \ \ast \\  &\mathsf{node\_contents}(n, c, \mathsf{range})
 \end{align*}
@@
 Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires a lock just before operating on a node, and holds at most one lock at any time.
-This means that a conflicting operation may invalidate our traversal, for instance by moving the next node to another part of the data structure before we acquire its lock.
-To guard against this, the \lstinline{traverse} function must explicitly check whether the target key is in the range of the current node to ensure that it is still on the right path.
+This means that a conflicting operation may invalidate a traversal, for instance by moving the next node to another part of the data structure before we acquire its lock.
+To guard against this, the \lstinline{traverse} function must explicitly check whether the target key is in the range of the current node.
 If a check fails, we give up and start the traversal over from the root node.
-The give-up template performs well in scenarios where operations generally do not conflict, either because they are mostly to independent parts of the data structure or because they mostly do not delete or relocate nodes.
+The give-up template performs well in scenarios where operations generally do not conflict, either because they are on independent parts of the data structure or because they do not delete or relocate nodes.
 
@@
 
-Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented.
+Figure \ref{traverse_giveup_a} shows the give-up template algorithm as originally presented.
 In addition to \lstinline{findNext}, the \lstinline{traverse} function uses a helper function called \lstinline{inRange}, which determines whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors.
@@
 If the \lstinline{inRange} call fails, we release our current lock and return to the root node by setting the current node to the root pointer \lstinline{p} (lines 21-22 in Figure \ref{traverse_giveup_b}).
-The \texttt{insert} operation (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version, with the one difference that it does not need to acquire a lock before calling \texttt{traverse}.
+The give-up version of the \texttt{insert} operation (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version, except that it does not acquire a lock before calling \texttt{traverse}.
 
@@
 Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations.
-However, in most sequential settings the key ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself.
-The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to as many target data structures as possible without modifying them.
-Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t} in our C implementation:
+However, in most sequential settings ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself.
+The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to data structures without modifying them.
+Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
@@
 %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
-The specification of \lstinline{inRange} is:
+Our specification of \lstinline{inRange} is:
 	\begin{mathpar}
@@
 			\begin{array}{c}
-				\texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2}  
+				\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2
 			\end{array}
@@
 		}
-		\\ 
+		\vspace{-6pt}\\ \vspace{-6pt}
 		\texttt{inRange(pn, k)} 
@@
 			\begin{array}{l}
-				\texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2} \ \ast  \ 
-				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\mathsf{n_1} < \texttt{k} < \mathsf{n_2})\ \mathsf{else}\ (\texttt{k} \leq \mathsf{n_1} \lor \texttt{k} \geq \mathsf{n_2}))
+				\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2 \ \ast  \ 
+				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2))
 			\end{array}
@@
 	\end{mathpar}
-It simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.
+It simply computes whether the input key \lstinline{k} is in the range $(n_1, n_2)$ associated with the node \texttt{n}, and returns 0 or 1 accordingly.
 
@@
 		}
-		\\ 
+		\vspace{-6pt}\\ \vspace{-6pt}
 		\texttt{traverse(pn, k)} 
@@
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
+\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
 \end{mathpar}
@@
 We were able to obtain verified \lstinline{insert} and \lstinline{lookup} operations using this technique; for \texttt{delete}, we found a mismatch between the template approach and BST deletion, which we discuss in Section \ref{internal_reorganize}.
-%The complete proofs in VST can be found online at \href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}; we discuss the overall development in Section \ref{proof_mech}.
 
@@
 	\centering
-	%\wm{I think this should be \lstinline{pn->n->t} $\mapsto$ \lstinline{(k',v', l, r)}, but I'm not sure.}
 	\begin{mathpar}
@@
 	\end{mathpar}
-	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)} %\wm{Shouldn't pn have two fields, p and n?}
+	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)}
 	\label{fig:findNext_lock}
@@
 
-%\wm{Should we recap the definition of the \lstinline{node} struct here?} 
 The contents of a BST node are defined in C as:
@@
 The proof is straightforward.
-%\wm{We should show the actual code for \lstinline{findNext} here!}
 
@@
 	\lstinputlisting[language=C, style=myStyle]{findnext.c} 
-	\caption{The \lstinline{findNext} method for binary search tree (all templates) written in C}
+	\caption{C implementation of \lstinline{findNext} for the binary search tree (all templates)}
 	\label{findnext}

### 2023-05-03 concurrency/CONCUR 2023/paper.tex
Complete font change before appendix.

@@
 			\begin{array}{c}
-				\texttt{pn->n->t}\mapsto \texttt{(k', v', l, r)} 
+				\texttt{pn->n->t}\mapsto (k', v', l, r)
 			\end{array}
@@
 		{\color{blue}
-			\left\{\mathit{res}. \ \exists \  \texttt{n'.}
+			\left\{\mathit{res}. \ \exists \  n'.
 			\begin{array}{c}
-				\texttt{pn->n->t}\mapsto \texttt{(k', v', l, r)}  \\\  
-				\ast \ (\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{l} = \texttt{n'} \land \texttt{k} < \texttt{k'}) \lor (\texttt{r} = \texttt{n'} \land \texttt{k} > \texttt{k'})\ \mathsf{else}\ (\texttt{n} = \texttt{n'} \land \texttt{k} = \texttt{k'}))
+				\texttt{pn->n->t}\mapsto (k', v', l, r)  \\\  
+				\ast \ (\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k'))
 			\end{array}
@@
 		{\color{blue}
-			\left\{\exists \  \texttt{l r lk1 lk2}.
+			\left\{\exists \  l\ r\ \mathit{lk}_1\ \mathit{lk}_2.
 			\begin{array}{l}
-				\ \texttt{pn->n->t} \mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
-				\ \texttt{l} \mapsto \texttt{(NULL, lk1)}\  \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2)}
+				\ \texttt{pn->n->t} \mapsto (\texttt{k}, \texttt{v}, l, r) \ \ast \ \\ 
+				\ l \mapsto (\texttt{NULL}, \mathit{lk}_1)\  \ast \ r \mapsto (\texttt{NULL}, \mathit{lk}_2)
 			\end{array}
@@
 	{\color{blue}
-		\left\{\exists \ \texttt{l r lk1 lk2}. 
+		\left\{\exists \ l\ r\ \mathit{lk}_1\ \mathit{lk}_2. 
 		\begin{array}{c}
-			\texttt{pn->n->t}\mapsto \texttt{(k, v, l, r)} \  \ast \ \texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2}  \ \ast \ 
-			\\ \texttt{l} \mapsto (\texttt{NULL}, \texttt{lk1}, (\mathsf{n_1}, \mathsf{k})) \ \ast \ \texttt{r} \mapsto (\texttt{NULL}, \texttt{lk2}, (\mathsf{k}, \mathsf{n_2}))
+			\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r) \  \ast \ \texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2}  \ \ast \ 
+			\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (\mathsf{n_1}, \texttt{k})) \ \ast \ r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, \mathsf{n_2}))
 		\end{array}
@@
 The BST's \lstinline{findNext} function (Figure \ref{findnext}) indicates whether the target key is in the current node, its left subtree, or its right subtree.
-The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes.
-The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the node's key \texttt{k'}, and the next node \lstinline{n'} is the left child \lstinline{l}; \lstinline{k} is greater than \lstinline{k'} and \lstinline{n'} is the right child \lstinline{r}; or \lstinline{k} is exactly \lstinline{k'} and \lstinline{n'} is the original node \lstinline{n}, in which case the return value is 0.
+The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with key $k'$, value $v'$, and pointers to the left $l$ and right $r$ child nodes.
+The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the node's key $k'$, and the next node $n'$ is the left child $l$; \lstinline{k} is greater than $k'$ and $n'$ is the right child $r$; or \lstinline{k} is exactly $k'$ and $n'$ is the original node $n$, in which case the return value is 0.
 The proof is straightforward.
@@
 The \texttt{insertOp} specification (Figure \ref{fig:insertOp_lock}) takes as input a node whose \lstinline{t} field is \lstinline{NULL}, i.e., an empty leaf node.
-It inserts a new node with the specified key \texttt{k} and value \texttt{v} at that leaf, complete with two new empty child nodes with corresponding locks, \texttt{(NULL, lk1)} and \texttt{(NULL, lk2)}, which are pointed to by the \texttt{l} and \texttt{r} of the current node.
+It inserts a new node with the specified key \texttt{k} and value \texttt{v} at that leaf, complete with two new empty child nodes with corresponding locks, $(\texttt{NULL}, \mathit{lk}_1)$ and $(\texttt{NULL}, \mathit{lk}_2)$, which are pointed to by the $l$ and $r$ of the current node.
 
@@
 As shown in Figure \ref{fig:insertOp_giveup}, a give-up template node (even an empty one!) has additional \lstinline{min} and \lstinline{max} fields, and these fields are needed to compute the bounds of the new child nodes.
-In a BST, if the current node's range is \lstinline{(min, max)} and its key is \lstinline{k}, its left child will have range \texttt{(min, k)} and its right child will have range \texttt{(k, max)}.
+In a BST, if the current node's range is (\lstinline{min, max}) and its key is \lstinline{k}, its left child will have range (\texttt{min, k}) and its right child will have range (\texttt{k, max}).
 It may be possible to remedy this discrepancy by adding an ``allocate empty node'' function to the templates, which is then called by \lstinline{insertOp}, but this would still complicate the approach: originally, we expected that templates would be parameterized by data structure implementations but not vice versa.

### 2023-05-03 concurrency/CONCUR 2023/proof_lock_traverse.c
tiny font tweak

@@ -14,7 +14,7 @@ int traverse(pn *pn, int k){
 						((\texttt{b} = 0 \ast k' = \texttt{k} \ast n'' = n')  \lor 						 (\texttt{b} = 1  \ast \texttt{k} \in \mathsf{range}(n'') \ast \nodeboxrep(n'')))   \end{array}\right\}$
 						if (b == 0){
 								$\color{blue} \left\{\begin{array}{l} 
-								\texttt{pn} \mapsto (n', n') \ast \mathsf{k' = \texttt{k}} \ast \cdots    \end{array}\right\}$
+								\texttt{pn} \mapsto (n', n') \ast k' = \texttt{k} \ast \cdots    \end{array}\right\}$
 								return 0;
 								$\color{violet} \left\{\begin{array}{l} 
 								\texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range'} \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range'})    \end{array}\right\}$

### 2023-05-03 concurrency/CONCUR 2023/paper.tex
small font tweaks

@@
 			\begin{array}{c}
-				\texttt{pn->n->t}\mapsto (k', v', l, r)  \\\  
-				\ast \ (\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k'))
+				\texttt{pn->n->t}\mapsto (k', v', l, r)  \ \ast \\\  
+				(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k'))
 			\end{array}
@@
 		\begin{array}{c}
-			\texttt{pn->n}\mapsto \texttt{NULL} \  \ast \ \texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2}
+			\texttt{pn->n}\mapsto \texttt{NULL} \  \ast \ \texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2
 		\end{array}
@@
 		\begin{array}{c}
-			\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r) \  \ast \ \texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2}  \ \ast \ 
-			\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (\mathsf{n_1}, \texttt{k})) \ \ast \ r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, \mathsf{n_2}))
+			\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r) \  \ast \ \texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2  \ \ast \ 
+			\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ \ast \ r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
 		\end{array}
@@
 	\begin{array}{c}
-		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \nodeboxrep(\texttt{n}) \ \big| \ \treerep\ m
+		\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
 	\end{array}
@@
 	$\color{blue}
-	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
-	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
-		\\ 
-		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, v), \mathsf{range})
+	\left\langle \mathit{res}. \ \exists \  n'.
+	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
+		\\ 
+		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v}), \mathsf{range})
 	\end{array}

### 2023-05-03 concurrency/CONCUR 2023/proof_giveup_traverse.c
making contents font consistent

@@ -8,25 +8,25 @@ int traverse(pn *pn, int k){
 				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
 				pn->p = pn->n; $\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
 				if (inRange(pn, k) == 1){
-						$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{node\_contents}(n', \mathsf{c}, \mathsf{range}) \ast \infp (\texttt{r}) \end{array}\right\}$
+						$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{node\_contents}(n', c, \mathsf{range}) \ast \infp (\texttt{r}) \end{array}\right\}$
 						if (pn->p->t == NULL)
 								return 1;
 								$\color{violet} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
 						else{
-								$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{node\_contents}(n', (\mathsf{k'}, \mathsf{v'}), \mathsf{range}) \ast \infp (\texttt{r}) \end{array}\right\}$
+								$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range}) \ast \infp (\texttt{r}) \end{array}\right\}$
 								int b = findNext(pn, k);
[…]

### 2023-05-03 concurrency/CONCUR 2023/proof_giveup_insert.c
Changed font in Figure 9.

@@ -1,23 +1,23 @@
 void insert (node_t **r, int k, void *v){
 		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
 		pn->n = *r;
-		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \infp(\texttt{r}) \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \infp(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-				$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \infp(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, (\texttt{k}, \mathsf{v'}), \mathsf{range'}) \end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range'}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, (\texttt{k}, \texttt{v}), \mathsf{range'}) \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range'}) \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
 				//Linearization point
 		}
 		else{
[…]

### 2023-05-03 concurrency/CONCUR 2023/proof_lock_insert.c

@@ -1,25 +1,25 @@
 void insert (node_t **r, int k, void *v){
 		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
 		pn->n = *r;
-		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \nodeboxrep(\texttt{r}) \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \end{array}\right\}$
 		acquire(pn->n->lock);
-		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  \nodeboxrep(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-		$\color{blue} \left\{\begin{array}{l}  \exists\ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, (\texttt{k}, \mathsf{v'}), \mathsf{range'}) \end{array}\right\}$
+		$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range'}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{n', n'}) \ast \nodeboxrep(\texttt{n'}) \ast \mathsf{node\_contents}(\texttt{n'}, (\texttt{k}, \texttt{v}), \mathsf{range'}) \end{array}\right\}$ 
[…]

### 2023-05-03 concurrency/CONCUR 2023/proof_giveup_insert.c
final proof formatting

@@ -3,17 +3,17 @@ void insert (node_t **r, int k, void *v){
 		pn->n = *r;
 		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \infp(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range'}) \end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range'}) \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
 				//Linearization point
 		}
 		else{
-				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \infp(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range'}) \ast \texttt{k} \in \mathsf{range'} \end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \infp(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$
 				insertOp(pn, k, v);
[…]

### 2023-05-03 concurrency/CONCUR 2023/proof_lock_insert.c

@@ -5,17 +5,17 @@ void insert (node_t **r, int k, void *v){
 		acquire(pn->n->lock);
 		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-		$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range'}) \end{array}\right\}$
+				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range'}) \end{array}\right\}$ 
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
 				//Linearization point
 		}
 		else{
-				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range'}) \ast \texttt{k} \in \mathsf{range'} \end{array}\right\}$
[…]

### 2023-06-05 concurrency/CONCUR 2023/response.txt
response outline

Thank you for your time and consideration!

A general note: several reviewers mentioned that they were disappointed in the limitations of the template approach, or wanted to see more evidence that it was effective in practice. In fact, we believe that one of our contributions is finding limitations that were not apparent in the original research [10], and will need to be addressed before putting the approach into practice on a larger scale.

Responses to Reviewer 1's questions:

"Moreover, the paper only gives one example... For the second template (the give-up template) it only discusses why this cannot be directly applied to the BST structure."

Actually, we have also verified the give-up template (for lookup and insert) on the BST. The text at the end of section 4.1 was only meant to imply that we had to break the template interface slightly in order to do so, not that we were unable to. So we do have an example where m = 2.

"Can you give a more explicit discussion of the challenges to use the template approach for C programs?"

"How do your correctness proofs compare with those for the flow interface templates?"

"Can you make your questions about the boundaries of the template/data structure division more explicit? And do you have ideas on how to address them?"

"Are there other examples which can be verified in your approach, e.g. other data structures for lock coupling, or data structures to which multiple templates can be applied?"

"How specific is your change to C? Could it also be used to reason about e.g. Java programs?"

Responses to Reviewer 3's questions:

"I was unable to understand how the updates of the abstract state of the CSS are obtained from the node-level updates at the linearization points. For instance, how do you obtain Node(m[k |-> v]) from Line 18 in Figure 8?"

"In the earlier work, the keysets are tied to the graph structure using an inductive invariant... What is the corresponding mechanism used in this paper?"

"I think the criticism of the template approach based on the proof of the delete operation is perhaps a bit overblown..."

(maybe we just agree with this one?)

Responses to Reviewer 4's questions:

"What exactly is the challenge in realizing this using VST as opposed to other systems like Iris?"

"While you say you'd like to verify structures such as Masstree and Wormhole, it's not clear whether you think the template-based approach would work, especially given your reservations for even "delete" for a bst."

### 2023-06-08 concurrency/CONCUR 2023/response.txt
filled in most of the response

@@
-Thank you for your time and consideration!
-
-A general note: several reviewers mentioned that they were disappointed in the limitations of the template approach, or wanted to see more evidence that it was effective in practice.
-In fact, we believe that one of our contributions is finding limitations that were not apparent in the original research [10], and will need to be addressed before putting the approach into practice on a larger scale.
+Thank you for your time and consideration! We'll take your feedback into account for any future version of the paper.
 
@@
 
-"Can you give a more explicit discussion of the challenges to use the template approach for C programs?"
+"Can you give a more explicit discussion of the challenges to use the template approach for C programs? ... How do your correctness proofs compare with those for the flow interface templates?"
 
-"How do your correctness proofs compare with those for the flow interface templates?"
+The biggest difference/challenge is caused by the switch from recursive calls to loops.
+In the proofs, this means we need to come up with loop invariants, where the previous template proofs were able to use simpler Löb induction.
+Coming up with the loop invariants was the hardest part of the proofs.
 
@@
 
+This will take some research, but we have ideas about how to formally separate the data structure and template specifications, e.g.
+by dividing the source code into multiple files and verifying each parameterized by a typeclass that describes the assumptions of that component.
+Something similar has been done in previous VST developments with Verified Software Units (Beringer, ESOP 2021).
+
 "Are there other examples which can be verified in your approach, e.g. other data structures for lock coupling, or data structures to which multiple templates can be applied?"
+
+We are currently working on a linked-list verification, and believe we can reproduce all the examples from [10].
+We have also started to look at the more complex data structures we mentioned (Wormhole, Masstree), which would use multiple templates in a single data structure.
 
@@
 
+The machinery of VST is completely specific to C, and there is no analogous (foundational, theorem-prover-based) system for Java.
+The overall shape of the specifications and proofs, however, should apply in exactly the same way to Java programs, and if we were given an analogous Java verification system we believe we could port all the work described to it.
+
 Responses to Reviewer 3's questions:
 
+"The decision to abandon the flow framework... restrict the ability of the verified templates..."
+
+We did this in large part to simplify the formalization, and in fact almost all the existing examples (minus the B-link tree) are tree-structured rather than graph-structured, so they don't really take advantage of flow interfaces anyway.
+If the templates are properly generic, they should be instantiatable with flow-interface-based invariants as well as the ones we use here, so we shouldn't lose any expressive power.
+
 "I was unable to understand how the updates of the abstract state of the CSS are obtained from the node-level updates at the linearization points.
-For instance, how do you obtain Node(m[k |-> v]) from Line 18 in Figure 8?"
+For instance, how do you obtain Node(m[k |-> v]) from Line 18 in Figure 8?" Also, "In the earlier work, the keysets are tied to the graph structure using an inductive invariant...
+What is the corresponding mechanism used in this paper?"
 
-"In the earlier work, the keysets are tied to the graph structure using an inductive invariant... What is the corresponding mechanism used in this paper?"
+We have essentially exactly the same style of inductive invariants in our give-up template: we just use a simpler form of ghost state than flow interfaces to express the contents of a node.
+Just as in the previous work, at linearization points we combine the ghost state of the affected node(s) with corresponding ghost state held in the global invariant for the data structure, and then use ghost state updates to change the global state to match the modified data structure.
+For instance, the Node assertion in Figure 9 is a separating conjunction over a Coq-level tree of the per-node assertions for each node; at the linearization point, we create new ghost state for the newly allocated nodes l and r', and update the existing ghost state for n', recombining them into a Node assertion for the updated map m[k |-> v].
+
+The lock-coupling template is factored slightly differently, with recursive lock invariants outside the global invariant, but this is for historical reasons: we could refactor it to use the same style of invariant as in the flow-interface proofs and our give-up example.
 
@@
 
-(maybe we just agree with this one?)
+This is a good point, and we intend to rephrase the criticism as follows.
+Essentially, the template approach requires dividing data structure operations into two categories: operations that can be templated, and "maintenance"-style operations that are no-ops on the abstract state but involve mixed synchronization and data structure reasoning.
+The existing work on templates says a lot about the first category, but the theoretical details of the second category, and the way the two are combined into a fully verified data structure, still need to be worked out more formally.
 
@@
 
+There are two main categories of challenges.
+The first comes from the details of real C code: proving absence of integer overflows, accurately representing struct-and-pointer-based data structures instead of functional-programming-style structured data, etc.
+The second, as mentioned above, is that idiomatic C tree implementations use loops instead of recursion, and so we had to come up with fairly complex loop invariants instead of using Löb induction.
+Some of these challenges are not inherent in using VST, but rather came from our goal of verifying (refactorings of) existing data structure implementations rather than writing data structures from scratch for the template approach.
+
 "While you say you'd like to verify structures such as Masstree and Wormhole, it's not clear whether you think the template-based approach would work, especially given your reservations for even "delete" for a bst."
+
+As stated in the last part of the response to Reviewer 3, we think that templates alone will not be enough to verify Masstree and Wormhole, but that the "template-based approach" actually quietly includes separate class of non-templated maintenance operations, and the combination of the two classes will be enough.
+This will require formalizing exactly how maintenance operations interact with template-verified operations.

### 2023-08-23 concurrency/CPP 2024/paper.tex
initial change for CPP24

--- 2023-05-03 concurrency/CONCUR 2023/paper.tex
@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
-%This is a template for producing LIPIcs articles. 
-%See lipics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
-
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
-\bibliographystyle{plainurl}% the mandatory bibstyle
-\usepackage{txfonts}
+%% Commands for TeXCount
+%TC:macro \cite [option:text,text]
+%TC:macro \citep [option:text,text]
+%TC:macro \citet [option:text,text]
+%TC:envir table 0 1
+%TC:envir table* 0 1
+%TC:envir tabular [ignore] word
+%TC:envir displaymath 0 word
+%TC:envir math 0 word
+%TC:envir comment 0 0
+%%
+%%
+%% The first command in your LaTeX source must be the \documentclass command.
+\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%%
+%% \BibTeX command to typeset BibTeX logo in the docs
+\AtBeginDocument{%
+  \providecommand\BibTeX{{%
+    Bib\TeX}}}
+
+%\usepackage{txfonts}
 \usepackage{subcaption} %% For complex figures with subfigures/subcaptions
@@
 \newcommand{\treerep}{\ensuremath{\mathsf{Node}}}
-\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Node\_ref}}}
+\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Node\_ref }}}
 \newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
-\newcommand{\infp}{\ensuremath{\mathsf{Node\_ref}}}
+\newcommand{\infp}{\ensuremath{\mathsf{Node\_ref }}}
 
@@
 
-\title{Compositional Verification of Concurrent C Programs with Search Structure Templates} %keep working on this
-
-%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line
-
-\author{Duc Than Nguyen}{University of Illinois at Chicago, USA \and \url{http://dnguye96.people.uic.edu} }{dnguye96@uic.edu}{https://orcid.org/0000-0002-6810-897X}{%(Optional) author-specific funding acknowledgements
-}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
-Please provide at least the name of the affiliation and the country.
-The full address is optional.
-Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.
-
-\author{Lennart Beringer}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~eberinge} }{eberinge@cs.princeton.edu}{https://orcid.org/0000-0002-1570-3492}{}
-
-\author{William Mansky}{University of Illinois at Chicago, USA\and \url{https://mansky.lab.uic.edu/} }{mansky1@uic.edu}{https://orcid.org/0000-0002-5351-895X}{}
-
-\author{Shengyi Wang}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~shengyiw} }{shengyiw@cs.princeton.edu}{https://orcid.org/0000-0002-2286-8703}{}
-
-\authorrunning{Duc Than Nguyen, Lennart Beringer, William Mansky, and Shengyi Wang} %TODO mandatory.
-First: Use abbreviated first/middle names.
-Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Duc Than Nguyen, Lennart Beringer, William Mansky, and Shengyi Wang} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011074.10011099.10011692</concept_id>
-<concept_desc>Software and its engineering~Formal software verification</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Formal software verification}
-
-\keywords{concurrent separation logic, fine-grained locking, logical atomicity, theorem proving,
-	Verified Software Toolchain, Iris} %TODO mandatory; please add comma-separated list of keywords
-
-\category{} %optional, e.g. invited paper
-
-\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-%\relatedversiondetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g.
-Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional
-
-%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-%\supplementdetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
-Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
-
-\funding{This work was funded in part by NSF Grant \#1811894.}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-%\acknowledgements{I want to thank \dots}%optional
-
-%\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+%% Rights management information.  This information is sent to you
+%% when you complete the rights form.  These commands have SAMPLE
+%% values in them; it is your responsibility as an author to replace
+%% the commands and values with those provided to you when you
+%% complete the rights form.
+\setcopyright{acmcopyright}
+\copyrightyear{2018}
+\acmYear{2018}
+\acmDOI{XXXXXXX.XXXXXXX}
+
+%% These commands are for a PROCEEDINGS abstract or paper.
+\acmConference[Conference acronym 'XX]{Make sure to enter the correct
+  conference title from your rights confirmation email}{June 03--05,
+  2018}{Woodstock, NY}
+\acmPrice{15.00}
+\acmISBN{978-1-4503-XXXX-X/18/06}
+
+%%
+%% Submission ID.
+%% Use this when submitting an article to a sponsored event. You'll
+%% receive a unique submission ID from the organizers
+%% of the event, and this ID should be used as the parameter to this command.
+%%\acmSubmissionID{123-A56-BU3}
+
+%%
+%% For managing citations, it is recommended to use bibliography
+%% files in BibTeX format.
+%%
+%% You can then either use BibTeX with the ACM-Reference-Format style,
+%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
+%% support for advanced citation of software artefact from the
+%% biblatex-software package, also separately available on CTAN.
+%%
+%% Look at the sample-*-biblatex.tex files for templates showcasing
+%% the biblatex styles.
+%%
+
+%%
+%% The majority of ACM publications use numbered citations and
+%% references.  The command \citestyle{authoryear} switches to the
+%% "author year" style.
+%%
+%% If you are preparing content for an event
+%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
+%% citations and references.
+%% Uncommenting
+%% the next command will enable that style.
+%%\citestyle{acmauthoryear}
+
+%%
+%% end of the preamble, start of the body of the document source.
 \begin{document}
 
-\maketitle
-
-%don't say ``real'' so much
+%%
+%% The "title" command has an optional parameter,
+%% allowing the author to define a "short title" to be used in page headers.
+\title{Compositional Verification of Concurrent C Programs with Search Structure Templates}
+
+%%
+%% The "author" command and its associated commands are used to define
+%% the authors and their affiliations.
+%% Of note is the shared affiliation of the first two authors, and the
+%% "authornote" and "authornotemark" commands
+%% used to denote shared contribution to the research.
+\author{Duc Than Nguyen}
+%\authornote{Both authors contributed equally to this research.}
+\email{dnguye96@uic.edu}
+\orcid{}
+\affiliation{%
+  \institution{University of Illinois Chicago}
+  \streetaddress{}
+  \city{Chicago}
+  \state{Illinois}
+  \country{USA}
+  \postcode{}
+}
+
+%%
+%% By default, the full list of authors will be used in the page
+%% headers. Often, this list is too long, and will overlap
+%% other information printed in the page headers. This command allows
+%% the author to define a more concise list
+%% of authors' names for this purpose.
+%\renewcommand{\shortauthors}{Trovato et al.}
+
+%%
+%% The abstract is a short summary of the work to be presented in the
+%% article.
 \begin{abstract}
-%a bit more of a lead-in?
 Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort.
@@
 
+%%
+%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
+%% Please copy and paste the code instead of the example below.
+%%
+
+%%
+%% Keywords. The author(s) should pick words that accurately describe
+%% the work being presented. Separate the keywords with commas.
+\keywords{}
+
+%%
+%% This command processes the author and affiliation and title
+%% information and builds the first part of the formatted document.
+\maketitle
+
 \section{Introduction}
 \label{sec:introduction}
-Krishna et al.~proposed concurrent search structure templates~\cite{templates} as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree).
+Krishna et al.
+~\cite{templates} proposed concurrent search structure templates as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree).
 The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency.
@@
 The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach.
-Some of the more complex data structures verified include the \lstinline{ConcurrentSkipListMap} from \lstinline{java.util.concurrent}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}.
+Some of the more complex data structures verified include the \lstinline{ConcurrentSkipListMap} from \lstinline{java.util.concurrent} ~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}.
 Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeted at an implementation language (C) rather than a toy language or algorithm-level representation.
@@
 
+\begin{figure*}
+
+\end{figure*}
+
 Figure~\ref{template-ex} shows an example search structure template.
@@
 
-As an example, a sequential stack \lstinline{push} operation could conventionally be specified as \\$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$.
-Its concurrent counterpart could be specified as
-$$\forall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle$$
+As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
+$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$. Its concurrent counterpart could be specified as 
+\begin{mathpar}
+	\forall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \\ \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
+\end{mathpar}
 indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution.
@@
 We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
 	typedef struct node_t {node *t; lock_t *lock;} node_t;
@@
 Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
-typedef struct node {int key; void *value; struct node_t *left, *right;} node;
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
+	typedef struct node {
+		int key; 
+		void *value; 
+		struct node_t *left, *right;} 
+	node;
 \end{lstlisting}
 while a linked list might use
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
-typedef struct node {int key; void *value; struct node_t *next;} node;
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
+	typedef struct node {
+		int key; 
+		void *value; 
+		struct node_t *next;
+	} node;
 \end{lstlisting}
@@
 The C implementation uses a struct
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
 	typedef struct pn {struct node_t *p; struct node_t *n;} pn;
@@
 
-\begin{figure}[h]
+\begin{figure*}[h]
 	\begin{subfigure}[t]{0.48\textwidth}
@@
 	\label{traverse_lock}
-\end{figure}
+\end{figure*}
 
@@
 Formally, we need to know that the following triples hold:
-\begin{align*}
-&{\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m \right\rangle}\ \texttt{acquire}(n\texttt{->lock})\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}\\
-&{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m \right\rangle}\ \texttt{release}(n\texttt{->lock})\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m\right\rangle}
-\end{align*}
+
+\begin{mathpar}
+	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}\\
+	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{release}(n\texttt{->lock})\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m\right\rangle}
+\end{mathpar}
+
 In other words, $\nodeboxrep(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant} for the node.
@@
 	{\color{blue}
-			\left\langle \mathit{res}. \ \exists \  n'.
-			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
-				\\ 
-				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
-			\end{array}
-			\ \Bigg| \ \treerep\ m \
+		\left\langle \mathit{res}. \ \exists \  n', v.
+		\begin{array}{l} \texttt{pn} \mapsto \left(n', n'\right) \ \ast \\ \nodeboxrep\left(n'\right) \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
+			\\ 
+			(\mathsf{if} \ \mathit{res} \ \\ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+			\\ \mathsf{else} \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right))
+		\end{array}
+		\ \Bigg | \ \treerep\ m \
 		\right\rangle
@@
 Formally, the lock invariant is defined by
-\begin{align*}
-	&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c)\\
-	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
+\begin{align*}\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
+	\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
+	\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
 where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.
@@
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(n)  \ \ast \\  &\mathsf{node\_contents}(n, c, \mathsf{range})
+\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \nodeboxrep(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
 \end{align*}
 
-\begin{figure}[!ht]
+\begin{figure*}[!ht]
 	$\color{blue}
@@
 	\label{proof_lock_traverse}
-\end{figure}
+\end{figure*}
 
@@
 	}
-	\ \texttt{insert(r, k, v)}\ 
+	\\ \texttt{insert(r, k, v)}\  \\
 	{\color{blue}
@@
 	\label{traverse_giveup}
-\end{figure}
+\end{figure*}
 
@@
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
-	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
+	typedef struct node_t {
+		node *t; 
+		lock_t *lock; 
+		int min; 
+		int max;
+	} node_t;
 \end{lstlisting}
@@
 	{\color{blue}
-			\left\langle \mathit{res}. \ \exists \  n'.
-			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast\ 
-				\\ 
-				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
-			\end{array}
-			\ \Bigg| \ \treerep\ m \
+		\left\langle \mathit{res}. \ \exists \  n', v.
+		\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \\ \texttt{k} \in \mathsf{range}\ \ast\ 
+			\\ 
+			(\mathsf{if} \ \mathit{res} \\ 
+			\ \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+			\\ 
+			\ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
+		\end{array}
+		\Bigg| \ \treerep\ m \
 		\right\rangle
@@
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
+\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
 \end{mathpar}
@@
 The top-level specifications we obtain for search structure operations are shown below: each operation takes effect on the data structure atomically, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
+\begin{figure*}[!ht]
+	\centering
 \begin{mathpar}
@@
 \end{mathpar}
+\end{figure*}
 As we saw in the previous section, these top-level specifications have already been verified at the template level; all we need to do now is provide implementations of data-structure-specific functions (e.g.
@@
 
-\begin{figure}[h]
+\begin{figure*}[!ht]
 	\centering
@@
 	\label{fig:findNext_lock}
-\end{figure}
-
-\begin{figure}[h]
+\end{figure*}
+
+\begin{figure*}[h]
 	\begin{subfigure}{\textwidth}
@@
 	\label{fig:insertOp}
-\end{figure}
+\end{figure*}
 
@@
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
-	typedef struct node {int key; void *value; struct node_t *left, *right;} node;
+	typedef struct node {
+		int key; 
+		void *value; 
+		struct node_t *left, *right;
+	} node;
 \end{lstlisting}
@@
 	\begin{center}
-		\begin{tabular}{l | c || l | c } 
+		\begin{tabular}{l | l || l | l } 
 			\hline
-			\textbf{Lock-coupling template} & \textbf{LoC} & \textbf{Give-up  template} & \textbf{LoC} \\
+			\textbf{Lock coupling} & \textbf{LoC} & \textbf{Give up} & \textbf{LoC} \\
 			\hline
-			Template specfications & 190 & Template specfications & 138 \\
+			Specfications & 190 & Specfications & 138 \\
 			\texttt{findNext} proof 	& 27  & \texttt{findNext} proof & 27 \\
@@
 The full development of our mechanization effort is available online at
+
 \href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.
@@
 
-%%
-%% Bibliography
-%% 
-%%
-
+\ignore{
+\begin{acks}
+Thanks to Roshan Sharma, Alex Oey, and Anastasiia Evdokimova for extensive work on the original binary search tree implementation and verification.
+\end{acks}
+}
+
+%12 pages
+
+%%
+%% The next two lines define the bibliography style to be used, and
+%% the bibliography file.
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{../sources}
-\newpage
+
+%%
+%% If your work has an appendix, this is the place to put it.
 \appendix
@@
 
-\begin{figure}[!ht]
+\begin{figure*}[!ht]
 	$\color{blue}
@@
 		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v}), \mathsf{range})
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
 	\end{array}
@@
 	\label{proof_giveup_traverse}
-\end{figure}
-
-\begin{figure}[!ht]
+\end{figure*}
+
+\begin{figure*}[!ht]
 	\begin{subfigure}{\textwidth}
@@
 		\right\rangle$
-		\caption{Proof outline of the lock-coupling \texttt{insert} function \\\\}
+		\caption{Proof outline of the lock-coupling \texttt{insert} function}
 		\label{proof_lock_insert}	
@@
 		\right\rangle$
-		\caption{Proof outline of the give-up \texttt{insert} function \\\\}
+		\caption{Proof outline of the give-up \texttt{insert} function}
 		\label{proof_giveup_insert}
@@
 	\label{proof_lock_giveup_insert}
-\end{figure}
+\end{figure*}
 
 \end{document}
+\endinput
+%%
+%% End of file `sample-sigplan.tex'.

### 2023-08-23 concurrency/CPP 2024/findnext.c

@@ -1,0 +1,13 @@
+int findNext (pn *pn, int k){
+  int y = pn->p->t->key;
+  if (x < y){
+    pn->n = pn->p->t->left;
+    return 1;
+  } 
+  else if (x > y){
+    pn->n = pn->p->t->right;
+    return 1;
+  } 
+  else
+    return 0;
+}

### 2023-08-23 concurrency/CPP 2024/giveup_insert.c

@@ -1,0 +1,12 @@
+void insert (node_t **r, int k, void *v){
+  struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+  pn->n = *r;
+  if (traverse(pn, k) == 0){
+    pn->p->t->value = v;
+  }
+  else{
+    insertOp(pn, k, v);
+  }
+  release(pn->n->lock);
+  free(pn);
+}

### 2023-08-23 concurrency/CPP 2024/giveup_traverse.c

@@ -1,0 +1,23 @@
+int traverse(pn *pn, int k){
+  node_t *r = (pn->n); 
+  for( ; ; ){
+    acquire(pn->n->lock); 
+    pn->p = pn->n;
+    if (inRange(pn, k) == 1){
+      if (pn->p->t == NULL)
+        return 1;
+      else{
+        int b = findNext(pn, k);
+        if (b == 0){
+          return 0;
+        }
+        else
+          release(pn->p->lock);
+      }
+    }
+    else{
+      release(pn->p->lock);
+      pn->n = r;
+    }
+  }
+}

### 2023-08-23 concurrency/CPP 2024/lock_insert.c

@@ -1,0 +1,13 @@
+void insert (node_t **r, int k, void *v){
+  struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+  pn->n = *r;
+  acquire(pn->n->lock);  
+  if (traverse(pn, k) == 0){
+    pn->p->t->value = v;
+  }
+  else{
+    insertOp(pn, k, v);
+  }
+  release(pn->n->lock);
+  free(pn);
+}

### 2023-08-23 concurrency/CPP 2024/lock_traverse.c

@@ -1,0 +1,17 @@
+int traverse(pn *pn, int k){
+  for( ; ; ){
+    pn->p = pn->n;
+    if (pn->p->t == NULL)
+      return 1;
+    else{
+      int b = findNext(pn, k);
+      if (b == 0){
+        return 0;
+      }
+      else{
+        acquire(pn->n->lock);
+        release(pn->p->lock);
+      }
+    }
+  }
+}

### 2023-08-23 concurrency/CPP 2024/proof_giveup_insert.c

@@ -1,0 +1,23 @@
+void insert (node_t **r, int k, void *v){
+		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+		pn->n = *r;
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \infp(\texttt{r}) \end{array}\right\}$
+		if (traverse(pn, k) == 0){
+				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
+				pn->p->t->value = v;
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
+				//Linearization point
+		}
+		else{
+				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \infp(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$
+				insertOp(pn, k, v);
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range}\ \ast \\ n'\texttt{->t}\mapsto (\texttt{k}, \texttt{v}, l, r')  \ast l \mapsto \texttt{NULL} \ast  r' \mapsto \texttt{NULL} \end{array}\right\}$ 
+				//Linearization point
+				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$ 
+		}
[…]

### 2023-08-23 concurrency/CPP 2024/proof_giveup_traverse.c

@@ -1,0 +1,40 @@
+int traverse(pn *pn, int k){
+		node_t *r = (pn->n); $\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  
+\infp (\texttt{r}) \end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
+			\exists\ n'.\ \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \infp (\texttt{r})
+		 \end{array}\right\}$
+				acquire(pn->n->lock);
+				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
+				pn->p = pn->n; $\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
+				if (inRange(pn, k) == 1){
+						$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{node\_contents}(n', c, \mathsf{range}) \ast \infp (\texttt{r}) \end{array}\right\}$
+						if (pn->p->t == NULL)
+								return 1;
+								$\color{violet} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
+						else{
+								$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range}) \ast \infp (\texttt{r}) \end{array}\right\}$
+								int b = findNext(pn, k);
[…]

### 2023-08-23 concurrency/CPP 2024/proof_lock_insert.c

@@ -1,0 +1,25 @@
+void insert (node_t **r, int k, void *v){
+		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+		pn->n = *r;
+		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \end{array}\right\}$
+		acquire(pn->n->lock);
+		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
+		if (traverse(pn, k) == 0){
+				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
+				pn->p->t->value = v;
+				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
+				//Linearization point
+		}
+		else{
+				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$
+				insertOp(pn, k, v);
+				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range}\ \ast \\ n'\texttt{->t}\mapsto (\texttt{k}, \texttt{v}, l, r')  \ast l \mapsto \texttt{NULL} \ast  r' \mapsto \texttt{NULL} \end{array}\right\}$ 
+				//Linearization point
[…]

### 2023-08-23 concurrency/CPP 2024/proof_lock_traverse.c

@@ -1,0 +1,31 @@
+int traverse(pn *pn, int k){
+		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (p, n) \ast  
+				\nodeboxrep(n)  \ast \mathsf{R}(n) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
+				pn->p = pn->n;
+				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', c, \mathsf{range})\end{array}\right\}$
+				if (pn->p->t == NULL)
+						return 1;						
+						$\color{violet} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
+				else{
+						$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range})\end{array}\right\}$
+						int b = findNext(pn, k);
+						$\color{blue} \left\{\begin{array}{l} \exists \  n''. \ \texttt{pn} \mapsto (n', n'') \ast \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range})\ \ast\\ 
+						((\texttt{b} = 0 \ast k' = \texttt{k} \ast n'' = n')  \lor 						 (\texttt{b} = 1  \ast \texttt{k} \in \mathsf{range}(n'') \ast \nodeboxrep(n'')))   \end{array}\right\}$
+						if (b == 0){
+								$\color{blue} \left\{\begin{array}{l} 
+								\texttt{pn} \mapsto (n', n') \ast k' = \texttt{k} \ast \cdots    \end{array}\right\}$
+								return 0;
[…]

### 2023-09-06 concurrency/CPP 2024/paper.tex
first pass at explaining maintenance ops

@@
 However, this two-way dependency seems fundamental when the layout of the node structure depends on the concurrency control mechanism we choose.
+For now, we break the abstraction boundary and verify two different versions of \lstinline{insertOp}, one for the lock-coupling template and one for the give-up template
 
@@
 
-\subsection{Templates vs. Internal Reorganization}
+\subsection{Templates and Internal Reorganization}
 \label{internal_reorganize}
-The \lstinline{delete} operation on binary search trees presents difficulties in applying the template approach.
+The \lstinline{delete} operation on binary search trees highlights a non-obvious aspect of the template approach.
 The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}).
-Likewise, in the data structures verified by Krishna et al., deletion can be performed by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation.
+In most of the data structures verified by Krishna et al., deletion can decomposed similarly, by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation.
 When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
@@
 
-It is instructive to compare BST deletion with Krishna et al.'s discussion of splitting in B-link trees.
-They assume the existence of a maintenance thread that constantly crawls the tree searching for full nodes and splitting them.
-The splitting can be defined as a sequential, data-structure-specific operation on a node and its parent; the crawl itself is a variant of the \lstinline{traverse} operation, and while it is not verified in their existing proofs, we expect that it could be defined as a template function.
-We might similarly imagine refactoring a lock-coupling \lstinline{pushdown_left} into a template-level traversal and a data-structure-level rotation operation.
-However, in the BST, rotation involves three nodes (the node to be deleted and both of its children)---and, crucially, \emph{we only know that it involves three nodes once we know the target data structure is a BST}.
-Thus, there is no clear way to define a template function that acquires enough locks (or otherwise performs enough synchronization) to call out to a sequential rotation operation without making the template specific to BSTs.
-
-We believe this represents a fundamental limitation in the template approach.
-Operations in the style of \lstinline{pushdown_left} are ubiquitous in tree structures (e.g., the rebalancing operations of AVL and red-black trees), and always weave together synchronization and restructuring in a way that cannot easily be decomposed.
-This is a serious impediment to the goal of verifying concurrency control separately from the data structure: even if we have a verified template and a verified single-threaded tree, we still need to verify a complex data-structure-specific concurrent function to support deletion, which is often the most complicated part of a concurrent search structure.
-%what's the conclusion here? can we see a way through?
+Krishna et al.
+also discuss an operation of this sort, the \lstinline{split} operation in B-link trees.
+Their approach is to assume the existence of a separate maintenance thread that constantly crawls the tree searching for full nodes and splitting them.
+The split itself is a no-op on the abstract state of the data structure: it makes internal structural changes but does not change the set of keys stored in the tree.
+This complicates the description of the template approach, but such a maintenance operation is necessary for many concurrent data structures.
+In general, we believe that the template approach should be thought of as the combination of \emph{three} kinds of operations: concurrent, data-structure-agnostic template operations; sequential data structure operations; and no-op maintenance operations, which are both concurrency-aware and data-structure-specific, but whose specifications leave the abstract state unchanged.
+For instance, rotation in the BST can be seen as a maintenance operation, and the deletion of a node that has already been rotated to a leaf position is a sequential, data-structure-specific operation.
+Rebalancing operations in AVL and red-black trees would also be considered maintenance operations.
+This can be seen as a limitation of the template approach, since concurrency cannot be completely decoupled from data-structure-specific reasoning, but because maintenance operations are logical no-ops, they should be easier to verify than concurrent data structure operations in general.
+In future work, we hope to provide a formal account of these three categories of operations, and a theorem describing how the proofs of each kind of operation combine to yield a top-level verified data structure.
+This will be essential for applying the template approach to the complexity of real-world high-performance data structure implementations.
 
@@
 While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future.
-The full development of our mechanization effort is available online at
-
-\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.
+The full development of our mechanization effort is available online at [link redacted for anonymity].
+%\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.
 

### 2023-09-06 concurrency/CPP 2024/paper.tex
first attempt at precisely defining a template

@@
 
+%https://tex.stackexchange.com/questions/78187/is-there-another-symbol-that-is-slightly-different-from-forall-likewise-for-e
+\makeatletter
+\newcommand*{\fforall}{%
+  {\mathpalette\fforallAux{}}%
+}
+\newcommand*{\fforallAuxx}[1]{%
+  \sbox0{$\m@th#1\forall$}%
+  \sbox2{%
+    \rlap{%
+      \raisebox{\depth}{$\m@th#1\backslash$}%
+    }%
+    \kern\ht0 %
+  }%
+  \sbox2{\resizebox{\ht2}{\height}{\copy2}}%
+  \sbox2{\resizebox{!}{\ht0}{\copy2}}%
+  \wd2=0pt %
+  \copy2
+  \forall
+}
+\newsavebox\forallBox
+\newdimen\forallLineWidth
+\newdimen\forallSep
+\newcommand*{\fforallAux}[1]{%
+  \sbox\forallBox{$\m@th#1\forall$}%
+  \setlength{\forallLineWidth}{.06\wd\forallBox}%
+  \setlength{\forallSep}{.09\wd\forallBox}%
+  \tikz[
+    inner sep=0pt,
+    line cap=round,
+    line width=\forallLineWidth,
+  ]
+  \draw
+    (0,0) node (A) {\copy\forallBox}
+    (A.south) ++(-\forallSep-\forallLineWidth,.4\forallLineWidth)
+    coordinate (A1)
+    (A.north west) ++(-\forallSep,-\forallLineWidth)
+    coordinate (A2)
+    (A1) -- (A2)
+  ;%
+}
+\makeatother
+
 \definecolor{codegreen}{rgb}{0,0.6,0}
@@
 	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
-	\item We highlight several questions about the boundaries of the template/data structure division, and describe a limitation of the template approach on tree data structures.
+	\item We highlight several questions about the boundaries of the template/data structure division, and clarify an element of the template approach that has not been formally addressed.
 \end{itemize}
@@
 Logical atomicity is a separation logic technique for concisely specifying the behavior of a concurrent operation.
-A logically atomic triple has the form $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+A logically atomic triple has the form $\fforall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
 Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread.
@@
 \begin{mathpar}
-	\forall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \\ \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
+	\fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \\ \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
 \end{mathpar}
@@
 \section{Search Structure Templates in VST}
+
+\subsection{What is a Search Structure Template?}
+Krishna et al.~\cite{templates} described the search structure template approach and used it to verify (parts of) several concurrent data structures.
+Their examples strongly suggest a systematic approach to separating sequential data structure logic from concurrent synchronization patterns.
+However, the approach is not formally defined, and on closer inspection its inputs and outputs are not identical across the example templates.
+In this section, we attempt to precisely describe the pieces of the search structure template approach, and how we can know whether we have successfully verified a data structure given a collection of verified sequential and concurrent functions.
+
+A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}.
+It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads.
+In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the logically atomic specifications shown in Figure~\ref{search-specs}.
+In other words, each operation takes effect atomically on an abstract map $m$, reading and/or updating it as appropriate to the operation.
+
+\begin{figure*}[!ht]
+	\centering
+\begin{mathpar}
+	{\color{blue}
+		\fforall m.\left\langle 
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
+		\right\rangle
+	}
+	\ \texttt{insert(r, k, v)}\ 
+	{\color{blue}
+		\left\langle 
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
+		\right\rangle
+	}
+	\\
+	{\color{blue}
+		\fforall m.\left\langle 
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
+		\right\rangle
+	}
+	\ \texttt{lookup(r, k)}\ 
+	{\color{blue}
+		\left\langle \texttt{v}. \ \
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
+		\right\rangle
+	}
+	\\
+	{\color{blue}
+		\fforall m.\left\langle 
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
+		\right\rangle
+	}
+	\ \texttt{delete(r, k)}\ 
+	{\color{blue}
+		\left\langle 
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
+		\right\rangle
+	}
+\end{mathpar}
+\caption{The logically atomic specifications for search structure operations}
+\label{search-specs}
+\end{figure*}
+
+A search structure template is an implementation of these three functions, written in a particularly restricted way.
+Aside from some top-level management code (e.g., acquiring and releasing an initial lock), all that each function should do is call two functions: a function called \lstinline{traverse}, which is specific to the template and implements its synchronization mechanism, and a concurrency-unaware function (e.g.
+\lstinline{insertOp}) provided by the target data structure.
+(In fact, Krishna et al.
+implemented all three operations as a single function \lstinline{decisiveOp}, parameterized by a function that takes the target operation type as input and performs the appropriate data structure operation; for ease of adaptation to C, we write three different functions, but their structures are identical.) A template consists of an implementation of the top-level operations and the \lstinline{traverse} function used to define them; a data structure consists of an implementation of \lstinline{insertOp}, etc., as well as a function \lstinline{findNext} that is used to implement \lstinline{traverse}.
+The interface between them is a set of standard specifications for \lstinline{insertOp}, etc., and \lstinline{findNext}: if a template's functions can be verified assuming these standard specifications, and a data structure's implementations can be shown to meet those specifications, then the template and the data structure can be combined to yield a verified concurrent search structure.
+
+This is the theory behind the template approach to verification.
+Unfortunately, it is not realized in practice.
+In the Coq development of Krishna et al., different templates use slightly different specifications for data structure operations and \lstinline{findNext}; the specifications proved for data structure implementations do not always match those used by templates (which is possible because they use separate provers for the sequential and concurrent proofs); and some data structures rely on other functions that do not appear in the templates (e.g., maintenance operations that split full nodes in B-trees).
+In the following presentation, we follow the template plan as far as possible, noting the places where we must deviate from the uniform approach; we discuss maintenance operations in further detail in Section~\ref{internal_reorganize}.
 
@@
 	{\color{blue}
-		\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
+		\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
 		\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
@@
 Formally, the lock invariant is defined by
-\begin{align*}\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
-	\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
-	\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
+\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
+	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
+	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
 where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.
@@
 	$\color{blue}
-	\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
+	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
 	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
@@
 	{\color{blue}
-		m.\left\langle 
+		\fforall m.\left\langle 
 		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
@@
 	{\color{blue}
-		\forall \  m. \left\langle
+		\fforall \  m. \left\langle
 		\begin{array}{c}
@@
 In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified C implementations of concurrent data structures. 
-The top-level specifications we obtain for search structure operations are shown below: each operation takes effect on the data structure atomically, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
-\begin{figure*}[!ht]
-	\centering
-\begin{mathpar}
-	{\color{blue}
-		\forall m.\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
-		\right\rangle
-	}
-	\ \texttt{insert(r, k, v)}\ 
-	{\color{blue}
-		\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
-		\right\rangle
-	}
-	\\
-	{\color{blue}
-		\forall m.\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
-		\right\rangle
-	}
-	\ \texttt{lookup(r, k)}\ 
-	{\color{blue}
-		\left\langle \texttt{v}. \ \
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
-		\right\rangle
-	}
-	\\
-	{\color{blue}
-		\forall m.\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
-		\right\rangle
-	}
-	\ \texttt{delete(r, k)}\ 
-	{\color{blue}
-		\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
-		\right\rangle
-	}
-\end{mathpar}
-\end{figure*}
-As we saw in the previous section, these top-level specifications have already been verified at the template level; all we need to do now is provide implementations of data-structure-specific functions (e.g.
+As we saw in the previous section, the top-level specifications of data structure operations have already been verified at the template level; all we need to do now is provide implementations of data-structure-specific functions (e.g.
 \lstinline{findNext}) that satisfy the right specifications to instantiate the template proofs.
@@
 	$\color{blue}
-	\forall \  m. \left\langle
+	\fforall \  m. \left\langle
 	\begin{array}{c}
@@
 		$\color{blue}
-		\forall m.\left\langle 
+		\fforall m.\left\langle 
 		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m

### 2023-09-07 concurrency/CPP 2024/paper.tex
fix breakage

@@
 	$\color{blue}
-	\left\langle \mathit{res}. \ \exists \  n'.
+	\left\langle \mathit{res}. \ \exists \  n', v.
 	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
 		\\ 
-		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
+		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
 	\end{array}
@@
 		lock_t *lock; 
-		int min; 
-		int max;
+		int min, max;
 	} node_t;

### 2023-09-07 concurrency/CPP 2024/paper.tex
fix breakage
edits throughout

--- 2023-09-06 concurrency/CPP 2024/paper.tex
@@
 \newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
-\newcommand{\treerep}{\ensuremath{\mathsf{Node}}}
-\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Node\_ref }}}
+\newcommand{\treerep}{\ensuremath{\mathsf{Abs}}}
+\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Ref }}}
 \newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
-\newcommand{\infp}{\ensuremath{\mathsf{Node\_ref }}}
+\newcommand{\infp}{\ensuremath{\mathsf{Ref }}}
 
@@
 	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
-	\item We highlight several questions about the boundaries of the template/data structure division, and clarify an element of the template approach that has not been formally addressed.
+	\item We give a precise description of search structure templates, and identify places where it is difficult in practice to preserve the boundary between template and data structure.
 \end{itemize}
@@
 The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach.
-Some of the more complex data structures verified include the \lstinline{ConcurrentSkipListMap} from \lstinline{java.util.concurrent} ~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}.
-Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeted at an implementation language (C) rather than a toy language or algorithm-level representation.
+Some of the more complex data structures verified include Java's \lstinline{ConcurrentSkipListMap}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}.
+Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeting an implementation language (C) rather than a toy language or algorithmic representation.
 %do we want to say this part?
@@
 	\end{subfigure}
-	\caption{An example search structure template (lock coupling)}
+	\caption{The lock-coupling search structure template}
 	\label{template-ex}
@@
 Logical atomicity is a separation logic technique for concisely specifying the behavior of a concurrent operation.
-A logically atomic triple has the form $\fforall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+A logically atomic triple has the form $\fforall a.\,\left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
 Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread.
@@
 As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
-$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$. Its concurrent counterpart could be specified as 
+\[\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}\] Its concurrent counterpart could be specified as 
 \begin{mathpar}
-	\fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \\ \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
+	\fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\  \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
 \end{mathpar}
@@
 It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads.
-In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the logically atomic specifications shown in Figure~\ref{search-specs}.
-In other words, each operation takes effect atomically on an abstract map $m$, reading and/or updating it as appropriate to the operation.
-
-\begin{figure*}[!ht]
-	\centering
+In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values:
 \begin{mathpar}
@@
 	}
-	\ \texttt{insert(r, k, v)}\ 
+	\ \texttt{insert(r, k, v)}\ \\
 	{\color{blue}
@@
 	}
-	\\
+\end{mathpar}
+\begin{mathpar}
 	{\color{blue}
@@
 	}
-	\ \texttt{lookup(r, k)}\ 
+	\ \texttt{lookup(r, k)}\ \\
 	{\color{blue}
@@
 	}
-	\ \texttt{delete(r, k)}\ 
+	\ \texttt{delete(r, k)}\ \\
 	{\color{blue}
@@
 \end{mathpar}
-\caption{The logically atomic specifications for search structure operations}
-\label{search-specs}
-\end{figure*}
+In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation.
+Krishna et al.
+use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.
 
@@
 This is the theory behind the template approach to verification.
-Unfortunately, it is not realized in practice.
+Unfortunately, it is not straightforward to realize in practice.
 In the Coq development of Krishna et al., different templates use slightly different specifications for data structure operations and \lstinline{findNext}; the specifications proved for data structure implementations do not always match those used by templates (which is possible because they use separate provers for the sequential and concurrent proofs); and some data structures rely on other functions that do not appear in the templates (e.g., maintenance operations that split full nodes in B-trees).
@@
 
-The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The specification of \texttt{insert} is:
+The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The desired specification of \texttt{insert} is:
 \begin{mathpar}
@@
 	}
-	\\ \texttt{insert(r, k, v)}\  \\
+	\ \texttt{insert(r, k, v)}\  \\
 	{\color{blue}
@@
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
-\end{mathpar}
+\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
+\end{align*}
 Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}).
@@
 \label{internal_reorganize}
-The \lstinline{delete} operation on binary search trees highlights a non-obvious aspect of the template approach.
-The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}).
-In most of the data structures verified by Krishna et al., deletion can decomposed similarly, by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation.
+The \lstinline{delete} operation on binary search trees highlights a limitation of the template approach as we have described it.
+The \lstinline{insert} and \lstinline{lookup} operations can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insertOp} or \lstinline{lookupOp}).
+In most of the prior template examples, deletion can decomposed similarly, by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation.
 When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
@@
 also discuss an operation of this sort, the \lstinline{split} operation in B-link trees.
-Their approach is to assume the existence of a separate maintenance thread that constantly crawls the tree searching for full nodes and splitting them.
+Their approach is to assume the existence of a separate maintenance thread that constantly crawls the tree searching for full nodes and splitting them; any operation that cannot take effect on full nodes can simply retry until the target node is split.
 The split itself is a no-op on the abstract state of the data structure: it makes internal structural changes but does not change the set of keys stored in the tree.
-This complicates the description of the template approach, but such a maintenance operation is necessary for many concurrent data structures.
-In general, we believe that the template approach should be thought of as the combination of \emph{three} kinds of operations: concurrent, data-structure-agnostic template operations; sequential data structure operations; and no-op maintenance operations, which are both concurrency-aware and data-structure-specific, but whose specifications leave the abstract state unchanged.
+Maintenance operations of this sort appear in almost every reasonably complex concurrent data structure, and so the template approach must account for their existence.
+In other words, in general the template approach must divide the operations of a target data structure into three categories: concurrent, data-structure-agnostic template operations, which can be verified once and applied to multiple data structures; sequential data structure operations, which can be verified once and plugged into multiple templates; and no-op maintenance operations, which are both concurrency-aware and data-structure-specific, but whose specifications leave the abstract state unchanged.
 For instance, rotation in the BST can be seen as a maintenance operation, and the deletion of a node that has already been rotated to a leaf position is a sequential, data-structure-specific operation.
 Rebalancing operations in AVL and red-black trees would also be considered maintenance operations.
-This can be seen as a limitation of the template approach, since concurrency cannot be completely decoupled from data-structure-specific reasoning, but because maintenance operations are logical no-ops, they should be easier to verify than concurrent data structure operations in general.
-In future work, we hope to provide a formal account of these three categories of operations, and a theorem describing how the proofs of each kind of operation combine to yield a top-level verified data structure.
-This will be essential for applying the template approach to the complexity of real-world high-performance data structure implementations.
+The existence of this third category means that we cannot completely separate concurrency reasoning from data structure reasoning, but because maintenance operations are logical no-ops, they should be easier to verify in a non-decomposed style than most concurrent data structure operations.
 

### 2023-09-07 concurrency/CPP 2024/paper.tex

@@
 	$\color{blue}
-	\left\langle \mathit{res}. \ \exists \  n'.
+	\left\langle \mathit{res}. \ \exists \  n', v.
 	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
 		\\ 
-		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
+		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
 	\end{array}

### 2023-09-11 concurrency/CPP 2024/paper.tex
clean up todos

@@
 The next step is to implement templates as a truly generic framework, with proofs of \lstinline{traverse}, etc.
-that can be freely combined with verified sequential data structures via a clear interface that defines the functions a data structure must provide and the assumptions the template makes on them.
+that can be freely combined with verified sequential data structures via a clear interface that defines the functions a data structure must provide and the assumptions the template makes on them (e.g., by verifying template functions with respect to a data-structure typeclass).
 Ultimately, we hope to use the template approach to verify data structures such as Masstree~\cite{masstree} and Wormhole~\cite{wormhole}, complicated real-world search structures that combine multiple concurrency patterns for maximum performance on multicore architectures.

### 2023-09-12 concurrency/CPP 2024/paper.tex
Shengyi's comments, more reviewer comments

@@
 The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency.
-In theory, this could allow us to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns, and immediately obtain $n \times m$ verified concurrent data structures.
+In theory, this could allow us to prove the correctness of $n$ concurrency patterns and $m$ (single-threaded) data structure implementations, and immediately obtain $n \times m$ verified concurrent data structures.
 In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.
@@
 \end{mathpar}
+
 \begin{mathpar}
@@
 \begin{mathpar}
-	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}\\
+	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}
+	\end{mathpar}
+	\begin{mathpar}
 	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{release}(n\texttt{->lock})\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m\right\rangle}
@@
 The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).
+Then $\treerep\ m$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.
 
@@
 Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}.
+Our definition of $\treerep$ for the give-up template closely follows that of Krishna et al.
 
@@
 While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future.
+
+Compared to prior template proofs in Iris and GRASShopper, our proof development is significantly larger (although direct comparison is difficult, since GRASShopper is an automatic prover).
+The biggest structural difference is that we need to directly state loop invariants for our imperative code, while the Iris proofs handle recursion with a simpler Löb-induction approach.
+Coming up with the loop invariants was the hardest part of the proofs.
+Our proofs also deal with the details of real C code: proving absence of integer overflows, accurately representing struct-and-pointer-based data structures instead of functional-programming-style structured data, etc.
+In exchange for this extra effort, the code we verify constitutes complete C programs that can be compiled and executed.
 The full development of our mechanization effort is available online at [link redacted for anonymity].

### 2023-09-18 concurrency/CPP 2024/paper.tex
link zenodo for artifact
rephrase flow interface mention in intro

@@
 \begin{itemize}
-	\item We reimplement the template approach without flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/subtree''.
+	\item We reimplement the template approach independently of flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/substructure''.
 	\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.

### 2023-11-22 concurrency/CPP 2024/paper.tex
initial revise

@@
 %% The first command in your LaTeX source must be the \documentclass command.
-\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %%
@@
 \email{dnguye96@uic.edu}
-\orcid{}
+\orcid{https://orcid.org/0000-0002-6810-897X}
 \affiliation{%
@@
   \city{Chicago}
-  \state{Illinois}
+  \state{IL}
+  \country{USA}
+  \postcode{}
+}
+\author{Lennart Beringer}
+%\authornote{Both authors contributed equally to this research.}
+\email{eberinge@cs.princeton.edu}
+\orcid{https://orcid.org/0000-0002-1570-3492}
+\affiliation{%
+	\institution{Princeton University}
+	\streetaddress{}
+	\city{Princeton}
+	\state{NJ}
+	\country{USA}
+	\postcode{}
+}
+
+\author{William Mansky}
+%\authornote{Both authors contributed equally to this research.}
+\email{mansky1@uic.edu}
+\orcid{https://orcid.org/0000-0002-5351-895X}
+\affiliation{%
+	\institution{University of Illinois Chicago}
+	\streetaddress{}
+	\city{Chicago}
+	\state{IL}
+	\country{USA}
+	\postcode{}
+}
+
+\author{Shengyi Wang}
+%\authornote{Both authors contributed equally to this research.}
+\email{shengyiw@princeton.edu}
+\orcid{https://orcid.org/0000-0002-2286-8703}
+\affiliation{%
+	\institution{Princeton University}
+	\streetaddress{}
+	\city{Princeton}
+	\state{NJ}
 	\country{USA}
@@
 		\fforall m.\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
 		\right\rangle
@@
 		\left\langle \texttt{v}. \ \
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m) \land m(\texttt{k}) = \texttt{v}
 		\right\rangle
@@
 \begin{mathpar}
-	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}
+	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
 	\end{mathpar}
 	\begin{mathpar}
-	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{release}(n\texttt{->lock})\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m\right\rangle}
+	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ \\ \texttt{release}(n\texttt{->lock})\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ (m)\right\rangle}
 \end{mathpar}
@@
 		\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
-		\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
+		\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
 		\right\rangle
@@
 The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).
-Then $\treerep\ m$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.
+Then $\treerep\ (m)$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.
 
@@
 	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
-	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
+	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
 	\right\rangle$
@@
 	\end{array}
-	\ \Bigg| \ \treerep\ m \
+	\ \Bigg| \ \treerep\ (m) \
 	\right\rangle$
@@
 		\begin{array}{c}
-			\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
+			\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ (m)
 		\end{array}
@@
 		\end{array}
-		\Bigg| \ \treerep\ m \
+		\Bigg| \ \treerep\ (m) \
 		\right\rangle
@@
 In exchange for this extra effort, the code we verify constitutes complete C programs that can be compiled and executed.
-The full development of our mechanization effort is available online at [link redacted for anonymity].
+The full development of our mechanization effort is available online at \href{https://zenodo.org/record/8337004}{\color{blue}{https://zenodo.org/record/8337004}}.
+
+%[link redacted for anonymity].
+
 %\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.
@@
 	\begin{array}{c}
-		\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
+		\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ (m)
 	\end{array}
@@
 		\fforall m.\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
+		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
 		\right\rangle$

### 2023-11-22 concurrency/CPP 2024/paper.tex
progress in revising

@@
 %% The first command in your LaTeX source must be the \documentclass command.
-\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=true,printacmref=true}
 %%
@@
 %% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Conference acronym 'XX]{Make sure to enter the correct
-  conference title from your rights confirmation email}{June 03--05,
-  2018}{Woodstock, NY}
+\acmConference[CPP '23]{Make sure to enter the correct
+  conference title from your rights confirmation email}{Jan 15--16,
+  2024}{Woodstock, NY}
 \acmPrice{15.00}
@@
 %% used to denote shared contribution to the research.
+
 \author{Duc Than Nguyen}
-%\authornote{Both authors contributed equally to this research.}
 \email{dnguye96@uic.edu}
 \orcid{https://orcid.org/0000-0002-6810-897X}
-\affiliation{%
-  \institution{University of Illinois Chicago}
-  \streetaddress{}
-  \city{Chicago}
-  \state{IL}
-  \country{USA}
-  \postcode{}
-}
+\affiliation{\institution{University of Illinois at Chicago}\streetaddress{}\city{Chicago}\state{IL}\country{USA}\postcode{}}
+
 \author{Lennart Beringer}
-%\authornote{Both authors contributed equally to this research.}
 \email{eberinge@cs.princeton.edu}
 \orcid{https://orcid.org/0000-0002-1570-3492}
-\affiliation{%
-	\institution{Princeton University}
-	\streetaddress{}
-	\city{Princeton}
-	\state{NJ}
-	\country{USA}
-	\postcode{}
-}
+\affiliation{\institution{Princeton University}\streetaddress{}\city{Princeton}\state{NJ}\country{USA}\postcode{}}
 
 \author{William Mansky}
-%\authornote{Both authors contributed equally to this research.}
 \email{mansky1@uic.edu}
 \orcid{https://orcid.org/0000-0002-5351-895X}
-\affiliation{%
-	\institution{University of Illinois Chicago}
-	\streetaddress{}
-	\city{Chicago}
-	\state{IL}
-	\country{USA}
-	\postcode{}
-}
+\affiliation{\institution{University of Illinois at Chicago}\streetaddress{}\city{Chicago}\state{IL}\country{USA}\postcode{}}
 
 \author{Shengyi Wang}
-%\authornote{Both authors contributed equally to this research.}
 \email{shengyiw@princeton.edu}
 \orcid{https://orcid.org/0000-0002-2286-8703}
-\affiliation{%
-	\institution{Princeton University}
-	\streetaddress{}
-	\city{Princeton}
-	\state{NJ}
-	\country{USA}
-	\postcode{}
-}
+\affiliation{\institution{Princeton University}\streetaddress{}\city{Princeton}\state{NJ}\country{USA}\postcode{}}
 
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{}
+\keywords{Concurrent Separation Logic, Fine-grained Locking, Iris, Logical Atomicity, Theorem proving, Verified Software Toolchain}
 
@@
 	\end{subfigure}\qquad
-	\begin{subfigure}[t]{0.48\textwidth}
+	\begin{subfigure}[t]{0.45\textwidth}
 		\begin{lstlisting}[language=caml, style=myStyle]
@@
 As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
-\[\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}\] Its concurrent counterpart could be specified as 
+$$
+ {\color{blue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(s, v)\  {\color{blue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
 \begin{mathpar}
-	\fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\  \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
+	{\color{blue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ } 
+	 \vspace{-0.8em}  \\ \texttt{push}(s, v)\   \vspace{-0.8em}  \\ 
+	{\color{blue} \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle }
 \end{mathpar}
@@
 		\right\rangle
-	}
-	\ \texttt{insert(r, k, v)}\ \\
+	} \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
 	{\color{blue}
@@
 	}
-\end{mathpar}
-\begin{mathpar}
+\\
 	{\color{blue}
@@
 		\right\rangle
-	}
-	\ \texttt{lookup(r, k)}\ \\
+	} \vspace{-0.8em}
+	\\ \texttt{lookup(r, k)}\  \vspace{-0.8em} \\
 	{\color{blue}
@@
 	}
-\end{mathpar}
-
-\begin{mathpar}
+	\\
 	{\color{blue}
@@
 		\right\rangle
-	}
-	\ \texttt{delete(r, k)}\ \\
+	}  \vspace{-0.8em} 
+	\\ \texttt{delete(r, k)}\ \vspace{-0.8em}\\
 	{\color{blue}
@@
 		void *value; 
-		struct node_t *left, *right;} 
-	node;
+		struct node_t *left, *right;
+	} node;
 \end{lstlisting}
@@
 \begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
-	typedef struct pn {struct node_t *p; struct node_t *n;} pn;
+	typedef struct pn {
+		struct node_t *p; struct node_t *n;
+	} pn;
 \end{lstlisting}
@@
 \begin{mathpar}
-	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
-	\end{mathpar}
-	\begin{mathpar}
-	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ \\ \texttt{release}(n\texttt{->lock})\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ (m)\right\rangle}
+	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
+	 \vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
+\\
+	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ 
+	 \vspace{-0.8em} \\ \texttt{release}(n\texttt{->lock})\  \vspace{-0.8em} \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ (m)\right\rangle}
 \end{mathpar}
@@
 			\\ 
-			(\mathsf{if} \ \mathit{res} \ \\ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-			\\ \mathsf{else} \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right))
-		\end{array}
-		\ \Bigg | \ \treerep\ m \
+			\mathsf{if} \ \mathit{res} \  \ \mathsf{then} \  \\ \ \ \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+			\\ \mathsf{else} \\ \ \ \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right)
+		\end{array}
+		\ \Bigg | \treerep\ (m) \
 		\right\rangle
@@
 		\\ 
-		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
-	\end{array}
-	\ \Bigg| \ \treerep\ (m) \
+		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
+	\end{array}
+	\ \Bigg| \treerep\ (m) \
 	\right\rangle$
@@
 	\right\rangle
-	}
-	\ \texttt{insert(r, k, v)}\  \\
+} \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
 {\color{blue}
@@
 		\right\}
-	}
-	\vspace{-6pt}\\ \vspace{-6pt}
-	\texttt{inRange(pn, k)} 
-	\\
+	} \vspace{-0.85em} 
+	\\ \texttt{inRange(pn, k)} 
+	\vspace{-0.85em}  \\
 	{\color{blue}
@@
 			\right\}
-		}
+		} \vspace{-0.85em} 
 		\\ 
 		\texttt{findNext(pn, k)} 
+		\vspace{-0.85em} 
 		\\
@@
 				\right\}
-			}
+			} \vspace{-0.85em} 
 			\\ 
 			\texttt{insertOp(pn, k, v)} 
+			\vspace{-0.85em} 
 			\\
@@
 				\right\}
-			}
+			} \vspace{-0.85em} 
 			\\
 			\texttt{insertOp(pn, k, v)}
+			\vspace{-0.85em} 
 			\\
@@
 The \lstinline{insert} and \lstinline{lookup} operations can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insertOp} or \lstinline{lookupOp}).
-In most of the prior template examples, deletion can decomposed similarly, by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation.
+In most of the prior template examples, deletion can be decomposed similarly, by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation.
 When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
@@
 
-\ignore{
 \begin{acks}
-Thanks to Roshan Sharma, Alex Oey, and Anastasiia Evdokimova for extensive work on the original binary search tree implementation and verification.
+We thank Roshan Sharma, Alex Oey, and Anastasiia Evdokimova for extensive work on the original binary search tree implementation and verification, and the anonymous reviewers for providing comprehensive and insightful reviews.
+This research was supported, by ... 
 \end{acks}
-}
 

### 2023-11-22 concurrency/CPP 2024/paper.tex
fix breakage
add CSS concepts

@@
 \usepackage{parcolumns}
-\usepackage{iris}
+%\usepackage{iris}
 %\usepackage{lstlangcoq}
@@
 
-\author{Duc Than Nguyen}
+\author{Duc-Than Nguyen}
 \email{dnguye96@uic.edu}
@@
 %%
+
+\begin{CCSXML}
+	<ccs2012>
+	<concept>
+	<concept_id>10011007.10011074.10011099.10011692</concept_id>
+	<concept_desc>Software and its engineering~Formal software verification</concept_desc>
+	<concept_significance>500</concept_significance>
+	</concept>
+	</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Formal software verification}
 

### 2023-11-22 concurrency/CPP 2024/paper.tex
add some acm format properties

@@
 %% The first command in your LaTeX source must be the \documentclass command.
-\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=true,printacmref=true}
+\documentclass[sigplan,10pt, screen]{acmart}
+\settopmatter{}
 %%
 %% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
+%\AtBeginDocument{%
+%  \providecommand\BibTeX{{%
+%    Bib\TeX}}}
+\startPage{1}
+\bibliographystyle{ACM-Reference-Format}
 
@@
 %% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
+\setcopyright{rightsretained}
+\acmPrice{}
 \acmDOI{XXXXXXX.XXXXXXX}
-
+\acmYear{2024}
+\copyrightyear{2024}
+\acmSubmissionID{}
+\acmISBN{}
 %% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[CPP '23]{Make sure to enter the correct
-  conference title from your rights confirmation email}{Jan 15--16,
-  2024}{Woodstock, NY}
-\acmPrice{15.00}
-\acmISBN{978-1-4503-XXXX-X/18/06}
+\acmConference[CPP '24]{Proceedings of the 13th ACM SIGPLAN International Conference on Certified Programs and Proofs}{Jan 15--16, 2024}{London, United Kingdom}
+\acmBooktitle{Proceedings of the 13th ACM SIGPLAN International Conference on Certified Programs and Proofs (CPP '24), Jan 15--16, 2024, London, United Kingdom}
 
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{Concurrent Separation Logic, Fine-grained Locking, Iris, Logical Atomicity, Theorem proving, Verified Software Toolchain}
+\keywords{Concurrent separation logic, fine-grained locking, Iris, logical atomicity, theorem proving, Verified Software Toolchain}
 

### 2023-11-23 concurrency/CPP 2024/paper.tex
resize table
first try to modify traverse spec

@@
 \usepackage{parcolumns}
-%\usepackage{iris}
+\usepackage{scalerel}
 %\usepackage{lstlangcoq}
@@
 \begin{mathpar}
-	% spec of inRange
-	{\color{blue}
-		\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
-		\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
-		\right\rangle
-	}\vspace{-6pt}\\
+{ \color{blue}
+	\fforall \  m. \ \scaleleftright[1.50ex]{\langle}
+	{\begin{array}{l}
+			\texttt{pn} \mapsto (p, n) \ \ast \ 
+			\nodeboxrep(n)  \ast \ \mathsf{R}(n) 
+		\end{array}  
+		\ \Bigg \vert  \treerep\ (m)
+	}  
+	{\rangle}
+}
+\vspace{-6pt}\\
 \vspace{-6pt}\texttt{traverse(pn, k)} \\
-	{\color{blue}
-		\left\langle \mathit{res}. \ \exists \  n', v.
-		\begin{array}{l} \texttt{pn} \mapsto \left(n', n'\right) \ \ast \\ \nodeboxrep\left(n'\right) \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
-			\\ 
-			\mathsf{if} \ \mathit{res} \  \ \mathsf{then} \  \\ \ \ \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-			\\ \mathsf{else} \\ \ \ \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right)
-		\end{array}
-		\ \Bigg | \treerep\ (m) \
-		\right\rangle
+{ \color{blue}
+\scaleleftright[1.50ex]{\langle}
+{\begin{array}{l}
+		\mathit{res}. \ \exists \  n', v, \mathsf{range}. \\ 
+		\ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
+		\\ 
+		\ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+		\\ 
+		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right)
+\end{array}  
+\Bigg \vert \treerep\ (m)
+}  
+{\rangle}
 }
 \end{mathpar}
+
 The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}.
@@
 \begin{table}[h]
-	\begin{center}
-		\begin{tabular}{l | l || l | l } 
+	\centering
+	\resizebox{\columnwidth}{!}{
+		\begin{tabular}{l | l || l | l} 
 			\hline
@@
 			\hline
-			Specfications & 190 & Specfications & 138 \\
+			Specifications & 190 & Specifications & 138 \\
 			\texttt{findNext} proof & 27 & \texttt{findNext} proof & 27 \\
@@
 			\texttt{delete} proof & 182 & & \\
-			supporting proofs    & 	1909	 & supporting proofs 	& 1876 \\
+			Supporting proofs & 1909 & Supporting proofs & 1876 \\
 			\hline
@@
 		\end{tabular}
+	}
 	\caption{Size of Coq definitions and proofs, by topic.}
 	\label{table_coq}
-	\end{center}
 \end{table}
@@
 In exchange for this extra effort, the code we verify constitutes complete C programs that can be compiled and executed.
-The full development of our mechanization effort is available online at \href{https://zenodo.org/record/8337004}{\color{blue}{https://zenodo.org/record/8337004}}.
+The full development of our mechanization effort is available online at \href{https://zenodo.org/record/8337004}{https://zenodo.org/record/8337004}.
 

### 2023-11-23 concurrency/CPP 2024/paper.tex
a bit hack for traverse spec

@@
 
-We prove correctness of the template by showing that \lstinline{traverse}, \lstinline{insert}, etc.
+We prove correctness of the template by showing that \lstinline{traverse}, \\ \lstinline{insert}, etc.
 meet logically atomic specifications describing their effects on the data structure.
@@
 { \color{blue}
-	\fforall \  m. \ \scaleleftright[1.50ex]{\langle}
+	\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
 	{\begin{array}{l}
@@
 		\end{array}  
-		\ \Bigg \vert  \treerep\ (m)
+		\Big \vert  \treerep\ (m)
 	}  
@@
 { \color{blue}
-\scaleleftright[1.50ex]{\langle}
+ \scaleleftright[1.4ex]{\langle}
 {\begin{array}{l}
-		\mathit{res}. \ \exists \  n', v, \mathsf{range}. \\ 
-		\ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
-		\\ 
-		\ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-		\\ 
-		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right)
+		\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
+		\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
+		\\ 
+		\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+		\\ 
+		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
 \end{array}  
@@
 		\begin{array}{c}
-			\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2
+			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2
 		\end{array}
@@
 	{\color{blue}
-		\left\{\mathit{res.} \ 
+		\left\{\mathit{res.} 
 		\begin{array}{l}
-			\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2 \ \ast  \ 
-			\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2))
+			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2 \ \ast  \ 
+			\\\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2)
 		\end{array}
@@
 \begin{mathpar}
-	{\color{blue}
-		\fforall \  m. \left\langle
-		\begin{array}{c}
-			\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ (m)
-		\end{array}
-		\right\rangle
-	}
-	\vspace{-6pt}\\ \vspace{-6pt}
-	\texttt{traverse(pn, k)} 
-	\\
-	{\color{blue}
-		\left\langle \mathit{res}. \ \exists \  n', v.
-		\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \\ \texttt{k} \in \mathsf{range}\ \ast\ 
-			\\ 
-			(\mathsf{if} \ \mathit{res} \\ 
-			\ \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-			\\ 
-			\ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
-		\end{array}
-		\Bigg| \ \treerep\ (m) \
-		\right\rangle
+	{ \color{blue}
+		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
+		{\begin{array}{l}
+				\texttt{pn} \mapsto (p, n) \ \ast \ 
+				\nodeboxrep(n)   
+			\end{array}  
+			\Big \vert  \treerep\ (m)
+		}  
+		{\rangle}
+	}
+	\vspace{-6pt}\\
+	\vspace{-6pt}\texttt{traverse(pn, k)} \\
+	{ \color{blue}
+		\scaleleftright[1.4ex]{\langle}
+		{\begin{array}{l}
+				\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
+				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range} \ \ast \ 
+				\\ 
+				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
+				\\ 
+				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
+			\end{array}  
+			\Bigg \vert \treerep\ (m)
+		}  
+		{\rangle}
 	}
 \end{mathpar}
+
 As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents.
@@
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
+\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast  \infp (r) \ast \infp (n)
 \end{align*}
@@
 				\texttt{pn->n->t}\mapsto (k', v', l, r)  \ \ast \\\  
-				(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k'))
+				\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k')
 			\end{array}
@@
 					\ \texttt{pn->n->t} \mapsto (\texttt{k}, \texttt{v}, l, r) \ \ast \ \\ 
-					\ l \mapsto (\texttt{NULL}, \mathit{lk}_1)\  \ast \ r \mapsto (\texttt{NULL}, \mathit{lk}_2)
+					\ l \mapsto (\texttt{NULL}, \mathit{lk}_1)  \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2)
 				\end{array}
@@
 				\begin{array}{c}
-					\texttt{pn->n}\mapsto \texttt{NULL} \  \ast \ \texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2
+					\texttt{pn->n}\mapsto \texttt{NULL}  \ast  \texttt{pn->n->min}\mapsto n_1  \ast  \texttt{pn->n->max}\mapsto n_2
 				\end{array}
@@
 				\begin{array}{c}
-					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r) \  \ast \ \texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2  \ \ast \ 
-					\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ \ast \ r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
+					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r)  \ast  \texttt{pn->n->min}\mapsto n_1 \ast  \texttt{pn->n->max}\mapsto n_2  \ \ast \ 
+					\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
 				\end{array}
@@
 	\begin{array}{c}
-		\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ (m)
+		\texttt{pn} \mapsto (p, n)  \ast  \nodeboxrep(n)  \big| \ \treerep\ (m)
 	\end{array}
@@
 	\left\langle \mathit{res}. \ \exists \  n', v.
-	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
+	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
 		\\ 

### 2023-11-24 concurrency/CPP 2024/paper.tex
WIP

@@
 	commentstyle=\color{codegreen},
+	keywordstyle=\bfseries\color{green!40!black},
+	commentstyle=\itshape\color{purple!40!black},
+	stringstyle=\color{codepurple},
 	basicstyle=\ttfamily\footnotesize,
@@
 	showtabs=false,                  
-	tabsize=1,
+	tabsize=2,
 }
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{Concurrent separation logic, fine-grained locking, Iris, logical atomicity, theorem proving, Verified Software Toolchain}
+\keywords{concurrent separation logic, fine-grained locking, Iris, logical atomicity, interactive theorem proving, Verified Software Toolchain}
 
@@
 	\begin{subfigure}[t]{0.45\textwidth}
-		\begin{lstlisting}[language=caml, style=myStyle]
-			let insert r k =
-			  lockNode r;
-			  let n = traverse r k in
-			  let res = insertOp n k in
-			  unlockNode n; 
-			  res\end{lstlisting}
+		\lstinputlisting[language=caml, style=myStyle]{insert.ml} 
+	\end{subfigure}
 		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
 		%		\label{traverse_lock_b}
-	\end{subfigure}
 	\caption{The lock-coupling search structure template}
@@
 \end{figure}
-
-\begin{figure*}
-
-\end{figure*}
-
 Figure~\ref{template-ex} shows an example search structure template.
@@
 We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
+\begin{lstlisting}[style=myStyle,language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, numbers=none, xleftmargin=0.5em]
 	typedef struct node_t {node *t; lock_t *lock;} node_t;
@@
 Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
+\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
 	typedef struct node {
@@
 while a linked list might use
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
+\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
 	typedef struct node {
@@
 The C implementation uses a struct
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
+\begin{lstlisting}[style=myStyle, language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, numbers=none, xleftmargin=0.5em]
 	typedef struct pn {
-		struct node_t *p; struct node_t *n;
+		struct node_t *p; 
+		struct node_t *n;
 	} pn;
@@
 		\begin{subfigure}[t]{\textwidth}
-			\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
+			\lstinputlisting[style=myStyle]{lock_insert.c} 
 			\caption{The \lstinline{insert} method of the lock-coupling template algorithm written in C}
@@
 	\renewcommand{\thesubfigure}{b}% New fixed/manual numbering
-	\begin{subfigure}[t]{0.48\textwidth}
+	\begin{subfigure}[t]{0.44\textwidth}
 		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
@@
 \begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
-	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
+	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ast \mathsf{node\_data}(n, c) \\
 	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
@@
 \end{align*}
-
 \begin{figure*}[!ht]
@@
 		\begin{subfigure}[t]{\textwidth}
-			\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
+			\lstinputlisting[style=myStyle]{giveup_insert.c} 
 			\caption{The \lstinline{insert} method of the give-up template algorithm written in C}
@@
 	\renewcommand{\thesubfigure}{b}% New fixed/manual numbering
-	\begin{subfigure}[t]{0.48\textwidth}
+	\begin{subfigure}[t]{0.44\textwidth}
 		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
@@
 Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+\begin{lstlisting}[style=myStyle,  numbers=none, xleftmargin=0.5em]
 	typedef struct node_t {
@@
 		\left\{\mathit{res.} 
-		\begin{array}{l}
-			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2 \ \ast  \ 
-			\\\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2)
+		\begin{array}{c}
+			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2 \ \ast  
+			\\ \mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2)
 		\end{array}
@@
 \label{traverse_proof_giveup}
-
 The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
@@
 		{\begin{array}{l}
-				\texttt{pn} \mapsto (p, n) \ \ast \ 
+				\texttt{pn} \mapsto (p, n) \ast  
 				\nodeboxrep(n)   
@@
 				\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
-				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range} \ \ast \ 
+				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast \texttt{k} \in \mathsf{range} \ \ast 
 				\\ 
@@
 As expected, their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples (see Figures \ref{fig:findNext_lock} and \ref{fig:insertOp}), and their proofs do not require any concurrency reasoning.
-
 \begin{figure*}[!ht]
@@
 The contents of a BST node are defined in C as:
-\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
+\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
 	typedef struct node {
@@
 The proof is straightforward.
-
 \begin{figure}[ht]
+	\begin{minipage}{0.45\textwidth}
 	\lstinputlisting[language=C, style=myStyle]{findnext.c} 
+	\end{minipage}
 	\caption{C implementation of \lstinline{findNext} for the binary search tree (all templates)}
@@
 The complete proofs can be found in our Coq mechanization (see Section \ref{proof_mech}).
-
-\begin{figure*}[!ht]
+\begin{figure*}[h]
 	$\color{blue}
@@
 
-\begin{figure*}[!ht]
+\begin{figure*}[h]
 	\begin{subfigure}{\textwidth}
+		\centering
 		$\color{blue}
-		\fforall m.\left\langle 
+		\fforall \  m. \left\langle
+		\begin{array}{c}
 			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
+		\end{array}
 		\right\rangle$
@@
 		\left\langle
+		\begin{array}{c}
 			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
+		\end{array}
 		\right\rangle$
@@
 	\end{subfigure}\qquad
+
 	\begin{subfigure}{\textwidth}
+		\centering
 		$\color{blue}
-		\fforall m.\left\langle 
+		\fforall \  m. \left\langle
+		\begin{array}{c}
 			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
+		\end{array}
 		\right\rangle$
@@
 \end{figure*}
-
 \end{document}

### 2023-11-24 concurrency/CPP 2024/paper.tex
WIP

@@
 
+\definecolor{specblue}{rgb}{0.0,0.3,0.55}
+
 \definecolor{codegreen}{rgb}{0,0.6,0}
@@
 	backgroundcolor=\color{white}, 
-	commentstyle=\color{codegreen},
 	keywordstyle=\bfseries\color{green!40!black},
-	commentstyle=\itshape\color{purple!40!black},
+	commentstyle=\itshape\color{violet!40!black},
 	stringstyle=\color{codepurple},
@@
 $$
- {\color{blue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(s, v)\  {\color{blue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
+ {\color{specblue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(s, v)\  {\color{specblue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
 \begin{mathpar}
-	{\color{blue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ } 
+	{\color{specblue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ } 
 	 \vspace{-0.8em}  \\ \texttt{push}(s, v)\   \vspace{-0.8em}  \\ 
-	{\color{blue} \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle }
+	{\color{specblue} \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle }
 \end{mathpar}
@@
 \begin{mathpar}
-	{\color{blue}
+	{\color{specblue}
 		\fforall m.\left\langle 
@@
 	} \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
-	{\color{blue}
+	{\color{specblue}
 		\left\langle 
@@
 \\
-	{\color{blue}
+	{\color{specblue}
 		\fforall m.\left\langle 
@@
 	\\ \texttt{lookup(r, k)}\  \vspace{-0.8em} \\
-	{\color{blue}
+	{\color{specblue}
 		\left\langle \texttt{v}. \ \
@@
 	\\
-	{\color{blue}
+	{\color{specblue}
 		\fforall m.\left\langle 
@@
 	\\ \texttt{delete(r, k)}\ \vspace{-0.8em}\\
-	{\color{blue}
+	{\color{specblue}
 		\left\langle 
@@
 \begin{mathpar}
-	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
-	 \vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
-\\
-	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ 
-	 \vspace{-0.8em} \\ \texttt{release}(n\texttt{->lock})\  \vspace{-0.8em} \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ (m)\right\rangle}
+	{\color{specblue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
+	 \vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
+\\
+	{\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ 
+	 \vspace{-0.8em} \\ \texttt{release}(n\texttt{->lock})\  \vspace{-0.8em} \\ {\color{specblue}\left\langle \nodeboxrep(n) \ |\ \treerep\ (m)\right\rangle}
 \end{mathpar}
@@
 \begin{mathpar}
-{ \color{blue}
+{ \color{specblue}
 	\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
@@
 \vspace{-6pt}\texttt{traverse(pn, k)} \\
-{ \color{blue}
+{ \color{specblue}
  \scaleleftright[1.4ex]{\langle}
@@
 \begin{figure*}[!ht]
-	$\color{blue}
+	$\color{specblue}
 	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
@@
 	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
-	$\color{blue}
+	$\color{specblue}
 	\left\langle \mathit{res}. \ \exists \  n', v.
@@
 \begin{mathpar}
-{\color{blue}
+{\color{specblue}
 	\fforall m.\left\langle 
@@
 } \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
-{\color{blue}
+{\color{specblue}
 	\left\langle 
@@
 	% inRange for give-up
-	{\color{blue}
+	{\color{specblue}
 		\left\{ 
@@
 	\vspace{-0.85em}  \\
-	{\color{blue}
+	{\color{specblue}
 		\left\{\mathit{res.} 
@@
 \begin{mathpar}
-	{ \color{blue}
+	{ \color{specblue}
 		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
@@
 	\vspace{-6pt}\texttt{traverse(pn, k)} \\
-	{ \color{blue}
+	{ \color{specblue}
 		\scaleleftright[1.4ex]{\langle}
@@
 	\begin{mathpar}
-		{\color{blue}
+		{\color{specblue}
 			\left\{ 
@@
 		\\
-		{\color{blue}
+		{\color{specblue}
 			\left\{\mathit{res}. \ \exists \  n'.
@@
 		\begin{mathpar}
-			{\color{blue}
+			{\color{specblue}
 				\left\{ 
@@
 			\\
-			{\color{blue}
+			{\color{specblue}
 				\left\{\exists \  l\ r\ \mathit{lk}_1\ \mathit{lk}_2.
@@
 			\\
-			{\color{blue}
+			{\color{specblue}
 				\left\{\exists \ l\ r\ \mathit{lk}_1\ \mathit{lk}_2. 
@@
 \begin{figure*}[h]
-	$\color{blue}
+	$\color{specblue}
 	\fforall \  m. \left\langle
@@
 	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
-	$\color{blue}
+	$\color{specblue}
 	\left\langle \mathit{res}. \ \exists \  n', v.
@@
 		\centering
-		$\color{blue}
+		$\color{specblue}
 		\fforall \  m. \left\langle
@@
 		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
-		$\color{blue}
+		$\color{specblue}
 		\left\langle
@@
 		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
-		$\color{blue}
+		$\color{specblue}
 		\left\langle

### 2023-11-24 concurrency/CPP 2024/proof_giveup_insert.c

@@ -1,23 +1,23 @@
 void insert (node_t **r, int k, void *v){
 		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
 		pn->n = *r;
-		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \infp(\texttt{r}) \end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \infp(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
+				$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
+				$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
 				//Linearization point
 		}
 		else{
-				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \infp(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$
[…]

### 2023-11-24 concurrency/CPP 2024/proof_giveup_traverse.c

@@ -1,40 +1,40 @@
 int traverse(pn *pn, int k){
-		node_t *r = (pn->n); $\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  
+		node_t *r = (pn->n); $\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  
 \infp (\texttt{r}) \end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
+		for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
 			\exists\ n'.\ \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \infp (\texttt{r})
 		 \end{array}\right\}$
 				acquire(pn->n->lock);
-				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
-				pn->p = pn->n; $\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
+				$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
+				pn->p = pn->n; $\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
 				if (inRange(pn, k) == 1){
[…]

### 2023-11-24 concurrency/CPP 2024/proof_lock_insert.c

@@ -1,25 +1,25 @@
 void insert (node_t **r, int k, void *v){
 		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
 		pn->n = *r;
-		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \end{array}\right\}$
 		acquire(pn->n->lock);
-		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
 		if (traverse(pn, k) == 0){
-				$\color{blue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
+				$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
 				pn->p->t->value = v;
-				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
+				$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
[…]

### 2023-11-24 concurrency/CPP 2024/proof_lock_traverse.c

@@ -1,30 +1,30 @@
 int traverse(pn *pn, int k){
-		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (p, n) \ast  
+		$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (p, n) \ast  
 				\nodeboxrep(n)  \ast \mathsf{R}(n) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-		for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
+		for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
 				pn->p = pn->n;
-				$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', c, \mathsf{range})\end{array}\right\}$
+				$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', c, \mathsf{range})\end{array}\right\}$
 				if (pn->p->t == NULL)
 						return 1;						
-						$\color{violet} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
+						$\color{purple!40!black} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
 				else{
-						$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range})\end{array}\right\}$
[…]

### 2023-11-24 concurrency/CPP 2024/paper.tex
WIP

@@
 	backgroundcolor=\color{white},
+	belowcaptionskip=1\baselineskip,
+	frame=none,
+	columns=[c]fixed,
 	keywordstyle=\bfseries\color{green!40!black},
@@
 		%		\label{traverse_lock_a}	
-	\end{subfigure}\qquad
+	\end{subfigure}
+	\qquad
 	\begin{subfigure}[t]{0.45\textwidth}
@@
 \end{figure}
+
 Figure~\ref{template-ex} shows an example search structure template.
@@
 $$
- {\color{specblue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(s, v)\  {\color{specblue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
+ {\color{specblue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(p, v)\  {\color{specblue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
 \begin{mathpar}
-	{\color{specblue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ } 
+	{\color{specblue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ s\ |\ \mathsf{stack}_g\ vs\right\rangle\ } 
 	 \vspace{-0.8em}  \\ \texttt{push}(s, v)\   \vspace{-0.8em}  \\ 
-	{\color{specblue} \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle }
+	{\color{specblue} \left\langle \mathsf{is\_stack}_g\ s\ |\ \mathsf{stack}_g\ (v::vs)\right\rangle }
 \end{mathpar}
-indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution.
+indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $vs$ to $v::vs$ at some point during its execution.
 The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion.
@@
 \begin{lstlisting}[style=myStyle,language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, numbers=none, xleftmargin=0.5em]
-	typedef struct node_t {node *t; lock_t *lock;} node_t;
+	typedef struct node_t {
+		node *t; 
+		lock_t *lock;
+	} node_t;
 \end{lstlisting}
@@
 
+\begin{minipage}{.2\textwidth}
+	\centering
 \begin{mathpar}
@@
 	\vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
-\\
-	{\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ 
-	 \vspace{-0.8em} \\ \texttt{release}(n\texttt{->lock})\  \vspace{-0.8em} \\ {\color{specblue}\left\langle \nodeboxrep(n) \ |\ \treerep\ (m)\right\rangle}
 \end{mathpar}
+\end{minipage}
+\begin{minipage}{.28\textwidth}
+		\centering
+\begin{mathpar}
+	{\color{specblue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
+	\vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
+\end{mathpar}
+\end{minipage} \\
 
@@
 	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ast \mathsf{node\_data}(n, c) \\
-	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
+	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})
+\end{align*}
 where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.
@@
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \nodeboxrep(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
+\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}, \texttt{k}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \nodeboxrep(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
 \end{align*}
+
 \begin{figure*}[!ht]
@@
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast  \infp (r) \ast \infp (n)
+\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\  & \ \ \ \ \ \ \ \ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast  \infp (r) \ast \infp (n)
 \end{align*}

### 2023-11-24 concurrency/CPP 2024/lock_insert.c

@@ -1,5 +1,5 @@
 void insert (node_t **r, int k, void *v){
-  struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+  struct pn *pn = (struct pn*)surely_malloc(sizeof *pn);
   pn->n = *r;
   acquire(pn->n->lock);  
   if (traverse(pn, k) == 0){

### 2023-11-24 concurrency/CPP 2024/proof_giveup_insert.c

@@ -1,23 +1,23 @@
 void insert (node_t **r, int k, void *v){
-		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
-		pn->n = *r;
-		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \infp(\texttt{r}) \end{array}\right\}$
-		if (traverse(pn, k) == 0){
-				$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
-				pn->p->t->value = v;
-				$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
-				//Linearization point
-		}
-		else{
-				$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \infp(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$
-				insertOp(pn, k, v);
-				$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range}\ \ast \\ n'\texttt{->t}\mapsto (\texttt{k}, \texttt{v}, l, r')  \ast l \mapsto \texttt{NULL} \ast  r' \mapsto \texttt{NULL} \end{array}\right\}$ 
-				//Linearization point
-				$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$ 
-		}
[…]

### 2023-11-24 concurrency/CPP 2024/proof_giveup_traverse.c

@@ -1,40 +1,41 @@
 int traverse(pn *pn, int k){
-		node_t *r = (pn->n); $\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  
-\infp (\texttt{r}) \end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-		for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
-			\exists\ n'.\ \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \infp (\texttt{r})
+	node_t *r = (pn->n); $\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  
+				\infp (\texttt{r}) \end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+	for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
+		\exists\ n'.\ \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \infp (\texttt{r})
 		 \end{array}\right\}$
-				acquire(pn->n->lock);
-				$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
-				pn->p = pn->n; $\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
-				if (inRange(pn, k) == 1){
-						$\color{specblue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{node\_contents}(n', c, \mathsf{range}) \ast \infp (\texttt{r}) \end{array}\right\}$
-						if (pn->p->t == NULL)
-								return 1;
[…]

### 2023-11-24 concurrency/CPP 2024/proof_lock_insert.c

@@ -1,25 +1,25 @@
 void insert (node_t **r, int k, void *v){
-		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
-		pn->n = *r;
-		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \end{array}\right\}$
-		acquire(pn->n->lock);
-		$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
-		if (traverse(pn, k) == 0){
-				$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
-				pn->p->t->value = v;
-				$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
-				//Linearization point
-		}
-		else{
-				$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$
-				insertOp(pn, k, v);
-				$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range}\ \ast \\ n'\texttt{->t}\mapsto (\texttt{k}, \texttt{v}, l, r')  \ast l \mapsto \texttt{NULL} \ast  r' \mapsto \texttt{NULL} \end{array}\right\}$ 
-				//Linearization point
[…]

### 2023-11-24 concurrency/CPP 2024/proof_lock_traverse.c

@@ -1,31 +1,31 @@
 int traverse(pn *pn, int k){
-		$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (p, n) \ast  
+	$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (p, n) \ast  
 				\nodeboxrep(n)  \ast \mathsf{R}(n) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-		for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
-				pn->p = pn->n;
-				$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', c, \mathsf{range})\end{array}\right\}$
-				if (pn->p->t == NULL)
-						return 1;						
-						$\color{purple!40!black} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
-				else{
-						$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range})\end{array}\right\}$
-						int b = findNext(pn, k);
-						$\color{specblue} \left\{\begin{array}{l} \exists \  n''. \ \texttt{pn} \mapsto (n', n'') \ast \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range})\ \ast\\ 
+	for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
+		pn->p = pn->n;
[…]

### 2023-11-24 concurrency/CPP 2024/paper.tex
resolved some TODO

@@
 \wm{Should this go in background instead?}
-
-Logical atomicity is a separation logic technique for concisely specifying the behavior of a concurrent operation.
-A logically atomic triple has the form $\fforall a.\,\left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
+We specify the concurrent behavior of search structures using logical atomicity~\cite{tada, iris, iris-vst-arxiv}, a separation logic technique that concisely defines the behavior of concurrent operations.
+A logically atomic triple has the form ${\color{specblue} \fforall a.\,\left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle } \ \texttt{c}\ {\color{specblue}\left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle}$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure.
 Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread.
@@
 	\right\rangle$
+		\begin{minipage}{0.9\textwidth}
 			\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
+		\end{minipage}
 	$\color{specblue}
@@
 				\begin{array}{c}
-					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r)  \ast  \texttt{pn->n->min}\mapsto n_1 \ast  \texttt{pn->n->max}\mapsto n_2  \ \ast \ 
+					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r)  \ast  \texttt{pn->n->min}\mapsto n_1 \ast  \texttt{pn->n->max}\mapsto n_2  \ \ast 
 					\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
@@
 	\right\rangle$
+	\begin{minipage}{0.9\textwidth}
 	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
+	\end{minipage}
 	$\color{specblue}
@@
 		\right\rangle$
+		\begin{minipage}{0.9\textwidth}
 		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
+		\end{minipage}
 		$\color{specblue}
@@
 		\right\rangle$
+		\begin{minipage}{0.9\textwidth}
 		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
+		\end{minipage}
 		$\color{specblue}

### 2023-11-28 concurrency/CPP 2024/paper.tex
distinguished nodeboxrep/Ref from in_tree/inFP
I'd somehow collapsed the two in my descriptions.

@@
 \newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
-\newcommand{\infp}{\ensuremath{\mathsf{Ref }}}
+\newcommand{\inFP}{\ensuremath{\mathsf{In }}}
 
@@
 \setcopyright{rightsretained}
-\acmPrice{}
 \acmDOI{XXXXXXX.XXXXXXX}
@@
 These specifications are defined in terms of assertions $\nodeboxrep$, representing a client thread's handle to the data structure, and $\treerep$, representing the data structure's abstract state.
-Holding $\nodeboxrep(n)$ for a node $n$ should allow us to acquire the lock on $n$, which in turn gives us access to the contents of the node.
-Formally, we need to know that the following triples hold:
+We also have an assertion $\inFP(n)$ that serves as a reference to an individual node; $\nodeboxrep$ should include, at minimum, the $\inFP$ assertion for the root node.
+Holding $\inFP(n)$ for a node $n$ should allow us to acquire the lock on $n$, which in turn gives us access to the contents of the node. Formally, we need to know that the following triples hold:
 
@@
 \begin{mathpar}
-	{\color{specblue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
-	\vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
+	{\color{specblue}\left\langle \inFP \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
+	\vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \inFP(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
 \end{mathpar}
@@
 \begin{mathpar}
-	{\color{specblue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
-	\vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
+	{\color{specblue}\left\langle \inFP \left(n\right) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ 
+	\vspace{-0.8em} \\ \texttt{release}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \inFP(n) \ |\ \treerep\ (m)\right\rangle}
 \end{mathpar}
@@
 
-In other words, $\nodeboxrep(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant} for the node.
+In other words, $\inFP(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant} for the node.
 %In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).
@@
 			\texttt{pn} \mapsto (p, n) \ \ast \ 
-			\nodeboxrep(n)  \ast \ \mathsf{R}(n) 
+			\inFP(n)  \ast \ \mathsf{R}(n) 
 		\end{array}  
@@
 		\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
-		\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
+		\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
 		\\ 
@@
 
-The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}.
+The local precondition of \lstinline{traverse} includes both the node handle $\inFP(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}.
 The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$.
-The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.
+The local postcondition then includes the handle of the new node $\inFP(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.
 
@@
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}, \texttt{k}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \nodeboxrep(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
+\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}, \texttt{k}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \inFP(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
 \end{align*}
@@
 	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
-	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
+	\inFP(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
 	\right\rangle$
@@
 	\left\langle \mathit{res}. \ \exists \  n', v.
-	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
+	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \inFP(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
 		\\ 
@@
 				\texttt{pn} \mapsto (p, n) \ast  
-				\nodeboxrep(n)   
+				\inFP(n)   
 			\end{array}  
@@
 				\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
-				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast \texttt{k} \in \mathsf{range} \ \ast 
+				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast \texttt{k} \in \mathsf{range} \ \ast 
 				\\ 
@@
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\  & \ \ \ \ \ \ \ \ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast  \infp (r) \ast \infp (n)
+\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\  & \ \ \ \ \ \ \ \ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast  \inFP (r) \ast \inFP (n)
 \end{align*}
 Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}).
-Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.
+Instead, we keep a reference $\inFP(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.
 
 With this invariant in hand, the proof of \texttt{traverse} proceeds as follows.
-We begin by using the $\infp$ predicate to acquire the lock on \lstinline{pn->n} (line 4 in Figure \ref{traverse_giveup_b}).
+We begin by using the $\inFP$ predicate to acquire the lock on \lstinline{pn->n} (line 4 in Figure \ref{traverse_giveup_b}).
 We then call \lstinline{inRange} to check whether we are still on the path to a node that can hold \lstinline{k}.
@@
 Otherwise, we proceed as in the lock-coupling template: if the current node's contents are \texttt{NULL}, we have found the empty node where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release the lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}.
-In this case, we once again hold no locks, and have only the $\infp$ assertion indicating that the new node is in the data structure.
+In this case, we once again hold no locks, and have only the $\inFP$ assertion indicating that the new node is in the data structure.
 % and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}.
@@
 	\begin{array}{c}
-		\texttt{pn} \mapsto (p, n)  \ast  \nodeboxrep(n)  \big| \ \treerep\ (m)
+		\texttt{pn} \mapsto (p, n)  \ast  \inFP(n)  \big| \ \treerep\ (m)
 	\end{array}
@@
 	\left\langle \mathit{res}. \ \exists \  n', v.
-	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
+	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \inFP(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
 		\\ 

### 2023-11-28 concurrency/CPP 2024/proof_giveup_insert.c

@@ -1,23 +1,23 @@
 void insert (node_t **r, int k, void *v){
 	struct pn *pn = (struct pn *)surely_malloc(sizeof *pn);
 	pn->n = *r;
-	$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \infp(\texttt{r}) \end{array}\right\}$
+	$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \inFP(\texttt{r}) \end{array}\right\}$
 	if (traverse(pn, k) == 0){
-		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \infp(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \inFP(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
 		pn->p->t->value = v;
 		$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
 		//Linearization point
 	}
 	else{
-		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \infp(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \inFP(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$
 		insertOp(pn, k, v);
[…]

### 2023-11-28 concurrency/CPP 2024/proof_giveup_traverse.c

@@ -1,41 +1,41 @@
 int traverse(pn *pn, int k){
 	node_t *r = (pn->n); $\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  
-				\infp (\texttt{r}) \end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+				\inFP (\texttt{r}) \end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
 	for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
-		\exists\ n'.\ \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \infp (\texttt{r})
+		\exists\ n'.\ \texttt{pn} \mapsto (\texttt{\_}, n') \ast \inFP (n') \ast \inFP (\texttt{r})
 		 \end{array}\right\}$
 		acquire(pn->n->lock);
-		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
-		pn->p = pn->n; $\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \infp (n') \ast \mathsf{R}(n') \ast \infp (\texttt{r}) \end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, n') \ast \inFP (n') \ast \mathsf{R}(n') \ast \inFP (\texttt{r}) \end{array}\right\}$
+		pn->p = pn->n; $\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \inFP (n') \ast \mathsf{R}(n') \ast \inFP (\texttt{r}) \end{array}\right\}$
 		if (inRange(pn, k) == 1){
[…]

### 2023-11-28 concurrency/CPP 2024/proof_lock_traverse.c

@@ -1,30 +1,30 @@
 int traverse(pn *pn, int k){
 	$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (p, n) \ast  
-				\nodeboxrep(n)  \ast \mathsf{R}(n) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+				\inFP(n)  \ast \mathsf{R}(n) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
 	for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
 		pn->p = pn->n;
-		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', c, \mathsf{range})\end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \inFP(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', c, \mathsf{range})\end{array}\right\}$
 		if (pn->p->t == NULL)
 			return 1;						
-			$\color{purple!40!black} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
+			$\color{purple!40!black} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \inFP(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
 		else{
-			$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \nodeboxrep(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range})\end{array}\right\}$
[…]

### 2023-11-29 concurrency/Modular/give_up_template.c
finished full implemetations of lock-coupling and giveup for linked-list and bst

@@ -1,49 +1,0 @@
-//
-//  give_up_template.c
-//  BST
-//
-//  Created by Duc Than Nguyen on 5/29/23.
-//
-
-#include "give_up_template.h"
-
-int inRange(node_t *p, int x){
-    if (x > p->min && x < p->max)
-        return 1;
-    else
-        return 0;
-}
-//    FOUND = 0, NOTFOUND = 1, NULLNEXT = 2
-Status traverse(pn *pn, int x) {
-    Status flag = 2;
-    void *p = (pn->n);
-    for( ; ; ){
-        acquire(pn->n->lock);
-        pn->p = pn->n;
-        if (inRange (pn->p, x) == 1){
-            if (pn->p->t == NULL){
-                break;
-            }
-            else{
-                Status status = findNext(pn->p, (void**)&pn->n, x);
-                if (status == FOUND){
-                    flag = 0;
-                    break;
-                }
-                else if (status == NOTFOUND){
-                    flag = 1;
-                    break;
-                }
-                else{
-                    release(pn->p->lock);
-                }
-            }
-        }
-        else{
-            release(pn->p->lock);
-            pn->n = p;
-        }
-            
-    }
-    return flag;
-}

### 2023-11-29 concurrency/Modular/implementations/bst_giveup.c

@@ -1,0 +1,55 @@
+#include "bst_giveup.h"
+
+void insertOp_giveup(node_t* p, int x, void* value, Status status){
+    node_t* p1 = (struct node_t*)surely_malloc(sizeof (node_t));
+    node_t* p2 = (struct node_t*)surely_malloc(sizeof (node_t));
+    p1->t = NULL;
+    p2->t = NULL;
+
+    lock_t* l1 = (lock_t*)surely_malloc(sizeof(lock_t));
+    makelock(l1);
+    p1->lock = l1;
+    release(l1);
+
+    lock_t* l2 = (lock_t*)surely_malloc(sizeof(lock_t));
+    makelock(l2);
+    p2->lock = l2;
+    release(l2);
+
+    DList dlist; 
+    dlist.size = 2; // For BST 
+    dlist.list = (void**)surely_malloc(dlist.size * sizeof(node_t));
+    dlist.list[0] = p1;
+    dlist.list[1] = p2;
+
+    insertOp(&p->t, x, value, status, &dlist);
+
+    ((node_t*)dlist.list[0])->min = p->min;
+    ((node_t*)dlist.list[0])->max = x;
+    ((node_t*)dlist.list[1])->min = x;
+    ((node_t*)dlist.list[1])->max = p->max;
+} 
+
+void printDS_giveup(void **t){
+    printf("For BST - GIVEUP\n");
+    struct node_t* tgt;
+    tgt = *t;
+
+    if (tgt == NULL) {
+        return;
+    }
+    stack s;
+    initStack(&s);
+    node_t * current = tgt;
+
+    while (current->t != NULL || !isEmpty(&s)) {
+        while (current->t != NULL) {
+            push(&s, current);
+            current = (node_t *)getLeftChild(current->t);
+        }
+
+        current = pop(&s);
+        printKey(current->t);
+        current = (node_t *)getRightChild(current->t);
+    }
+}

### 2023-11-29 concurrency/Modular/implementations/common.h

@@ -1,0 +1,29 @@
+//
+//  data_structure.h
+//  BST
+//
+//  Created by Duc Than Nguyen on 6/15/23.
+//
+#ifndef COMMON_H
+#define COMMON_H
+
+typedef enum {
+    FOUND,
+    NOTFOUND,
+    NULLNEXT
+} Status;
+
+lock_t thread_lock;
+
+// Define the DynamicList struct
+typedef struct {
+    void** list;         // Double pointer to store elements (can be BSTNode or ListNode)
+    unsigned int size;   // Size of the list
+} DList;
+
+static void *surely_malloc (size_t n) {
+    void *p = malloc(n);
+    if (!p) exit(1);
+    return p;
+}
+#endif

### 2023-11-29 concurrency/Modular/implementations/giveup_template.c

@@ -1,0 +1,163 @@
+//
+//  give_up_template.c
+//  BST
+//
+//  Created by Duc Than Nguyen on 5/29/23.
+//
+
+#include "giveup_template.h"
+
+int inRange(node_t *p, int x){
+    if (x > p->min && x < p->max)
+        return 1;
+    else
+        return 0;
+}
+//FOUND == 0, NOTFOUND = 1, NULLNEXT = 2
+Status traverse(pn *pn, int x) {
+    Status status = 2; //NULLNEXT
+    void *p = (pn->n);
+    for( ; ; ){
+        acquire(pn->n->lock);
+        pn->p = pn->n;
+        if (inRange (pn->p, x) == 1){
+            if (pn->p->t == NULL){
+                status = NULLNEXT;
+                break;
+            }
+            else{
+                status = findNext(pn->p->t, (void**)&pn->n, x);
+                if (status == FOUND){
+                    status = FOUND;
+                    break;
+                }
+                else if (status == NOTFOUND){
+                    status = NOTFOUND;
+                    break;
+                }
+                else{
+                    release(pn->p->lock);
+                }
+            }
+        }
+        else{
+            release(pn->p->lock);
+            pn->n = p;
+        }
+            
+    }
+    return status;
+}
+
+/*
+void insertOp_list(node_t* p, int x, void* value, Status status){
+    struct node_t* pl; 
+    pl = (struct node_t*)surely_malloc(sizeof (node_t));
+    if (status == NULLNEXT) {
+        pl->t = NULL;
+        lock_t* l = (lock_t*)surely_malloc(sizeof(lock_t));
+        makelock(l);
+        pl->lock = l;
+        release(l); 
+
+        insertOp_llist(&p->t, x, value, status, pl);
+        pl->max = INT_MAX;
+        pl->min = x;
+    }
+    else{
+        lock_t* l = (lock_t*)surely_malloc(sizeof(lock_t));
+        makelock(l);
+        pl->lock = l;
+        release(l);
+        pl->t = p->t;
+        pl->max = p->max; 
+        pl->min = x; 
+        insertOp_llist(&(p->t), x, value, status, pl);
+    }
+} 
+*/
+/*
+void insertOp_list(node_t* p, int x, void* value, Status status){
+    DList dlist; 
[…]

### 2023-11-29 concurrency/Modular/implementations/giveup_template.h

@@ -1,0 +1,54 @@
+//
+//  give_up_template.h
+//  BST
+//
+//  Created by Duc Than Nguyen on 5/29/23.
+//
+
+#include "data_structure.h"
+#include "template.h"
+
+typedef struct node_t {node *t; lock_t *lock; int min; int max; } node_t;
+/*
+//void insertOp_bst(node_t* p, int x, void* value, Status status);
+//void insertOp_list(node_t* p, int x, void* value, Status status);
+*/
+
+void insertOp_giveup(node_t* p, int x, void* value, Status status);
+void insertOp_helper(node_t* p, int x, void* value, Status status);
+int inRange(node_t *p, int x);
+Status traverse(pn *pn, int x);
+node_t * treebox_helper(node_t *newt);
+lock_t* getLock(node_t *p);
+void changeValue_helper(node_t * p, void * value);
+void *getValue_helper(node_t* p_ds);
+void printDS_giveup (void **t);
+void printDS_helper (void **t);
+
+
+typedef struct stack {
+    struct node_t * items[100]; // Assuming a maximum of 100 nodes
+    int top;
+} stack;
+
+static void initStack(stack* s) {
+    s->top = -1;
+}
+
+// Function to push a node onto the stack
+static void push(stack* s, node_t * item) {
+    s->items[++s->top] = item;
+}
+
+// Function to check if the stack is empty
+static int isEmpty(stack* s) {
+    return s->top == -1;
+}
+
+// Function to pop a node from the stack
+static node_t* pop(stack* s) {
+    if (!isEmpty(s)) {
+        return s->items[s->top--];
+    }
+    return NULL;
+}

### 2023-11-29 concurrency/Modular/implementations/link_template.h

@@ -7,7 +7,7 @@
 
 #include "data_structure.h"
 
-typedef struct node_t {node *t; lock_t lock; int min; int max; } node_t;
+typedef struct node_t {node *t; lock_t *lock; int min; int max; } node_t;
 
 //Template style
 typedef struct pn {
@@ -15,6 +15,6 @@ typedef struct pn {
     struct node_t *n;
 } pn;
 
-int inRange(node_t *p, int x);
+void insertOp_helper(void* p, int x, void* value, Status status);
 Status traverse(pn *pn, int x);
-
+node_t * treebox_helper(node_t *newt);

### 2023-11-29 concurrency/Modular/implementations/list_giveup.c

@@ -1,0 +1,49 @@
+#include "list_giveup.h"
+
+void insertOp_giveup(node_t* p, int x, void* value, Status status){
+    DList dlist; 
+    dlist.size = 1; // For Linked-list 
+    dlist.list = (void**)surely_malloc(dlist.size * sizeof(node_t));
+    dlist.list[0] = (struct node_t*)surely_malloc(sizeof (node_t));
+    if (status == NULLNEXT) {
+        ((node_t*)dlist.list[0])->t = NULL;
+        lock_t* l = (lock_t*)surely_malloc(sizeof(lock_t));
+        makelock(l);
+        ((node_t*)dlist.list[0])->lock = l;
+        release(l);
+        insertOp(&p->t, x, value, status, &dlist);
+        ((node_t*)dlist.list[0])->min = x;
+        ((node_t*)dlist.list[0])->max = INT_MAX;
+    }
+    else{
+        lock_t* l = (lock_t*)surely_malloc(sizeof(lock_t));
+        makelock(l);
+        ((node_t*)dlist.list[0])->lock = l;
+        release(l);
+        ((node_t*)dlist.list[0])->t = p->t;
+        ((node_t*)dlist.list[0])->max = p->max; 
+        ((node_t*)dlist.list[0])->min = x;
+        insertOp(&(p->t), x, value, status, &dlist);
+    }
+} 
+
+void printDS_giveup(void **t){
+    printf("For Linked-list - GIVEUP\n");
+    struct node_t* tgt;
+    tgt = *t;
+
+    if (tgt == NULL) {
+        return;
+    }
+    stack s;
+    initStack(&s);
+    node_t * current = tgt;
+    while (current->t != NULL) {
+        push(&s, current);
+        current = (node_t *)getNext(current->t);
+    }
+    while (!isEmpty(&s)){   
+        current = pop(&s);
+        printKey(current->t);
+    }
+}

### 2023-11-29 concurrency/Modular/implementations/lock_template.h

@@ -1,0 +1,54 @@
+//
+//  give_up_template.h
+//  BST
+//
+//  Created by Duc Than Nguyen on 5/29/23.
+//
+
+#include "data_structure.h"
+#include "template.h"
+
+typedef struct node_t {node *t; lock_t *lock; } node_t;
+/*
+//void insertOp_bst(node_t* p, int x, void* value, Status status);
+//void insertOp_list(node_t* p, int x, void* value, Status status);
+*/
+
+void insertOp_lock(node_t* p, int x, void* value, Status status);
+void insertOp_helper(node_t* p, int x, void* value, Status status);
+int inRange(node_t *p, int x);
+Status traverse(pn *pn, int x);
+node_t * treebox_helper(node_t *newt);
+lock_t* getLock(node_t *p);
+void changeValue_helper(node_t * p, void * value);
+void *getValue_helper(node_t* p_ds);
+void printDS_lock (void **t);
+void printDS_helper (void **t);
+
+
+typedef struct stack {
+    struct node_t * items[100]; // Assuming a maximum of 100 nodes
+    int top;
+} stack;
+
+static void initStack(stack* s) {
+    s->top = -1;
+}
+
+// Function to push a node onto the stack
+static void push(stack* s, node_t * item) {
+    s->items[++s->top] = item;
+}
+
+// Function to check if the stack is empty
+static int isEmpty(stack* s) {
+    return s->top == -1;
+}
+
+// Function to pop a node from the stack
+static node_t* pop(stack* s) {
+    if (!isEmpty(s)) {
+        return s->items[s->top--];
+    }
+    return NULL;
+}

### 2023-11-29 concurrency/Modular/implementations/template.c

@@ -1,0 +1,62 @@
+//
+//  template.c
+//  BST
+//
+//  Created by Duc Than Nguyen on 6/22/23.
+//
+
+#include <stdio.h>
+#include <limits.h>
+#include <stdlib.h>
+#include "threads.h"
+#include "template.h"
+
+void insert (treebox t, int x, void *value) {
+    struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+    pn->n = *t;
+    Status status = traverse(pn, x);
+    if (status == 0){
+        changeValue_helper(pn->p, value);
+    }
+    else
+    {
+        insertOp_helper(pn->p, x, value, status);
+    }
+    release(getLock(pn->n));
+    free(pn);
+}
+
+
+void *lookup (treebox t, int x) {
+    struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
+    void *v;
+    pn->n = *t;
+    Status status = traverse(pn, x);
+    if (status == FOUND){
+        v = getValue_helper(pn->p);
+    }
+    else
+    {
+        v = NULL;
+    }
+    release(getLock(pn->n));
+    free(pn);
+    return v;
+}
+
+
+treebox treebox_new(void) {
+    treebox p = (treebox) surely_malloc(sizeof (*p));
+    node_t *newt;
+    //*p = newt;
+    *p = treebox_helper(newt);
+    return p;
+}
+
+/*
+void treebox_free(treebox b) {
+    struct node_t *t = *b;
+    freeDS(t);
+    free(b);
+}
+*/

### 2023-11-29 concurrency/Modular/implementations/template.h

@@ -1,0 +1,39 @@
+//
+//  template.h
+//  BST
+//
+//  Created by Duc Than Nguyen on 6/22/23.
+//
+
+#include <stdio.h>
+#include <limits.h>
+#include <stdlib.h>
+#include "threads.h"
+#include "common.h"
+
+typedef struct node_t node_t;
+typedef struct node_t **treebox;
+treebox tb;
+
+//Template style
+typedef struct pn {
+    struct node_t *p;
+    struct node_t *n;
+} pn;
+
+Status traverse(pn *pn, int x);
+void insertOp_helper(node_t* p, int x, void* value, Status status);
+lock_t * getLock(node_t *p);
+node_t * treebox_helper(node_t *newt);
+void changeValue(void* p_ds, void* value);
+void *getValue_helper(node_t* p_ds);
+void changeValue_helper(node_t * p, void * value);
+//void freeDS(void *p_ds);
+
+void insert (treebox t, int x, void *value);
+void *lookup (treebox t, int x);
+treebox treebox_new(void);
+void treebox_free(treebox b);
+void printDS_helper (void **tgt);
+void printDS_1 (node_t *tgt);
+

### 2023-12-08 concurrency/CPP 2024/paper.tex
finished Conference/Publisher-Specific Requirements

@@
 %% The first command in your LaTeX source must be the \documentclass command.
-\documentclass[sigplan,10pt, screen]{acmart}
+\documentclass[sigplan,screen]{acmart}
 \settopmatter{}
@@
 
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{rightsretained}
-\acmDOI{XXXXXXX.XXXXXXX}
+%%% The following is specific to CPP '24 and the paper
+%%% 'Compositional Verification of Concurrent C Programs with Search Structure Templates'
+%%% by Duc-Than Nguyen, Lennart Beringer, William Mansky, and Shengyi Wang.
+%%%
+\setcopyright{acmlicensed}
+\acmDOI{10.1145/3636501.3636940}
 \acmYear{2024}
 \copyrightyear{2024}
-\acmSubmissionID{}
-\acmISBN{}
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[CPP '24]{Proceedings of the 13th ACM SIGPLAN International Conference on Certified Programs and Proofs}{Jan 15--16, 2024}{London, United Kingdom}
-\acmBooktitle{Proceedings of the 13th ACM SIGPLAN International Conference on Certified Programs and Proofs (CPP '24), Jan 15--16, 2024, London, United Kingdom}
+\acmSubmissionID{poplws24cppmain-p2-p}
+\acmISBN{979-8-4007-0488-8/24/01}
+\acmConference[CPP '24]{Proceedings of the 13th ACM SIGPLAN International Conference on Certified Programs and Proofs}{January 15--16, 2024}{London, UK}
+\acmBooktitle{Proceedings of the 13th ACM SIGPLAN International Conference on Certified Programs and Proofs (CPP '24), January 15--16, 2024, London, UK}
+\received{2023-09-19}
+\received[accepted]{2023-11-25}
 
@@
 
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
-
 \author{Duc-Than Nguyen}
+\orcid{0000-0002-6810-897X}
+\affiliation{%
+	\institution{University of Illinois at Chicago}
+	\city{}
+	\country{USA}
+}
 \email{dnguye96@uic.edu}
-\orcid{https://orcid.org/0000-0002-6810-897X}
-\affiliation{\institution{University of Illinois at Chicago}\streetaddress{}\city{Chicago}\state{IL}\country{USA}\postcode{}}
 
 \author{Lennart Beringer}
+\orcid{0000-0002-1570-3492}
+\affiliation{%
+	\institution{Princeton University}
+	\city{}
+	\country{USA}
+}
 \email{eberinge@cs.princeton.edu}
-\orcid{https://orcid.org/0000-0002-1570-3492}
-\affiliation{\institution{Princeton University}\streetaddress{}\city{Princeton}\state{NJ}\country{USA}\postcode{}}
 
 \author{William Mansky}
+\orcid{0000-0002-5351-895X}
+\affiliation{%
+	\institution{University of Illinois at Chicago}
+	\city{}
+	\country{USA}
+}
 \email{mansky1@uic.edu}
-\orcid{https://orcid.org/0000-0002-5351-895X}
-\affiliation{\institution{University of Illinois at Chicago}\streetaddress{}\city{Chicago}\state{IL}\country{USA}\postcode{}}
 
 \author{Shengyi Wang}
+\orcid{0000-0002-2286-8703}
+\affiliation{%
+	\institution{Princeton University}
+	\city{}
+	\country{USA}
+}
 \email{shengyiw@princeton.edu}
-\orcid{https://orcid.org/0000-0002-2286-8703}
-\affiliation{\institution{Princeton University}\streetaddress{}\city{Princeton}\state{NJ}\country{USA}\postcode{}}
-
 %%

### 2023-12-08 concurrency/CPP 2024/paper.tex
working on camera-ready

--- 2023-11-28 concurrency/CPP 2024/paper.tex
@@
 \startPage{1}
-\bibliographystyle{ACM-Reference-Format}
 
@@
 \usepackage{amsthm}
+\usepackage{tikzit}
+\input{box.tikzstyles}
+\usetikzlibrary{fit}
 \usepackage{mathpartir}
@@
 
+%https://tex.stackexchange.com/questions/112634/how-to-draw-a-diagram-with-boxes-of-text-inside-each-other-and-have-them-laid-ou
+\newcommand\TBox[2][]{%
+  \tikz\node[draw,align=left,#1] {#2};\hskip2pt}
+
+\usetikzlibrary{positioning}
+
 \definecolor{specblue}{rgb}{0.0,0.3,0.55}
@@
 \label{sec:introduction}
+Concurrent search structure templates~\cite{shasha1988concurrent} are a technique for describing and implementing concurrency patterns (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) that can then be specialized to various search data structures (e.g., linked list, hashtable, B-tree) compositionally.
 Krishna et al.
-~\cite{templates} proposed concurrent search structure templates as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree).
-The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency.
+~\cite{templates} repurposed templates as a compositional \emph{verification} technique, separating the proof of a concurrent access method from the proof of the underlying data structure.
+The concurrency templates are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency.
 In theory, this could allow us to prove the correctness of $n$ concurrency patterns and $m$ (single-threaded) data structure implementations, and immediately obtain $n \times m$ verified concurrent data structures.
@@
 	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
-	\item We give a precise description of search structure templates, and identify places where it is difficult in practice to preserve the boundary between template and data structure.
+	\item We give a precise description of search structure templates, and identify places where it is difficult in practice to preserve the boundary between template and data structure: for instance, both node creation and rotation in binary search trees operate on concurrent and data-structure-specific aspects in ways that do not cleanly decompose.
 \end{itemize}
@@
 Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeting an implementation language (C) rather than a toy language or algorithmic representation.
-%do we want to say this part?
+For instance, Iris's RustBelt instance~\cite{rustbelt} targets a core calculus based on Rust's intermediate language and has been used to verify several interesting concurrent programs, but does not provide guarantees about actual running Rust code.
 
@@
 	\begin{subfigure}[t]{0.45\textwidth}
-		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
+		\lstinputlisting[language=caml, style=myStyle, numbers=none]{lock_traverse.ml} 
 		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
@@
 	\begin{subfigure}[t]{0.45\textwidth}
-		\lstinputlisting[language=caml, style=myStyle]{insert.ml} 
+		\lstinputlisting[language=caml, style=myStyle, numbers=none]{insert.ml} 
 	\end{subfigure}
@@
 \subsubsection{Logical atomicity} 
+\label{logatom}
 %In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification.
@@
 $$
- {\color{specblue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(p, v)\  {\color{specblue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
+ {\color{specblue} \{\mathsf{stack}(s, p)\} } \ \texttt{push}(p, v)\  {\color{specblue} \{\mathsf{stack}(v :: s, p)\}}$$  Its concurrent counterpart is canonically specified as 
 \begin{mathpar}
-	{\color{specblue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ s\ |\ \mathsf{stack}_g\ vs\right\rangle\ } 
-	 \vspace{-0.8em}  \\ \texttt{push}(s, v)\   \vspace{-0.8em}  \\ 
-	{\color{specblue} \left\langle \mathsf{is\_stack}_g\ s\ |\ \mathsf{stack}_g\ (v::vs)\right\rangle }
+	{\color{specblue} \fforall s.\ \left\langle \mathsf{is\_stack}_g(p)\ |\ \mathsf{stack}_g(s)\right\rangle\ } 
+	 \vspace{-0.8em}  \\ \texttt{push}(p, v)\   \vspace{-0.8em}  \\ 
+	{\color{specblue} \left\langle \mathsf{is\_stack}_g(p)\ |\ \mathsf{stack}_g(v::s)\right\rangle }
 \end{mathpar}
@@
 The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context.
-Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.
+Logical atomicity can be used to prove linearizability, the standard correctness condition for concurrent data structures---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}.
+It can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}, but in this paper we will focus on linearizable data structures and canonical specifications.
 
@@
 It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads.
-In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values:
+In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values (we omit the ghost identifier $g$ connecting $\nodeboxrep$ and $\treerep$, as described in section~\ref{logatom}):
 \begin{mathpar}
@@
 
-A search structure template is an implementation of these three functions, written in a particularly restricted way.
-Aside from some top-level management code (e.g., acquiring and releasing an initial lock), all that each function should do is call two functions: a function called \lstinline{traverse}, which is specific to the template and implements its synchronization mechanism, and a concurrency-unaware function (e.g.
+\begin{figure}[htb]
+\ctikzfig{templates}
+	\caption{Diagram of the template approach. Each pair of template and data structure should yield a verified concurrent data structure.}
+	\label{templates-diagram}
+\end{figure}
+
+A search structure template is an implementation of these three functions, written in a particularly restricted way outlined in Figure~\ref{templates-diagram}.
+Aside from some top-level management code (e.g., acquiring and releasing an initial lock), all that each function should do is call two functions: a helper function called \lstinline{traverse}, which is specific to the template and implements its synchronization mechanism, and a concurrency-unaware function (e.g.
 \lstinline{insertOp}) provided by the target data structure.
@@
 
-In other words, $\inFP(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant} for the node.
+In other words, $\inFP(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant} for the node, which we will soon describe in detail.
 %In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).
@@
 		\end{array}  
-		\Big \vert  \treerep\ (m)
+		\Big \vert  \ \treerep\ (m)
 	}  
@@
 {\begin{array}{l}
-		\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
-		\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
-		\\ 
-		\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-		\\ 
-		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
-\end{array}  
-\Bigg \vert \treerep\ (m)
+		\mathit{res}.\ \exists \  n', v, \mathit{range}. \\ 
+		\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast\ \texttt{k} \in \mathit{range}\ \ast \ 
+		\\ 
+		\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ 
+		\\ 
+		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathit{range})
+\end{array}  
+\Bigg \vert \ \treerep\ (m)
 }  
@@
 The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$.
-The local postcondition then includes the handle of the new node $\inFP(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.
+The local postcondition then includes the handle of the new node $\inFP(n')$, its contents $\mathsf{node\_contents}(n', ...)$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k} (which is reflected in the remaining arguments to $\mathsf{node\_contents}$).
 
@@
 Formally, the lock invariant is defined by
-\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
-	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ast \mathsf{node\_data}(n, c) \\
-	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})
+\begin{align*}&\mathsf{node\_contents}(n, c, \mathit{range}) \triangleq \\ 
+	&\qquad\mathsf{ghost\_node}(n, c,\mathit{range}) \ast \mathsf{node\_data}(n, c) \\
+	&\mathsf{R}(n) \triangleq \exists\ c, \mathit{range}.\ \mathsf{node\_contents}(n, c, \mathit{range})
 \end{align*}
@@
 The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
-\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}, \texttt{k}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \inFP(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
+\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}, \texttt{k}) \triangleq \\ \exists \ p, n, c, \mathit{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathit{range}\ \ast \ \\  \inFP(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathit{range})
 \end{align*}
@@
 	\left\langle \mathit{res}. \ \exists \  n', v.
-	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \inFP(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
-		\\ 
-		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
-	\end{array}
-	\ \Bigg| \treerep\ (m) \
+	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \inFP(n') \ \ast\ \texttt{k} \in \mathit{range}\ \ast\ 
+		\\ 
+		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ 
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathit{range})
+	\end{array}
+	\ \Bigg| \ \treerep\ (m) \
 	\right\rangle$
@@
 			\end{array}  
-			\Big \vert  \treerep\ (m)
+			\Big \vert \ \treerep\ (m)
 		}  
@@
 		{\begin{array}{l}
-				\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
-				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast \texttt{k} \in \mathsf{range} \ \ast 
-				\\ 
-				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-				\\ 
-				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
-			\end{array}  
-			\Bigg \vert \treerep\ (m)
+				\mathit{res}.\ \exists \  n', v, \mathit{range}. \\ 
+				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast \texttt{k} \in \mathit{range} \ \ast 
+				\\ 
+				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ 
+				\\ 
+				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathit{range})
+			\end{array}  
+			\Bigg \vert \ \treerep\ (m)
 		}  
@@
 Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
-\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\  & \ \ \ \ \ \ \ \ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast  \inFP (r) \ast \inFP (n)
+\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\  & \ \ \ \ \ \ \ \ \exists \ p, n.\ \texttt{pn} \mapsto (p, n) \ast  \inFP (\texttt{r}) \ast \inFP (n)
 \end{align*}
 Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}).
-Instead, we keep a reference $\inFP(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.
+Instead, we keep a reference $\inFP(\texttt{r})$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.
 
@@
 We then call \lstinline{inRange} to check whether we are still on the path to a node that can hold \lstinline{k}.
-If the check fails (lines 18-20), we immediately release the lock and start over from the root node $r$.
+If the check fails (lines 18-20), we immediately release the lock and start over from the root node \texttt{r}.
 Otherwise, we proceed as in the lock-coupling template: if the current node's contents are \texttt{NULL}, we have found the empty node where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release the lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}.
@@
 However, this two-way dependency seems fundamental when the layout of the node structure depends on the concurrency control mechanism we choose.
-For now, we break the abstraction boundary and verify two different versions of \lstinline{insertOp}, one for the lock-coupling template and one for the give-up template
+For now, we break the abstraction boundary and verify two different versions of \lstinline{insertOp}, one for the lock-coupling template and one for the give-up template, which reduces the proof-reuse benefits of the template approach.
 
@@
 However, the translation also exposed some ambiguities and limitations in the approach: for instance, the \lstinline{inRange} function of the give-up template must either be assumed to exist in the data structure or incorporated into the template, and the \lstinline{delete} operation of binary search trees does not decompose naturally into a concurrent part and a data-structure-specific part.
-%\wm{Do we have an idea of how to fix this?}
+Templates still allow us to reuse proofs across data structures and concurrency patterns, but not as completely as the theoretical approach suggests.
+
 The next step is to implement templates as a truly generic framework, with proofs of \lstinline{traverse}, etc.
 that can be freely combined with verified sequential data structures via a clear interface that defines the functions a data structure must provide and the assumptions the template makes on them (e.g., by verifying template functions with respect to a data-structure typeclass).
+We are also interested in verifying templates that use atomic operations instead of locks, as is common in optimistic concurrency control.
 Ultimately, we hope to use the template approach to verify data structures such as Masstree~\cite{masstree} and Wormhole~\cite{wormhole}, complicated real-world search structures that combine multiple concurrency patterns for maximum performance on multicore architectures.
@@
 We thank Roshan Sharma, Alex Oey, and Anastasiia Evdokimova for extensive work on the original binary search tree implementation and verification, and the anonymous reviewers for providing comprehensive and insightful reviews.
-This research was supported, by ... 
+This work was partially funded by the National Science Foundation under the award 1005849 (Veriﬁed High Performance Data Structure Implementations, Beringer \& Mansky) and the Defense Advanced Research Projects Agency (DARPA) under contract HR001120C0160.
 \end{acks}
@@
 	\begin{array}{c}
-		\texttt{pn} \mapsto (p, n)  \ast  \inFP(n)  \big| \ \treerep\ (m)
+		\texttt{pn} \mapsto (p, n)  \ast  \inFP(n) \ \big| \ \treerep\ (m)
 	\end{array}
@@
 	\left\langle \mathit{res}. \ \exists \  n', v.
-	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \inFP(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
-		\\ 
-		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
+	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \inFP(n') \ \ast \ \texttt{k} \in \mathit{range}\ \ast \ 
+		\\ 
+		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ 
+		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathit{range}))
 	\end{array}

### 2023-12-08 concurrency/CPP 2024/proof_giveup_insert.c

@@ -3,17 +3,17 @@ void insert (node_t **r, int k, void *v){
 	pn->n = *r;
 	$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \inFP(\texttt{r}) \end{array}\right\}$
 	if (traverse(pn, k) == 0){
-		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \inFP(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \inFP(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathit{range}) \end{array}\right\}$
 		pn->p->t->value = v;
-		$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
+		$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathit{range}) \end{array}\right\}$ 
 		//Linearization point
 	}
 	else{
-		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \inFP(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \inFP(n') \ast \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ast \texttt{k} \in \mathit{range} \end{array}\right\}$
 		insertOp(pn, k, v);
[…]

### 2023-12-08 concurrency/CPP 2024/proof_giveup_traverse.c

@@ -1,6 +1,6 @@
 int traverse(pn *pn, int k){
 	node_t *r = (pn->n); $\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_, r}) \ast  
-				\inFP (\texttt{r}) \end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
+				\inFP (\texttt{r}) \end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv}(\texttt{pn}) \end{array}\right\}$
 	for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} 
 		\exists\ n'.\ \texttt{pn} \mapsto (\texttt{\_}, n') \ast \inFP (n') \ast \inFP (\texttt{r})
 		 \end{array}\right\}$

### 2023-12-08 concurrency/CPP 2024/proof_lock_insert.c

@@ -5,17 +5,17 @@ void insert (node_t **r, int k, void *v){
 	acquire(pn->n->lock);
 	$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{\_}, \texttt{r}) \ast  \nodeboxrep(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \end{array}\right\}$
 	if (traverse(pn, k) == 0){
-		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathsf{range}) \end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, v'), \mathit{range}) \end{array}\right\}$
 		pn->p->t->value = v;
-		$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathsf{range}) \end{array}\right\}$ 
+		$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', (\texttt{k}, \texttt{v}), \mathit{range}) \end{array}\right\}$ 
 		//Linearization point
 	}
 	else{
-		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ast \texttt{k} \in \mathsf{range} \end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l}  \exists\ n'.\ \texttt{pn} \mapsto (n', n') \ast n'\texttt{->t} = \texttt{NULL}  \ast \nodeboxrep(n') \ast \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ast \texttt{k} \in \mathit{range} \end{array}\right\}$
[…]

### 2023-12-08 concurrency/CPP 2024/proof_lock_traverse.c

@@ -1,7 +1,7 @@
 int traverse(pn *pn, int k){
 	$\color{specblue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (p, n) \ast  
 				\inFP(n)  \ast \mathsf{R}(n) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
-	for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} $
+	for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv}(\texttt{pn}) \end{array}\right\} $
 		pn->p = pn->n;
 		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \inFP(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', c, \mathsf{range})\end{array}\right\}$
 		if (pn->p->t == NULL)

### 2023-12-09 concurrency/CPP 2024/proof_giveup_traverse.c
edit font

@@ -8,25 +8,25 @@ int traverse(pn *pn, int k){
 		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{\_}, n') \ast \inFP (n') \ast \mathsf{R}(n') \ast \inFP (\texttt{r}) \end{array}\right\}$
 		pn->p = pn->n; $\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \inFP (n') \ast \mathsf{R}(n') \ast \inFP (\texttt{r}) \end{array}\right\}$
 		if (inRange(pn, k) == 1){
-			$\color{specblue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \inFP (n') \ast \mathsf{node\_contents}(n', c, \mathsf{range}) \ast \inFP (\texttt{r}) \end{array}\right\}$
+			$\color{specblue} \left\{\begin{array}{l} \texttt{k} \in \mathit{range} \ast \texttt{pn} \mapsto (n', n') \ast \inFP (n') \ast \mathsf{node\_contents}(n', c, \mathit{range}) \ast \inFP (\texttt{r}) \end{array}\right\}$
 			if (pn->p->t == NULL)
 				return 1;
-				$\color{purple!40!black} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \inFP (n') \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
+				$\color{purple!40!black} \left\{\begin{array}{l} \texttt{k} \in \mathit{range} \ast \texttt{pn} \mapsto (n', n') \ast \inFP (n') \ast \mathsf{node\_contents}(n', \cdot, \mathit{range}) \end{array}\right\}$
 			else{
-				$\color{specblue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range} \ast \texttt{pn} \mapsto (n', n') \ast \inFP (n') \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range}) \ast \inFP (\texttt{r}) \end{array}\right\}$
[…]

### 2023-12-09 concurrency/CPP 2024/proof_lock_traverse.c

@@ -3,28 +3,28 @@ int traverse(pn *pn, int k){
 				\inFP(n)  \ast \mathsf{R}(n) 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
 	for( ; ; ){ 			$\color{specblue} \left\{\begin{array}{l} \mathsf{traverse\_inv}(\texttt{pn}) \end{array}\right\} $
 		pn->p = pn->n;
-		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \inFP(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', c, \mathsf{range})\end{array}\right\}$
+		$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \inFP(n') \ast  \texttt{k} \in \mathit{range} \ast \mathsf{node\_contents}(n', c, \mathit{range})\end{array}\right\}$
 		if (pn->p->t == NULL)
 			return 1;						
-			$\color{purple!40!black} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \inFP(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \end{array}\right\}$
+			$\color{purple!40!black} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \inFP(n') \ast  \texttt{k} \in \mathit{range} \ast \mathsf{node\_contents}(n', \cdot, \mathit{range}) \end{array}\right\}$
 		else{
-			$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \inFP(n') \ast  \texttt{k} \in \mathsf{range} \ast \mathsf{node\_contents}(n', (k', v'), \mathsf{range})\end{array}\right\}$
+			$\color{specblue} \left\{\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast  \inFP(n') \ast  \texttt{k} \in \mathit{range} \ast \mathsf{node\_contents}(n', (k', v'), \mathit{range})\end{array}\right\}$
 			int b = findNext(pn, k);
[…]

### 2024-11-14 concurrency/template paper/paper.tex
outlining paper on formal templates

--- 2023-12-08 concurrency/CPP 2024/paper.tex
@@
 \usepackage[edges]{forest}
+\usepackage{balance}
 \renewcommand{\lstlistingname}{Figure}
@@
 
-%https://tex.stackexchange.com/questions/112634/how-to-draw-a-diagram-with-boxes-of-text-inside-each-other-and-have-them-laid-ou
-\newcommand\TBox[2][]{%
-  \tikz\node[draw,align=left,#1] {#2};\hskip2pt}
-
 \usetikzlibrary{positioning}
@@
 
-%%% The following is specific to CPP '24 and the paper
-%%% 'Compositional Verification of Concurrent C Programs with Search Structure Templates'
-%%% by Duc-Than Nguyen, Lennart Beringer, William Mansky, and Shengyi Wang.
-%%%
-\setcopyright{acmlicensed}
-\acmDOI{10.1145/3636501.3636940}
-\acmYear{2024}
-\copyrightyear{2024}
-\acmSubmissionID{poplws24cppmain-p2-p}
-\acmISBN{979-8-4007-0488-8/24/01}
-\acmConference[CPP '24]{Proceedings of the 13th ACM SIGPLAN International Conference on Certified Programs and Proofs}{January 15--16, 2024}{London, UK}
-\acmBooktitle{Proceedings of the 13th ACM SIGPLAN International Conference on Certified Programs and Proofs (CPP '24), January 15--16, 2024, London, UK}
-\received{2023-09-19}
-\received[accepted]{2023-11-25}
-
 %%
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Compositional Verification of Concurrent C Programs with Search Structure Templates}
+\title{A Formal Interface for Concurrent Search Structure Templates}
 
@@
 \affiliation{%
-	\institution{University of Illinois at Chicago}
+	\institution{University of Illinois Chicago}
 	\city{}
@@
 \email{dnguye96@uic.edu}
-
-\author{Lennart Beringer}
-\orcid{0000-0002-1570-3492}
-\affiliation{%
-	\institution{Princeton University}
-	\city{}
-	\country{USA}
-}
-\email{eberinge@cs.princeton.edu}
 
@@
 
-\author{Shengyi Wang}
-\orcid{0000-0002-2286-8703}
-\affiliation{%
-	\institution{Princeton University}
-	\city{}
-	\country{USA}
-}
-\email{shengyiw@princeton.edu}
 %%
@@
 \begin{abstract}
-Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort.
-In this paper, we implement the template approach in the Verified Software Toolchain (VST), and use it to prove correctness of C implementations of fine-grained concurrent data structures.
-This involves translating code, specifications, and proofs to the idiom of C and VST, and gives us another look at the requirements and limitations of the template approach.
-We encounter several questions about the boundaries between template and data structure, as well as some common data structure operations that cannot naturally be decomposed into templates.
-Nonetheless, the approach appears promising for modular verification of real-world concurrent data structures.
+
 \end{abstract}
@@
 \label{sec:introduction}
+Search structure templates~\cite{templates} were introduced by Krishna et al.
+as an approach to modular verification of concurrent data structures.
+The approach consists of a programming pattern and specification style that decompose concurrent data structures into \emph{concurrency templates} (lock coupling, give-up) and \emph{sequential data structures} (linked list, hash table).
+Later work in the same line extends the template approach to adding multicopy support (e.g., log-structure merging)~\cite{}, or combining thread-safe node implementations into larger data structures~\cite{}.
+However, these applications are qualitatively different from the original promise of the approach: that we can decompose both the implementation and the proof of a concurrent data structure into a \emph{concurrency-unaware} data structure component and a \emph{data-structure-unaware} concurrency component.
+This form of modularity is, to the best of our knowledge, unique in the literature, at least when the concurrent component is fine-grained (i.e., implemented with a lock per node or lock-free operations rather than one big lock).
+It also poses unique challenges in both implementation and specification/verification.
+
+However, while \citet{templates} lay out the high-level ideas of the template approach and work through several examples, they do not formally define the approach.
+In particular, a modularization technique should have a well-defined \emph{interface} for each of its components, so that we know that components written and verified according to the technique can be successfully combined.
+In theory, given $m$ data structure components and $n$ concurrency templates, we should immediately be able to obtain $m \times n$ verified concurrent data structures.
+The original paper by Krishna et al.
+does not realize this (nor does it claim to): it presents three verified templates and five verified data structures, but each data structure is combined with only one template.
+In fact, close examination of the examples shows that different templates use different and incompatible specifications for the data structure component.
+In this paper, we rectify this by presenting a \textbf{formal interface for concurrent search structure templates}, implemented as a library in C and a collection of typeclasses in Coq, that guarantees $m \times n$ verified data structures from $m + n$ components.
+More specifically:
+\begin{itemize}
+\item We describe an approach to \textbf{implementing} data structures and synchronization mechanisms so that they can be freely combined.
+This is a surprisingly challenging task: ideally, the data structure implementations should not explicitly leave space for concurrency metadata like locks, and the concurrency mechanism should not come between pieces of the data structure.
+We discuss several undesirable approaches and their relationships to object-orientation patterns, and show that the most desirable approach corresponds to \emph{traits} or \emph{aspects}, though it can also be implemented in C.
+\item We give a formal \textbf{interface} for data structure components and concurrency templates, in the form of separation logic specifications for each of the relevant functions.
+These specifications follow the style of \citet{templates}, but we clarify which specifications should be considered the official interface and which are derived or implementation-specific, as well as ensuring that data structure components are never assumed to include concurrent features and concurrency templates never rely on data structure details.
+We verify the top-level concurrent data structure operations (\lstinline{insert} and \lstinline{lookup}) against these specifications, guaranteeing that they work correctly for any data structure component and template that satisfy the interface.
+\item We exhibit two data structures (linked list and BST) and two concurrency templates (lock coupling and give-up) that satisfy our interface, and from them \textbf{freely obtain four verified concurrent data structures}, implemented in C and verified in VST.
+%The BST is particularly interesting because...
+\end{itemize}
+
+\ignore{
 Concurrent search structure templates~\cite{shasha1988concurrent} are a technique for describing and implementing concurrency patterns (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) that can then be specialized to various search data structures (e.g., linked list, hashtable, B-tree) compositionally.
@@
 \end{itemize}
-
-\subsection*{Related Work}
+}
+
+\ignore{
+\paragraph{Related Work}
 Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}.
@@
 Feldman et al.~\cite{feldman2020proving} take an approach similar to templates to verify search structures with highly optimistic concurrency patterns, where the data structure may be restructured by other threads during traversal.
-Their technique could potentially be applied to prove correctness of much more complex \lstinline{traverse} operations than those we describe.
-
+Their technique could potentially be applied to prove correctness of much more complex \lstinline{traverse} operations than those we describe.}
+
+\ignore{
 \section{Background}
@@
 
-\subsubsection{Logical atomicity} 
+\paragraph{Logical atomicity} 
 \label{logatom}
@@
 %clean this up a little
-
-\section{Search Structure Templates in VST}
+}
+
+\section{Coding Search Structure Templates}
+2-by-2 grid of concurrent node implementations (bst+lock, bst+lock+range, list+lock, list+lock+range)
+
+which pieces are data structure and which are template? (paper assumes locks and ranges are provided by data structure)
+
+difficulty of actually implementing things this way -- even in C++, we'd have to write subclasses for BST-coupling-node, BST-giveup-node, LL-coupling-node, LL-giveup-node, and in Java those classes would have to have code in them
+
+sketch the four approaches, describe why only one works for us?
+
+some of the tricky pieces to separate: node creation/allocation of initial metadata, updating metadata on insert, ...
+
+highlight the interface: which functions must a data structure provide?
+which functions must a template provide?
+which functions are written once generically, parameterized by both data structure and template?
+
+\section{Specifying the Template Interface}
+standard Hoare triples for each data structure function
+
+atomic triples for each template function, with data structure instance as parameter
+
+top-level specs proved once-and-for-all, with data structure and template instance as parameters
+
+\section{Data Structure Instances: Linked List and Binary Search Tree}
+
+\section{Template Instances: Give-Up and Lock Coupling}
+
+\section{Evaluation}
+
+\ignore{\section{Search Structure Templates in VST}
 
@@
 
-A search structure template is an implementation of these three functions, written in a particularly restricted way outlined in Figure~\ref{templates-diagram}.
+A search structure template is an implementation of these three functions, written in a restricted way outlined in Figure~\ref{templates-diagram}.
 Aside from some top-level management code (e.g., acquiring and releasing an initial lock), all that each function should do is call two functions: a helper function called \lstinline{traverse}, which is specific to the template and implements its synchronization mechanism, and a concurrency-unaware function (e.g.
@@
 
-\subsubsection{Verifying the Lock-Coupling Template}
+\paragraph{Verifying the Lock-Coupling Template}
 \label{traverse_proof_lock}
@@
 	typedef struct node_t {
-		node *t; 
-		lock_t *lock; 
+		node *t; lock_t *lock; 
 		int min, max;
@@
 
-\subsubsection{Verifying the Give-Up Template}
+\paragraph{Verifying the Give-Up Template}
 \label{traverse_proof_giveup}
@@
 
-\section{Verifying Binary Search Trees using Templates}
-\label{BST_proof}
-In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified C implementations of concurrent data structures. 
-As we saw in the previous section, the top-level specifications of data structure operations have already been verified at the template level; all we need to do now is provide implementations of data-structure-specific functions (e.g.
-\lstinline{findNext}) that satisfy the right specifications to instantiate the template proofs.
-Once we have chosen a template and a data structure, we should not need to do any further reasoning about specific concurrency patterns.
-
-Our target data structure is a binary search tree (BST) implemented in C.
-A lock-coupling BST was already written verified as a demonstration of VST's concurrency capabilities~\cite{bst-conc}.
-We refactored it into \lstinline{findNext}, \lstinline{traverse}, \lstinline{insertOp}, etc.
-as required by the template approach, and then re-verified it using the template; we then replaced \lstinline{traverse} with the give-up version and verified the resulting give-up BST as well.
-We were able to obtain verified \lstinline{insert} and \lstinline{lookup} operations using this technique; for \texttt{delete}, we found a mismatch between the template approach and BST deletion, which we discuss in Section \ref{internal_reorganize}.
-
-\subsection{Instantiating \lstinline{findNext} and \lstinline{insertOp}}
-We instantiate the templates with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}).
-As expected, their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples (see Figures \ref{fig:findNext_lock} and \ref{fig:insertOp}), and their proofs do not require any concurrency reasoning.
-\begin{figure*}[!ht]
-	\centering
-	\begin{mathpar}
-		{\color{specblue}
-			\left\{ 
-			\begin{array}{c}
-				\texttt{pn->n->t}\mapsto (k', v', l, r)
-			\end{array}
-			\right\}
-		} \vspace{-0.85em} 
-		\\ 
-		\texttt{findNext(pn, k)} 
-		\vspace{-0.85em} 
-		\\
-		{\color{specblue}
-			\left\{\mathit{res}. \ \exists \  n'.
-			\begin{array}{c}
-				\texttt{pn->n->t}\mapsto (k', v', l, r)  \ \ast \\\  
-				\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k')
-			\end{array}
-			\right\}
-		}
-	\end{mathpar}
-	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)}
-	\label{fig:findNext_lock}
-\end{figure*}
-
-\begin{figure*}[h]
-	\begin{subfigure}{\textwidth}
-		\centering
-		\begin{mathpar}
-			{\color{specblue}
-				\left\{ 
-				\begin{array}{c}
-					\ \texttt{pn->n} \mapsto \texttt{NULL}  
-				\end{array}
-				\right\}
-			} \vspace{-0.85em} 
-			\\ 
-			\texttt{insertOp(pn, k, v)} 
-			\vspace{-0.85em} 
-			\\
-			{\color{specblue}
-				\left\{\exists \  l\ r\ \mathit{lk}_1\ \mathit{lk}_2.
-				\begin{array}{l}
-					\ \texttt{pn->n->t} \mapsto (\texttt{k}, \texttt{v}, l, r) \ \ast \ \\ 
-					\ l \mapsto (\texttt{NULL}, \mathit{lk}_1)  \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2)
-				\end{array}
-				\right\}
-			}
-		\end{mathpar}
-		\caption{Specification of \texttt{insertOp} for the lock-coupling template}
-		\label{fig:insertOp_lock}
-	\end{subfigure}
-	%insertOp for give-up of give-up
-	\begin{subfigure}{\textwidth}
-		\begin{mathpar}
-			{\color{specblue}
-				\left\{ 
-				\begin{array}{c}
-					\texttt{pn->n}\mapsto \texttt{NULL}  \ast  \texttt{pn->n->min}\mapsto n_1  \ast  \texttt{pn->n->max}\mapsto n_2
-				\end{array}
-				\right\}
-			} \vspace{-0.85em} 
-			\\
-			\texttt{insertOp(pn, k, v)}
-			\vspace{-0.85em} 
-			\\
-			{\color{specblue}
-				\left\{\exists \ l\ r\ \mathit{lk}_1\ \mathit{lk}_2. 
-				\begin{array}{c}
-					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r)  \ast  \texttt{pn->n->min}\mapsto n_1 \ast  \texttt{pn->n->max}\mapsto n_2  \ \ast 
-					\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
-				\end{array}
-				\right\}
-			}
-		\end{mathpar}
-		\caption{Specification of \texttt{insertOp} for the give-up template}
-		\label{fig:insertOp_giveup}
-	\end{subfigure}
-	\caption{Specification of \texttt{insertOp} for the lock-coupling and give-up templates}
-	\label{fig:insertOp}
-\end{figure*}
-
-The contents of a BST node are defined in C as:
-\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
-	typedef struct node {
-		int key; 
-		void *value; 
-		struct node_t *left, *right;
-	} node;
-\end{lstlisting}
-The BST's \lstinline{findNext} function (Figure \ref{findnext}) indicates whether the target key is in the current node, its left subtree, or its right subtree.
-The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with key $k'$, value $v'$, and pointers to the left $l$ and right $r$ child nodes.
-The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the node's key $k'$, and the next node $n'$ is the left child $l$; \lstinline{k} is greater than $k'$ and $n'$ is the right child $r$; or \lstinline{k} is exactly $k'$ and $n'$ is the original node $n$, in which case the return value is 0.
-The proof is straightforward.
-\begin{figure}[ht]
-	\begin{minipage}{0.45\textwidth}
-	\lstinputlisting[language=C, style=myStyle]{findnext.c} 
-	\end{minipage}
-	\caption{C implementation of \lstinline{findNext} for the binary search tree (all templates)}
-	\label{findnext}
-\end{figure}
-
-The BST's \lstinline{insertOp} function allocates a new \lstinline{node} struct at an empty leaf, as well as two new empty leaves to serve as its children.
-The \texttt{insertOp} specification (Figure \ref{fig:insertOp_lock}) takes as input a node whose \lstinline{t} field is \lstinline{NULL}, i.e., an empty leaf node.
-It inserts a new node with the specified key \texttt{k} and value \texttt{v} at that leaf, complete with two new empty child nodes with corresponding locks, $(\texttt{NULL}, \mathit{lk}_1)$ and $(\texttt{NULL}, \mathit{lk}_2)$, which are pointed to by the $l$ and $r$ of the current node.
-
-Unfortunately, while this specification suffices for the lock-coupling template, it cannot be used as is for the give-up template: because we allocate new empty leaf nodes, we must allocate \lstinline{node_t} structures, whose implementation in the give-up template has extra fields that do not appear in the lock-coupling template.
-As shown in Figure \ref{fig:insertOp_giveup}, a give-up template node (even an empty one!) has additional \lstinline{min} and \lstinline{max} fields, and these fields are needed to compute the bounds of the new child nodes.
-In a BST, if the current node's range is (\lstinline{min, max}) and its key is \lstinline{k}, its left child will have range (\texttt{min, k}) and its right child will have range (\texttt{k, max}).
-It may be possible to remedy this discrepancy by adding an ``allocate empty node'' function to the templates, which is then called by \lstinline{insertOp}, but this would still complicate the approach: originally, we expected that templates would be parameterized by data structure implementations but not vice versa.
-However, this two-way dependency seems fundamental when the layout of the node structure depends on the concurrency control mechanism we choose.
-For now, we break the abstraction boundary and verify two different versions of \lstinline{insertOp}, one for the lock-coupling template and one for the give-up template, which reduces the proof-reuse benefits of the template approach.
-
-%The \texttt{insert} specification for the give-up template is identical to that of the lock-coupling template, up to changes in the definitions of $\nodeboxrep\ \texttt{p}$ and $\treerep \ \texttt{m}$.
-%Essentially, the \texttt{insert} function's specification comprises two main components: local and public preconditions and postconditions, as previously mentioned.
-Moreover, we must demonstrate that it is possible to atomically update the tree's abstract state from an unknown \texttt{m} to $\texttt{m[k} \mapsto \texttt{v])}$.
-%The local preconditions and postconditions, containing $\nodeboxrep \ \texttt{p}$, depict the binary search tree's concrete state, where the root node $\texttt{p}$ points to the node's lock and footprint predicate, signifying that the root node is located within the search structure's footprint.
-In contrast, the public preconditions and postconditions are denoted by the predicate  $\treerep \ \texttt{m}$, which is identical to the public predicates of the \texttt{traverse} function in the give-up template.
-
-%The proof for the \texttt{insert} operation begins with the preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$.
-We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_giveup}).
-In Section \ref{traverse_proof_giveup}, we discuss providing the identical preconditions for the \texttt{traverse} function, which encompasses the local precondition $\infp\texttt{(pn->n)}$ and the public precondition $\treerep\ \texttt{m}$.
-Subsequently, we apply the \texttt{traverse} specification from Section \ref{traverse_proof_giveup}.
-
-%Analogous to the lock-coupling template's \texttt{insert} operation, there are two possible branches representing the outcome following the execution of \texttt{traverse}.
-Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ \texttt{(m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.
-
-\subsection{Templates and Internal Reorganization}
-\label{internal_reorganize}
-The \lstinline{delete} operation on binary search trees highlights a limitation of the template approach as we have described it.
-The \lstinline{insert} and \lstinline{lookup} operations can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insertOp} or \lstinline{lookupOp}).
-In most of the prior template examples, deletion can be decomposed similarly, by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation.
-When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
-rotating three-node sections of the tree until the node to be deleted is at a leaf.
-This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them).
-
-Krishna et al.
-also discuss an operation of this sort, the \lstinline{split} operation in B-link trees.
-Their approach is to assume the existence of a separate maintenance thread that constantly crawls the tree searching for full nodes and splitting them; any operation that cannot take effect on full nodes can simply retry until the target node is split.
-The split itself is a no-op on the abstract state of the data structure: it makes internal structural changes but does not change the set of keys stored in the tree.
-Maintenance operations of this sort appear in almost every reasonably complex concurrent data structure, and so the template approach must account for their existence.
-In other words, in general the template approach must divide the operations of a target data structure into three categories: concurrent, data-structure-agnostic template operations, which can be verified once and applied to multiple data structures; sequential data structure operations, which can be verified once and plugged into multiple templates; and no-op maintenance operations, which are both concurrency-aware and data-structure-specific, but whose specifications leave the abstract state unchanged.
-For instance, rotation in the BST can be seen as a maintenance operation, and the deletion of a node that has already been rotated to a leaf position is a sequential, data-structure-specific operation.
-Rebalancing operations in AVL and red-black trees would also be considered maintenance operations.
-The existence of this third category means that we cannot completely separate concurrency reasoning from data structure reasoning, but because maintenance operations are logical no-ops, they should be easier to verify in a non-decomposed style than most concurrent data structure operations.
-
-\section{Proof Mechanization}
-\label{proof_mech}
-All the proofs described above have been mechanized in VST, using its extensions for logical atomicity~\cite{iris-vst-arxiv}. Statistics on the verification effort are shown in Table~\ref{table_coq}.
-
-\begin{table}[h]
-	\centering
-	\resizebox{\columnwidth}{!}{
-		\begin{tabular}{l | l || l | l} 
-			\hline
-			\textbf{Lock coupling} & \textbf{LoC} & \textbf{Give up} & \textbf{LoC} \\
-			\hline
-			Specifications & 190 & Specifications & 138 \\
-			\texttt{findNext} proof & 27 & \texttt{findNext} proof & 27 \\
-			\texttt{insertOp} proof & 32 & \texttt{insertOp} proof & 31 \\
-			& & \texttt{inRange} proof & 30 \\
-			\texttt{traverse} proof & 360 & \texttt{traverse} proof & 336 \\ 
-			\texttt{insert} proof & 450 & \texttt{insert} proof & 284 \\
-			\texttt{lookup} proof & 235 & \texttt{lookup} proof & 304 \\ 
-			\texttt{pushdown\_left} proof & 282 & & \\	
-			\texttt{delete} proof & 182 & & \\
-			Supporting proofs & 1909 & Supporting proofs & 1876 \\
-			\hline
-			\textbf{Total} & \textbf{3667} & & \textbf{3026}  
-		\end{tabular}
-	}
-	\caption{Size of Coq definitions and proofs, by topic.}
-	\label{table_coq}
-\end{table}
-
-%By utilizing VST, we can create programs in standard C, automatically generate abstract syntax trees (ASTs) for them in Coq, and subsequently state and interactively prove specifications for these programs with the support of automated tactics.
-Table \ref{table_coq} presents a summary of our Coq development.
-We formalized all previously described proofs in Coq, taking advantage of VST's features for verifying separation logic specifications of C functions.
-This encompasses supplementary proofs, template specifications (encompassing all helper functions such as \texttt{findNext}, \texttt{insertOp}, and others, in addition to \texttt{traverse}, \texttt{insert}, \texttt{lookup}, and \texttt{delete}), as well as proofs against with lock-coupling and give-up template specifications.
-%\wm{Rather than re-stating what we proved, this is the place to discuss the development: e.g., which parts took the most effort and LoC, which parts we expect to be reusable and in what way, how much bigger or smaller the template proofs were than non-template versions, etc.}
-
-The total lines of code required were about the same as for the original VST verification of the BST.
-The vast majority of the proof effort was in the templates, in both \lstinline{traverse} and the top-level data structure operations.
-This is fairly encouraging, since this proof effort should be reusable for other data structures that use the same templates.
-For lock-coupling deletion, we retained the original (non-templated) proofs; we did not verify give-up deletion, since optimistic deletion is considerably harder to reason about and did not stand to benefit from the template approach.
-%Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process.
-Otherwise, the construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, necessitated a considerable amount of time and effort.
-%However, we expect that functions such as \texttt{traverse}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.
-The methods \texttt{findNext} and \texttt{insertOp} may also be adapted to other data structures, such as linked lists, with some additional effort.
-Proving the correctness of e.g.
-a lock-coupling linked list should be as simple as defining a new \lstinline{node} struct, and then implementing and verifying \lstinline{findNext} and \lstinline{insertOp} (plus the appropriate functions for deletion).
-%Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext}, \texttt{inRange}, and \texttt{insertOp} to restructure the \texttt{node} struct.
-While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future.
-
-Compared to prior template proofs in Iris and GRASShopper, our proof development is significantly larger (although direct comparison is difficult, since GRASShopper is an automatic prover).
-The biggest structural difference is that we need to directly state loop invariants for our imperative code, while the Iris proofs handle recursion with a simpler Löb-induction approach.
-Coming up with the loop invariants was the hardest part of the proofs.
-Our proofs also deal with the details of real C code: proving absence of integer overflows, accurately representing struct-and-pointer-based data structures instead of functional-programming-style structured data, etc.
-In exchange for this extra effort, the code we verify constitutes complete C programs that can be compiled and executed.
-The full development of our mechanization effort is available online at \href{https://zenodo.org/record/8337004}{https://zenodo.org/record/8337004}.
-
-%[link redacted for anonymity].
-
-%\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.
+}
 
 \section{Conclusion and Future Work}
-Concurrent search structure templates are a promising approach for proving the correctness of concurrency patterns and data structures separately, and then combining them to obtain verified concurrent data structures.
-We have translated the approach to apply to C programs using VST, with appropriate imperative versions of the key \lstinline{traverse} functions that define the templates, and used it to prove the correctness of an (appropriately refactored) existing concurrent data structure.
-However, the translation also exposed some ambiguities and limitations in the approach: for instance, the \lstinline{inRange} function of the give-up template must either be assumed to exist in the data structure or incorporated into the template, and the \lstinline{delete} operation of binary search trees does not decompose naturally into a concurrent part and a data-structure-specific part.
-Templates still allow us to reuse proofs across data structures and concurrency patterns, but not as completely as the theoretical approach suggests.
-
-The next step is to implement templates as a truly generic framework, with proofs of \lstinline{traverse}, etc.
-that can be freely combined with verified sequential data structures via a clear interface that defines the functions a data structure must provide and the assumptions the template makes on them (e.g., by verifying template functions with respect to a data-structure typeclass).
-We are also interested in verifying templates that use atomic operations instead of locks, as is common in optimistic concurrency control.
-Ultimately, we hope to use the template approach to verify data structures such as Masstree~\cite{masstree} and Wormhole~\cite{wormhole}, complicated real-world search structures that combine multiple concurrency patterns for maximum performance on multicore architectures.
-
+The framework we have presented fulfills the true promise of the concurrency template approach: each data structure instance is implemented and verified without any awareness of concurrency or metadata, each template instance is implemented and proved without any knowledge of the implementation of the data structure, and we obtain verified concurrent data structures immediately by instantiating our top-level theorems with any combination of data structure and template instances.
+We accomplished this by careful attention to which components belong to the data structure and which to the template, and by working to disentangle the two in both the C implementation and the Coq specification and proof.
+Our interfaces are general enough that it should be easy to add more data structures and templates, including concurrency patterns that do not depend on locks (e.g., optimistic concurrency control with fine-grained atomics).
+We have seen that the concurrency template must be implemented in a very specific way, with metadata stored separately from the data structure itself, in order to sufficiently separate concerns; in future work, we are interested to see whether languages with more flexible composition mechanisms, like traits in Rust and Scala, lend themselves to more natural implementations of this composition.
+
+\ignore{
 \begin{acks}
-We thank Roshan Sharma, Alex Oey, and Anastasiia Evdokimova for extensive work on the original binary search tree implementation and verification, and the anonymous reviewers for providing comprehensive and insightful reviews.
-This work was partially funded by the National Science Foundation under the award 1005849 (Veriﬁed High Performance Data Structure Implementations, Beringer \& Mansky) and the Defense Advanced Research Projects Agency (DARPA) under contract HR001120C0160.
+We thank Nisarg Patel, ...
+This work was partially funded by the National Science Foundation under the award ???.
 \end{acks}
-
-%12 pages
+}
+
+\clearpage
 
@@
 
-%%
-%% If your work has an appendix, this is the place to put it.
-\appendix
-
-\section{Proof Outlines for Templates (Sections \ref{traverse_proof_lock} and \ref{traverse_proof_giveup})}
-\label{sec:apd_proof}
-Below are proof outlines for the \texttt{traverse} function for the give-up template (Figure \ref{proof_giveup_traverse}) as well as the \texttt{insert} function for both templates.
-The complete proofs can be found in our Coq mechanization (see Section \ref{proof_mech}).
-\begin{figure*}[h]
-	$\color{specblue}
-	\fforall \  m. \left\langle
-	\begin{array}{c}
-		\texttt{pn} \mapsto (p, n)  \ast  \inFP(n) \ \big| \ \treerep\ (m)
-	\end{array}
-	\right\rangle$
-	\begin{minipage}{0.9\textwidth}
-	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
-	\end{minipage}
-	$\color{specblue}
-	\left\langle \mathit{res}. \ \exists \  n', v.
-	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \inFP(n') \ \ast \ \texttt{k} \in \mathit{range}\ \ast \ 
-		\\ 
-		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ 
-		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathit{range}))
-	\end{array}
-	\ \Bigg| \ \treerep\ (m) \
-	\right\rangle$
-	\caption{Proof outline of the give-up \texttt{traverse} function}
-	\label{proof_giveup_traverse}
-\end{figure*}
-
-\begin{figure*}[h]
-	\begin{subfigure}{\textwidth}
-		\centering
-		$\color{specblue}
-		\fforall \  m. \left\langle
-		\begin{array}{c}
-			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
-		\end{array}
-		\right\rangle$
-		\begin{minipage}{0.9\textwidth}
-		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
-		\end{minipage}
-		$\color{specblue}
-		\left\langle
-		\begin{array}{c}
-			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
-		\end{array}
-		\right\rangle$
-		\caption{Proof outline of the lock-coupling \texttt{insert} function}
-		\label{proof_lock_insert}	
-	\end{subfigure}\qquad
-
-	\begin{subfigure}{\textwidth}
-		\centering
-		$\color{specblue}
-		\fforall \  m. \left\langle
-		\begin{array}{c}
-			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
-		\end{array}
-		\right\rangle$
-		\begin{minipage}{0.9\textwidth}
-		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
-		\end{minipage}
-		$\color{specblue}
-		\left\langle
-		\begin{array}{c}
-			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
-		\end{array}
-		\right\rangle$
-		\caption{Proof outline of the give-up \texttt{insert} function}
-		\label{proof_giveup_insert}
-	\end{subfigure}
-	\caption{Proof outlines for the \texttt{insert} function}
-	\label{proof_lock_giveup_insert}
-\end{figure*}
 \end{document}
-\endinput
-%%
-%% End of file `sample-sigplan.tex'.

### 2025-02-11 concurrency/template paper/paper.tex
background and motivation

@@
 This is a surprisingly challenging task: ideally, the data structure implementations should not explicitly leave space for concurrency metadata like locks, and the concurrency mechanism should not come between pieces of the data structure.
-We discuss several undesirable approaches and their relationships to object-orientation patterns, and show that the most desirable approach corresponds to \emph{traits} or \emph{aspects}, though it can also be implemented in C.
+We discuss several undesirable approaches and their relationships to object-orientation patterns, and show that the most desirable approach corresponds to \emph{traits}, though it can also be implemented in C.
 \item We give a formal \textbf{interface} for data structure components and concurrency templates, in the form of separation logic specifications for each of the relevant functions.
@@
 
-\ignore{
 \section{Background}
@@
 
-A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria.
-In the abstract, a search structure implements a map from keys to values, providing operations such as search, insertion, deletion, and traversal on that map.
-For efficiency, many search structures allow concurrent access and modification, often employing fine-grained or lock-free concurrency to allow as many threads as possible to operate on separate parts of the data structure.
-Designing these search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory, including cache behavior~\cite{masstree}.
-The complex interplay between concurrency and data structure design in concurrent search structures makes them challenging targets for formal verification.
-
-The Concurrent Search Structure Template approach of Krishna et al.~\cite{templates} aims to make the problem tractable by separating verification of concurrency control patterns from verification of the underlying data structure.
-Addressing the proof of each component separately makes the individual proofs easier, and also offers the possibility of proof reuse: each verified concurrency pattern (``template'') can be applied to many different data structures, and each verified sequential data structure can be outfitted with many different templates.
-In theory, verifying $n$ templates and $m$ data structures might yield $n \times m$ verified concurrent data structures; in practice, as we will see, both templates and data structures make assumptions that may invalidate certain combinations.
-%how much more can we say about this?
-As a side note, the example templates verified by Krishna et al.~and the ones we present here focus on synchronization via locks, but the approach also applies to lock-free synchronization (as long as it is sequentially consistent; weak-memory template reasoning has not yet been investigated).
-
+Concurrent search structure templates~\cite{templates} are an abstraction technique for decomposing the implementation/verification of a concurrent search structure (i.e., a data structure that supports lookup, insert, and possibly delete operations) into a concurrency part (the \emph{template} and a data structure part.
+A data structure implements a \lstinline{node} type and core \emph{local} operations on those nodes, including lookup, insert, etc.
+but also some helper functions.
+A template implements traversal and top-level data structure operations, interacting with the underlying nodes only via the specified functions.
+In this way, any data structure that implements the appropriate functions can be plugged into the template to yield a concurrent data structure.
 \begin{figure}[h]
@@
 Once the appropriate node for the key has been found, the template returns it to a top-level function such as \lstinline{insert} that calls out to the data structure to perform the actual insertion on the node.
-Thus, the \lstinline{traverse} and \lstinline{insert} functions can be verified without knowing anything about the target data structure other than its synchronization mechanism, as long as the data structure implements \lstinline{findNext} and \lstinline{insertOp} operations with the required semantics.
-The concurrency functions in the template and the sequential \lstinline{findNext} and \lstinline{insertOp} functions provided by the data structure combine into a fully operational concurrent data structure.
-Krishna et al.~specify templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures (i.e., those in which a node may be reachable along multiple paths), but many common search structures do not require flow-style reasoning; in this paper, we present a version of templates that is independent of flow interfaces.
-%is this a contribution in itself?
-%what more do we want to say here?
-
+Thus, the \lstinline{traverse} and \lstinline{insert} functions can be written and verified without knowing anything about the target data structure other than its synchronization mechanism, as long as the data structure implements \lstinline{findNext} and \lstinline{insertOp} operations with the required semantics.
+
+The verification mechanism for search structure mechanisms is concurrent separation logic, and the interface between data structures and templates relies on two key constructs: \emph{flow interfaces} and \emph{logical atomicity}.
+Flow interfaces~\cite{krishna2017flow} are a generic mechanism for capturing the relationship of a single node to a larger data structure, and can be implemented in separation logic using ghost state.
+The interface provided by a data structure implementation includes a predicate of the form $\mathsf{node}\ n\ I\ C$, where $n$ is the node itself (i.e., a pointer to the node data structure), $I$ is $n$'s flow interface, and $C$ is $n$'s contribution to the state of the overall data structure, e.g., the set of keys contained in $n$.
+Data structure operations such as \lstinline{insert}, \lstinline{lookup}, and \lstinline{findNext} are specified in terms of the $\mathsf{node}$ predicate.
+%example?
+
+Logical atomicity~\cite{tada} is used to lift a sequential data structure specification to the concurrent setting.
+A logically atomic triple ${\color{specblue} \fforall a.\,\left\langle \texttt{P}_p(a) \right\rangle } \ \texttt{c}\ {\color{specblue}\left\langle \texttt{Q}_p(a)\right\rangle}$ asserts that the program \texttt{c} atomically updates the abstract data $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread.
+For instance, the top-level specification for the \lstinline{insert} operation on a (linearizable) concurrent data structure can be written as 	\[{\color{specblue}
+		\fforall m.\left\langle 
+		\treerep(m)
+		\right\rangle
+	}\ \texttt{insert(r, k, v)}\ 
+	{\color{specblue}
+		\left\langle 
+		\treerep(m[\texttt{k} \mapsto \texttt{v}])
+		\right\rangle
+	}\]
+saying that \lstinline{insert} atomically updates the state of the data structure from $\treerep(m)$ to $\treerep(m[\texttt{k} \mapsto \texttt{v}])$, without mentioning the details of either the synchronization mechanism or the underlying data structure implementation.
+The \lstinline{traverse} function for a given template is proved to satisfy a logically atomic specification that says roughly ``this function finds the node where key \lstinline{k} belongs''.
+The \lstinline{traverse} specification can then be used to prove atomic specifications for the data structure operations, lifting the sequential specifications for insert, lookup, etc.
+to the concurrent setting.
+
+Krishna et al., following Shasha and Goodman's original work on template algorithms~\cite{shasha1988concurrent}, define three templates: lock coupling, give-up, and link.
+The lock coupling template acquires locks on both parent and child node when moving from one to the other, guaranteeing that the link followed is never out of date.
+The link template only acquires a lock on the current node, and its \lstinline{traverse} may return an invalid node; in this case, the operation (insert, lookup, etc.) may fail, and if it does then the operation restarts from the root of the data structure.
+The give-up template stores an explicit range of expected keys in each node, and checks this range at each node it traverses, restarting the traversal if it ever reaches a node whose range does not include the target key.
+For each of these templates, they implement and verify a \lstinline{traverse} function and use it to implement and verify the top-level data structure operations.
+
+\ignore{
 \paragraph{Logical atomicity} 
@@
 \section{Coding Search Structure Templates}
-2-by-2 grid of concurrent node implementations (bst+lock, bst+lock+range, list+lock, list+lock+range)
-
-which pieces are data structure and which are template? (paper assumes locks and ranges are provided by data structure)
-
-difficulty of actually implementing things this way -- even in C++, we'd have to write subclasses for BST-coupling-node, BST-giveup-node, LL-coupling-node, LL-giveup-node, and in Java those classes would have to have code in them
-
-sketch the four approaches, describe why only one works for us?
+\begin{figure}
+  \begin{subfigure}[t]{.4\textwidth}
+  \begin{lstlisting}[language=C, style=myStyle]
+  struct node{ int key; int value;
+    node* next;
+    lock l; }
+  \end{lstlisting}
+    \caption{Linked-list node with lock}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[t]{.4\textwidth}
+  \begin{lstlisting}[language=C, style=myStyle]
+  struct node{ int key; int value;
+    node* next;
+    lock l; int min; int max; }
+  \end{lstlisting}
+    \caption{Linked-list node with lock and range}
+  \end{subfigure}
+
+  \medskip
+
+  \begin{subfigure}[t]{.4\textwidth}
+  \begin{lstlisting}[language=C, style=myStyle,numbers=none]
+  struct node{ int key; int value;
+    node* left; node* right;
+    lock l; }
+  \end{lstlisting}
+    \caption{BST node with lock}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[t]{.4\textwidth}
+  \begin{lstlisting}[language=C, style=myStyle]
+  struct node{ int key; int value;
+    node* left; node* right;
+    lock l; int min; int max; }
+  \end{lstlisting}
+    \caption{BST node with lock and range}
+  \end{subfigure}
+  \caption{Four different \lstinline{node} implementations}
+  \label{node-impls}
+\end{figure}
+
+Suppose we have implemented four concurrent data structures in C: a linked list and a binary search tree, each with both the lock-coupling and give-up synchronization patterns.
+Figure~\ref{node-impls} shows the definition of the \lstinline{node} type for each of these implementations.
+All four node types include a key and value field; linked-list nodes have a \lstinline{next} field, while BST nodes have a \lstinline{left} and \lstinline{right} child.
+Both kinds of nodes have a lock for synchronization; give-up nodes also have a range (\lstinline{min} and \lstinline{max}) indicating the range of keys allowed in this node and its children.
+In a modular approach, we should be able to define each component separately (linked list, BST, lock-coupling, give-up), and then freely combine them to yield these four node implementations.
+
+This modular approach already stands in contrast to Krishna et al.~\cite{templates}, where functions like \lstinline{lockNode} and \lstinline{inRange} are assumed to be provided by the data structure.
+This is convenient from one perspective, since the data structure is entirely responsible for implementing and manipulating the \lstinline{node} struct, and the template can interact with it only via interface functions; on the other hand, it runs counter to the idea that the data structure implementation is ``concurrency-unaware'', since it must include locks in its implementation.
+Furthermore, requiring the data structure to provide template-specific features like \lstinline{inRange} rules out any possibility of modularity, since the template can then only be applied to data structures that implement its specific requirements (and a large enough set of templates may have mutually inconsistent requirements).
+Instead, we propose that each template should be responsible for providing whatever fields it requires, and each data structure should only implement the fields that make up the core sequential data structure (e.g., \lstinline{key}, \lstinline{value}, and \lstinline{next}).
+
+Writing code that composes in this way is quite difficult in most languages.
+The most direct analogue is Rust and Scala's traits...
+In languages with multiple inheritance like C++, we could define a \lstinline{BST_giveup_node} that inherits fields and methods from \lstinline{BST_node} and \lstinline{giveup_node}, but we would still have to declare a class for each combination of data structure and template.
+Writing the template node as a wrapper around the data structure node, or vice versa, entangles the two in a way that makes both programming and proving less modular.
+For our purposes, and working in C, we settle for a highly compositional but theoretically unsatisfying approach: we store template and data structure fields separately, with template maintaining a hash table that maps each data structure node to its associated template fields.
 
@@
 We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
-\begin{lstlisting}[style=myStyle,language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, numbers=none, xleftmargin=0.5em]
+\begin{lstlisting}[style=myStyle,language=C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, numbers=none, xleftmargin=0.5em]
 	typedef struct node_t {

### 2025-02-12 concurrency/template paper/paper.tex
describe code decomposition for templates

@@
 
-This modular approach already stands in contrast to Krishna et al.~\cite{templates}, where functions like \lstinline{lockNode} and \lstinline{inRange} are assumed to be provided by the data structure.
+This modular approach stands in contrast to Krishna et al.~\cite{templates}, where functions like \lstinline{lockNode} and \lstinline{inRange} are assumed to be provided by the data structure.
 This is convenient from one perspective, since the data structure is entirely responsible for implementing and manipulating the \lstinline{node} struct, and the template can interact with it only via interface functions; on the other hand, it runs counter to the idea that the data structure implementation is ``concurrency-unaware'', since it must include locks in its implementation.
@@
 Instead, we propose that each template should be responsible for providing whatever fields it requires, and each data structure should only implement the fields that make up the core sequential data structure (e.g., \lstinline{key}, \lstinline{value}, and \lstinline{next}).
+
+\begin{figure}
+  \begin{subfigure}[t]{.4\textwidth}
+  \begin{lstlisting}[language=C, style=myStyle]
+  struct node{ int key; int value;
+    node* next; }
+  \end{lstlisting}
+    \caption{Linked-list node}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[t]{.4\textwidth}
+  \begin{lstlisting}[language=C, style=myStyle]
+  struct md_entry{ lock l;
+    int min; int max; }
+  \end{lstlisting}
+    \caption{Give-up metadata}
+  \end{subfigure}
+
+  \medskip
+
+  \begin{subfigure}[t]{.4\textwidth}
+  \begin{lstlisting}[language=C, style=myStyle,numbers=none]
+  struct css{ node* root;
+    md_table* metadata; }
+  \end{lstlisting}
+    \caption{Top-level concurrent data structure}
+  \end{subfigure}
+  \caption{Modular definition of concurrent search structures}
+  \label{node-md}
+\end{figure}
 
@@
 Writing the template node as a wrapper around the data structure node, or vice versa, entangles the two in a way that makes both programming and proving less modular.
-For our purposes, and working in C, we settle for a highly compositional but theoretically unsatisfying approach: we store template and data structure fields separately, with template maintaining a hash table that maps each data structure node to its associated template fields.
-
-some of the tricky pieces to separate: node creation/allocation of initial metadata, updating metadata on insert, ...
-
-highlight the interface: which functions must a data structure provide?
-which functions must a template provide?
-which functions are written once generically, parameterized by both data structure and template?
+For our purposes, and working in C, we settle for a nonlocal but highly compositional approach: we store template and data structure fields separately, with template maintaining a hash table that maps each data structure node to its associated template fields, as shown in Figure~\ref{node-md}.
+Each data structure implements a \lstinline{node} type, each template implements an \lstinline{md_entry} type, and the top-level \lstinline{css} type is defined once and for all.
+
+Once we have decomposed the data structure type, we can then implement the functions for each component.
+The data structure defines the local \lstinline{findNext}, \lstinline{lookupOp}, and \lstinline{insertOp} functions as operations on \lstinline{node}s; the template defines \lstinline{traverse}, as well as an \lstinline{insertHelper} function to maintain metadata on updates, using the details of \lstinline{md_entry} but treating the \lstinline{node} type as a black box; and the top-level functions \lstinline{insert}, \lstinline{lookup}, etc.
+are defined once and for all on \lstinline{css} by calling the template functions, generic in the the implementation of both \lstinline{node} (the data structure) and \lstinline{md_entry} (the template).
 

### 2025-02-12 concurrency/template paper/paper.tex
fixed css code
outline template specs

@@
 \section{Specifying the Template Interface}
-standard Hoare triples for each data structure function
-
-atomic triples for each template function, with data structure instance as parameter
-
-top-level specs proved once-and-for-all, with data structure and template instance as parameters
+Our verification process for concurrent search structure templates follows the same modular architecture as the code itself:
+\begin{itemize}
+\item The interface for data structures is an abstract $\mathsf{node}$ predicate, plus sequential Hoare triples for the data structure functions (\lstinline{findNext}, \lstinline{insertOp}, \lstinline{lookupOp}) that characterize their behavior at the level of $\mathsf{node}$s.
+\item The interface for templates takes an arbitrary data structure as a parameter, and gives logically atomic triples for the template functions (\lstinline{traverse}, \lstinline{insertHelper}) in terms of their effects on an abstract $\mathsf{CSS}$ predicate.
+\item The top-level \lstinline{insert} and \lstinline{lookup} functions are specified and verified once and for all, taking both a data structure and a template as a parameter, and providing proofs that the top-level \lstinline{insert} and \lstinline{lookup} functions satisfy atomic specifications on $\mathsf{CSS}$:
+\[{\color{specblue}
+		\fforall m.\left\langle 
+		\mathsf{CSS}(m)
+		\right\rangle
+	}\ \texttt{insert(r, k, v)}\ 
+	{\color{specblue}
+		\left\langle 
+		\mathsf{CSS}(m[\texttt{k} \mapsto \texttt{v}])
+		\right\rangle
+	}\]
+\[{\color{specblue}
+		\fforall m.\left\langle 
+		\mathsf{CSS}(m)
+		\right\rangle
+	}\ \texttt{lookup(r, k)}\ 
+	{\color{specblue}
+		\left\langle v.\ 
+		\mathsf{CSS}(m) \land m(k) = v
+		\right\rangle
+	}\]
+\end{itemize}
+In this section, we present the precise specifications for the data structure and template functions.
+The structure of the specifications guarantees compositionality: if we have $m$ data structures that satisfy the data structure interface, and $n$ templates that satisfy the template interface, we can freely combine them to get $m \times n$ verified concurrent search structures.
+As is typical when defining interfaces, the key challenge is to define specifications for the data structure functions that give enough information to verify any template, but are general enough to be satisfied by any data structure.
+...
 

### 2025-02-24 concurrency/template paper/paper.tex
cite prior work

@@
   struct css{ node* root;
-    md_table* metadata; }
+    md_entry* metadata[TABLE_SIZE]; }
   \end{lstlisting}
@@
 In languages with multiple inheritance like C++, we could define a \lstinline{BST_giveup_node} that inherits fields and methods from \lstinline{BST_node} and \lstinline{giveup_node}, but we would still have to declare a class for each combination of data structure and template.
-Writing the template node as a wrapper around the data structure node, or vice versa, entangles the two in a way that makes both programming and proving less modular.
+Writing the template node as a wrapper around the data structure node (as done by \citet{vst-templates}), or vice versa, entangles the two in a way that makes both programming and proving less modular.
 For our purposes, and working in C, we settle for a nonlocal but highly compositional approach: we store template and data structure fields separately, with template maintaining a hash table that maps each data structure node to its associated template fields, as shown in Figure~\ref{node-md}.
@@
 
-\ignore{\section{Search Structure Templates in VST}
-
-\subsection{What is a Search Structure Template?}
-Krishna et al.~\cite{templates} described the search structure template approach and used it to verify (parts of) several concurrent data structures.
-Their examples strongly suggest a systematic approach to separating sequential data structure logic from concurrent synchronization patterns.
-However, the approach is not formally defined, and on closer inspection its inputs and outputs are not identical across the example templates.
-In this section, we attempt to precisely describe the pieces of the search structure template approach, and how we can know whether we have successfully verified a data structure given a collection of verified sequential and concurrent functions.
-
-A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}.
-It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads.
-In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values (we omit the ghost identifier $g$ connecting $\nodeboxrep$ and $\treerep$, as described in section~\ref{logatom}):
-\begin{mathpar}
-	{\color{specblue}
-		\fforall m.\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
-		\right\rangle
-	} \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
-	{\color{specblue}
-		\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
-		\right\rangle
-	}
-\\
-	{\color{specblue}
-		\fforall m.\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
-		\right\rangle
-	} \vspace{-0.8em}
-	\\ \texttt{lookup(r, k)}\  \vspace{-0.8em} \\
-	{\color{specblue}
-		\left\langle \texttt{v}. \ \
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m) \land m(\texttt{k}) = \texttt{v}
-		\right\rangle
-	}
-	\\
-	{\color{specblue}
-		\fforall m.\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
-		\right\rangle
-	}  \vspace{-0.8em} 
-	\\ \texttt{delete(r, k)}\ \vspace{-0.8em}\\
-	{\color{specblue}
-		\left\langle 
-		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
-		\right\rangle
-	}
-\end{mathpar}
-In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation.
-Krishna et al.
-use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.
-
-\begin{figure}[htb]
-\ctikzfig{templates}
-	\caption{Diagram of the template approach. Each pair of template and data structure should yield a verified concurrent data structure.}
-	\label{templates-diagram}
-\end{figure}
-
-A search structure template is an implementation of these three functions, written in a restricted way outlined in Figure~\ref{templates-diagram}.
-Aside from some top-level management code (e.g., acquiring and releasing an initial lock), all that each function should do is call two functions: a helper function called \lstinline{traverse}, which is specific to the template and implements its synchronization mechanism, and a concurrency-unaware function (e.g.
-\lstinline{insertOp}) provided by the target data structure.
-(In fact, Krishna et al.
-implemented all three operations as a single function \lstinline{decisiveOp}, parameterized by a function that takes the target operation type as input and performs the appropriate data structure operation; for ease of adaptation to C, we write three different functions, but their structures are identical.) A template consists of an implementation of the top-level operations and the \lstinline{traverse} function used to define them; a data structure consists of an implementation of \lstinline{insertOp}, etc., as well as a function \lstinline{findNext} that is used to implement \lstinline{traverse}.
-The interface between them is a set of standard specifications for \lstinline{insertOp}, etc., and \lstinline{findNext}: if a template's functions can be verified assuming these standard specifications, and a data structure's implementations can be shown to meet those specifications, then the template and the data structure can be combined to yield a verified concurrent search structure.
-
-This is the theory behind the template approach to verification.
-Unfortunately, it is not straightforward to realize in practice.
-In the Coq development of Krishna et al., different templates use slightly different specifications for data structure operations and \lstinline{findNext}; the specifications proved for data structure implementations do not always match those used by templates (which is possible because they use separate provers for the sequential and concurrent proofs); and some data structures rely on other functions that do not appear in the templates (e.g., maintenance operations that split full nodes in B-trees).
-In the following presentation, we follow the template plan as far as possible, noting the places where we must deviate from the uniform approach; we discuss maintenance operations in further detail in Section~\ref{internal_reorganize}.
-
-\subsection{Defining a Template}
-\label{templates}
-
-The core of a search structure template is a \lstinline{traverse} function that navigates a data structure using the chosen synchronization mechanism.
-A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well.
-The defining feature of all these functions is that they abstract away from the data structure being traversed: they operate on a generic \lstinline{node} type and take data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters.
-We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
-\begin{lstlisting}[style=myStyle,language=C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, numbers=none, xleftmargin=0.5em]
-	typedef struct node_t {
-		node *t; 
-		lock_t *lock;
-	} node_t;
-\end{lstlisting}
-There are two things worth noting about this data structure.
-First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding locks on as few nodes as possible to allow other parts of the data structure to be modified concurrently.
-Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as
-\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
-	typedef struct node {
-		int key; 
-		void *value; 
-		struct node_t *left, *right;
-	} node;
-\end{lstlisting}
-while a linked list might use
-\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
-	typedef struct node {
-		int key; 
-		void *value; 
-		struct node_t *next;
-	} node;
-\end{lstlisting}
-This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component.
-Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns, and then prove (assuming appropriate specifications for data-structure-specific functions) that they correctly implement the data structure operations.
-
+\ignore{
 \subsection{Lock-Coupling Template}
@@
 Our interfaces are general enough that it should be easy to add more data structures and templates, including concurrency patterns that do not depend on locks (e.g., optimistic concurrency control with fine-grained atomics).
-We have seen that the concurrency template must be implemented in a very specific way, with metadata stored separately from the data structure itself, in order to sufficiently separate concerns; in future work, we are interested to see whether languages with more flexible composition mechanisms, like traits in Rust and Scala, lend themselves to more natural implementations of this composition.
+%We have seen that the concurrency template must be implemented in a very specific way, with metadata stored separately from the data structure itself, in order to sufficiently separate concerns; in future work, we are interested to see whether languages with more flexible composition mechanisms, like traits in Rust and Scala, lend themselves to more natural implementations of this composition.
 

### 2025-04-28 concurrency/template paper/paper.tex
outlined interfaces

@@
 \newcommand{\inFP}{\ensuremath{\mathsf{In }}}
+\newcommand{\node}[3]{\ensuremath{\mathsf{node}(#1, #2, #3)}}
+\newcommand{\mdentry}[3]{\ensuremath{\mathsf{md\_node}(#1, #2, #3)}} %will need more arguments -- what are the arguments to md_entry_rep?
 
@@
 In fact, close examination of the examples shows that different templates use different and incompatible specifications for the data structure component.
-In this paper, we rectify this by presenting a \textbf{formal interface for concurrent search structure templates}, implemented as a library in C and a collection of typeclasses in Coq, that guarantees $m \times n$ verified data structures from $m + n$ components.
+In this paper, we rectify this by presenting a \textbf{formal interface for concurrent search structure templates}, implemented as a library in C and a collection of typeclasses in Rocq, that guarantees $m \times n$ verified data structures from $m + n$ components.
 More specifically:
@@
 Flow interfaces~\cite{krishna2017flow} are a generic mechanism for capturing the relationship of a single node to a larger data structure, and can be implemented in separation logic using ghost state.
-The interface provided by a data structure implementation includes a predicate of the form $\mathsf{node}\ n\ I\ C$, where $n$ is the node itself (i.e., a pointer to the node data structure), $I$ is $n$'s flow interface, and $C$ is $n$'s contribution to the state of the overall data structure, e.g., the set of keys contained in $n$.
+The interface provided by a data structure implementation includes a predicate of the form $\node{n}{I}{C}$, where $n$ is the node itself (i.e., a pointer to the node data structure), $I$ is $n$'s flow interface, and $C$ is $n$'s contribution to the state of the overall data structure, e.g., the set of keys contained in $n$.
 Data structure operations such as \lstinline{insert}, \lstinline{lookup}, and \lstinline{findNext} are specified in terms of the $\mathsf{node}$ predicate.
 %example?
+explain flow interfaces further?
 
@@
 The search structure template approach uses concurrent separation logic to specify and prove pre- and postconditions for the template and data structure functions.
-Krishna et al.~implemented their framework in Iris~\cite{iris}, a language-independent CSL framework built in the Coq proof assistant~\cite{coq}, with flexible support for ghost state, invariants, and atomic specifications.
+Krishna et al.~implemented their framework in Iris~\cite{iris}, a language-independent CSL framework built in the Rocq proof assistant~\cite{rocq}, with flexible support for ghost state, invariants, and atomic specifications.
 This makes it easy to describe the effects of concurrency control functions independently of the underlying data structure, e.g., as ``this function atomically finds a node that contains the target key.'' %forward reference to where we show the specs?
@@
 To apply the template approach to real-world code, we instead use the Verified Software Toolchain (VST)~\cite{plcc}, a separation-logic-based verifier for C programs.
-VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, allowing it to guarantee that proved properties will hold on compiled code.
+VST is also built in Rocq and is connected to the CompCert verified C compiler~\cite{compcert}, allowing it to guarantee that proved properties will hold on compiled code.
 Recent work on VST~\cite{iris-vst-arxiv} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications.
@@
 
-\section{Specifying the Template Interface}
+\section{Specifying Data Structures and Templates as Interfaces}
 Our verification process for concurrent search structure templates follows the same modular architecture as the code itself:
@@
 As is typical when defining interfaces, the key challenge is to define specifications for the data structure functions that give enough information to verify any template, but are general enough to be satisfied by any data structure.
-...
+Each of our interfaces is implemented as a typeclass in Rocq, which makes it easy to do proofs over a generic instance of the interface.
+
+\subsection{Data Structure Interface}
+
+The data structure interface is built around an abstract predicate $\node{n}{I}{C}$ representing a node in the data structure, where $n$ is the concrete pointer to the node, $I$ is the flow interface for the node (an abstraction of its position in the data structure), and $C$ is the contents of the note as a map from keys to values.
+In practice, $\mathsf{node}$ will be implemented as a combination of concrete points-to predicates for the physical node in memory, and ghost state representing the node's contribution to the abstract state of the data structure.
+%discuss how this relates to the acsys proofs?
+We then specify the key data structure operations as follows: %fill these in
+\begin{mathpar}
+\texttt{findNext}
+
+\texttt{insertOp}
+
+\texttt{lookupOp}
+\end{mathpar}
+
+\subsection{Template Interface}
+The template interface includes two abstract predicates: \treerep(p, C), which describes the abstract state of the entire concurrent data structure at pointer $p$ as a key-value map $C$, and $\mdentry{n}{I}{C}$, %more arguments
+which represents the combination of a node and the metadata attached to it in the template.
+Prior work does not clearly distinguish between $\mathsf{node}$ and $\mathsf{md\_node}$---for instance, \citet{templates} assume that the $\mathsf{node}$ predicate includes the lock and, in the give-up template, the range for the node---but this distinction is necessary for truly modular specification: the data structure interface should not include concurrency metadata, since it is both concurrency-specific and different for each template instance.
+Thus, while data structure functions act on $\mathsf{node}$s, the template functions are defined entirely at the level of $\mathsf{md\_node}$s.
+
+The template operations are specified as follows: %fill these in
+\begin{mathpar}
+\texttt{traverse}
+
+\texttt{insertHelper}
+
+\texttt{lookupHelper}
+\end{mathpar}
+Each top-level operation will be implemented with a combination of \texttt{traverse}, which traverses the data structure to find the $\mathsf{md\_node}$ where a key belongs, and one of the helper functions, which completes the chosen operation given the target $\mathsf{md\_node}$.
+The helper functions are responsible for both performing the actual data structure operation (using \texttt{insertOp} or \texttt{lookupOp} from the data structure interface) and updating the metadata (creating new locks, modifying ranges, etc.) to create a consistent concurrent data structure for the new state.
+
+\subsection{Top-Level Operations}
+Given instances of the data structure and template interfaces, we can define and verify the top-level \texttt{insert} and \texttt{lookup} operations for concurrent search structures.
+(implementations and proofs here)
+
+In Rocq, these proofs are parameterized by instances of the $\mathsf{DataStructure}$ and $\mathsf{Template}$ classes.
+It is precisely this that gives us true modularity: we know that for \emph{any} data structure implementation meeting the data structure interface, and \emph{any} template meeting the template interface, we can instantiate these parameterized proofs and obtain verified \texttt{insert} and \texttt{lookup} functions.
+In other words, this guarantees that given $m$ data structure implementations and $n$ template implementations, we can immediately and with no additional effort obtain $m \times n$ verified concurrent data structures.
 
@@
 The framework we have presented fulfills the true promise of the concurrency template approach: each data structure instance is implemented and verified without any awareness of concurrency or metadata, each template instance is implemented and proved without any knowledge of the implementation of the data structure, and we obtain verified concurrent data structures immediately by instantiating our top-level theorems with any combination of data structure and template instances.
-We accomplished this by careful attention to which components belong to the data structure and which to the template, and by working to disentangle the two in both the C implementation and the Coq specification and proof.
+We accomplished this by careful attention to which components belong to the data structure and which to the template, and by working to disentangle the two in both the C implementation and the Rocq specification and proof.
 Our interfaces are general enough that it should be easy to add more data structures and templates, including concurrency patterns that do not depend on locks (e.g., optimistic concurrency control with fine-grained atomics).
@@
 \begin{acks}
-We thank Nisarg Patel, ...
+We thank Nisarg Patel for detailed discussions about the original search structure template proofs.
 This work was partially funded by the National Science Foundation under the award ???.

### 2025-04-28 concurrency/template paper/paper.tex
sketched implementation layout

@@
 \section{Coding Search Structure Templates}
+\label{implementation}
 \begin{figure}
@@
 \section{Template Instances: Give-Up and Lock Coupling}
-
-\section{Evaluation}
 
@@
 
+\section{Implementation and Evaluation}
+The code for our concurrent data structures is written in C, and our specifications and proofs are written in the Verified Software Toolchain (VST)~\cite{plcc}.
+Using C forces us to confront the implementation challenges described in section~\ref{implementation}, which we might accidentally circumvent in a core calculus.
+VST allows us to prove separation logic specifications of C programs in Rocq, and newer versions of VST~\cite{iris-compcert} are built on Iris, allowing us to directly reuse the implementation of flow interfaces from \citet{templates}.
+Our C code and our Rocq proofs follow a common structure that ensures modularity.
+%diagram:                       %should we put some of this structural stuff earlier?
+%                   data_structure.h
+%             /              |                 \
+%         bst.c          list.c             template.h
+%                                       /            |             \
+%                                give-up.c   coupling.c   template.c
+
+The files \texttt{data\_structure.h} and \texttt{template.h} define the interfaces for data structures and templates respectively, and the corresponding VST files define the typeclasses for these interfaces, including the specification of each interface function.
+\texttt{bst.c} and \texttt{list.c} implement the data structure interface, and \texttt{give-up.c} and \texttt{coupling.c} implement the template interface.
+Finally, \texttt{template.c} implements the top-level operations, relying on the interfaces for data structures and templates but not on any details of their implementation (and the corresponding VST file verifies the operations with the interface typeclasses as parameters).
+Importantly, there is no file in either C or Rocq that is specific to a combination of data structure and template, such as BST+give-up or list+coupling: we can freely combine any data structure and concurrency template and obtain a working concurrent data structure with no further effort, simply by compiling \texttt{template.c} along with the chosen data structure and template file.
+
+(LoC/LoP numbers for each part)
+(comparison of proof size to acsys proofs and/or CPP paper?)
+
 \section{Conclusion and Future Work}
